### 2021-07-16 doc/paper.Rmd
Initial commit

---
title: Fast and Flexible Spatial and Spatiotemporal Generalized Linear Mixed-Effects Models with sdmTMB
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
  $^2$Other addresses\n"
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage[left]{lineno}
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
editor_options:
  markdown:
    wrap: sentence
abstract: >
  The sdmTMB package implements spatiotemporal predictive-process GLMMs (Generalized Linear Mixed Effect Models) using TMB, INLA, and Gaussian Markov random fields. One common application is for spatial or spatiotemporal species distribution models (SDMs). ...
---

```{r, echo = FALSE, include = FALSE, message = FALSE, warnings = FALSE}
library(here)
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warnings = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Main body here.

-   [@anderson2019]
-   [@maureaud2021]
-   [@anderson2020]
-   [@osgood-zimmerman2021]
-   [@breivik2021]
-   [@monnahan2021]
-   [@shelton2014]
-   [@lindgren2011]
-   [@latimer2009]

Content:

-   outline the general need for this type of modelling

-   outline the need for this type of package to make the models accessible and rapid to iterate on

-   emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference

-   outline the philosophy and how the functions work together

    -   modular
    -   familiar (similar to glm, lme4, glmmTMB, mgcv, etc.)
    -   separate fitting and predicting
    -   avoid expensive calculations unless needed

-   table comparison to related packages

-   case studies

    -   an SDM with time-varying depth?
    -   an index calculation?
    -   stitching surveys?

-   how it achieves the steps Jim described [@thorson2019] in what a VAST-related package has to be able to do (and how VAST does it vs. how sdmTMB does it)

-   speed comparisons (INLA [inlabru?], VAST, mgcv, glmmTMB without predictive process?)

-   intended audience: ecology, but much broader too

-   self simulation testing --- appendix

-   package comparison fitting --- appendix

\clearpage

## References

::: {#refs}
:::

```{=tex}
\clearpage
\beginsupplement
```
# Supplemetal Materials

### 2021-07-20 doc/paper.Rmd
Typo
Add to paper notes + very rough comparison stuff

@@
 ---
-title: Fast and Flexible Spatial and Spatiotemporal Generalized Linear Mixed-Effects Models with sdmTMB
+title: "`sdmTMB`: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
 author: "|
@@
   \usepackage[left]{lineno}
+  \linenumbers
   \newcommand{\beginsupplement}{
@@
     wrap: sentence
-abstract: >
-  The sdmTMB package implements spatiotemporal predictive-process GLMMs (Generalized Linear Mixed Effect Models) using TMB, INLA, and Gaussian Markov random fields.
+abstract: |
+    | \noindent Point 1: set the context for and purpose of the work; \smallskip 
+    | Point 2: indicate the approach and methods; \smallskip 
+    | Point 3: outline the main results; \smallskip 
+    | Point 4: identify the conclusions and the wider implications. \smallskip 
+---
+
+<!-- # The sdmTMB package implements spatiotemporal predictive-process GLMMs (Generalized Linear Mixed Effect Models) using TMB, INLA, and Gaussian Markov random fields.
 One common application is for spatial or spatiotemporal species distribution models (SDMs).
-...
----
+-->
 
@@
 
-Main body here.
+```{=html}
+<!--
+
+Application: Short (no more than 3000-4000 words total) descriptions of new software that are intended to describe and promote the software as well as act as a citeable resource for developers.
+Application papers are made free to access for all readers upon publication in order to encourage uptake of the methodologies they describe.
+Uploading a package to a site such as [**CRAN**](http://cran.r-project.org/) or [**sourceforge** ](http://sourceforge.net/)in advance is not considered prior publication and will not hinder your submission to MEE.
+View some example Applications [**here**](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/features/applicationpapers).
+
+<https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/features/applicationpapers>
+
+-   Single column, double line spaced
+
+-   Within the word count (no more than 7000-8000 words for Standard Articles, 3000-4000 words for Applications and Practical Tools)
+
+-   Continuous line and page numbering throughout
+
+-   Clearly defined manuscript structure as standard: Author details, Abstract (must be numbered according to [**Manuscript Specifications**](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210x/author-guidelines#MS_spec)), Keywords, Introduction, Materials and Methods, Results, Discussion, Figures and Tables with captions
+
+-   Figures and Tables can be embedded within the text where referenced to facilitate reviewing
+
+-   Author Contributions ([[if \>1 author -- see Authorship Policy]{.ul}](https://besjournals.onlinelibrary.wiley.com/hub/editorial-policies#authorship)).
+
+-   Statement of where you *intend* to [**archive your data**](https://besjournals.onlinelibrary.wiley.com/hub/data_archiving_policy)
+
+Title Page should include:
+
+A concise and informative title. Do not include the authorities for taxonomic names.
+A list of all authors' names with names and addresses of Institutions.
+The name, address and e-mail address of the correspondence author.
+A running headline of not more than 45 characters.
+
+Abstract
+
+The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:
+
+Key-words: A list in alphabetical order not exceeding eight words or short phrases.
+The most important key-words should appear in the title and the abstract as well as the key-word list.
+More advice on selecting good keywords can be found here.
+
+Introduction
+
+This should state the reason for doing the work, the nature of the hypothesis or hypotheses under consideration, and should outline the essential background.
+
+Materials and Methods
+
+Include sufficient details for the work to be repeated.
+Where specific equipment and materials are named, the manufacturer’s details (name, city and country) should be given so that readers can trace specifications by contacting the manufacturer.
+Where commercially available software has been used, details of the supplier should be given in brackets or the reference given in full in the reference list.
+Do not describe or refer to commonplace statistical tests in this section but allude to them briefly in Results.
+
+Results
+
+State the results and draw attention in the text to important details shown in tables and figures.
+
+Discussion
+
+This should point out the significance of the results in relation to the reasons for doing the work, and place them in the context of other work.
+
+Conclusions (optional)
+
+Acknowledgements (optional)
+
+A brief statement acknowledging collaborators and research assistants who do not meet the criteria for authorship described above, or acknowledging funding sources, providing relevant permit numbers (including institutional animal use permits), or giving recognition to nature reserves or other organizations that made the work possible.
+
+Conflict of Interest statement (required for all article types)
+
+-->
+```
+
+# Introduction
 
@@
 -   [@latimer2009]
+
+# Examples of sections:
+
+# Materials and Methods
+
+# Description
+
+# Model
+
+# Application Examples
+
+# Implementation and Availability
+
+# Performance
+
+# Discussion
+
+# Conclusion
+
+Optional
+
+# Acknowledgements
+
+# Conflict of Interest statement
+
+# Author Contributions
+
+<!-- 
+All submissions with more than one author must include an Authors’ contributions statement. All persons listed as authors on a paper are expected to meet ALL of the following criteria for authorship:
+
+substantial contributions to conception and design, or acquisition of data, or analysis and interpretation of data, or drafting the article or revising it critically for important intellectual content;
+final approval of the version to be published;
+agreement to be accountable for the aspects of the work that they conducted and ensuring that questions related to the accuracy or integrity of any part of their work are appropriately investigated and resolved.
+Acquisition of funding, provision of facilities, or supervising the research group of authors without additional contribution are not usually sufficient justifications for authorship.
+The statement should include an explanation of the contribution of each author.
+We suggest the following format for the Authors’ contributions statement:
+
+AB and CD conceived the ideas and designed methodology; CD and EF collected the data; EF and GH analysed the data; AB and CD led the writing of the manuscript.
+All authors contributed critically to the drafts and gave final approval for publication.
+-->
+
+# Data Availability
+
+<!--
+To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
+For example, “Data available from the Dryad Digital Repository http://dx.doi.org/10.5061/dryad.41qh7 (Kiere & Drummond 2016).” When a DOI is available for the data, the full data citation should also be given in the reference list (see below).
+For further information see our data archiving policy.
+-->
 
@@
 
-## References
+# References
 
@@
 ```
-# Supplemetal Materials
+
+# Supplemental Materials

### 2021-07-20 analysis/compare-glmmTMB.R

@@ -1,0 +1,86 @@
+library(glmmTMB)
+library(sdmTMB)
+
+# all pcod2011 data with Matern far too slow with glmmTMB (45 minutes, not really converged)
+# try a version on just 2017 as an example
+
+(pos <- numFactor(d$X, d$Y))
+parseNumLevels(levels(pos))
+# d$times <- numFactor(d$year)
+# levels(d$times)
+d$pos <- pos
+d$group <- factor(rep(1, nrow(d)))
+
+tictoc::tic("glmmTMB")
+d <- pcod
+d2017 <- subset(d, year == 2017)
+d2017$pos <- numFactor(d2017$X, d2017$Y)
+d2017$group <- factor(rep(1, nrow(d2017)))
+m_glmmTMB <- glmmTMB(present ~ 1 + mat(pos + 0 | group), data = d2017,
+  family = binomial(link = "logit"), verbose = TRUE,
+  map = list(theta = factor(c(1, 2, NA))),
+  start = list(theta = c(log(1.2), log(30), log(1))) # SD, range, smoothness v (fixed at 1 usually)
+)
+tictoc::toc()
+
+mesh <- make_mesh(d2017, c("X", "Y"), cutoff = 1.5)
+mesh$mesh$n
+m <- sdmTMB(present ~ 1, data = d2017, spde = mesh,
+  family = binomial(link = "logit"), silent = FALSE
+  # priors = sdmTMBpriors(
+  #   matern_st = pc_matern(range_gt = 5, range_prob = 0.05, sigma_lt = 5, sigma_prob = 0.05),
+  #   ar1_rho = normal(0, 1),
+  #   b = normal(0, 100)
+)
+
+summary(m_glmmTMB)
+# confint(m_glmmTMB, "theta")
+m_glmmTMB$sdr
+# exp(theta(0)) = SD?
+# exp(theta(1)) /* range */,
+# exp(theta(2)) /* smoothness */) );
+m
+
+tidy(m, "ran_pars", conf.int = TRUE)
+s <- as.list(m_glmmTMB$sdr, "Estimate")
+exp(s$theta)
+m
+# https://kaskr.github.io/adcomp/group__special__functions.html#ga4bdabb9b3e59feb7c552b30c46c8959c
+
+
+library(INLA)
+
+tictoc::tic("INLA")
+d <- d2017
+d$i_year <- 1L
+mesh_inla <- inla.mesh.2d(cbind(d$X, d$Y), max.edge = c(10, 20))
+plot(mesh_inla)
[…]

### 2021-07-20 analysis/compare-inla.R

@@ -24,7 +24,7 @@ sdat <- inla.stack(
 # h.spec <- list(theta = list(prior = 'pccor1', param = c(0, 0.9)))
 h.spec <- list(theta = list(prior = 'normal', param = c(0, 1)))
 formulae <- y ~ 0 + year_factor + f(i, model = spde, group = i.group,
-  control.group = list(model = 'ar1', hyper = h.spec))
+  control.group = list(model = 'iid', hyper = h.spec))
 # PC prior on the autoreg. param.
 # prec.prior <- list(prior = 'pc.prec', param = c(1, 0.01))
 m_inla <- inla(formulae,
@@ -49,7 +49,7 @@ summary(m_inla)
 
 tictoc::tic("sdmTMB")
 m <- sdmTMB(present ~ 0 + as.factor(year), data = d, spde = mesh, time = "year",
-  fields = "AR1", include_spatial = FALSE, family = binomial(link = "logit"),
+  fields = "IID", include_spatial = FALSE, family = binomial(link = "logit"),
   reml = FALSE,
   priors = sdmTMBpriors(
     matern_st = pc_matern(range_gt = 5, range_prob = 0.05, sigma_lt = 5, sigma_prob = 0.05),

### 2021-07-20 analysis/compare-mgcv.R

@@ -1,0 +1,27 @@
+library(sdmTMB)
+library(mgcv)
+
+d <- pcod_2011
+mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 5)
+mesh$mesh$n
+tictoc::tic("mgcv")
+d$year_factor <- as.factor(d$year)
+m_mgcv <- gam(present ~ 0 + year_factor + s(X, Y, by = year_factor, k = 60), # FIXME: k has huge influence on timing...
+  data = d, family = binomial(link = "logit"))
+# m_mgcv
+# mgcv::plot.gam(m_mgcv)
+tictoc::toc()
+p1 <- predict(m_mgcv)
+
+tictoc::tic("sdmTMB")
+m <- sdmTMB(present ~ 0 + as.factor(year), data = d, spde = mesh, time = "year",
+  fields = "AR1", include_spatial = FALSE, family = binomial(link = "logit"),
+  priors = sdmTMBpriors(
+    matern_st = pc_matern(range_gt = 5, range_prob = 0.05, sigma_lt = 5, sigma_prob = 0.05),
+    ar1_rho = normal(0, 1),
+    b = normal(0, 100)
+  ))
+tictoc::toc()
+p2 <- predict(m)
+
+plot(p1, p2$est);abline(a = 0, b = 1)

### 2021-07-20 doc/paper.Rmd
add more notes

@@
 ---
+
+<!-- # For submission to MEE as an Application paper -->
 
@@
 
+# Tables
+
+1. Comparison to related software?
+
+- VAST
+- INLA
+- mgcv
+- glmmTMB
+- glmmfields?
+- other stuff on CRAN, spBayes etc.
+
+# Figures
+
+1. Flow diagram of functions? Not needed?
+1. Speed with various knots and/or spatial complexity in comparison to INLA + VAST + ...?
+1. Example application 1 - SDM, show mesh(?) and/or raw data, show predictions and SD in space,
+1. Example application 2 - population index with stitched survey?
+
 Content:

### 2021-08-04 doc/paper.Rmd
Add some VAST notes

@@
 -   how it achieves the steps Jim described [@thorson2019] in what a VAST-related package has to be able to do (and how VAST does it vs. how sdmTMB does it)
+    -   "VAST represents four main design principles.
+Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include."
+    -   Area weighting (check, helper functions to sum up weighted grid cells)
+    -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`)
+    -   Condition on missing covariates (check, random fields)
+    -   Bridge between univariate and multivariate applications (no! should we?)
 

### 2021-08-13 doc/paper.Rmd
couple ideas of things to highlight

@@
 # Description
+- maybe some overview of spatial models, either here in intro. Contrast with GAMs, and highlight explicit estimation of flexible spatial covariance functions
 
@@
 
+Highlights?
+- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances) 
+- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
+- Spatiotemporal models (RW/AR v IID)
+- Time varying parameters
+- Non-linear effects (smooth splines, threshold/breakpoint functions)
+- Variance estimation and derived quantities (spatial range, area occupied etc)
+
 # Application Examples
 
+- Thinking of application with broad interest or a dataset that is pretty well known. Trawl surveys are great obv, and there's lots of examples worked out
+- gorillas dataset from spatstat/inlabru: https://rdrr.io/cran/spatstat.data/man/gorillas.html -- only prob is it's not time-varying
+- Spatial rainfall dataset from INLA? https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html
+- Simulating a dataset could be used to highlight sim() function 
+- 
 # Implementation and Availability
 
+- CRAN before?  
+- 
 # Performance
+
+- Probably some discussion of speed. Comparison of one of the models in the application above to INLA? Or other software, e.g. Stan -- though that's not as fair :)
+- Probably not extensive space for simulation testing, but could do some minimal example and point to vignettes
 

### 2021-08-27 doc/paper.Rmd
some initial equations and demos

@@
 
+Though `sdmTMB` has been developed with spatial modeling as a primary focus, all of the statistical models fall under the class of generalized linear mixed models (GLMMs).
+Non-spatial GLMMs are available in several widely used packages, including `glmmTMB` (Brooks et al.
+2017) and `lme4` (Bates et al.
+2015), and to allow comparison we also allow non-spatial models to be constructed with `sdmTMB`.
+The general form of a GLMM is  $g(\bf{\mu}) = \bf{Xb} + \bf{Zu}$, where the inverse link function $g(x)$ relating the mean $\bf{\mu}$ to fixed effect predictors in the design matrix $\mathbf{X}$ with estimated coefficients $\bf{b}$ and random effects with the design matrix $\bf{Z}$ with estimated coefficients $\bf{u}$.
+The matrix $\bf{Z}$ allows random effects to be structured (e.g.
+intercepts varying by group) or unstructured, with $\bf{u}$ representing random variation.
+
+For clarity, we partition the random effects $\bf{Zu}$ into a number of more interpretable components, each of which can optionally be included in the model.
+Using our notation, the full `sdmTMB` model can be written as $$g(\mu_{s,t}) = X_{s,t}\bf{b} + \omega_{s} + \epsilon_{s,t} + \bf{\alpha_{g}} + X_{s,t}\gamma_{t} + X_{s,t}\zeta_{s}$$.
+In this form, $\omega_{s}$ represents a random spatial component (spatial field with mean 0), $\epsilon_{s,t}$ represents a spatiotemporal component (spatial field with mean 0), $\alpha_{g}$ represents random intercepts by group $\alpha_{g}\sim N(0,\sigma_\alpha)$, $\gamma_{t}$ represents time-varying coefficients modeled as a random walk, $\gamma_{t} \sim N(\gamma_{t-1},\sigma_\gamma)$, and $\zeta_{s}$ represents spatially varying coefficients (modeled as a spatial field).
+All of the spatial fields in this model are constructed using Gaussian random fields, where the random effects are structured as multivariate normal: $\omega_s \sim MVN(\bf{0},\bf{\Sigma}_\omega)$, $\epsilon_{s,t} \sim MVN(\bf{0},\bf{\Sigma}_{\epsilon})$, and $\zeta_s \sim MVN(\bf{0},\bf{\Sigma}_\zeta)$.
+By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in `sdmTMB` allow them to be modeled as a random walk or AR(1) process.
+Additional flexibility is included in relaxing assumptions about linear fixed effects $\bf{Xb}$; sdmTMB allows any covariates to be modeled as smooth functions using the same generalized additive model syntax in `mgcv` (Wood 2003), or modeled as threshold (e.g.
+hockey stick models; Barrowman et al.
+2003).
+
+[paragraph about spde and gaussian process / Matern covariance here]
+
 Highlights?
@@
 
-- Thinking of application with broad interest or a dataset that is pretty well known. Trawl surveys are great obv, and there's lots of examples worked out
-- gorillas dataset from spatstat/inlabru: https://rdrr.io/cran/spatstat.data/man/gorillas.html -- only prob is it's not time-varying
-- Spatial rainfall dataset from INLA? https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html
-- Simulating a dataset could be used to highlight sim() function 
-- 
+*EW: These are just placeholders -- not sure what order is best or what to highlight in very limited space*
+
+As a first example, we use a spatial dataset (`parana`) from the `geoR` package (Ribeiro Jr and Diggle 2018) that describes average May-June rainfall recorded at 143 stations in the state of Paraná, Brazil.
+This dataset is also well-known because it is used as an example of spatiotemporal modeling with INLA (cite INLA) book.
+Because the dataset doesn't include covariates and only includes a single year, we can model rainfall using an intercept and spatial component, $g(\mu_{s,t}) = b_{0} + \omega_{s}$.
+We assume the observation model for the rainfall data is normally distributed.
+The first step in fitting a model with `sdmTMB` is constructing the mesh to approximate the spatial process.
+This step can be done with the `make_mesh` function, which accepts the data (coordinates) and at minimum, the minimum allowed distance between points (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
+Custom built meshes with INLA can also be passed in (additional examples of this are included in the SI).
+For example, an initial mesh with a cutoff distance of 0.1km could be made with
+
+```{r eval=FALSE, echo=TRUE}
+rain_spde = make_mesh(data = parana, c("east","north"), cutoff=0.1)
+```
+
+Second, we can fit the model using TMB with the `sdmTMB` function, whose formula syntax should be familiar to most R users,
+
+```{r eval = FALSE, echo=TRUE}
+fit = sdmTMB(data ~ 1, data=parana, spde = rain_spde)
+```
+
+As a second example, we use `sdmTMB` to analyze a 2-year dataset of Eurasian cranes (*Grus grus*) from England, UK (Soriano-Redondo et al.
+2019).
+In the original analysis, Soriano-Redondo et al.
+2019 used log Gaussian Cox process models to jointly model crane presence--absence and the spatiotemporal distribution of wetlands.
+We use this dataset as an example of modeling presence--absence data with sdmTMB, and only model the occurence of cranes given covariates.
+With any multi-year dataset, a common question of interest is often whether observations can be best described by spatial variation, or spatial and spatiotemporal variation.
+With two years of data, we fit two models with `sdmTMB`: a model with a spatial component but no spatiotemporal variation, and a model with only spatiotemporal variation (allowing each year to have a different spatial field).
+We included the same fixed effect covariates in both models: different intercepts by year, and two covariates found to be important by Soriano-Redondo et al.
+2019 (wetland area, perimeter to area ratio).
+As a coarse comparison of data support, we compare the Akaike's Information Criterion (Akaike 1973) from each model.
+The model with a shared spatial field appeared to be more supported ($\Delta AIC$=25.23).
+The sdmTMB package allows for additional model comparisons, including k-fold cross validation with the `sdmTMB_cv` function.
+
+Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?
+
 # Implementation and Availability
@@
 
-- Probably some discussion of speed. Comparison of one of the models in the application above to INLA? Or other software, e.g. Stan -- though that's not as fair :)
+- Probably some discussion of speed.
+Comparison of one of the models in the application above to INLA?
+Or other software, e.g.
+Stan -- though that's not as fair :).
+[EW: at least for the rainfall example I found inlabru to be about twice as fast as sdmTMB]
 - Probably not extensive space for simulation testing, but could do some minimal example and point to vignettes

### 2021-08-27 crane_example.R

@@ -1,0 +1,31 @@
+devtools::install_github("cmjt/lgcpSPDE")
+library(lgcpSPDE)
+# https://onlinelibrary.wiley.com/doi/epdf/10.1111/ecog.03771
+# https://github.com/cmjt/examples/blob/master/species_distribution.md
+data(cranes)
+locs <- as.matrix(cranes[,2:3]) ## matrix of wetland epicentre locations
+mesh <- inla.mesh.2d(loc = locs, cutoff = 0.15, max.edge = c(0.15,2))
+
+covariates <- data.frame(Area_sc = scale(cranes$Area), PA_ratio_sc = scale(cranes$PA_ratio),
+                         Wet_density_nosea_sc = scale(cranes$Wet_density_buf_NoSea),
+                         Urb_density_nosea_sc = scale(cranes$Urb_density_buf_NoSea))
+cranes <- cbind(cranes, covariates)
+
+# use sdmTMB to fit the model -- spatial only
+cranes$Year = as.factor(cranes$Year)
+fit1 = sdmTMB(mark ~ -1 + Year + Area_sc + PA_ratio_sc,
+             data=cranes,
+             time="Year",
+             spde = make_mesh(data = cranes, c("Lon","Lat"),
+                              mesh=mesh), family = binomial(),
+             spatial_only = TRUE)
+
+# fit second model with different fields by year
+fit2 = sdmTMB(mark ~ -1 + Year + Area_sc + PA_ratio_sc,
+             data=cranes,
+             time="Year",
+             spde = make_mesh(data = cranes, c("Lon","Lat"),
+                              mesh=mesh), family = binomial(),
+             include_spatial = FALSE)
+
+round(AIC(fit1) - AIC(fit2),2)

### 2021-08-27 rainfall_example.R

@@ -1,0 +1,46 @@
+library(INLA)
+library(geoR)
+library(sdmTMB)
+library(matlab) # just for timing
+library(inlabru)
+
+data(parana)
+# from inla book, https://www.paulamoraga.com/book-geospatial/sec-geostatisticaldatatheory.html#spatial-modeling-of-rainfall-in-paran%C3%A1-brazil
+parana = as.data.frame(parana)
+coo <- as.matrix(parana[,c("east","north")])
+bnd <- inla.nonconvex.hull(coo)
+meshb <- inla.mesh.2d(
+  boundary = bnd, offset = c(50, 100),
+  cutoff = 1, max.edge = c(30, 60)
+)
+
+# use sdmTMB to fit the model
+fit = sdmTMB(data ~ 1,
+             data=parana,
+             spde = make_mesh(data = parana, c("east","north"),cutoff=0.1))
+
+# use sdmTMB to fit the model, using custom mesh above
+fit = sdmTMB(data ~ 1,
+             data=parana,
+             spde = make_mesh(data = parana, c("east","north"),
+                                        mesh=meshb))
+
+# Add PC priors
+fit = sdmTMB(data ~ 1,
+             data=parana,
+             spde = make_mesh(data = parana, c("east","north"),
+                              mesh=meshb),
+             priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 5, range_prob = 0.05,
+                                                        sigma_lt = 10, sigma_prob = 0.05)))
+
+
+# use inlabru to fit the same model/mesh
+matern <- inla.spde2.pcmatern(meshb, prior.sigma=c(5,0.05),
+                              prior.range=c(10,0.05))
+components <- data ~ field(main=coordinates,model=matern)
+
+# add coordinates for dataframe
+parana <- as.data.frame(parana)
+coordinates(parana) = c("east","north")
+
+fit2 <- bru(components, data = parana)

### 2021-10-13 doc/paper.Rmd
Develop model equations

@@
   \usepackage[left]{lineno}
+  \usepackage{bm}
   \linenumbers
@@
 
-Though `sdmTMB` has been developed with spatial modeling as a primary focus, all of the statistical models fall under the class of generalized linear mixed models (GLMMs).
-Non-spatial GLMMs are available in several widely used packages, including `glmmTMB` (Brooks et al.
+Though `sdmTMB` has been developed with spatial or spatiotemporal modelling as a primary focus, all of the statistical models it fits fall under the class of generalized linear mixed models (GLMMs).
+GLMMs are available in several widely used packages, including `glmmTMB` (Brooks et al.
 2017) and `lme4` (Bates et al.
 2015), and to allow comparison we also allow non-spatial models to be constructed with `sdmTMB`.
-The general form of a GLMM is  $g(\bf{\mu}) = \bf{Xb} + \bf{Zu}$, where the inverse link function $g(x)$ relating the mean $\bf{\mu}$ to fixed effect predictors in the design matrix $\mathbf{X}$ with estimated coefficients $\bf{b}$ and random effects with the design matrix $\bf{Z}$ with estimated coefficients $\bf{u}$.
-The matrix $\bf{Z}$ allows random effects to be structured (e.g.
-intercepts varying by group) or unstructured, with $\bf{u}$ representing random variation.
 
-For clarity, we partition the random effects $\bf{Zu}$ into a number of more interpretable components, each of which can optionally be included in the model.
-Using our notation, the full `sdmTMB` model can be written as $$g(\mu_{s,t}) = X_{s,t}\bf{b} + \omega_{s} + \epsilon_{s,t} + \bf{\alpha_{g}} + X_{s,t}\gamma_{t} + X_{s,t}\zeta_{s}$$.
-In this form, $\omega_{s}$ represents a random spatial component (spatial field with mean 0), $\epsilon_{s,t}$ represents a spatiotemporal component (spatial field with mean 0), $\alpha_{g}$ represents random intercepts by group $\alpha_{g}\sim N(0,\sigma_\alpha)$, $\gamma_{t}$ represents time-varying coefficients modeled as a random walk, $\gamma_{t} \sim N(\gamma_{t-1},\sigma_\gamma)$, and $\zeta_{s}$ represents spatially varying coefficients (modeled as a spatial field).
-All of the spatial fields in this model are constructed using Gaussian random fields, where the random effects are structured as multivariate normal: $\omega_s \sim MVN(\bf{0},\bf{\Sigma}_\omega)$, $\epsilon_{s,t} \sim MVN(\bf{0},\bf{\Sigma}_{\epsilon})$, and $\zeta_s \sim MVN(\bf{0},\bf{\Sigma}_\zeta)$.
-By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in `sdmTMB` allow them to be modeled as a random walk or AR(1) process.
-Additional flexibility is included in relaxing assumptions about linear fixed effects $\bf{Xb}$; sdmTMB allows any covariates to be modeled as smooth functions using the same generalized additive model syntax in `mgcv` (Wood 2003), or modeled as threshold (e.g.
-hockey stick models; Barrowman et al.
+We use the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time.
+The general form of a GLMM is 
+$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
+with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
+<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
+For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
+The full `sdmTMB` model is
+$$
+g(\mu_{s,t}) = 
+\bm{X}_{1,\bm{s},t} \bm{\beta} + 
+\bm{\alpha}_g +
+\bm{X}_{2,\bm{s},t} \gamma_t + 
+\bm{X}_{3,\bm{s},t} \zeta_{\bm{s}} +
+\omega_{\bm{s}} + 
+\epsilon_{\bm{s},t},
+$$
+where
+
+* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
+* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
+* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
+* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
+* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
+* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
+* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
+
+The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function. 
+The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
+(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
+<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
+where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
+a scaling parameter that is estimated (REF). 
+The parameter $\nu$ controls the smoothness of the covariance function. 
+We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+
+By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in `sdmTMB` allow them to be modelled as a random walk or first-order autoregressive (AR1) process.
+We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: `sdmTMB` allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in `mgcv` (Wood 2003), or to be modelled with threshold shapes (e.g., hockey stick models; Barrowman et al.
 2003).
@@
 Highlights?
+
 - Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances) 

### 2021-10-18 doc/paper.Rmd
Add example

@@
 
-```{r, echo = FALSE, include = FALSE, message = FALSE, warnings = FALSE}
+```{r, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
 library(here)
@@
 knitr::opts_chunk$set(
-  echo = FALSE, message = FALSE, warnings = FALSE,
+  echo = FALSE, message = FALSE, warning = FALSE,
   collapse = TRUE,
@@
 $$
-g(\mu_{s,t}) = 
+g(\mu_{\bm{s},t}) = 
 \bm{X}_{1,\bm{s},t} \bm{\beta} + 
@@
 # Application Examples
+
+```{r libs, echo = TRUE}
+library(sdmTMB)
+library(ggplot2)
+```
+
+```{r}
+theme_set(ggsidekick::theme_sleek())
+```
+
+First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
+The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between v vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
+Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in (examples included in `?make_mesh`).
+For example, an mesh with a cutoff distance of 10km can be made with
+
+```{r mesh, echo = TRUE, cache=FALSE}
+mesh <- make_mesh(pcod_2011, xy_cols = c("X", "Y"), cutoff = 10)
+```
+
+```{r plot-mesh, echo = TRUE, fig.height=5, fig.width=5, out.width="3.25in"}
+plot(mesh)
+```
+
+```{r fit-pcod, cache=TRUE, echo = TRUE}
+fit <- sdmTMB(density ~ 0 + as.factor(year) + s(depth), data = pcod_2011,
+ time = "year", spde = mesh, family = tweedie(link = "log")
+)
+```
+
+```{r print-fit, echo = TRUE}
+print(fit)
+```
+
+```{r tidy, eval=TRUE, echo=TRUE}
+tidy(fit, conf.int = TRUE)
+```
+
+```{r, eval=TRUE, echo=TRUE}
+tidy(fit, effects = "ran_pars", conf.int = TRUE)
+```
+
+```{r predict, cache=TRUE, echo = TRUE}
+pred <- predict(fit)
+```
+
+```{r qqnorm, cache=TRUE, fig.width=4, fig.height=4, out.width="3.5in", echo = TRUE}
+pred$resid <- residuals(fit)
+# qqnorm(pred$resid);qqline(pred$resid) # not shown for space
+```
+
+```{r plot-resids, echo = FALSE, eval=FALSE}
+ggplot(subset(pred, year == 2017), aes(X, Y, colour = resid)) +
+    geom_point() + facet_wrap(~year) +
+    coord_fixed() + scale_colour_gradient2()
+```
+
+```{r predict-newdata, cache=TRUE, echo = TRUE}
+qcs_grid_2011 <- subset(qcs_grid, year >= 2011)
+pred_qcs <- predict(fit, newdata = qcs_grid_2011)
+```
+
+```{r plot-predictions, echo = TRUE, fig.height=5, fig.width=5, out.width="4.5in", cache=TRUE}
+ggplot(pred_qcs, aes(X, Y, fill = exp(est))) +
+    geom_raster() + facet_wrap(~year) +
+    coord_fixed() + scale_fill_viridis_c(trans = "sqrt")
+```
+
+```{r marginal, cache=TRUE, echo = TRUE, fig.height=2.5, fig.width=4, out.width="3.5in", echo = TRUE}
+nd <- data.frame(depth = seq(50, 300, length.out = 100), year = 2017L)
+pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = ~ 0)
+ggplot(pred_depth, aes(depth, exp(est),
+  ymin = exp(est - 1.96 * est_se), ymax = exp(est + 1.96 * est_se))) +
+  geom_line() + geom_ribbon(alpha = 0.4) + 
+  ylab("Expected density\n(kg/TODO)") +
+  xlab("Depth (m)")
+```
+
+```{r sims, cache=TRUE, echo=TRUE}
+pred_sims <- predict(fit, newdata = qcs_grid_2011, sims = 500)
+dim(pred_sims)
+```
+
+```{r sims-cv-plot, cache=FALSE, fig.height=4, fig.width=4, out.width="3.5in", echo=TRUE}
+qcs_grid_2011$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
+ggplot(subset(qcs_grid_2011, year == 2011), aes(X, Y, fill = cv)) +
+    geom_raster() + facet_wrap(~year) +
+    coord_fixed() + scale_fill_viridis_c(trans = "log10")
+```
+
+```{r index, cache=TRUE, fig.height=2.5, fig.width=4, out.width="3.5in"}
+qcs_grid_2011$area <- 4 # all 2 x 2km
+pred2 <- predict(fit, newdata = qcs_grid_2011, return_tmb_object = TRUE, 
+  area = qcs_grid_2011$area)
+ind <- get_index(pred2)
+ggplot(ind, aes(year, est)) + geom_line() +
+  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
+  ylab("Biomass (kg)")
+```
 

### 2021-10-30 analysis/compare-bruinla.R
Catchup

@@ -1,0 +1,75 @@
+# from MEE inlabru paper
+
+library(INLA)
+library(inlabru)
+library(RColorBrewer)
+data(gorillas)
+
+gnests <- gorillas$nests
+mesh <- gorillas$mesh
+boundary <- gorillas$boundary
+gcounts <- gorillas$plotsample$counts
+plots <- gorillas$plotsample$plots
+plotnests <- gorillas$plotsample$nests
+
+fig.plots <- ggplot() +
+  gg(boundary) +
+  gg(plots) +
+  gg(gnests, size=0.07,color = "blue") +
+  gg(plotnests,size=0.07,color = "red") +
+  geom_label(aes(label=gcounts$count, x=gcounts$x+0.35, y=gcounts$y-0.3),
+    fill = "white", size = 1.5, label.padding = unit(0.1, "lines")) +
+  coord_fixed() + xlab("x") + ylab("y")
+fig.plots
+
+# Inference using Poisson count likelihood
+# Define random variables
+
+constr <- FALSE
+mdl <- count ~ spat(main = coordinates,
+  model = inla.spde2.matern(mesh, constr = constr)) +
+  Intercept
+
+# The `constr = TRUE` parameter imposes an integrate-to-zero constraint on the random field to make the model more identifiable.
+# Fit the model
+
+fit <- bru(components = mdl, family = "poisson", data = gcounts,
+  options = list(E = gcounts$exposure), formula = ~ spat + Intercept)
+print(fit)
+
+# Results
+# Predict the intensity
+
+pxl <- pixels(mesh, mask = boundary, nx = 500, ny = 400)
+dens <- predict(fit, pxl, ~ exp(spat+Intercept))
+
+# Plot it
+
+fig.brudens <- ggplot() +
+  gg(dens) +
+  gg(boundary, alpha = 0) +
+  gg(plots) +
+  scale_fill_gradientn(colours = brewer.pal(9,"YlOrRd"),
+    limits = c(0,max(dens$mean))) +
+  gg(gcounts, mapping = aes(size = count)) +
+  scale_size_area(max_size = 2)
+fig.brudens
+
+# sdmTMB
+
+library(sdmTMB)
+d <- as.data.frame(gcounts)
+# mesh <- make_mesh(d, xy_cols = c("x", "y"), )
+mesh2 <- make_mesh(d, xy_cols = c("x", "y"), mesh = gorillas$mesh)
[…]

### 2021-10-30 analysis/compare-geostatsp.R

@@ -1,0 +1,79 @@
+library(geostatsp)
+data("swissRain")
+names(swissRain)
+names(swissAltitude)
+
+library(sdmTMB)
+library(ggplot2)
+parana <- geoR::parana
+d <- data.frame(rain = parana$data, x = parana$coords[, 1], y = parana$coords[, 2])
+head(d)
+rain_mesh <- make_mesh(d, xy_cols = c("x", "y"), cutoff = 10)
+rain_mesh$mesh$n
+ggplot(d, aes(x, y, size = rain)) +
+  geom_point() +
+  scale_size_area()
+plot(rain_mesh)
+
+parana <- as.data.frame(parana)
+coo <- as.matrix(parana[, c("east", "north")])
+bnd <- inla.nonconvex.hull(coo)
+meshb <- inla.mesh.2d(
+  boundary = bnd, offset = c(50, 100),
+  cutoff = 5, max.edge = c(75, 150)
+)
+plot(meshb)
+
+rain_mesh <- make_mesh(d, xy_cols = c("x", "y"), mesh = meshb)
+tictoc::tic()
+fit <- sdmTMB(rain ~ 1,
+  data = d, family = Gamma(link = "log"),
+  spde = rain_mesh, silent = TRUE
+)
+tictoc::toc()
+fit
+
+bb <- bbox(parana$border)
+x <- seq(bb[1, "min"] - 1, bb[1, "max"] + 1, length.out = 50)
+y <- seq(bb[2, "min"] - 1, bb[2, "max"] + 1, length.out = 50)
+coop <- as.matrix(expand.grid(x, y))
+ind <- point.in.polygon(
+  coop[, 1], coop[, 2],
+  parana$border[, 1], parana$border[, 2]
+)
+coop <- coop[which(ind == 1), ]
+coop <- as.data.frame(coop)
+names(coop) <- c("x", "y")
+
+p <- predict(fit, newdata = coop)
+
+ggplot(p, aes(x, y, fill = exp(est))) +
+  geom_raster() +
+  scale_fill_viridis_c(option = "C") +
+  geom_point(aes(size = rain), data = d)
+
+psims <- predict(fit, newdata = coop, sims = 1000)
+p$cv <- apply(psims, 1, function(x) sd(exp(x)) / mean(exp(x)))
+
+ggplot(p, aes(x, y, fill = cv)) +
+  geom_raster() +
+  scale_fill_viridis_c(option = "B")
+
+
+library(inlabru)
+library(INLA)
+
[…]

### 2021-10-30 analysis/compare-inla.R

@@ -48,9 +48,8 @@ summary(m_inla)
 # https://www.paulamoraga.com/book-geospatial/sec-inla.html
 
 tictoc::tic("sdmTMB")
-m <- sdmTMB(present ~ 0 + as.factor(year), data = d, spde = mesh, time = "year",
-  fields = "IID", include_spatial = FALSE, family = binomial(link = "logit"),
-  reml = FALSE,
+m <- sdmTMB(present ~ 0 + as.factor(year), data = d, mesh = mesh, time = "year",
+  spatiotemporal = "IID", spatial = "off", family = binomial(link = "logit"),
   priors = sdmTMBpriors(
     matern_st = pc_matern(range_gt = 5, range_prob = 0.05, sigma_lt = 5, sigma_prob = 0.05),
     ar1_rho = normal(0, 1),

### 2021-10-30 analysis/compare-mgcv.R

@@ -25,3 +25,26 @@ tictoc::toc()
 p2 <- predict(m)
 
 plot(p1, p2$est);abline(a = 0, b = 1)
+
+
+mesh$mesh$n
+
+mesh2 <- inla.mesh.2d(
+  loc = d[,c("X", "Y")],
+  # boundary = boundary,
+  # offset = c(-0.05, -0.05),
+  # offset = c(-0.05, -0.05),
+  max.n = 100,
+  # max.n.strict = 5000,
+  cutoff = 5
+  # max.edge = c(100, 500)
+)
+plot(mesh2)
+
+library(INLA)
+fit <- gam(present ~ 0 + as.factor(year) + s(X, Y, bs = "spde", k = mesh2$n,
+  xt = list(mesh = mesh2)),
+  data = d,
+  family= binomial(),
+  method = "ML",
+  control =  gam.control(scalePenalty = FALSE))

### 2021-10-30 analysis/inlabru-examples/02_random_fields.R

@@ -1,0 +1,403 @@
+#' Fitting georeferenced data in 1D and 2D
+#'=========================================================
+
+#' Set things up
+#+results="hide",warning=FALSE,message=FALSE
+library(INLA)
+library(inlabru)
+library(mgcv)
+
+#' Set default INLA values for tutorial session,
+#' and make a shortcut to a nicer colour scale:
+#' 
+#+results="hide",warning=FALSE,message=FALSE
+
+init.tutorial()
+colsc <- function(...) {
+  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11,"RdYlBu")),
+                       limits = range(..., na.rm=TRUE))
+}
+
+#' This is precisely what `init.tutorial` does:
+#+echo=FALSE
+
+print(init.tutorial)
+
+#' `iinla.verbose=TRUE` prints iterated inla progression information
+#'
+#' `int.strategy="eb"` limits INLA to Empirical Bayes so that the
+#' examples run faster; no posterior integration over $\theta$ is performed.
+#' See `?INLA::control.inla` for some more information.
+#' 
+#' The `control.compute` options turn on information needed for posterior sampling, and
+#' tells INLA to compute DIC and WAIC. See `?INLA::control.compute` for more information.
+
+
+
+#' Fitting to 1-dimensional gridded data (counts)
+#'=========================================================
+
+#' Get the data
+#'-----------------------------------
+#+results="hide",warning=FALSE,message=FALSE
+
+data(Poisson2_1D)
+
+#' Put the count data in `cd` (just because '`cd`' is less to type than '`countdata2`'.)
+
+cd <- countdata2
+
+#' Take a look at the count data.
+#' 
+#+warning=FALSE,message=FALSE
+
+cd
+ggplot(cd) + geom_point(aes(x, y = count)) + ylim(0, max(cd$count))
+
+#' _Tip_: `RStudio > Help > Cheatsheets > Data visualisation with ggplot2` is a useful 
+#' reference for `ggplot2` syntax.
+
+
+ 
+#' Fit a Generalised Additive Model (GAM)
+#'-----------------------------------
+#' If you're not familiar with GAMs and the syntax of `gam` don't worry, the point of 
+#' this is just to provide something to which we can compare the `inlabru` model fit.
+#' 
+#+results="hide",warning=FALSE,message=FALSE
+
+fit2.gam <- gam(count ~ s(x, k = 10) + offset(log(exposure)), family = poisson(), data = cd)
+
[…]

### 2021-10-30 analysis/inlabru-examples/08_spatio_temporal.R

@@ -1,0 +1,175 @@
+#' Spatio-temporal modelling
+#'=========================================================
+#+results="hide",warning=FALSE,message=FALSE
+# rm(list=ls())
+
+library(inlabru)
+init.tutorial()
+library(rgdal)
+library(ggmap) # for fancy plotting
+library(INLA)
+library(RColorBrewer)
+
+#' Make a shortcut to a nicer colour scale:
+#+results="hide",warning=FALSE,message=FALSE
+colsc <- function(...) {
+  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11,"RdYlBu")),
+                       limits = range(..., na.rm=TRUE))
+}
+
+#' Read in the data
+#+results="hide"
+
+load("Nigeria/terror_data.RData")
+terror.data<-data[data$country=='NGA',]
+#' look at it
+#+warning=FALSE,message=FALSE
+head(terror.data)
+class(terror.data)
+
+studyarea<-studyarea[studyarea$sov_a3=='NGA',]
+data<-data[data$country=='NGA',]
+bdry <- inla.sp2segment(studyarea)
+bdry$loc <- inla.mesh.map(bdry$loc)
+
+#' this is quite a rough mesh; used here to avoid long running times
+#' during the practical
+#+warning=FALSE,message=FALSE
+mesh3<-inla.mesh.2d(boundary=bdry, max.edge=c(80,6000)/180,cutoff=80/180,
+                    crs=CRS(proj4string(studyarea))) #nv=330
+plot(mesh3)
+
+terror.data <- terror.data[terror.data$iyear>=2010 & terror.data$iyear<=2015,]
+round.to <- function(x, b) {round(x/b)*b}
+terror.data$originalnkill<-round.to(terror.data$nkill, 1)#some observations have non-integer values
+terror.data$originallethal<-ifelse(terror.data$originalnkill==0,0,1)
+resp.bern<- terror.data$originallethal
+
+#' The data are stored as a `data.frame` and the columns holding the coorinates are
+#' called `longitude` and `latitude`. Using this information we can convert the data into a
+#' `SpatialPointsDataFrame` object
+#+warning=FALSE,message=FALSE
+coordinates(terror.data) = c("longitude","latitude")
+
+#' The object `terror.data` now is in the proper format:
+#+warning=FALSE,message=FALSE
+
+class(terror.data)
+
+#' and the coordinates of each point are equivalent to the previous `Lat` and `Long`:
+
[…]

### 2021-10-30 analysis/meuse.R

@@ -1,0 +1,145 @@
+library("spatstat")
+library("sp")
+library("maptools")
+library(INLA)
+
+library("gstat")
+data(meuse)
+
+coordinates(meuse) <- ~ x + y
+proj4string(meuse) <- CRS("+init=epsg:28992")
+
+# Code from gstat
+data(meuse.grid)
+coordinates(meuse.grid) <- ~ x + y
+proj4string(meuse.grid) <- CRS("+init=epsg:28992")
+gridded(meuse.grid) <- TRUE
+
+meuse.bdy <- unionSpatialPolygons(
+  as(meuse.grid, "SpatialPolygons"), rep(1, length(meuse.grid))
+)
+
+pts <- meuse.bdy@polygons[[1]]@Polygons[[1]]@coords
+pts <- pts
+mesh <- inla.mesh.2d(
+  loc.domain = pts, max.edge = c(300, 600),
+  offset = c(100, 250)
+)
+plot(mesh, asp = 1, main = "")
+lines(pts, col = 3, lwd = 2)
+
+pts2 <- pts / 1000
+mesh2 <- inla.mesh.2d(
+  loc.domain = pts2, max.edge = c(.300, .600),
+  offset = c(.100, .250)
+)
+
+# Create SPDE
+meuse.spde <- inla.spde2.matern(mesh = mesh, alpha = 2)
+A.meuse <- inla.spde.make.A(mesh = mesh, loc = coordinates(meuse))
+s.index <- inla.spde.make.index(
+  name = "spatial.field",
+  n.spde = meuse.spde$n.spde
+)
+
+meuse.stack <- inla.stack(
+  data = list(zinc = meuse$zinc),
+  A = list(A.meuse, 1),
+  effects = list(
+    c(s.index, list(Intercept = 1)),
+    list(dist = meuse$dist)
+  ),
+  tag = "meuse.data"
+)
+
+A.pred <- inla.spde.make.A(mesh = mesh, loc = coordinates(meuse.grid))
+meuse.stack.pred <- inla.stack(
+  data = list(zinc = NA),
+  A = list(A.pred, 1),
+  effects = list(
+    c(s.index, list(Intercept = 1)),
+    list(dist = meuse.grid$dist)
+  ),
+  tag = "meuse.pred"
+)
+
+# Join stack
+join.stack <- inla.stack(meuse.stack, meuse.stack.pred)
+
+# Fit model
[…]

### 2021-11-01 doc/paper.Rmd
Catchup
Rough draft of intro

@@
 header-includes:
+  \usepackage{setspace}\onehalfspacing
   \usepackage[left]{lineno}
@@
 
+*[quick draft outline of intro; may be too long]*
+
+Ecological data are often collected in space or in space repeatedly over time.
+For example...
+?
+While such data are a rich source of information about ecological process, it is challenging to properly modelling them.
+Data points closer in space and time are usually more correlated to each other than those farther apart, both due to measured and unmeasured variables.
+While measured variables can be accounted for via predictors in a model, unmeasured ("latent") variables cause residual spatial/spatiotemporal correlation that (1) needs be accounted for valid statistical inference, (2) can improve predictions compared to ignoring it, and (3) can be of ecological interest in itself.
+
+Geostatistical GLMMs (Generalized Linear Mixed Effects Models) with spatially correlated random effects is a class of models appropriate for these data.
+Similarly to how random intercepts or slopes can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
+A common approach to modelling the random effects is via Gaussian Random fields (GRFs), where the random effects are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn.
+
+While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data points.
+A number of solutions have been proposed, such as predictive processes (e.g., REF), NNGP (e.g., REF), and SPDE + GMRF (e.g., REF).
+These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
+This can make an enormous difference to computational efficiency (REFs).
+The SPDE approach is a solution popularized via INLA and an implementation in TMB (Template Model Builder) that partially relies on INLA.
+The SPDE… yada yada GMRF yada yada.
+
+INLA and TMB are template systems for specifying statistical models that can include the SPDE, but they are challenging to use for many applied ecologists and it can be slow to iterate on and experiment with many models when writing bespoke code.
+Packages such as lme4 and glmmTMB let users quickly iterate and explore statistical models, evaluating fit and comparing models, but do not have built-in SPDE functionality.
+Packages such as VAST and inlabru are powerful user interfaces to spatial models that use the SPDE, but either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some of the flexibility for the particular class of models discussed here.
+We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
+<!-- User interfaces through software such as VAST or inlabru can greatly reduce these challenges.
+VAST is built on TMB, and is extremely powerful (e.g., fitting many multivariate models not described here), but does not provide a simple, familiar, modular interface for users with basic spatial or spatiotemporal data.
+inlabru provides a somewhat more familiar interface to a large class of models including GRMF models, but is strictly Bayesian, lacks some flexibility for the particular class of models discussed here.
+It also relies on INLA, and at least for some problems, TMB can be much faster [@osgood-zimmerman2021].
+We provide a more detailed comparison of related software packages in Table 1 and the discussion.
+-->
+
+<!-- User interfaces to classes of models fit by these template systems can greatly  -->
+
+<!-- reduce these challenges for the vast majority of models. -->
+
+<!-- keep the above even briefer? -->
+
+<!-- inlabru is "very good" but not entirely familiar to those used to `glm()`, lme4, glmmTMB, or mgcv; -->
+
+<!-- *TODO: the differences are many and more than can be listed here. Keep brief add Table or discussion later about differences and when to pick one over the other?* -->
+
+Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB, INLA, and Gaussian Markov random fields.
+One common application is for species distribution models (SDMs), hence the package name.
+This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through X case studies, and concludes with a discussion of additional functionality and related software.
+
 -   [@anderson2019]
@@
 -   [@latimer2009]
+
+Packages:
+
+spaMM: @rousset2014
+spNNGP: @finley2021
 
@@
 
-```{r print-fit, echo = TRUE}
+```{r print-fit, echo = TRUE, cache=TRUE}
 print(fit)
@@
 
-```{r sims-cv-plot, cache=FALSE, fig.height=4, fig.width=4, out.width="3.5in", echo=TRUE}
+```{r sims-cv-plot, cache=TRUE, fig.height=4, fig.width=4, out.width="3.5in", echo=TRUE}
 qcs_grid_2011$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))

### 2021-11-02 doc/paper.Rmd
Adding to intro

@@
 Ecological data are often collected in space or in space repeatedly over time. 
-For example...
-?
-While such data are a rich source of information about ecological process, it is challenging to properly modelling them.
-Data points closer in space and time are usually more correlated to each other than those farther apart, both due to measured and unmeasured variables.
-While measured variables can be accounted for via predictors in a model, unmeasured ("latent") variables cause residual spatial/spatiotemporal correlation that (1) needs be accounted for valid statistical inference, (2) can improve predictions compared to ignoring it, and (3) can be of ecological interest in itself.
+While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model [e.g., @cressie1993; @cressie2011]. 
+Data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables. 
+While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
+This residual correlation is ideally dealt with because doing so (1) results in valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].
 
-Geostatistical GLMMs (Generalized Linear Mixed Effects Models) with spatially correlated random effects is a class of models appropriate for these data.
-Similarly to how random intercepts or slopes can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling the random effects is via Gaussian Random fields (GRFs), where the random effects are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn.
+Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data. 
+Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
+A common approach to modelling such random effects is with Gaussian Random fields (GRFs), where the random effects are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
 
-While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data points.
-A number of solutions have been proposed, such as predictive processes (e.g., REF), NNGP (e.g., REF), and SPDE + GMRF (e.g., REF).
+While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009]. 
+Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
 These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
-This can make an enormous difference to computational efficiency (REFs).
-The SPDE approach is a solution popularized via INLA and an implementation in TMB (Template Model Builder) that partially relies on INLA.
-The SPDE… yada yada GMRF yada yada.
+This can make an enormous difference to computational efficiency [@heaton2017].
+The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA.
+Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].
 
-INLA and TMB are template systems for specifying statistical models that can include the SPDE, but they are challenging to use for many applied ecologists and it can be slow to iterate on and experiment with many models when writing bespoke code.
-Packages such as lme4 and glmmTMB let users quickly iterate and explore statistical models, evaluating fit and comparing models, but do not have built-in SPDE functionality.
-Packages such as VAST and inlabru are powerful user interfaces to spatial models that use the SPDE, but either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some of the flexibility for the particular class of models discussed here.
+Template systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists. 
+It can also be slow to iterate on and experiment with many models when writing bespoke template code. 
+Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to spatial models that use the SPDE, but either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some of the flexibility for the particular class of models discussed here.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
@@
 
-Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB, INLA, and Gaussian Markov random fields.
+Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA in a fast, flexible, and user-friendly interface that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
 One common application is for species distribution models (SDMs), hence the package name.
@@
 spNNGP: @finley2021
+
+millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies."
+
+legendre1989a: "Spatial pattern and ecological analysis": "We can expect the spatial approach to ecological problems to bring about a quantum jump for ecologists and population geneticists who had learned a type of statistics where one had to hide space or time structures.
+It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables."
+
+ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany.
 

### 2021-11-02 doc/paper.Rmd
Edit intro

@@
 ---
-title: "`sdmTMB`: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
+title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
 author: "|
@@
 Ecological data are often collected in space or in space repeatedly over time. 
-While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model [e.g., @cressie1993; @cressie2011]. 
-Data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables. 
+While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [e.g., @cressie1993; @cressie2011].
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
@@
 Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling such random effects is with Gaussian Random fields (GRFs), where the random effects are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
+A common approach to modelling such random effects is with Gaussian random fields (GRFs), where the random effects are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
 
@@
 This can make an enormous difference to computational efficiency [@heaton2017].
-The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA.
+The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA [e.g., @osgood-zimmerman2021].
 Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].
@@
 
-Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA in a fast, flexible, and user-friendly interface that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
+Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA.
+Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
 One common application is for species distribution models (SDMs), hence the package name.
@@
 
--   [@anderson2019]
--   [@maureaud2021]
--   [@anderson2020]
--   [@osgood-zimmerman2021]
--   [@breivik2021]
--   [@monnahan2021]
--   [@shelton2014]
--   [@lindgren2011]
--   [@latimer2009]
+<!-- -   [@anderson2019] -->
+<!-- -   [@maureaud2021] -->
+<!-- -   [@anderson2020] -->
+<!-- -   [@osgood-zimmerman2021] -->
+<!-- -   [@breivik2021] -->
+<!-- -   [@monnahan2021] -->
+<!-- -   [@shelton2014] -->
+<!-- -   [@lindgren2011] -->
+<!-- -   [@latimer2009] -->
 
-Packages:
+<!-- Packages: -->
 
-spaMM: @rousset2014
-spNNGP: @finley2021
+<!-- spaMM: @rousset2014 -->
+<!-- spNNGP: @finley2021 -->
 
-millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies."
+<!-- millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies." -->
 
-legendre1989a: "Spatial pattern and ecological analysis": "We can expect the spatial approach to ecological problems to bring about a quantum jump for ecologists and population geneticists who had learned a type of statistics where one had to hide space or time structures.
-It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables."
+<!-- legendre1989a: "Spatial pattern and ecological analysis": "We can expect the spatial approach to ecological problems to bring about a quantum jump for ecologists and population geneticists who had learned a type of statistics where one had to hide space or time structures.
+It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables." -->
 
-ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany.
+<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
 

### 2021-11-02 doc/paper.Rmd
add more text

@@
 # Description
-- maybe some overview of spatial models, either here in intro. Contrast with GAMs, and highlight explicit estimation of flexible spatial covariance functions
+
+Spatial and spatiotemporal models constructed with `sdmTMB` may be viewed as spatially expicit models, in the sense that they estimate parameters of a user-specified spatial covariance function, and these estimates may be converted into interpretable quantities common in spatial statistics (spatial decay, range).
+We contrast these approaches with semi- and non-parametric approaches that do not estimate spatial covariance functions (e.g.
+`randomForest` [@liaw2002], `MaxEnt` [@phillips2006], some smooths in `mgcv`).
+By default, sdmTMB implements the Matérn covariance (kernel) (cite Rue et al?) -- this function can accomodate a range of shapes, but also be anisotropic (covariance in a latitudinal direction may be different than in longitudinal direction).
 
@@
 GLMMs are available in several widely used packages, including `glmmTMB` (Brooks et al.
-2017) and `lme4` (Bates et al.
-2015), and to allow comparison we also allow non-spatial models to be constructed with `sdmTMB`.
+2017), `mgcv` [@wood2004], and `lme4` (Bates et al.
+2015), and to allow comparison with these packages, `sdmTMB` allows for spatial and spatiotemporal effects to be removed.
 
@@
 For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model.
-The full `sdmTMB` model is
+Though in practice it is rare to include all of these parameters, the full `sdmTMB` model is
 $$
@@
 The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function. 
+<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
 The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
@@
 [paragraph about spde and gaussian process / Matern covariance here]
+
+<!-- Paragraph on model selection here-->
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model (Akaike XXXX).
+Because of the well documented biases of AIC with hierarchical or mixed-effects models (can cite Vaida and Blanchard or another) we also provide functionality to perform estimation via k-fold  cross-validation.
+Our wrapper function `sdmTMB_cv` allows for any number of folds, with assignments being user-specified or chosen randomly.
+The predictive log-density is calculated for each fold, and then summarized across folds for a total measure of the predictive accuracy of a model.
+
+<!-- Paragraph on derived quantities here-->
+The results from a model fit with sdmTMB can be used to generate quantities of interest, especially relevant to population status.
+Estimated parameters from the Matérn covariance function can be used to describe the spatial variance and Matérn range (= $\sqrt{8} / \kappa$).
+We provide additional functionality for generating population-level summaries for each time slice; these metrics include the fraction of area occupied (?), 2-dimensional center of gravity (via our `get_cog()` function), and total population trend (via our `get_index()` function).
 

### 2021-11-02 diagram-figs.R
Add diagram figs

@@ -1,0 +1,94 @@
+library(sdmTMB)
+library(dplyr)
+library(ggplot2)
+
+predictor_dat <- expand.grid(X = seq(0, 1, length.out = 100), Y = seq(0, 1, length.out = 100))
+predictor_dat <- bind_rows(
+  mutate(predictor_dat, year = 1L),
+  mutate(predictor_dat, year = 2L),
+  mutate(predictor_dat, year = 3L),
+  mutate(predictor_dat, year = 4L)
+  )
+mesh <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), cutoff = 0.03)
+plot(mesh$mesh)
+
+head(predictor_dat)
+nrow(predictor_dat)
+
+d <- sdmTMB_simulate(
+  formula = ~ 1,
+  data = predictor_dat,
+  time = "year",
+  mesh = mesh,
+  family = gaussian(link = "identity"),
+  range = 0.3,
+  phi = 0.1,
+  sigma_O = 0.2,
+  sigma_E = 0.2,
+  seed = 123,
+  B = c(0)
+)
+
+dir.create("figs", showWarnings = FALSE)
+ggplot(filter(d, year == 1L), aes(X, Y, fill = omega_s)) +
+  geom_raster() +
+  scale_fill_gradient2() +
+  theme_void() +
+  theme(legend.position = "none") +
+  coord_fixed()
+ggsave("figs/omega_s.png", width = 3, height = 3)
+
+ggplot(filter(d, year == 1L), aes(X, Y, fill = epsilon_st)) +
+  geom_raster() + scale_fill_gradient2() + theme_void() +
+  theme(legend.position = "none") +
+  coord_fixed()
+ggsave("figs/eps_st1.png", width = 3, height = 3)
+
+ggplot(filter(d, year == 2L), aes(X, Y, fill = epsilon_st)) +
+  geom_raster() + scale_fill_gradient2() + theme_void() +
+  theme(legend.position = "none") +
+  coord_fixed()
+ggsave("figs/eps_st2.png", width = 3, height = 3)
+
+ggplot(filter(d, year == 3L), aes(X, Y, fill = epsilon_st)) +
+  geom_raster() + scale_fill_gradient2() + theme_void() +
+  theme(legend.position = "none") +
+  coord_fixed()
+ggsave("figs/eps_st3.png", width = 3, height = 3)
+
[…]

### 2021-11-02 doc/paper.Rmd
X_3,t
Clean up;edit

--- 2021-11-02 doc/paper.Rmd
@@
 <!-- # For submission to MEE as an Application paper -->
-
-<!-- # The sdmTMB package implements spatiotemporal predictive-process GLMMs (Generalized Linear Mixed Effect Models) using TMB, INLA, and Gaussian Markov random fields.
-One common application is for spatial or spatiotemporal species distribution models (SDMs).
--->
 
@@
 
-*[quick draft outline of intro; may be too long]*
-
 Ecological data are often collected in space or in space repeatedly over time.
@@
 Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA.
-Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
+Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
 One common application is for species distribution models (SDMs), hence the package name.
@@
 # Description
-- maybe some overview of spatial models, either here in intro. Contrast with GAMs, and highlight explicit estimation of flexible spatial covariance functions
+
+Spatial and spatiotemporal models constructed with sdmTMB may be viewed as spatially explicit models, in the sense that they estimate parameters of a user-specified spatial covariance function, and these estimates may be converted into interpretable quantities common in spatial statistics (spatial decay, range).
+We contrast these approaches with semi- and non-parametric approaches that do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], some smooths in mgcv [@wood2017a]).
+By default, sdmTMB implements the Matérn covariance function (REF) (cite Rue et al?)---this function can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in a latitudinal direction may be different than in the longitudinal direction).
 
@@
 
-Though `sdmTMB` has been developed with spatial or spatiotemporal modelling as a primary focus, all of the statistical models it fits fall under the class of generalized linear mixed models (GLMMs).
-GLMMs are available in several widely used packages, including `glmmTMB` (Brooks et al.
-2017) and `lme4` (Bates et al.
-2015), and to allow comparison we also allow non-spatial models to be constructed with `sdmTMB`.
+Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs.
+GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.
+
+By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
+We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000B].
+
+[paragraph about spde and gaussian process / Matern covariance here]
+
+<!-- Paragraph on model selection here-->
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model (Akaike XXXX). 
+Because of well-documented biases of AIC with hierarchical or mixed-effects models (can cite Vaida and Blanchard or another), we also provide functionality to perform estimation via k-fold cross-validation.
+Our function `sdmTMB_cv()` allows for any number of folds or 'blocks', with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold, and then summarized across folds for a total measure of the predictive accuracy of a model.
+
+<!-- Paragraph on derived quantities here-->
+The results from a model fit with sdmTMB can be used to generate quantities of interest, especially relevant to population status. 
+Estimated parameters from the Matérn covariance function can be used to describe the spatial variance and Matérn range. 
+We provide additional functionality for generating population-level summaries for each time slice; these metrics include the fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index ( `sdmTMB::get_index()`).
+
+<!--
+Highlights?
+
+- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances)
+- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
+- Spatiotemporal models (RW/AR v IID)
+- Time varying parameters
+- Non-linear effects (smooth splines, threshold/breakpoint functions)
+- Variance estimation and derived quantities (spatial range, area occupied etc)
+-->
+
+# Supporting Information
 
@@
 \bm{X}_{2,\bm{s},t} \gamma_t +
-\bm{X}_{3,\bm{s},t} \zeta_{\bm{s}} +
+\bm{X}_{3,t} \zeta_{\bm{s}} +
 \omega_{\bm{s}} +
@@
 The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function.
+<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
 The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
@@
 We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
-
-By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in `sdmTMB` allow them to be modelled as a random walk or first-order autoregressive (AR1) process.
-We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: `sdmTMB` allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in `mgcv` (Wood 2003), or to be modelled with threshold shapes (e.g., hockey stick models; Barrowman et al.
-2003).
-
-[paragraph about spde and gaussian process / Matern covariance here]
-
-Highlights?
-
-- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances) 
-- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
-- Spatiotemporal models (RW/AR v IID)
-- Time varying parameters
-- Non-linear effects (smooth splines, threshold/breakpoint functions)
-- Variance estimation and derived quantities (spatial range, area occupied etc)
 

### 2021-11-02 doc/paper.Rmd
Update paper.Rmd

@@
 <!-- Paragraph on model selection here-->
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model (Akaike XXXX). 
-Because of well-documented biases of AIC with hierarchical or mixed-effects models (can cite Vaida and Blanchard or another), we also provide functionality to perform estimation via k-fold cross-validation.
-Our function `sdmTMB_cv()` allows for any number of folds or 'blocks', with assignments being user-specified or chosen randomly. 
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974].
+Because of the well documented biases of AIC with hierarchical or mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold  cross-validation.
+Our wrapper function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly.
 The predictive log-density is calculated for each fold, and then summarized across folds for a total measure of the predictive accuracy of a model.
@@
 The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function.
-<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
-The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
-(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
-<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
-where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
-a scaling parameter that is estimated (REF).
-The parameter $\nu$ controls the smoothness of the covariance function.
-We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+Additional details are provided in the SI.
 
@@
 ```
-
-*EW: These are just placeholders -- not sure what order is best or what to highlight in very limited space*
-
-As a first example, we use a spatial dataset (`parana`) from the `geoR` package (Ribeiro Jr and Diggle 2018) that describes average May-June rainfall recorded at 143 stations in the state of Paraná, Brazil.
-This dataset is also well-known because it is used as an example of spatiotemporal modeling with INLA (cite INLA) book.
-Because the dataset doesn't include covariates and only includes a single year, we can model rainfall using an intercept and spatial component, $g(\mu_{s,t}) = b_{0} + \omega_{s}$.
-We assume the observation model for the rainfall data is normally distributed.
-The first step in fitting a model with `sdmTMB` is constructing the mesh to approximate the spatial process.
-This step can be done with the `make_mesh` function, which accepts the data (coordinates) and at minimum, the minimum allowed distance between points (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with INLA can also be passed in (additional examples of this are included in the SI).
-For example, an initial mesh with a cutoff distance of 0.1km could be made with
-
-```{r eval=FALSE, echo=TRUE}
-rain_spde = make_mesh(data = parana, c("east","north"), cutoff=0.1)
-```
-
-Second, we can fit the model using TMB with the `sdmTMB` function, whose formula syntax should be familiar to most R users,
-
-```{r eval = FALSE, echo=TRUE}
-fit = sdmTMB(data ~ 1, data=parana, spde = rain_spde)
-```
-
-As a second example, we use `sdmTMB` to analyze a 2-year dataset of Eurasian cranes (*Grus grus*) from England, UK (Soriano-Redondo et al.
-2019).
-In the original analysis, Soriano-Redondo et al.
-2019 used log Gaussian Cox process models to jointly model crane presence--absence and the spatiotemporal distribution of wetlands.
-We use this dataset as an example of modeling presence--absence data with sdmTMB, and only model the occurence of cranes given covariates.
-With any multi-year dataset, a common question of interest is often whether observations can be best described by spatial variation, or spatial and spatiotemporal variation.
-With two years of data, we fit two models with `sdmTMB`: a model with a spatial component but no spatiotemporal variation, and a model with only spatiotemporal variation (allowing each year to have a different spatial field).
-We included the same fixed effect covariates in both models: different intercepts by year, and two covariates found to be important by Soriano-Redondo et al.
-2019 (wetland area, perimeter to area ratio).
-As a coarse comparison of data support, we compare the Akaike's Information Criterion (Akaike 1973) from each model.
-The model with a shared spatial field appeared to be more supported ($\Delta AIC$=25.23).
-The sdmTMB package allows for additional model comparisons, including k-fold cross validation with the `sdmTMB_cv` function.
-
-Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?
 
@@
 # Supplemental Materials
+
+### Matern parameterization
+<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
+The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
+(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
+<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
+where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
+a scaling parameter that is estimated (REF).
+The parameter $\nu$ controls the smoothness of the covariance function.
+We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+
+### Junk to delete
+*EW: These are just placeholders -- not sure what order is best or what to highlight in very limited space*
+
+As a first example, we use a spatial dataset (`parana`) from the `geoR` package (Ribeiro Jr and Diggle 2018) that describes average May-June rainfall recorded at 143 stations in the state of Paraná, Brazil.
+This dataset is also well-known because it is used as an example of spatiotemporal modeling with INLA (cite INLA) book.
+Because the dataset doesn't include covariates and only includes a single year, we can model rainfall using an intercept and spatial component, $g(\mu_{s,t}) = b_{0} + \omega_{s}$.
+We assume the observation model for the rainfall data is normally distributed.
+The first step in fitting a model with `sdmTMB` is constructing the mesh to approximate the spatial process.
+This step can be done with the `make_mesh` function, which accepts the data (coordinates) and at minimum, the minimum allowed distance between points (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
+Custom built meshes with INLA can also be passed in (additional examples of this are included in the SI).
+For example, an initial mesh with a cutoff distance of 0.1km could be made with
+
+```{r eval=FALSE, echo=TRUE}
+rain_spde = make_mesh(data = parana, c("east","north"), cutoff=0.1)
+```
+
+Second, we can fit the model using TMB with the `sdmTMB` function, whose formula syntax should be familiar to most R users,
+
+```{r eval = FALSE, echo=TRUE}
+fit = sdmTMB(data ~ 1, data=parana, spde = rain_spde)
+```
+
+As a second example, we use `sdmTMB` to analyze a 2-year dataset of Eurasian cranes (*Grus grus*) from England, UK (Soriano-Redondo et al.
+2019).
+In the original analysis, Soriano-Redondo et al.
+2019 used log Gaussian Cox process models to jointly model crane presence--absence and the spatiotemporal distribution of wetlands.
+We use this dataset as an example of modeling presence--absence data with sdmTMB, and only model the occurence of cranes given covariates.
+With any multi-year dataset, a common question of interest is often whether observations can be best described by spatial variation, or spatial and spatiotemporal variation.
+With two years of data, we fit two models with `sdmTMB`: a model with a spatial component but no spatiotemporal variation, and a model with only spatiotemporal variation (allowing each year to have a different spatial field).
+We included the same fixed effect covariates in both models: different intercepts by year, and two covariates found to be important by Soriano-Redondo et al.
+2019 (wetland area, perimeter to area ratio).
+As a coarse comparison of data support, we compare the Akaike's Information Criterion (Akaike 1973) from each model.
+The model with a shared spatial field appeared to be more supported ($\Delta AIC$=25.23).
+The sdmTMB package allows for additional model comparisons, including k-fold cross validation with the `sdmTMB_cv` function.
+
+Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?

### 2021-11-02 doc/paper.Rmd
Edit; add Fig 1 caption

@@
 
-<!-- # For submission to MEE as an Application paper -->
-
 ```{r, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
@@
 )
-```
-
-```{=html}
-<!--
-
-Application: Short (no more than 3000-4000 words total) descriptions of new software that are intended to describe and promote the software as well as act as a citeable resource for developers.
-Application papers are made free to access for all readers upon publication in order to encourage uptake of the methodologies they describe.
-Uploading a package to a site such as [**CRAN**](http://cran.r-project.org/) or [**sourceforge** ](http://sourceforge.net/)in advance is not considered prior publication and will not hinder your submission to MEE.
-View some example Applications [**here**](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/features/applicationpapers).
-
-<https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/features/applicationpapers>
-
--   Single column, double line spaced
-
--   Within the word count (no more than 7000-8000 words for Standard Articles, 3000-4000 words for Applications and Practical Tools)
-
--   Continuous line and page numbering throughout
-
--   Clearly defined manuscript structure as standard: Author details, Abstract (must be numbered according to [**Manuscript Specifications**](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210x/author-guidelines#MS_spec)), Keywords, Introduction, Materials and Methods, Results, Discussion, Figures and Tables with captions
-
--   Figures and Tables can be embedded within the text where referenced to facilitate reviewing
-
--   Author Contributions ([[if \>1 author -- see Authorship Policy]{.ul}](https://besjournals.onlinelibrary.wiley.com/hub/editorial-policies#authorship)).
-
--   Statement of where you *intend* to [**archive your data**](https://besjournals.onlinelibrary.wiley.com/hub/data_archiving_policy)
-
-Title Page should include:
-
-A concise and informative title. Do not include the authorities for taxonomic names.
-A list of all authors' names with names and addresses of Institutions.
-The name, address and e-mail address of the correspondence author.
-A running headline of not more than 45 characters.
-
-Abstract
-
-The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:
-
-Key-words: A list in alphabetical order not exceeding eight words or short phrases.
-The most important key-words should appear in the title and the abstract as well as the key-word list.
-More advice on selecting good keywords can be found here.
-
-Introduction
-
-This should state the reason for doing the work, the nature of the hypothesis or hypotheses under consideration, and should outline the essential background.
-
-Materials and Methods
-
-Include sufficient details for the work to be repeated.
-Where specific equipment and materials are named, the manufacturer’s details (name, city and country) should be given so that readers can trace specifications by contacting the manufacturer.
-Where commercially available software has been used, details of the supplier should be given in brackets or the reference given in full in the reference list.
-Do not describe or refer to commonplace statistical tests in this section but allude to them briefly in Results.
-
-Results
-
-State the results and draw attention in the text to important details shown in tables and figures.
-
-Discussion
-
-This should point out the significance of the results in relation to the reasons for doing the work, and place them in the context of other work.
-
-Conclusions (optional)
-
-Acknowledgements (optional)
-
-A brief statement acknowledging collaborators and research assistants who do not meet the criteria for authorship described above, or acknowledging funding sources, providing relevant permit numbers (including institutional animal use permits), or giving recognition to nature reserves or other organizations that made the work possible.
-
-Conflict of Interest statement (required for all article types)
-
--->
 ```
@@
 
+(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
+An sdmTMB model can be built from any combination of these components.
+The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
+Notation:
+We refer to design matrices as $\bm{X}$.
+The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and 'group', respectively.
+The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
+All other symbols refer to the  described model components.
+Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
+
+```{r diagram, fig.cap="(ref:diagram-fig)"}
+if (file.exists(here::here("figs/diagram.pdf"))) {
+  knitr::include_graphics(here::here("figs/diagram.pdf"))
+} else {
+  plot(1, type = "n", main = "figs/diagram.pdf missing; will be synced when complete")
+}
+```
+
 # Examples of sections:
@@
 By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
-We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000B].
+We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000].
 

### 2021-11-02 doc/paper.Rmd
Edit

@@
 
-Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs.
-GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.
+<!-- Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs. -->
+<!-- GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.
+-->
 
-By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
-We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000].
+By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
+Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000] (Fig.\ \@ref(fig:diagram)).
 
@@
 Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
-Because of the well documented biases of AIC with hierarchical or mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold  cross-validation. 
-Our wrapper function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold, and then summarized across folds for a total measure of the predictive accuracy of a model.
+Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
+Our function `sdmTMB::sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
 
 <!-- Paragraph on derived quantities here-->
-The results from a model fit with sdmTMB can be used to generate quantities of interest, especially relevant to population status.
-Estimated parameters from the Matérn covariance function can be used to describe the spatial variance and Matérn range.
-We provide additional functionality for generating population-level summaries for each time slice; these metrics include the fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index ( `sdmTMB::get_index()`).
+The results from a model fit with sdmTMB can be used to generate quantities of interest.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index (`sdmTMB::get_index()`).
+Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
 

### 2021-11-02 doc/paper.Rmd
Outline structure

@@
 
+- high-level paragraph or two about what the model includes, referencing Fig 1
+- high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?)
+- high-level paragraph about simulation capabilities and + testing performed (reference supplement)
+- model selection/cross validation paragraph
+- derived quantities paragraph (or just put this in the examples?)
+
+- example 1 + figs
+- example 2 + figs
+
+- discussion
+    - take home, re-iterate
+    - how to pick among packages/tools including a bit outside the SPDE world? (table)
+    - touch on other stuff possible: priors, Bayes, barrier meshes, residuals, 
+      simulation from joint precision, ...?
+
 # Model
@@
 The results from a model fit with sdmTMB can be used to generate quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index (`sdmTMB::get_index()`).
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`sdmTMB::get_cog()`) and total population index (`sdmTMB::get_index()`).
 Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).

### 2021-11-02 doc/paper.Rmd
Edits
Add notes and discussion

@@
 
-- discussion
-    - take home, re-iterate
-    - how to pick among packages/tools including a bit outside the SPDE world? (table)
-    - touch on other stuff possible: priors, Bayes, barrier meshes, residuals, 
-      simulation from joint precision, ...?
+<!--     - perhaps a paragraph on how it achieves what Jim outlined @thorson2019 as necessary: -->
+<!--         -   "VAST represents four main design principles.
+Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
+<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
+<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
+<!--         -   Condition on missing covariates (check, random fields) -->
+<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
+
+<!-- sdmTMB in its current form achieves 3 of the 4 design principles outlined by @thorson2019 for index standardization:  -->
+
+<!-- (e.g., 'catchability' and 'habitat' covariates in VAST merely depend on the values supplied to `predict()` in sdmTMB). -->
 
@@
 
+# Discussion
+
+Take home, re-iterate paragraph?
+
+There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru.
+VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality (e.g., VAST lacks penalized smoothers and the Matérn PC prior).
+The difference lies mainly in user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
+Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists.
+For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1`).
+However, for users who require multivariate spatiotemporal models or the Poisson-link delta model (REF), there is no substitute for VAST.
+
+inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
+For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of models specification (Supp fig?).
+Two major functions included in sdmTMB but not INLA/inlabru are anisotropy and non-linear threshold covariates (TODO: actually true??).
+We also think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
+However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
+
+Final paragraph touching on other stuff possible in one paragraph: priors, Bayes, barrier meshes, randomized quantile residuals, simulation from joint precision, ...?
+
+# Acknowledgements
+
+# Conflict of Interest statement
+
+# Author Contributions
+
 # Supporting Information
+
+## Model description
 
@@
 ```
-
-# Implementation and Availability
-
-- CRAN before?
--
-# Performance
-
-- Probably some discussion of speed.
-Comparison of one of the models in the application above to INLA?
-Or other software, e.g.
-Stan -- though that's not as fair :).
-[EW: at least for the rainfall example I found inlabru to be about twice as fast as sdmTMB]
-- Probably not extensive space for simulation testing, but could do some minimal example and point to vignettes
-
-# Discussion
-
-# Conclusion
-
-Optional
-
-# Acknowledgements
-
-# Conflict of Interest statement
-
-# Author Contributions
 

### 2021-11-02 doc/paper.Rmd
Edit

@@
 inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
-For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of models specification (Supp fig?).
+For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
 Two major functions included in sdmTMB but not INLA/inlabru are anisotropy and non-linear threshold covariates (TODO: actually true??).

### 2021-11-02 doc/paper.Rmd
mgcv brain dump; likely not much space

@@
 However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
+
+mgcv, but, yikes, do we have space:
+
+- mgcv is capable of fitting with the SPDE (REF)
+- typical smoothers don't offer MVN interpretability
+- Tweedie is a big factor... must fix power 'p' to fit a GAMM.
+- Barrier approaches are different
+- doesn't have functionality built in for indexes/COG and uncertainty on such derived parameters
+- for this class of models, we think sdmTMB has advantages, but mgcv is an immensely powerful modelling package that extends well beyond spatial/spatiotemporal data.
 

### 2021-11-03 doc/paper.Rmd
movign things + couple edits

@@
 Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling such random effects is with Gaussian random fields (GRFs), where the random effects are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
 
@@
 Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
+
+<!-- Paragraph on advanced options here?
+Specifically, this could address
+- PC-priors -- and other penalized priors
+- Non-linear stuff / etc
+- Linkages to Rstan
+- Simulation
+-->
 

### 2021-11-03 bbs_example.R

@@ -1,0 +1,85 @@
+library(bbsBayes)
+
+fetch_bbs_data() # takes a few minutes, run once
+# break BBS data into lon-lat 1 deg squares
+stratified_data <- stratify(by = "latlong")
+# Pull out species of interest, e.g. Smith et al. PLoS
+# https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0130768
+
+#Olive-sided Flycatcher in Quebec/BCR 12
+indx <- grep("Olive-sided Flycatcher", stratified_data$species_strat$english)
+
+# Pull out info from all routes
+all_routes <- stratified_data$route_strat
+# join data
+stratified_data$bird_strat <- dplyr::filter(stratified_data$bird_strat,
+                                            AOU == as.numeric(stratified_data$species_strat$aou[indx]))
+# join in route info
+dat <- dplyr::left_join(stratified_data$route_strat, stratified_data$bird_strat)
+
+# grab routes for prediction later
+on_routes = dplyr::filter(stratified_data$route_strat, St_Abrev=="ON") %>%
+  dplyr::group_by(Route) %>%
+  dplyr::summarize(Latitude=Latitude[1],Longitude=Longitude[1])
+
+# pick out canadian data
+dat <- dplyr::filter(dat, St_Abrev %in% c("AB", "BC","MB","NB","NS","NT",
+                                          "NU","SK",
+                                          "ON","QC","YT"))
+
+# StopTotal is the total for all stops
+dat=dplyr::select(dat, Year,StopTotal,
+                  Latitude,Longitude,Month,Day,ObsN)
+
+
+# filter based on longitude
+dat = dplyr::filter(dat,Year >= 1970)
+
+library(sdmTMB)
+mesh = make_mesh(data = dat,
+                 xy_cols = c("Longitude","Latitude"),
+                 cutoff = 1)
+
+dat$date = paste0(dat$Month,"-",dat$Day,"-",dat$Year)
+dat$date = lubridate::parse_date_time(dat$date,orders="mdy")
+dat$jday = lubridate::yday(dat$date)
+dat$StopTotal[which(is.na(dat$StopTotal))] <- 0
+dat$present = ifelse(dat$StopTotal>0,1,0)
+
+ggplot(dat, aes(Longitude, Latitude,col=present)) +
+  geom_point() +
+  facet_wrap(~Year)
+# full model with smooths -- does better than fixed
+# effects or modeling them separate
+
[…]

### 2021-11-05 doc/paper.Rmd
Example of paragraph on advanced uses

@@
 -->
+
+While not of interest to all users, the sdmTMB package also implements several advanced options for model fitting.
+First, models can include penalized likelihoods via assigning priors to all model parameters.
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces.
+Following other recent SPDE approaches [@osgood-zimmerman2021; @simpson2017], these 'penalized complexity' priors may be particularly useful in assigning priors to the range and variance.
+Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to user of `mgcv`.
+Third, for users interested in Bayesian inference, we allow models constructed with `sdmTMB` to be passed to Stan (cite) via the package tmbstan [@monnahan2018].
 

### 2021-11-15 doc/appendix-pcod-sdmTMB.Rmd
Initial draft of pcod vignette

---
title: "Annotated example of index standardization with sdmTMB"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
theme_set(theme_minimal())
```

Let's perform index standardization with the built-in data for Pacific cod. The density units should be kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between v vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in (examples included in `?make_mesh`).
For example, a mesh with a cutoff distance of 10km can be made with

```{r mesh, echo = TRUE, cache=FALSE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(mesh)
```

Let's fit a GLMM. We include `0 + as.factor(year)` so that there is a factor predictor that represents the mean estimate for each time slice. Setting `silent = FALSE` would give us optimization details as the model fits but clutter the Rmd output.

```{r fit-pcod, warning=FALSE, message=FALSE}
fit <- sdmTMB(density ~ 0 + as.factor(year), data = pcod,
  time = "year", mesh = mesh, family = tweedie(link = "log"),
  silent = TRUE
)
```

Print the model:

```{r print-fit}
print(fit)
```

Extract the parameter estimates as a data frame:

```{r tidy}
tidy(fit, conf.int = TRUE)
tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

We can predict on the original data locations:

```{r predict}
pred <- predict(fit)
```

We can inspect randomized quantile residuals:

```{r qqnorm, out.width="5in"}
pred$resid <- residuals(fit)
qqnorm(pred$resid)
qqline(pred$resid)
```

```{r plot-resids}
ggplot(pred, aes(X, Y, colour = resid)) +
    geom_point() + facet_wrap(~year) +
    coord_fixed() + scale_colour_gradient2()
```

Now we want to predict on a fine-scale grid over the entire survey domain. There is a grid built into the sdmTMB package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above.

Now make the predictions on new data.

```{r predict-newdata}
pred_qcs <- predict(fit, newdata = qcs_grid)
```

Let's make a small function to make maps.

```{r plot-map}
plot_map <- function(dat, column) {
  ggplot(dat, aes_string("X", "Y", fill = column)) +
    geom_raster() +
    facet_wrap(~year) +
    coord_fixed()
}
```

There are four kinds of predictions that we get out of the model. First we will show the predictions that incorporate all fixed effects and random effects:

```{r plot-all-effects}
plot_map(pred_qcs, "exp(est)") +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("Prediction (fixed effects + all random effects)")
```

We can also look at just the fixed effects, here year:

```{r plot-fix-defects}
plot_map(pred_qcs, "exp(est_non_rf)") +
  ggtitle("Prediction (fixed effects only)") +
  scale_fill_viridis_c(trans = "sqrt")
```

We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

```{r plot-spatial-effects}
plot_map(pred_qcs, "omega_s") +
  ggtitle("Spatial random effects only") +
  scale_fill_gradient2()
```

And finally we can look at the spatiotemporal random effects that represent deviation from the fixed effect predictions and the spatial random effect deviations. These represent biotic and abiotic factors that are changing through time and are not accounted for in the model.

```{r plot-spatiotemporal-effects}
plot_map(pred_qcs, "epsilon_st") +
  ggtitle("Spatiotemporal random effects only") +
  scale_fill_gradient2()
```

The fastest way to assess uncertainty on our spatial predictions is with simulation from the ...

```{r sims}
pred_sims <- predict(fit, newdata = qcs_grid, sims = 500)
dim(pred_sims)
```

```{r sims-cv-plot}
qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
ggplot(qcs_grid, aes(X, Y, fill = cv)) +
    geom_raster() + facet_wrap(~year) +
    coord_fixed() + scale_fill_viridis_c(trans = "log10", option = "D")
```

```{r index, out.width="5in"}
qcs_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit, newdata = qcs_grid, return_tmb_object = TRUE,
  area = qcs_grid$area
)
ind <- get_index(pred2)
ggplot(ind, aes(year, est/1000)) + geom_line() +
  geom_ribbon(aes(ymin = lwr/1000, ymax = upr/1000), alpha = 0.4) +
  ylab("Biomass (t)")
```

```{r index-sims, out.width="5in"}
ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)
ggplot(ind, aes(year, est/1000)) + geom_line() +
  geom_ribbon(aes(ymin = lwr/1000, ymax = upr/1000), alpha = 0.4) +
  ylab("Biomass (t)")
```

### 2021-11-15 doc/paper.Rmd
Initial draft of pcod vignette

@@
 
-```{r libs, echo = TRUE}
-library(sdmTMB)
-library(ggplot2)
-```
-
-```{r}
-theme_set(ggsidekick::theme_sleek())
-```
-
-First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
-The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between v vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in (examples included in `?make_mesh`).
-For example, an mesh with a cutoff distance of 10km can be made with
-
-```{r mesh, echo = TRUE, cache=FALSE}
-mesh <- make_mesh(pcod_2011, xy_cols = c("X", "Y"), cutoff = 10)
-```
-
-```{r plot-mesh, echo = TRUE, fig.height=5, fig.width=5, out.width="3.25in"}
-plot(mesh)
-```
-
-```{r fit-pcod, cache=TRUE, echo = TRUE}
-fit <- sdmTMB(density ~ 0 + as.factor(year) + s(depth), data = pcod_2011,
- time = "year", spde = mesh, family = tweedie(link = "log")
-)
-```
-
-```{r print-fit, echo = TRUE, cache=TRUE}
-print(fit)
-```
-
-```{r tidy, eval=TRUE, echo=TRUE}
-tidy(fit, conf.int = TRUE)
-```
-
-```{r, eval=TRUE, echo=TRUE}
-tidy(fit, effects = "ran_pars", conf.int = TRUE)
-```
-
-```{r predict, cache=TRUE, echo = TRUE}
-pred <- predict(fit)
-```
-
-```{r qqnorm, cache=TRUE, fig.width=4, fig.height=4, out.width="3.5in", echo = TRUE}
-pred$resid <- residuals(fit)
-# qqnorm(pred$resid);qqline(pred$resid) # not shown for space
-```
-
-```{r plot-resids, echo = FALSE, eval=FALSE}
-ggplot(subset(pred, year == 2017), aes(X, Y, colour = resid)) +
-    geom_point() + facet_wrap(~year) +
-    coord_fixed() + scale_colour_gradient2()
-```
-
-```{r predict-newdata, cache=TRUE, echo = TRUE}
-qcs_grid_2011 <- subset(qcs_grid, year >= 2011)
-pred_qcs <- predict(fit, newdata = qcs_grid_2011)
-```
-
-```{r plot-predictions, echo = TRUE, fig.height=5, fig.width=5, out.width="4.5in", cache=TRUE}
-ggplot(pred_qcs, aes(X, Y, fill = exp(est))) +
-    geom_raster() + facet_wrap(~year) +
-    coord_fixed() + scale_fill_viridis_c(trans = "sqrt")
-```
-
-```{r marginal, cache=TRUE, echo = TRUE, fig.height=2.5, fig.width=4, out.width="3.5in", echo = TRUE}
-nd <- data.frame(depth = seq(50, 300, length.out = 100), year = 2017L)
-pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = ~ 0)
-ggplot(pred_depth, aes(depth, exp(est),
-  ymin = exp(est - 1.96 * est_se), ymax = exp(est + 1.96 * est_se))) +
-  geom_line() + geom_ribbon(alpha = 0.4) +
-  ylab("Expected density\n(kg/TODO)") +
-  xlab("Depth (m)")
-```
-
-```{r sims, cache=TRUE, echo=TRUE}
-pred_sims <- predict(fit, newdata = qcs_grid_2011, sims = 500)
-dim(pred_sims)
-```
-
-```{r sims-cv-plot, cache=TRUE, fig.height=4, fig.width=4, out.width="3.5in", echo=TRUE}
-qcs_grid_2011$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
-ggplot(subset(qcs_grid_2011, year == 2011), aes(X, Y, fill = cv)) +
-    geom_raster() + facet_wrap(~year) +
-    coord_fixed() + scale_fill_viridis_c(trans = "log10")
-```
-
-```{r index, cache=TRUE, fig.height=2.5, fig.width=4, out.width="3.5in"}
-qcs_grid_2011$area <- 4 # all 2 x 2km
-pred2 <- predict(fit, newdata = qcs_grid_2011, return_tmb_object = TRUE,
-  area = qcs_grid_2011$area)
-ind <- get_index(pred2)
-ggplot(ind, aes(year, est)) + geom_line() +
-  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
-  ylab("Biomass (kg)")
-```
-
 <!--

### 2021-11-15 analysis/compare-bruinla.R
Here's 2 examples of using the inlabru() approach on the presence-absence pcod data. I can get the 2 approaches to have the same coefficients, but haven't figured out why they're slightly off from sdmTMB

@@ -3,73 +3,140 @@
 library(INLA)
 library(inlabru)
 library(RColorBrewer)
-data(gorillas)
+library(sdmTMB)
 
-gnests <- gorillas$nests
-mesh <- gorillas$mesh
-boundary <- gorillas$boundary
-gcounts <- gorillas$plotsample$counts
-plots <- gorillas$plotsample$plots
-plotnests <- gorillas$plotsample$nests
+d <- pcod_2011
 
-fig.plots <- ggplot() +
-  gg(boundary) +
-  gg(plots) +
-  gg(gnests, size=0.07,color = "blue") +
-  gg(plotnests,size=0.07,color = "red") +
-  geom_label(aes(label=gcounts$count, x=gcounts$x+0.35, y=gcounts$y-0.3),
-    fill = "white", size = 1.5, label.padding = unit(0.1, "lines")) +
-  coord_fixed() + xlab("x") + ylab("y")
-fig.plots
+# mesh in inla
+# mesh <- inla.mesh.2d(loc = d[,c("X","Y")], cutoff = 5)
+mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 5)
+plot(mesh)
+# spde with same priors as in the INLA example
+spde <- inla.spde2.pcmatern(mesh = mesh$mesh,
+                            prior.range = c(10, 0.05), # P(range < 10) = 0.05
+                            prior.sigma = c(2, 0.05)) # P(sigma > 2) = 0.05
 
-# Inference using Poisson count likelihood
-# Define random variables
+# fit sdmtmb model
+fit_sdmtmb = sdmTMB(present ~ 0 + as.factor(year), data = d, spde = mesh, time = "year",
+              fields = "IID", include_spatial = FALSE, family = binomial(link = "logit"),
+              priors = sdmTMBpriors(
+                matern_st = pc_matern(range_gt = 10, range_prob = 0.05,
+                                      sigma_lt = 2, sigma_prob = 0.05)
+              ))
 
 constr <- FALSE
-mdl <- count ~ spat(main = coordinates,
-  model = inla.spde2.matern(mesh, constr = constr)) +
-  Intercept
+# set this up so year effects are independent fixed effects. This first approach is similar
+# to this thread: https://githubmemory.com/repo/inlabru-org/inlabru/issues/87
+components_1 <- present ~ 0 +
[…]

### 2021-11-15 doc/appendix-pcod-sdmTMB.Rmd

@@
 
-Let's perform index standardization with the built-in data for Pacific cod. The density units should be kg/km^2^. Here, X and Y are coordinates in UTM zone 9.
+We will fit a geostatistical spatiotemporal model with sdmTMB for the purposes of index standardization.
+We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
+The density units are kg/km^2^.
+Here, X and Y are coordinates in UTM zone 9.
 
 First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
-The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between v vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in (examples included in `?make_mesh`).
-For example, a mesh with a cutoff distance of 10km can be made with
+The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
+Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in. See examples included in `?make_mesh`.
+
+For example, a mesh with a cutoff distance of 10 km can be made with
 
@@
 
-Let's fit a GLMM.
-We include `0 + as.factor(year)` so that there is a factor predictor that represents the mean estimate for each time slice.
-Setting `silent = FALSE` would give us optimization details as the model fits but clutter the Rmd output.
+Next we will fit a GLMM (generalized linear mixed effects model).
+For this example, we will include `0 + as.factor(year)` so that there is a factor predictor that represents the mean estimate for each time slice.
+Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.
 
@@
 fit <- sdmTMB(density ~ 0 + as.factor(year), data = pcod,
-  time = "year", mesh = mesh, family = tweedie(link = "log"),
-  silent = TRUE
+  time = "year", mesh = mesh, family = tweedie(link = "log")
 )
@@
 
-Print the model:
+Print the model to inspect the fit:
 
@@
 
-Now we want to predict on a fine-scale grid over the entire survey domain.
+Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year.
 There is a grid built into the sdmTMB package for Queen Charlotte Sound (named `qcs_grid`).
 Our prediction grid also needs to have all the covariates that we used in the model above.
-
-Now make the predictions on new data.
+In this case, that is a `year` column as well as `X` and `Y`.
 
@@
 
-Let's make a small function to make maps.
+Next we will make a small function to help plot maps.
 
@@
 
-We can also look at just the fixed effects, here year:
+We can also look at just the fixed effects, here `year`:
 
@@
 
-And finally we can look at the spatiotemporal random effects that represent deviation from the fixed effect predictions and the spatial random effect deviations.
-These represent biotic and abiotic factors that are changing through time and are not accounted for in the model.
+Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations.
+These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model.
 
@@
 
-The fastest way to assess uncertainty on our spatial predictions is with simulation from the ...
+The fastest way to assess uncertainty on our spatial predictions is with simulation from the joint precision matrix.
 
@@
 ```
+
+This returns a matrix with a row for each row of the `newdata` data frame and a column for each replicate draw. We can use this matrix to calculate the standard deviation or coefficient of variation:
 
@@
 ```
+
+We can calculate our standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE`.
+This returns the TMB objective function, which is required to pass the output to `get_index()`.
 
@@
 
+The above example used the default `bias_correct = FALSE` in `get_index()`.
+This is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
+Setting `bias_correct = TRUE` implements the generic by his correction algorithm that should correct for this, but can take considerable time and memory.
+
+As an alternative, we can use the output from the joint precision matrix simulation draws to calculate our index.
+This should approximate the output of `get_index(x, bias_correct = TRUE)`.
+However, we note that there may be situations where this approach may fail or be biased.
+
 ```{r index-sims, out.width="5in"}

### 2021-11-15 analysis/compare-inlabru.R
inlabru examples

@@ -18,18 +18,18 @@ spde <- inla.spde2.pcmatern(mesh = mesh$mesh,
 
 # fit sdmtmb model
 fit_sdmtmb = sdmTMB(present ~ 0 + as.factor(year), data = d, spde = mesh, time = "year",
-              fields = "IID", include_spatial = FALSE, family = binomial(link = "logit"),
-              priors = sdmTMBpriors(
-                matern_st = pc_matern(range_gt = 10, range_prob = 0.05,
-                                      sigma_lt = 2, sigma_prob = 0.05)
-              ))
+                    fields = "IID", include_spatial = FALSE, family = binomial(link = "logit"),
+                    priors = sdmTMBpriors(
+                      matern_st = pc_matern(range_gt = 10, range_prob = 0.05,
+                                            sigma_lt = 2, sigma_prob = 0.05)
+                    ))
 
 constr <- FALSE
 # set this up so year effects are independent fixed effects. This first approach is similar
 # to this thread: https://githubmemory.com/repo/inlabru-org/inlabru/issues/87
 components_1 <- present ~ 0 +
   fac(main = year, model = "factor_full") +
-      #hyper = list(prec = list(fixed = TRUE))) +
+  #hyper = list(prec = list(fixed = TRUE))) +
   spatrf(main = coordinates,model = spde, group=year, ngroup=4,
          control.group = list(model = "iid"))
 
@@ -40,7 +40,7 @@ components_1 <- present ~ 0 +
 # as.factor(year)2015     0.18    0.45
 # as.factor(year)2017    -0.65    0.45
 fit_1 <- bru(components = components_1,
-           family = "binomial", data = pcod_2011)
+             family = "binomial", data = pcod_2011)
 #print(fit)
 
 # look at effects

### 2021-11-18 doc/paper.Rmd
Edit; develop basic paper example

@@
       fig_caption: true
+      highlight: "monochrome"
 csl: mee.csl
@@
 library(ggplot2)
+theme_set(theme_minimal())
+# theme_set(ggsidekick::theme_sleek())
 library(dplyr)
@@
 knitr::opts_chunk$set(
-  echo = FALSE, message = FALSE, warning = FALSE,
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
   collapse = TRUE,
@@
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-This residual correlation is ideally dealt with because doing so (1) results in valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].
+This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].
 
 Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
-Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial/spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
+Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
 A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
@@
 Template systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
-It can also be slow to iterate on and experiment with many models when writing bespoke template code.
+For example, TMB requires the user to program in a C++ like template and it can be slow to experiment with multiple models when writing bespoke model code.
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
@@
 
-Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA.
+Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
 Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
 One common application is for species distribution models (SDMs), hence the package name.
-This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through X case studies, and concludes with a discussion of additional functionality and related software.
+This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of additional functionality and related software.
 
@@
 
-Spatial and spatiotemporal models constructed with sdmTMB may be viewed as spatially explicit models, in the sense that they estimate parameters of a user-specified spatial covariance function, and these estimates may be converted into interpretable quantities common in spatial statistics (spatial decay, range).
-We contrast these approaches with semi- and non-parametric approaches that do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], some smooths in mgcv [@wood2017a]).
-By default, sdmTMB implements the Matérn covariance function (REF) (cite Rue et al?)---this function can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in a latitudinal direction may be different than in the longitudinal direction).
+Models constructed with sdmTMB may be viewed as spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+We contrast these approaches with semi- and non-parametric approaches that do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
+sdmTMB implements the Matérn covariance function [@matern1960], which can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
 
@@
 
+# An introductory example
+
+An sdmTMB model is fit by supplying a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
+It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region.
+
+Here, we illustrate a simple spatial model fit to Pacific Cod (TODO) trawl survey data from Queen Charlotte Sound.
+Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
+Our data frame `d` has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in m of that tow, and spatial coordinates X and Y, which are UTM coordinates in km:
+
+```{r, include=FALSE}
+dat <- dplyr::select(pcod, year, density, depth, X, Y)
+```
+
+```{r, echo=TRUE, eval=FALSE}
+head(dat)
+```
+
+```{}
+#>    year density depth     X     Y
+#>   <int>   <dbl> <dbl> <dbl> <dbl>
+#> 1  2003   113.    201  446. 5793.
+#> 2  2003    41.7   212  446. 5800.
+#> 3  2003     0     220  449. 5802.
+```
+
+We must then create a mesh object that contains matrices to apply the SPDE approach:
+
+```{r, echo=TRUE}
+mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 10)
+```
+
+Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`.
+We can then inspect our mesh object with the associated `plot()` method (see Fig.
+\@ref(fig:pcod-fig)a):
+
+```{r, eval=FALSE, echo=TRUE}
+plot(mesh)
+```
+
+And use our data frame and mesh to fit a spatial model:
+
+```{r pcod-eg1-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE}
+fit <- sdmTMB(
+  density ~ s(depth),
+  data = dat,
+  family = tweedie(link = "log"),
+  mesh = mesh,
+  spatial = 'on'
+)
+```
+
+We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space):
+
+```{r,eval=FALSE, echo=TRUE}
+print(fit)
+```
+
+<!-- tidy(fit, conf.int = TRUE) -->
+<!-- tidy(fit, effects = 'ran_pars', conf.int = TRUE) -->
+
+```{}
+#> Spatiotemporal model fit by ML ['sdmTMB']
+#> Formula: density ~ s(depth)
+#> Mesh: mesh
+#> Data: d
+#> Family: tweedie(link = 'log')
+#>             coef.est coef.se
+#> (Intercept)     2.37    0.21
+#>
+#> Dispersion parameter: 12.69
+#> Tweedie p: 1.58
+#> Matern range: 16.39
+#> Spatial SD: 1.86
+#> ML criterion at convergence: 6402.136
+```
+
+The output indicates our model was fit by (marginal) maximum likelihood ('ML').
+We also see the formula, mesh, fitted data, and family.
+Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the \mbox{Mat\'ern} range distance, marginal spatial standard deviation, and the negative log likelihood at convergence.
+
+We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
+The data frame `qcs_grid` is a grid that extends over the full studied domain and contains all predictors used in the model set at values we wish to predict at.
+
+```{r, echo=TRUE, eval=FALSE}
+p_new <- predict(fit, newdata = qcs_grid)
+```
+
+We show a basic plot of spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
+
+```{r, echo=FALSE, eval=TRUE}
+p_new <- predict(fit, newdata = subset(qcs_grid, year == 2003))
+```
+
+```{r}
+g_pred <- ggplot(p_new, aes(X, Y, fill = exp(est))) + geom_raster() +
+  scale_fill_viridis_c(trans = "sqrt") + coord_fixed() +
+  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
+  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+```
+
+```{r, echo=FALSE}
+blues <- RColorBrewer::brewer.pal(5, "Blues")
+g_mesh <- ggplot(dat, aes(X, Y)) +
+  inlabru::gg(mesh$mesh, edge.color = "grey60") +
+  geom_point(alpha = 0.6, mapping = aes(size = density), colour = blues[4]) +
+  theme_minimal() + coord_fixed() +
+  labs(x = "UTMs E (km)", y = "UTMs N (km)") +
+  scale_size_area(max_size = 10) +
+  theme(legend.position = "none")
+```
+
+```{r}
+nd <- data.frame(depth =
+  seq(min(dat$depth), max(dat$depth), length.out = 300))
+p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
+g_smooth <- ggplot(p_smooth,
+  aes(depth, exp(est), ymin = exp(est - 1.96 * est_se),
+    ymax = exp(est + 1.96 * est_se))) +
+  geom_ribbon(fill = "grey65") +
+  geom_line(colour = "grey10") +
+  labs(x = "Depth (m)", y = "Density (units)") +
+  xlim(min(dat$depth), 300) +
+  coord_fixed(expand = FALSE, ratio = 1.2)
+```
+
+```{r pcod-fig, fig.cap="example output", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
+g_omega <- ggplot(p_new, aes(X, Y, fill = omega_s)) +
+  geom_raster() +
+  scale_fill_gradient2(high = scales::muted("red"), low = scales::muted("blue")) + coord_fixed() +
+  labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
+  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
+```
+
+We could extend our model to a spatiotemporal one simply by supplying the `'year'` column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
+
+```{r, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=TRUE}
+fit_spatiotemporal <- sdmTMB(
+  density ~ s(depth), family = tweedie(link = "log"), data = dat, mesh = mesh
+  time = "year", spatial = 'off', spatiotemporal = 'ar1'
+)
+```
+
 <!--
@@
 There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru.
-VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality (e.g., VAST lacks penalized smoothers and the Matérn PC prior).
-The difference lies mainly in user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
+VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
+The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
 Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists.
@@
 For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
-Two major functions included in sdmTMB but not INLA/inlabru are anisotropy and non-linear threshold covariates (TODO: actually true??).
-We also think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
+One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
+We think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB, and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
 However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

### 2021-11-18 doc/paper.Rmd
Work on example
Much to be deleted I think

@@
 
-Here, we illustrate a simple spatial model fit to Pacific Cod (TODO) trawl survey data from Queen Charlotte Sound.
+Here, we illustrate a spatial model fit to Pacific Cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `d` has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in m of that tow, and spatial coordinates X and Y, which are UTM coordinates in km:
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in m of that tow, and spatial coordinates X and Y, which are UTM coordinates in km:
 
 ```{r, include=FALSE}
-dat <- dplyr::select(pcod, year, density, depth, X, Y)
+pcod <- dplyr::select(pcod, year, density, depth, X, Y)
 ```
@@
 ```{r, echo=TRUE, eval=FALSE}
-head(dat)
+head(pcod)
 ```
@@
 ```{r, echo=TRUE}
-mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 10)
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
 ```
 
-Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`.
-We can then inspect our mesh object with the associated `plot()` method (see Fig.
-\@ref(fig:pcod-fig)a):
+Here, `cutoff` defines the minimum allowed distance between points.
+Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh(mesh = ...)`. 
+We can then inspect our mesh object with the associated `plot()` method: `plot(mesh)` (see Fig. \@ref(fig:pcod-fig)a):
 
-```{r, eval=FALSE, echo=TRUE}
+```{r, eval=FALSE, echo=FALSE}
 plot(mesh)
@@
   density ~ s(depth),
-  data = dat,
+  data = pcod,
   family = tweedie(link = "log"),
@@
 #> Mesh: mesh
-#> Data: d
+#> Data: pcod
 #> Family: tweedie(link = 'log')
@@
 ```{r, echo=TRUE, eval=FALSE}
-p_new <- predict(fit, newdata = qcs_grid)
+p <- predict(fit, newdata = qcs_grid)
 ```
 
-We show a basic plot of spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
 
 ```{r, echo=FALSE, eval=TRUE}
-p_new <- predict(fit, newdata = subset(qcs_grid, year == 2003))
+p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
 ```
@@
 ```{r}
-g_pred <- ggplot(p_new, aes(X, Y, fill = exp(est))) + geom_raster() +
-  scale_fill_viridis_c(trans = "sqrt") + coord_fixed() +
+g_pred <- ggplot(p, aes(X, Y, fill = exp(est))) + geom_raster() +
+  scale_fill_viridis_c(trans = "sqrt", option = "C") + coord_fixed() +
   labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
@@
 
+```{r, eval=FALSE}
+p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
+se <- apply(p3, 1, sd)
+p$se <- se
+g_pred_se <- ggplot(p, aes(X, Y, fill = se)) + geom_raster() +
+  scale_fill_viridis_c(trans = "log10", option = "C") + coord_fixed() +
+  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
+  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+```
+
+```{r, warning=FALSE, message=FALSE, eval=FALSE}
+library(sf)
+map_data <- rnaturalearth::ne_countries(
+  scale = "large",
+  returnclass = "sf", country = "canada")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+bc_coast <- suppressWarnings(suppressMessages(
+  st_crop(map_data,
+    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
+utm_zone9 <- 3156
+bc_coast_proj <- sf::st_transform(bc_coast, crs = utm_zone9)
+# ggplot(bc_coast_proj) + geom_sf()
+```
+
+```{r, eval=FALSE}
+ggplot(bc_coast_proj) + geom_sf() +
+  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
+  xlim(230957.7 + 100000, 1157991 - 500000) +
+  ylim(5366427 + 200000, 6353456 - 500000) +
+  scale_fill_viridis_c() +
+  theme_light() +
+  labs(fill = "Predicted\ndensity") +
+  labs(x = "Longitude", y = "Latitude")
+```
+
 ```{r, echo=FALSE}
 blues <- RColorBrewer::brewer.pal(5, "Blues")
-g_mesh <- ggplot(dat, aes(X, Y)) +
+reds <- RColorBrewer::brewer.pal(5, "Reds")
+g_mesh <- ggplot(pcod, aes(X, Y)) +
   inlabru::gg(mesh$mesh, edge.color = "grey60") +
@@
 nd <- data.frame(depth =
-  seq(min(dat$depth), max(dat$depth), length.out = 300))
+  seq(min(pcod$depth), max(pcod$depth), length.out = 300))
 p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
@@
     ymax = exp(est + 1.96 * est_se))) +
-  geom_ribbon(fill = "grey65") +
-  geom_line(colour = "grey10") +
+  geom_ribbon(fill = blues[2]) +
+  geom_line(colour = blues[5], lwd = 1) +
   labs(x = "Depth (m)", y = "Density (units)") +
-  xlim(min(dat$depth), 300) +
-  coord_fixed(expand = FALSE, ratio = 1.2)
+  xlim(min(pcod$depth), 300) +
+  coord_fixed(expand = FALSE, ratio = 1.28)
 ```
 
-```{r pcod-fig, fig.cap="example output", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(p_new, aes(X, Y, fill = omega_s)) +
+```{r, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=TRUE}
+fit2 <- sdmTMB(
+  density ~ s(depth), 
+  family = tweedie(link = "log"), data = pcod, mesh = mesh,
+  time = 'year', spatial = 'off', spatiotemporal = 'ar1'
+)
+```
+
+```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
+fit2 <- sdmTMB(
+  density ~ 0 + as.factor(year), 
+  family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
+  time = 'year', spatial = 'on', spatiotemporal = 'iid', silent = TRUE
+)
+p2 <- predict(fit2, newdata = qcs_grid, sims = 500L)
+ind <- get_index_sims(p2)
+```
+
+```{r plot-ind, eval=FALSE}
+g_ind <- ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) + 
+  geom_ribbon(fill = blues[2]) +
+  geom_line(colour = blues[5], lwd = 1) +
+  ylim(0, max(ind$upr)) +
+  coord_fixed(expand = FALSE, ratio = 0.3) +
+  labs(x = "Year", y = "Biomass (units)")
+```
+
+(ref:pcod-fig-cap) Plots from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada.
+(a) The SPDE mesh (lines) combined with the trawl survey observations.
+Finer will be slower to fit.
+Circle area corresponds to density of Pacific Cod for individual trawls.
+(b) Conditional effect of depth modelled as a penalized smoother, `s(depth)`.
+Line indicates estimate and ribbon indicates a 95\% confidence interval (+/- 2 SEs).
+These predictions are made omitting the spatial random field.
+(c) The spatial random field:
+these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
+(d) The overall prediction: these estimates come from combining all fixed and random effect estimates.
+
+```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
+g_omega <- ggplot(p, aes(X, Y, fill = omega_s)) +
   geom_raster() +
-  scale_fill_gradient2(high = scales::muted("red"), low = scales::muted("blue")) + coord_fixed() +
+  # scale_fill_gradient2(high = scales::muted("red"), low = scales::muted("blue")) + coord_fixed() +
+  scale_fill_gradient2(high = reds[4], low = blues[4]) + coord_fixed() +
   labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
@@
 
-We could extend our model to a spatiotemporal one simply by supplying the `'year'` column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
 Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
@@
 fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth), family = tweedie(link = "log"), data = dat, mesh = mesh
-  time = "year", spatial = 'off', spatiotemporal = 'ar1'
+  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh
+  time = 'year', spatial = 'off', spatiotemporal = 'ar1'
 )

### 2021-11-18 doc/paper.Rmd
Develop example

@@
 
-We must then create a mesh object that contains matrices to apply the SPDE approach:
+We start by creating a mesh object that contains matrices to apply the SPDE approach:
 
@@
 
-Here, `cutoff` defines the minimum allowed distance between points.
+Here, `cutoff` defines the minimum allowed distance between points, here in km.
 Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh(mesh = ...)`. 
-We can then inspect our mesh object with the associated `plot()` method: `plot(mesh)` (see Fig. \@ref(fig:pcod-fig)a):
+We can then inspect our mesh object with the associated `plot()` method: `plot(mesh)` (see Fig. \@ref(fig:pcod-fig)a).
 
-```{r, eval=FALSE, echo=FALSE}
-plot(mesh)
-```
-
-And use our data frame and mesh to fit a spatial model:
+We can then fit a model with spatial random fields via the function `sdmTMB()`. 
+We use a penalized smoother for depth as a main effect via `mgcv::s()`.
+We specify the family as Tweedie to account for the positive continuous density values that also have zeros.
 
@@
   mesh = mesh,
-  spatial = 'on'
+  spatial = "on"
 )
@@
 
-The output indicates our model was fit by (marginal) maximum likelihood ('ML').
+The output indicates our model was fit by (marginal) maximum likelihood (`'ML'`).
 We also see the formula, mesh, fitted data, and family.
-Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the \mbox{Mat\'ern} range distance, marginal spatial standard deviation, and the negative log likelihood at convergence.
+Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the \mbox{Mat\'ern} range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
 
@@
 The data frame `qcs_grid` is a grid that extends over the full studied domain and contains all predictors used in the model set at values we wish to predict at.
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
 
@@
 ```
-
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
 
@@
 ```{r}
-g_pred <- ggplot(p, aes(X, Y, fill = exp(est))) + geom_raster() +
-  scale_fill_viridis_c(trans = "sqrt", option = "C") + coord_fixed() +
+g_pred <- ggplot(p, aes(X, Y, fill = exp(est))) +
+  geom_raster() +
+  scale_fill_viridis_c(trans = "sqrt", option = "C") +
+  coord_fixed() +
   labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
 ```
@@
 p$se <- se
-g_pred_se <- ggplot(p, aes(X, Y, fill = se)) + geom_raster() +
-  scale_fill_viridis_c(trans = "log10", option = "C") + coord_fixed() +
+g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
+  geom_raster() +
+  scale_fill_viridis_c(trans = "log10", option = "C") +
+  coord_fixed() +
   labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
 ```
@@
 ```{r}
-nd <- data.frame(depth =
-  seq(min(pcod$depth), max(pcod$depth), length.out = 300))
+nd <- data.frame(
+  depth =
+    seq(min(pcod$depth), max(pcod$depth), length.out = 300)
+)
 p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
-g_smooth <- ggplot(p_smooth,
-  aes(depth, exp(est), ymin = exp(est - 1.96 * est_se),
-    ymax = exp(est + 1.96 * est_se))) +
+g_smooth <- ggplot(
+  p_smooth,
+  aes(depth, exp(est),
+    ymin = exp(est - 1.96 * est_se),
+    ymax = exp(est + 1.96 * est_se)
+  )
+) +
   geom_ribbon(fill = blues[2]) +
@@
 
-```{r, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=TRUE}
-fit2 <- sdmTMB(
-  density ~ s(depth), 
-  family = tweedie(link = "log"), data = pcod, mesh = mesh,
-  time = 'year', spatial = 'off', spatiotemporal = 'ar1'
-)
-```
-
 ```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
@@
   family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
-  time = 'year', spatial = 'on', spatiotemporal = 'iid', silent = TRUE
+  time = "year", spatial = "on", spatiotemporal = "iid", silent = TRUE
 )
@@
 
-```{r, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=TRUE}
+```{r, eval=FALSE, echo=TRUE}
 fit_spatiotemporal <- sdmTMB(
   density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh
-  time = 'year', spatial = 'off', spatiotemporal = 'ar1'
+  time = "year", spatial = "off", spatiotemporal = "ar1"
 )

### 2021-11-18 doc/appendix-pcod-inlabru.Rmd
inlabru

---
title: "Annotated example of index standardization with inlabru"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(INLA)
library(inlabru)
theme_set(theme_minimal())
```

This model is an extension of the same vignette using sdmTMB.

```{r mesh, echo = TRUE, cache=FALSE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)

spde <- inla.spde2.pcmatern(mesh = mesh$mesh,
                            prior.range = c(10, 0.05), # P(range < 10) = 0.05
                            prior.sigma = c(2, 0.05)) # P(sigma > 2) = 0.05
```

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(mesh)
```

There are a couple ways to include the fixed effects with `inlabru`. Unfortunately, using factor variables in the standard equation syntax does not work. In the first case, we can create a design matrix and add year effects individually to the formula.

We can fit the model first with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.

```{r}
sdmtmb_fit <- sdmTMB(present ~ -1 + as.factor(year), data = pcod,
   time = "year", spde = mesh, 
   family = binomial(link="logit"), 
   spatial_only=TRUE,
   priors = sdmTMBpriors(
                      matern_st = pc_matern(range_gt = 10, range_prob = 0.05,
                                            sigma_lt = 2, sigma_prob = 0.05)
                    ))
sdmtmb_coefs = tidy(sdmtmb_fit, effects="fixed")
```

Now we can set up the design matrix for `inlabru`

```{r fit-pcod, warning=FALSE, message=FALSE}
m <- lm(present ~ -1 + as.factor(year), data = pcod)
model_mat = model.matrix(m)
colnames(model_mat) = paste0("x",unique(pcod$year))
dat <- cbind(pcod, model_mat)
```

Next we set up the `components` object for the model. This includes a formula with fixed or random effects (here only fixed) and the design of the spatial random field. The name `spatrf` can be anything, but includes the coordinates, spde model, and how the year effects are shared.

```{r}
components <- present ~ -1 + x2003 + x2004 + x2005 + x2007 +
  x2009 + x2011 + x2013 + x2015 + x2017 +
  spatrf(main = coordinates,model = spde) + 
  #spatrf(main = coordinates,model = spde, group=year, ngroup=ncol(model_mat),
  #       control.group = list(model = "iid"))

fit <- bru(components = components,
             family = "binomial", data = dat)
bru_coefs <- fit$summary.fixed[,c("mean","sd")]
```

Comparing the fixed effects between `sdmTMB` and `inlabru` shows the estimates are correlated, but not perfectly so

```{r}
plot(sdmtmb_coefs$estimate, bru_coefs$mean)
```

The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`

```{r}
components <- present ~ 0 +
  fac(main = year, model = "factor_full") + 
  spatrf(main = coordinates,model = spde)

fit2 <- bru(components = components,
             family = "binomial", data = dat)
bru_coefs2 <- fit2$summary.random$fac[,c("mean","sd")]
```

These estimates are perfectly correlated with the estimates above,

```{r}
plot(bru_coefs$mean, bru_coefs2$mean)
```

### 2021-11-18 analysis/compare-inlabru.R

@@ -1,142 +1,0 @@
-# from MEE inlabru paper
-
-library(INLA)
-library(inlabru)
-library(RColorBrewer)
-library(sdmTMB)
-
-d <- pcod_2011
-
-# mesh in inla
-# mesh <- inla.mesh.2d(loc = d[,c("X","Y")], cutoff = 5)
-mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 5)
-plot(mesh)
-# spde with same priors as in the INLA example
-spde <- inla.spde2.pcmatern(mesh = mesh$mesh,
-                            prior.range = c(10, 0.05), # P(range < 10) = 0.05
-                            prior.sigma = c(2, 0.05)) # P(sigma > 2) = 0.05
-
-# fit sdmtmb model
-fit_sdmtmb = sdmTMB(present ~ 0 + as.factor(year), data = d, spde = mesh, time = "year",
-                    fields = "IID", include_spatial = FALSE, family = binomial(link = "logit"),
-                    priors = sdmTMBpriors(
-                      matern_st = pc_matern(range_gt = 10, range_prob = 0.05,
-                                            sigma_lt = 2, sigma_prob = 0.05)
-                    ))
-
-constr <- FALSE
-# set this up so year effects are independent fixed effects. This first approach is similar
-# to this thread: https://githubmemory.com/repo/inlabru-org/inlabru/issues/87
-components_1 <- present ~ 0 +
-  fac(main = year, model = "factor_full") +
-  #hyper = list(prec = list(fixed = TRUE))) +
-  spatrf(main = coordinates,model = spde, group=year, ngroup=4,
-         control.group = list(model = "iid"))
-
-# note that 'd' not passed in -- b/c inlabru doesn't recognize tbl_df etc
-# sdmTMB estimates (code below) are
-# as.factor(year)2011    -0.69    0.45
-# as.factor(year)2013     0.48    0.45
-# as.factor(year)2015     0.18    0.45
-# as.factor(year)2017    -0.65    0.45
-fit_1 <- bru(components = components_1,
-             family = "binomial", data = pcod_2011)
-#print(fit)
-
-# look at effects
-fit_1$summary.random$fac
-
-
-# Second option is to use model.matrix() and have fixed effects
[…]

### 2021-11-18 doc/appendix-pcod-inlabru.Rmd
Fix a few small things; run styler

@@
 
-spde <- inla.spde2.pcmatern(mesh = mesh$mesh,
+spde <- inla.spde2.pcmatern(
+  mesh = mesh$mesh,
   prior.range = c(10, 0.05), # P(range < 10) = 0.05
-                            prior.sigma = c(2, 0.05)) # P(sigma > 2) = 0.05
+  prior.sigma = c(2, 0.05)
+) # P(sigma > 2) = 0.05
 ```
@@
 ```{r}
-sdmtmb_fit <- sdmTMB(present ~ -1 + as.factor(year), data = pcod,
+sdmtmb_fit <- sdmTMB(present ~ -1 + as.factor(year),
+  data = pcod,
   time = "year", spde = mesh,
-   family = binomial(link="logit"), 
-   spatial_only=TRUE,
+  family = binomial(link = "logit"),
+  spatiotemporal = "off",
   priors = sdmTMBpriors(
-                      matern_st = pc_matern(range_gt = 10, range_prob = 0.05,
-                                            sigma_lt = 2, sigma_prob = 0.05)
-                    ))
-sdmtmb_coefs = tidy(sdmtmb_fit, effects="fixed")
+    matern_st = pc_matern(
+      range_gt = 10, range_prob = 0.05,
+      sigma_lt = 2, sigma_prob = 0.05
+    )
+  )
+)
+sdmtmb_coefs <- tidy(sdmtmb_fit, effects = "fixed")
 ```
@@
 m <- lm(present ~ -1 + as.factor(year), data = pcod)
-model_mat = model.matrix(m)
-colnames(model_mat) = paste0("x",unique(pcod$year))
+model_mat <- model.matrix(m)
+colnames(model_mat) <- paste0("x", unique(pcod$year))
 dat <- cbind(pcod, model_mat)
@@
   x2009 + x2011 + x2013 + x2015 + x2017 +
-  spatrf(main = coordinates,model = spde) + 
-  #spatrf(main = coordinates,model = spde, group=year, ngroup=ncol(model_mat),
+  spatrf(main = coordinates, model = spde)
+# spatrf(main = coordinates,model = spde, group=year, ngroup=ncol(model_mat),
 #       control.group = list(model = "iid"))
 
-fit <- bru(components = components,
-             family = "binomial", data = dat)
-bru_coefs <- fit$summary.fixed[,c("mean","sd")]
+fit <- bru(
+  components = components,
+  family = "binomial", data = dat
+)
+bru_coefs <- fit$summary.fixed[, c("mean", "sd")]
 ```
@@
 plot(sdmtmb_coefs$estimate, bru_coefs$mean)
+abline(a = 0, b = 1)
 ```
@@
   fac(main = year, model = "factor_full") +
-  spatrf(main = coordinates,model = spde)
+  spatrf(main = coordinates, model = spde)
 
-fit2 <- bru(components = components,
-             family = "binomial", data = dat)
-bru_coefs2 <- fit2$summary.random$fac[,c("mean","sd")]
+fit2 <- bru(
+  components = components,
+  family = "binomial", data = dat
+)
+bru_coefs2 <- fit2$summary.random$fac[, c("mean", "sd")]
 ```

### 2021-11-18 doc/paper.Rmd
Word
Editing

@@
 
-While not of interest to all users, the sdmTMB package also implements several advanced options for model fitting.
-First, models can include penalized likelihoods via assigning priors to all model parameters.
+While not of interest to all users, sdmTMB also implements several advanced options for model fitting. 
+First, models can include penalized likelihoods via assigning priors to model parameters.
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces.
-Following other recent SPDE approaches [@osgood-zimmerman2021; @simpson2017], these 'penalized complexity' priors may be particularly useful in assigning priors to the range and variance.
-Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to user of `mgcv`.
-Third, for users interested in Bayesian inference, we allow models constructed with `sdmTMB` to be passed to Stan (cite) via the package tmbstan [@monnahan2018].
+Following other recent SPDE approaches [@simpson2017; @osgood-zimmerman2021], 'penalized complexity' priors may be particularly useful in assigning priors to the spatial range and variance.
+Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to users of `mgcv`.
+Third, for users interested in Bayesian inference, we allow models to be passed to Stan (cite) via the package tmbstan [@monnahan2018].
 
@@
 
-An sdmTMB model is fit by supplying a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region.
+At a minimum, an sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
+It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
 
@@
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in m of that tow, and spatial coordinates X and Y, which are UTM coordinates in km:
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
 
 ```{r, include=FALSE}
+# simplify for paper
 pcod <- dplyr::select(pcod, year, density, depth, X, Y)
@@
 
-Here, `cutoff` defines the minimum allowed distance between points, here in km.
-Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh(mesh = ...)`. 
-We can then inspect our mesh object with the associated `plot()` method: `plot(mesh)` (see Fig. \@ref(fig:pcod-fig)a).
+Here, `cutoff` defines the minimum allowed distance between points in the units of `X` and `Y` (km).
+Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
+We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)a).
 
 We can then fit a model with spatial random fields via the function `sdmTMB()`. 
-We use a penalized smoother for depth as a main effect via `mgcv::s()`.
-We specify the family as Tweedie to account for the positive continuous density values that also have zeros.
+We use a penalized smoother for depth as a main effect via `s()` from the mgcv package.
+We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
 
@@
 
-The output indicates our model was fit by (marginal) maximum likelihood (`'ML'`).
+The output indicates our model was fit by maximum (marginal) likelihood (`'ML'`).
 We also see the formula, mesh, fitted data, and family.
-Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the \mbox{Mat\'ern} range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
+Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
 
@@
 
-(ref:pcod-fig-cap) Plots from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada.
-(a) The SPDE mesh (lines) combined with the trawl survey observations.
-Finer will be slower to fit.
-Circle area corresponds to density of Pacific Cod for individual trawls.
-(b) Conditional effect of depth modelled as a penalized smoother, `s(depth)`.
-Line indicates estimate and ribbon indicates a 95\% confidence interval (+/- 2 SEs).
+(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
+(a) SPDE mesh (lines) combined with the trawl survey observations (points).
+Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
+The circle area corresponds to density of Pacific Cod for individual trawls.
+(b) Conditional effect of depth modelled as a penalized smoother.
+The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
 These predictions are made omitting the spatial random field.
-(c) The spatial random field:
+(c) Spatial random field:
 these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
-(d) The overall prediction: these estimates come from combining all fixed and random effect estimates.
+(d) Overall prediction: these estimates represent the combination of all fixed and random effects.
 
@@
 
-We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+We could extend our model to a spatiotemporal model simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
 Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
@@
 
-The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function.
+The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a Matérn covariance function.
 Additional details are provided in the SI.
@@
 <!-- EW: Because of space limits, I might suggest moving all this to the SI -->
-The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
+The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
 (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,

### 2021-11-18 doc/paper.Rmd
Edits to paper

@@
 
+basic optimization details - random effects conditional...
+
 [paragraph about spde and gaussian process / Matern covariance here]
@@
 We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
-The data frame `qcs_grid` is a grid that extends over the full studied domain and contains all predictors used in the model set at values we wish to predict at.
+The data frame `qcs_grid` is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
+The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
 We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.

### 2021-11-18 doc/paper.Rmd
Developing paper

@@
 
-# Examples of sections:
+# Model description
 
-# Materials and Methods
+sdmTMB fits spatially explicit geostatistical GLMMs to point data.
+Point data simply means the collected data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
+The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
+\@ref(fig:diagram)).
+This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
 
-# Description
+Using the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time,
+the general form of a GLMM is
+$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
+with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
+<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
+For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
+Although in practice it is rare to include all of these components, the full sdmTMB model is
+$$
+g(\mu_{\bm{s},t}) =
+\bm{X}_{1,\bm{s},t} \bm{\beta} +
+\bm{\alpha}_g +
+\bm{X}_{2,\bm{s},t} \gamma_t +
+\bm{X}_{3,t} \zeta_{\bm{s}} +
+\omega_{\bm{s}} +
+\epsilon_{\bm{s},t},
+$$
+where
+
+* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
+* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
+* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
+* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
+* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
+* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
+* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
 
 Models constructed with sdmTMB may be viewed as spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-We contrast these approaches with semi- and non-parametric approaches that do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
-sdmTMB implements the Matérn covariance function [@matern1960], which can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
+In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
+The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
+The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
+The Matérn standard deviations are estimated separately for the various components and the range can be shared or estimated separately.
 
-- high-level paragraph or two about what the model includes, referencing Fig 1
-- high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?)
-- high-level paragraph about simulation capabilities and + testing performed (reference supplement)
-- model selection/cross validation paragraph
-- derived quantities paragraph (or just put this in the examples?)
+<!-- - high-level paragraph or two about what the model includes, referencing Fig 1 -->
+<!-- - high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?) -->
+<!-- - high-level paragraph about simulation capabilities and + testing performed (reference supplement) -->
+<!-- - model selection/cross validation paragraph -->
+<!-- - derived quantities paragraph (or just put this in the examples?) -->
 
-- example 1 + figs
-- example 2 + figs
+<!-- - example 1 + figs -->
+<!-- - example 2 + figs -->
 
@@
 
-# Model
+<!-- # Model -->
 
@@
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000] (Fig.\ \@ref(fig:diagram)).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] (Fig.\ \@ref(fig:diagram)).
 
-basic optimization details - random effects conditional...
+The sdmTMB model is fit by maximum marginal likelihood.
+Internally, a TMB model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the optimization routine `stats::nlminb()` in R (REF).
+Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation (REF).
 
-[paragraph about spde and gaussian process / Matern covariance here]
+<!-- [paragraph about spde and gaussian process / Matern covariance here] -->
 
 <!-- Paragraph on model selection here-->
+
+## Model selection
+
 Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
 Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
-Our function `sdmTMB::sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
+Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).
 
-<!-- Paragraph on derived quantities here-->
-The results from a model fit with sdmTMB can be used to generate quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`sdmTMB::get_cog()`) and total population index (`sdmTMB::get_index()`).
+## Quantities of interest
+
+The results from a model fit with sdmTMB can be used to generate various quantities of interest.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
 Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
+
+## Model evaluation
+
+Validation of mixed-effects models is non-trivial (that REF).
+
+We provide three tools to assist with model validation:
+
+1. The `residuals()` method returns randomized quantile residuals (e.g., as in DHARMa REF).
+These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects.
+These residuals have known statistical issues (REF).
+One-step-ahead residuals are also possible or residuals from a single MCMC draw, but both approaches are much slower.
+
+2.
+The `sdmTMB_simulate()` function can simulate new data, or simulate from a fitted model, to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
+
+3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation. 
+We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).
 
@@
 -->
+
+## Advanced options
 
@@
 
-At a minimum, an sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
+An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
 It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
-
 Here, we illustrate a spatial model fit to Pacific Cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
@@
 
-## Model description
-
-We use the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time.
-The general form of a GLMM is
-$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
-with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
-<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
-For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model.
-Though in practice it is rare to include all of these parameters, the full `sdmTMB` model is
-$$
-g(\mu_{\bm{s},t}) =
-\bm{X}_{1,\bm{s},t} \bm{\beta} +
-\bm{\alpha}_g +
-\bm{X}_{2,\bm{s},t} \gamma_t +
-\bm{X}_{3,t} \zeta_{\bm{s}} +
-\omega_{\bm{s}} +
-\epsilon_{\bm{s},t},
-$$
-where
-
-* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
-* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
-* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
-* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
-* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
-* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
-* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
-
-The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a Matérn covariance function.
-Additional details are provided in the SI.
-
 # Application Examples

### 2021-11-18 doc/paper.Rmd
Develop paper

@@
 
-## Model selection
-
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
-Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
-Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).
-
-## Quantities of interest
-
-The results from a model fit with sdmTMB can be used to generate various quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
-
 ## Model evaluation
@@
 
+## Model selection
+
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
+Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
+Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).
+
+## Parameter priors/penalties
+
+sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
+These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
+Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.
+
+## Derived quantities of interest
+
+The results from a model fit with sdmTMB can be used to generate various quantities of interest.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
+Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
+
 <!-- Paragraph on advanced options here?
@@
 
-## Advanced options
+<!-- ## Advanced options -->
 
-While not of interest to all users, sdmTMB also implements several advanced options for model fitting. 
-First, models can include penalized likelihoods via assigning priors to model parameters.
-These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces.
-Following other recent SPDE approaches [@simpson2017; @osgood-zimmerman2021], 'penalized complexity' priors may be particularly useful in assigning priors to the spatial range and variance.
-Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to users of `mgcv`.
-Third, for users interested in Bayesian inference, we allow models to be passed to Stan (cite) via the package tmbstan [@monnahan2018].
+<!-- While not of interest to all users, sdmTMB also implements several advanced options for model fitting.  -->
+<!-- First, models can include penalized likelihoods via assigning priors to model parameters. -->
+<!-- These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces. -->
+<!-- Following other recent SPDE approaches [@simpson2017; @osgood-zimmerman2021], 'penalized complexity' priors may be particularly useful in assigning priors to the spatial range and variance. -->
+<!-- Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to users of `mgcv`.
+-->
+<!-- Third, for users interested in Bayesian inference, we allow models to be passed to Stan (cite) via the package tmbstan [@monnahan2018]. -->
 
@@
 It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
-Here, we illustrate a spatial model fit to Pacific Cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
+Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
@@
 We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
-The data frame `qcs_grid` is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
+The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
 The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
@@
 
-We could extend our model to a spatiotemporal model simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
 Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):

### 2021-11-18 doc/paper.Rmd
Develop paper

@@
 sdmTMB fits spatially explicit geostatistical GLMMs to point data.
-Point data simply means the collected data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
+Point data simply means the data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
 The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
@@
 
-Using the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time,
+Using the notation of a bold capital letter to represent a data vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
 the general form of a GLMM is
@@
 
-Models constructed with sdmTMB may be viewed as spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+sdmTMB models are spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
 In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
 The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
-The Matérn standard deviations are estimated separately for the various components and the range can be shared or estimated separately.
+The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) (REF).
+The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim$ 0.13 (REF)---can be shared or estimated separately.
 
@@
 
-By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
+By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] (Fig.\ \@ref(fig:diagram)).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.
 
 The sdmTMB model is fit by maximum marginal likelihood.
-Internally, a TMB model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the optimization routine `stats::nlminb()` in R (REF).
-Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation (REF).
+Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
+Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
 
@@
 
+## Model selection
+
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`aic()`, @akaike1974]. 
+Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
+The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
+
 ## Model evaluation
 
-Validation of mixed-effects models is non-trivial (that REF).
-
-We provide three tools to assist with model validation:
+Validation of mixed-effects models is non-trivial (that REF). We provide three approaches to assist with model validation:
 
@@
 These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects.
-These residuals have known statistical issues (REF).
-One-step-ahead residuals are also possible or residuals from a single MCMC draw, but both approaches are much slower.
+These residuals have known statistical issues (REF) but are quick to calculate.
+One-step-ahead residuals (REF) or residuals from a single MCMC draw (REF) are also possible, but both approaches are much slower.
 
@@
 
-3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation. 
+3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation (see `?extract_mcmc`). 
 We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).
-
-## Model selection
-
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
-Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
-Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).
 
@@
 sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
 
@@
 
-The results from a model fit with sdmTMB can be used to generate various quantities of interest.
+The results from an sdmTMB model can be used to generate various quantities of interest.
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
+Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
 
@@
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
 
@@
 ```{r, echo=TRUE, eval=FALSE}
+library(sdmTMB)
 head(pcod)
@@
 Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
+Internally, `s(depth)` creates random effect spline weights and so is not shown here.
 
 We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
-The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
+The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
 The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
@@
 VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
-The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
+The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
 Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists.
-For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1`).
-However, for users who require multivariate spatiotemporal models or the Poisson-link delta model (REF), there is no substitute for VAST.
+For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
+sdmTMB achieves 3 of the 4 design principles outlined in Thorson (REF) for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
+Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.
+
+<!--         -   "VAST represents four main design principles.
+Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
+<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
+<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
+<!--         -   Condition on missing covariates (check, random fields) -->
+<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
+<!-- However, VAST can fit multivariate spatiotemporal models (REF) and Poisson-link delta model (REF), whereas sdmTMB cannot. -->
 
 inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
-For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
+For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (TODO Supp fig?).
 One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
-We think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB, and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
-However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
+Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
+On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
 
-mgcv, but, yikes, do we have space:
+mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields (REF); however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
+First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions (REF) do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
+Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
+Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
+Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.
 
-- mgcv is capable of fitting with the SPDE (REF)
-- typical smoothers don't offer MVN interpretability
-- Tweedie is a big factor... must fix power 'p' to fit a GAMM.
-- Barrier approaches are different
-- doesn't have functionality built in for indexes/COG and uncertainty on such derived parameters
-- for this class of models, we think sdmTMB has advantages, but mgcv is an immensely powerful modelling package that extends well beyond spatial/spatiotemporal data.
-
-Final paragraph touching on other stuff possible in one paragraph: priors, Bayes, barrier meshes, randomized quantile residuals, simulation from joint precision, ...?
+Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management (REF).
+These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.
 
@@
 # Supporting Information
-
-# Application Examples
 
@@
 
+<!--
 # Tables
@@
 -   package comparison fitting --- appendix
+
+-->
 

### 2021-11-18 doc/paper.Rmd
Edits

@@
 Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [e.g., @cressie1993; @cressie2011].
+While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].
+This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [@shelton2014], and (3) can be of ecological interest itself [REFs, @barnett2021].
 
@@
 Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
 
@@
 Template systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
-For example, TMB requires the user to program in a C++ like template and it can be slow to experiment with multiple models when writing bespoke model code.
+For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
@@
 One common application is for species distribution models (SDMs), hence the package name.
-This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of additional functionality and related software.
+This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of related software.
 
@@
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
+Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
 
@@
 
+Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
 We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space):

### 2021-11-18 doc/paper.Rmd
Clean up refs
Start model description appendix

@@
 
-Take home, re-iterate paragraph?
+<!-- Take home, re-iterate paragraph? -->
 
@@
 
-# Supplemental Materials
-
-### Matern parameterization
-<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
-The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
-(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
-<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
-where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
-a scaling parameter that is estimated (REF).
-The parameter $\nu$ controls the smoothness of the covariance function.
-We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+<!--
 
@@
 Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?
+-->

### 2021-11-19 doc/paper.Rmd
Start adding table

@@
   \usepackage{bm}
+  \usepackage{amssymb}
+  \usepackage{tablefootnote}
   \linenumbers
@@
 
+```{r compare-table, results='asis'}
+library(kableExtra)
+d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE)
+d <- t(d)
+d <- cbind(row.names(d), d)
+d[d == "1"] <- "\\checkmark"
+d[d == "1*"] <- "\\checkmark\\footnote{a}"
+d[d == "0"] <- "--"
+d[d == "0*"] <- "--\\footnote{a}"
+d <- as.data.frame(d)
+names(d) <- d[1,]
+d <- d[-1,]
+head(d)
+names(d)[1] <- ""
+row.names(d) <- NULL
+d <- d[,c(1, 2, 3, 5, 6, 4)]
+d <- d[!grepl("SPDE", d[,1]), ]
+knitr::kable(d, format = "latex", caption = "Title of the table", 
+  booktabs = TRUE, linesep = "", escape = FALSE, row.names = FALSE) %>%
+  kableExtra::kable_styling(font_size = 8) %>% 
+  kableExtra::footnote(
+    number = c("Footnote 1; ", "Footnote 2; ")
+  )
+```
+
 # Discussion

### 2021-11-19 doc/paper.Rmd
Formatting fixes

--- 2021-11-18 doc/paper.Rmd
@@
 abstract: |
-    | \noindent Point 1: set the context for and purpose of the work; \smallskip
-    | Point 2: indicate the approach and methods; \smallskip
-    | Point 3: outline the main results; \smallskip
-    | Point 4: identify the conclusions and the wider implications. \smallskip
+    | \noindent Point 1: set the context for and purpose of the work;
+    | Point 2: indicate the approach and methods;
+    | Point 3: outline the main results;
+    | Point 4: identify the conclusions and the wider implications.
 ---
@@
 
+```{r compare-table}
+library(kableExtra)
+suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE)))
+d <- t(d)
+d <- cbind(row.names(d), d)
+d[d == "1"] <- "\\checkmark"
+d[d == "1*"] <- "\\checkmark\\footnote{a}"
+d[d == "0"] <- "--"
+d[d == "0*"] <- "--\\footnote{a}"
+d <- as.data.frame(d)
+names(d) <- d[1,]
+d <- d[-1,]
+names(d)[1] <- ""
+row.names(d) <- NULL
+d <- d[,c(1, 2, 3, 5, 6, 4)]
+d <- d[!grepl("SPDE", d[,1]), ]
+knitr::kable(d, format = "latex", caption = "Title of the table", 
+  booktabs = TRUE, linesep = "", escape = FALSE, row.names = FALSE) %>%
+  kableExtra::kable_styling(font_size = 8) %>% 
+  kableExtra::footnote(
+    number = c("Footnote 1; ", "Footnote 2; ")
+  )
+```
+
 # Discussion

### 2021-11-19 doc/paper.Rmd
Formatting
add white coastline to maps

--- 2021-11-18 doc/paper.Rmd
@@
 Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
+While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [e.g., @cressie1993; @cressie2011].
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [@shelton2014], and (3) can be of ecological interest itself [REFs, @barnett2021].
+This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].
 
@@
 Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
 
@@
 Template systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
-For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
+For example, TMB requires the user to program in a C++ like template and it can be slow to experiment with multiple models when writing bespoke model code.
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
@@
 One common application is for species distribution models (SDMs), hence the package name.
-This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of related software.
+This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of additional functionality and related software.
 
@@
 sdmTMB fits spatially explicit geostatistical GLMMs to point data.
-Point data simply means the data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
+Point data simply means the collected data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
 The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
@@
 
-Using the notation of a bold capital letter to represent a data vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
+Using the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time,
 the general form of a GLMM is
@@
 
-sdmTMB models are spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+Models constructed with sdmTMB may be viewed as spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
 In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
 The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) (REF).
-The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim$ 0.13 (REF)---can be shared or estimated separately.
+The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
+The Matérn standard deviations are estimated separately for the various components and the range can be shared or estimated separately.
 
@@
 
-By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
+By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] (Fig.\ \@ref(fig:diagram)).
 
 The sdmTMB model is fit by maximum marginal likelihood.
-Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
-Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
+Internally, a TMB model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the optimization routine `stats::nlminb()` in R (REF).
+Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation (REF).
 
@@
 
-## Model selection
-
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`aic()`, @akaike1974]. 
-Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
-The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
-
 ## Model evaluation
 
-Validation of mixed-effects models is non-trivial (that REF). We provide three approaches to assist with model validation:
+Validation of mixed-effects models is non-trivial (that REF).
+
+We provide three tools to assist with model validation:
 
@@
 These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects.
-These residuals have known statistical issues (REF) but are quick to calculate.
-One-step-ahead residuals (REF) or residuals from a single MCMC draw (REF) are also possible, but both approaches are much slower.
+These residuals have known statistical issues (REF).
+One-step-ahead residuals are also possible or residuals from a single MCMC draw, but both approaches are much slower.
 
@@
 
-3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation (see `?extract_mcmc`). 
+3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation. 
 We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).
+
+## Model selection
+
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
+Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
+Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).
 
@@
 sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
+These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
+Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.
 
@@
 
-The results from an sdmTMB model can be used to generate various quantities of interest.
+The results from a model fit with sdmTMB can be used to generate various quantities of interest.
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
+Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
 
@@
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
 
@@
 ```{r, echo=TRUE, eval=FALSE}
-library(sdmTMB)
 head(pcod)
@@
 
-Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
 We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space):
@@
 Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
-Internally, `s(depth)` creates random effect spline weights and so is not shown here.
 
 We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
-The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
+The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
 The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
@@
 p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
-```
-
-```{r}
-g_pred <- ggplot(p, aes(X, Y, fill = exp(est))) +
-  geom_raster() +
-  scale_fill_viridis_c(trans = "sqrt", option = "C") +
-  coord_fixed() +
-  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
 ```
@@
 
-```{r, warning=FALSE, message=FALSE, eval=FALSE}
+```{r, warning=FALSE, message=FALSE}
 library(sf)
+# devtools::install_github("ropensci/rnaturalearthhires") # install if needed
 map_data <- rnaturalearth::ne_countries(
@@
 
-```{r, eval=FALSE}
-ggplot(bc_coast_proj) + geom_sf() +
+```{r}
+g_pred <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
+                                          fill="white"
+                                          # fill="darkgray"
+                                          ) +
   geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
-  xlim(230957.7 + 100000, 1157991 - 500000) +
-  ylim(5366427 + 200000, 6353456 - 500000) +
-  scale_fill_viridis_c() +
-  theme_light() +
-  labs(fill = "Predicted\ndensity") +
-  labs(x = "Longitude", y = "Latitude")
+  coord_sf(datum = st_crs(utm_zone9)) +
+  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
+    labels = function(x) paste0(x/1000)
+  ) +
+  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
+    labels = function(x) paste0(x/1000)
+  ) +  scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
+  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
+  theme(
+    panel.background = element_rect(fill = "slategray1", colour= "white"), 
+    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
 ```
@@
 ```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(p, aes(X, Y, fill = omega_s)) +
-  geom_raster() +
-  # scale_fill_gradient2(high = scales::muted("red"), low = scales::muted("blue")) + coord_fixed() +
-  scale_fill_gradient2(high = reds[4], low = blues[4]) + coord_fixed() +
+g_omega <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
+                                           fill="white"
+                                           # fill="darkgray"
+                                           ) +
+  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
+  coord_sf(datum = st_crs(utm_zone9)) +
+  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
+    labels = function(x) paste0(x/1000)
+  ) +
+  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
+    labels = function(x) paste0(x/1000)) +
+  scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
+  # scale_fill_gradient2(high = reds[4], low = blues[4]) + 
   labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+  theme(
+    panel.background = element_rect(fill = "slategray1", colour= "white"),
+    legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
+
 cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
@@
 
-<!-- Take home, re-iterate paragraph? -->
+Take home, re-iterate paragraph?
 
@@
 VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
-The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
+The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
 Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists.
-For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
-sdmTMB achieves 3 of the 4 design principles outlined in Thorson (REF) for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
-Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.
-
-<!--         -   "VAST represents four main design principles.
-Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
-<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
-<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
-<!--         -   Condition on missing covariates (check, random fields) -->
-<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
-<!-- However, VAST can fit multivariate spatiotemporal models (REF) and Poisson-link delta model (REF), whereas sdmTMB cannot. -->
+For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1`).
+However, for users who require multivariate spatiotemporal models or the Poisson-link delta model (REF), there is no substitute for VAST.
 
 inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
-For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (TODO Supp fig?).
+For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
 One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
-Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
-On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
+We think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB, and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
+However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
 
-mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields (REF); however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
-First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions (REF) do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
-Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
-Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
-Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.
+mgcv, but, yikes, do we have space:
 
-Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management (REF).
-These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
-We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.
+- mgcv is capable of fitting with the SPDE (REF)
+- typical smoothers don't offer MVN interpretability
+- Tweedie is a big factor... must fix power 'p' to fit a GAMM.
+- Barrier approaches are different
+- doesn't have functionality built in for indexes/COG and uncertainty on such derived parameters
+- for this class of models, we think sdmTMB has advantages, but mgcv is an immensely powerful modelling package that extends well beyond spatial/spatiotemporal data.
+
+Final paragraph touching on other stuff possible in one paragraph: priors, Bayes, barrier meshes, randomized quantile residuals, simulation from joint precision, ...?
 
@@
 # Supporting Information
+
+# Application Examples
 
@@
 
-<!--
 # Tables
@@
 
--->
-
 \clearpage
@@
 
-<!--
+# Supplemental Materials
+
+### Matern parameterization
+<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
+The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
+(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
+<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
+where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
+a scaling parameter that is estimated (REF).
+The parameter $\nu$ controls the smoothness of the covariance function.
+We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
 
@@
 Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?
--->

### 2021-11-19 doc/paper.Rmd
version with original colours and darkgrey coast outlines

@@
 ```{r}
-g_pred <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
-                                          fill="white"
-                                          # fill="darkgray"
+g_pred <- ggplot(bc_coast_proj) + geom_sf(
+                                           # colour=NA, fill="white"
+                                           fill=NA, colour="gray30"
                                           ) +
@@
     labels = function(x) paste0(x/1000)
-  ) +  scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
+  ) +  
+  scale_fill_viridis_c(trans = "sqrt", option = "C") +
+  # scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
   labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
   theme(
-    panel.background = element_rect(fill = "slategray1", colour= "white"), 
+    # panel.background = element_rect(fill = "slategray1", colour= "white"), 
     legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
@@
 ```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
-                                           fill="white"
-                                           # fill="darkgray"
+g_omega <- ggplot(bc_coast_proj) + geom_sf(
+                                           # colour=NA, fill="white"
+                                           fill=NA, colour="gray30"
                                            ) +
@@
     labels = function(x) paste0(x/1000)) +
-  scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
-  # scale_fill_gradient2(high = reds[4], low = blues[4]) + 
+  # scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
+  scale_fill_gradient2(high = reds[4], low = blues[4]) +
   labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
   theme(
-    panel.background = element_rect(fill = "slategray1", colour= "white"),
+    # panel.background = element_rect(fill = "slategray1", colour= "white"),
     legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))

### 2021-11-19 doc/paper.Rmd
add unused option for solid gray land

--- 2021-11-19 doc/paper.Rmd
@@
 ```{r}
-g_pred <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
-                                          fill="white"
-                                          # fill="darkgray"
+g_pred <- ggplot(bc_coast_proj) + geom_sf(
+                                           # colour=NA, fill="white"
+                                           fill=NA, colour="gray30"
+                                           # fill="gray70", colour=NA
                                           ) +
@@
 ```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
-                                           fill="white"
-                                           # fill="darkgray"
+g_omega <- ggplot(bc_coast_proj) + geom_sf(
+                                           # colour=NA, fill="white"
+                                           fill=NA, colour="gray30"
+                                           # fill="gray70", colour=NA
                                            ) +

### 2021-11-19 doc/appendix-model.Rmd

--- 2021-11-19 doc/paper.Rmd
@@
 ---
-title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
-author: "|
-  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
-  $^2$Other addresses\n"
+title: Detailed model description
 output:
@@
 bibliography: refs.bib
-link-citations: yes
-header-includes:
-  \usepackage{setspace}\onehalfspacing
-  \usepackage[left]{lineno}
-  \usepackage{bm}
-  \linenumbers
-  \newcommand{\beginsupplement}{
-  \setcounter{equation}{0}
-  \renewcommand{\theequation}{S.\arabic{equation}}
-  \setcounter{table}{0}
-  \renewcommand{\thetable}{S\arabic{table}}
-  \setcounter{figure}{0}
-  \renewcommand{\thefigure}{S\arabic{figure}}}
-editor_options:
-  markdown:
-    wrap: sentence
-abstract: |
-    | \noindent Point 1: set the context for and purpose of the work; \smallskip
-    | Point 2: indicate the approach and methods; \smallskip
-    | Point 3: outline the main results; \smallskip
-    | Point 4: identify the conclusions and the wider implications. \smallskip
 ---
 
-```{r, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
-library(here)
-library(ggplot2)
-theme_set(theme_minimal())
-# theme_set(ggsidekick::theme_sleek())
-library(dplyr)
-library(sdmTMB)
-```
+### Matern parameterization
 
-```{r, include=FALSE}
-options(tinytex.verbose = TRUE)
-knitr::opts_chunk$set(
-  echo = FALSE,
-  message = FALSE,
-  warning = FALSE,
-  collapse = TRUE,
-  comment = "#>"
-)
-```
-
-# Introduction
-
-Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [e.g., @cressie1993; @cressie2011].
-While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [e.g., @legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @barnett2021].
-
-Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
-Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
-
-While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
-Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
-These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
-This can make an enormous difference to computational efficiency [@heaton2017].
-The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA [e.g., @osgood-zimmerman2021].
-Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].
-
-Template systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
-For example, TMB requires the user to program in a C++ like template and it can be slow to experiment with multiple models when writing bespoke model code.
-Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to spatial models that use the SPDE, but either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some of the flexibility for the particular class of models discussed here.
-We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
-<!-- User interfaces through software such as VAST or inlabru can greatly reduce these challenges.
-VAST is built on TMB, and is extremely powerful (e.g., fitting many multivariate models not described here), but does not provide a simple, familiar, modular interface for users with basic spatial or spatiotemporal data.
-inlabru provides a somewhat more familiar interface to a large class of models including GRMF models, but is strictly Bayesian, lacks some flexibility for the particular class of models discussed here.
-It also relies on INLA, and at least for some problems, TMB can be much faster [@osgood-zimmerman2021].
-We provide a more detailed comparison of related software packages in Table 1 and the discussion.
--->
-
-<!-- User interfaces to classes of models fit by these template systems can greatly  -->
-
-<!-- reduce these challenges for the vast majority of models. -->
-
-<!-- keep the above even briefer? -->
-
-<!-- inlabru is "very good" but not entirely familiar to those used to `glm()`, lme4, glmmTMB, or mgcv; -->
-
-<!-- *TODO: the differences are many and more than can be listed here. Keep brief add Table or discussion later about differences and when to pick one over the other?* -->
-
-Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
-Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
-One common application is for species distribution models (SDMs), hence the package name.
-This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of additional functionality and related software.
-
-<!-- -   [@anderson2019] -->
-<!-- -   [@maureaud2021] -->
-<!-- -   [@anderson2020] -->
-<!-- -   [@osgood-zimmerman2021] -->
-<!-- -   [@breivik2021] -->
-<!-- -   [@monnahan2021] -->
-<!-- -   [@shelton2014] -->
-<!-- -   [@lindgren2011] -->
-<!-- -   [@latimer2009] -->
-
-<!-- Packages: -->
-
-<!-- spaMM: @rousset2014 -->
-<!-- spNNGP: @finley2021 -->
-
-<!-- millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies." -->
-
-<!-- legendre1989a: "Spatial pattern and ecological analysis": "We can expect the spatial approach to ecological problems to bring about a quantum jump for ecologists and population geneticists who had learned a type of statistics where one had to hide space or time structures.
-It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables." -->
-
-<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
-
-(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
-An sdmTMB model can be built from any combination of these components.
-The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
-Notation:
-We refer to design matrices as $\bm{X}$.
-The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and 'group', respectively.
-The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
-All other symbols refer to the  described model components.
-Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
-
-```{r diagram, fig.cap="(ref:diagram-fig)"}
-if (file.exists(here::here("figs/diagram.pdf"))) {
-  knitr::include_graphics(here::here("figs/diagram.pdf"))
-} else {
-  plot(1, type = "n", main = "figs/diagram.pdf missing; will be synced when complete")
-}
-```
-
-# Model description
-
-sdmTMB fits spatially explicit geostatistical GLMMs to point data.
-Point data simply means the collected data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
-The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
-\@ref(fig:diagram)).
-This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
-
-Using the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time,
-the general form of a GLMM is
-$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
-with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
-<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
-For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
-Although in practice it is rare to include all of these components, the full sdmTMB model is
-$$
-g(\mu_{\bm{s},t}) =
-\bm{X}_{1,\bm{s},t} \bm{\beta} +
-\bm{\alpha}_g +
-\bm{X}_{2,\bm{s},t} \gamma_t +
-\bm{X}_{3,t} \zeta_{\bm{s}} +
-\omega_{\bm{s}} +
-\epsilon_{\bm{s},t},
-$$
-where
-
-* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
-* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
-* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
-* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
-* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
-* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
-* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
-
-Models constructed with sdmTMB may be viewed as spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
-The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
-The Matérn standard deviations are estimated separately for the various components and the range can be shared or estimated separately.
-
-<!-- - high-level paragraph or two about what the model includes, referencing Fig 1 -->
-<!-- - high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?) -->
-<!-- - high-level paragraph about simulation capabilities and + testing performed (reference supplement) -->
-<!-- - model selection/cross validation paragraph -->
-<!-- - derived quantities paragraph (or just put this in the examples?) -->
-
-<!-- - example 1 + figs -->
-<!-- - example 2 + figs -->
-
-<!--     - perhaps a paragraph on how it achieves what Jim outlined @thorson2019 as necessary: -->
-<!--         -   "VAST represents four main design principles.
-Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
-<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
-<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
-<!--         -   Condition on missing covariates (check, random fields) -->
-<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
-
-<!-- sdmTMB in its current form achieves 3 of the 4 design principles outlined by @thorson2019 for index standardization:  -->
-
-<!-- (e.g., 'catchability' and 'habitat' covariates in VAST merely depend on the values supplied to `predict()` in sdmTMB). -->
-
-<!-- # Model -->
-
-<!-- Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs. -->
-<!-- GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.
--->
-
-By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
-Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] (Fig.\ \@ref(fig:diagram)).
-
-The sdmTMB model is fit by maximum marginal likelihood.
-Internally, a TMB model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the optimization routine `stats::nlminb()` in R (REF).
-Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation (REF).
-
-<!-- [paragraph about spde and gaussian process / Matern covariance here] -->
-
-<!-- Paragraph on model selection here-->
-
-## Model evaluation
-
-Validation of mixed-effects models is non-trivial (that REF).
-
-We provide three tools to assist with model validation:
-
-1. The `residuals()` method returns randomized quantile residuals (e.g., as in DHARMa REF).
-These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects.
-These residuals have known statistical issues (REF).
-One-step-ahead residuals are also possible or residuals from a single MCMC draw, but both approaches are much slower.
-
-2.
-The `sdmTMB_simulate()` function can simulate new data, or simulate from a fitted model, to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
-
-3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation. 
-We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).
-
-## Model selection
-
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
-Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
-Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).
-
-## Parameter priors/penalties
-
-sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.
-
-## Derived quantities of interest
-
-The results from a model fit with sdmTMB can be used to generate various quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
-
-<!-- Paragraph on advanced options here?
-Specifically, this could address
-- PC-priors -- and other penalized priors
-- Non-linear stuff / etc
-- Linkages to Rstan
-- Simulation
--->
-
-<!-- ## Advanced options -->
-
-<!-- While not of interest to all users, sdmTMB also implements several advanced options for model fitting.  -->
-<!-- First, models can include penalized likelihoods via assigning priors to model parameters. -->
-<!-- These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces. -->
-<!-- Following other recent SPDE approaches [@simpson2017; @osgood-zimmerman2021], 'penalized complexity' priors may be particularly useful in assigning priors to the spatial range and variance. -->
-<!-- Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to users of `mgcv`.
--->
-<!-- Third, for users interested in Bayesian inference, we allow models to be passed to Stan (cite) via the package tmbstan [@monnahan2018]. -->
-
-# An introductory example
-
-An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
-Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
-Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
-
-```{r, include=FALSE}
-# simplify for paper
-pcod <- dplyr::select(pcod, year, density, depth, X, Y)
-```
-
-```{r, echo=TRUE, eval=FALSE}
-head(pcod)
-```
-
-```{}
-#>    year density depth     X     Y
-#>   <int>   <dbl> <dbl> <dbl> <dbl>
-#> 1  2003   113.    201  446. 5793.
-#> 2  2003    41.7   212  446. 5800.
-#> 3  2003     0     220  449. 5802.
-```
-
-We start by creating a mesh object that contains matrices to apply the SPDE approach:
-
-```{r, echo=TRUE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
-```
-
-Here, `cutoff` defines the minimum allowed distance between points in the units of `X` and `Y` (km).
-Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
-We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)a).
-
-We can then fit a model with spatial random fields via the function `sdmTMB()`. 
-We use a penalized smoother for depth as a main effect via `s()` from the mgcv package.
-We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
-
-```{r pcod-eg1-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE}
-fit <- sdmTMB(
-  density ~ s(depth),
-  data = pcod,
-  family = tweedie(link = "log"),
-  mesh = mesh,
-  spatial = "on"
-)
-```
-
-We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space):
-
-```{r,eval=FALSE, echo=TRUE}
-print(fit)
-```
-
-<!-- tidy(fit, conf.int = TRUE) -->
-<!-- tidy(fit, effects = 'ran_pars', conf.int = TRUE) -->
-
-```{}
-#> Spatiotemporal model fit by ML ['sdmTMB']
-#> Formula: density ~ s(depth)
-#> Mesh: mesh
-#> Data: pcod
-#> Family: tweedie(link = 'log')
-#>             coef.est coef.se
-#> (Intercept)     2.37    0.21
-#>
-#> Dispersion parameter: 12.69
-#> Tweedie p: 1.58
-#> Matern range: 16.39
-#> Spatial SD: 1.86
-#> ML criterion at convergence: 6402.136
-```
-
-The output indicates our model was fit by maximum (marginal) likelihood (`'ML'`).
-We also see the formula, mesh, fitted data, and family.
-Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
-
-We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
-The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
-The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
-
-```{r, echo=TRUE, eval=FALSE}
-p <- predict(fit, newdata = qcs_grid)
-```
-
-```{r, echo=FALSE, eval=TRUE}
-p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
-```
-
-```{r, eval=FALSE}
-p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
-se <- apply(p3, 1, sd)
-p$se <- se
-g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
-  geom_raster() +
-  scale_fill_viridis_c(trans = "log10", option = "C") +
-  coord_fixed() +
-  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
-```
-
-```{r, warning=FALSE, message=FALSE}
-library(sf)
-# devtools::install_github("ropensci/rnaturalearthhires") # install if needed
-map_data <- rnaturalearth::ne_countries(
-  scale = "large",
-  returnclass = "sf", country = "canada")
-# Crop the polygon for plotting and efficiency:
-# st_bbox(map_data) # find the rough coordinates
-bc_coast <- suppressWarnings(suppressMessages(
-  st_crop(map_data,
-    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
-utm_zone9 <- 3156
-bc_coast_proj <- sf::st_transform(bc_coast, crs = utm_zone9)
-# ggplot(bc_coast_proj) + geom_sf()
-```
-
-```{r}
-g_pred <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
-                                          fill="white"
-                                          # fill="darkgray"
-                                          ) +
-  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
-  coord_sf(datum = st_crs(utm_zone9)) +
-  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
-    labels = function(x) paste0(x/1000)
-  ) +
-  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
-    labels = function(x) paste0(x/1000)
-  ) +  scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
-  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(
-    panel.background = element_rect(fill = "slategray1", colour= "white"), 
-    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
-```
-
-```{r, echo=FALSE}
-blues <- RColorBrewer::brewer.pal(5, "Blues")
-reds <- RColorBrewer::brewer.pal(5, "Reds")
-g_mesh <- ggplot(pcod, aes(X, Y)) +
-  inlabru::gg(mesh$mesh, edge.color = "grey60") +
-  geom_point(alpha = 0.6, mapping = aes(size = density), colour = blues[4]) +
-  theme_minimal() + coord_fixed() +
-  labs(x = "UTMs E (km)", y = "UTMs N (km)") +
-  scale_size_area(max_size = 10) +
-  theme(legend.position = "none")
-```
-
-```{r}
-nd <- data.frame(
-  depth =
-    seq(min(pcod$depth), max(pcod$depth), length.out = 300)
-)
-p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
-g_smooth <- ggplot(
-  p_smooth,
-  aes(depth, exp(est),
-    ymin = exp(est - 1.96 * est_se),
-    ymax = exp(est + 1.96 * est_se)
-  )
-) +
-  geom_ribbon(fill = blues[2]) +
-  geom_line(colour = blues[5], lwd = 1) +
-  labs(x = "Depth (m)", y = "Density (units)") +
-  xlim(min(pcod$depth), 300) +
-  coord_fixed(expand = FALSE, ratio = 1.28)
-```
-
-```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
-fit2 <- sdmTMB(
-  density ~ 0 + as.factor(year),
-  family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
-  time = "year", spatial = "on", spatiotemporal = "iid", silent = TRUE
-)
-p2 <- predict(fit2, newdata = qcs_grid, sims = 500L)
-ind <- get_index_sims(p2)
-```
-
-```{r plot-ind, eval=FALSE}
-g_ind <- ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) + 
-  geom_ribbon(fill = blues[2]) +
-  geom_line(colour = blues[5], lwd = 1) +
-  ylim(0, max(ind$upr)) +
-  coord_fixed(expand = FALSE, ratio = 0.3) +
-  labs(x = "Year", y = "Biomass (units)")
-```
-
-(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
-(a) SPDE mesh (lines) combined with the trawl survey observations (points).
-Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
-The circle area corresponds to density of Pacific Cod for individual trawls.
-(b) Conditional effect of depth modelled as a penalized smoother.
-The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
-These predictions are made omitting the spatial random field.
-(c) Spatial random field:
-these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
-(d) Overall prediction: these estimates represent the combination of all fixed and random effects.
-
-```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(bc_coast_proj) + geom_sf(colour=NA, 
-                                           fill="white"
-                                           # fill="darkgray"
-                                           ) +
-  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
-  coord_sf(datum = st_crs(utm_zone9)) +
-  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
-    labels = function(x) paste0(x/1000)
-  ) +
-  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
-    labels = function(x) paste0(x/1000)) +
-  scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
-  # scale_fill_gradient2(high = reds[4], low = blues[4]) + 
-  labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(
-    panel.background = element_rect(fill = "slategray1", colour= "white"),
-    legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
-
-cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
-```
-
-We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
-Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
-
-```{r, eval=FALSE, echo=TRUE}
-fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh
-  time = "year", spatial = "off", spatiotemporal = "ar1"
-)
-```
-
-<!--
-Highlights?
-
-- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances)
-- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
-- Spatiotemporal models (RW/AR v IID)
-- Time varying parameters
-- Non-linear effects (smooth splines, threshold/breakpoint functions)
-- Variance estimation and derived quantities (spatial range, area occupied etc)
--->
-
-# Discussion
-
-Take home, re-iterate paragraph?
-
-There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru.
-VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
-The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
-Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists.
-For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1`).
-However, for users who require multivariate spatiotemporal models or the Poisson-link delta model (REF), there is no substitute for VAST.
-
-inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
-For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
-One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
-We think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB, and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
-However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
-
-mgcv, but, yikes, do we have space:
-
-- mgcv is capable of fitting with the SPDE (REF)
-- typical smoothers don't offer MVN interpretability
-- Tweedie is a big factor... must fix power 'p' to fit a GAMM.
-- Barrier approaches are different
-- doesn't have functionality built in for indexes/COG and uncertainty on such derived parameters
-- for this class of models, we think sdmTMB has advantages, but mgcv is an immensely powerful modelling package that extends well beyond spatial/spatiotemporal data.
-
-Final paragraph touching on other stuff possible in one paragraph: priors, Bayes, barrier meshes, randomized quantile residuals, simulation from joint precision, ...?
-
-# Acknowledgements
-
-# Conflict of Interest statement
-
-# Author Contributions
-
-# Supporting Information
-
-# Application Examples
-
-<!--
-All submissions with more than one author must include an Authors’ contributions statement. All persons listed as authors on a paper are expected to meet ALL of the following criteria for authorship:
-
-substantial contributions to conception and design, or acquisition of data, or analysis and interpretation of data, or drafting the article or revising it critically for important intellectual content;
-final approval of the version to be published;
-agreement to be accountable for the aspects of the work that they conducted and ensuring that questions related to the accuracy or integrity of any part of their work are appropriately investigated and resolved.
-Acquisition of funding, provision of facilities, or supervising the research group of authors without additional contribution are not usually sufficient justifications for authorship.
-The statement should include an explanation of the contribution of each author.
-We suggest the following format for the Authors’ contributions statement:
-
-AB and CD conceived the ideas and designed methodology; CD and EF collected the data; EF and GH analysed the data; AB and CD led the writing of the manuscript.
-All authors contributed critically to the drafts and gave final approval for publication.
--->
-
-# Data Availability
-
-<!--
-To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
-For example, “Data available from the Dryad Digital Repository http://dx.doi.org/10.5061/dryad.41qh7 (Kiere & Drummond 2016).” When a DOI is available for the data, the full data citation should also be given in the reference list (see below).
-For further information see our data archiving policy.
--->
-
-# Tables
-
-1. Comparison to related software?
-
-- VAST
-- INLA
-- mgcv
-- glmmTMB
-- glmmfields?
-- other stuff on CRAN, spBayes etc.
-
-# Figures
-
-1. Flow diagram of functions? Not needed?
-1. Speed with various knots and/or spatial complexity in comparison to INLA + VAST + ...?
-1. Example application 1 - SDM, show mesh(?) and/or raw data, show predictions and SD in space,
-1. Example application 2 - population index with stitched survey?
-
-Content:
-
--   outline the general need for this type of modelling
-
--   outline the need for this type of package to make the models accessible and rapid to iterate on
-
--   emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference
-
--   outline the philosophy and how the functions work together
-
-    -   modular
-    -   familiar (similar to glm, lme4, glmmTMB, mgcv, etc.)
-    -   separate fitting and predicting
-    -   avoid expensive calculations unless needed
-
--   table comparison to related packages
-
--   case studies
-
-    -   an SDM with time-varying depth?
-    -   an index calculation?
-    -   stitching surveys?
-
--   how it achieves the steps Jim described [@thorson2019] in what a VAST-related package has to be able to do (and how VAST does it vs. how sdmTMB does it)
-    -   "VAST represents four main design principles.
-Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include."
-    -   Area weighting (check, helper functions to sum up weighted grid cells)
-    -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`)
-    -   Condition on missing covariates (check, random fields)
-    -   Bridge between univariate and multivariate applications (no! should we?)
-
--   speed comparisons (INLA [inlabru?], VAST, mgcv, glmmTMB without predictive process?)
-
--   intended audience: ecology, but much broader too
-
--   self simulation testing --- appendix
-
--   package comparison fitting --- appendix
-
-\clearpage
-
-# References
-
-::: {#refs}
-:::
-
-```{=tex}
-\clearpage
-\beginsupplement
-```
-
-# Supplemental Materials
-
-### Matern parameterization
 <!-- EW: Because of space limits, I might suggest moving all this to the SI -->
@@
   We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
-
-### Junk to delete
-*EW: These are just placeholders -- not sure what order is best or what to highlight in very limited space*
-
-As a first example, we use a spatial dataset (`parana`) from the `geoR` package (Ribeiro Jr and Diggle 2018) that describes average May-June rainfall recorded at 143 stations in the state of Paraná, Brazil.
-This dataset is also well-known because it is used as an example of spatiotemporal modeling with INLA (cite INLA) book.
-Because the dataset doesn't include covariates and only includes a single year, we can model rainfall using an intercept and spatial component, $g(\mu_{s,t}) = b_{0} + \omega_{s}$.
-We assume the observation model for the rainfall data is normally distributed.
-The first step in fitting a model with `sdmTMB` is constructing the mesh to approximate the spatial process.
-This step can be done with the `make_mesh` function, which accepts the data (coordinates) and at minimum, the minimum allowed distance between points (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with INLA can also be passed in (additional examples of this are included in the SI).
-For example, an initial mesh with a cutoff distance of 0.1km could be made with
-
-```{r eval=FALSE, echo=TRUE}
-rain_spde = make_mesh(data = parana, c("east","north"), cutoff=0.1)
-```
-
-Second, we can fit the model using TMB with the `sdmTMB` function, whose formula syntax should be familiar to most R users,
-
-```{r eval = FALSE, echo=TRUE}
-fit = sdmTMB(data ~ 1, data=parana, spde = rain_spde)
-```
-
-As a second example, we use `sdmTMB` to analyze a 2-year dataset of Eurasian cranes (*Grus grus*) from England, UK (Soriano-Redondo et al.
-2019).
-In the original analysis, Soriano-Redondo et al.
-2019 used log Gaussian Cox process models to jointly model crane presence--absence and the spatiotemporal distribution of wetlands.
-We use this dataset as an example of modeling presence--absence data with sdmTMB, and only model the occurence of cranes given covariates.
-With any multi-year dataset, a common question of interest is often whether observations can be best described by spatial variation, or spatial and spatiotemporal variation.
-With two years of data, we fit two models with `sdmTMB`: a model with a spatial component but no spatiotemporal variation, and a model with only spatiotemporal variation (allowing each year to have a different spatial field).
-We included the same fixed effect covariates in both models: different intercepts by year, and two covariates found to be important by Soriano-Redondo et al.
-2019 (wetland area, perimeter to area ratio).
-As a coarse comparison of data support, we compare the Akaike's Information Criterion (Akaike 1973) from each model.
-The model with a shared spatial field appeared to be more supported ($\Delta AIC$=25.23).
-The sdmTMB package allows for additional model comparisons, including k-fold cross validation with the `sdmTMB_cv` function.
-
-Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?

### 2021-11-19 doc/paper.Rmd

@@
 sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
 
@@
 
+\clearpage
+
+```{r compare-table}
+library(kableExtra)
+suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE)))
+d <- t(d)
+d <- cbind(row.names(d), d)
+d[d == "1"] <- "\\checkmark"
+d[d == "1*"] <- "\\checkmark\\footnote{a}"
+d[d == "0"] <- "--"
+d[d == "0*"] <- "--\\footnote{a}"
+d <- as.data.frame(d)
+names(d) <- d[1,]
+d <- d[-1,]
+names(d)[1] <- ""
+row.names(d) <- NULL
+d <- d[,c(1, 2, 3, 5, 6, 4)]
+d <- d[!grepl("SPDE", d[,1]), ]
+knitr::kable(d, format = "latex", caption = "Title of the table", 
+  booktabs = TRUE, linesep = "", escape = FALSE, row.names = FALSE) %>%
+  kableExtra::kable_styling(font_size = 8) %>% 
+  kableExtra::footnote(
+    number = c("Footnote 1; ", "Footnote 2; ")
+  )
+```
+
 # Discussion
 
-Take home, re-iterate paragraph?
+<!-- Take home, re-iterate paragraph? -->
 
@@
 
+-->
+
 \clearpage

### 2021-11-19 doc/paper.Rmd
finer coastlines

@@
                                            # colour=NA, fill="white"
-                                           fill=NA, colour="gray30"
+                                           fill=NA, colour="gray30", lwd = 0.3
                                            # fill="gray70", colour=NA

### 2021-11-23 doc/paper.Rmd
Add timing fig

--- 2021-11-19 doc/paper.Rmd
@@
 ```{r diagram, fig.cap="(ref:diagram-fig)"}
-if (file.exists(here::here("figs/diagram.pdf"))) {
-  knitr::include_graphics(here::here("figs/diagram.pdf"))
-} else {
-  plot(1, type = "n", main = "figs/diagram.pdf missing; will be synced when complete")
+f <- here::here("figs", "diagram.pdf")
+if (!file.exists(f)) {
+  download.file("https://www.dropbox.com/s/j8km3hv8xz9b63k/diagram.pdf?dl=1", 
+    destfile = f)
 }
+knitr::include_graphics(f)
 ```
@@
 ```{r}
-g_pred <- ggplot(bc_coast_proj) + geom_sf(
-                                           # colour=NA, fill="white"
-                                           fill=NA, colour="gray30"
-                                           # fill="gray70", colour=NA
-                                          ) +
+g_pred <- ggplot(bc_coast_proj) +
+  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
   geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
   coord_sf(datum = st_crs(utm_zone9)) +
-  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
-    labels = function(x) paste0(x/1000)
+  scale_x_continuous(
+    limits = c(230957.7 + 105000, 1157991 - 570000),
+    labels = function(x) paste0(x / 1000)
   ) +
-  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
-    labels = function(x) paste0(x/1000)
+  scale_y_continuous(
+    limits = c(5366427 + 270000, 6353456 - 513000),
+    labels = function(x) paste0(x / 1000)
   ) +
@@
     # panel.background = element_rect(fill = "slategray1", colour= "white"),
-    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
+    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
+  )
 ```
@@
 ```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(bc_coast_proj) + geom_sf(
-                                           # colour=NA, fill="white"
-                                           fill=NA, colour="gray30"
-                                           # fill="gray70", colour=NA
-                                           ) +
+g_omega <- ggplot(bc_coast_proj) +
+  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
   geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
   coord_sf(datum = st_crs(utm_zone9)) +
-  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
-    labels = function(x) paste0(x/1000)
+  scale_x_continuous(
+    limits = c(230957.7 + 105000, 1157991 - 570000),
+    labels = function(x) paste0(x / 1000)
   ) +
-  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
-    labels = function(x) paste0(x/1000)) +
+  scale_y_continuous(
+    limits = c(5366427 + 270000, 6353456 - 513000),
+    labels = function(x) paste0(x / 1000)
+  ) +
   # scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
@@
     # panel.background = element_rect(fill = "slategray1", colour= "white"),
-    legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
-
+    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
+  )
 cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
@@
 
+## Speed comparison
+
+We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+\@ref(fig:timing)).
+In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
+inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
+mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
+With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (note that this option may not be possible or as fast for all models).
+Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern (not shown).
+
+(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
+Lines represent means and ribbons ranges across 20 random iterations.
+Note the log x and y axes.
+sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
+inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
+mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
+inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library.
+All platforms were restricted to one core and could be faster with parallel computation.
+<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
+
+```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
+f <- here::here("figs", "timing2-logx.pdf")
+if (!file.exists(f)) {
+  download.file("https://www.dropbox.com/s/3aco85mizuplz7z/timing2-logx.pdf?dl=1", 
+    destfile = f)
+}
+knitr::include_graphics(f)
+```
+
 # Discussion
@@
 inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
-For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (TODO Supp fig?).
+For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig.
+\@ref(fig:timing), @osgood-zimmerman2021].
 One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
@@
 
-mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields (REF); however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
-First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions (REF) do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
+mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields [@miller2019]; however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
+First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions [@miller2019; @pedersen2019] do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
 Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.

### 2021-11-23 doc/paper.Rmd
Add examples to paper

@@
 fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh
+  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
   time = "year", spatial = "off", spatiotemporal = "ar1"
+)
+```
+
+We could generate a standardized index from that fit with:
+
+```{r, eval=FALSE, echo=TRUE}
+p <- predict(fit_spatiotemporal, newdata = qcs_grid, return_tmb_object = TRUE)
+ind <- get_index(p)
+```
+
+As a third example, we could model presence/absence of Pacific cod (`present`) with a time-varying quadratic effect of depth, an independent mean each year, a spatial random field, and IID spatiotemporal fields:
+
+```{r, eval=TRUE, echo=FALSE}
+# restore 'present' column
+pcod <- sdmTMB::pcod
+```
+
+```{r, eval=FALSE, echo=TRUE}
+fit_spatiotemporal_tv <- sdmTMB(
+  present ~ 0 + as.factor(year), 
+  time_varying = ~ 0 + poly(depth, 2), 
+  family = binomial(), data = pcod, mesh = mesh,
+  time = "year", spatial = "on", spatiotemporal = "iid"
 )
@@
 
-## Speed comparison
+# Speed comparison
 

### 2021-11-23 doc/paper.Rmd
Adjust table

@@
 suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE)))
-d <- t(d)
-d <- cbind(row.names(d), d)
+d <- as.matrix(d)
 d[d == "1"] <- "\\checkmark"
@@
 d <- as.data.frame(d)
-names(d) <- d[1,]
-d <- d[-1,]
 names(d)[1] <- ""

### 2021-11-23 doc/paper.Rmd
Develop table

@@
 library(kableExtra)
-suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE)))
+suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE, comment = "#")))
 d <- as.matrix(d)
@@
 d <- d[!grepl("SPDE", d[,1]), ]
-knitr::kable(d, format = "latex", caption = "Title of the table", 
-  booktabs = TRUE, linesep = "", escape = FALSE, row.names = FALSE) %>%
+names(d)[names(d) == "mgcv"] <- "mgcv\\footnote{a}"
+
+knitr::kable(d, format = "latex", caption = "Comparison of functionality between five R packages that can fit SPDE-based geostatistical GLMMs.", 
+  booktabs = TRUE, 
+  linesep = c(
+    rep('', 4), # coefs
+    '\\addlinespace',
+    rep('', 3), # correlation stuff
+    '\\addlinespace',
+    rep('', 10), # likelihood
+    '\\addlinespace',
+    rep('', 2), # bayes
+    '\\addlinespace'
+  ),
+  escape = FALSE, row.names = FALSE) %>%
   kableExtra::kable_styling(font_size = 8) %>% 
   kableExtra::footnote(
-    number = c("Footnote 1; ", "Footnote 2; ")
+    number = c(
+      "SPDE approach as in Miller et al. (2019).",
+      "Technically possible but non-trivial.",
+      "Zero-inflated NB2 only.",
+      "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
+      "Technically possible but non-trivial.",
+      "Technically possible but non-trivial."
+    ),
+    escape = FALSE
   )

### 2021-11-23 doc/paper.Rmd
Add to paper

@@
 
+<!-- TODO: cite:  -->
+<!-- @article{gelfand2017, -->
+<!--   title = {Bayesian Modeling and Analysis of Geostatistical Data}, -->
+<!--   author = {Gelfand, Alan E. and Banerjee, Sudipto}, -->
+<!--   year = {2017}, -->
+<!--   journal = {Annual Review of Statistics and Its Application}, -->
+<!--   volume = {4}, -->
+<!--   number = {1}, -->
+<!--   pages = {245--266}, -->
+<!--   doi = {10.1146/annurev-statistics-060116-054155}, -->
+<!--   annotation = {\_eprint: https://doi.org/10.1146/annurev-statistics-060116-054155} -->
+<!-- } -->
+
 Ecological data are often collected in space or in space repeatedly over time.
@@
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [@shelton2014], and (3) can be of ecological interest itself [REFs, @barnett2021].
+This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].
 
@@
 This can make an enormous difference to computational efficiency [@heaton2017].
-The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA [e.g., @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
 Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].
 
-Template systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
+Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
 For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
@@
 
-Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
+Here, we introduce our R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
 Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
@@
 (ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
-An sdmTMB model can be built from any combination of these components.
+An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
 The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
@@
 
-sdmTMB fits spatially explicit geostatistical GLMMs to point data.
-Point data simply means the data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
+sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
+Geostatistical data simply refers to data observed at specific spatial coordinates (vs. areal data that might represent polygons or grids).
 The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
@@
 
-Using the notation of a bold capital letter to represent a data vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
+Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
 the general form of a GLMM is
@@
 
-There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru.
+There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
 VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
 The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
-Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists.
+Interface-wise, VAST is a less modular interface (e.g., by default automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
 For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
-sdmTMB achieves 3 of the 4 design principles outlined in Thorson (REF) for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
+sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
 Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.
@@
 mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields [@miller2019]; however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
-First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions [@miller2019; @pedersen2019] do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
+First, the smoothers typically used with mgcv (e.g., thin plate splines), while capable of making equivalent predictions [@miller2019; @pedersen2019], do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance estimates).
 Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.

### 2021-11-23 doc/paper.Rmd
Try adding spBayes

@@
 <!--   annotation = {\_eprint: https://doi.org/10.1146/annurev-statistics-060116-054155} -->
+
+<!-- @book{diggle2007, -->
+<!--   title = {Model-Based {{Geostatistics}}}, -->
+<!--   author = {Diggle, Peter and Ribeiro, Paulo Justiniano}, -->
+<!--   year = {2007}, -->
+<!--   publisher = {{Springer}}, -->
+<!--   googlebooks = {qCqOm39OuFUC}, -->
+<!--   isbn = {978-0-387-48536-2}, -->
+<!--   langid = {english} -->
+<!-- } -->
+
 <!-- } -->
@@
 Geostatistical data simply refers to data observed at specific spatial coordinates (vs. areal data that might represent polygons or grids).
+The data can be collected across discrete points in time.
 The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
@@
 row.names(d) <- NULL
-d <- d[,c(1, 2, 3, 5, 6, 4)]
+d <- d[,c(1, 2, 3, 5, 6, 4, 7)]
 d <- d[!grepl("SPDE", d[,1]), ]
@@
 
-knitr::kable(d, format = "latex", caption = "Comparison of functionality between five R packages that can fit SPDE-based geostatistical GLMMs.", 
+knitr::kable(d, format = "latex", caption = "Comparison of functionality between six R packages that can fit geostatistical GLMMs.", 
   booktabs = TRUE, 
@@
     '\\addlinespace',
-    rep('', 3), # correlation stuff
+    rep('', 4), # correlation stuff
     '\\addlinespace',

### 2021-11-23 doc/paper.Rmd
Add some packages

@@
 row.names(d) <- NULL
-d <- d[,c(1, 2, 3, 5, 6, 4, 7)]
-d <- d[!grepl("SPDE", d[,1]), ]
+d <- d[,c(1, 2, 3, 5, 6, 4, 7, 8, 9)]
 names(d)[names(d) == "mgcv"] <- "mgcv\\footnote{a}"
 
-knitr::kable(d, format = "latex", caption = "Comparison of functionality between six R packages that can fit geostatistical GLMMs.", 
+knitr::kable(d, format = "latex", caption = "Comparison of functionality between eight R packages that can fit geostatistical GLMMs.", 
   booktabs = TRUE, 
@@
     '\\addlinespace',
-    rep('', 4), # correlation stuff
+    rep('', 5), # correlation stuff
     '\\addlinespace',
-    rep('', 10), # likelihood
+    rep('', 11), # likelihood
     '\\addlinespace',

### 2021-11-24 doc/appendix-pcod-VAST.Rmd
init

---
title: "Annotated example of index standardization with VAST"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(TMB)               
library(VAST)
library(sp)
library(sdmTMB)
library(magick)
```

We will fit a geostatistical spatiotemporal model with VAST for the purposes of index standardization, following the sdmTMB example. We will use a data set built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

To specify the mesh used to approximate the spatial process, which is used in the SPDE calculations, we use the k-means method in VAST. Rather than specifying the cutoff distance, meshes in VAST are typically generated by specifying only the number of knots, which we will later pass, along with other model settings to the function make_settings. We will use 205 knots, the same number in the mesh created in the other versions of this appendix.

For this example, we will include a factor predictor that represents the mean estimate for each time slice. Settings used for index standardization are specified partially by specifying `purpose = "index2"` but we also explicitly provide arguments for these and other key settings here.

```{r, echo = TRUE}
FieldConfig = matrix(c("IID","IID","IID","IID","IID","IID"), ncol=2, nrow=3, 
                     dimnames = list(c("Omega","Epsilon","Beta"),
                                     c("Component_1","Component_2")))
RhoConfig = c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
```

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).  Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid"=input_grid` in `fit_model`.

```{r settings, echo = TRUE}
# get coordinates in geographic coordinates from utm projection, after changing units to m
qcs_grid_ll <- qcs_grid
qcs_grid_ll$Y <- qcs_grid_ll$Y * 1000
qcs_grid_ll$X <- qcs_grid_ll$X * 1000

# with sp
coordinates(qcs_grid_ll) = ~X + Y
proj4string(qcs_grid_ll) <- CRS("+proj=utm +zone=9")
qcs_grid_ll <- as.data.frame(spTransform(qcs_grid_ll, CRS("+proj=longlat +datum=WGS84")))

## or with sf
# qcs_grid_ll <- st_as_sf(x = qcs_grid_ll,
#                         coords = c("X", "Y"),
#                         crs = "+proj=utm +zone=9")
# qcs_grid_ll <- st_transform(qcs_grid_ll, crs = "+proj=longlat +datum=WGS84")

# remove replicate locations for each year and format for VAST
qcs_grid_ll <- dplyr::filter(qcs_grid_ll, year == min(qcs_grid_ll$year))
input_grid <- cbind(Lat = qcs_grid_ll$Y, Lon = qcs_grid_ll$X, Area_km2 = 4)

settings <- make_settings(n_x = 205,
                          Region = "User", 
                          purpose = "index2", 
                          fine_scale = TRUE, 
                          zone = 9,
                          FieldConfig = FieldConfig,
                          RhoConfig = RhoConfig,
                          ObsModel = c(2,0), # this is for conventional logit-linked delta-gamma; c(10,2) for Tweedie
                          bias.correct = FALSE,
                          use_anisotropy = FALSE,
                          max_cells = Inf, # or set higher or lower to predict to greater/fewer locations
                          knot_method = "grid") # or "samples"
```

Next we will fit a GLMM (generalized linear mixed effects model).

```{r fit, echo = TRUE, results = 'hide'}
dir.create(paste0(here::here("doc", "appendix-VAST"))) # create folder for saved output

pcod$effort = 1 # effort is 1 when using CPUE instead of observed weight as the response

fit <- fit_model(settings = settings, 
                 Lat_i = pcod[,'lat'], 
                 Lon_i = pcod[,'lon'], 
                 t_i = pcod[,'year'], 
                 b_i = pcod[,'density'], 
                 a_i = pcod[,'effort'], 
                 input_grid = input_grid,
                 working_dir = paste0(here::here("doc", "appendix-VAST"), "/"))

```

We can look at parameter estimates and their standard errors.

```{r check-parameters, echo = TRUE}
fit$parameter_estimates$SD
```

While making custom plots of individual elements requires a fair bit of work to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want. While here we stick with the default set of plots, where `plot_set = 3`, one can specify different standard plots to make by changing the setting of this argument (see `?FishStatsUtils::plot_maps`).

```{r save-plots, echo = TRUE, warning = FALSE}
plot(fit, 
     check_residuals = FALSE,
     working_dir = paste0(here::here("doc", "appendix-VAST"), "/"))
```

The saved plots can be perused by browsing the exported image files in the working directory, but here we will read in some key plots as an example. We can start by looking at the location of samples and knots.

```{r plot-knots, echo = TRUE}
knots <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "Data_and_knots.png"))
plot(knots)
```

Then we can look at maps of the predicted population densities (here on the log scale).

```{r plot-density, echo = TRUE}
density <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "ln_density-predicted.png"))
plot(density)
```

And finally the biomass index.

```{r plot-index, echo = TRUE}
index <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "Index-Biomass.png"))
plot(index)
```

### 2021-11-24 doc/paper.Rmd

@@
   linesep = c(
-    rep('', 4), # coefs
+    rep('', 5), # coefs
     '\\addlinespace',

### 2021-11-24 doc/appendix-pcod-VAST.Rmd
make effort variable numeric

@@
 ```{r fit, echo = TRUE, results = 'hide'}
-dir.create(paste0(here::here("doc", "appendix-VAST"))) # create folder for saved output
+dir.create(paste0(here::here("doc", "appendix-VAST")), showWarnings = FALSE) # create folder for saved output
 
-pcod$effort = 1 # effort is 1 when using CPUE instead of observed weight as the response
+pcod$effort <- as.numeric(1) # effort is 1 when using CPUE instead of observed weight as the response
 

### 2021-11-24 doc/appendix-pcod-VAST.Rmd
Tidy VAST example

--- 2021-11-24 doc/appendix-pcod-VAST.Rmd
@@
   cache = TRUE,
-  autodep = TRUE
+  autodep = TRUE,
+  cache.comments = FALSE
 )
@@
 library(sdmTMB)
-library(magick)
+library(here)
 ```
@@
 ```{r, echo = TRUE}
-FieldConfig = matrix(c("IID","IID","IID","IID","IID","IID"), ncol=2, nrow=3, 
-                     dimnames = list(c("Omega","Epsilon","Beta"),
-                                     c("Component_1","Component_2")))
-RhoConfig = c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
+FieldConfig <- matrix(c("IID", "IID", "IID", "IID", "IID", "IID"),
+  ncol = 2, nrow = 3,
+  dimnames = list(
+    c("Omega", "Epsilon", "Beta"),
+    c("Component_1", "Component_2")
+  )
+)
+RhoConfig <- c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
 ```
@@
 Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
-Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid"=input_grid` in `fit_model`.
+Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.
 
 ```{r settings, echo = TRUE}
-# get coordinates in geographic coordinates from utm projection, after changing units to m
+# get coordinates in geographic coordinates from UTM projection, after changing units to m
 qcs_grid_ll <- qcs_grid
@@
 
-# with sp
-coordinates(qcs_grid_ll) = ~X + Y
+# with sp:
+coordinates(qcs_grid_ll) <- ~ X + Y
 proj4string(qcs_grid_ll) <- CRS("+proj=utm +zone=9")
@@
 
-## or with sf
-# qcs_grid_ll <- st_as_sf(x = qcs_grid_ll,
+# or with sf:
+# qcs_grid_ll <- sf::st_as_sf(
+#   x = qcs_grid_ll,
 #   coords = c("X", "Y"),
-#                         crs = "+proj=utm +zone=9")
-# qcs_grid_ll <- st_transform(qcs_grid_ll, crs = "+proj=longlat +datum=WGS84")
+#   crs = "+proj=utm +zone=9"
+# )
+# qcs_grid_ll <- sf::st_transform(qcs_grid_ll, crs = "+proj=longlat +datum=WGS84")
 
 # remove replicate locations for each year and format for VAST
-qcs_grid_ll <- dplyr::filter(qcs_grid_ll, year == min(qcs_grid_ll$year))
+qcs_grid_ll <- subset(qcs_grid_ll, year == min(qcs_grid_ll$year))
 input_grid <- cbind(Lat = qcs_grid_ll$Y, Lon = qcs_grid_ll$X, Area_km2 = 4)
 
-settings <- make_settings(n_x = 205,
+settings <- make_settings(
+  n_x = 205, # number of vertices in the SPDE mesh
   Region = "User",
-                          purpose = "index2", 
-                          fine_scale = TRUE, 
+  purpose = "index2", # index of abundance with Gamma for positive catches
+  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
   zone = 9,
@@
   RhoConfig = RhoConfig,
-                          ObsModel = c(2,0), # this is for conventional logit-linked delta-gamma; c(10,2) for Tweedie
+  ObsModel = c(2, 0), # this is for conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
   bias.correct = FALSE,
   use_anisotropy = FALSE,
-                          max_cells = Inf, # or set higher or lower to predict to greater/fewer locations
-                          knot_method = "grid") # or "samples"
+  max_cells = Inf, # use all grid cells from the extrapolation grid
+  knot_method = "grid" # or "samples"
+)
 ```
@@
 ```{r fit, echo = TRUE, results = 'hide'}
-dir.create(paste0(here::here("doc", "appendix-VAST"))) # create folder for saved output
+ # create folder for saved output:
+dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)
 
-pcod$effort = 1 # effort is 1 when using CPUE instead of observed weight as the response
+# effort is 1 when using CPUE instead of observed weight as the response:
+pcod$effort <- as.numeric(1)
 
-fit <- fit_model(settings = settings, 
-                 Lat_i = pcod[,'lat'], 
-                 Lon_i = pcod[,'lon'], 
-                 t_i = pcod[,'year'], 
-                 b_i = pcod[,'density'], 
-                 a_i = pcod[,'effort'], 
+fit <- fit_model(
+  settings = settings,
+  Lat_i = pcod[, "lat"],
+  Lon_i = pcod[, "lon"],
+  t_i = pcod[, "year"],
+  b_i = pcod[, "density"],
+  a_i = pcod[, "effort"],
   input_grid = input_grid,
-                 working_dir = paste0(here::here("doc", "appendix-VAST"), "/"))
-
+  working_dir = paste0(here("doc", "appendix-VAST"), "/")
+)
 ```
@@
 ```{r save-plots, echo = TRUE, warning = FALSE}
-plot(fit, 
+plot(
+  fit,
   check_residuals = FALSE,
-     working_dir = paste0(here::here("doc", "appendix-VAST"), "/"))
+  working_dir = paste0(here("doc", "appendix-VAST"), "/")
+)
 ```
@@
 ```{r plot-knots, echo = TRUE}
-knots <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "Data_and_knots.png"))
-plot(knots)
+knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
 ```
@@
 ```{r plot-density, echo = TRUE}
-density <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "ln_density-predicted.png"))
-plot(density)
+knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
 ```
@@
 ```{r plot-index, echo = TRUE}
-index <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "Index-Biomass.png"))
-plot(index)
+if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
+}
+if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
+}
 ```

### 2021-11-24 doc/appendix-pcod-VAST.Rmd
Shrink fig

@@
 
-```{r plot-index, echo = TRUE}
+```{r plot-index, echo = TRUE, out.width="3in"}
 if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {

### 2021-11-24 doc/appendix-pcod-sdmTMB.Rmd
Clean up appendix

@@
 The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in. See examples included in `?make_mesh`.
+Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in.
+See examples included in `?make_mesh`.
 
@@
 ```{r fit-pcod, warning=FALSE, message=FALSE}
-fit <- sdmTMB(density ~ 0 + as.factor(year), data = pcod,
+fit <- sdmTMB(density ~ 0 + as.factor(year),
+  data = pcod,
   time = "year", mesh = mesh, family = tweedie(link = "log")
@@
 ggplot(pred, aes(X, Y, colour = resid)) +
-    geom_point() + facet_wrap(~year) +
-    coord_fixed() + scale_colour_gradient2()
+  geom_point() +
+  facet_wrap(~year) +
+  coord_fixed() +
+  scale_colour_gradient2()
 ```
@@
 ggplot(qcs_grid, aes(X, Y, fill = cv)) +
-    geom_raster() + facet_wrap(~year) +
-    coord_fixed() + scale_fill_viridis_c(trans = "log10", option = "D")
+  geom_raster() +
+  facet_wrap(~year) +
+  coord_fixed() +
+  scale_fill_viridis_c(trans = "log10", option = "D")
 ```
@@
 pred2 <- predict(
-  fit, newdata = qcs_grid, return_tmb_object = TRUE,
+  fit,
+  newdata = qcs_grid, return_tmb_object = TRUE,
   area = qcs_grid$area
@@
 ind <- get_index(pred2)
-ggplot(ind, aes(year, est/1000)) + geom_line() +
-  geom_ribbon(aes(ymin = lwr/1000, ymax = upr/1000), alpha = 0.4) +
+ggplot(ind, aes(year, est / 1000)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000), alpha = 0.4) +
   ylab("Biomass (t)")
@@
 ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)
-ggplot(ind, aes(year, est/1000)) + geom_line() +
-  geom_ribbon(aes(ymin = lwr/1000, ymax = upr/1000), alpha = 0.4) +
+ggplot(ind, aes(year, est / 1000)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000), alpha = 0.4) +
   ylab("Biomass (t)")

### 2021-11-24 doc/paper.Rmd
dir.create

@@
 ```{r diagram, fig.cap="(ref:diagram-fig)"}
+dir.create("figs", showWarnings = FALSE)
 f <- here::here("figs", "diagram.pdf")

### 2021-11-24 doc/appendix-pcod-VAST.Rmd
Minor VAST tweaks

@@
 
-```{r, echo = TRUE}
+```{r config, echo = TRUE}
 FieldConfig <- matrix(c("IID", "IID", "IID", "IID", "IID", "IID"),
@@
 )
+FieldConfig
 RhoConfig <- c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
@@
 ```
+
+This document was built using:
+
+```{r, echo=TRUE}
+R.Version()$version.string
+packageVersion("VAST")
+packageVersion("FishStatsUtils")
+```

### 2021-11-24 doc/appendix-pcod-VAST.Rmd

@@
 # effort is 1 when using CPUE instead of observed weight as the response:
-pcod$effort <- as.numeric(1)
-
+pcod$effort <- 1
+pcod <- as.data.frame(pcod) # ensure not a tibble
 fit <- fit_model(

### 2021-11-24 doc/paper.Rmd

@@
 
+```{r rnaturalearthhires, include=FALSE, echo=FALSE}
+if (!require("rnaturalearthhires", quietly = TRUE)) {
+  stop("Please install 'rnaturalearthhires'.\n",
+  "`remotes::install_github('ropensci/rnaturalearthhires')`")
+}
+```
+
 # Introduction
@@
 -->
-
-<!-- User interfaces to classes of models fit by these template systems can greatly  -->
-
-<!-- reduce these challenges for the vast majority of models. -->
-
-<!-- keep the above even briefer? -->
-
-<!-- inlabru is "very good" but not entirely familiar to those used to `glm()`, lme4, glmmTMB, or mgcv; -->
-
-<!-- *TODO: the differences are many and more than can be listed here. Keep brief add Table or discussion later about differences and when to pick one over the other?* -->
 
@@
 
-<!-- - high-level paragraph or two about what the model includes, referencing Fig 1 -->
 <!-- - high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?) -->
-<!-- - high-level paragraph about simulation capabilities and + testing performed (reference supplement) -->
-<!-- - model selection/cross validation paragraph -->
-<!-- - derived quantities paragraph (or just put this in the examples?) -->
-
-<!-- - example 1 + figs -->
-<!-- - example 2 + figs -->
-
-<!--     - perhaps a paragraph on how it achieves what Jim outlined @thorson2019 as necessary: -->
-<!--         -   "VAST represents four main design principles.
-Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
-<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
-<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
-<!--         -   Condition on missing covariates (check, random fields) -->
-<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
-
-<!-- sdmTMB in its current form achieves 3 of the 4 design principles outlined by @thorson2019 for index standardization:  -->
-
-<!-- (e.g., 'catchability' and 'habitat' covariates in VAST merely depend on the values supplied to `predict()` in sdmTMB). -->
-
-<!-- # Model -->
 
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-
-<!-- [paragraph about spde and gaussian process / Matern covariance here] -->
-
-<!-- Paragraph on model selection here-->
 
@@
 Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
-
-<!-- Paragraph on advanced options here?
-Specifically, this could address
-- PC-priors -- and other penalized priors
-- Non-linear stuff / etc
-- Linkages to Rstan
-- Simulation
--->
 
@@
 
-<!--         -   "VAST represents four main design principles.
-Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
-<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
-<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
-<!--         -   Condition on missing covariates (check, random fields) -->
-<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
-<!-- However, VAST can fit multivariate spatiotemporal models (REF) and Poisson-link delta model (REF), whereas sdmTMB cannot. -->
-
 inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
@@
 <!--
-# Tables
-
-1. Comparison to related software?
-
-- VAST
-- INLA
-- mgcv
-- glmmTMB
-- glmmfields?
-- other stuff on CRAN, spBayes etc.
-
-# Figures
-
-1. Flow diagram of functions? Not needed?
-1. Speed with various knots and/or spatial complexity in comparison to INLA + VAST + ...?
-1. Example application 1 - SDM, show mesh(?) and/or raw data, show predictions and SD in space,
-1. Example application 2 - population index with stitched survey?
-
-Content:
-
--   outline the general need for this type of modelling
-
--   outline the need for this type of package to make the models accessible and rapid to iterate on
 
@@
 
--   table comparison to related packages
-
 -   case studies
@@
     -   stitching surveys?
-
--   how it achieves the steps Jim described [@thorson2019] in what a VAST-related package has to be able to do (and how VAST does it vs. how sdmTMB does it)
-    -   "VAST represents four main design principles.
-Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include."
-    -   Area weighting (check, helper functions to sum up weighted grid cells)
-    -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`)
-    -   Condition on missing covariates (check, random fields)
-    -   Bridge between univariate and multivariate applications (no! should we?)
 

### 2021-11-25 doc/appendix-snow-sdmTMB.Rmd
draft of owl appendix

---
title: "Annotated example of a spatially varying effect with sdmTMB"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
theme_set(theme_minimal())
```

This illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
We will use counts of Snow Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts (CBCs) between 1979 and 2020. 
This citizen science survey is organized by the by the National Audubon Society (data available by request from cbcadmin@audubon.org). 
Each survey is conducted between December 14 to January 5 in 24.14 kilometre diameter circle. 
Circle locations are chosen by local volunteers and therefore reflect both locations of higher human densities and known bird watching locations.
Each circle (given a unique "CBCID") is generally surveyed annually by at least 10 volunteers, or deemed inactive.

Snow Owls are an arctic breeding species is an irruptive migrant meaning that the number of individuals choosing to migrate varies dramatically from year to year. 
The reasons for this variation are still not well understood, but seem to be at least partially related to very high abundances of food during the breeding season and therefore higher productivity [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the University of East Anglia's [Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution and the only portion well surveyed by CBCs.

```{r load-data, cache=FALSE} 
dat <- readRDS(here::here("owls/SNOW_data.rds"))
glimpse(dat)
```
Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km.

```{r proj, echo=TRUE}
proj <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
```

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations. The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`). This mesh uses a cutoff distance of 1.5 units (or 150 km).

```{r mesh, echo=TRUE, cache=FALSE}
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 1.5)
```

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(mesh)
```
The result is a mesh with `r mesh$mesh$n` knots.

Next, we will plot the counts through time to see if there is an obvious overall trend, or relationship with NAO.

```{r explore, fig.width = 8}
ggplot(dat, aes(year, log(count+1), colour = nao)) + 
  geom_jitter(alpha = 0.5) + 
  scale_colour_viridis_c()
```

Because there does not appear to be much correlation between years, year will be included as a random factor: `(1|year_f)`.

We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "IID"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance. 
The random intercept for year allows the mean estimate for each of these spatiotemporal fields to vary. 
If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.

Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America. 
Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.

First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
We will use `normalize = TRUE` to speed up optimization. 
Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.

```{r fit-poisson, echo=TRUE, eval=FALSE}
m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
             time = "year",
             spatial_varying = ~ 0 + nao,
             spatial = "on", 
             spatiotemporal = "IID",
             family = poisson(link = "log"),
             control = sdmTMBcontrol(normalize = TRUE),
             data = dat, mesh = mesh)
```

```{r fit-negb1, echo=TRUE, eval=FALSE}
m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom1(link = "log"),
            control = sdmTMBcontrol(normalize = TRUE),
            data = dat, mesh = mesh)
```

```{r fit-negb2, echo=TRUE, eval=FALSE}
m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            control = sdmTMBcontrol(normalize = TRUE),
            data = dat, mesh = mesh)
```

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE}
# m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
# m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
# m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
```

```{r aic, echo=TRUE, cache=TRUE}
# AIC(m0,m1,m2)
```

The `nbinom2` distribution appears to have the best fit. 
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and then check our residuals. 
This method of checking residuals is not guarenteed to be accurate, so an additional step using STAN could be added.

```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            reml = TRUE,
            data = dat, mesh = mesh)
```

We can predict on the original data locations:

```{r predict}
pred <- predict(m)
```

We can inspect randomized quantile residuals:

```{r qqnorm, out.width="5in"}
pred$resid <- residuals(m)
qqnorm(pred$resid)
qqline(pred$resid)
```

```{r}
library(tmbstan)
s <- tmbstan(m$tmb_obj, iter = 200, chains = 1, warmup = 199)
pstan <- predict(m, tmbstan_model = s)
resid <- as.numeric(pstan) - m$data$observed
# or do any PIT residuals randomization here...
qqnorm(resid);qqline(resid) # should be right
```

```{r plot-resids}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2()
```

Print the model to inspect the fit:

```{r print-fit}
print(m)
```

Extract the parameter estimates as a data frame:

```{r tidy}
tidy(m, conf.int = TRUE)
tidy(m, effects = "ran_pars", conf.int = TRUE)
```

In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and reproject. 
One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from   http://www.naturalearthdata.com/downloads/10m-physical-vectors/ and transform them to the correct projection using `sf::st_transform(crs = proj)`.

```{r proj-p, echo=TRUE, eval=FALSE}
p <- pred %>% mutate(
  X = X * 100000,
  Y = Y * 100000
)

p_proj <- p %>% mutate(x = X, y = Y) %>% st_as_sf(., coords = c("x", "y"), crs = proj)
```

We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO. 
First, we extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.

```{r p-mean, echo=TRUE, eval=FALSE}
b <- tidy(m)

p_mean <- p %>% group_by(CBCID) %>% summarise(
  X = mean(X), Y= mean(Y), 
  mean_est_count = exp(mean(est)), 
  nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=FALSE}
ggplot(data = p_proj) +
  geom_point(data = p_mean, aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), # adjust buffers slightly
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "top", label.position = "bottom")) +
  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

There are four other kinds of predictions that we get out of the model. 
We can plot our predictions of winter owl counts that incorporate all fixed effects and random effects:

```{r plot-all-effects, echo=TRUE, eval=FALSE}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(
    # transform to make variation in low number visible 
    trans="sqrt", 
    # and also trim the top extremes of the colour axis 
    limits = c(0, quantile(exp(p_proj$est), 0.995)), na.value = "yellow") +
  labs(x = "Longitude", y = "Latitude", colour = "Predicted\ncount") +
  # theme_bw() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
  facet_wrap(~year_f)
```

Note that for these annually varying estimates, we are showing only the most recent 25 years. 
We can also look at just the fixed effects, but in this case these only include the overall effect of mean annual NAO:

```{r plot-fixed-effects}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c() + 
  # theme_bw() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
  facet_wrap(~year_f)
```

We can look at the spatial random effects that represent consistent deviations in space that are not accounted for by our fixed effects. 
In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers like coasts.

```{r plot-spatial-effects, echo=TRUE, eval=FALSE}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), # adjust buffers slightly
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "top", label.position = "bottom")) +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be things like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_fill_gradient2() +
  # theme_bw() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
  facet_wrap(~year_f)
```

### 2021-11-25 snow.R
make owl mesh coarser, make reml version, start testing stan residuals, edit fig code

@@ -11,13 +11,14 @@ library(tidyverse)
 sdat <- readRDS("owls/SNOW_data.rds")
 
 # length(unique(sdat$CBCID))
+ggplot(sdat, aes(year, count)) + geom_point() + geom_smooth()
 
 
 #### spatially varying model ####
 
 library(sdmTMB)
 
-mesh <- make_mesh(sdat, xy_cols = c("X", "Y"), cutoff = 1)
+mesh <- make_mesh(sdat, xy_cols = c("X", "Y"), cutoff = 1.5)
 plot(mesh)
 mesh$mesh$n
 
@@ -25,49 +26,84 @@ m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
              time = "year",
              spatial_varying = ~ 0 + nao,
              family = poisson(link = "log"),
-             control = sdmTMBcontrol(normalize = TRUE),
+             # control = sdmTMBcontrol(normalize = TRUE),
              spatial = "on", spatiotemporal = "IID",
              data = sdat, mesh = mesh)
 m0
-saveRDS(m0, "owls/snow_w_main_effect_0.rds")
+saveRDS(m0, "owls/snow_w_main_effect_0_150km.rds")
 
 m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
             time = "year",
             spatial_varying = ~ 0 + nao,
             family = nbinom1(link = "log"),
-            control = sdmTMBcontrol(normalize = TRUE),
+            # control = sdmTMBcontrol(normalize = TRUE),
             spatial = "on", spatiotemporal = "IID",
             data = sdat, mesh = mesh)
 m1
-saveRDS(m1, "owls/snow_w_main_effect_1.rds")
+saveRDS(m1, "owls/snow_w_main_effect_1_150km.rds")
+
 
 m <- sdmTMB(count ~ 1 + nao + (1|year_f),
-            time = "year",
-            # time_varying = ~ 1,
-            spatial_varying = ~ 0 + nao,
-            family = nbinom2(link = "log"),
-            control = sdmTMBcontrol(normalize = TRUE),
-            # this helps estimate ranges by constraining complexity/preventing ranges getting too small
-            # share_range = F,
-            # priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 1, sigma_lt = 3),
[…]

### 2021-11-26 doc/paper.Rmd
draft owl para

@@
 
+# Citizen science count data with a spatially varying effect
+
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
+The reasons for the interannual variation in the number of individuals migrating south are still not well understood, but seem to be related to very high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
+The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
+Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially variable effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
+We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial distribution, random intercepts for year, both spatial and spatiotemporal random fields, and a spatially varying effect of NAO (**see appendix for more detail**):
+
+```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
+fit_owls <- sdmTMB(
+  count ~ 1 + nao + (1|year_f),
+  spatial_varying = ~ 0 + nao,     
+  family = nbinom2(link = "log"), data = snow, mesh = mesh, reml = TRUE,  
+  time = "year", spatial = "on", spatiotemporal = "IID"
+  )
+```
+
+We found a weak positive effect of annual mean NAO on overall counts, but that there was also a southeast to northwest gradient in the intensity of this effect that would be consistent with owls closer to the Atlantic coast and migrating the furthest south being the most effected by NAO (Fig.
+\@ref(fig:owl-nao)).
+
+(ref:owl-nao-fig) A spatially varying effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts between 1979 and 2020. 
+Circles represent all count locations and are roughly scaled to the mean number of owls observed each year (range: 0 to 8).
+
+```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="3.5in", fig.align='center'}
+owl <- here::here("figs", "owl_nao_effect.pdf")
+if (!file.exists(owl)) {
+  download.file("https://www.dropbox.com/s/3aco85mizuplz7z/owl_nao_effect.pdf?dl=1", 
+    destfile = owl)
+}
+knitr::include_graphics(owl)
+```
+
 <!--

### 2021-11-26 doc/appendix-snow-sdmTMB.Rmd
clean up owl appendix

@@
 library(sdmTMB)
-theme_set(theme_minimal())
+# library(sf)
+theme_set(theme_bw())
+# theme_set(theme_minimal())
 ```
@@
 Each circle (given a unique "CBCID") is generally surveyed annually by at least 10 volunteers, or deemed inactive. 
+While these locations are not random, the consistent sampling through time still make them useful for exploring spatiotemporal processes.
 
@@
 ```{r load-data, cache=FALSE} 
-dat <- readRDS(here::here("owls/SNOW_data.rds"))
-glimpse(dat)
+snow <- readRDS(here::here("owls/SNOW_data.rds"))
+glimpse(snow)
 ```
+
 Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km.
@@
 ```{r proj, echo=TRUE}
-proj <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
+proj <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 
++datum=NAD83 +units=m +no_defs"
 ```
@@
 ```{r mesh, echo=TRUE, cache=FALSE}
-mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 1.5)
+mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
 ```
 
-```{r plot-mesh, fig.asp=0.9, out.width="4in"}
+```{r plot-mesh, fig.asp=0.9, out.width="6in"}
 plot(mesh)
 ```
+
 The result is a mesh with `r mesh$mesh$n` knots.
 
-Next, we will plot the counts through time to see if there is an obvious overall trend, or relationship with NAO.
+Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO.
 
-```{r explore, fig.width = 8}
-ggplot(dat, aes(year, log(count+1), colour = nao)) + 
+```{r explore, fig.asp=0.4, fig.width = 6}
+ggplot(snow, aes(year, count, colour = nao)) + 
   geom_jitter(alpha = 0.5) + 
@@
 
+# Model fitting
+
+First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
+<!-- We will use `normalize = TRUE` to speed up optimization.  -->
+Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.
+
+Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America.
+Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
 Because there does not appear to be much correlation between years, year will be included as a random factor: `(1|year_f)`. 
-
 We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "IID"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance.
@@
 If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
-
-Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America.
-Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
-
-First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
-We will use `normalize = TRUE` to speed up optimization. 
-Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.
 
@@
              family = poisson(link = "log"),
-             control = sdmTMBcontrol(normalize = TRUE),
-             data = dat, mesh = mesh)
+             data = snow, mesh = mesh)
 ```
@@
             family = nbinom1(link = "log"),
-            control = sdmTMBcontrol(normalize = TRUE),
-            data = dat, mesh = mesh)
+            data = snow, mesh = mesh)
 ```
@@
             family = nbinom2(link = "log"),
-            control = sdmTMBcontrol(normalize = TRUE),
-            data = dat, mesh = mesh)
+            data = snow, mesh = mesh)
 ```
@@
 ```{r load-models, warning=FALSE, message=FALSE, echo=FALSE}
-# m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
-# m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
-# m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
+m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
+m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
+m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
 m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
@@
 ```{r aic, echo=TRUE, cache=TRUE}
-# AIC(m0,m1,m2)
+AIC(m0,m1,m2)
 ```
@@
 We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and then check our residuals.
-This method of checking residuals is not guarenteed to be accurate, so an additional step using STAN could be added.
 
@@
             reml = TRUE,
-            data = dat, mesh = mesh)
+            data = snow, mesh = mesh)
 ```
 
-We can predict on the original data locations:
+Print the model to inspect the fit:
 
-```{r predict}
+```{r print-fit, message=FALSE}
+print(m)
+```
+
+View the confidence intervals on both fixed and random parameter estimates using `tidy()`.
+
+```{r tidy}
+tidy(m, conf.int = TRUE)
+tidy(m, effects = "ran_pars", conf.int = TRUE)
+```
+
+We can also renerate predictions and residuals for the original data locations:
+
+```{r predict, message=FALSE}
 pred <- predict(m)
@@
 
-We can inspect randomized quantile residuals:
+We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate. 
+An additional step using STAN could be added.
 
-```{r qqnorm, out.width="5in"}
+```{r qqnorm, fig.asp=0.9, out.width="3in"}
 pred$resid <- residuals(m)
@@
 
-```{r}
-library(tmbstan)
-s <- tmbstan(m$tmb_obj, iter = 200, chains = 1, warmup = 199)
-pstan <- predict(m, tmbstan_model = s)
-resid <- as.numeric(pstan) - m$data$observed
-# or do any PIT residuals randomization here...
-qqnorm(resid);qqline(resid) # should be right
-```
-
-```{r plot-resids}
+```{r plot-resids, fig.width = 7}
 ggplot(pred, aes(X, Y, colour = resid)) +
@@
 
-Print the model to inspect the fit:
+# Plotting spatial predictions
 
-```{r print-fit}
-print(m)
-```
+In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will appear in latitude and longitude.
 
-Extract the parameter estimates as a data frame:
+```{r proj-p, echo=TRUE, eval=T}
+library(sf)
 
-```{r tidy}
-tidy(m, conf.int = TRUE)
-tidy(m, effects = "ran_pars", conf.int = TRUE)
-```
-
-In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and reproject. 
-One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from   http://www.naturalearthdata.com/downloads/10m-physical-vectors/ and transform them to the correct projection using `sf::st_transform(crs = proj)`.
-
-```{r proj-p, echo=TRUE, eval=FALSE}
 p <- pred %>% mutate(
@@
 
+One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transform them to the correct projection using `sf::st_transform(crs = proj)`.
+
+```{r shapes, echo=FALSE, message=FALSE}
+library("rnaturalearth")
+library("rnaturalearthdata")
+coast <- ne_coastline(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(crs = proj)
+lakes <- sf::st_read(here::here("owls/ne_10m_lakes"))
+lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
+```
+
+## Spatially varying effect
+
 We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
@@
 
-```{r p-mean, echo=TRUE, eval=FALSE}
+```{r p-mean, message=FALSE, echo=TRUE, eval=T}
 b <- tidy(m)
@@
 
-```{r plot-map, echo=TRUE, eval=FALSE}
+```{r plot-map, echo=TRUE, eval=T}
 ggplot(data = p_proj) +
-  geom_point(data = p_mean, aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
+  geom_point(data = p_mean, 
+     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
   geom_sf(data = coast, fill = NA) +
   geom_sf(data = lakes, fill = NA) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), # adjust buffers slightly
+  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
            ylim = c(min(p_proj$Y), max(p_proj$Y))) +
-  scale_colour_viridis_c(
-    guide = guide_colourbar(direction = "horizontal", title.position = "top", label.position = "bottom")) +
-  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
+  scale_colour_viridis_c(guide = guide_colourbar(
+    direction = "horizontal", title.position = "top", label.position = "bottom")) +
+  guides(size = "none") + 
+  labs(colour = "NAO effect\non Snowy Owl count") +
   theme_bw() + theme(
@@
 
-There are four other kinds of predictions that we get out of the model. 
-We can plot our predictions of winter owl counts that incorporate all fixed effects and random effects:
+## Predicted counts
 
-```{r plot-all-effects, echo=TRUE, eval=FALSE}
+These are the predictions of winter owl counts that incorporate all fixed effects and random effects:
+
+```{r plot-all-effects, echo=TRUE, eval=T, fig.width = 7}
 ggplot(data = filter(p_proj, year > 1995)) +
@@
              size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
+  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
+           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
   scale_colour_viridis_c(
@@
   labs(x = "Longitude", y = "Latitude", colour = "Predicted\ncount") +
-  # theme_bw() + 
-  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank()) +
+  facet_wrap(~year_f)
+```
+Note that for these annually varying estimates, we are showing only the most recent 25 years.
+
+## Other components of predicted counts
+
+In addition to the spatially varying effect we started with, there are three other components that go into the overall predictions that we can plot spatially.
+
+1. Just fixed effects
+
+<!-- confirm that the random effect of year is not included here? -->
+In this case these only include the overall effect of mean annual NAO:
+
+```{r plot-fixed-effects, echo=TRUE, eval=T}
+ggplot(data = filter(p_proj, year > 1995)) +
+  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
+             size = 0.4, alpha = 0.5) +
+  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
+           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
+  scale_colour_viridis_c() + 
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank()) +
   facet_wrap(~year_f)
@@
 
-Note that for these annually varying estimates, we are showing only the most recent 25 years. 
-We can also look at just the fixed effects, but in this case these only include the overall effect of mean annual NAO:
+2. Static spatial random effects
 
-```{r plot-fixed-effects}
-ggplot(data = filter(p_proj, year > 1995)) +
-  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
-             size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_colour_viridis_c() + 
-  # theme_bw() + 
-  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
-  facet_wrap(~year_f)
-```
-
-We can look at the spatial random effects that represent consistent deviations in space that are not accounted for by our fixed effects. 
+We can look at the spatial random effects (`omega_s`) that represent consistent deviations in space that are not accounted for by our fixed effects. 
 In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=FALSE}
+```{r plot-spatial-effects, echo=TRUE, eval=T}
 ggplot(data = p_proj) +
@@
   geom_sf(data = lakes, fill = NA) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), # adjust buffers slightly
+  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
            ylim = c(min(p_proj$Y), max(p_proj$Y))) +
-  scale_colour_viridis_c(
-    guide = guide_colourbar(direction = "horizontal", title.position = "top", label.position = "bottom")) +
+  scale_colour_viridis_c(guide = guide_colourbar(
+    direction = "horizontal", title.position = "top", label.position = "bottom")) +
   theme_bw() + theme(
@@
 
-Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
+3. Spatiotemporal random effects
+
+Finally, we can look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
 These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
@@
 
-```{r plot-spatiotemporal-effects}
+```{r plot-spatiotemporal-effects, echo=TRUE, eval=T}
 ggplot(data = filter(p_proj, year > 1995)) +
@@
              size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_fill_gradient2() +
-  # theme_bw() + 
-  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
+  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
+           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
+  scale_colour_viridis_c() +
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank()) +
   facet_wrap(~year_f)

### 2021-11-26 doc/appendix-snow-sdmTMB.Rmd
add refs
edits to owl appendix

@@
     highlight: tango
+bibliography: refs.bib
 ---
@@
   fig.width = 7,
-  fig.asp = 0.618,
+  fig.asp = 1,
   cache = TRUE,
@@
 library(sdmTMB)
-# library(sf)
-theme_set(theme_bw())
-# theme_set(theme_minimal())
+library(sf)
+theme_set(theme_bw() + theme(
+  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
+  legend.position = "top", legend.justification = c(0, 0), 
+  axis.title = element_blank()))
 ```
@@
 Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
+This projection was achieved with the following crs code using `sf::st_transform(crs = proj)`.
 
@@
 
-```{r plot-mesh, fig.asp=0.9, out.width="6in"}
+```{r plot-mesh, out.width="6in"}
 plot(mesh)
@@
 
-```{r explore, fig.asp=0.4, fig.width = 6}
+```{r explore, fig.asp = 0.3}
 ggplot(snow, aes(year, count, colour = nao)) + 
-  geom_jitter(alpha = 0.5) + 
-  scale_colour_viridis_c()
+  geom_jitter() + 
+  scale_colour_viridis_c() +
+  theme_bw() 
 ```
@@
 
-```{r load-models, warning=FALSE, message=FALSE, echo=FALSE}
+```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
 m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
@@
 The `nbinom2` distribution appears to have the best fit. 
-We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and then check our residuals.
+We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model.
 
@@
 
-```{r print-fit, message=FALSE}
+```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE}
 print(m)
@@
 
-We can also renerate predictions and residuals for the original data locations:
+We can then generate predictions and residuals for the original data locations:
 
@@
 pred <- predict(m)
+pred$resid <- residuals(m)
 ```
@@
 We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate.
-An additional step using STAN could be added.
 
-```{r qqnorm, fig.asp=0.9, out.width="3in"}
-pred$resid <- residuals(m)
+```{r qqnorm, out.width = "33%"}
 qqnorm(pred$resid)
@@
 
-```{r plot-resids, fig.width = 7}
+More accurate residuals can be generated using STAN, but this is a very slow process!
+In this case, it took more than 2.5 hrs.
+
+```{r stan-resids, eval=FALSE}
+library(tmbstan)
+stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
+eta <- predict(m, tmbstan_model = stan_fit)
+mu <- m$family$linkinv(as.numeric(eta))
+r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
+saveRDS(r_mcmc, "mcmc_residuals.rds")
+```
+
+```{r stan-dat, echo=FALSE, cache=FALSE}
+r_mcmc <- readRDS(here::here("owls/mcmc_residuals.rds"))
+```
+
+```{r stan-qqnorm, out.width = "33%"}
+qqnorm(r_mcmc)
+qqline(r_mcmc) 
+```
+
+<!-- \clearpage -->
+
+```{r plot-resids, fig.width = 8}
 ggplot(pred, aes(X, Y, colour = resid)) +
@@
   coord_fixed() +
-  scale_colour_gradient2()
+  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
+  scale_colour_gradient2(guide = guide_colourbar(title.position = "top"))
 ```
+
+<!-- \clearpage -->
 
@@
 
-```{r shapes, echo=FALSE, message=FALSE}
+```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
 library("rnaturalearth")
@@
   sf::st_transform(crs = proj)
-lakes <- sf::st_read(here::here("owls/ne_10m_lakes"))
+lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
 lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
@@
   scale_colour_viridis_c(guide = guide_colourbar(
-    direction = "horizontal", title.position = "top", label.position = "bottom")) +
+    direction = "horizontal", title.position = "top")) +
   guides(size = "none") + 
@@
 
+\clearpage
+
 ## Predicted counts
 
-These are the predictions of winter owl counts that incorporate all fixed effects and random effects:
+These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
+The colour axis has been transformed and trimmed to improve the visibility of variation in count estimates.
 
-```{r plot-all-effects, echo=TRUE, eval=T, fig.width = 7}
+```{r plot-all-effects, echo=TRUE, eval=T}
 ggplot(data = filter(p_proj, year > 1995)) +
@@
            ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_colour_viridis_c(
-    # transform to make variation in low number visible 
-    trans="sqrt", 
-    # and also trim the top extremes of the colour axis 
-    limits = c(0, quantile(exp(p_proj$est), 0.995)), na.value = "yellow") +
-  labs(x = "Longitude", y = "Latitude", colour = "Predicted\ncount") +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank()) +
+  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
+    # can also trim the extreme high values
+    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
+    guide = guide_colourbar(title.position = "top")) +
+  labs(colour = "Predicted counts") + 
+  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
   facet_wrap(~year_f)
 ```
-Note that for these annually varying estimates, we are showing only the most recent 25 years.
 
+Note that for all of these annually varying predictions, we are showing only the most recent 25 years.
+
+\clearpage
 ## Other components of predicted counts
 
-In addition to the spatially varying effect we started with, there are three other components that go into the overall predictions that we can plot spatially.
+In addition to the spatially varying effect that we started with, there are three other components that go into the overall predictions that we can plot spatially.
 
-1. Just fixed effects
+### 1. Just fixed effects
 
@@
            ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_colour_viridis_c() + 
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank()) +
+  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
+    guide = guide_colourbar(title.position = "top")) +
+  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
   facet_wrap(~year_f)
@@
 
-2. Static spatial random effects
+\clearpage
+
+### 2. Static spatial random effects
 
@@
   scale_colour_viridis_c(guide = guide_colourbar(
-    direction = "horizontal", title.position = "top", label.position = "bottom")) +
+    direction = "horizontal", title.position = "top")) +
   theme_bw() + theme(
@@
 
-3. Spatiotemporal random effects
+\clearpage
+
+### 3. Spatiotemporal random effects
 
@@
            ylim = c(min(p_proj$Y), max(p_proj$Y))) +
-  scale_colour_viridis_c() +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank()) +
+  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
+  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
   facet_wrap(~year_f)
 ```
+
+\clearpage
+
+# References
+
+::: {#refs}
+:::

### 2021-11-26 snow.R
misc extra snow.R code

@@ -108,11 +108,46 @@ qqnorm(sdat$residuals0)
 sdat$residuals1 <- residuals(m1)
 qqnorm(sdat$residuals1);abline(a = 0, b = 1)
 
+# ## TODO:add other residual checks
+# mini_sim <- function(fit) {
+#   params <- fit$tmb_obj$env$parList()
+#   tmb_data <- fit$tmb_data
+#   # tmb_data$sim_re <- rep(1, length(tmb_data$sim_re))
+#   newobj <- TMB::MakeADFun(data = tmb_data, map = fit$tmb_map,
+#                            random = fit$tmb_random, parameters = params, DLL = "sdmTMB")
+#   s <- newobj$simulate()
+#   s$y_i
+# }
+#
+# nd <- matrix(NA, ncol = 20L, nrow = nrow(m$data))
+# for (i in seq_len(ncol(nd))) {
+#   nd[,i] <- mini_sim(m)
+# }
+#
+# s <- DHARMa::createDHARMa(
+#   simulatedResponse = nd,
+#   observedResponse = m$data$count,
+#   fittedPredictedResponse = m$family$linkinv(predict(m, newdata = NULL)$est),
+#   method = "PIT"
+# )
+#
+# plot(s)
 
-## TODO:add other residual checks
+# ### or this, but it took 2.5 hrs!
 # library(rstan)
 # library(tmbstan)
-# s <- tmbstan(m$tmb_obj, iter = 200, chains = 1, warmup = 199)
+#
+# stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
+# eta <- predict(m, tmbstan_model = stan_fit)
+# mu <- m$family$linkinv(as.numeric(eta))
+# r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
+# qqnorm(r_mcmc);qqline(r_mcmc) # MCMC
+# saveRDS(r_mcmc, "owls/mcmc_residuals.rds")
+#
+# r <- residuals(m)
+# qqnorm(r);qqline(r) # naive
+
+# s <- tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 199)
 # pstan <- predict(m, tmbstan_model = s)
 # resid <- as.numeric(pstan) - m$data$observed
 # # or do any PIT residuals randomization here...
@@ -282,4 +317,23 @@ ggplot(data = filter(p_proj, year > 1995)) +
[…]

### 2021-11-26 doc/paper.Rmd
Edit owl text

--- 2021-11-24 doc/paper.Rmd
@@
 
+```{r pkg-checks, include=FALSE, echo=FALSE}
+if (!require("rnaturalearthhires", quietly = TRUE)) {
+  stop("Please install 'rnaturalearthhires'.\n",
+  "`remotes::install_github('ropensci/rnaturalearthhires')`")
+}
+if (!require("ggsidekick", quietly = TRUE)) {
+  stop("Please install 'ggsidekick'.\n",
+  "`remotes::install_github('seananderson/ggsidekick')`")
+}
+```
+
 ```{r, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
@@
 library(ggplot2)
-theme_set(theme_minimal())
-# theme_set(ggsidekick::theme_sleek())
+# theme_set(theme_minimal())
+theme_set(ggsidekick::theme_sleek())
 library(dplyr)
@@
 )
-```
-
-```{r rnaturalearthhires, include=FALSE, echo=FALSE}
-if (!require("rnaturalearthhires", quietly = TRUE)) {
-  stop("Please install 'rnaturalearthhires'.\n",
-  "`remotes::install_github('ropensci/rnaturalearthhires')`")
-}
 ```
@@
 
+# Citizen science count data with a spatially varying effect
+
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
+The reasons for the interannual variation in the number of individuals migrating south are still not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
+The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
+Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
+We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial ('NB2') distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with NAO (**see appendix for more detail**):
+
+```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
+mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
+fit_owls <- sdmTMB(
+  count ~ 1 + nao + (1 | year_factor),
+  spatial_varying = ~ 0 + nao,     
+  family = nbinom2(link = "log"), data = snow, mesh = mesh, 
+  time = "year", spatial = "on", spatiotemporal = "IID"
+  )
+```
+
+We found a weak positive effect of annual mean NAO on overall counts, but also a southeast to northwest gradient in the intensity of this effect.
+This result would be consistent with owls closer to the Atlantic coast and those migrating the furthest south being the most effected by NAO (Fig. \@ref(fig:owl-nao)).
+
+(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020. 
+Circles represent all count locations and are their area is scaled to the mean number of owls observed per year (range: 0 to 8).
+
+```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="3.5in", fig.align='center'}
+owl <- here::here("figs", "owl_nao_effect.pdf")
+if (!file.exists(owl)) {
+  download.file("https://www.dropbox.com/s/fz1iw0a0diohyxc/owl_nao_effect.pdf?dl=1", 
+    destfile = owl)
+}
+knitr::include_graphics(owl)
+```
+
 <!--

### 2021-11-26 doc/paper.Rmd
Fig tweaks

@@
   coord_sf(datum = st_crs(utm_zone9)) +
+  theme(panel.grid.major = element_line(colour = "grey90")) +
   scale_x_continuous(
@@
   geom_point(alpha = 0.6, mapping = aes(size = density), colour = blues[4]) +
-  theme_minimal() + coord_fixed() +
+  coord_fixed() +
   labs(x = "UTMs E (km)", y = "UTMs N (km)") +
   scale_size_area(max_size = 10) +
+  theme(panel.grid.major = element_line(colour = "grey90")) +
   theme(legend.position = "none")
@@
   xlim(min(pcod$depth), 300) +
-  coord_fixed(expand = FALSE, ratio = 1.28)
+  coord_fixed(expand = FALSE, ratio = 1.28) +
+  theme(panel.grid.major = element_line(colour = "grey90"))
 ```
@@
     legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
-  )
-cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
+  ) +
+  theme(panel.grid.major = element_line(colour = "grey90"))
+cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto", align = "hv")
 ```
@@
 
-```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="3.5in", fig.align='center'}
+```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="3.9in", fig.align='center'}
 owl <- here::here("figs", "owl_nao_effect.pdf")

### 2021-11-26 doc/appendix-snow-sdmTMB.Rmd
text edits to owl appendix

@@
 library(sf)
-theme_set(theme_bw() + theme(
-  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
-  legend.position = "top", legend.justification = c(0, 0), 
-  axis.title = element_blank()))
 ```
@@
 This illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
-We will use counts of Snow Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts (CBCs) between 1979 and 2020. 
-This citizen science survey is organized by the by the National Audubon Society (data available by request from cbcadmin@audubon.org). 
+We will use counts of Snow Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts [CBCs, @cbc]. 
 Each survey is conducted between December 14 to January 5 in 24.14 kilometre diameter circle. 
 Circle locations are chosen by local volunteers and therefore reflect both locations of higher human densities and known bird watching locations.
-Each circle (given a unique "CBCID") is generally surveyed annually by at least 10 volunteers, or deemed inactive. 
+Each circle (given a unique ID) is generally surveyed annually by at least 10 volunteers, or deemed inactive. 
 While these locations are not random, the consistent sampling through time still make them useful for exploring spatiotemporal processes.
@@
 For these years, we have retrieved mean annual climate indices from the University of East Anglia's [Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
-Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution and the only portion well surveyed by CBCs.
+Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance.
+Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.
 
@@
 Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO.
+The dataframe also contains the Southern Oscillation Index (SOI), which is related to El Niño cycles, but we will only consider NAO in this example because any effect of SOI was even weaker.
 
@@
 ```{r plot-resids, fig.width = 8}
+theme_set(theme_bw() + theme(
+  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
+  legend.position = "top", legend.justification = c(0, 0), 
+  axis.title = element_blank()))
+
 ggplot(pred, aes(X, Y, colour = resid)) +
@@
 These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
-The colour axis has been transformed and trimmed to improve the visibility of variation in count estimates.
+The mean estimated count is `r round(mean(exp(p$est)), 2)`, which is essentially equal to the observed mean of `r round(mean(snow$count), 2)`.
+The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates.
 
@@
 
-<!-- confirm that the random effect of year is not included here? -->
-In this case these only include the overall effect of mean annual NAO:
+In this case these only include the overall effect of mean annual NAO and the random intercepts for year.
 

### 2021-11-26 doc/paper.Rmd
Tweak table

@@
 d[d == "1*"] <- "\\checkmark\\footnote{a}"
+d[d == "GAMs*"] <- "GAMs\\footnote{a}"
 d[d == "0"] <- "--"
@@
 d <- d[,c(1, 2, 3, 5, 6, 4, 7, 8, 9)]
-names(d)[names(d) == "mgcv"] <- "mgcv\\footnote{a}"
+# names(d)[names(d) == "mgcv"] <- "mgcv\\footnote{a}"
+d$Hmsc <- NULL
 
@@
   linesep = c(
-    rep('', 5), # coefs
+    rep('', 4), # coefs
     '\\addlinespace',
@@
     '\\addlinespace',
-    rep('', 11), # likelihood
+    rep('', 12), # likelihood
     '\\addlinespace',
@@
     number = c(
-      "SPDE approach as in Miller et al. (2019).",
-      "Technically possible but non-trivial.",
-      "Zero-inflated NB2 only.",
-      "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
-      "Technically possible but non-trivial.",
-      "Technically possible but non-trivial."
+      "Penalized smoother GAMs",
+      "SPDE approach as in Miller et al. (2019)",
+      "Technically possible but non-trivial",
+      "Zero-inflated NB2 only",
+      "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}",
+      "Technically possible but non-trivial",
+      "Technically possible but non-trivial"
     ),

### 2021-11-26 doc/paper.Rmd
Remove 3rd pcod eg

@@
 
-As a third example, we could model presence/absence of Pacific cod (`present`) with a time-varying quadratic effect of depth, an independent mean each year, a spatial random field, and IID spatiotemporal fields:
+<!-- As a third example, we could model presence/absence of Pacific cod (`present`) with a time-varying quadratic effect of depth, an independent mean each year, a spatial random field, and IID spatiotemporal fields: -->
 
@@
 
-```{r, eval=FALSE, echo=TRUE}
+```{r, eval=FALSE, echo=FALSE}
 fit_spatiotemporal_tv <- sdmTMB(

### 2021-11-27 doc/paper.Rmd
Add index to pcod fig

--- 2021-11-26 doc/paper.Rmd
@@
 ```{r pkg-checks, include=FALSE, echo=FALSE}
-if (!require("rnaturalearthhires", quietly = TRUE)) {
-  stop("Please install 'rnaturalearthhires'.\n",
-  "`remotes::install_github('ropensci/rnaturalearthhires')`")
-}
-if (!require("ggsidekick", quietly = TRUE)) {
-  stop("Please install 'ggsidekick'.\n",
-  "`remotes::install_github('seananderson/ggsidekick')`")
-}
+# if (!require("rnaturalearthhires", quietly = TRUE)) {
+#   stop("Please install 'rnaturalearthhires'.\n",
+#   "`remotes::install_github('ropensci/rnaturalearthhires')`")
+# }
+# if (!require("ggsidekick", quietly = TRUE)) {
+#   stop("Please install 'ggsidekick'.\n",
+#   "`remotes::install_github('seananderson/ggsidekick')`")
+# }
 ```
@@
 
-```{r, echo=TRUE}
+```{r, echo=TRUE, eval=FALSE}
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
@@
 
-```{r pcod-eg1-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE}
+```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
 fit <- sdmTMB(
@@
 ```
-
-<!-- tidy(fit, conf.int = TRUE) -->
-<!-- tidy(fit, effects = 'ran_pars', conf.int = TRUE) -->
 
@@
 
-```{r, echo=FALSE, eval=TRUE}
-p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
-```
+We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
 
-```{r, eval=FALSE}
-p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
-se <- apply(p3, 1, sd)
-p$se <- se
-g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
-  geom_raster() +
-  scale_fill_viridis_c(trans = "log10", option = "C") +
-  coord_fixed() +
-  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
-```
-
-```{r, warning=FALSE, message=FALSE}
-library(sf)
-# devtools::install_github("ropensci/rnaturalearthhires") # install if needed
-map_data <- rnaturalearth::ne_countries(
-  scale = "large",
-  returnclass = "sf", country = "canada")
-# Crop the polygon for plotting and efficiency:
-# st_bbox(map_data) # find the rough coordinates
-bc_coast <- suppressWarnings(suppressMessages(
-  st_crop(map_data,
-    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
-utm_zone9 <- 3156
-bc_coast_proj <- sf::st_transform(bc_coast, crs = utm_zone9)
-# ggplot(bc_coast_proj) + geom_sf()
-```
-
-```{r}
-g_pred <- ggplot(bc_coast_proj) +
-  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
-  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
-  coord_sf(datum = st_crs(utm_zone9)) +
-  theme(panel.grid.major = element_line(colour = "grey90")) +
-  scale_x_continuous(
-    limits = c(230957.7 + 105000, 1157991 - 570000),
-    labels = function(x) paste0(x / 1000)
-  ) +
-  scale_y_continuous(
-    limits = c(5366427 + 270000, 6353456 - 513000),
-    labels = function(x) paste0(x / 1000)
-  ) +
-  scale_fill_viridis_c(trans = "sqrt", option = "C") +
-  # scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
-  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(
-    # panel.background = element_rect(fill = "slategray1", colour= "white"),
-    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
+```{r fit-tv, eval=FALSE, echo=TRUE}
+fit_spatiotemporal <- sdmTMB(
+  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
+  time = "year", spatial = "off", spatiotemporal = "ar1"
 )
@@
 
-```{r, echo=FALSE}
-blues <- RColorBrewer::brewer.pal(5, "Blues")
-reds <- RColorBrewer::brewer.pal(5, "Reds")
-g_mesh <- ggplot(pcod, aes(X, Y)) +
-  inlabru::gg(mesh$mesh, edge.color = "grey60") +
-  geom_point(alpha = 0.6, mapping = aes(size = density), colour = blues[4]) +
-  coord_fixed() +
-  labs(x = "UTMs E (km)", y = "UTMs N (km)") +
-  scale_size_area(max_size = 10) +
-  theme(panel.grid.major = element_line(colour = "grey90")) +
-  theme(legend.position = "none")
-```
+We could generate a standardized index from that fit on a 2x2 km grid:
 
-```{r}
-nd <- data.frame(
-  depth =
-    seq(min(pcod$depth), max(pcod$depth), length.out = 300)
-)
-p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
-g_smooth <- ggplot(
-  p_smooth,
-  aes(depth, exp(est),
-    ymin = exp(est - 1.96 * est_se),
-    ymax = exp(est + 1.96 * est_se)
-  )
-) +
-  geom_ribbon(fill = blues[2]) +
-  geom_line(colour = blues[5], lwd = 1) +
-  labs(x = "Depth (m)", y = "Density (units)") +
-  xlim(min(pcod$depth), 300) +
-  coord_fixed(expand = FALSE, ratio = 1.28) +
-  theme(panel.grid.major = element_line(colour = "grey90"))
-```
-
-```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
-fit2 <- sdmTMB(
-  density ~ 0 + as.factor(year),
-  family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
-  time = "year", spatial = "on", spatiotemporal = "iid", silent = TRUE
-)
-p2 <- predict(fit2, newdata = qcs_grid, sims = 500L)
-ind <- get_index_sims(p2)
-```
-
-```{r plot-ind, eval=FALSE}
-g_ind <- ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) + 
-  geom_ribbon(fill = blues[2]) +
-  geom_line(colour = blues[5], lwd = 1) +
-  ylim(0, max(ind$upr)) +
-  coord_fixed(expand = FALSE, ratio = 0.3) +
-  labs(x = "Year", y = "Biomass (units)")
+```{r, eval=FALSE, echo=TRUE}
+p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
+ind <- get_index(p)
 ```
@@
 
-```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
-g_omega <- ggplot(bc_coast_proj) +
-  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
-  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
-  coord_sf(datum = st_crs(utm_zone9)) +
-  scale_x_continuous(
-    limits = c(230957.7 + 105000, 1157991 - 570000),
-    labels = function(x) paste0(x / 1000)
-  ) +
-  scale_y_continuous(
-    limits = c(5366427 + 270000, 6353456 - 513000),
-    labels = function(x) paste0(x / 1000)
-  ) +
-  # scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
-  scale_fill_gradient2(high = reds[4], low = blues[4]) +
-  labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
-  theme(
-    # panel.background = element_rect(fill = "slategray1", colour= "white"),
-    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
-  ) +
-  theme(panel.grid.major = element_line(colour = "grey90"))
-cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto", align = "hv")
+```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", out.width="5.2in", fig.align='center'}
+pcodfig <- here::here("figs", "pcod-fig.pdf")
+if (!file.exists(pcodfig)) {
+  download.file("https://www.dropbox.com/s/s8kl4u49p27qqhy/pcod-fig.pdf?dl=1", 
+    destfile = pcodfig)
+}
+knitr::include_graphics(pcodfig)
 ```
 
-We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
-Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
-
-```{r, eval=FALSE, echo=TRUE}
-fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
-  time = "year", spatial = "off", spatiotemporal = "ar1"
-)
-```
-
-We could generate a standardized index from that fit with:
-
-```{r, eval=FALSE, echo=TRUE}
-p <- predict(fit_spatiotemporal, newdata = qcs_grid, return_tmb_object = TRUE)
-ind <- get_index(p)
-```
-
-As a third example, we could model presence/absence of Pacific cod (`present`) with a time-varying quadratic effect of depth, an independent mean each year, a spatial random field, and IID spatiotemporal fields:
-
-```{r, eval=TRUE, echo=FALSE}
-# restore 'present' column
-pcod <- sdmTMB::pcod
-```
-
-```{r, eval=FALSE, echo=TRUE}
-fit_spatiotemporal_tv <- sdmTMB(
-  present ~ 0 + as.factor(year), 
-  time_varying = ~ 0 + poly(depth, 2), 
-  family = binomial(), data = pcod, mesh = mesh,
-  time = "year", spatial = "on", spatiotemporal = "iid"
-)
-```
+\clearpage
 
@@
 
-```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="3.9in", fig.align='center'}
+```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="4.1in", fig.align='center'}
 owl <- here::here("figs", "owl_nao_effect.pdf")

### 2021-11-27 analysis/pcod-fig.R

@@ -1,0 +1,190 @@
+library(sdmTMB)
+library(dplyr)
+library(ggplot2)
+library(sf)
+library(patchwork)
+
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
+
+fit <- sdmTMB(
+  density ~ s(depth),
+  data = pcod,
+  family = tweedie(link = "log"),
+  mesh = mesh,
+  spatial = "on", silent = FALSE
+)
+print(fit)
+p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
+
+# p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
+# se <- apply(p3, 1, sd)
+# p$se <- se
+# g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
+#   geom_raster() +
+#   scale_fill_viridis_c(trans = "log10", option = "C") +
+#   coord_fixed() +
+#   labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
+#   theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
+
+nd <- data.frame(
+  depth =
+    seq(min(pcod$depth), max(pcod$depth), length.out = 300)
+)
+p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
+
+fit_spatiotemporal <- sdmTMB(
+  density ~ s(depth),
+  family = tweedie(link = "log"), data = pcod, mesh = mesh,
+  time = "year", spatial = "off", spatiotemporal = "ar1", silent = FALSE
+)
+p_tv <- predict(fit_spatiotemporal,
+  newdata = qcs_grid, return_tmb_object = TRUE,
+  area = 4
+)
+ind <- get_index(p_tv)
+
+# just for plotting in UTMs
+pcod$X1000 <- pcod$X * 1000
+pcod$Y1000 <- pcod$Y * 1000
+mesh_1000 <- make_mesh(pcod, xy_cols = c("X1000", "Y1000"), cutoff = 10 * 1000)
+
+# plot -------------------
+
+lims_x <- c(230957.7 + 105000, 1157991 - 570000) + c(-10000, 10000)
+lims_y <- c(5366427 + 270000, 6353456 - 513000) + c(-10000, 10000)
+land <- "grey86"
+land_border <- "grey86"
+
+# devtools::install_github("ropensci/rnaturalearthhires") # install if needed
[…]

### 2021-11-27 doc/paper.Rmd
Edit

@@
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.
 
@@
 
-We start by creating a mesh object that contains matrices to apply the SPDE approach:
+We start by creating a mesh object that contains matrices to apply the SPDE approach.
 
@@
 Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
-We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)a).
+We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).
 
@@
 Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
-We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space):
+We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space).
 
@@
 The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)B--D.
 
@@
 We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
-Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):
+Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).
 
@@
 
-We could generate a standardized index from that fit on a 2x2 km grid:
+We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary TMB object (Fig.
+\@ref(fig:pcod-fig)E).
+`get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations.
 
@@
 Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
-The reasons for the interannual variation in the number of individuals migrating south are still not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
+The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
 Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial ('NB2') distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with NAO (**see appendix for more detail**):
+We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):
 
@@
 
-We found a weak positive effect of annual mean NAO on overall counts, but also a southeast to northwest gradient in the intensity of this effect.
-This result would be consistent with owls closer to the Atlantic coast and those migrating the furthest south being the most effected by NAO (Fig. \@ref(fig:owl-nao)).
+We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao)).
+This result would be consistent with owls closer to the Atlantic coast and those migrating the furthest south being the most effected by NAO.
 
@@
 Circles represent all count locations and are their area is scaled to the mean number of owls observed per year (range: 0 to 8). 
+The effect is multiplicative on owl count.
 

### 2021-11-27 analysis/pcod-fig.R

@@ -4,6 +4,17 @@ library(ggplot2)
 library(sf)
 library(patchwork)
 
+if (!require("rnaturalearthhires", quietly = TRUE)) {
+  stop("Please install 'rnaturalearthhires'.\n",
+  "`remotes::install_github('ropensci/rnaturalearthhires')`")
+}
+if (!require("ggsidekick", quietly = TRUE)) {
+  stop("Please install 'ggsidekick'.\n",
+  "`remotes::install_github('seananderson/ggsidekick')`")
+}
+
+theme_set(ggsidekick::theme_sleek())
+
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
 
 fit <- sdmTMB(
@@ -186,5 +197,7 @@ layout <- "
       12
       34
 "
-g <- wrap_plots(g_mesh, g_omega, g_pred, g3) + plot_layout(design = layout)
+g <- wrap_plots(g_mesh, g_omega, g_pred, g3) + plot_layout(design = layout) +
+  plot_annotation(tag_levels = c('A'))
+# g
 ggsave("figs/pcod-fig.pdf", width = 9, height = 7)

### 2021-11-27 doc/paper.Rmd
Edit

@@
 With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (note that this option may not be possible or as fast for all models).
-Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern (not shown).
+Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern substantially (not shown).
+<!-- Results with optimized math libraries (openBLAS and PARDISO (REFs)) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv. -->
+Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
 

### 2021-11-27 doc/paper.Rmd
Edit

@@
 (ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020. 
-Circles represent all count locations and are their area is scaled to the mean number of owls observed per year (range: 0 to 8). 
-The effect is multiplicative on owl count.
+Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). 
+The effect is multiplicative on owl count per NOA unit.
 

### 2021-11-27 doc/appendix-snow-sdmTMB.Rmd
change theme, rename projection...

@@
 library(sf)
+theme_set(ggsidekick::theme_sleek() + theme( # modifications for plotting sf data
+  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
+  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"), 
+  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
+  axis.title = element_blank())) # lat/lon is clear from axis text
 ```
@@
 Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
-This projection was achieved with the following crs code using `sf::st_transform(crs = proj)`.
+This projection was achieved with the following crs code using `sf::st_transform(crs = Albers)`.
 
 ```{r proj, echo=TRUE}
-proj <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 
+Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 
 +datum=NAD83 +units=m +no_defs"
@@
   scale_colour_viridis_c() +
-  theme_bw() 
+  ggsidekick::theme_sleek()
 ```
@@
 m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
+             spatial_varying = ~ 0 + nao,
              time = "year",
-             spatial_varying = ~ 0 + nao,
              spatial = "on", 
@@
 m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
+            spatial_varying = ~ 0 + nao,  
             time = "year",
-            spatial_varying = ~ 0 + nao,            
             spatial = "on", 
@@
 m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
+            spatial_varying = ~ 0 + nao,  
             time = "year",
-            spatial_varying = ~ 0 + nao,            
             spatial = "on", 
@@
 ```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
-m <- sdmTMB(count ~ 1 + nao + (1|year_f),
-            time = "year",
-            spatial_varying = ~ 0 + nao,            
-            spatial = "on", 
-            spatiotemporal = "IID",
+m <- sdmTMB(count ~ 1 + nao + (1|year_f), spatial_varying = ~ 0 + nao,
+            time = "year", spatial = "on", spatiotemporal = "IID",
             family = nbinom2(link = "log"),
@@
 Print the model to inspect the fit:
+```{r check, include = FALSE, cache=FALSE, eval=TRUE}
+m 
+```
 
-```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE}
+```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
 print(m)
@@
 
-We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate.
+\clearpage
+We can inspect randomized quantile residuals, although this method is not guaranteed to be accurate.
 
@@
 
+We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
+<!-- If our data was zero-inflated ... maybe make suggestions or even link to an example for how to do a delta model in sdmTMB?-->
+
+```{r zero-test, echo=TRUE, message=FALSE, cache=FALSE}
+s_nb2 <- simulate(m, nsim = 500) 
+pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
+r_nb2 <- DHARMa::createDHARMa(
+  simulatedResponse = s_nb2,
+  observedResponse = m$data$count,
+  fittedPredictedResponse = pred_fixed
+)
+DHARMa::testZeroInflation(r_nb2, plot = FALSE)
+```
+
 <!-- \clearpage -->
+Check the spatiotemporal distribution of the residuals.
 
-```{r plot-resids, fig.width = 8}
-theme_set(theme_bw() + theme(
-  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
-  legend.position = "top", legend.justification = c(0, 0), 
-  axis.title = element_blank()))
-
-ggplot(pred, aes(X, Y, colour = resid)) +
-  geom_point(size = 0.5) +
-  facet_wrap(~year) +
-  coord_fixed() +
-  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
-  scale_colour_gradient2(guide = guide_colourbar(title.position = "top"))
+```{r plot-resids, fig.width = 8, cache=FALSE}
+ggplot(pred, aes(X, Y, colour = resid)) + geom_point(size = 0.5) +
+  facet_wrap(~year) + coord_fixed() +
+  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) + 
+  theme(axis.text = element_blank(), axis.ticks = element_blank())
 ```
@@
 
-In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will appear in latitude and longitude.
+In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.
 
 ```{r proj-p, echo=TRUE, eval=T}
-library(sf)
-
-p <- pred %>% mutate(
-  X = X * 100000,
-  Y = Y * 100000
-)
-
-p_proj <- p %>% mutate(x = X, y = Y) %>% st_as_sf(., coords = c("x", "y"), crs = proj)
+p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
+p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
 ```
 
-One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transform them to the correct projection using `sf::st_transform(crs = proj)`.
+We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transformed them to the correct projection.
 
 ```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
-library("rnaturalearth")
-library("rnaturalearthdata")
-coast <- ne_coastline(scale = "medium", returnclass = "sf") %>% 
-  sf::st_transform(crs = proj)
+coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(crs = Albers) 
 lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
-lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
+lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
 ```
@@
 
-```{r plot-map, echo=TRUE, eval=T}
+```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
 ggplot(data = p_proj) + 
@@
      aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
-  geom_sf(data = coast, fill = NA) +
-  geom_sf(data = lakes, fill = NA) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
   coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
@@
   labs(colour = "NAO effect\non Snowy Owl count") +
-  theme_bw() + theme(
-    legend.key.height = unit(0.3, "cm"),
-    legend.position = c(0.25,0.13),
+  theme(legend.key.height = unit(0.3, "cm"),
+        legend.position = c(0.1,0.1),
         axis.title = element_blank())
@@
 
-```{r plot-all-effects, echo=TRUE, eval=T}
+```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
 ggplot(data = filter(p_proj, year > 1995)) + 
@@
 
-```{r plot-fixed-effects, echo=TRUE, eval=T}
+```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
 ggplot(data = filter(p_proj, year > 1995)) + 
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=T}
+```{r plot-spatial-effects, echo=TRUE, eval=T, cache=FALSE}
 ggplot(data = p_proj) + 
   geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
-  geom_sf(data = coast, fill = NA) +
-  geom_sf(data = lakes, fill = NA) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
   coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
@@
     direction = "horizontal", title.position = "top")) +
-  theme_bw() + theme(
-    legend.key.height = unit(0.3, "cm"),
-    legend.position = c(0.25,0.13),
+  theme(legend.key.height = unit(0.3, "cm"),
+        legend.position = c(0.1,0.1),
         axis.title = element_blank())
@@
 
-```{r plot-spatiotemporal-effects, echo=TRUE, eval=T}
+```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
 ggplot(data = filter(p_proj, year > 1995)) + 

### 2021-11-27 doc/paper.Rmd

@@
 
-# Citizen science count data with a spatially varying effect
+# Spatially varying coefficients applied to citizen-science owl counts
 
@@
   family = nbinom2(link = "log"), data = snow, mesh = mesh, 
-  time = "year", spatial = "on", spatiotemporal = "IID"
+  time = "year", spatial = "on", spatiotemporal = "iid"
 )

### 2021-11-27 snow.R
general owl code tweaks

@@ -109,29 +109,20 @@ sdat$residuals1 <- residuals(m1)
 qqnorm(sdat$residuals1);abline(a = 0, b = 1)
 
 # ## TODO:add other residual checks
-# mini_sim <- function(fit) {
-#   params <- fit$tmb_obj$env$parList()
-#   tmb_data <- fit$tmb_data
-#   # tmb_data$sim_re <- rep(1, length(tmb_data$sim_re))
-#   newobj <- TMB::MakeADFun(data = tmb_data, map = fit$tmb_map,
-#                            random = fit$tmb_random, parameters = params, DLL = "sdmTMB")
-#   s <- newobj$simulate()
-#   s$y_i
-# }
-#
-# nd <- matrix(NA, ncol = 20L, nrow = nrow(m$data))
-# for (i in seq_len(ncol(nd))) {
-#   nd[,i] <- mini_sim(m)
-# }
-#
-# s <- DHARMa::createDHARMa(
-#   simulatedResponse = nd,
-#   observedResponse = m$data$count,
-#   fittedPredictedResponse = m$family$linkinv(predict(m, newdata = NULL)$est),
-#   method = "PIT"
-# )
-#
-# plot(s)
+
+# m <- readRDS("owls/snow_w_main_effect_0_150km.rds") # poisson for comparison
+
+s_nb2 <- simulate(m, nsim = 500)
+
+pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
+r_nb2 <- DHARMa::createDHARMa(
+  simulatedResponse = s_nb2,
+  observedResponse = m$data$count,
+  fittedPredictedResponse = pred_fixed
+)
+plot(r_nb2)
+
+DHARMa::testZeroInflation(r_nb2)
 
 # ### or this, but it took 2.5 hrs!
 # library(rstan)
@@ -147,11 +138,6 @@ qqnorm(sdat$residuals1);abline(a = 0, b = 1)
 # r <- residuals(m)
 # qqnorm(r);qqline(r) # naive
 
-# s <- tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 199)
-# pstan <- predict(m, tmbstan_model = s)
-# resid <- as.numeric(pstan) - m$data$observed
-# # or do any PIT residuals randomization here...
-# qqnorm(resid);qqline(resid) # should be right
 
 # # try model with SOI for comparison -- a little worse
[…]

### 2021-11-27 doc/paper.Rmd
Major re-arrange

@@
 
-## Model selection
-
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`aic()`, @akaike1974]. 
-Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
-The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
-
-## Model evaluation
-
-Validation of mixed-effects models is non-trivial (that REF). We provide three approaches to assist with model validation:
-
-1. The `residuals()` method returns randomized quantile residuals (e.g., as in DHARMa REF).
-These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects.
-These residuals have known statistical issues (REF) but are quick to calculate.
-One-step-ahead residuals (REF) or residuals from a single MCMC draw (REF) are also possible, but both approaches are much slower.
-
-2.
-The `sdmTMB_simulate()` function can simulate new data, or simulate from a fitted model, to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
-
-3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation (see `?extract_mcmc`). 
-We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).
-
-## Parameter priors/penalties
-
-sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
-
-## Derived quantities of interest
-
-The results from an sdmTMB model can be used to generate various quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
-
 <!-- ## Advanced options -->
@@
 
-We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary TMB object (Fig.
-\@ref(fig:pcod-fig)E).
-`get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations.
-
-```{r, eval=FALSE, echo=TRUE}
-p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
-ind <- get_index(p)
-```
-
 (ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
@@
 
-(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020. 
+(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
 Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). 
@@
 knitr::include_graphics(owl)
+```
+
+## Model selection and validation
+
+Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]. 
+Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
+The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
+The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
+
+Validation of state-space models is non-trivial, particularly those using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation:
+
+1. The cross validation function `sdmTMB_cv()` can be used to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
+
+2. The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
+These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.
+These residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
+One-step-ahead residuals [@thygesen2017a] or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are much slower.
+
+3.
+The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
+Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE).
+
+4.
+An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017] to evaluate the accuracy of the Laplace approximation or evaluate residuals (see `?extract_mcmc`).
+<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->
+
+## Parameter priors/penalties
+
+sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
+
+## Derived quantities of interest
+
+The results from an sdmTMB model can be used to generate various quantities of interest.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`; Fig.
+\@ref(fig:pcod-fig)E).
+Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
+
+<!-- We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary TMB object (Fig.
+\@ref(fig:pcod-fig)E).
+`get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations.
+-->
+
+```{r, eval=FALSE, echo=FALSE}
+p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
+ind <- get_index(p)
 ```
@@
 mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
-inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library.
-All platforms were restricted to one core and could be faster with parallel computation.
+<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
+<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
 <!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
@@
 The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
-Interface-wise, VAST is a less modular interface (e.g., by default automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
+Interface-wise, VAST is a less modular interface (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
 For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).

### 2021-11-27 doc/paper.Rmd
Edit

@@
 An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
+It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
 Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
@@
 
+<!-- Future directions para? -->
+<!-- multivariate, but only if can keep interface simple -->
+<!-- optimized Bayes sampling -->
+<!-- residual diagnostics -->
+<!-- zero inflation -->
+<!-- MVT? time-varying epsilon -->
+
 Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management (REF).

### 2021-11-27 doc/paper.Rmd
Sections; lme4
Add missing E from Fig 2

@@
 This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].
+<!-- I'm finding the above a bit awkward.
+What about: Ideally, this residual correlation should be accounted for because doing so...
+OR Accounting for this residual correlation is important because doing so...
+-->
 
@@
 Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
+<!-- the e.g. here isn't intuitive for me... usually you only get one count from one plot so what is being grouped? Maybe heights of trees on different plots of land? -->
 A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
 
-While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
+While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
 Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
@@
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to spatial models that use the SPDE, but either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some of the flexibility for the particular class of models discussed here.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or some of the flexibility for the particular class of models discussed here.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
@@
 Here, we introduce our R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
-Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
+Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv, for a specific class of spatial and spatiotemporal models [@wood2017a].
 One common application is for species distribution models (SDMs), hence the package name.
@@
 (ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
-(a) SPDE mesh (lines) combined with the trawl survey observations (points).
+(A) SPDE mesh (lines) combined with the trawl survey observations (points).
 Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
 The circle area corresponds to density of Pacific Cod for individual trawls.
-(b) Conditional effect of depth modelled as a penalized smoother.
 The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
 These predictions are made omitting the spatial random field.
-(c) Spatial random field:
+(B) Spatial random field:
 these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
-(d) Overall prediction: these estimates represent the combination of all fixed and random effects.
+(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
+(D) Conditional effect of depth modelled as a penalized smoother.
+(E) Standardized area-weighted population index derived from a spatiotemporal version of this model.
 
@@
 Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):
+We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):
 
@@
 
-## Model selection and validation
+# Model selection and validation
 
@@
 
-## Parameter priors/penalties
+# Parameter priors/penalties
 
@@
 
-## Derived quantities of interest
+# Derived quantities of interest
 
@@
 The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
-Interface-wise, VAST is a less modular interface (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
+Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
 For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).

### 2021-11-27 doc/paper.Rmd
minor edits/comments on intro

@@
 (ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
-(A) SPDE mesh (lines) combined with the trawl survey observations (points).
+(a) SPDE mesh (lines) combined with the trawl survey observations (points).
 Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
 The circle area corresponds to density of Pacific Cod for individual trawls.
+(b) Conditional effect of depth modelled as a penalized smoother.
 The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
 These predictions are made omitting the spatial random field.
-(B) Spatial random field:
+(c) Spatial random field:
 these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
-(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
-(D) Conditional effect of depth modelled as a penalized smoother.
-(E) Standardized area-weighted population index derived from a spatiotemporal version of this model.
+(d) Overall prediction: these estimates represent the combination of all fixed and random effects.
 
@@
 Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):
+We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):
 
@@
 The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
-Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
+Interface-wise, VAST is a less modular interface (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
 For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).

### 2021-11-27 doc/appendix-snow-sdmTMB.Rmd
consolidate pages some

@@
 
-```{r mesh, echo=TRUE, cache=FALSE}
+```{r mesh, echo=TRUE, cache=FALSE, out.width="5.5in"}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
+plot(mesh)
 ```
 
-```{r plot-mesh, out.width="6in"}
-plot(mesh)
+```{r plot-mesh}
+
 ```
@@
 
-```{r explore, fig.asp = 0.3}
-ggplot(snow, aes(year, count, colour = nao)) + 
-  geom_jitter() + 
-  scale_colour_viridis_c() +
-  ggsidekick::theme_sleek()
+```{r explore, fig.asp = 0.3, cache=FALSE}
+ggplot(snow, aes(year, count, colour = nao)) + geom_jitter() + 
+  scale_colour_viridis_c() + ggsidekick::theme_sleek()
 ```
@@
 ```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
-m <- sdmTMB(count ~ 1 + nao + (1|year_f), spatial_varying = ~ 0 + nao,
+m <- sdmTMB(count ~ 1 + nao + (1|year_f), 
+            spatial_varying = ~ 0 + nao,
             time = "year", spatial = "on", spatiotemporal = "IID",
@@
 
-\clearpage
+<!-- \clearpage -->
 We can inspect randomized quantile residuals, although this method is not guaranteed to be accurate.
@@
 
-More accurate residuals can be generated using STAN, but this is a very slow process!
+More accurate residuals can be generated using tmbstan, but this is a very slow process!
 In this case, it took more than 2.5 hrs.
@@
 ```{r stan-resids, eval=FALSE}
-library(tmbstan)
 stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
@@
 
-In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.
+To produce pretty maps of these predictions, we back transformed the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.
 
@@
 
-```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
+```{r shapes, cache=FALSE}
 coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
@@
 
-```{r p-mean, message=FALSE, echo=TRUE, eval=T}
+```{r p-mean, message=FALSE, echo=TRUE, eval=T, cache=FALSE}
 b <- tidy(m)
-
 p_mean <- p %>% group_by(CBCID) %>% summarise(
@@
 ```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
-ggplot(data = p_proj) + 
-  geom_point(data = p_mean, 
+ggplot(data = p_proj) + geom_point(data = p_mean, 
      aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
@@
     direction = "horizontal", title.position = "top")) +
-  guides(size = "none") + 
-  labs(colour = "NAO effect\non Snowy Owl count") +
-  theme(legend.key.height = unit(0.3, "cm"),
-        legend.position = c(0.1,0.1),
-        axis.title = element_blank())
+  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
+  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
 ```
@@
 ```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
-ggplot(data = filter(p_proj, year > 1995)) + 
-  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
+ggplot(data = filter(p_proj, year > 1978)) + 
+  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est)), 
              size = 0.4, alpha = 0.5) +
@@
 
-Note that for all of these annually varying predictions, we are showing only the most recent 25 years.
+<!-- Note that for all of these annually varying predictions, we are showing only the most recent 25 years.  -->
 
@@
 
-### 1. Just fixed effects
-
-In this case these only include the overall effect of mean annual NAO and the random intercepts for year.
-
-```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
-ggplot(data = filter(p_proj, year > 1995)) + 
-  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
-             size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
-           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
-    guide = guide_colourbar(title.position = "top")) +
-  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
-  facet_wrap(~year_f)
-```
-
-\clearpage
-
-### 2. Static spatial random effects
+### 1. Static spatial random effects
 
@@
     direction = "horizontal", title.position = "top")) +
-  theme(legend.key.height = unit(0.3, "cm"),
-        legend.position = c(0.1,0.1),
-        axis.title = element_blank())
+  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
 ```
@@
 
-### 3. Spatiotemporal random effects
+### 2. Spatiotemporal random effects
 
@@
 ```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
-ggplot(data = filter(p_proj, year > 1995)) + 
-  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
+ggplot(data = filter(p_proj, year > 1978)) + 
+  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st), 
              size = 0.4, alpha = 0.5) +
@@
 
+<!-- Note that for these annually varying predictions, we are showing only the most recent 25 years.  -->
+
 \clearpage
 
+### 3. Just fixed effects
+
+In this case these only include the overall effect of mean annual NAO and the random intercepts for year.
+
+```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
+ggplot(data = filter(p_proj, year > 1978)) + 
+  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)), 
+             size = 0.4, alpha = 0.5) +
+  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
+           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
+  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
+    guide = guide_colourbar(title.position = "top")) +
+  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
+  facet_wrap(~year_f)
+```
+
+\clearpage
 # References

### 2021-11-27 doc/appendix-snow-sdmTMB.Rmd
minor edits to owls

@@
 
-Snow Owls are an arctic breeding species is an irruptive migrant meaning that the number of individuals choosing to migrate varies dramatically from year to year. 
-The reasons for this variation are still not well understood, but seem to be at least partially related to very high abundances of food during the breeding season and therefore higher productivity [@robillard2016].
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
+The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
@@
 
-```{r aic, echo=TRUE, cache=TRUE}
+```{r aic, echo=TRUE, cache=FALSE}
 AIC(m0,m1,m2)
@@
 ```{r check, include = FALSE, cache=FALSE, eval=TRUE}
-m 
+m # loading this twice just prevents some excess messages from cluttering the output when printing below
 ```
@@
 
-```{r proj-p, echo=TRUE, eval=T}
+```{r proj-p}
 p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
@@
 
-```{r shapes, cache=FALSE}
+```{r shapes}
 coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
@@
 
-```{r p-mean, message=FALSE, echo=TRUE, eval=T, cache=FALSE}
+```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=FALSE}
 b <- tidy(m)
+
 p_mean <- p %>% group_by(CBCID) %>% summarise(
@@
 
-```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
+```{r plot-map, echo=TRUE, eval=T, cache=T}
 ggplot(data = p_proj) + geom_point(data = p_mean, 
@@
 
-```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
+```{r plot-all-effects, echo=TRUE, eval=T, cache=T}
 ggplot(data = filter(p_proj, year > 1978)) + 
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=T, cache=FALSE}
+```{r plot-spatial-effects, echo=TRUE, eval=T, cache=T}
 ggplot(data = p_proj) + 
@@
 
-```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
+```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=T}
 ggplot(data = filter(p_proj, year > 1978)) + 
@@
 
-```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
+```{r plot-fixed-effects, echo=TRUE, eval=T, cache=T}
 ggplot(data = filter(p_proj, year > 1978)) + 

### 2021-11-27 SNOWdataprep.R

@@ -9,7 +9,7 @@ library(tidyverse)
 
 #### get owl data ####
 
-d <- read_excel("owls/1-121-CBC_Count_History_Report.xlsx") %>%
+d <- read_excel("owls/PE-CBC_data/1-121-CBC_Count_History_Report.xlsx") %>%
   select(CBCID = Abbrev, Latitude, Longitude, surveyID = Count_yr, TotalSpecies) %>%
   filter(surveyID > 79) %>%
   # filter(TotalSpecies == TRUE) %>% # remove survey years not conducted?
@@ -20,7 +20,8 @@ a <- read_excel("owls/PE-CBC_data/PE-CBC_Circle_Species_Report_SQL_updated.xlsx"
   select(CBCID = Abbrev, Species = COM_NAME, Latitude, Longitude, surveyID = Count_yr,
          count = how_many, SppCount = TotalSpecies) %>%
   mutate(year = ifelse(surveyID <= 100, paste0(19, surveyID-1),
-                       ifelse(surveyID <= 110, paste0(200, surveyID-101), paste0(20, surveyID-101)))) %>%
+                       ifelse(surveyID <= 110, paste0(200, surveyID-101), paste0(20, surveyID-101))),
+         year_f = as.factor(year)) %>%
   filter(Longitude > -125 & Longitude < -60 & Latitude > 32 & Latitude < 55)
 
 
@@ -63,7 +64,7 @@ s <- bind_cols(s, sxy)
 
 # select relevant variables and scale X and Y for mesh
 # use unit of 100 kms
-sdat <- s %>% select(X, Y, count, year, CBCID, Latitude, Longitude,TotalSpecies, SppCount) %>%
+sdat <- s %>% select(X, Y, count, year, year_f, CBCID, Latitude, Longitude, TotalSpecies, SppCount) %>%
   mutate(year = as.integer(year),
          year_f = as.factor(year),
          X = X / 100000,

### 2021-11-27 doc/paper.Rmd
Edits

--- 2021-11-27 doc/paper.Rmd
@@
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].
-<!-- I'm finding the above a bit awkward.
-What about: Ideally, this residual correlation should be accounted for because doing so...
-OR Accounting for this residual correlation is important because doing so...
--->
+Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].
 
 Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
-Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-<!-- the e.g. here isn't intuitive for me... usually you only get one count from one plot so what is being grouped? Maybe heights of trees on different plots of land? -->
+Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
 A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].

### 2021-11-29 doc/paper.Rmd
Eric edits

@@
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].
+Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
 
@@
 
-sdmTMB models are spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+The GLMMs underpinning sdmTMB models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
 In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
 The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) (REF).
+The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim$ 0.13 (REF)---can be shared or estimated separately.
@@
 
-We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
+<!--
+EW: TODO I think this could use a bit more hand holding — things to highlight are (1) predictions can be made for a new dataframe with any locations, 
+(2) more detail about why you’re predicting on a grid. 
+(3) other uses for predict — e.g. might be helpful to understand marginal effects of covariates for complicated models
+-->
+
+We can make predictions with the `predict()` method and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
+<!-- Here, we will predict on a 2x2 km grid that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
+-->
 The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
@@
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`; Fig.
-\@ref(fig:pcod-fig)E).
+\@ref(fig:pcod-fig)E) given a user supplied grid.
 Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
@@
 <!-- Take home, re-iterate paragraph? -->
+
+<!-- EW: TODO Not sure if Heaton included CAR models, but we might want to talk about them briefly because there’s a lot of packages that include that functionality - including NIMBLE, etc.  -->
 

### 2021-11-30 analysis/timing.R
Fix spaMM environment issue

@@ -78,6 +78,7 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
   cat("n_obs:", n_obs, "\n")
   cat("iter:", iter, "\n")
   cat("simulating...\n")
+
   s <- simulate_dat(n_obs = n_obs, cutoff = cutoff, seed = seed,
     iter = iter, family = family, max.edge = max.edge, tweedie_p = tweedie_p, phi = phi,
     sigma_O = sigma_O)
@@ -215,32 +216,20 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
     max.edge = c(max.edge, 0.2),
     offset = c(0.1, 0.05)
   )
-  spde2 <- INLA::inla.spde2.pcmatern(
+  # note <<- below:
+  # Avoids: `Hmmm... it looks like you should put some object(s) in control.HLfit$formula_env`
+  spde_spaMM <<- INLA::inla.spde2.pcmatern(
     mesh = mesh2,
     prior.range = c(0.05, 0.05),
     prior.sigma = c(2, 0.05)
   )
-
   out <- system.time({
     fit_spaMM <- tryCatch({
-      # browser()
-      fitme(observed~a1 + IMRF(1|X+Y, model = spde2),
+      fitme(observed~a1 + IMRF(1|X+Y, model = spde_spaMM),
             family = family, data = sim_dat)
     }, error = function(e) NA)
   })
-  # Error in eval(parse(text = paste("list(", pars, ")")), envir = env) :
-  #   object 'spde2' not found
-  # Error in eval(parse(text = paste("list(", pars, ")")), envir = env) :
-  #   object 'spde' not found
-  # Error in .process_IMRF_bar(term, env = env) :
-  #   Hmmm... it looks like you should put some object(s) in control.HLfit$formula_env.
-  # if (!is.null(fit_spaMM$error)) {
-  #   spaMM_error <- TRUE
-  # } else {
-  #   spaMM_error <- FALSE
-  # }
   times$spaMM <- if (all(is.na(fit_spaMM))) NA else out[["elapsed"]]
-
 
   cat("mgcv disc.\n")
     loc.bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)

### 2021-11-30 analysis/sim-test-families.R
Build sim testing

@@ -61,6 +61,7 @@ fake_NAs <- function(iter, family, link) {
 est <- purrr::map_dfr(families, function(.fam) {
   cat(.fam$family, "\n")
   furrr::future_map_dfr(seq_len(8 * 16), function(i) {
+  # furrr::future_map_dfr(seq_len(1), function(i) {
   # purrr::map_dfr(seq_len(1), function(i) {
     if (.fam$family == "Beta") {
       phi <- phi * 5 # small phi's can cause ~0.0 and ~1.0
@@ -90,7 +91,7 @@ est <- purrr::map_dfr(families, function(.fam) {
 
     if (max(m$gradients) > 0.001) {
       m <- tryCatch({run_extra_optimization(m,
-        nlminb_loops = 0, newton_steps = 1)},
+        nlminb_loops = 0, newton_loops = 1)},
         error = function(e) return(NULL))
     }
     if (is.null(m)) return(fake_NAs(i, .fam$family, .fam$link))
@@ -171,7 +172,7 @@ worm_plot <- function(dat, lwr, est, upr, true, title) {
     geom_point(size = 1.5) +
     geom_linerange(size = 0.3) +
     scale_shape_manual(values = c("TRUE" = 19, "FALSE" = 21)) +
-    guides(shape = FALSE) +
+    guides(shape = "none") +
     ggtitle(title) +
     labs(x = "Parameter value") +
     theme(axis.title.y = element_blank(), axis.text.y = element_blank(),

### 2021-11-30 analysis/sim-test-index.R

@@ -9,42 +9,52 @@ plan(multisession, workers = floor(availableCores() / 2))
 
 SEED <- 1
 set.seed(SEED)
-x <- seq(-1, 1, length.out = 25)
-y <- seq(-1, 1, length.out = 25)
-loc <- expand.grid(x = x, y = y)
-x <- loc$x
-y <- loc$y
-N <- length(x)
+N <- 625 # per year
 time_steps <- 20
 sigma_O <- 0.6
 sigma_E <- 0.3
-phi <- 1.2
-tweedie_p <- 1.7 # Tweedie p
+phi <- 6
+tweedie_p <- 1.6 # Tweedie p
 .range <- 1
 
+x <- seq(-1, 1, length.out = 25)
+y <- seq(-1, 1, length.out = 25)
+loc <- expand.grid(x = x, y = y)
+x <- rep(loc$x, time_steps)
+y <- rep(loc$y, time_steps)
 loc <- data.frame(x = x, y = y)
 spde <- make_mesh(loc, xy_cols = c("x", "y"), cutoff = 0.2)
 plot(spde)
 
 sim_test_index <- function(i) {
   betas <- rnorm(time_steps, 1, 1)
-  dat <- data.frame(year = as.factor(rep(seq_len(time_steps), each = N)))
-  s <- sdmTMB_sim(
-    x = x, y = y, mesh = spde, X = model.matrix(~ 0 + year, dat),
-    betas = betas, time_steps = time_steps,
-    phi = phi, range = .range, sigma_O = sigma_O, sigma_E = sigma_E,
-    seed = SEED * i, family = tweedie(), tweedie_p = tweedie_p
+  dat <- data.frame(year = rep(seq_len(time_steps), each = N))
+  s <- sdmTMB_simulate(
+    formula = ~ 0 + as.factor(year),
+    data = dat,
+    mesh = spde,
+    B = betas,
+    phi = phi,
+    range = .range,
+    sigma_O = sigma_O,
+    sigma_E = sigma_E,
+    seed = SEED * i,
+    time = "year",
+    family = tweedie(),
+    tweedie_p = tweedie_p
   )
   s <- mutate(s, year = dat$year)
+  s <- mutate(s, x = loc$x)
+  s <- mutate(s, y = loc$y)
   s_sampled <- s %>% group_by(year) %>% slice_sample(n = 25L)
 
   if (FALSE) {
     ggplot(s, aes(x, y, fill = mu)) +
[…]

### 2021-11-30 doc/appendix-model.Rmd
Start model appendix

@@
 
-### Matern parameterization
+\newcommand{\s}{\boldsymbol{s}}
 
-<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
-  The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
-  (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
-  <!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
-    where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
-  a scaling parameter that is estimated (REF).
-  The parameter $\nu$ controls the smoothness of the covariance function.
-  We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+## Notation conventions
+
+This appendix uses the following notation conventions:
+
+* Bold lowercase for vectors
+
+* Bold subscript $\s$ since x, y
+
+* Bold uppercase for matrices
+
+* $\phi$ for all 'dispersion' parameters for consistency with code
+
+* Family titles link to TMB or sdmTMB source code
+
+* Attempt to link math symbols to argument options in model and code symbols
+
+```{r, results='asis'}
+symbols <- tibble::tribble(
+  ~Symbol, ~Description,
+  "$\\phi$",   "A dispersion parameter for a distribution",
+  "$\\mu$",   "Mean",
+  "$f$",   "Link function"
+)
+knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex",  booktabs = TRUE)
+```
+
+```{r}
+subscripts <- tibble::tribble(
+  ~Symbol, ~Description,
+  "$\\s$",   "Index for space; a vector of x and y coordinates",
+  "$t$",   "Index for time",
+  "$g$",   "Group"
+)
+knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex",  booktabs = TRUE)
+```

### 2021-11-30 doc/appendix-model.Rmd
Start model appendix

--- 2021-11-19 doc/appendix-model.Rmd
@@
 
-### Matern parameterization
+```{r setup, echo=FALSE, cache=FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>",
+  echo = FALSE,
+  error = FALSE,
+  message = FALSE,
+  warning = FALSE
+)
+```
+
+\newcommand{\s}{\boldsymbol{s}}
+
+# Notation conventions
+
+This appendix uses the following notation conventions:
+
+* Bold lowercase for vectors
+
+* Bold subscript $\s$ since x, y
+
+* Bold uppercase for matrices
+
+* $\phi$ for all 'dispersion' parameters for consistency with code
+
+* Family titles link to TMB or sdmTMB source code
+
+* Attempt to link math symbols to argument options in model and code symbols
+
+```{r, results='asis'}
+symbols <- tibble::tribble(
+  ~Symbol, ~Description,
+  "$\\phi$", "A dispersion parameter for a distribution",
+  "$\\mu$", "Mean",
+  "$y$", "Observed response data",
+  "$f$", "Link function",
+  "$\\omega_{\\s}$", "Spatial random field at point $\\s$",
+  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$",
+  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field",
+  "$\\boldsymbol{\\Sigma}_\\omega$", "Spatial random field covariance matrix",
+  "$\\boldsymbol{\\Sigma}_\\epsilon$", "Spatiotemporal random field covariance matrix",
+  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
+  "$\\boldsymbol{\\beta}$", "Parameter vector",
+  "$\\boldsymbol{X}$", "Model predictor matrix",
+  "$\\alpha_g$", "IID random intercept deviations",
+  "$\\sigma_\\alpha^2$", "IID random intercept variance",
+  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations",
+  "$\\rho$", "Correlation between random fields in subsequent time steps"
+
+)
+knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex", booktabs = TRUE)
+```
+
+```{r}
+subscripts <- tibble::tribble(
+  ~Symbol, ~Description,
+  "$\\s$", "Index for space; a vector of x and y coordinates",
+  "$t$", "Index for time",
+  "$g$", "Group"
+)
+knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex", booktabs = TRUE)
+```
+
+# Basic model structure
+
+$$
+\begin{aligned}
+  \mathbb{E}(y_{\s,t}) &= \mu_{\s,t},\\
+  \mu_{\s,t} &= f^{-1}
+  \left( \boldsymbol{X} \boldsymbol{\beta} + \omega_s + \epsilon_{\s,t} \right),\\
+  \boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
+  \boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
+\end{aligned}
+$$
+where $f$ is a link function (e.g., log or logit) and $f^{-1}$ is its inverse link.
+
+* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
+  represents the fixed effect model matrix and a corresponding vector of
+  coefficients
+
+* $\omega_s$ are included if `spatial = 'on'`
+
+* $\epsilon_{\s,t}$ are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`IID`; default), AR1 (`AR1`), random walk (`RW`), or off (`'off`) via the `spatiotemporal` argument
+
+# Random or multi-level intercepts
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
+  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
+\end{aligned}
+$$
+
+where $\alpha_g$ is an example optional "random" intercept---an intercept
+centered on zero that varies by level $g$ and is constrained by
+$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
+syntax as in lme4 or glmmTMB.
+
+# Time-varying regression parameters
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} x_{\s,t} + \ldots \right),\\
+  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
+  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
+\end{aligned}
+$$
+
+where $\gamma_t$ is an example optional time-varying regression parameter and
+$x_{\s,t}$ is the corresponding predictor value. This is defined via the
+`time_varying` argument. E.g., `~ 0 + x`. The first value is given an implicit
+uniform prior. I.e., the same variable should not appear in the fixed effect
+formula since the initial value is estimated as part of the time-varying formula.
+
+# Spatial regression parameters
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \zeta_s x_t + \ldots \right),\\
+  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
+\end{aligned}
+$$
+
+where $\zeta_s$ is a regression parameter that varies spatially and $x_t$ is a predictor that is constant for a given time $t$.
+Currently, $x_t$ must be the time variable itself.
+If you are interested in using a different predictor, contact the developers.
+
+# AR1 spatiotemporal random fields
+
+Dropping the optional $\omega_s$ for simplicity:
+
+$$
+\begin{aligned}
+\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
+\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
+\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
+\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\end{aligned}
+$$
+
+where $\rho$ is the correlation between subsequent spatiotemporal random
+fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
+
+# Random walk spatiotemporal random fields
+
+Dropping the optional $\omega_s$ for simplicity:
+
+$$
+\begin{aligned}
+\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
+\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
+\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_t,  \:
+\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\end{aligned}
+$$
+
+# Offset terms
+
+Offset terms can be included with the reserved word `offset` in the formula.
+E.g., `y ~ x + offset`.
+
+These are included in the linear predictor as
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + O_{\s,t} + \ldots \right),
+\end{aligned}
+$$
+where $O_{\s,t}$ is an offset term---a **log transformed** variable without
+a coefficient (assuming a log link).
+
+# Threshold models
+
+## Linear breakpoint threshold models
+
+TODO
+
+These models can be fit by including `+ breakpt(x)` in the model formula, where
+`x` is a covariate.
+
+## Logistic threshold models
+
+The form is
+
+$$
+s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
+     / \left(s95 - s50 \right) } \right] }^{-1},
+$$
+where $\psi$ is a scaling parameter (controlling the height of the y-axis for
+the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
+a parameter controlling the point at which the function reaches 50% of the
+maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
+function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
+constrained to be larger than $s50$.
+
+These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.
+
+# Observation model families
+
+## [Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaee11f805f02bc1febc6d7bf0487671be)
+
+Internally parameterized as the [robust
+version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5).
+
+## [Beta](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga5324c83759d5211c7c2fbbad37fa8e59)
+
+$$
+\operatorname{Beta} \left(\mu \phi, 1 - \mu \phi \right)
+$$
+
+where $\phi$ is variance.
+
+$$
+\operatorname{Binomial} \left( N, \mu \right)
+$$
+
+$N = 1$ (i.e., 'size' currently fixed) and $\mu$ is probability
+
+## [Gamma](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gab0e2205710a698ad6a0ed39e0652c9a3)
+
+As shape, scale:
+
+$$
+\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
+$$
+
+where $\phi$ represents the shape and $\frac{\mu}{\phi}$ represents the scale.
+
+## [Gaussian](https://kaskr.github.io/adcomp/dnorm_8hpp.html)
+
+$$
+\operatorname{Normal} \left( \mu, \phi \right)
+$$
+where $\phi$ is the standard deviation (following Stan convention of SD not variance).
+
+## [Lognormal](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L22-L30)
+
+$$
+\operatorname{Lognormal} \left( \log \mu - \frac{\phi^2}{2}, \phi \right)
+$$
+
+## [Negative Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
+
+Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c)
+
+$$
+\operatorname{NB2} \left( \mu, \phi \right)
+$$
+
+Variance scales quadratically with mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.
+
+## [Poisson](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa1ed15503e1441a381102a8c4c9baaf1)
+
+$$
+\operatorname{Poisson} \left( \mu \right)
+$$
+
+## [Student-t](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L11-L20)
+
+$$
+\operatorname{Student-t} \left( \mu, \phi, \nu \right)
+$$
+
+where $\nu$, the degrees of freedom, is currently fixed at 3.
+
+## [Tweedie](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga262f3c2d1cf36f322a62d902a608aae0)
+
+[Source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) as in [cplm](https://CRAN.R-project.org/package=cplm) package. <https://link.springer.com/article/10.1007/s11222-012-9343-7>
+
+$$
+\operatorname{Tweedie} \left(\mu, p, \phi \right), \: 1 < p < 2
+$$
+
+## Matern parameterization
 
@@
   We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+
+# References

### 2021-11-30 doc/appendix-model.Rmd
Develop model description

@@
 ---
-title: Detailed model description
+title: sdmTMB statistical model description
 output:
@@
 bibliography: refs.bib
+header-includes:
+  \usepackage[left]{lineno}
+  \usepackage{bm}
+  \usepackage{amssymb}
 ---
@@
 
+As described in the main paper, the complete sdmTMB model can be written as:
+
 $$
@@
 \mathbb{E}(y_{\s,t}) &= \mu_{\s,t},\\
-  \mu_{\s,t} &= f^{-1}
-  \left( \boldsymbol{X} \boldsymbol{\beta} + \omega_s + \epsilon_{\s,t} \right),\\
+\mu_{\bm{s},t} &=
+f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} +
+\bm{\alpha}_g +
+\bm{X}_{2,\bm{s},t} \gamma_t +
+\bm{X}_{3,t} \zeta_{\bm{s}} +
+\omega_{\bm{s}} +
+\epsilon_{\bm{s},t} \right),
+\end{aligned}
+$$
+
+where
+
+* $y$ represents the response data;
+* $\mu$ represents the mean;
+* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
+* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
+* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
+* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
+* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
+* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
+* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
+* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
+
+We now split this model apart to describe the various parts in more detail
+
+## Fixed- or main-effect portion:
+
+$$
+\begin{aligned}
+\mu_{\bm{s},t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} \ldots \right)
+\end{aligned}
+$$
+
+Within `sdmTMB()`,
+
+* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
+  represents the fixed effect model matrix and a corresponding vector of
+  coefficients.
+
+* Get into smoothers, breakpoints, etc. here
+
+## Spatial random fields
+
+$$
+\begin{aligned}
+\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_s + \ldots \right),\\
 \boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
+\end{aligned}
+$$
+
+* $\omega_s$ are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
+
+## Spatiotemporal random fields
+
+Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
+
+### IID spatiotemporal random fields
+
+$$
+\begin{aligned}
+\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
 \boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
@@
 $$
-where $f$ is a link function (e.g., log or logit) and $f^{-1}$ is its inverse link.
 
-* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
-  represents the fixed effect model matrix and a corresponding vector of
-  coefficients
+* $\epsilon_{\s,t}$ are \ldots
 
-* $\omega_s$ are included if `spatial = 'on'`
+### AR1 spatiotemporal random fields
 
-* $\epsilon_{\s,t}$ are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`IID`; default), AR1 (`AR1`), random walk (`RW`), or off (`'off`) via the `spatiotemporal` argument
+Dropping other components for simplicity:
 
-# Random or multi-level intercepts
+$$
+\begin{aligned}
+\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
+\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
+\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
+\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\end{aligned}
+$$
+where $\rho$ is the correlation between subsequent spatiotemporal random fields.
+The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
+
+### Random walk spatiotemporal random fields
+
+Dropping other components for simplicity:
+
+$$
+\begin{aligned}
+\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
+\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
+\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_t,  \:
+\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\end{aligned}
+$$
+
+TODO
+
+<!-- * $\alpha_g$, are defined by an lme4-style formula, e.g. `formula = y ~ x + (1 | g)`, where `g` represents a column with grouping factors. -->
+
+<!-- * $\gamma_{t}$, time-varying coefficients, are defined by the `time_varying` formula argument. -->
+
+<!-- * $\zeta_{\bm{s}}$, IID random intercepts, are defined by the `spatial_varying` formula argument. -->
+
+<!-- * $f$ and $f^-1$ functions are defined by the `family` argument. E.g., `family = Gamma(link = "log")`. -->
+
+## Spatially varying coefficients
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \bm{X}_{3,t} \zeta_{\bm{s}} + \ldots \right),\\
+  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
+\end{aligned}
+$$
+
+where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
+Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant for a given time $t$.
+Currently, $\bm{X}_{3,t}$ is restricted to a single covariate, supplied by a formula to `spatial_varying`.
+
+## IID random or multi-level intercepts
 
@@
 $\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
-syntax as in lme4 or glmmTMB.
+syntax as in lme4 or glmmTMB. There can be multiple random intercepts. 
+E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.
 
-# Time-varying regression parameters
+## Time-varying regression parameters
 
@@
 
-# Spatial regression parameters
+If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept.
 
-$$
-\begin{aligned}
-  \mu_{\s,t} &= f^{-1} \left( \ldots + \zeta_s x_t + \ldots \right),\\
-  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
-\end{aligned}
-$$
-
-where $\zeta_s$ is a regression parameter that varies spatially and $x_t$ is a predictor that is constant for a given time $t$.
-Currently, $x_t$ must be the time variable itself.
-If you are interested in using a different predictor, contact the developers.
-
-# AR1 spatiotemporal random fields
-
-Dropping the optional $\omega_s$ for simplicity:
-
-$$
-\begin{aligned}
-\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
-\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
-\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
-\end{aligned}
-$$
-
-where $\rho$ is the correlation between subsequent spatiotemporal random
-fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
-
-# Random walk spatiotemporal random fields
-
-Dropping the optional $\omega_s$ for simplicity:
-
-$$
-\begin{aligned}
-\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
-\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_t,  \:
-\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
-\end{aligned}
-$$
-
-# Offset terms
+## Offset terms
 
@@
 $$
+
 where $O_{\s,t}$ is an offset term---a **log transformed** variable without
@@
 
-$N = 1$ (i.e., 'size' currently fixed) and $\mu$ is probability
+where $N$ is the size or number of trials and $\mu$ is the probability.
+
+Following the structure of `stats::glm()` and glmmTMB, a binomial family can be specified in one of 4 ways:
+
+1. the response may be a factor (and the model classifies the first level versus all others)
+2. the response may be binomial (0/1)
+3. the response can be a matrix of form `cbind(success, failure)`, and 
+4. the response may be the observed proportions, and the `weights` argument is used to specify the Binomial size ($N$) parameter (`prob ~ ..., weights = N`).
 
@@
 $$
-where $\phi$ is the standard deviation (following Stan convention of SD not variance).
+where $\mu$ is the mean and $\phi$ is the standard deviation (following Stan convention of SD not variance).
 
@@
 
-## [Negative Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
+TODO
 
-Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c)
+## [Negative Binomial 2 (NB2)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
 
@@
 
-Variance scales quadratically with mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.
+Where $\mu$ is the mean and $\phi$ is the dispersion parameter.
+The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.
+
+Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
+
+## [Negative Binomial 1 (NB1)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
+
+$$
+\operatorname{NB1} \left( \mu, \phi \right)
+$$
+Where $\mu$ is the mean and $\phi$ is the dispersion parameter.
+The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$.
+
+Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
 
@@
 
+where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 1, and $\phi$ is the dispersion parameter.
+
 ## Matern parameterization
 
-<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
-  The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
-  (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
+The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as 
+$$
+\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),$$
 <!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
-    where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
-  a scaling parameter that is estimated (REF).
+where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
 The parameter $\nu$ controls the smoothness of the covariance function.

### 2021-11-30 analysis/timing.R
working timing version with spaMM, hopefully...

@@ -283,7 +283,7 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
   out
 }
 
-test <- sim_fit_time(n_obs = 1000, max.edge = 0.1, family = gaussian(), seed = 1)
+test <- sim_fit_time(n_obs = 1000, max.edge = 0.2, family = gaussian(), seed = 1)
 test
 # test1 <- sim_fit_time(n_obs = 1000, max.edge = 0.2, family = gaussian(), seed = 1)
 # test1
@@ -294,6 +294,14 @@ to_run <- expand.grid(
   max.edge = c(0.06, 0.075, 0.1, 0.15, 0.2),
   iter = seq_len(20L)
 )
+
+# short test version
+# to_run <- expand.grid(
+#   n_obs = c(1000L),
+#   max.edge = c(0.1, 0.2),
+#   iter = seq_len(5L)
+# )
+
 to_run$seed <- to_run$iter * 29212
 nrow(to_run)
 
@@ -303,6 +311,8 @@ nrow(to_run)
 to_run <- to_run[sample(seq_len(nrow(to_run)), replace = FALSE), ]
 
 # out <- purrr::pmap_dfr(to_run, sim_fit_time)
+
+
 plan(multisession, workers = 6L)
 out <- furrr::future_pmap_dfr(
 # out <- purrr::pmap_dfr(
@@ -313,7 +323,7 @@ out <- furrr::future_pmap_dfr(
   .options = furrr::furrr_options(seed = TRUE,
     globals = c('simulate_dat', 'sim_fit_time',
       'Predict.matrix.spde.smooth', 'smooth.construct.spde.smooth.spec'),
-    packages = c('mgcv', 'inlabru', 'INLA', 'ggplot2', 'dplyr', 'sdmTMB', 'spaMM'))
+    packages = c('mgcv', 'inlabru', 'INLA', 'ggplot2', 'dplyr', 'sdmTMB'))
 )
 saveRDS(out, file = "analysis/timing-cache-parallel-openblas-spaMM.rds")
 out <- readRDS("analysis/timing-cache-parallel-openblas-spaMM.rds")

### 2021-11-30 doc/appendix-model.Rmd
Add to model appendix

@@
 
-# Basic model structure
+# Combined model structure
 
@@
 
-We now split this model apart to describe the various parts in more detail
+We now split this model apart to describe the various parts in more detail.
 
-## Fixed- or main-effect portion:
+## Main effects:
 
@@
 * Get into smoothers, breakpoints, etc. here
+
+### Smoothers
+
+TODO
+
+### Linear breakpoint threshold models
+
+TODO
+
+These models can be fit by including `+ breakpt(x)` in the model formula, where
+`x` is a covariate.
+
+### Logistic threshold models
+
+The form is
+
+$$
+s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
+     / \left(s95 - s50 \right) } \right] }^{-1},
+$$
+
+where $\psi$ is a scaling parameter (controlling the height of the y-axis for
+the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
+a parameter controlling the point at which the function reaches 50% of the
+maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
+function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
+constrained to be larger than $s50$.
+
+These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.
 
@@
 
-If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept.
+If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).
 
@@
 a coefficient (assuming a log link).
-
-# Threshold models
-
-## Linear breakpoint threshold models
-
-TODO
-
-These models can be fit by including `+ breakpt(x)` in the model formula, where
-`x` is a covariate.
-
-## Logistic threshold models
-
-The form is
-
-$$
-s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
-     / \left(s95 - s50 \right) } \right] }^{-1},
-$$
-where $\psi$ is a scaling parameter (controlling the height of the y-axis for
-the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
-a parameter controlling the point at which the function reaches 50% of the
-maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
-function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
-constrained to be larger than $s50$.
-
-These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.
 
@@
 
+# Random field details
+
 ## Matern parameterization
@@
 The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as
+
 $$
@@
 <!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
+
 where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
@@
 
+## Projection 'A' matrix
+
+## Anisoptropy
+
+## Barrier mesh
+
+# Optimization details
+
+The sdmTMB model is fit by maximum marginal likelihood.
+Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
+Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
+
+TODO two-stage fitting
+
+TODO extra loops and Newton loops
+
+# Assessing convergence
+
+TODO
+
 # References

### 2021-11-30 doc/paper.Rmd
Update table

@@
     '\\addlinespace',
-    rep('', 5), # correlation stuff
+    rep('', 6), # correlation stuff
     '\\addlinespace',
@@
     number = c(
+      "Technically possible but non-trivial",
       "Penalized smoother GAMs",
+      "Spatiotemporal fields as random walk only",
       "SPDE approach as in Miller et al. (2019)",
-      "Technically possible but non-trivial",
+      "Does have predictive process knots",
       "Zero-inflated NB2 only",

### 2021-12-01 doc/paper.Rmd
suggested edits to VAST para

@@
 
-# Spatially varying coefficients applied to citizen-science owl counts
+# Spatially varying coefficients from citizen-science count data
 
@@
 
-\clearpage
+<!-- \clearpage  -->
 
@@
 ```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
-f <- here::here("figs", "timing2-logx.pdf")
+f <- here::here("figs", "timing2-logx-PE.pdf")
 if (!file.exists(f)) {
@@
 There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
-VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
-The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
+VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.
+<!-- could add: 'and functionality' to the above if next sentence deemed redundant. -->
+<!-- The major difference lies in VAST's considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF). -->
 Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
 For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
-sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
-Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.
+From a fisheries perspective, sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
+Unlike VAST, however, sdmTMB does not currently extend to multivariate applications or support delta/hurdle models.
+<!-- I think either this final sentence or the 3rd sentence are a bit redundant. I suggest removing the 3rd one here. -->
 

### 2021-12-02 doc/paper.Rmd
CAR sentence
Add CAR sentences (w/out adding another ref) +

cleaning up

@@
 sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
-Geostatistical data simply refers to data observed at specific spatial coordinates (vs. areal data that might represent polygons or grids).
+Geostatistical data simply refers to data observed at specific spatial coordinates.
 The data can be collected across discrete points in time.
+Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
+sdmTMB can also fit models with areal data if each polygon has an associated centroid (observations are treated as replicates).
+A benefit of the sdmTMB approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g.
+range).
+
 The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.

### 2021-12-02 bbs_example.R

@@ -1,150 +1,0 @@
-library(bbsBayes)
-
-fetch_bbs_data() # takes a few minutes, run once
-# break BBS data into lon-lat 1 deg squares
-stratified_data <- stratify(by = "latlong")
-# Pull out species of interest, e.g. Smith et al. PLoS
-# https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0130768
-route_strat = stratified_data$route_strat
-#Olive-sided Flycatcher in Quebec/BCR 12
-indx <- grep("Olive-sided Flycatcher", stratified_data$species_strat$english)
-
-# Pull out info from all routes
-all_routes <- stratified_data$route_strat
-# join data
-stratified_data$bird_strat <- dplyr::filter(stratified_data$bird_strat,
-                                            AOU == as.numeric(stratified_data$species_strat$aou[indx]))
-# join in route info
-dat <- dplyr::left_join(stratified_data$route_strat, stratified_data$bird_strat)
-
-# grab routes for prediction later
-on_routes = dplyr::filter(stratified_data$route_strat, St_Abrev=="ON") %>%
-  dplyr::group_by(Route) %>%
-  dplyr::summarize(Latitude=Latitude[1],Longitude=Longitude[1])
-
-# pick out canadian data
-dat <- dplyr::filter(dat, St_Abrev %in% c("AB", "BC","MB","NB","NS","NT",
-                                          "NU","SK",
-                                          "ON","QC","YT"))
-
-# StopTotal is the total for all stops
-dat=dplyr::select(dat, Year,StopTotal,
-                  Latitude,Longitude,Month,Day,ObsN)
-
-
-# filter based on longitude
-dat = dplyr::filter(dat,Year >= 1970)
-
-library(sdmTMB)
-mesh = make_mesh(data = dat,
-                 xy_cols = c("Longitude","Latitude"),
-                 cutoff = 1)
-
-dat$date = paste0(dat$Month,"-",dat$Day,"-",dat$Year)
-dat$date = lubridate::parse_date_time(dat$date,orders="mdy")
-dat$jday = lubridate::yday(dat$date)
-dat$StopTotal[which(is.na(dat$StopTotal))] <- 0
-dat$present = ifelse(dat$StopTotal>0,1,0)
-
-ggplot(dat, aes(Longitude, Latitude,col=present)) +
-  geom_point() +
-  facet_wrap(~Year)
[…]

### 2021-12-02 doc/appendix-model.Rmd
Cross off some TODOs in the model appendix

@@
 
-* Get into smoothers, breakpoints, etc. here
-
 ### Smoothers
 
-TODO
+Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to `mgcv` users fitting GAMs (generalized additive models).
+Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
+Within these smooths, the same syntax commonly used in mgcv::s() can be applied, e.g.
+2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; the basis function dimensions may be specified, e.g.
+`+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, `+ s(month, bs = "cc", k = 12)`.
+
+While `mgcv` can fit unpenalized (e.g. B-splines, and others) or penalized splines (P-splines), sdmTMB only uses penalized splines. 
+The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed effects modeling framework.
+This is a similar approach implemented in `gamm4` and `brms`, among other packages.
 
@@
 
-TODO
+The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses. This function consists of two pieces, so that for $x < b_{1}$
+
+$$s(x) = x \cdot b_{0}$$
+and for $x > b_{1}$
+
+$$s(x) = b_{1} \cdot b_{0}$$
+In both cases, $b_{0}$ represents the slope of the function up to a threshold, and the product $b_{1} \cdot b_{0}$ represents the value at the asymptote.
+No constraints are placed on parameters $b_{0}$ or $b_{1}$.
 
@@
 The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
+The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is generally expected to be $-1 < \rho < 1$.
 
-### Random walk spatiotemporal random fields
+### Random walk spatiotemporal random fields (RW)
 
@@
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_t,  \:
-\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
+\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 
-TODO
+where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.
 
@@
 
-## Spatially varying coefficients
+## Spatially varying coefficients (SVC)
 
@@
 
-TODO
+where $\mu$ and $\phi$ are the mean and standard deviation of the natural log of the random variable.
+If $X$ is lognormally distributed, $E[log(X)] = \mu$ and $Var(log(X)) = \phi$.
+The expectation of $X$ is $exp(u + \frac{\phi^2}{2})$ and variance of X is $\left(exp(\phi^2) - 1 \right)exp(2\mu + \phi^2)$.
 

### 2021-12-02 neon_beetles.R

@@ -1,139 +1,0 @@
-
-library(neonstore)
-library(tidyverse)
-library(ISOweek)
-library(neonUtilities)
-# access data using EW's token
-#neon_download("DP1.10022.001", type = "expanded",.token=NEON_TOKEN)
-
-beetles <- neon_download("DP1.10022.001", type = "expanded",.token=NEON_TOKEN)
-neon_store(product = "DP1.10022.001")
-
-# Update `scientificName`, `taxonID`, `taxonRank` and `morphospeciesID` using assignments from parataxonomy and expert taxonomy.
-#
-library(dplyr)
-library(stringi)
-
-clean_names <- function (x)
-{
-  s <- stringi::stri_split_regex(x, "/", simplify = TRUE)[,1]
-  s <- stringi::stri_extract_all_words(s, simplify = TRUE)
-  if (dim(s)[2] > 1)
-    stringi::stri_trim(paste(s[, 1], s[, 2]))
-  else stringi::stri_trim(s[, 1])
-}
-
-resolve_taxonomy <- function(sorting, para, expert){
-
-  taxonomy <-
-    left_join(sorting,
-              select(para, subsampleID, individualID, scientificName, taxonRank, taxonID, morphospeciesID),
-              by = "subsampleID")  %>%
-    ## why are there so many other shared columns (siteID, collectDate, etc?  and why don't they match!?)
-    ## we use `select` to avoid these
-    left_join(
-      select(expert, -uid, -namedLocation, -domainID, -siteID, -collectDate, -plotID, -setDate, -collectDate),
-      by = "individualID") %>%
-    distinct() %>%
-    ## Prefer the para table cols over the sorting table cols only for sampleType=="other carabid"
-    mutate(taxonRank.x = ifelse(is.na(taxonRank.y) | sampleType != "other carabid", taxonRank.x, taxonRank.y),
-           scientificName.x = ifelse(is.na(scientificName.y) | sampleType != "other carabid", scientificName.x, scientificName.y),
-           taxonID.x = ifelse(is.na(taxonID.y) | sampleType != "other carabid", taxonID.x, taxonID.y),
-           morphospeciesID.x =  ifelse(is.na(morphospeciesID.y) | sampleType != "other carabid", morphospeciesID.x, morphospeciesID.y)) %>%
-    ## Prefer expert values where available
-    mutate(taxonRank = ifelse(is.na(taxonRank), taxonRank.x, taxonRank),
-           scientificName = ifelse(is.na(scientificName), scientificName.x, scientificName),
[…]

### 2021-12-02 doc/paper.Rmd
add notes

@@
 
+<!-- TODO cite: -->
+<!-- Dormann, C.F., McPherson, J.M., Araújo, M.B., Bivand, R., Bolliger, J., Carl, G., Davies, R.G., Hirzel, A., Jetz, W., Kissling, W.D., Kühn, I., Ohlemüller, R., Peres‐Neto, P.R., Reineking, B., Schröder, B., Schurr, F.M., and Wilson, R.
+(n.d.).
+Methods to account for spatial autocorrelation in the analysis of species distributional data: a review.
+Ecography 30(5): 609–628.
+doi:10.1111/j.2007.0906-7590.05171.x.
+-->
+
 Ecological data are often collected in space or in space repeatedly over time.
@@
 
+<!-- or areal/grid data with centroids instead of CAR/SAR models -->
+<!-- add CAR/SAR to table -->
 sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
@@
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
+
+<!-- highlight where each one excels ... -->
+<!-- no CV built into VAST or priors -->
+<!-- discussion integrates across speed, familiarity and table 1 -->
+<!-- results paragraph on Table 1 -->
+<!-- familiarity is an important word here -->
+<!-- add section on functionality comparison -->
 
@@
 <!-- zero inflation -->
-<!-- MVT? time-varying epsilon -->
+<!-- MVT? time-varying epsilon, non-gaussian fields -->
+
+<!-- limitations -->
+
+<!-- 3d models, continuous time, -->
+<!-- areal models, neighbours or not, must be CAR or SAR, ...
+but if have centroid, could use sdmTMB...
+why not CAR/SAR, spatial process...
+not 1:1 relationship between distance and spatial correlation?
+neibour based only...
+-->
+<!-- https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.701.1636&rep=rep1&type=pdf -->
 
@@
 -->
+
+<!-- CAR/SAR models? and why not if have...-->
+<!-- EOF paper from Jim-->
+<!-- testing that has been done... -->
+<!-- https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.15149 -->
+<!-- https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.15149 -->

### 2021-12-02 analysis/timing.R
Use pardiso if me

@@ -6,13 +6,15 @@ library(inlabru)
 library(mgcv)
 library(spaMM)
 # library(tictoc)
-library(future)
-plan(multisession)
+# library(future)
+# plan(multisession)
 INLA::inla.setOption(num.threads = "1:1")
 source(here::here("analysis/mgcv_spde_smooth.R"))
 
-# INLA::inla.setOption("pardiso.license", "~/Dropbox/licenses/pardiso.lic")
-# INLA::inla.setOption(inla.mode="experimental")
+if (Sys.info()[["user"]] == "seananderson") {
+  INLA::inla.setOption("pardiso.license", "~/Dropbox/licenses/pardiso.lic")
+  # INLA::inla.setOption(inla.mode="experimental")
+}
 
 # INLA expectation: linear growth in n_obs and O(n_mesh^(3/2)) growth in n_mesh
 

### 2021-12-03 doc/paper.Rmd
Edit and add to paper

@@
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or some of the flexibility for the particular class of models discussed here.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or lack some of the flexibility for the particular class of models discussed here.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
@@
 
-Here, we introduce our R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
+Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
 Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv, for a specific class of spatial and spatiotemporal models [@wood2017a].
 One common application is for species distribution models (SDMs), hence the package name.
-This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of related software.
+This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
 
@@
 Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
-sdmTMB can also fit models with areal data if each polygon has an associated centroid (observations are treated as replicates).
-A benefit of the sdmTMB approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g.
-range).
+sdmTMB can also fit models with areal data if each polygon has an associated centroid.
+<!-- (observations are treated as replicates). -->
+A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range).
 
-The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
+The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
 \@ref(fig:diagram)).
@@
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.
 
@@
 Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
-We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space).
+We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method.
+<!-- (omitted for space). -->
 
@@
 
-We can make predictions with the `predict()` method and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
-<!-- Here, we will predict on a 2x2 km grid that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
+We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
+Here, we will predict on a 2x2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
+The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
+<!-- The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
 -->
-The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
 The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
@@
 
-# Model selection and validation
+# Additional functionality
+
+## Model selection and validation
 
@@
 
-# Parameter priors/penalties
+## Parameter priors/penalties
 
@@
 
-# Derived quantities of interest
+## Derived quantities of interest
 
@@
 
-# Speed comparison
+# Package comparisons
+
+## Speed
 
@@
 mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
-With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (note that this option may not be possible or as fast for all models).
+With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
 Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern substantially (not shown).
-<!-- Results with optimized math libraries (openBLAS and PARDISO (REFs)) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv. -->
 Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
+Results with optimized math libraries (openBLAS [REF] and PARDISO [REFs]) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
 
@@
 
+## Functionality
+
+There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
+sdmTMB, VAST, INLA/inlabru, and spaMM are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
+sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
+To our knowledge, VAST is the only package to implement spatial (REF) and spatial dynamic factor analysis (REF) and spatial empirical orthogonal functional (EOF) analysis (REF).
+Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
+spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
+<!-- TODO: still true? -->
+There is considerable variability in the available observation likelihoods across packages (Table 1).
+
 ```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
@@
 
+\clearpage
+
 # Discussion
@@
 
+How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
+Assuming a given package can fit the model of interest (Table 1), we suggest the primary difference is the user interface.
+We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
+Users familiar with INLA will find inlabru approachable.
+Users already familiar with mgcv can adapt mgcv to fit similar models with custom code (REF) and INLA/inlabru and mgcv are also general purpose modelling packages.
+VAST is the sole option for fitting some multivariate models;
+alternatively, because VAST focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
+Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
+Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets (e.g., REFs).
+INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.
+
+<!-- However, with the SPDE, mgcv is typically slower than sdmTMB for similar models without optimized matrix libraries and lacks certain  -->
+
+<!--
 There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
 VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.
-<!-- could add: 'and functionality' to the above if next sentence deemed redundant. -->
-<!-- The major difference lies in VAST's considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF). -->
 Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
@@
 Unlike VAST, however, sdmTMB does not currently extend to multivariate applications or support delta/hurdle models.
-<!-- I think either this final sentence or the 3rd sentence are a bit redundant. I suggest removing the 3rd one here. -->
 
@@
 Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.
+-->
 
@@
 <!-- https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.701.1636&rep=rep1&type=pdf -->
+
+Additional functionality and future directions here?
 

### 2021-12-03 analysis/timing.R
Style .R

@@ -40,7 +40,7 @@ simulate_dat <- function(n_obs = 100,
   segm.bnd <- inla.mesh.segment(loc.bnd)
 
   # bnd <- INLA::inla.nonconvex.hull(cbind(predictor_dat$X, predictor_dat$Y),
-    # convex = -0.2)
+  # convex = -0.2)
   me <- INLA::inla.mesh.2d(
     boundary = segm.bnd,
     max.edge = c(max.edge, 0.2),
@@ -69,25 +69,23 @@ simulate_dat <- function(n_obs = 100,
   list(mesh = mesh, dat = sim_dat)
 }
 
-
-
-
 sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedie_p = 1.5,
-  seed = sample.int(.Machine$integer.max, 1), family = gaussian(), sigma_O = 0.2,
-  max.edge = 0.1) {
-
+                         seed = sample.int(.Machine$integer.max, 1), family = gaussian(), sigma_O = 0.2,
+                         max.edge = 0.1) {
   cat("max.edge:", max.edge, "\n")
   cat("n_obs:", n_obs, "\n")
   cat("iter:", iter, "\n")
   cat("simulating...\n")
 
-  s <- simulate_dat(n_obs = n_obs, cutoff = cutoff, seed = seed,
+  s <- simulate_dat(
+    n_obs = n_obs, cutoff = cutoff, seed = seed,
     iter = iter, family = family, max.edge = max.edge, tweedie_p = tweedie_p, phi = phi,
-    sigma_O = sigma_O)
+    sigma_O = sigma_O
+  )
   sim_dat <- s$dat
   mesh <- s$mesh
   times <- list()
-# browser()
+  # browser()
   cat("fitting...\n")
 
   cat("sdmTMB\n")
@@ -110,8 +108,9 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
   times$sdmTMB_norm <- out[["elapsed"]]
 
   dat_sp <- sp::SpatialPointsDataFrame(cbind(sim_dat$X, sim_dat$Y),
-    proj4string = sp::CRS('+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
[…]

### 2021-12-03 doc/paper.Rmd
8->7

@@
 
-knitr::kable(d, format = "latex", caption = "Comparison of functionality between eight R packages that can fit geostatistical GLMMs.", 
+knitr::kable(d, format = "latex", caption = "Comparison of functionality between seven R packages that can fit geostatistical GLMMs.", 
   booktabs = TRUE, 

### 2021-12-03 doc/paper.Rmd
Tweak timing fig

@@
 ```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
-f <- here::here("figs", "timing2-logx-PE.pdf")
-if (!file.exists(f)) {
-  download.file("https://www.dropbox.com/s/3aco85mizuplz7z/timing2-logx.pdf?dl=1", 
-    destfile = f)
-}
+f <- here::here("figs", "timing-spatial.pdf")
 knitr::include_graphics(f)

### 2021-12-03 analysis/timing.R

@@ -6,8 +6,8 @@ library(inlabru)
 library(mgcv)
 library(spaMM)
 # library(tictoc)
-# library(future)
-# plan(multisession)
+library(future)
+plan(multisession)
 INLA::inla.setOption(num.threads = "1:1")
 source(here::here("analysis/mgcv_spde_smooth.R"))
 
@@ -399,7 +399,7 @@ clean_names <- tribble(
   "INLA_eb_nolike", "INLA EB no like()",
   "spaMM", "spaMM",
   "mgcv_ml", "mgcv::bam\n(discretize = F) SPDE",
-  "mgcv_disc", "mgcv::bam\n(discretize = T) SPDE "
+  "mgcv_disc", "mgcv::bam\n(discretize = TRUE) SPDE"
 )
 clean_names$model_clean <- factor(clean_names$model_clean, levels = clean_names$model_clean)
 out_long <- left_join(out_long, clean_names)
@@ -434,7 +434,25 @@ colour_vect <- c(
   "#8E2043" # bordeaux
 )
 
-library(colorblindr)
+cols <- RColorBrewer::brewer.pal(6, "Dark2")
+colour_vect <- c(
+  "sdmTMB" = cols[1],
+  "sdmTMB(normalize = TRUE)" = cols[5],
+  "inlabru EB" = cols[3],
+  "spaMM" = cols[4],
+  "mgcv::bam\n(discretize = TRUE) SPDE" = cols[2]
+)
+
+# library(colorblindr)
+
+leg <- tribble(
+  ~model_clean, ~mean_mesh_n, ~time,
+  "sdmTMB", 170, 0.8,
+  "sdmTMB(normalize = TRUE)", 400, 0.43,
+  "inlabru EB", 150, 8,
+  "spaMM", 500, 4.7,
+  "mgcv::bam\n(discretize = TRUE) SPDE", 245, 40
+)
 
 g <- out_long_sum %>%
   filter(model_clean != "mgcv::bam\n(discretize = F) SPDE") %>%
@@ -451,22 +469,24 @@ g <- out_long_sum %>%
   # theme(legend.position = "bottom") +
   # theme(legend.position = c(0.35, 0.87), legend.background = element_rect(fill = "white")) +
   theme(legend.position = c(0.4, 0.85)) +
-  scale_fill_OkabeIto(order = c(2, 5, 7, 3, 4)) +
-  scale_colour_OkabeIto(order = c(2, 5, 7, 3, 4)) +
[…]

### 2021-12-03 analysis/timing.R
ylim

@@ -476,7 +476,7 @@ g <- out_long_sum %>%
   scale_fill_manual(values = colour_vect) +
   scale_colour_manual(values = colour_vect) +
   labs(y = "Time (s)", x = "Mesh nodes", colour = "Model", fill = "Model") +
-  coord_cartesian(expand = FALSE) +
+  coord_cartesian(expand = FALSE, ylim = c(0.2, 102)) +
   theme(panel.spacing.x = unit(20, "pt"), legend.position = "none")
   # guides(
     # colour = guide_legend(nrow = 3, byrow = TRUE, title.theme = element_blank()),

### 2021-12-03 doc/paper.Rmd
add footnote

@@
       "Technically possible but non-trivial",
-      "Technically possible but non-trivial"
+      "Technically possible but non-trivial",
+      "Available in a companion package INLAutils"
     ),

### 2021-12-03 doc/paper.Rmd
Add a rough future directions paragraph

@@
 
-Additional functionality and future directions here?
+Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, spatial covariance barriers, time-varying spatiotemporal variance, and simulation from the parameter joint precision matrix.
+Future development may include zero-inflated models, improvements to Bayesian sampling efficiency, continuous time models, and non-gaussian random fields (REF).
+We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
+We may introduce capability for multivariate response data, but VAST already focusses on such models with a TMB backend.
+Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data.
 
@@
 
+TODO
+Would not be possible without TMB and INLA.
+Heavily inspired by and in some places code has been adapted from VAST and glmmTMB...
+
 # Conflict of Interest statement
 
+None?
+
 # Author Contributions
+
+TODO
 

### 2021-12-03 doc/paper.Rmd
No clearpage
EOF

@@
 sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
-To our knowledge, VAST is the only package to implement spatial (REF) and spatial dynamic factor analysis (REF) and spatial empirical orthogonal functional (EOF) analysis (REF).
+To our knowledge, VAST is the only package to implement spatial (REF) and spatial dynamic factor analysis (REF) and spatial empirical orthogonal function (EOF) regression (REF).
 Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
@@
 
-\clearpage
-
 # References

### 2021-12-03 doc/appendix-model.Rmd
add optimization details

@@
 
+TODO
+
 ## Anisoptropy
 
+TODO
+
 ## Barrier mesh
+
+TODO
 
@@
 
-TODO two-stage fitting
+Like AD Model Builder, TMB allows for parameters to be fit in phases and we inlcude the `multiphase` argument in `sdmTMB::sdmTMBcontrol()` to allow this.
+For high dimensional models (lots of fixed and random effects), phased estimation may be faster and result in more stable convergence.
+In sdmTMB, phased estimation proceeds by first estimating all fixed effect parameters contributing to the likelihood (holding random effects constant at initial values).
+In the second phase, the random effect parameters (and their variances) are also estimated.
+Fixed effect parameters are also estimated in the second phase, and initialized at their estimates from the first phase.
 
-TODO extra loops and Newton loops
+In some cases, a single call to `stats::nlminb()` may not be result in convergence (or a small maximum gradient), and the algorithm may need to be run multiple times.
+In the `sdmTMB::sdmTMBcontrol()` function, we include an argument `nlminb_loops` that will restart the optimization at the previous best values.
+The number of `nlminb_loops` should generally be small (e.g.
+2 or 3 initially), and defaults to 1.
+For some sdmTMB models, the Hessian may also be unstable and need to be re-evaluated.
+We do this optionally with the `stats::optimHess()` routine after the call to `stats::nlminb()`.
+The `stats::optimHess()` function implements a Newton optimization routine to find the Hessian, and we include the argument `newton_loops` in `sdmTMB::sdmTMBcontrol()` to allow for multiple function evaluations (each starting at the previous best value).
+By default, this is not included and `newton_loops` is set to 0.
 

### 2021-12-08 doc/paper.Rmd
cite dormann2007

@@
 output:
-    bookdown::pdf_document2:
+    bookdown::word_document2:  
+    # bookdown::pdf_document2:
       toc: false
@@
 
-<!-- TODO cite: -->
-<!-- Dormann, C.F., McPherson, J.M., Araújo, M.B., Bivand, R., Bolliger, J., Carl, G., Davies, R.G., Hirzel, A., Jetz, W., Kissling, W.D., Kühn, I., Ohlemüller, R., Peres‐Neto, P.R., Reineking, B., Schröder, B., Schurr, F.M., and Wilson, R.
-(n.d.).
-Methods to account for spatial autocorrelation in the analysis of species distributional data: a review.
-Ecography 30(5): 609–628.
-doi:10.1111/j.2007.0906-7590.05171.x.
--->
-
 Ecological data are often collected in space or in space repeatedly over time.
@@
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
 

### 2021-12-08 doc/paper.Rmd
fix ref formatting and add optimizer refs to text

@@
 output:
-    bookdown::word_document2:  
-    # bookdown::pdf_document2:
+    # bookdown::word_document2:  
+    bookdown::pdf_document2:
       toc: false
@@
 Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries (openBLAS [REF] and PARDISO [REFs]) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
+Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2a, @pardiso-7.2b, @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
+<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
 
@@
 sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
-To our knowledge, VAST is the only package to implement spatial (REF) and spatial dynamic factor analysis (REF) and spatial empirical orthogonal function (EOF) regression (REF).
+To our knowledge, VAST is the only package to implement spatial (REF) and spatial dynamic factor analysis (REF). 
+<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
 Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.

### 2021-12-08 doc/paper.Rmd
add a couple thorson refs and gelfand2017

@@
 <!-- TODO: cite: -->
-<!-- @article{gelfand2017, -->
-<!--   title = {Bayesian Modeling and Analysis of Geostatistical Data}, -->
-<!--   author = {Gelfand, Alan E. and Banerjee, Sudipto}, -->
-<!--   year = {2017}, -->
-<!--   journal = {Annual Review of Statistics and Its Application}, -->
-<!--   volume = {4}, -->
-<!--   number = {1}, -->
-<!--   pages = {245--266}, -->
-<!--   doi = {10.1146/annurev-statistics-060116-054155}, -->
-<!--   annotation = {\_eprint: https://doi.org/10.1146/annurev-statistics-060116-054155} -->
 
@@
 Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
-These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
+These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
 This can make an enormous difference to computational efficiency [@heaton2017].
@@
 sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
-To our knowledge, VAST is the only package to implement spatial (REF) and spatial dynamic factor analysis (REF). 
+To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
 <!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
@@
 We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
-We may introduce capability for multivariate response data, but VAST already focusses on such models with a TMB backend.
+We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend.
 Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data.

### 2021-12-09 doc/appendix-model.Rmd
censored-poisson?
Add to the model appendix

@@
 
-* Bold subscript $\s$ since x, y
+* Bold subscript $\s$ (a point in space) since it represents a vector of length 2
 
@@
 
-* $\phi$ for all 'dispersion' parameters for consistency with code
-
-* Family titles link to TMB or sdmTMB source code
-
-* Attempt to link math symbols to argument options in model and code symbols
+* $\phi$ for all 'dispersion' parameters for consistency with the code
 
@@
   "$\\rho$", "Correlation between random fields in subsequent time steps"
-
 )
@@
 * $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
-* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
+* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
 * $\bm{\beta}$ represents a vector of fixed-effect coefficients;
@@
 
-## Main effects:
+## Main effects
 
@@
 \begin{aligned}
-\mu_{\bm{s},t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} \ldots \right)
+\mu_{\bm{s},t} &= f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} \ldots \right)
 \end{aligned}
@@
 
-Within `sdmTMB()`,
-
-* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
-  represents the fixed effect model matrix and a corresponding vector of
-  coefficients.
+Within `sdmTMB()`, $\bm{X}_{1,\bm{s},t} \bm{\beta}$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients.
+This main effect formula can contain optional penalized smoothers or non-linear functions as defined below.
 
@@
 
-Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to `mgcv` users fitting GAMs (generalized additive models).
+Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
 Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. 
-Within these smooths, the same syntax commonly used in mgcv::s() can be applied, e.g.
+Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g.
 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; the basis function dimensions may be specified, e.g.
-`+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, `+ s(month, bs = "cc", k = 12)`.
+`+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g.
+`+ s(month, bs = "cc", k = 12)`.
 
-While `mgcv` can fit unpenalized (e.g. B-splines, and others) or penalized splines (P-splines), sdmTMB only uses penalized splines. 
-The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed effects modeling framework.
+While mgcv can fit unpenalized (e.g., B-splines) or penalized splines (P-splines), sdmTMB only implements penalized splines. 
+The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed-effects modelling framework.
 This is a similar approach implemented in `gamm4` and `brms`, among other packages.
@@
 
-The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses. This function consists of two pieces, so that for $x < b_{1}$
-
-$$s(x) = x \cdot b_{0}$$
-and for $x > b_{1}$
-
-$$s(x) = b_{1} \cdot b_{0}$$
+The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses.
+This function consists of two pieces, so that for $x < b_{1}$, $s(x) = x \cdot b_{0}$, and for $x > b_{1}$, $s(x) = b_{1} \cdot b_{0}$.
 In both cases, $b_{0}$ represents the slope of the function up to a threshold, and the product $b_{1} \cdot b_{0}$ represents the value at the asymptote.
@@
 
-The form is
+Models with logistic threshold relationships between a predictor and the response can be fit with the form
 
@@
 
-where $\psi$ is a scaling parameter (controlling the height of the y-axis for
+where $s$ represents the logistic function, 
+$\psi$ is a scaling parameter (controlling the height of the y-axis for
 the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
@@
 
+Spatial random fields, $\omega_s$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
+
 $$
@@
 
-* $\omega_s$ are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
-
 ## Spatiotemporal random fields
 
-Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
+Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
 
 ### IID spatiotemporal random fields
+
+IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as
 
@@
 
-* $\epsilon_{\s,t}$ are \ldots
+where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
+The random fields are assumed independent across time slices.
 
@@
 
-Dropping other components for simplicity:
+AR1 spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. The are defined as
 
@@
 \begin{aligned}
-\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
+\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} \ldots \right),\\
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
@@
 The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
-The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is generally expected to be $-1 < \rho < 1$.
+The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
 
@@
 
-Dropping other components for simplicity:
+Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time slice to the next are IID. There defined as
 
@@
 \begin{aligned}
-\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
+\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} + \ldots \right),\\
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
@@
 where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.
+Note that, in contrast to the AR1 parametrization, the variance is no longer the steady-state marginal variance.
 
@@
 ## Spatially varying coefficients (SVC)
+
+Spatially varying coefficient (SVC) models are defined as
 
@@
 
+Parameters can be modelled as time varying (a random walk) according to the form
+
 $$
@@
 
-where $\gamma_t$ is an example optional time-varying regression parameter and
-$x_{\s,t}$ is the corresponding predictor value. This is defined via the
-`time_varying` argument. E.g., `~ 0 + x`. The first value is given an implicit
-uniform prior. I.e., the same variable should not appear in the fixed effect
-formula since the initial value is estimated as part of the time-varying formula.
-
-If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).
+where $\gamma_t$ is an example optional time-varying regression parameter and $x_{\s,t}$ is the corresponding predictor value.
+This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
+For example, `time_varying ~ 0 + x`.
+The first value is given an implicit uniform prior. 
+I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
+The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).
 
@@
 E.g., `y ~ x + offset`.
-
 These are included in the linear predictor as
@@
 
-where $O_{\s,t}$ is an offset term---a **log transformed** variable without
-a coefficient (assuming a log link).
+where $O_{\s,t}$ is an offset term---a **log transformed** variable without a coefficient (assuming a log link).
 
@@
 
+## [Censored-poisson] TODO
+
 ## [Student-t](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L11-L20)

### 2021-12-09 doc/paper.Rmd

@@
 <!-- TODO: cite: -->
-
 <!-- @book{diggle2007, -->
@@
 To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
+However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020].
 <!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
@@
 Users familiar with INLA will find inlabru approachable.
-Users already familiar with mgcv can adapt mgcv to fit similar models with custom code (REF) and INLA/inlabru and mgcv are also general purpose modelling packages.
+Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [e.g., @camp2020] and INLA/inlabru and mgcv are also general purpose modelling packages.
 VAST is the sole option for fitting some multivariate models;
@@
 Future development may include zero-inflated models, improvements to Bayesian sampling efficiency, continuous time models, and non-gaussian random fields (REF).
+<!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
 We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
@@
 
-Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management (REF).
+Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management [@sofaer2019.
+<!-- Not sure if you have ideas here? @anderson2019 and @english2021 are my best ideas at the moment ;) -->
 These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.

### 2021-12-09 doc/paper.Rmd
Comment out brodie2020 for now

@@
 To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
-However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020].
+<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
 <!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->

### 2021-12-09 doc/paper.Rmd
Switch to miller2019

@@
 Users familiar with INLA will find inlabru approachable.
-Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [e.g., @camp2020] and INLA/inlabru and mgcv are also general purpose modelling packages.
+Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
 VAST is the sole option for fitting some multivariate models;

### 2021-12-09 doc/paper.Rmd
Remove 2 pardiso refs

@@
 Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2a, @pardiso-7.2b, @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
+Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
 <!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

### 2021-12-09 doc/paper.Rmd
Removed note

@@
 # Introduction
-
-<!-- TODO: cite: -->
-<!-- @book{diggle2007, -->
-<!--   title = {Model-Based {{Geostatistics}}}, -->
-<!--   author = {Diggle, Peter and Ribeiro, Paulo Justiniano}, -->
-<!--   year = {2007}, -->
-<!--   publisher = {{Springer}}, -->
-<!--   googlebooks = {qCqOm39OuFUC}, -->
-<!--   isbn = {978-0-387-48536-2}, -->
-<!--   langid = {english} -->
-<!-- } -->
-
-<!-- } -->
 

### 2021-12-09 doc/appendix-model.Rmd
anisotropy

@@
 
-TODO
+TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longtidue ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)).
+Anistropy can be turned on and off with the boolean `anisotropy` argument to `sdmTMB()`.
+There are a number of ways to implement anisotropic covariance (Fuglstad et al.
+2015), and we adopt a 2-parameter rotation matrix $\textbf{H}$.
+The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$
+
+Once a model is fitted with `sdmTMB()`, the anisotropy relationships may be plotted using the `plot_anisotropy()` function, which takes the fitted object as an argument.
+Note: if you use a barrier mesh, anisotropy is disabled.
 

### 2021-12-09 doc/isec_abstract.Rmd
isec

---
title: "ISEC_abstract"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Analyzing geostatistical data (coordinate referenced observations) is becoming increasingly common in ecology; these data often include a discrete temporal dimension, such as monthly or yearly samples. Computational tools to model spatial and spatiotemporal processes have advanced rapidly over the last decade; examples of R packages for analyzing geostatistical data include `mgcv`, `INLA` and `inlabru`, `VAST`, and `spBayes`. These packages are 
In this talk, we present a new R package `sdmTMB`. The goals of `sdmTMB` are not to replace functionality in existing packages, but provide a user friendly implementation of SPDE models -- similar to the popular package `glmmTMB`, but with the addition of spatial models. We extend the generalized linear mixed models (GLMMs) familiar to ecologists to include the following optional features:
* spatial fields
* spatiotemporal fields, that may be independent by year or modeled with random walks or autoregressive processes
* smooth terms for covariates, using the familiar `s()` notation in `mgcv`
* spatially varying coefficient models (SVC)
* non-stationary covariates (coefficients modeled as random walks)

Like `glmmTMB`, estimation is performed in `sdmTMB` via maximum likelihood, implemented  using the Template Model Builder (TMB) software.

We provide a general overview of the package structure, and demonstrate features using several examples (Snowy Owl distribution in North America, and a dataset describing the distribution of Pacific cod in British Columbia).

### 2021-12-09 doc/isec_abstract.Rmd
clarify owl model

@@
 
-We provide a general overview of the package structure, and demonstrate features using several examples (Snowy Owl distribution in North America, and a dataset describing the distribution of Pacific cod in British Columbia).
+We provide a general overview of the package structure, and demonstrate features using several examples (Snowy Owl winter irruptions in North America, and a dataset describing the distribution of Pacific cod in British Columbia).

### 2021-12-09 doc/appendix-pcod-VAST-tweedie.Rmd
init Tweedie version

--- 2021-11-24 doc/appendix-pcod-VAST.Rmd
@@
 For this example, we will include a factor predictor that represents the mean estimate for each time slice.
-Settings used for index standardization are specified partially by specifying `purpose = "index2"` but we also explicitly provide arguments for these and other key settings here.
+Settings used for index standardization are specified partially by defining `purpose = "index2"` but we also explicitly provide arguments for these and other key settings here.
+Specifically, we set the elements of `RhoConfig` to 0 to specify that we want the intercepts and spatiotemporal fields to be independent (which is the default).
+
+VAST is constructed to accommodate many different model structures, including delta or hurdle models, and thus requires the user to specify arguments for the spatial and spatiotemporal fields of two model components.
+In this case, we want to fit a model with a single component, using the Tweedie distribution.
+To do so, we must turn off estimation of omega and epsilon for 1st linear predictor by entering 0 for those elements in the object we'll use to define the configuration of these fields `FieldConfig`.
 
 ```{r config, echo = TRUE}
-FieldConfig <- matrix(c("IID", "IID", "IID", "IID", "IID", "IID"),
-  ncol = 2, nrow = 3,
+FieldConfig <- matrix(c("0", "0", "IID", "Identity", "IID", "IID", "IID", "Identity"),
+  ncol = 2, nrow = 4,
   dimnames = list(
-    c("Omega", "Epsilon", "Beta"),
+    c("Omega", "Epsilon", "Beta", "Epsilon_year"),
     c("Component_1", "Component_2")
@@
 FieldConfig
+
 RhoConfig <- c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
@@
   Region = "User",
-  purpose = "index2", # index of abundance with Gamma for positive catches
+  purpose = "index2", # use recommended defaults for an index of abundance
   fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
@@
   RhoConfig = RhoConfig,
-  ObsModel = c(2, 0), # this is for conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
+  ObsModel = c(10, 2), # use the Tweedie distribution as the observation model
   bias.correct = FALSE,
@@
  # create folder for saved output:
-dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)
+dir.create(paste0(here("doc", "appendix-VAST-tweedie")), showWarnings = FALSE)
 
 # effort is 1 when using CPUE instead of observed weight as the response:
-pcod$effort <- as.numeric(1)
-
+pcod$effort <- 1
+pcod <- as.data.frame(pcod) # ensure not a tibble
 fit <- fit_model(
@@
   input_grid = input_grid,
-  working_dir = paste0(here("doc", "appendix-VAST"), "/")
+  working_dir = paste0(here("doc", "appendix-VAST-tweedie"), "/")
 )
@@
   check_residuals = FALSE,
-  working_dir = paste0(here("doc", "appendix-VAST"), "/")
+  working_dir = paste0(here("doc", "appendix-VAST-tweedie"), "/")
 )
@@
 ```{r plot-knots, echo = TRUE}
-knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
+knitr::include_graphics(here("doc", "appendix-VAST-tweedie", "Data_and_knots.png"))
 ```
@@
 ```{r plot-density, echo = TRUE}
-knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
+knitr::include_graphics(here("doc", "appendix-VAST-tweedie", "ln_density-predicted.png"))
 ```
@@
 ```{r plot-index, echo = TRUE, out.width="3in"}
-if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
-  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
+if (file.exists(here("doc", "appendix-VAST-tweedie", "Index-Biomass.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST-tweedie", "Index-Biomass.png"))
 }
-if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
-  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
+if (file.exists(here("doc", "appendix-VAST-tweedie", "Index.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST-tweedie", "Index.png"))
 }

### 2021-12-09 doc/appendix-pcod-VAST-tweedie.Rmd
fix formatting

@@
 
-```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
+```{r packages, message = FALSE, warning = FALSE, cache = FALSE}
 library(TMB)
@@
 
-```{r fit, echo = TRUE, results = 'hide'}
+```{r fit, echo = TRUE, results = FALSE, message = FALSE, warning = FALSE}
  # create folder for saved output:
@@
 
-```{r save-plots, echo = TRUE, warning = FALSE}
+```{r save-plots, echo = TRUE, message = FALSE, warning = FALSE}
 plot(
@@
 
-```{r plot-density, echo = TRUE}
+```{r plot-density, echo = TRUE, out.width = "6in"}
 knitr::include_graphics(here("doc", "appendix-VAST-tweedie", "ln_density-predicted.png"))
@@
 
-```{r plot-index, echo = TRUE, out.width="3in"}
+```{r plot-index, echo = TRUE, out.width = "3in"}
 if (file.exists(here("doc", "appendix-VAST-tweedie", "Index-Biomass.png"))) {

### 2021-12-09 doc/paper.Rmd
Edits

@@
 
-## Model selection and validation
+## Model validation and selection
 
-Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]. 
-Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
-The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
-The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.
-
-Validation of state-space models is non-trivial, particularly those using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation:
-
-1. The cross validation function `sdmTMB_cv()` can be used to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
-
-2. The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
-These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.
-These residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
-One-step-ahead residuals [@thygesen2017a] or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are much slower.
-
-3.
-The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
-Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE).
-
-4.
-An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017] to evaluate the accuracy of the Laplace approximation or evaluate residuals (see `?extract_mcmc`).
+Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. 
+We provide several approaches to assist this process: 
+(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
+<!-- We allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]; however, because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
+-->
+<!-- The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. -->
+<!-- The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model. -->
+<!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
+(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection (e.g., via expected log predictive density, ELPD, REF) or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
+(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
+(4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
+<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.
+-->
+For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
+Simulation-based residuals [@DHARMa], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
+(5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
+<!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
 <!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->

### 2021-12-09 doc/isec_abstract.Rmd
edits

@@
 Like `glmmTMB`, estimation is performed in `sdmTMB` via maximum likelihood, implemented  using the Template Model Builder (TMB) software.
+The `sdmTMB` package also allows for models to be passed to Stan, allowing for efficient Bayesian estimation.
 
 We provide a general overview of the package structure, and demonstrate features using several examples (Snowy Owl winter irruptions in North America, and a dataset describing the distribution of Pacific cod in British Columbia).
+We highlight the advantages of using `sdmTMB` for these classes of models: (1) estimation in TMB is often faster than with alternative estimation routines, (2) `sdmTMB` provides a user friendly interface that is familiar to many R users, (3) `sdmTMB` allows for easy out of sample model validation, not found in other packages, (4) non-stationary processes are easily constructed with the formula interface (time varying coefficients, spatially varying coefficient models).
+For analysts fitting species distribution models (SDMs) or estimating population abundance trends, `sdmTMB` provides specific functions for performing these calculations.

### 2021-12-09 doc/paper.Rmd
Edits

@@
 <!-- (observations are treated as replicates). -->
-A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range).
+A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range---the distance at which points are effectively independent).
 
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
+After rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+
+sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017] (`?pc_matern`) can constrain the spatial range and variance parameters.
+
+The results from an sdmTMB model can be used to generate various quantities of interest.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
+Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
 
@@
 
-# An introductory example
+# Introductory example with fish survey biomass
 
@@
 
-# Spatially varying coefficients from citizen-science count data
+# Spatially varying coefficient example with citizen science data
 
@@
 
-# Additional functionality
-
-## Model validation and selection
+# Model validation and selection
 
@@
 <!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->
-
-## Parameter priors/penalties
-
-sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.
-
-## Derived quantities of interest
-
-The results from an sdmTMB model can be used to generate various quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`; Fig.
-\@ref(fig:pcod-fig)E) given a user supplied grid.
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
 
@@
 
-## Speed
+There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
+sdmTMB, VAST, INLA/inlabru, and spaMM are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
+sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
+To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
+<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
+<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
+Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
+spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
+<!-- TODO: still true? -->
+There is considerable variability in the available observation likelihoods across packages (Table 1).
+
+```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
+f <- here::here("figs", "timing-spatial.pdf")
+knitr::include_graphics(f)
+```
 
@@
 <!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
-
-## Functionality
-
-There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
-sdmTMB, VAST, INLA/inlabru, and spaMM are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
-sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
-sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
-To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
-<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
-<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
-Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
-spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
-<!-- TODO: still true? -->
-There is considerable variability in the available observation likelihoods across packages (Table 1).
-
-```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
-f <- here::here("figs", "timing-spatial.pdf")
-knitr::include_graphics(f)
-```
 

### 2021-12-09 doc/paper.Rmd
Edits

@@
   \renewcommand{\thefigure}{S\arabic{figure}}}
+  \widowpenalty10000
+  \clubpenalty10000
 editor_options:
@@
 
-\clearpage
-
 # Spatially varying coefficient example with citizen science data
@@
 
-\clearpage
-
 # Discussion
@@
 We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
-Users familiar with INLA will find inlabru approachable.
-Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
+Users familiar with INLA may find inlabru approachable.
+Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
 VAST is the sole option for fitting some multivariate models;

### 2021-12-09 doc/paper.Rmd
Edits

@@
 Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, spatial covariance barriers, time-varying spatiotemporal variance, and simulation from the parameter joint precision matrix.
-Future development may include zero-inflated models, improvements to Bayesian sampling efficiency, continuous time models, and non-gaussian random fields (REF).
+Future development may include zero-inflated models, improvements to Bayesian sampling efficiency, continuous time models (REF?), and non-gaussian random fields [e.g., @anderson2019].
 <!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
@@
 
-Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management [@sofaer2019.
-<!-- Not sure if you have ideas here? @anderson2019 and @english2021 are my best ideas at the moment ;) -->
+Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @english2021] (OTHER REF?) and improve ecological management [@sofaer2019].
 These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.

### 2021-12-09 doc/paper.Rmd
Add ref

@@
 
-Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @english2021] (OTHER REF?) and improve ecological management [@sofaer2019].
+Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2021] and improve ecological management [@sofaer2019].
 These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.

### 2021-12-09 doc/paper.Rmd
Edits

@@
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017] (`?pc_matern`) can constrain the spatial range and variance parameters.
+Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
 
@@
 Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
-Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets (e.g., REFs).
+Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @english2021].
 INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.

### 2021-12-09 doc/isec_abstract.Rmd
Edits

--- 2021-12-09 doc/isec_abstract.Rmd
@@
 title: "ISEC_abstract"
-output: pdf_document
+output: word_document
 ---
@@
 
-Analyzing geostatistical data (coordinate referenced observations) is becoming increasingly common in ecology; these data often include a discrete temporal dimension, such as monthly or yearly samples.
-Computational tools to model spatial and spatiotemporal processes have advanced rapidly over the last decade; examples of R packages for analyzing geostatistical data include `mgcv`, `INLA` and `inlabru`, `VAST`, and `spBayes`.
-These packages are
-In this talk, we present a new R package `sdmTMB`.
-The goals of `sdmTMB` are not to replace functionality in existing packages, but provide a user friendly implementation of SPDE models -- similar to the popular package `glmmTMB`, but with the addition of spatial models.
+Analyzing geostatistical data (coordinate-referenced observations) is becoming increasingly common in ecology; these data often include a discrete temporal dimension, such as monthly or yearly samples.
+Computational tools to model spatial and spatiotemporal processes have advanced rapidly over the last decade; examples of R packages for analyzing geostatistical data include mgcv, INLA and inlabru, VAST, and spBayes.
+SOMETHING HERE...
+In this talk, we present a new R package sdmTMB.
+The goals of sdmTMB are not to replace functionality in existing packages, but provide a user-friendly implementation of SPDE models---similar to the popular R package glmmTMB, but with the addition of spatial and spatiotemporal models with an SPDE approach.
 We extend the generalized linear mixed models (GLMMs) familiar to ecologists to include the following optional features:
+
 * spatial fields
-* spatiotemporal fields, that may be independent by year or modeled with random walks or autoregressive processes
-* smooth terms for covariates, using the familiar `s()` notation in `mgcv`
-* spatially varying coefficient models (SVC)
+* spatiotemporal fields that may be independent by year or modeled with random walks or autoregressive processes
+* smooth terms for covariates, using the familiar `s()` notation in mgcv
+* spatially varying coefficient models (SVCs)
 * non-stationary covariates (coefficients modeled as random walks)
 
-Like `glmmTMB`, estimation is performed in `sdmTMB` via maximum likelihood, implemented  using the Template Model Builder (TMB) software.
-
-We provide a general overview of the package structure, and demonstrate features using several examples (Snowy Owl distribution in North America, and a dataset describing the distribution of Pacific cod in British Columbia).
+Estimation is performed in sdmTMB via maximum marginal likelihood, implemented using Template Model Builder (TMB).
+The sdmTMB package also allows for models to be passed to Stan, allowing for efficient Bayesian estimation.
+We provide a general overview of the package structure, and demonstrate features using several examples including Snowy Owl winter irruptions in North America and a dataset describing the distribution of Pacific Cod in British Columbia.
+We highlight advantages of using sdmTMB for these classes of models including (1) estimation in TMB is often faster than alternative estimation routines, (2) sdmTMB provides a user-friendly interface that is familiar to many R users, (3) sdmTMB allows for easy out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficient models).
+For analysts fitting species distribution models (SDMs) or estimating population abundance trends, sdmTMB provides specific functions for performing these calculations.

### 2021-12-09 doc/paper.Rmd
Edits

@@
 
+We could generate a population index that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
+\@ref(fig:pcod-fig)E, APPENDIX).
+
 (ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
@@
 Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, spatial covariance barriers, time-varying spatiotemporal variance, and simulation from the parameter joint precision matrix.
-Future development may include zero-inflated models, improvements to Bayesian sampling efficiency, continuous time models (REF?), and non-gaussian random fields [e.g., @anderson2019].
+Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-gaussian random fields [e.g., @anderson2019].
 <!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->

### 2021-12-10 doc/appendix-model.Rmd
number appendices and cite them in paper

@@
 ---
-title: sdmTMB statistical model description
+title: "Appendix 1: sdmTMB statistical model description"
 output:

### 2021-12-10 doc/appendix-snow-sdmTMB.Rmd
number appendices and cite them in paper

@@
 ---
-title: "Annotated example of a spatially varying effect with sdmTMB"
+title: "Appendix 3: example of a spatially varying effect of climate on citizen science count data with sdmTMB"
 output:

### 2021-12-10 doc/paper.Rmd
number appendices and cite them in paper

@@
 We could generate a population index that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
-\@ref(fig:pcod-fig)E, APPENDIX).
+\@ref(fig:pcod-fig)E, Appendix 2).
 
@@
 Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):
+We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3):
 
@@
 We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
-Users familiar with INLA may find inlabru approachable.
+Users familiar with INLA may find inlabru approachable (Appendix 4).
 Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
 VAST is the sole option for fitting some multivariate models;
-alternatively, because VAST focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
+alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
 Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).

### 2021-12-10 doc/appendix-pcod-inlabru.Rmd
number appendices and cite them in paper

@@
 ---
-title: "Annotated example of index standardization with inlabru"
+title: "Appendix 4: example of index standardization of fishery-independent survey data with inlabru"
 output:
@@
   data = pcod,
-  time = "year", spde = mesh,
+  time = "year", mesh = mesh,
   family = binomial(link = "logit"),

### 2021-12-10 doc/paper.Rmd
maureaud2021

@@
 Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
-Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @english2021].
+Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2021].
 INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.

### 2021-12-10 doc/paper.Rmd
Edits

@@
 While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
+<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
 Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
@@
 While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
-Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
+Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
 These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
@@
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
-<!-- User interfaces through software such as VAST or inlabru can greatly reduce these challenges.
-VAST is built on TMB, and is extremely powerful (e.g., fitting many multivariate models not described here), but does not provide a simple, familiar, modular interface for users with basic spatial or spatiotemporal data.
-inlabru provides a somewhat more familiar interface to a large class of models including GRMF models, but is strictly Bayesian, lacks some flexibility for the particular class of models discussed here.
-It also relies on INLA, and at least for some problems, TMB can be much faster [@osgood-zimmerman2021].
-We provide a more detailed comparison of related software packages in Table 1 and the discussion.
--->
 
@@
 
-<!-- -   [@anderson2019] -->
-<!-- -   [@maureaud2021] -->
-<!-- -   [@anderson2020] -->
-<!-- -   [@osgood-zimmerman2021] -->
-<!-- -   [@breivik2021] -->
-<!-- -   [@monnahan2021] -->
-<!-- -   [@shelton2014] -->
-<!-- -   [@lindgren2011] -->
-<!-- -   [@latimer2009] -->
-
-<!-- Packages: -->
-
-<!-- spaMM: @rousset2014 -->
-<!-- spNNGP: @finley2021 -->
-
 <!-- millard1985: Space–Time Correlation and Its Effects on Methods for Detecting Aquatic Ecological Change "In practice, spatial correlation is more likely to be a problem than is temporal correlation, given typical monitoring frequencies." -->
@@
 It is now possible to use these structures and to integrate them into our analyses as fully-fledged controlled variables." -->
-
-<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
 
@@
 There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
-sdmTMB, VAST, INLA/inlabru, and spaMM are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
 sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.

### 2021-12-10 doc/appendix-model.Rmd
Convergence and barrier meshes

@@
 
-TODO
+In some cases the spatial domain of interest may be complex, and bordered by land or water (coastlines, islands, lakes, etc).
+SPDE models allow for barriers to be included, via barrier meshes.
+With  `sdmTMB()` models, the mesh construction occurs in two steps: (1) the user constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
+The barriers must be constructed as `sf` objects with polygons defining the barriers.
+See `?sdmTMB::add_barrier_mesh` for examples.
 
@@
 
-TODO
+Much of the guidance around diagnostics and `glmmTMB` also applies to `sdmTMB`, e.g.
+[https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html).
+Warnings of non-positive-definite Hessian matrices (accompanied by parameters with NAs for standard errors) often mean models are improperly specified given the data used.
+Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`) -- more iterations and smaller tolerance thresholds ensure that the optimal solution is found, but more evaluations translates into longer computation time.
+Additionally, multiple function calls to `stats::nlminb()` may aid convergence; we include `nlmbinb_loops` as an argument to `sdmTMB()`, where each call is initialized at the best parameter vector from the previous iteration.
+Alternatively, `sdmTMB()` may be called repeatedly, using different random seeds.
 

### 2021-12-10 doc/paper.Rmd
a possible way to get von Humboldt in

@@
 <!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
+<!-- While measured variables can be accounted for via predictors in a model (e.g., including latitude, recognized as an important ecological gradient as far back as von Humboldt (1808)), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can still cause residual spatial correlation.
+-->
 Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

### 2021-12-10 doc/paper.Rmd
sf citation?

@@
 An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
+It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`, @pebesma2018).
+<!-- Since this is not particular to your package, maybe the e.g.
+for how to do this should be moved to the appendix?
+I do explain the transformation for the owl data, but it could be added to the pcod appendix too?
+-->
 Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.

### 2021-12-11 doc/paper.Rmd
Editing

--- 2021-12-10 doc/paper.Rmd
@@
 While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
-While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
-Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
+While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
+Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
 
@@
 
-While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
+While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
 Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
 These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
-This can make an enormous difference to computational efficiency [@heaton2017].
+This can greatly improve computational efficiency [@heaton2017].
 The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
@@
 We refer to design matrices as $\bm{X}$.
-The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and 'group', respectively.
+The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
 The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
-All other symbols refer to the  described model components.
+All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
 Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
@@
 dir.create("figs", showWarnings = FALSE)
-f <- here::here("figs", "diagram.pdf")
+f <- here::here("figs", "diagram-2021-12-10.pdf")
 if (!file.exists(f)) {
-  download.file("https://www.dropbox.com/s/j8km3hv8xz9b63k/diagram.pdf?dl=1", 
+  download.file("https://www.dropbox.com/s/lvoem3ahwkftzjy/diagram-2021-12-10.pdf?dl=1", 
     destfile = f)
@@
 
-<!-- or areal/grid data with centroids instead of CAR/SAR models -->
-<!-- add CAR/SAR to table -->
 sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
@@
 sdmTMB can also fit models with areal data if each polygon has an associated centroid.
-<!-- (observations are treated as replicates). -->
-A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range---the distance at which points are effectively independent).
+A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).
 
-The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
+The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
 \@ref(fig:diagram)).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
+See Appendix TODO for a full description of the statistical model.
 
-<!-- highlight where each one excels ... -->
-<!-- no CV built into VAST or priors -->
-<!-- discussion integrates across speed, familiarity and table 1 -->
-<!-- results paragraph on Table 1 -->
-<!-- familiarity is an important word here -->
-<!-- add section on functionality comparison -->
+<!-- Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time, -->
+<!-- the general form of a GLMM is -->
+<!-- $g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$, -->
+<!-- with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
+-->
+<!-- We can partition the random effects $\bm{Z \psi}$ into subcomponents, each of which can optionally be included in the model (Appendix TODO). -->
+<!-- Although, in practice it is rare to include all of these components, the full sdmTMB model is -->
+<!-- $$ -->
+<!-- g(\mu_{\bm{s},t}) = -->
+<!-- \bm{X}_{1,\bm{s},t} \bm{\beta} + -->
+<!-- \bm{\alpha}_g + -->
+<!-- \bm{X}_{2,\bm{s},t} \gamma_t + -->
+<!-- \bm{X}_{3,t} \zeta_{\bm{s}} + -->
+<!-- \omega_{\bm{s}} + -->
+<!-- \epsilon_{\bm{s},t}, -->
+<!-- $$ -->
+<!-- where -->
 
-Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
-the general form of a GLMM is
-$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
-with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
-<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
-For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
-Although in practice it is rare to include all of these components, the full sdmTMB model is
-$$
-g(\mu_{\bm{s},t}) =
-\bm{X}_{1,\bm{s},t} \bm{\beta} +
-\bm{\alpha}_g +
-\bm{X}_{2,\bm{s},t} \gamma_t +
-\bm{X}_{3,t} \zeta_{\bm{s}} +
-\omega_{\bm{s}} +
-\epsilon_{\bm{s},t},
-$$
-where
-
-* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
-* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
-* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
-* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
-* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
-* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
-* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
+<!-- * $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices; -->
+<!-- * $\bm{\beta}$ represents a vector of fixed-effect coefficients; -->
+<!-- * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; -->
+<!-- * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$; -->
+<!-- * $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$; -->
+<!-- * $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and -->
+<!-- * $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field). For the IID case, $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$ (Appendix TODO). -->
 
@@
 
-By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
+By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix TODO).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix TODO).
 
 The sdmTMB model is fit by maximum marginal likelihood.
-Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
+Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
@@
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
+Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
 
@@
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
-Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
+Derived parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
 

### 2021-12-11 doc/appendix-snow-sdmTMB.Rmd

@@
 For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
-For these years, we have retrieved mean annual climate indices from the University of East Anglia's [Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
+For these years, we have retrieved mean annual climate indices from the University of East Anglia's Climatic Research Unit ([ https://crudata.uea.ac.uk/cru/data/nao/](https://crudata.uea.ac.uk/cru/data/nao/)).
 Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance.
@@
 ```{r load-data, cache=FALSE} 
-snow <- readRDS(here::here("owls/SNOW_data.rds"))
+snow <- readRDS(here::here("data/SNOW_data.rds"))
 glimpse(snow)
@@
 Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
-This projection was achieved with the following crs code using `sf::st_transform(crs = Albers)`.
+This projection was achieved with the following crs code using `sf::st_transform(crs = Albers)` [@pebesma2018].
 
@@
 plot(mesh)
-```
-
-```{r plot-mesh}
-
 ```
@@
 
+```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
+saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
+saveRDS(m1, here::here("data/snow_w_main_effect_1_150km.rds"))
+saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
+```
+
 ```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
-m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
-m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
-m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
-m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
+m0 <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
+m1 <- readRDS(here::here("data/snow_w_main_effect_1_150km.rds"))
+m2 <- readRDS(here::here("data/snow_w_main_effect_150km.rds"))
 ```
@@
 
+```{r save-reml-model, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
+saveRDS(m, here::here("data/snow_w_main_effect_150km_reml.rds"))
+```
+
 Print the model to inspect the fit:
 ```{r check, include = FALSE, cache=FALSE, eval=TRUE}
+m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
 m # loading this twice just prevents some excess messages from cluttering the output when printing below
@@
 r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
-saveRDS(r_mcmc, "mcmc_residuals.rds")
+saveRDS(r_mcmc, "data/mcmc_residuals.rds")
 ```
@@
 ```{r stan-dat, echo=FALSE, cache=FALSE}
-r_mcmc <- readRDS(here::here("owls/mcmc_residuals.rds"))
+r_mcmc <- readRDS(here::here("data/mcmc_residuals.rds"))
 ```
@@
 
-We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transformed them to the correct projection.
+We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http//www.naturalearthdata.com/download/10m/physical/ne_10m_lakes.zip) and transformed them to the correct projection.
 
@@
   sf::st_transform(crs = Albers) 
-lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes"),  quiet = TRUE)
 lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes

### 2021-12-11 doc/paper.Rmd

@@
 Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
 

### 2021-12-11 doc/paper.Rmd
Trivial edits

@@
 An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`, @pebesma2018).
+It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
 <!-- Since this is not particular to your package, maybe the e.g.
@@
 
-Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
+<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
 We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method.

### 2021-12-11 doc/appendix-pcod-inlabru.Rmd
start adding detail to inlabru example

@@
 
-This model is an extension of the same vignette using sdmTMB.
+This model will attempt to replicate the model fit in Appendix 1 as an example of how index standardization can also be achieved with inlabru.
+We will start by building an identical mesh using INLA.
+These same functions that are used internally in the sdmTMB `make_mesh()` function.
 
 ```{r mesh, echo = TRUE, cache=FALSE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
+loc_xy <- as.matrix(pcod[, c("X", "Y"), drop = FALSE])
+inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
 
@@
   prior.range = c(10, 0.05), # P(range < 10) = 0.05
-  prior.sigma = c(2, 0.05)
-) # P(sigma > 2) = 0.05
+  prior.sigma = c(2, 0.05) # P(sigma > 2) = 0.05
+) 
 ```
@@
 ```{r plot-mesh, fig.asp=0.9, out.width="4in"}
-plot(mesh)
+plot(spde$mesh)
+```
+
+```{r mesh-sdmTMB, echo = TRUE, cache=FALSE}
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
 ```

### 2021-12-11 doc/paper.Rmd

@@
 We can then fit a model with spatial random fields via the function `sdmTMB()`. 
-We use a penalized smoother for depth as a main effect via `s()` from the mgcv package.
+We use a penalized smoother for depth as a main effect via `s()` from the mgcv package and choose a basis dimension of 5.
 We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
@@
 fit <- sdmTMB(
-  density ~ s(depth),
+  density ~ s(depth, k = 5),
   data = pcod,
@@
 #> Spatiotemporal model fit by ML ['sdmTMB']
-#> Formula: density ~ s(depth)
+#> Formula: density ~ s(depth, k = 5)
 #> Mesh: mesh
@@
 #>             coef.est coef.se
-#> (Intercept)     2.37    0.21
+#> (Intercept)     2.34    0.23
+#> sdepth        -26.96   21.52
 #>
-#> Dispersion parameter: 12.69
+#> Smooth terms:
+#>            Std. Dev.
+#> sds(depth)     34.03
+#>
+#> Dispersion parameter: 12.70
 #> Tweedie p: 1.58
-#> Matern range: 16.39
+#> Matern range: 17.47
 #> Spatial SD: 1.86
-#> ML criterion at convergence: 6402.136
+#> ML criterion at convergence: 6403.637
 ```
 
-The output indicates our model was fit by maximum (marginal) likelihood (`'ML'`).
+The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
 We also see the formula, mesh, fitted data, and family.
-Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
-Internally, `s(depth)` creates random effect spline weights and so is not shown here.
+Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
 
@@
 fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
+  density ~ s(depth, k = 5), family = tweedie(link = "log"), data = pcod, mesh = mesh,
   time = "year", spatial = "off", spatiotemporal = "ar1"

### 2021-12-13 doc/appendix-pcod-inlabru.Rmd
a little more inlabru organization

@@
 
-```{r mesh-sdmTMB, echo = TRUE, cache=FALSE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
-```
-
 There are a couple ways to include the fixed effects with `inlabru`.
@@
 In the first case, we can create a design matrix and add year effects individually to the formula.
-
-We can fit the model first with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.
-
-```{r}
-sdmtmb_fit <- sdmTMB(present ~ -1 + as.factor(year),
-  data = pcod,
-  time = "year", mesh = mesh,
-  family = binomial(link = "logit"),
-  spatiotemporal = "off",
-  priors = sdmTMBpriors(
-    matern_st = pc_matern(
-      range_gt = 10, range_prob = 0.05,
-      sigma_lt = 2, sigma_prob = 0.05
-    )
-  )
-)
-sdmtmb_coefs <- tidy(sdmtmb_fit, effects = "fixed")
-```
 
@@
 
+We can also fit the model with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.
+
+```{r sdmtmb}
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
+
+sdmtmb_fit <- sdmTMB(present ~ -1 + as.factor(year),
+  data = pcod,
+  time = "year", mesh = mesh,
+  family = binomial(link = "logit"),
+  spatiotemporal = "off",
+  priors = sdmTMBpriors(
+    matern_st = pc_matern(
+      range_gt = 10, range_prob = 0.05,
+      sigma_lt = 2, sigma_prob = 0.05
+    )
+  )
+)
+sdmtmb_coefs <- tidy(sdmtmb_fit, effects = "fixed")
+```
+
 Comparing the fixed effects between `sdmTMB` and `inlabru` shows the estimates are correlated, but not perfectly so

### 2021-12-13 doc/appendix-pcod-inlabru.Rmd
inlabru not working with tweedie

@@
 This model will attempt to replicate the model fit in Appendix 1 as an example of how index standardization can also be achieved with inlabru. 
+Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
+The density units are kg/km^2^. 
+This can be converted to tonnes per km2 by dividing by 1000, which can sometimes scale better.
+
+```{r}
+pcod$density_tpkm <- pcod$density/1000
+```
+
 We will start by building an identical mesh using INLA.
+Here, X and Y are coordinates in UTM zone 9.
 These same functions that are used internally in the sdmTMB `make_mesh()` function.
@@
 ```{r fit-pcod, warning=FALSE, message=FALSE}
-m <- lm(present ~ -1 + as.factor(year), data = pcod)
+m <- lm(density_tpkm ~ -1 + as.factor(year), data = pcod)
 model_mat <- model.matrix(m)
@@
 ```{r}
-components <- present ~ -1 + x2003 + x2004 + x2005 + x2007 +
+components <- density_tpkm ~ -1 + x2003 + x2004 + x2005 + x2007 +
   x2009 + x2011 + x2013 + x2015 + x2017 +
-  spatrf(main = coordinates, model = spde)
-# spatrf(main = coordinates,model = spde, group=year, ngroup=ncol(model_mat),
-#       control.group = list(model = "iid"))
+  # spatrf(main = coordinates, model = spde)
+  spatrf(main = coordinates, model = spde, group=year, ngroup=ncol(model_mat),
+      control.group = list(model = "iid"))
 
@@
   components = components,
-  family = "binomial", data = dat
+  family = "tweedie",
+  # family = "binomial",
+  options = bru_options(bru_verbose = TRUE), 
+  data = dat
 )
@@
 
-sdmtmb_fit <- sdmTMB(present ~ -1 + as.factor(year),
+sdmtmb_fit <- sdmTMB(density_tpkm ~ -1 + as.factor(year),
   data = pcod,
   time = "year", mesh = mesh,
-  family = binomial(link = "logit"),
-  spatiotemporal = "off",
-  priors = sdmTMBpriors(
-    matern_st = pc_matern(
-      range_gt = 10, range_prob = 0.05,
-      sigma_lt = 2, sigma_prob = 0.05
-    )
-  )
+  family = tweedie(link = "log"),
+  # family = binomial(link = "logit"),
+  # priors = sdmTMBpriors(
+  #   matern_st = pc_matern(
+  #     range_gt = 10, range_prob = 0.05,
+  #     sigma_lt = 2, sigma_prob = 0.05
+  #   )
+  # ),
+  spatiotemporal = "IID"
 )
@@
 ```{r}
-components <- present ~ 0 +
-  fac(main = year, model = "factor_full") +
-  spatrf(main = coordinates, model = spde)
-
-fit2 <- bru(
-  components = components,
-  family = "binomial", data = dat
-)
-bru_coefs2 <- fit2$summary.random$fac[, c("mean", "sd")]
+# components <- present ~ 0 +
+#   fac(main = year, model = "factor_full") +
+#   spatrf(main = coordinates, model = spde)
+# 
+# fit2 <- bru(
+#   components = components,
+#   family = "binomial", data = dat
+# )
+# bru_coefs2 <- fit2$summary.random$fac[, c("mean", "sd")]
 ```
@@
 ```{r}
-plot(bru_coefs$mean, bru_coefs2$mean)
+# plot(bru_coefs$mean, bru_coefs2$mean)
 ```

### 2021-12-14 doc/appendix-pcod-VAST-tweedie.Rmd
number appendices and cite them in paper
save sdmTMB indices and compare with VAST

@@
 ---
-title: "Annotated example of index standardization with VAST"
+title: "Appendix 5: example of index standardization of fishery-independent survey data with VAST"
 output:
@@
 
-```{r plot-knots, echo = TRUE}
+```{r plot-knots, echo = TRUE, out.width = "6in"}
 knitr::include_graphics(here("doc", "appendix-VAST-tweedie", "Data_and_knots.png"))
@@
 
+```{r compare-w-sdmTMB, out.width = "6in"}
+library(dplyr)
+library(ggplot2)
+vast_i <- read.csv(here("doc/appendix-VAST-tweedie/Index.csv")) %>% 
+  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, se = Std..Error.for.Estimate) %>% select(index, year, est, se)
+sdm_i <- readRDS("sdmTMB-index.rds") %>% mutate(index = "sdmTMB")%>% select(index, year, est, se, lwr,upr)
+both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0 )
+
+ggplot(both_i, aes(year, est / 1000, group = index)) +
+  geom_line(aes(year, est / 1000, colour = index)) +
+    geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000, fill= index), alpha = 0.4) +
+  geom_ribbon(aes(ymin = (est - (se*1.96)) / 1000, ymax = (est + (se*1.96)) / 1000, 
+                  fill= index), alpha = 0.4) + 
+  ggsidekick::theme_sleek() +
+  ylab("Biomass (t)")
+```
+
 This document was built using:

### 2021-12-14 doc/appendix-pcod-sdmTMB.Rmd
number appendices and cite them in paper
save sdmTMB indices and compare with VAST

@@
 ---
-title: "Annotated example of index standardization with sdmTMB"
+title: "Appendix 2: example of index standardization of fishery-independent survey data with sdmTMB"
 output:
@@
 ind <- get_index(pred2)
+saveRDS(ind, "sdmTMB-index.rds")
 ggplot(ind, aes(year, est / 1000)) +
@@
 ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)
+saveRDS(ind, "sdmTMB-index-sim.rds")
 ggplot(ind, aes(year, est / 1000)) +

### 2021-12-14 doc/appendix-pcod-VAST.Rmd
Develop VAST comparison

--- 2021-11-24 doc/appendix-pcod-VAST.Rmd
@@
 
-Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
-Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.
+Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function `fit_model()` and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
+Several options are built in for commonly used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.
 
@@
   RhoConfig = RhoConfig,
-  ObsModel = c(2, 0), # this is for conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
-  bias.correct = FALSE,
+  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
+  bias.correct = TRUE,
   use_anisotropy = FALSE,
@@
 # effort is 1 when using CPUE instead of observed weight as the response:
-pcod$effort <- as.numeric(1)
-
+pcod$effort <- 1
+pcod <- as.data.frame(pcod) # ensure not a tibble
+f <- here("doc", "appendix-VAST", "vast-cache.rds")
+if (!file.exists(f)) {
   fit <- fit_model(
@@
   )
+  saveRDS(fit, file = f)
+} else {
+  fit <- readRDS(f)
+}
 ```
 
-We can look at parameter estimates and their standard errors.
+We can look at parameter estimates and their standard errors. First we see estimates from the binomial component and second we see estimates from the positive Gamma component.
 
@@
 
-While making custom plots of individual elements requires a fair bit of work to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want.
-While here we stick with the default set of plots, where `plot_set = 3`, one can specify different standard plots to make by changing the setting of this argument (see `?FishStatsUtils::plot_maps`).
+sdmTMB does not have built-in delta models; however, we can fit the two components separately and combine the predictions.
+Here, we will fit the binomial presence-absence model and then a model for the positive catches only with a Gamma observation distribution and a log link.
+We will use the VAST-created mesh so the results are comparable.
 
-```{r save-plots, echo = TRUE, warning = FALSE}
+```{r sdmTMB-example, echo = TRUE, cache=TRUE}
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
+  mesh = fit$spatial_list$MeshList$isotropic_mesh)
+m1 <- sdmTMB(present ~ 0 + as.factor(year), 
+  data = pcod, mesh = mesh, family = binomial(link = "logit"),
+  time = "year", spatiotemporal = "iid", spatial = "on")
+m1
+
+dat2 <- subset(pcod, density > 0)
+mesh2 <- make_mesh(dat2, xy_cols = c("X", "Y"),
+  mesh = fit$spatial_list$MeshList$isotropic_mesh)
+m2 <- sdmTMB(density ~ 0 + as.factor(year), 
+  data = dat2, mesh = mesh2, family = Gamma(link = "log"),
+  time = "year", spatiotemporal = "iid", spatial = "on")
+m2
+```
+
+```{r compare-vast-sdmTMB1, echo = FALSE}
+plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft") {
+  s <- vast_model$parameter_estimates$SD
+  vast_est1 <- as.list(s, "Estimate", report = FALSE)
+  vast_est2 <- as.list(s, "Estimate", report = TRUE)
+  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
+  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
+  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
+  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
+  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
+  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
+  years <- sort(unique(pcod$year))
+  lwr_vast <- b_year_vast - 2 * b_year_vast_se
+  upr_vast <- b_year_vast + 2 * b_year_vast_se
+  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
+  segments(years, lwr_vast, years, upr_vast)
+  years <- years + 0.05
+  points(years, sdmTMB_est$b_j)
+  segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
+    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
+    col = "red")
+  legend("topright", legend = c("VAST", "sdmTMB"), 
+    col = c("black", "red"), bty = "n", lty = c(1, 1))
+}
+```
+
+We wrote some custom code to extract comparable parameters (not shown above). Here are the annual mean estimates in link space with 95% confidence intervals for the two components to the delta model:
+
+```{r compare-vast-sdmTMB2, echo = TRUE}
+plot_betas(fit, m1, "beta1_ft")
+plot_betas(fit, m2, "beta2_ft")
+```
+
+While making custom plots of individual elements would require considerable additional code to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want.
+Here we stick with the default set of plots (`plot_set = 3`); however, one can specify different standard plots to make by changing the setting of this argument (see `?FishStatsUtils::plot_maps` and `?FishStatsUtils::plot_results`).
+
+```{r save-plots, echo = TRUE, warning = FALSE, cache=TRUE}
 plot(
@@
 
-```{r plot-knots, echo = TRUE}
+```{r plot-knots, echo = TRUE, out.width="6in"}
 knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
@@
 
-```{r plot-density, echo = TRUE}
+```{r plot-density, echo = TRUE, out.width="6in"}
 knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
@@
 
+We can compare the index we would get using sdmTMB.
+Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models.
+Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws.
+This should be approximately comparable with the bias-corrected index calculated above with VAST.
+
+```{r sdmTMB-sim-index, cache=TRUE, echo=TRUE}
+set.seed(28239)
+p_bin <- predict(m1, newdata = qcs_grid, sims = 1000)
+p_pos <- predict(m2, newdata = qcs_grid, sims = 1000)
+
+p_bin_prob <- plogis(p_bin)
+p_pos_exp <- exp(p_pos)
+p_combined <- log(p_bin_prob * p_pos_exp)
+ind_combined_sims <- get_index_sims(p_combined, area = rep(4, nrow(p_combined)))
+```
+
+Now, we can compare the indices. We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above.
+
+```{r index-compare, message=FALSE, warning=FALSE}
+library(dplyr)
+library(ggplot2)
+vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
+  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
+    se = Std..Error.for.ln.Estimate.) %>% 
+  select(index, year, est, se) %>% 
+  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
+  mutate(upr = exp(log(est) + qnorm(0.975) * se))
+sdm_i <- ind_combined_sims %>% mutate(index = "sdmTMB")
+both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
+ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
+  geom_ribbon(alpha = 0.3) +
+  geom_line() + 
+  ylim(0, max(both_i$upr)) +
+  coord_cartesian(expand = FALSE)
+```
+
 This document was built using:

### 2021-12-14 analysis/sim-test-families.R
Add to family test

@@ -5,7 +5,6 @@ library(sdmTMB)
 library(future)
 theme_set(ggsidekick::theme_sleek())
 plan(multisession, workers = floor(availableCores() / 2))
-# options(future.rng.onMisuse = "ignore")
 
 SEED <- 1
 set.seed(SEED)
@@ -144,6 +143,14 @@ est %>% filter(sigma_O.est > 2)
 est <- est %>% filter(max_gradient < 0.002) %>%
   filter(sigma_O.est < 2)
 
+meds <- est %>% reshape2::melt() %>%
+  right_join(true, by = "variable") %>%
+  filter(!(family == "binomial" & variable == "phi.est")) %>%
+  filter(!(family == "poisson" & variable == "phi.est")) %>%
+  mutate(family_link = paste(family, link)) %>%
+  group_by(family_link, variable) %>%
+  mutate(median_value = median(value))
+
 est %>%
   reshape2::melt() %>%
   right_join(true, by = "variable") %>%
@@ -152,7 +159,9 @@ est %>%
   ggplot(aes(value)) +
   facet_grid(vars(paste(family, link)), vars(variable), scales = "free") +
   geom_histogram(bins = 15) +
-  geom_vline(aes(xintercept = true_value), colour = "red")
+  geom_vline(aes(xintercept = true_value), colour = "red") +
+  geom_vline(aes(xintercept = median_value), colour = "blue", data = meds)
+ggsave("figs/sim-test-families-pars.pdf", width = 11, height = 11)
 
 coverage <- est %>%
   filter(family != "poisson") %>%
@@ -184,6 +193,45 @@ est %>%
   filter(family != "binomial") %>%
   worm_plot(phi.lwr, phi.est, phi.upr, phi, expression(phi)) +
   coord_cartesian(xlim = c(phi-0.2, phi+0.2)) +
-  facet_wrap(~paste(family, link), scales = "free")
+  facet_wrap(~paste(family, link), scales = "free_y", ncol = 5)
+ggsave("figs/sim-test-families-phi.pdf", width = 12, height = 7)
+
+est %>%
+  worm_plot(b1.lwr, b1.est, b1.upr, betas[2], "b1") +
+  facet_wrap(~paste(family, link), scales = "free_y", ncol = 5) +
+  coord_cartesian(xlim = c(0.5, 0.9))
+ggsave("figs/sim-test-families-b1.pdf", width = 12, height = 7)
+
[…]

### 2021-12-14 doc/appendix-validation.Rmd
Add validation appendix

---
title: "Appendix X: sdmTMB validation"
output:
  bookdown::pdf_document2:
    toc: false
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

```{r}
library(sdmTMB)
```

The sdmTMB has a number of validation processes to ensure the code is working as expected. This includes unit testing using the testthat package (REF) embedded within the package. This unit testing validates a large number of model configurations against simulated data and comparable models from other packages. We highlight some model validation in particular in this appendix.

In addition to the validation discussed here, Appendix X contains a comparison with VAST and Appendix Y contains a comparison with INLA/inlabru.

This appendix illustrates a comparison of penalized smoothers with mgcv and a comparison of IID random intercepts with glmmTMB. Additionally, it illustrates some simulation testing of a spatial random field model with the main observation families within the sdmTMB package.

# Penalized smoothers

Here, we compare a number of smoother configurations against mgcv. Throughout the section we will turn off spatial random fields so the models should be identical.

## `s(x)`

First, a simple thin plate regression spline with a single variable. We will use simulated data from the mgcv package.

```{r, echo=TRUE}
set.seed(19203)
# examples from ?mgcv::gam.models
dat <- mgcv::gamSim(3, n = 800)
m_mgcv <- mgcv::gam(y ~ s(x2), data = dat)
p_mgcv <- predict(m_mgcv)
dat$X <- runif(nrow(dat)) # fake
dat$Y <- runif(nrow(dat)) # fake
spde <- make_mesh(dat, c("X", "Y"), cutoff = 0.1)
m <- sdmTMB(y ~ s(x2),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```

We can see the predictions match nearly exactly. Furthermore, we can use the in-development function `sdmTMB::plot_smooth()` to compared the smoothers themselves. Note that the sdmTMB version is showing the marginal effect of a prediction including the intercept whereas the mgcv version is showing only the smooth component. Nonetheless, here the smoothers look very similar except for these scale on the y axis.

```{r, echo=TRUE, fig.asp=1}
par(mfrow = c(2, 1))
plot_smooth(m)
plot(m_mgcv)
```

## `s(x1, by = x2)`

Next, we compare a model with a continuous `by` argument. This lets the `x2` smoother change depending on the value of `x1`.

```{r, echo=TRUE}
set.seed(19203)
# examples from ?mgcv::gam.models
# continuous by example:
dat <- mgcv::gamSim(3, n = 800)
m_mgcv <- mgcv::gam(y ~ s(x2, by = x1), data = dat)
p_mgcv <- predict(m_mgcv)
dat$X <- runif(nrow(dat)) # fake
dat$Y <- runif(nrow(dat)) # fake
spde <- make_mesh(dat, c("X", "Y"), cutoff = 0.1)
m <- sdmTMB(y ~ s(x2, by = x1),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```

Again, the results are nearly identical.

## `s(x, y)`

Next, we compare a bivariate smoother, with near identical predictions.

```{r, echo=TRUE}
m_mgcv <- mgcv::gam(y ~ s(x2, x1), data = dat)
p_mgcv <- predict(m_mgcv)
m <- sdmTMB(y ~ s(x2, x1),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```

# IID random intercepts

Here we compare a model that includes two IID random intercepts against the same model fit via glmmTMB. Again, we will turn off random fields so the models are comparable.

First, simulate some data. The variables `g` and `h` will represent groups g and h. g will have 50 groups. h will have 20 groups.

```{r, echo=TRUE}
predictor_dat <- data.frame(X = runif(1000), Y = runif(1000))
mesh <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), cutoff = 0.1)
sim_dat <- sdmTMB_simulate(
  formula = ~ 1,
  data = predictor_dat,
  mesh = mesh,
  family = gaussian(),
  range = 0.5,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 42,
  B = c(0.5) # B0 = intercept, B1 = a1 slope
)
sim_dat$g <- gl(50, 20)
iid_re_vals_g <- rnorm(50, 0, 0.3)
sim_dat$h <- gl(20, 50)
iid_re_vals_h <- rnorm(20, 0, 0.3)
sim_dat$observed <- sim_dat$observed + iid_re_vals_g[sim_dat$g] + iid_re_vals_h[sim_dat$h]
```

```{r, echo=TRUE}
m <- sdmTMB(observed ~ 1 + (1 | g) + (1 | h), mesh = mesh, data = sim_dat, 
  spatial = "off")

m_glmmTMB <- glmmTMB::glmmTMB(observed ~ 1 + (1 | g) + (1 | h), data = sim_dat)
```

We can extract the random effect estimates and compare them. We find them to be nearly identical:

```{r, echo=TRUE}
sdmTMB_pars <- tidy(m, "ran_pars")
glmmTMB_varcorr <- glmmTMB::VarCorr(m_glmmTMB)

sdmTMB_pars$estimate[sdmTMB_pars$term == "tau_G"][1]
sqrt(as.numeric(glmmTMB_varcorr$cond$g))

sdmTMB_pars$estimate[sdmTMB_pars$term == "tau_G"][2]
sqrt(as.numeric(glmmTMB_varcorr$cond$h))

sdmTMB_re <- as.list(m$sd_report, "Estimate")
glmmTMB_re <- glmmTMB::ranef(m_glmmTMB)$cond
plot(c(glmmTMB_re$g$`(Intercept)`, glmmTMB_re$h$`(Intercept)`), sdmTMB_re$RE,
  col = c(rep(1, 50), rep(2, 20)))
abline(0, 1)
```

The last plot illustrates the IID random effect deviations themselves with glmmTMB on the x-axis and sdmTMB on the y-axis. The 'g' deviations are black and the 'h' ones are red.

# Simulation-testing with various observation families

```{r}
knitr::include_graphics(here::here("figs", "sim-test-families-pars.pdf"))
```

```{r}
knitr::include_graphics(here::here("figs", "sim-test-families-b1.pdf"))
```

```{r}
knitr::include_graphics(here::here("figs", "sim-test-families-phi.pdf"))
```

```{r}
knitr::include_graphics(here::here("figs", "sim-test-families-range.pdf"))
```

```{r}
knitr::include_graphics(here::here("figs", "sim-test-families-sigmaO.pdf"))
```

### 2021-12-15 doc/appendix-pcod-VAST-tweedie.Rmd
change rhoconfig to obtain conventional Tweedie model
Add to Tweedie appendix

--- 2021-12-09 doc/appendix-pcod-VAST-tweedie.Rmd
@@
 
-RhoConfig <- c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
+RhoConfig <- c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
 ```
@@
 pcod <- as.data.frame(pcod) # ensure not a tibble
+f <- here("doc", "appendix-VAST-tweedie", "vast-cache.rds")
+if (!file.exists(f)) {
+  tictoc::tic()
   fit <- fit_model(
@@
   )
+  tictoc::toc()
+  saveRDS(fit, file = f)
+} else {
+  fit <- readRDS(f)
+}
 ```
@@
 
+Fit sdmTMB model and compare:
+
+```{r sdmTMB-example, echo = TRUE, cache=TRUE}
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
+  mesh = fit$spatial_list$MeshList$isotropic_mesh)
+tictoc::tic()
+fit_sdmTMB <- sdmTMB(density ~ 0 + as.factor(year), 
+  data = pcod, mesh = mesh, family = tweedie(link = "log"),
+  time = "year", spatiotemporal = "iid", spatial = "on")
+tictoc::toc()
+
+s1 <- fit$ParHat$beta1_ft[fit$ParHat$beta2_ft != 0]
+s2 <- fit$ParHat$beta2_ft[fit$ParHat$beta2_ft != 0]
+b_VAST <- as.numeric(s1 + s2)
+
+b_sdmTMB <- tidy(fit_sdmTMB)
+
+plot(b_sdmTMB$estimate, b_VAST);abline(0, 1)
+cor(b_sdmTMB$estimate, b_VAST)
+
+s <- fit$parameter_estimates$SD
+vast_est1 <- as.list(s, "Estimate", report = FALSE)
+vast_est2 <- as.list(s, "Estimate", report = TRUE)
+vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
+vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
+
+sdmtmb_est <- tidy(fit_sdmTMB, "ran_pars")
+
+# range
+sdmtmb_est$estimate[sdmtmb_est$term == "range"]
+vast_est2$Range_raw2
+
+# sigma_O
+sdmtmb_est$estimate[sdmtmb_est$term == "sigma_O"]
+vast_est1$L_omega2_z
+
+# sigma_E
+sdmtmb_est$estimate[sdmtmb_est$term == "sigma_E"]
+vast_est1$L_epsilon2_z
+
+# Tweedie p
+sdmtmb_est$estimate[sdmtmb_est$term == "tweedie_p"]
+1 + plogis(fit$ParHat$logSigmaM[1,1] )
+
+# Tweedie dispersion
+as.numeric(exp(fit$ParHat$beta1_ft))[1]
+sdmtmb_est$estimate[sdmtmb_est$term == "phi"]
+```
+
 While making custom plots of individual elements requires a fair bit of work to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want.
@@
 ```{r save-plots, echo = TRUE, message = FALSE, warning = FALSE}
+if (!file.exists(here("doc", "appendix-VAST-tweedie", "Data_and_knots.png")) {
   plot(
@@
   )
+}
 ```
@@
 
+We can compare the index we would get using sdmTMB.
+Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models.
+Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws.
+This should be approximately comparable with the bias-corrected index calculated above with VAST.
+
+```{r sdmTMB-index, cache=TRUE, echo=TRUE}
+pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE, area = 4)
+ind_sdmTMB <- get_index(pred)
+```
+
+Now, we can compare the indices. We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above.
+
+```{r index-compare, message=FALSE, warning=FALSE}
+library(dplyr)
+library(ggplot2)
+vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
+  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
+    se = Std..Error.for.ln.Estimate.) %>% 
+  select(index, year, est, se) %>% 
+  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
+  mutate(upr = exp(log(est) + qnorm(0.975) * se))
+sdm_i <- ind_sdmTMB %>% mutate(index = "sdmTMB")
+both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
+ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
+  geom_ribbon(alpha = 0.3) +
+  geom_line() + 
+  ylim(0, max(both_i$upr)) +
+  coord_cartesian(expand = FALSE) +
+  ylab("Biomass (t)")
+```
+
 This document was built using:

### 2021-12-15 analysis/sim-test-index-delta.R
Sim-test delta

@@ -1,0 +1,193 @@
+library(dplyr)
+library(testthat)
+library(ggplot2)
+library(sdmTMB)
+library(future)
+theme_set(ggsidekick::theme_sleek())
+plan(multisession, workers = floor(availableCores() / 2))
+# options(future.rng.onMisuse = "ignore")
+
+SEED <- 1
+set.seed(SEED)
+N <- 625 # per year
+time_steps <- 20
+sigma_O <- 0
+sigma_E <- 0.6
+phi <- 3
+tweedie_p <- 1.5 # Tweedie p
+.range <- 1
+
+x <- seq(-1, 1, length.out = 25)
+y <- seq(-1, 1, length.out = 25)
+loc <- expand.grid(x = x, y = y)
+x <- rep(loc$x, time_steps)
+y <- rep(loc$y, time_steps)
+loc <- data.frame(x = x, y = y)
+spde <- make_mesh(loc, xy_cols = c("x", "y"), cutoff = 0.2)
+plot(spde)
+
+sim_test_index <- function(i) {
+  betas <- rnorm(time_steps, 1, 1)
+  dat <- data.frame(year = rep(seq_len(time_steps), each = N))
+  s <- sdmTMB_simulate(
+    formula = ~ 0 + as.factor(year),
+    data = dat,
+    mesh = spde,
+    B = betas,
+    phi = phi,
+    range = .range,
+    sigma_O = sigma_O,
+    sigma_E = sigma_E,
+    seed = SEED * i,
+    time = "year",
+    family = tweedie(),
+    tweedie_p = tweedie_p
+  )
+  s <- mutate(s, year = dat$year)
+  s <- mutate(s, x = loc$x)
+  s <- mutate(s, y = loc$y)
+  s_sampled <- s %>% group_by(year) %>% slice_sample(n = 100L)
+
+  if (FALSE) {
+    ggplot(s, aes(x, y, fill = mu)) +
+      geom_raster() +
+      geom_point(aes(size = observed), data = s_sampled, pch = 21, fill = NA) +
+      scale_fill_viridis_c() +
+      scale_size_area() +
+      facet_wrap(vars(year)) +
+      coord_cartesian(expand = FALSE)
+  }
+
+  mesh <- make_mesh(s_sampled, xy_cols = c("x", "y"), cutoff = 0.2)
+  s_sampled$present <- ifelse(s_sampled$observed > 0, 1, 0)
+  m1 <- tryCatch({sdmTMB(
+    data = s_sampled,
[…]

### 2021-12-15 doc/appendix-pcod-VAST-tweedie.Rmd
Finish draft tweedie comparison

@@
 
+Note that `"Beta1" = 3` results in a traditional Tweedie model where the dispersion parameter is held constant across years.
+
 Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
-Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.
+Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `"input_grid" = input_grid` in `fit_model()`.
 
@@
 
-Next we will fit a GLMM (generalized linear mixed effects model).
+Next we will fit the model:
 
@@
 
-Fit sdmTMB model and compare:
+We will fit an equivalent sdmTMB model and compare. We will use the mesh already created in VAST so the models are identical.
 
@@
   mesh = fit$spatial_list$MeshList$isotropic_mesh)
-tictoc::tic()
+
 fit_sdmTMB <- sdmTMB(density ~ 0 + as.factor(year), 
@@
   time = "year", spatiotemporal = "iid", spatial = "on")
-tictoc::toc()
+```
 
+We can compare the various parameter estimates.
+Note that VAST adds the dispersion parameter (`phi` in sdmTMB) in the mean of the Tweedie distribution, so we need to account for that when comparing main-effect estimates.
+
+```{r sdmTMB-VAST-par-compare}
 s1 <- fit$ParHat$beta1_ft[fit$ParHat$beta2_ft != 0]
@@
 
+All estimates are effectively identical.
+
 While making custom plots of individual elements requires a fair bit of work to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want.
@@
 ```{r save-plots, echo = TRUE, message = FALSE, warning = FALSE}
-if (!file.exists(here("doc", "appendix-VAST-tweedie", "Data_and_knots.png")) {
+if (!file.exists(here::here("doc", "appendix-VAST-tweedie", "Data_and_knots.png"))) {
   plot(
@@
 We can compare the index we would get using sdmTMB.
-Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models.
-Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws.
-This should be approximately comparable with the bias-corrected index calculated above with VAST.
+To get and equivalent index (not bias corrected), we predict on the survey grid (`qcs_grid`), tell it the area of each grid cell (4 km^2^), and pass the output to `get_index()`.
 
@@
 
-Now, we can compare the indices. We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above.
+Now, we can compare the indices.
 
@@
 library(ggplot2)
-vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
-  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
-    se = Std..Error.for.ln.Estimate.) %>% 
-  select(index, year, est, se) %>% 
-  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
-  mutate(upr = exp(log(est) + qnorm(0.975) * se))
+vast_i <- data.frame(
+  index = "VAST",
+  log_est = vast_est2$ln_Index_ctl[,,1],
+  se = vast_sd2$ln_Index_ctl[,,1],
+  stringsAsFactors = FALSE
+  )
+no_data <- vast_est1$beta2_ft[1,] == 0
+vast_i <- vast_i[!no_data,]
+vast_i <- vast_i %>%
+  mutate(year = unique(pcod$year)) %>% 
+  mutate(lwr = exp(log_est + qnorm(0.025) * se)) %>% 
+  mutate(upr = exp(log_est + qnorm(0.975) * se)) %>% 
+  mutate(est = exp(log_est))
+
 sdm_i <- ind_sdmTMB %>% mutate(index = "sdmTMB")
@@
   ylab("Biomass (t)")
+
+plot(ind_sdmTMB$est, vast_i$est);abline(0, 1)
+plot(ind_sdmTMB$upr, vast_i$upr);abline(0, 1)
+plot(ind_sdmTMB$lwr, vast_i$lwr);abline(0, 1)
+
+(ind_sdmTMB$est - vast_i$est) / vast_i$est
+(ind_sdmTMB$upr - vast_i$upr) / vast_i$upr
+(ind_sdmTMB$lwr - vast_i$lwr) / vast_i$lwr
 ```
+
+The results are effectively identical. In the ribbon plot above, the lines perfectly cover each other.
 

### 2021-12-15 doc/appendix-pcod-VAST.Rmd
Edit delta-gamma compare

@@
 if (!file.exists(f)) {
+  tictoc::tic()
   fit <- fit_model(
@@
   )
+  tictoc::toc()
   saveRDS(fit, file = f)
@@
   mesh = fit$spatial_list$MeshList$isotropic_mesh)
+tictoc::tic()
 m1 <- sdmTMB(present ~ 0 + as.factor(year), 
@@
   time = "year", spatiotemporal = "iid", spatial = "on")
+tictoc::toc()
 m1
@@
   mesh = fit$spatial_list$MeshList$isotropic_mesh)
+tictoc::tic()
 m2 <- sdmTMB(density ~ 0 + as.factor(year), 
@@
   time = "year", spatiotemporal = "iid", spatial = "on")
+tictoc::toc()
 m2
@@
 ```{r sdmTMB-sim-index, cache=TRUE, echo=TRUE}
+tictoc::tic()
 set.seed(28239)
-p_bin <- predict(m1, newdata = qcs_grid, sims = 1000)
-p_pos <- predict(m2, newdata = qcs_grid, sims = 1000)
+p_bin <- predict(m1, newdata = qcs_grid, sims = 500)
+p_pos <- predict(m2, newdata = qcs_grid, sims = 500)
 
@@
 ind_combined_sims <- get_index_sims(p_combined, area = rep(4, nrow(p_combined)))
+tictoc::toc()
 ```
 
-Now, we can compare the indices. We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above.
+Now, we can compare the indices. 
+<!-- We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above. -->
 
@@
 library(ggplot2)
-vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
+vast_i1 <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
   mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
@@
 
+```{r}
+plot(ind_sdmTMB$est, vast_i$est);abline(0, 1)
+plot(ind_sdmTMB$upr, vast_i$upr);abline(0, 1)
+plot(ind_sdmTMB$lwr, vast_i$lwr);abline(0, 1)
+
+(ind_sdmTMB$est - vast_i$est) / vast_i$est
+(ind_sdmTMB$upr - vast_i$upr) / vast_i$upr
+(ind_sdmTMB$lwr - vast_i$lwr) / vast_i$lwr
+```
+
 This document was built using:

### 2021-12-15 doc/appendix-binomial-inla.Rmd
start draft of bionomial inla example; inlabru not matching inla and sdmtmb

---
title: "Appendix 4: species distribution model comparison between inlabru and sdmTMB"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(INLA)
library(inlabru)
theme_set(theme_minimal())
```

This model will attempt to replicate the model fit in Appendix 1 as an example of how index standardization can also be achieved with inlabru. 
Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
The density units are kg/km^2^. 
<!-- This can be converted to tonnes per km2 by dividing by 1000, which can sometimes scale better. -->

```{r}
d <- pcod %>% filter(year > 2008)
yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
d <- dplyr::left_join(d, yr_lu)
```

We will start by building an identical mesh using INLA.
Here, X and Y are coordinates in UTM zone 9.
These same functions that are used internally in the sdmTMB `make_mesh()` function.

```{r mesh, echo = TRUE, cache=FALSE}
mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 7)

loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 7)

spde <- inla.spde2.pcmatern(
  # mesh = inla_mesh,
  mesh = mesh$mesh,
  prior.range = c(5, 0.05), # P(range < 5) = 0.05
  prior.sigma = c(5, 0.05)  # P(sigma > 5) = 0.05
) 
```

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(spde$mesh)
```

# INLA

```{r}

tictoc::tic("INLA")
nyear <- length(unique(d$year))

iset <- inla.spde.make.index('i', n.spde = spde$n.spde, n.group = nyear)

A <- inla.spde.make.A(mesh = inla_mesh, loc = loc_xy, group = d$i_year)

sdat <- inla.stack(
  data = list(y = d$present),
  A = list(A, 1),
  effects = list(iset, year_factor = as.factor(d$year)),
  tag = 'stdata')
# Pr(cor > 0) = 0.9:

h.spec <- list(theta = list(prior = 'pccor1', param = c(0, 0.9)))

formulae <- y ~ 0 + year_factor + f(i, model = spde, group = i.group,
  control.group = list(model = 'ar1', hyper = h.spec))

# PC prior on the autoreg. param.
# prec.prior <- list(prior = 'pc.prec', param = c(1, 0.01))
m_inla <- inla(formulae,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.fixed = list(expand.factor.strategy = 'inla', mean = 0, prec = 1/(100*100)),
  verbose = FALSE,
  control.inla = list(int.strategy = "eb", strategy = "gaussian"),
  control.compute = list(config = TRUE)
  )

tictoc::toc()

post_inla <- inla.posterior.sample(result = m_inla, n = 400L)

inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
summary(m_inla)
```

# inlabru

*** NOT WORKING YET ***

There are a couple ways to include the fixed effects with `inlabru`. Unfortunately, using factor variables in the standard equation syntax does not work. In the first case, we can create a design matrix and add year effects individually to the formula.

Now we can set up the design matrix for `inlabru`

```{r fit-pcod, warning=FALSE, message=FALSE}
m <- lm(present ~ -1 + as.factor(year), data = d)
model_mat <- model.matrix(m)
colnames(model_mat) <- paste0("x", unique(d$year))
dat <- cbind(d, model_mat)
```

Next we set up the `components` object for the model. This includes a formula with fixed or random effects (here only fixed) and the design of the spatial random field. The name `spatrf` can be anything, but includes the coordinates, spde model, and how the year effects are shared.

```{r}
tictoc::tic("bru")
components <- present ~ 0 + 
  x2009 + x2011 + x2013 + x2015 + x2017 +
  # fac(main = year,  model = "factor_full"
  #   # , mean = 0, prec = 0.001
  # ) +
  # x2009(main=x2009, model="factor_full") +
  # x2011(main=x2011, model="factor_full") +
  # x2013(main=x2013, model="factor_full") +
  # x2015(main=x2015, model="factor_full") +
  # x2017(main=x2017, model="factor_full") +
  spatrf(main = coordinates, model = spde, group=i_year, #ngroup=ncol(model_mat),
      control.group = list(model = 'ar1', hyper = h.spec))

m_bru <- bru(
  components = components,
  family = "binomial",
  options = 
    bru_options(bru_verbose = TRUE,
        control.inla = list(int.strategy = "eb", strategy = "gaussian"),
        control.compute = list(config = TRUE)
  ),
  data = dat
)
tictoc::toc()

bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
summary(m_bru)
```

# sdmTMB

We can also fit the model with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.

```{r sdmtmb}
# mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 7)
tictoc::tic("sdmTMB")
m <- sdmTMB(present ~ 0 + as.factor(year), 
  data = d, 
  mesh = mesh, 
  time = "year",
  spatial = "off", 
  spatiotemporal = "AR1", 
  family = binomial(link = "logit"),
  priors = sdmTMBpriors(
    matern_st = pc_matern(range_gt = 5, range_prob = 0.05, sigma_lt = 5, sigma_prob = 0.05),
    ar1_rho = normal(0, 1),
    b = normal(0, 100)
  ))
tictoc::toc()

post_sdmTMB <- gather_sims(m, n_sims = 2000L)

sdmtmb_coefs <- tidy(m, effects = "fixed")

summary(m)
```

```{r}
tidy(m, conf.int = TRUE)
tidy(m, "ran_pars", conf.int = TRUE)
```

Comparing the fixed effects between `sdmTMB` and `inlabru` shows the estimates are correlated, but not perfectly so

```{r}
# plot(sdmtmb_coefs$estimate, bru_coefs$mean)
plot(sdmtmb_coefs$estimate, inla_coefs$mean)
abline(a = 0, b = 1)
```

```{r}
library(dplyr)
library(ggplot2)
year_factor2011_inla <- unlist(lapply(post_inla, function(s) s$latent["year_factor2017:1", 1]))
year_factor2011_tmb <- filter(post_sdmTMB, .variable == "as.factor.year.2017") %>% pull(.value)

ggplot(bind_rows(data.frame(type = "INLA", post = year_factor2011_inla),
  data.frame(type = "sdmTMB", post = year_factor2011_tmb)), aes(post, colour = type)) +
  geom_density()
```

```{r}
ggplot(post_sdmTMB, aes(.value)) + geom_histogram() + facet_wrap(~.variable, scales = "free")
```

```{r}
x <- inla.smarginal(m_inla$marginals.fixed$year_factor2013)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2013") %>% pull(.value)
g2 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2013")

x <- inla.smarginal(m_inla$marginals.fixed$year_factor2011)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2011") %>% pull(.value)
g1 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2011")

x <- inla.smarginal(m_inla$marginals.fixed$year_factor2015)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2015") %>% pull(.value)
g2015 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2015")

x <- inla.smarginal(m_inla$marginals.fixed$year_factor2017)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2017") %>% pull(.value)
g2017 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2017")

x <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`)
range_post <- filter(post_sdmTMB, .variable == "range") %>% pull(.value)
g3 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = range_post), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 1.5, fill = "blue", alpha = 0.2) +
  ggtitle("Range")

x <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`)
sigma_E_post <- filter(post_sdmTMB, .variable == "sigma_E") %>% pull(.value)
g4 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = sigma_E_post), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Random Field SD")

x <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`)
rho_post <- filter(post_sdmTMB, .variable == "ar1_rho") %>% pull(.value)
g5 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = rho_post), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.02, fill = "blue", alpha = 0.2) +
  ggtitle("AR1 rho")

cowplot::plot_grid(g1, g2, g2015, g2017, g3, g4, g5)

# ------------------

```

```{r}

```

The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`

```{r}
# components <- present ~ 0 +
#   fac(main = year, model = "factor_full") +
#   spatrf(main = coordinates, model = spde)
# 
# fit2 <- bru(
#   components = components,
#   family = "binomial", data = dat
# )
# bru_coefs2 <- fit2$summary.random$fac[, c("mean", "sd")]
```

These estimates are perfectly correlated with the estimates above,

```{r}
# plot(bru_coefs$mean, bru_coefs2$mean)
```

### 2021-12-15 doc/appendix-binomial-inla.Rmd
more consise comparison with inla

@@
 
-This model will attempt to replicate the model fit in Appendix 1 as an example of how index standardization can also be achieved with inlabru. 
+In this example, we fit a presence-absence species distribution model with only the fixed year effects and spatiotemporal random fields using AR1.  
 Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
@@
 
-```{r plot-mesh, fig.asp=0.9, out.width="4in"}
+```{r plot-mesh, fig.asp=0.9, out.width="3in"}
 plot(spde$mesh)
@@
 tictoc::toc()
-
-post_inla <- inla.posterior.sample(result = m_inla, n = 400L)
 
@@
 
-bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
+# bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
 bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
@@
 
+Comparing the spatial random effects between `INLA` and `inlabru` shows us that the latter is not estimating well.
+
+```{r}
+m_inla$summary.hyperpar[, c("mean", "sd")]
+m_bru$summary.hyperpar[, c("mean", "sd")]
+```
+
+As a result the fixed effect estimates are also quite different.
+
+```{r}
+plot(inla_coefs$mean, bru_coefs$mean)
+abline(a = 0, b = 1)
+```
+
 # sdmTMB
 
-We can also fit the model with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.
+We can also fit the model with `sdmTMB`. 
+The priors only differ in that... TODO?
 
@@
 
-post_sdmTMB <- gather_sims(m, n_sims = 2000L)
-
 sdmtmb_coefs <- tidy(m, effects = "fixed")
@@
 
-Comparing the fixed effects between `sdmTMB` and `inlabru` shows the estimates are correlated, but not perfectly so
+Comparing the fixed effects between `sdmTMB` and `inla` shows the estimates are very tightly correlated, but not perfectly so
 
@@
 
-```{r}
-library(dplyr)
-library(ggplot2)
-year_factor2011_inla <- unlist(lapply(post_inla, function(s) s$latent["year_factor2017:1", 1]))
-year_factor2011_tmb <- filter(post_sdmTMB, .variable == "as.factor.year.2017") %>% pull(.value)
+With sdmTMB, we can extract parameter simulations from the joint precision matrix using `spread_sims()` or `gather_sims()` functions.
 
-ggplot(bind_rows(data.frame(type = "INLA", post = year_factor2011_inla),
-  data.frame(type = "sdmTMB", post = year_factor2011_tmb)), aes(post, colour = type)) +
-  geom_density()
+```{r sdm-sims}
+post_sdmTMB <- gather_sims(m, n_sims = 2000L)
+
+post_sdmTMB2 <- post_sdmTMB %>% mutate(variable = factor(`.variable`, 
+                    levels = unique(post_sdmTMB$`.variable`), 
+                    labels = c("2009", "2011", "2013", "2015", "2017", "Range", "AR1 rho", "Random Field SD")))
 ```
 
-```{r}
-ggplot(post_sdmTMB, aes(.value)) + geom_histogram() + facet_wrap(~.variable, scales = "free")
+We then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA. 
+From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
+
+```{r inla-densities}
+# post_inla <- inla.posterior.sample(result = m_inla, n = 400L)
+
+x <- list()
+x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% as_tibble() %>% mutate(variable = "as.factor.year.2009")
+x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% mutate(variable = "as.factor.year.2011") 
+x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% mutate(variable = "as.factor.year.2013") 
+x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% mutate(variable = "as.factor.year.2015") 
+x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% mutate(variable = "as.factor.year.2017") 
+x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% mutate(variable = "range") 
+x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% mutate(variable = "sigma_E") 
+x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% mutate(variable = "ar1_rho") 
+post_inla1 <- do.call(rbind, x)
+
+post_inla2 <- post_inla1 %>% mutate(variable = factor(variable, 
+                    levels = unique(post_inla1$variable), 
+                    labels = c("2009", "2011", "2013", "2015", "2017", "Range", "Random Field SD", "AR1 rho")))
 ```
 
-```{r}
-x <- inla.smarginal(m_inla$marginals.fixed$year_factor2013)
-b <- filter(post_sdmTMB, .variable == "as.factor.year.2013") %>% pull(.value)
-g2 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
-  ggtitle("Year 2013")
-
-x <- inla.smarginal(m_inla$marginals.fixed$year_factor2011)
-b <- filter(post_sdmTMB, .variable == "as.factor.year.2011") %>% pull(.value)
-g1 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
-  ggtitle("Year 2011")
-
-x <- inla.smarginal(m_inla$marginals.fixed$year_factor2015)
-b <- filter(post_sdmTMB, .variable == "as.factor.year.2015") %>% pull(.value)
-g2015 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
-  ggtitle("Year 2015")
-
-x <- inla.smarginal(m_inla$marginals.fixed$year_factor2017)
-b <- filter(post_sdmTMB, .variable == "as.factor.year.2017") %>% pull(.value)
-g2017 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
-  ggtitle("Year 2017")
-
-x <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`)
-range_post <- filter(post_sdmTMB, .variable == "range") %>% pull(.value)
-g3 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = range_post), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 1.5, fill = "blue", alpha = 0.2) +
-  ggtitle("Range")
-
-x <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`)
-sigma_E_post <- filter(post_sdmTMB, .variable == "sigma_E") %>% pull(.value)
-g4 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = sigma_E_post), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
-  ggtitle("Random Field SD")
-
-x <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`)
-rho_post <- filter(post_sdmTMB, .variable == "ar1_rho") %>% pull(.value)
-g5 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
-  geom_histogram(data = data.frame(x = rho_post), mapping = aes(x, after_stat(density)),
-    inherit.aes = FALSE, binwidth = 0.02, fill = "blue", alpha = 0.2) +
-  ggtitle("AR1 rho")
-
-cowplot::plot_grid(g1, g2, g2015, g2017, g3, g4, g5)
-
-# ------------------
-
+```{r compare-plot, out.width="6in"}
+ggplot(post_inla2) + 
+  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), binwidth = 0.1, fill = "blue", alpha = 0.2) + 
+  geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
+  facet_wrap(~variable, scales = "free") +
+  ggsidekick::theme_sleek()
 ```
 
-```{r}
-
-```
-
-The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`
+<!-- The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac` -->
 
@@
 
-These estimates are perfectly correlated with the estimates above,
+<!-- These estimates are perfectly correlated with the estimates above, -->
 

### 2021-12-15 doc/appendix-binomial-inla.Rmd
inlabru still not working

@@
 library(inlabru)
-theme_set(theme_minimal())
+theme_set(ggsidekick::theme_sleek())
 ```
 
-In this example, we fit a presence-absence species distribution model with only the fixed year effects and spatiotemporal random fields using AR1.  
+In this example, we fit a presence-absence species distribution model with fixed effect intercepts for year and spatiotemporal random fields using AR1.  
 Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
@@
 ```{r}
-d <- pcod %>% filter(year > 2008)
+d <- pcod %>% filter(year >= 2011)
 yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
@@
 
-We will start by building an identical mesh using INLA.
+We will start by building an identical meshes using both sdmTMB and INLA.
 Here, X and Y are coordinates in UTM zone 9.
 These same functions that are used internally in the sdmTMB `make_mesh()` function. 
+```{r sdm-mesh, echo = TRUE, cache=FALSE}
+sdm_mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 10)
+```
 
-```{r mesh, echo = TRUE, cache=FALSE}
-mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 7)
+```{r plot-sdm-mesh, fig.asp=0.9, out.width="3in"}
+plot(sdm_mesh)
+```
 
+```{r inla-mesh, echo = TRUE, cache=FALSE}
 loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
-inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 7)
+inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
 
 spde <- inla.spde2.pcmatern(
-  # mesh = inla_mesh,
-  mesh = mesh$mesh,
-  prior.range = c(5, 0.05), # P(range < 5) = 0.05
-  prior.sigma = c(5, 0.05)  # P(sigma > 5) = 0.05
-) 
+  mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
+    prior.range = c(1, 0.05), # P(range < 1) = 0.05
+    prior.sigma = c(10, 0.05))  # P(sigma > 0.1) = 0.05
+
 ```
 
-```{r plot-mesh, fig.asp=0.9, out.width="3in"}
+```{r plot-inla-mesh, fig.asp=0.9, out.width="3in"}
 plot(spde$mesh)
+```
+In this example, a PC-prior is used for the temporal autoregressive parameter, i.e., the autocorrelation parameter. 
+This prior is defines as P(cor>0) = 0.9:
+
+```{r}
+h.spec <- list(rho = list(prior = 'pccor1', param = c(0.7, 0.7)))
 ```
@@
 
-```{r}
+```{r inla}
+tictoc::tic("INLA")
 
-tictoc::tic("INLA")
 nyear <- length(unique(d$year))
@@
   tag = 'stdata')
-# Pr(cor > 0) = 0.9:
-
-h.spec <- list(theta = list(prior = 'pccor1', param = c(0, 0.9)))
 
 formulae <- y ~ 0 + year_factor + f(i, model = spde, group = i.group,
-  control.group = list(model = 'ar1', hyper = h.spec))
+  control.group = list(model = 'ar1'))
+  # control.group = list(model = 'ar1', hyper = h.spec))
 
-# PC prior on the autoreg. param.
-# prec.prior <- list(prior = 'pc.prec', param = c(1, 0.01))
 m_inla <- inla(formulae,
@@
   family = "binomial",
+  control.family = list(link = "logit"),
   control.fixed = list(expand.factor.strategy = 'inla', mean = 0, prec = 1/(100*100)),
@@
 
-```{r fit-pcod, warning=FALSE, message=FALSE}
+```{r echo = TRUE}
 m <- lm(present ~ -1 + as.factor(year), data = d)
@@
 colnames(model_mat) <- paste0("x", unique(d$year))
-dat <- cbind(d, model_mat)
+dat <- cbind(d, model_mat) %>% mutate(x = X, y = Y)
+
+# dat <- dat %>% sf::st_as_sf(coords = c("x", "y"))
+
+coordinates(dat) <- c("x", "y")
+dat <- as(dat, "SpatialPointsDataFrame")
 ```
@@
 
-```{r}
-tictoc::tic("bru")
+```{r eval = F, echo = TRUE}
 components <- present ~ 0 + 
     x2009 + x2011 + x2013 + x2015 + x2017  + 
-  # fac(main = year,  model = "factor_full"
-  #   # , mean = 0, prec = 0.001
-  # ) +
-  # x2009(main=x2009, model="factor_full") +
-  # x2011(main=x2011, model="factor_full") +
-  # x2013(main=x2013, model="factor_full") +
-  # x2015(main=x2015, model="factor_full") +
-  # x2017(main=x2017, model="factor_full") +
-  spatrf(main = coordinates, model = spde, group=i_year, #ngroup=ncol(model_mat),
+    # x2009(main=x2009, model="linear") +
+    # x2011(main=x2011, model="linear") +
+    # x2013(main=x2013, model="linear") +
+    # x2015(main=x2015, model="linear") +
+    # x2017(main=x2017, model="linear") +
+    spatrf(coordinates, model = spde, group = year, ngroup=length(unique(d$year)),
       control.group = list(model = 'ar1', hyper = h.spec))
@@
   options = 
-    bru_options(bru_verbose = TRUE,
+   list(bru_verbose = TRUE, 
+        control.fixed = list(expand.factor.strategy = 'inla'),
+        control.family = list(link = "logit"),
+        control.inla = list(int.strategy = "eb", strategy = "gaussian"),
+        control.compute = list(config = TRUE)
+  ),
+  data = dat
+)
+
+bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
+summary(m_bru)
+```
+
+The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`
+
+```{r}
+tictoc::tic("bru")
+
+components <- present ~ 0 +
+  spatrf(main = coordinates, model = spde, group = dat$year, ngroup=length(unique(dat$year)),
+      # control.group = list(model = 'iid')) +
+      control.group = list(model = 'ar1')) +
+      # control.group = list(model = 'ar1', hyper = h.spec)) +
+  # fac(main = year,  model = "iid",
+  #     hyper = list(prec = list(initial = log(1e-6),
+  #                            fixed = TRUE))
+  # )
+  fac(main = year,  model = "factor_full")
+
+m_bru <- bru(
+  components = components,
+  family = "binomial",
+  options = 
+   list(bru_verbose = TRUE, 
+        control.fixed = list(expand.factor.strategy = 'inla'),
+        control.family = list(link = "logit"),
         control.inla = list(int.strategy = "eb", strategy = "gaussian"),
@@
 
-# bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
-bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
+bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
+summary(m_bru)
+```
+
+```{r}
+m_bru <- bru_rerun(m_bru, options = list())
 summary(m_bru)
@@
   data = d, 
-  mesh = mesh, 
+  mesh = sdm_mesh, 
   time = "year",
@@
   priors = sdmTMBpriors(
-    matern_st = pc_matern(range_gt = 5, range_prob = 0.05, sigma_lt = 5, sigma_prob = 0.05),
+    matern_st = pc_matern(range_gt = 1, range_prob = 0.05, # P(range > 1) = 1-0.05
+                          sigma_lt = 10, sigma_prob = 0.05), # P(sigma < 10) = 1-0.05
     ar1_rho = normal(0, 1),
@@
                     levels = unique(post_sdmTMB$`.variable`), 
-                    labels = c("2009", "2011", "2013", "2015", "2017", "Range", "AR1 rho", "Random Field SD")))
+                    labels = c( "2011", "2013", "2015", "2017", "Range", "AR1 rho", "Random Field SD")))
 ```
@@
 # post_inla <- inla.posterior.sample(result = m_inla, n = 400L)
-
 x <- list()
-x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% as_tibble() %>% mutate(variable = "as.factor.year.2009")
-x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% mutate(variable = "as.factor.year.2011") 
-x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% mutate(variable = "as.factor.year.2013") 
-x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% mutate(variable = "as.factor.year.2015") 
-x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% mutate(variable = "as.factor.year.2017") 
-x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% mutate(variable = "range") 
-x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% mutate(variable = "sigma_E") 
-x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% mutate(variable = "ar1_rho") 
+# x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% as_tibble() %>% 
+#   mutate(variable = "as.factor.year.2009")
+x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% 
+  mutate(variable = "as.factor.year.2011") 
+x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% 
+  mutate(variable = "as.factor.year.2013") 
+x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% 
+  mutate(variable = "as.factor.year.2015") 
+x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% 
+  mutate(variable = "as.factor.year.2017") 
+x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% 
+  mutate(variable = "range") 
+x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% 
+  mutate(variable = "sigma_E") 
+x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% 
+  mutate(variable = "ar1_rho") 
 post_inla1 <- do.call(rbind, x)
@@
                     levels = unique(post_inla1$variable), 
-                    labels = c("2009", "2011", "2013", "2015", "2017", "Range", "Random Field SD", "AR1 rho")))
+                    labels = c( "2011", "2013", "2015", "2017", "Range", "Random Field SD", "AR1 rho")))
 ```
@@
   geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
-  facet_wrap(~variable, scales = "free") +
-  ggsidekick::theme_sleek()
+  facet_wrap(~variable, scales = "free") 
 ```

### 2021-12-16 doc/appendix-binomial-inla.Rmd
inla draft

@@
 ---
-title: "Appendix 4: species distribution model comparison between inlabru and sdmTMB"
+title: "Appendix 4: species distribution model comparison between inla and sdmTMB"
 output:
@@
 In this example, we fit a presence-absence species distribution model with fixed effect intercepts for year and spatiotemporal random fields using AR1. 
+This will be fit with a binomial distribution and a logit link. 
 Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
-The density units are kg/km^2^. 
-<!-- This can be converted to tonnes per km2 by dividing by 1000, which can sometimes scale better. -->
+INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`.
 
@@
 
+In this example, we will use identical priors for the range and random field SD (aka. sigma) for models built in INLA and with sdmTMB.
+While being described different in each package, the numbers amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).
+
+```{r set-priors, echo = TRUE, cache=FALSE}
+range_min <- 5
+range_prob <- 0.05
+sigma_max <- 5
+sigma_prob <- 0.05
+```
+
+In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., the which defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now sdmTMB only takes a mean and SD (e.g., `ar1_rho = normal(0, 1)`) so we will not set a prior for either model.
+
+```{r}
+h.spec <- list(rho = list(prior = 'pccor1', param = c(0, 0.9)))
+```
+
 We will start by building an identical meshes using both sdmTMB and INLA.
@@
 These same functions that are used internally in the sdmTMB `make_mesh()` function.
+
 ```{r sdm-mesh, echo = TRUE, cache=FALSE}
 sdm_mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 10)
-```
-
-```{r plot-sdm-mesh, fig.asp=0.9, out.width="3in"}
-plot(sdm_mesh)
 ```
@@
   mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
-    prior.range = c(1, 0.05), # P(range < 1) = 0.05
-    prior.sigma = c(10, 0.05))  # P(sigma > 0.1) = 0.05
-
+    prior.range = c(range_min, range_prob),
+    prior.sigma = c(sigma_max, sigma_prob)) 
 ```
 
-```{r plot-inla-mesh, fig.asp=0.9, out.width="3in"}
-plot(spde$mesh)
-```
-In this example, a PC-prior is used for the temporal autoregressive parameter, i.e., the autocorrelation parameter. 
-This prior is defines as P(cor>0) = 0.9:
-
-```{r}
-h.spec <- list(rho = list(prior = 'pccor1', param = c(0.7, 0.7)))
+```{r plot-inla-mesh, fig.asp=0.9, out.height="3in"}
+plot(sdm_mesh);title(main = "sdmTMB mesh");plot(spde$mesh)
 ```
@@
 # INLA
+
+Data preparation for INLA models includes a formulae, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
+TODO: explain further, including `control.inla = list(int.strategy = "eb", strategy = "gaussian")` 
+TODO: We use very weak priors on the fixed effects (`mean = 0, prec = 1/(100*100)`)?
 
@@
 
+summary(m_inla)
+# saves fixed effect coefs
 inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
-summary(m_inla)
 ```
 
-# inlabru
+<!-- # inlabru -->
 
-*** NOT WORKING YET ***
+<!-- *** NOT WORKING YET *** -->
 
-There are a couple ways to include the fixed effects with `inlabru`.
+<!-- There are a couple ways to include the fixed effects with `inlabru`.
 Unfortunately, using factor variables in the standard equation syntax does not work.
 In the first case, we can create a design matrix and add year effects individually to the formula.
+-->
 
-Now we can set up the design matrix for `inlabru`
+<!-- Now we can set up the design matrix for `inlabru`   -->
 
-```{r echo = TRUE}
-m <- lm(present ~ -1 + as.factor(year), data = d)
-model_mat <- model.matrix(m)
-colnames(model_mat) <- paste0("x", unique(d$year))
-dat <- cbind(d, model_mat) %>% mutate(x = X, y = Y)
+<!-- ```{r echo = F} -->
+<!-- m <- lm(present ~ -1 + as.factor(year), data = d) -->
+<!-- model_mat <- model.matrix(m) -->
+<!-- colnames(model_mat) <- paste0("x", unique(d$year)) -->
+<!-- dat <- cbind(d, model_mat) %>% mutate(x = X, y = Y) -->
 
-# dat <- dat %>% sf::st_as_sf(coords = c("x", "y"))
+<!-- # dat <- dat %>% sf::st_as_sf(coords = c("x", "y")) -->
 
-coordinates(dat) <- c("x", "y")
-dat <- as(dat, "SpatialPointsDataFrame")
-```
+<!-- coordinates(dat) <- c("x", "y") -->
+<!-- dat <- as(dat, "SpatialPointsDataFrame") -->
+<!-- ``` -->
 
-Next we set up the `components` object for the model.
+<!-- Next we set up the `components` object for the model.
 This includes a formula with fixed or random effects (here only fixed) and the design of the spatial random field.
 The name `spatrf` can be anything, but includes the coordinates, spde model, and how the year effects are shared.
+-->
 
-```{r eval = F, echo = TRUE}
-components <- present ~ 0 + 
-    x2009 + x2011 + x2013 + x2015 + x2017  + 
-    # x2009(main=x2009, model="linear") +
-    # x2011(main=x2011, model="linear") +
-    # x2013(main=x2013, model="linear") +
-    # x2015(main=x2015, model="linear") +
-    # x2017(main=x2017, model="linear") +
-    spatrf(coordinates, model = spde, group = year, ngroup=length(unique(d$year)),
-      control.group = list(model = 'ar1', hyper = h.spec))
+<!-- ```{r eval = F, echo = F} -->
+<!-- components <- present ~ 0 +  -->
+<!--     x2009 + x2011 + x2013 + x2015 + x2017  +  -->
+<!--     # x2009(main=x2009, model="linear") + -->
+<!--     # x2011(main=x2011, model="linear") + -->
+<!--     # x2013(main=x2013, model="linear") + -->
+<!--     # x2015(main=x2015, model="linear") + -->
+<!--     # x2017(main=x2017, model="linear") + -->
+<!--     spatrf(coordinates, model = spde, group = year, ngroup=length(unique(d$year)), -->
+<!--       control.group = list(model = 'ar1', hyper = h.spec)) -->
 
-m_bru <- bru(
-  components = components,
-  family = "binomial",
-  options = 
-   list(bru_verbose = TRUE, 
-        control.fixed = list(expand.factor.strategy = 'inla'),
-        control.family = list(link = "logit"),
-        control.inla = list(int.strategy = "eb", strategy = "gaussian"),
-        control.compute = list(config = TRUE)
-  ),
-  data = dat
-)
+<!-- m_bru <- bru( -->
+<!--   components = components, -->
+<!--   family = "binomial", -->
+<!--   options =  -->
+<!--    list(bru_verbose = TRUE,  -->
+<!--         control.fixed = list(expand.factor.strategy = 'inla'), -->
+<!--         control.family = list(link = "logit"), -->
+<!--         control.inla = list(int.strategy = "eb", strategy = "gaussian"), -->
+<!--         control.compute = list(config = TRUE) -->
+<!--   ), -->
+<!--   data = dat -->
+<!-- ) -->
 
-bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
-summary(m_bru)
-```
+<!-- bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")] -->
+<!-- summary(m_bru) -->
+<!-- ``` -->
 
-The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`
+<!-- The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac` -->
 
-```{r}
-tictoc::tic("bru")
+<!-- ```{r } -->
+<!-- tictoc::tic("bru") -->
 
-components <- present ~ 0 +
-  spatrf(main = coordinates, model = spde, group = dat$year, ngroup=length(unique(dat$year)),
-      # control.group = list(model = 'iid')) +
-      control.group = list(model = 'ar1')) +
-      # control.group = list(model = 'ar1', hyper = h.spec)) +
-  # fac(main = year,  model = "iid",
-  #     hyper = list(prec = list(initial = log(1e-6),
-  #                            fixed = TRUE))
-  # )
-  fac(main = year,  model = "factor_full")
+<!-- components <- present ~ 0 + -->
+<!--   spatrf(main = coordinates, model = spde, group = dat$year, ngroup=length(unique(dat$year)), -->
+<!--       control.group = list(model = 'ar1')) + -->
+<!--       # control.group = list(model = 'ar1', hyper = h.spec)) + -->
+<!--   # fac(main = year,  model = "iid", -->
+<!--   #     hyper = list(prec = list(initial = log(1e-6), -->
+<!--   #                            fixed = TRUE)) -->
+<!--   # ) -->
+<!--   fac(main = year,  model = "factor_full") -->
 
-m_bru <- bru(
-  components = components,
-  family = "binomial",
-  options = 
-   list(bru_verbose = TRUE, 
-        control.fixed = list(expand.factor.strategy = 'inla'),
-        control.family = list(link = "logit"),
-        control.inla = list(int.strategy = "eb", strategy = "gaussian"),
-        control.compute = list(config = TRUE)
-  ),
-  data = dat
-)
-tictoc::toc()
+<!-- m_bru <- bru( -->
+<!--   components = components, -->
+<!--   family = "binomial", -->
+<!--   options =  -->
+<!--    list(bru_verbose = TRUE,  -->
+<!--         control.fixed = list(expand.factor.strategy = 'inla'), -->
+<!--         control.family = list(link = "logit"), -->
+<!--         control.inla = list(int.strategy = "eb", strategy = "gaussian"), -->
+<!--         control.compute = list(config = TRUE) -->
+<!--   ), -->
+<!--   data = dat -->
+<!-- ) -->
+<!-- tictoc::toc() -->
 
-bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
-summary(m_bru)
-```
+<!-- bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")] -->
+<!-- summary(m_bru) -->
+<!-- ``` -->
 
-```{r}
-m_bru <- bru_rerun(m_bru, options = list())
-summary(m_bru)
-```
+<!-- ```{r} -->
+<!-- m_bru <- bru_rerun(m_bru, options = list()) -->
+<!-- summary(m_bru) -->
+<!-- ``` -->
 
-Comparing the spatial random effects between `INLA` and `inlabru` shows us that the latter is not estimating well.
+<!-- Comparing the spatial random effects between `INLA` and `inlabru` shows us that the latter is not estimating well. -->
 
-```{r}
-m_inla$summary.hyperpar[, c("mean", "sd")]
-m_bru$summary.hyperpar[, c("mean", "sd")]
-```
+<!-- ```{r} -->
+<!-- m_inla$summary.hyperpar[, c("mean", "sd")] -->
+<!-- m_bru$summary.hyperpar[, c("mean", "sd")] -->
+<!-- ``` -->
 
-As a result the fixed effect estimates are also quite different.
+<!-- As a result the fixed effect estimates are also quite different.  -->
 
-```{r}
-plot(inla_coefs$mean, bru_coefs$mean)
-abline(a = 0, b = 1)
-```
+<!-- ```{r} -->
+<!-- plot(inla_coefs$mean, bru_coefs$mean) -->
+<!-- abline(a = 0, b = 1) -->
+<!-- ``` -->
 
@@
 
-We can also fit the model with `sdmTMB`. 
-The priors only differ in that... TODO?
+We then fit a very similar model with `sdmTMB()`. 
+We again use very weak priors on the fixed effects.
 
-```{r sdmtmb}
-# mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 7)
+```{r sdmtmb, warning=FALSE, message=FALSE}
 tictoc::tic("sdmTMB")
+
 m <- sdmTMB(present ~ 0 + as.factor(year), 
@@
   priors = sdmTMBpriors(
-    matern_st = pc_matern(range_gt = 1, range_prob = 0.05, # P(range > 1) = 1-0.05
-                          sigma_lt = 10, sigma_prob = 0.05), # P(sigma < 10) = 1-0.05
-    ar1_rho = normal(0, 1),
+    matern_st = pc_matern(
+      range_gt = range_min, range_prob = range_prob, #P(range > 5) = 1-0.05
+      sigma_lt = sigma_max, sigma_prob = sigma_prob #P(sigma < 5) = 1-0.05
+    ),
     b = normal(0, 100)
   ))
+
 tictoc::toc()
-
+summary(m)
+tidy(m, "ran_pars", conf.int = TRUE)
+# saves fixed effect coefs
 sdmtmb_coefs <- tidy(m, effects = "fixed")
-
-summary(m)
 ```
 
-```{r}
-tidy(m, conf.int = TRUE)
-tidy(m, "ran_pars", conf.int = TRUE)
-```
+# Comparison
 
-Comparing the fixed effects between `sdmTMB` and `inla` shows the estimates are very tightly correlated, but not perfectly so
+Comparing the fixed effects between `sdmTMB` and `inla` shows the estimates are very tightly correlated.
 
-```{r}
-# plot(sdmtmb_coefs$estimate, bru_coefs$mean)
+```{r , out.width="4in"}
 plot(sdmtmb_coefs$estimate, inla_coefs$mean)
@@
                     levels = unique(post_sdmTMB$`.variable`), 
-                    labels = c( "2011", "2013", "2015", "2017", "Range", "AR1 rho", "Random Field SD")))
+                    labels = c( "2011", "2013", "2015", "2017", 
+                                "Range", "AR1 rho", "Random Field SD")))
 ```
@@
 From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
+TODO: maybe say something about why not using `inla.posterior.sample(result = m_inla, n = 400L)`?
 
@@
 x <- list()
-# x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% as_tibble() %>% 
-#   mutate(variable = "as.factor.year.2009")
-x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% 
+x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% 
   mutate(variable = "as.factor.year.2011") 
-x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% 
+x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% 
   mutate(variable = "as.factor.year.2013") 
-x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% 
+x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% 
   mutate(variable = "as.factor.year.2015") 
-x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% 
+x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% 
   mutate(variable = "as.factor.year.2017") 
-x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% 
+x[[5]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% 
   mutate(variable = "range") 
-x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% 
+x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% 
   mutate(variable = "sigma_E") 
-x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% 
+x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% 
   mutate(variable = "ar1_rho") 
@@
                     levels = unique(post_inla1$variable), 
-                    labels = c( "2011", "2013", "2015", "2017", "Range", "Random Field SD", "AR1 rho")))
+                    labels = c( "2011", "2013", "2015", "2017", 
+                                "Range", "Random Field SD", "AR1 rho")))
 ```
@@
 ggplot(post_inla2) + 
-  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), binwidth = 0.1, fill = "blue", alpha = 0.2) + 
+  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), bins = 30, 
+                 fill = "blue", alpha = 0.2) + 
   geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
@@
 ```
+# Distribution map
 
-<!-- The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac` -->
+We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to produce a map of occurrence probabilities in each year.
 
-```{r}
-# components <- present ~ 0 +
-#   fac(main = year, model = "factor_full") +
-#   spatrf(main = coordinates, model = spde)
-# 
-# fit2 <- bru(
-#   components = components,
-#   family = "binomial", data = dat
-# )
-# bru_coefs2 <- fit2$summary.random$fac[, c("mean", "sd")]
+```{r sdm-predictions}
+nd <- filter(qcs_grid, year >= 2011)
+pred <- predict(m, newdata = nd)
+
+plot_map <- function(dat, column) {
+ggplot(dat, aes_string("X", "Y", fill = column)) +
+    geom_raster() +
+    facet_wrap(~year) +
+    coord_fixed()
+}
 ```
 
-<!-- These estimates are perfectly correlated with the estimates above, -->
+```{r maps, cache=FALSE, out.width="6in"}
+plot_map(pred, "plogis(est)") +
+  scale_fill_viridis_c() +
+  ggtitle("Prediction (fixed effects + all random effects)")
+```
+Similar predictions are more complicated to extract from INLA, but can be plotted for the original sample locations.
 
-```{r}
-# plot(bru_coefs$mean, bru_coefs2$mean)
+```{r inla-pred, out.width="6in"}
+idx <- inla.stack.index(sdat, 'stdata')$data
+
+d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
+
+ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + facet_wrap(~year) + 
+  scale_colour_viridis_c() +
+  coord_fixed() +
+  ggtitle("Prediction (fixed effects + all random effects)")
 ```
+
+<!-- predict.inla? -->
+
+<!-- ```{r} -->
+<!-- pgrid <- inla.mesh.projector(inla_mesh) -->
+
+<!-- predict(m_inla, pgrid) -->
+<!-- ``` -->
+
+<!-- # ```{r epsilon, cache=FALSE, out.width="6in"} -->
+<!-- # plot_map(pred, "epsilon_st") + -->
+<!-- #   ggtitle("Spatiotemporal random effects only") + -->
+<!-- #   scale_fill_gradient2() -->
+<!-- # ``` -->
+<!-- #  -->
+<!-- # ```{r resids, out.width="3in"} -->
+<!-- # p <- predict(m) -->
+<!-- # p$resid <- residuals(m) -->
+<!-- # qqnorm(p$resid) -->
+<!-- # qqline(p$resid) -->
+<!-- # ``` -->

### 2021-12-16 doc/appendix-binomial-inla.Rmd
a little more refining of inla example

@@
 range_min <- 5
-range_prob <- 0.05
 sigma_max <- 5
-sigma_prob <- 0.05
+prior_prob <- 0.05
 ```
@@
 inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
-
 spde <- inla.spde2.pcmatern(
   mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
-    prior.range = c(range_min, range_prob),
-    prior.sigma = c(sigma_max, sigma_prob)) 
+    prior.range = c(range_min, prior_prob),
+    prior.sigma = c(sigma_max, prior_prob)) 
 ```
 
-```{r plot-inla-mesh, fig.asp=0.9, out.height="3in"}
+```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in"}
 plot(sdm_mesh);title(main = "sdmTMB mesh");plot(spde$mesh)
@@
     matern_st = pc_matern(
-      range_gt = range_min, range_prob = range_prob, #P(range > 5) = 1-0.05
-      sigma_lt = sigma_max, sigma_prob = sigma_prob #P(sigma < 5) = 1-0.05
+      range_gt = range_min, range_prob = prior_prob, #P(range > 5) = 1-0.05
+      sigma_lt = sigma_max, sigma_prob = prior_prob #P(sigma < 5) = 1-0.05
     ),
@@
 x <- list()
-x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% 
-  mutate(variable = "as.factor.year.2011") 
-x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% 
-  mutate(variable = "as.factor.year.2013") 
-x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% 
-  mutate(variable = "as.factor.year.2015") 
-x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% 
-  mutate(variable = "as.factor.year.2017") 
-x[[5]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% 
-  mutate(variable = "range") 
-x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% 
-  mutate(variable = "sigma_E") 
-x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% 
-  mutate(variable = "ar1_rho") 
+x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% 
+  as_tibble() %>% mutate(variable = "as.factor.year.2011") 
+x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% 
+  as_tibble() %>% mutate(variable = "as.factor.year.2013") 
+x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% 
+  as_tibble() %>% mutate(variable = "as.factor.year.2015") 
+x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% 
+  as_tibble() %>% mutate(variable = "as.factor.year.2017") 
+x[[5]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% 
+  as_tibble() %>% mutate(variable = "range") 
+x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% 
+  as_tibble() %>% mutate(variable = "sigma_E") 
+x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% 
+  as_tibble() %>% mutate(variable = "ar1_rho") 
 post_inla1 <- do.call(rbind, x)
@@
 ```
-# Distribution map
 
-We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to produce a map of occurrence probabilities in each year.
+# Annual distribution maps
+
+We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year.
 
@@
 pred <- predict(m, newdata = nd) 
-
-plot_map <- function(dat, column) {
-ggplot(dat, aes_string("X", "Y", fill = column)) +
-    geom_raster() +
-    facet_wrap(~year) +
-    coord_fixed()
-}
 ```
 
-```{r maps, cache=FALSE, out.width="6in"}
-plot_map(pred, "plogis(est)") +
-  scale_fill_viridis_c() +
-  ggtitle("Prediction (fixed effects + all random effects)")
+```{r maps, cache=FALSE, out.width="4in"}
+ggplot(pred) + geom_raster(aes(X, Y, fill = plogis(est))) + 
+  facet_wrap(~year) +   scale_fill_viridis_c() + coord_fixed() 
 ```
+
 Similar predictions are more complicated to extract from INLA, but can be plotted for the original sample locations.
 
-```{r inla-pred, out.width="6in"}
+```{r inla-pred, out.width="4in"}
 idx <- inla.stack.index(sdat, 'stdata')$data
-
 d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
-
-ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + facet_wrap(~year) + 
-  scale_colour_viridis_c() +
-  coord_fixed() +
-  ggtitle("Prediction (fixed effects + all random effects)")
+ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + 
+  facet_wrap(~year) + scale_colour_viridis_c() + coord_fixed() 
 ```

### 2021-12-16 doc/paper.Rmd
add results from appendix 4 and 5 to paper

@@
 There is considerable variability in the available observation likelihoods across packages (Table 1).
+<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
+We provide examples of the syntax and the reproducibility of results achieved when similar models were built using INLA and sdmTMB (Appendix 4) and VAST and sdmTMB (Appendix 5). 
+We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax. <!-- Change this if we manage to solve the inlabru issue  -->
 
@@
 
-We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Appendix 6 TODO, Fig.
 \@ref(fig:timing)).
@@
 We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
-Users familiar with INLA may find inlabru approachable (Appendix 4).
+Users familiar with INLA may find inlabru approachable.
+<!-- ; however, we ran into problems with AR1 spatiotemporal fields in inlabru, so we only provide a comparison with INLA instead (Appendix 4).
+-->
 Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.

### 2021-12-16 doc/appendix-binomial-inla.Rmd
add a couple more years to analysis

@@
 ```{r}
-d <- pcod %>% filter(year >= 2011)
+d <- pcod %>% filter(year >= 2007)
 yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
@@
 
-```{r , out.width="4in"}
+```{r , out.width="5.5in"}
 plot(sdmtmb_coefs$estimate, inla_coefs$mean)
@@
                   levels = unique(post_sdmTMB$`.variable`), 
-                    labels = c( "2011", "2013", "2015", "2017", 
+                  labels = c( "2007", "2009", "2011", "2013", "2015", "2017", 
                                 "Range", "AR1 rho", "Random Field SD")))
@@
 x <- list()
-x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% 
+x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>% 
+  as_tibble() %>% mutate(variable = "as.factor.year.2007") 
+x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% 
+  as_tibble() %>% mutate(variable = "as.factor.year.2009") 
+x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% 
   as_tibble() %>% mutate(variable = "as.factor.year.2011") 
-x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% 
+x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% 
   as_tibble() %>% mutate(variable = "as.factor.year.2013") 
-x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% 
+x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% 
   as_tibble() %>% mutate(variable = "as.factor.year.2015") 
-x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% 
+x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% 
   as_tibble() %>% mutate(variable = "as.factor.year.2017") 
-x[[5]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% 
+x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% 
   as_tibble() %>% mutate(variable = "range") 
-x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% 
+x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% 
   as_tibble() %>% mutate(variable = "sigma_E") 
-x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% 
+x[[9]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% 
   as_tibble() %>% mutate(variable = "ar1_rho") 
@@
                   levels = unique(post_inla1$variable), 
-                    labels = c( "2011", "2013", "2015", "2017", 
+                  labels = c( "2007", "2009", "2011", "2013", "2015", "2017", 
                                 "Range", "Random Field SD", "AR1 rho")))
@@
 ```{r sdm-predictions}
-nd <- filter(qcs_grid, year >= 2011)
+nd <- filter(qcs_grid, year >= 2007)
 pred <- predict(m, newdata = nd) 
@@
 
-```{r maps, cache=FALSE, out.width="4in"}
+```{r maps, cache=FALSE, out.width="6in"}
 ggplot(pred) + geom_raster(aes(X, Y, fill = plogis(est))) + 
@@
 
-```{r inla-pred, out.width="4in"}
+```{r inla-pred, out.width="6in"}
 idx <- inla.stack.index(sdat, 'stdata')$data

### 2021-12-16 doc/appendix-binomial-inla.Rmd
minor edit
remove inlabru

@@
 ---
-title: "Appendix 4: species distribution model comparison between inla and sdmTMB"
+title: "Appendix 4: species distribution model comparison between INLA and sdmTMB"
 output:
@@
 library(INLA)
-library(inlabru)
+# library(inlabru)
 theme_set(ggsidekick::theme_sleek())
@@
 
-Similar predictions are more complicated to extract from INLA, but can be plotted for the original sample locations.
+Similar predictions are more complicated to extract from INLA, but are easy to plot for the original sample locations.
 

### 2021-12-17 doc/paper.Rmd
update citation for english2022

@@
 Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
-Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2021].
+Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
 INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.
@@
 
-Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2021] and improve ecological management [@sofaer2019].
+Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
 These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.

### 2021-12-21 doc/appendix-model.Rmd
Add to model appendix

@@
 
+* Greek symbols for parameters
+
 * Bold lowercase for vectors
-
-* Bold subscript $\s$ (a point in space) since it represents a vector of length 2
 
@@
 
-* $\phi$ for all 'dispersion' parameters for consistency with the code
+* $\phi$ for all distribution dispersion parameters for consistency with the code
 
@@
 
-As described in the main paper, the complete sdmTMB model can be written as:
+The complete sdmTMB model can be written as:
 
@@
 
-* $y$ represents the response data;
+* $\mathbb{E}$ indicates expected value;
+* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
 * $\mu$ represents the mean;
@@
 
-We now split this model apart to describe the various parts in more detail.
+Next, we will split this model apart to describe the various parts in more detail using $\ldots$ to represent the other optional components.
 
@@
 
-Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
+Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
 Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. 
 Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g.
-2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; the basis function dimensions may be specified, e.g.
+2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g.
 `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g.
@@
 The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed-effects modelling framework.
-This is a similar approach implemented in `gamm4` and `brms`, among other packages.
+This is the same approach as is implemented in the R packages gamm4 (REF) and brms (REF).
 
-### Linear breakpoint threshold models
+### Linear break-point threshold models
 
-The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses.
+The linear break-point or "hockey stick" model can be used to describe threshold or asymptotic responses. 
 This function consists of two pieces, so that for $x < b_{1}$, $s(x) = x \cdot b_{0}$, and for $x > b_{1}$, $s(x) = b_{1} \cdot b_{0}$.
@@
 where $s$ represents the logistic function, 
-$\psi$ is a scaling parameter (controlling the height of the y-axis for
-the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
-a parameter controlling the point at which the function reaches 50% of the
-maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
-function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
-constrained to be larger than $s50$.
+$\psi$ is a scaling parameter (controlling the height of the y-axis for the response; unconstrained), 
+$\tau$ is an intercept, 
+$s50$ is a parameter controlling the point at which the function reaches 50% of the maximum ($\psi$), 
+and $s95$ is a parameter controlling the point at which the function reaches 95%. 
+The parameter $s50$ is unconstrained but $s95$ is constrained to be larger than $s50$.
 
@@
 $$
+The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
 
@@
 Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
+These text values are case insensitive.
+
+Spatiotemporal random fields are represented by $\boldsymbol{\epsilon}_t$ within sdmTMB. 
+This has been chosen to match the representation in VAST (REF). 
+The marginal standard deviation of $\boldsymbol{\epsilon}_t$ is indicated by `Spatiotemporal SD` in the model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
 
@@
 
-### AR1 spatiotemporal random fields
+### AR(1) spatiotemporal random fields
 
-AR1 spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. The are defined as
+AR(1) spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. 
+They are defined as
 
@@
 The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
+Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.
 
@@
 where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.
-Note that, in contrast to the AR1 parametrization, the variance is no longer the steady-state marginal variance.
+Note that, in contrast to the AR(1) parametrization, the variance is no longer the steady-state marginal variance.
 
@@
 where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
-Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant for a given time $t$.
-Currently, $\bm{X}_{3,t}$ is restricted to a single covariate, supplied by a formula to `spatial_varying`.
+Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant spatially for a given time $t$.
+Currently, $\bm{X}_{3,t}$ is restricted to a single covariate, supplied by a one-sided formula to `spatial_varying`.
+For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.
 
@@
 
-Internally parameterized as the [robust
-version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5).
+$$
+\operatorname{Binomial} \left(N, \mu \right)
+$$
+where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
+If $N = 1$, the distribution becomes the Bernoulli distribution.
+Internally, the distribution is parameterized as the [robust
+version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
+
+Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:
+
+1. the response may be a factor (and the model classifies the first level versus all others)
+2. the response may be binomial (0/1)
+3. the response can be a matrix of form `cbind(success, failure)`, or
+4. the response may be the observed proportions, and the `weights` argument is used to specify the Binomial size ($N$) parameter (`probabilty ~ ..., weights = N`).
+
+Example: `family = binomial(link = "logit")`
 
@@
 $$
-\operatorname{Beta} \left(\mu \phi, 1 - \mu \phi \right)
+\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
 $$
 
-where $\phi$ is variance.
+where $\mu$ is the mean and $\phi$ is a precision parameter.
+This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
+The variance is $\mu (1 - \mu) / (\phi + 1)$.
 
-$$
-\operatorname{Binomial} \left( N, \mu \right)
-$$
-
-where $N$ is the size or number of trials and $\mu$ is the probability.
-
-Following the structure of `stats::glm()` and glmmTMB, a binomial family can be specified in one of 4 ways:
-
-1. the response may be a factor (and the model classifies the first level versus all others)
-2. the response may be binomial (0/1)
-3. the response can be a matrix of form `cbind(success, failure)`, and 
-4. the response may be the observed proportions, and the `weights` argument is used to specify the Binomial size ($N$) parameter (`prob ~ ..., weights = N`).
+Example: `family = Beta(link = "logit")`
 
 ## [Gamma](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gab0e2205710a698ad6a0ed39e0652c9a3)
-
-As shape, scale:
 
@@
 
-where $\phi$ represents the shape and $\frac{\mu}{\phi}$ represents the scale.
+where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
+The variance is $\mu \cdot \phi$.
+
+Example: `family = Gamma(link = "log")`
 
@@
 $$
-where $\mu$ is the mean and $\phi$ is the standard deviation (following Stan convention of SD not variance).
+where $\mu$ is the mean and $\phi$ is the standard deviation (following Stan convention of standard deviation not variance).
+The variance is $\phi^2$.
+
+Example: `family = Gaussian(link = "identity")`
 
@@
 where $\mu$ and $\phi$ are the mean and standard deviation of the natural log of the random variable. 
-If $X$ is lognormally distributed, $E[log(X)] = \mu$ and $Var(log(X)) = \phi$.
-The expectation of $X$ is $exp(u + \frac{\phi^2}{2})$ and variance of X is $\left(exp(\phi^2) - 1 \right)exp(2\mu + \phi^2)$.
+If $\boldsymbol{x}$ is lognormally distributed, $\mathbb{E}[\log(\boldsymbol{x})] = \mu$ and $\mathrm{Var}[\log(\boldsymbol{x})] = \phi$.
+This translates to $\mathbb{E}[\boldsymbol{x}] = \exp(\mu + \phi^2 / 2)$ and $\mathrm{Var}[\boldsymbol{x}] = \left(\exp(\phi^2) - 1 \right) \cdot \exp(2 \mu + \phi^2)$.
+
+TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
+
+Example: `family = lognormal(link = "log")`
 
@@
 
-Where $\mu$ is the mean and $\phi$ is the dispersion parameter.
-The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.
+where $\mu$ is the mean and $\phi$ is the dispersion parameter.
+The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].
 
 Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
+
+Example: `family = nbinom2(link = "log")`
 
@@
 $$
-Where $\mu$ is the mean and $\phi$ is the dispersion parameter.
-The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$.
+
+where $\mu$ is the mean and $\phi$ is the dispersion parameter.
+The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$ [@hilbe2011].
 
 Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
+
+Example: `family = nbinom1(link = "log")`
 
@@
 $$
+where $\mu$ represents the mean and $\mathrm{Var}[y] = \mu$.
 
-## [Censored-poisson] TODO
+Example: `family = poisson(link = "log")`
 
@@
 
-where $\nu$, the degrees of freedom, is currently fixed at 3.
+where $\nu$, the degrees of freedom (`df`), is a user-supplied fixed parameter.
+Lower values of $\nu$ result in heavier tails compared to the Gaussian distribution.
+Above approximately `df = 20`, the distribution becomes very similar to the Gaussian.
+The Student-t distribution with a low degrees of freedom (e.g., $\nu \le 7$) can be helpful for modelling data that would otherwise be suitable for Gaussian but needs an approach that is robust to outliers [e.g., @anderson2017c].
+
+Example: `family = student(link = "log", df = 7)`
 
@@
 
-where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 1, and $\phi$ is the dispersion parameter.
+where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
+The Tweedie distribution can be helpful for modelling data that positive and continuous but also contains zeros.
+
+Internally, $p$ is transformed from $\mathrm{logit}^{-1} (\texttt{thetaf}) + 1$ to constrain it between 1 and 2 and \texttt{thetaf} is fit as an unconstrained variable.
 
@@
 
-TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longtidue ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)).
-Anistropy can be turned on and off with the boolean `anisotropy` argument to `sdmTMB()`.
+TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)).
+Anisotropy can be turned on and off with the boolean `anisotropy` argument to `sdmTMB()`.
 There are a number of ways to implement anisotropic covariance (Fuglstad et al.
@@
 Once a model is fitted with `sdmTMB()`, the anisotropy relationships may be plotted using the `plot_anisotropy()` function, which takes the fitted object as an argument.
-Note: if you use a barrier mesh, anisotropy is disabled.
+If a barrier mesh is used, anisotropy is disabled.
 
@@
 
-In some cases the spatial domain of interest may be complex, and bordered by land or water (coastlines, islands, lakes, etc).
-SPDE models allow for barriers to be included, via barrier meshes.
-With  `sdmTMB()` models, the mesh construction occurs in two steps: (1) the user constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
+In some cases the spatial domain of interest may be complex and bordered by land or water (e.g., coastlines, islands, lakes).
+SPDE models allow for barriers to be included via barrier meshes.
+With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
 The barriers must be constructed as `sf` objects with polygons defining the barriers.
 See `?sdmTMB::add_barrier_mesh` for examples.
+
+TODO DETAILS ON  METHOD HERE
 
@@
 
-Like AD Model Builder, TMB allows for parameters to be fit in phases and we inlcude the `multiphase` argument in `sdmTMB::sdmTMBcontrol()` to allow this.
+Like AD Model Builder, TMB allows for parameters to be fit in phases and we include the `multiphase` argument in `sdmTMB::sdmTMBcontrol()` to allow this.
 For high dimensional models (lots of fixed and random effects), phased estimation may be faster and result in more stable convergence.

### 2021-12-21 analysis/sim-test-index-delta.R
Expand delta sim testing

@@ -4,7 +4,7 @@ library(ggplot2)
 library(sdmTMB)
 library(future)
 theme_set(ggsidekick::theme_sleek())
-plan(multisession, workers = floor(availableCores() / 2))
+plan(multisession, workers = floor(availableCores() / 2) - 0)
 # options(future.rng.onMisuse = "ignore")
 
 SEED <- 1
@@ -13,20 +13,24 @@ N <- 625 # per year
 time_steps <- 20
 sigma_O <- 0
 sigma_E <- 0.6
-phi <- 3
+phi <- 9
 tweedie_p <- 1.5 # Tweedie p
-.range <- 1
+.range <- 70
 
-x <- seq(-1, 1, length.out = 25)
-y <- seq(-1, 1, length.out = 25)
+# x <- seq(-1, 1, length.out = 25)
+# y <- seq(-1, 1, length.out = 25)
+x <- seq(340, 580, length.out = 25)
+y <- seq(5500, 5800, length.out = 25)
 loc <- expand.grid(x = x, y = y)
 x <- rep(loc$x, time_steps)
 y <- rep(loc$y, time_steps)
 loc <- data.frame(x = x, y = y)
-spde <- make_mesh(loc, xy_cols = c("x", "y"), cutoff = 0.2)
+spde <- sdmTMB::make_mesh(loc, xy_cols = c("x", "y"), cutoff = 10)
 plot(spde)
+plot(spde$mesh, asp = 1)
 
 sim_test_index <- function(i) {
+  set.seed(i * 1029)
   betas <- rnorm(time_steps, 1, 1)
   dat <- data.frame(year = rep(seq_len(time_steps), each = N))
   s <- sdmTMB_simulate(
@@ -46,7 +50,7 @@ sim_test_index <- function(i) {
   s <- mutate(s, year = dat$year)
   s <- mutate(s, x = loc$x)
   s <- mutate(s, y = loc$y)
-  s_sampled <- s %>% group_by(year) %>% slice_sample(n = 100L)
+  s_sampled <- s %>% group_by(year) %>% slice_sample(n = 200L)
 
   if (FALSE) {
     ggplot(s, aes(x, y, fill = mu)) +
@@ -58,7 +62,87 @@ sim_test_index <- function(i) {
       coord_cartesian(expand = FALSE)
   }
 
-  mesh <- make_mesh(s_sampled, xy_cols = c("x", "y"), cutoff = 0.2)
[…]

### 2021-12-21 doc/appendix-model.Rmd
Add to appendix

@@
 $$
-\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),$$
+\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),
+$$
 <!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
@@
 
-## Projection 'A' matrix
+Internally, the random fields are assumed to follow a Gaussian Markov random field. For example, the spatial random field case:
 
-TODO
+$$
+\boldsymbol{\omega} \sim \mathrm{MVN}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}\right)
+$$
+where $\boldsymbol{Q}$ is a sparse precision matrix.
+
+## Projection '$\boldsymbol{A}$' matrix
+
+The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates the values at the locations of the observed or predicted data [@lindgren2015]:
+
+$$
+\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega}
+$$
+The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
+Three non-zero elements on each row define the weight of the neighbouring 3 knot locations.
+The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$.
 
@@
 Anisotropy can be turned on and off with the boolean `anisotropy` argument to `sdmTMB()`.
-There are a number of ways to implement anisotropic covariance (Fuglstad et al.
-2015), and we adopt a 2-parameter rotation matrix $\textbf{H}$.
+There are a number of ways to implement anisotropic covariance REF, and we adopt a 2-parameter rotation matrix $\textbf{H}$.
 The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$
@@
 With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`. 
-The barriers must be constructed as `sf` objects with polygons defining the barriers.
+The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers. 
 See `?sdmTMB::add_barrier_mesh` for examples.

### 2021-12-21 doc/appendix-model.Rmd
Add to model appendix

@@
 
+Internally, the random fields follow a Gaussian Markov random field (GMRF):
+
+$$
+\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
+$$
+where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
+
 ## Spatiotemporal random fields
@@
 
+Similarly to the spatial random fields, these spatiotemporal random fields (and all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):
+
+$$
+\boldsymbol{\epsilon}_t \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
+$$
+
 ### AR(1) spatiotemporal random fields
 
-AR(1) spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. 
+First-order auto regressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. 
 They are defined as
@@
 where $\rho$ is the correlation between subsequent spatiotemporal random fields. 
-The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
+The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ term scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
 The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
-Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.
+Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. 
+The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.
 
@@
 
+Here we describe the main observation families that are available in sdmTMB and comment on their parametrization, statistical properties, utility, and code representation in sdmTMB.
+
 ## [Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaee11f805f02bc1febc6d7bf0487671be)
@@
 $$
-\operatorname{Normal} \left( \mu, \phi \right)
+\operatorname{Normal} \left( \mu, \phi^2 \right)
 $$
-where $\mu$ is the mean and $\phi$ is the standard deviation (following Stan convention of standard deviation not variance).
+where $\mu$ is the mean and $\phi$ is the standard deviation.
 The variance is $\phi^2$.
@@
 The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].
-
-Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
+Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
 
@@
 The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$ [@hilbe2011].
-
-Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
+Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
 
@@
 
-Internally, $p$ is transformed from $\mathrm{logit}^{-1} (\texttt{thetaf}) + 1$ to constrain it between 1 and 2 and \texttt{thetaf} is fit as an unconstrained variable.
+Internally, $p$ is transformed from $\mathrm{logit}^{-1} (\texttt{thetaf}) + 1$ to constrain it between 1 and 2 and \texttt{thetaf} is estimated as an unconstrained variable.
+
+Example: `family = tweedie(link = "log")`
 
@@
 
-where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
-The parameter $\nu$ controls the smoothness of the covariance function.
-We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
-
-Internally, the random fields are assumed to follow a Gaussian Markov random field. For example, the spatial random field case:
-
-$$
-\boldsymbol{\omega} \sim \mathrm{MVN}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}\right)
-$$
-where $\boldsymbol{Q}$ is a sparse precision matrix.
+where $\tau^2$ controls the spatial variance,
+$\nu$ controls the smoothness,
+$\Gamma$ represents the Gamma function,
+$d_{jk}$ represents the distance between locations $s_j$ and $s_k$,
+$K_\nu$ represents the modified Bessel function of the second kind,
+and $\kappa$ controls the spatial correlation range.
+The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
 
@@
 $$
-\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega}
+\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega},
 $$
+where $\boldsymbol{\omega}^*$ represents the values of the spatial random fields at the observed locations or predicted data locations.
 The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
-Three non-zero elements on each row define the weight of the neighbouring 3 knot locations.
-The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$.
+Three non-zero elements on each row define the weight of the neighbouring 3 knot locations for location $s$.
+The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$:
+
+$$
+\boldsymbol{\epsilon}_t^* = \boldsymbol{A} \boldsymbol{\epsilon}_t.
+$$
 
@@
 TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)).
-Anisotropy can be turned on and off with the boolean `anisotropy` argument to `sdmTMB()`.
+Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`.
 There are a number of ways to implement anisotropic covariance REF, and we adopt a 2-parameter rotation matrix $\textbf{H}$.
@@
 In some cases the spatial domain of interest may be complex and bordered by land or water (e.g., coastlines, islands, lakes). 
-SPDE models allow for barriers to be included via barrier meshes. 
+SPDE models allow for barriers to be included via barrier meshes [@bakka2019]. 
 With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`. 
@@
 
-TODO DETAILS ON  METHOD HERE
+The barrier mesh implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019]. 
+For example, if the range was estimated at 10 km, `range_fraction = 0.2` would assume that the range was 2 km across the barrier. 
+This would let the spatial correlation decay 5 times faster with distance. 
+From experimentation, values around 0.1 or 0.2 seem to work well but values much lower than 0.1 can result in convergence issues.
+
+The website <https://haakonbakkagit.github.io/btopic128.html> provides an illustration with INLA.
+The implementation within TMB was borrowed from code written by Olav Nikolai Breivik and Hans Skaug at <https://github.com/skaug/tmb-case-studies>.
 
@@
 
+# Optimization
+
+The sdmTMB model is fit by maximum marginal likelihood.
+Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
+Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
+After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+
+Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`)---a greater number of iterations and smaller tolerance thresholds increase the chance that the optimal solution is found, but more evaluations translates into longer computation time.
+Additionally, multiple function calls to `stats::nlminb()` may aid convergence; we include `nlminb_loops` as an argument to the control list in `sdmTMB()` (e.g., `sdmTMB(..., control = sdmTMBcontrol(nlmbinb_loops = 2))`), where each call is initialized at the best parameter vector from the previous iteration.
+Similarly, additional optimization with a Newton solver (`stats::optimHess()`) can be implemented by adding `newton_loops` in the control list.
+If a model is already fit, the function `sdmTMB::run_extra_optimization()` can run additional optimization loops with either routine to further reduce the maximum gradient.
+
 # Assessing convergence
 
-Much of the guidance around diagnostics and `glmmTMB` also applies to `sdmTMB`, e.g.
+Much of the guidance around diagnostics and glmmTMB also applies to `sdmTMB`, e.g.
 [https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html).
-Warnings of non-positive-definite Hessian matrices (accompanied by parameters with NAs for standard errors) often mean models are improperly specified given the data used.
-Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`) -- more iterations and smaller tolerance thresholds ensure that the optimal solution is found, but more evaluations translates into longer computation time.
-Additionally, multiple function calls to `stats::nlminb()` may aid convergence; we include `nlmbinb_loops` as an argument to `sdmTMB()`, where each call is initialized at the best parameter vector from the previous iteration.
-Alternatively, `sdmTMB()` may be called repeatedly, using different random seeds.
+Warnings of non-positive-definite Hessian matrices (accompanied by parameters with `NA`s for standard errors) often mean models are improperly specified given the data.
+Standard errors can be observed in the output of `print.sdmTMB()` or by checking `fit$sd_report`.
+The maximum gradient of the marginal likelihood with respect to fixed-effect parameters can be checked by inspecting (`fit$gradients`).
+Guidance varies, but the maximum gradient should likely be at least $< 0.001$ before assuming the fitting routine is consistent with convergence.
+If maximum gradients are already relatively small, they can often be reduced further with additional optimization calls beginning at the previous best parameter vector as described above with `sdmTMB::run_extra_optimization()`.
 

### 2021-12-21 doc/appendix-model.Rmd
Add to model description

@@
 fig_caption: true
+urlcolor: blue
 highlight: "monochrome"
@@
 
-* Bold lowercase for vectors
-
-* Bold uppercase for matrices
+* Bold symbols for vectors or matrices
 
 * $\phi$ for all distribution dispersion parameters for consistency with the code
+
+* $\mathbb{E}[y]$ to define the expected value (mean) of variable $y$
+
+* $\mathrm{Var}[y]$ to define the expected variance of the variable $y$
+
+* An $^*$ superscript to represent interpolated or projected values as opposed to values at knot locations (e.g., $\omega$ vs. $\omega^*$).
 
@@
   ~Symbol, ~Description,
+  "$y$", "Observed response data",
+  "$\\mu$", "Mean",
   "$\\phi$", "A dispersion parameter for a distribution",
-  "$\\mu$", "Mean",
-  "$y$", "Observed response data",
   "$f$", "Link function",
-  "$\\omega_{\\s}$", "Spatial random field at point $\\s$",
-  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$",
-  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field",
+  "$f^{-1}$", "Inverse link function",
+  "$\\boldsymbol{\\beta}$", "Parameter vector",
+  "$\\boldsymbol{X}$", "A predictor model matrix",
+  "$O_{\\s, t}$", "An offset variable at point $\\s$ and time $t$",
+
+  "$\\omega_{\\s}$", "Spatial random field at point $\\s$ (knot)",
+  "$\\omega_{\\s}^*$", "Spatial random field at point $\\s$ (interpolated)",
+  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$ (knot)",
+  "$\\zeta_{\\s, t}^*$", "Spatially varying coefficient random field at point $\\s$ and time $t$ (interpolated)",
+  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
+  "$\\epsilon_{\\s, t}^*$", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
+  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations (knot)",
+  "$\\alpha_g$", "IID random intercept deviation for group $g$",
+
   "$\\boldsymbol{\\Sigma}_\\omega$", "Spatial random field covariance matrix",
+  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
   "$\\boldsymbol{\\Sigma}_\\epsilon$", "Spatiotemporal random field covariance matrix",
-  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
-  "$\\boldsymbol{\\beta}$", "Parameter vector",
-  "$\\boldsymbol{X}$", "Model predictor matrix",
-  "$\\alpha_g$", "IID random intercept deviations",
+  "$\\boldsymbol{Q}_\\omega$", "Spatial random field precision matrix",
+  "$\\boldsymbol{Q}_\\zeta$", "Spatially varying coefficient random field precision matrix",
+  "$\\boldsymbol{Q}_\\epsilon$", "Spatiotemporal random field precision matrix",
+
   "$\\sigma_\\alpha^2$", "IID random intercept variance",
-  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations",
+  "$\\sigma_\\epsilon^2$", "Spatiotemporal random field marginal variance",
+  "$\\sigma_\\omega^2$", "Spatial random field marginal variance",
+  "$\\sigma_\\zeta^2$", "Spatially varying coefficient random field marginal variance",
   "$\\rho$", "Correlation between random fields in subsequent time steps"
 )
-knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex", booktabs = TRUE)
+knitr::kable(symbols, caption = "Symbol notation and descriptions.", escape = FALSE, format = "latex", booktabs = TRUE,
+   linesep = c(
+    rep('', 7), # y, mean etc.
+    '\\addlinespace',
+    rep('', 7), # fields
+    '\\addlinespace',
+    rep('', 5), # covariance stuff
+    '\\addlinespace',
+    rep('', 2), # bayes
+    '\\addlinespace'
+  ))
 ```
@@
 
-# Combined model structure
+# sdmTMB model structure
 
@@
 \begin{aligned}
-\mathbb{E}(y_{\s,t}) &= \mu_{\s,t},\\
+\mathbb{E}[y_{\s,t}] &= \mu_{\s,t},\\
 \mu_{\bm{s},t} &=
 f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} +
+O_{\s,t} +
 \bm{\alpha}_g +
@@
 
-* $\mathbb{E}$ indicates expected value;
 * $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
@@
 * $\bm{\beta}$ represents a vector of fixed-effect coefficients;
+* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
 * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
@@
 
-Next, we will split this model apart to describe the various parts in more detail using $\ldots$ to represent the other optional components.
+A single sdmTMB model will rarely, if ever, contain all of the above components.
+Next, we will split the model apart to describe the various parts in more detail using $\ldots$ to represent the other optional components.
 
@@
 $$
-where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
+where $\boldsymbol{Q}_\omega$ is a sparse precision matrix and $\sigma_\omega^2$ is the marginal variance.
 
@@
 
-## [Lognormal](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L22-L30)
+## [Lognormal](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54)
 
@@
 
-## [Negative Binomial 2 (NB2)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
-
-$$
-\operatorname{NB2} \left( \mu, \phi \right)
-$$
-
-where $\mu$ is the mean and $\phi$ is the dispersion parameter.
-The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].
-Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
-
-Example: `family = nbinom2(link = "log")`
-
-## [Negative Binomial 1 (NB1)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
+## [Negative Binomial 1 (NB1)](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/sdmTMB.cpp#L577-L582)
 
@@
 
+## [Negative Binomial 2 (NB2)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
+
+$$
+\operatorname{NB2} \left( \mu, \phi \right)
+$$
+
+where $\mu$ is the mean and $\phi$ is the dispersion parameter.
+The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].
+The NB2 parametrization is more commonly seen in ecology than the NB1.
+Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.
+
+Example: `family = nbinom2(link = "log")`
+
 ## [Poisson](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa1ed15503e1441a381102a8c4c9baaf1)
@@
 
-## [Student-t](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L11-L20)
+## [Student-t](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L37-L45)
 
@@
 
-# Random field details
+# Gaussian random fields
 
@@
 
-The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates the values at the locations of the observed or predicted data [@lindgren2015]:
+The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates from the knot locations to the values at the locations of the observed or predicted data [@lindgren2015]:
 
@@
 
-The website <https://haakonbakkagit.github.io/btopic128.html> provides an illustration with INLA.
-The implementation within TMB was borrowed from code written by Olav Nikolai Breivik and Hans Skaug at <https://github.com/skaug/tmb-case-studies>.
+[This website](https://haakonbakkagit.github.io/btopic128.html) by Francesco Serafini and Haakon Bakka provides an illustration with INLA.
+The implementation within TMB was borrowed from code written by Olav Nikolai Breivik and Hans Skaug at the [TMB Case Studies](https://github.com/skaug/tmb-case-studies) Github site.
 
-# Optimization details
+# Optimization
+
+## Optimization details
 
@@
 Like AD Model Builder, TMB allows for parameters to be fit in phases and we include the `multiphase` argument in `sdmTMB::sdmTMBcontrol()` to allow this. 
-For high dimensional models (lots of fixed and random effects), phased estimation may be faster and result in more stable convergence.
-In sdmTMB, phased estimation proceeds by first estimating all fixed effect parameters contributing to the likelihood (holding random effects constant at initial values).
-In the second phase, the random effect parameters (and their variances) are also estimated.
-Fixed effect parameters are also estimated in the second phase, and initialized at their estimates from the first phase.
+For high-dimensional models (many of fixed and random effects), phased estimation may be faster and result in more stable convergence. 
+In sdmTMB, phased estimation proceeds by first estimating all fixed-effect parameters contributing to the likelihood (holding random effects constant at initial values). 
+In the second phase, the random-effect parameters (and their variances) are also estimated. 
+Fixed-effect parameters are also estimated in the second phase and are initialized at their estimates from the first phase.
 
-In some cases, a single call to `stats::nlminb()` may not be result in convergence (or a small maximum gradient), and the algorithm may need to be run multiple times.
+In some cases, a single call to `stats::nlminb()` may not be result in convergence (e.g., the maximum gradient of the marginal likelihood with respect to fixed-effect parameters is not small enough yet), and the algorithm may need to be run multiple times.
 In the `sdmTMB::sdmTMBcontrol()` function, we include an argument `nlminb_loops` that will restart the optimization at the previous best values. 
-The number of `nlminb_loops` should generally be small (e.g.
-2 or 3 initially), and defaults to 1.
+The number of `nlminb_loops` should generally be small (e.g., 2 or 3 initially), and defaults to 1. 
 For some sdmTMB models, the Hessian may also be unstable and need to be re-evaluated. 
@@
 By default, this is not included and `newton_loops` is set to 0.
-
-# Optimization
-
-The sdmTMB model is fit by maximum marginal likelihood.
-Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
-Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
-
-Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`)---a greater number of iterations and smaller tolerance thresholds increase the chance that the optimal solution is found, but more evaluations translates into longer computation time.
-Additionally, multiple function calls to `stats::nlminb()` may aid convergence; we include `nlminb_loops` as an argument to the control list in `sdmTMB()` (e.g., `sdmTMB(..., control = sdmTMBcontrol(nlmbinb_loops = 2))`), where each call is initialized at the best parameter vector from the previous iteration.
-Similarly, additional optimization with a Newton solver (`stats::optimHess()`) can be implemented by adding `newton_loops` in the control list.
 If a model is already fit, the function `sdmTMB::run_extra_optimization()` can run additional optimization loops with either routine to further reduce the maximum gradient.
 
-# Assessing convergence
+## Assessing convergence
 
-Much of the guidance around diagnostics and glmmTMB also applies to `sdmTMB`, e.g.
-[https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html).
+Much of the guidance around diagnostics and glmmTMB also applies to sdmTMB, e.g.
+[the glmmTMB vignette on troubleshooting](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html).
+Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`)---a greater number of iterations and smaller tolerance thresholds increase the chance that the optimal solution is found, but more evaluations translates into longer computation time.
 Warnings of non-positive-definite Hessian matrices (accompanied by parameters with `NA`s for standard errors) often mean models are improperly specified given the data.

### 2021-12-21 doc/appendix-model.Rmd
Edit

@@
 
-* Greek symbols for parameters
+* Greek symbols for parameters,
 
-* Bold symbols for vectors or matrices
+* the Latin/Roman alphabet for data,
 
-* $\phi$ for all distribution dispersion parameters for consistency with the code
+* bold symbols for vectors or matrices,
 
-* $\mathbb{E}[y]$ to define the expected value (mean) of variable $y$
+* $\phi$ for all distribution dispersion parameters for consistency with the code,
 
-* $\mathrm{Var}[y]$ to define the expected variance of the variable $y$
+* $\mathbb{E}[y]$ to define the expected value (mean) of variable $y$,
 
-* An $^*$ superscript to represent interpolated or projected values as opposed to values at knot locations (e.g., $\omega$ vs. $\omega^*$).
+* $\mathrm{Var}[y]$ to define the expected variance of the variable $y$, and
 
-```{r, results='asis'}
+* a $^*$ superscript to represent interpolated or projected values as opposed to values at knot locations (e.g., $\omega$ vs. $\omega^*$).
+
+We include tables of all major indices (Table \@ref(tab:indices)) and symbols (Table \@ref(tab:symbols)).
+
+```{r indices}
+library(magrittr)
+subscripts <- tibble::tribble(
+  ~Symbol, ~Description,
+  "$\\s$", "Index for space; a vector of x and y coordinates",
+  "$t$", "Index for time",
+  "$g$", "Group"
+)
+knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex", booktabs = TRUE) %>% 
+  kableExtra::kable_styling(latex_options = "hold_position")
+```
+
+```{r symbols, results='asis'}
 symbols <- tibble::tribble(
@@
 
-```{r}
-subscripts <- tibble::tribble(
-  ~Symbol, ~Description,
-  "$\\s$", "Index for space; a vector of x and y coordinates",
-  "$t$", "Index for time",
-  "$g$", "Group"
-)
-knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex", booktabs = TRUE)
-```
+\clearpage
 
@@
 
+The random fields are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\zeta$):
+
+$$
+\boldsymbol{\zeta} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\zeta^2 \boldsymbol{Q}^{-1}_\zeta\right).
+$$
+
 ## IID random or multi-level intercepts
@@
 where $\alpha_g$ is an example optional "random" intercept---an intercept
-centered on zero that varies by level $g$ and is constrained by
+with mean zero that varies by level $g$ and is constrained by
 $\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
@@
 
-[Source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) as in [cplm](https://CRAN.R-project.org/package=cplm) package. <https://link.springer.com/article/10.1007/s11222-012-9343-7>
+[Source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) as in the [cplm](https://CRAN.R-project.org/package=cplm) package [@zhang2013a].
 

### 2021-12-21 doc/appendix-model.Rmd
Add refs

@@
 
-[Source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) as in the [cplm](https://CRAN.R-project.org/package=cplm) package [@zhang2013a].
-
 $$
@@
 Internally, $p$ is transformed from $\mathrm{logit}^{-1} (\texttt{thetaf}) + 1$ to constrain it between 1 and 2 and \texttt{thetaf} is estimated as an unconstrained variable.
+
+The [source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) is implemented as in the [cplm](https://CRAN.R-project.org/package=cplm) package [@zhang2013a] and is based on @dunn2005.
 
@@
 Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`.
-There are a number of ways to implement anisotropic covariance REF, and we adopt a 2-parameter rotation matrix $\textbf{H}$.
-The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$
+There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$.
+The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.
 

### 2021-12-22 doc/appendix-model.Rmd
Develop appendix

@@
   bookdown::pdf_document2:
-    toc: false
+    toc: true
     number_sections: false
@@
 \newcommand{\s}{\boldsymbol{s}}
+
+\clearpage
 
@@
   "$\\sigma_\\zeta^2$", "Spatially varying coefficient random field marginal variance",
-  "$\\rho$", "Correlation between random fields in subsequent time steps"
+  "$\\kappa_\\omega$", "Spatial decorrelation rate",
+  "$\\kappa_\\epsilon$", "Spatiotemporal decorrelation rate",
+
+  "$\\rho$", "Correlation between random fields in subsequent time steps",
+  "$\\boldsymbol{A}$", "A sparse projection matrix to bilinearly interpolate between knot and data locations",
+  "$\\boldsymbol{H}$", "A 2-parameter rotation matrix used to define anisotropy"
 )
@@
     '\\addlinespace',
-    rep('', 2), # bayes
-    '\\addlinespace'
+    rep('', 5), # more covariance stuff
+    '\\addlinespace',
+    rep('', 99) # end
   ))
@@
 
-## IID random or multi-level intercepts
-
-$$
-\begin{aligned}
-  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
-  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
-\end{aligned}
-$$
-
-where $\alpha_g$ is an example optional "random" intercept---an intercept
-with mean zero that varies by level $g$ and is constrained by
-$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
-syntax as in lme4 or glmmTMB. There can be multiple random intercepts. 
-E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.
-
 ## Time-varying regression parameters
@@
 The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).
+
+## IID random or multi-level intercepts
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
+  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
+\end{aligned}
+$$
+
+where $\alpha_g$ is an example optional "random" intercept---an intercept
+with mean zero that varies by level $g$ and is constrained by
+$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
+syntax as in lme4 or glmmTMB. There can be multiple random intercepts. 
+E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.
 
@@
 $K_\nu$ represents the modified Bessel function of the second kind,
-and $\kappa$ controls the spatial correlation range.
+and $\kappa$ controls the spatial correlation range (decorrelation rate).
 The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+Internally, the parameter $\kappa$ and $\tau$ are converted to range and marginal standard deviation $\sigma$ as $\textrm{range} = \sqrt{8} / \kappa$ and $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
 
-## Projection '$\boldsymbol{A}$' matrix
+In the case of a spatiotemporal model with both spatial and spatiotemporal fields, if `share_range` is `TRUE` in `sdmTMB()` (the default), then a single $\kappa$ and range are estimated with separate $\sigma_\omega$ and $\sigma_\epsilon$.
+This often makes sense since the data are often only weakly informative about $\kappa$.
+If `share_range = FALSE`, then separate $\kappa_\omega$ and $\kappa_\epsilon$ are estimated.
+The spatially varying coefficient field always shares $\kappa$ with the spatial random field.
+
+## Projection $\boldsymbol{A}$ matrix
 

### 2021-12-22 doc/appendix-model.Rmd
Change distr formatting

@@
 
-## [Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaee11f805f02bc1febc6d7bf0487671be)
+## Binomial
 
@@
 
+Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaee11f805f02bc1febc6d7bf0487671be).
+
 Example: `family = binomial(link = "logit")`
 
-## [Beta](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga5324c83759d5211c7c2fbbad37fa8e59)
+## Beta
 
@@
 
+Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga5324c83759d5211c7c2fbbad37fa8e59).
+
 Example: `family = Beta(link = "logit")`
 
-## [Gamma](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gab0e2205710a698ad6a0ed39e0652c9a3)
+## Gamma
 
@@
 
+Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gab0e2205710a698ad6a0ed39e0652c9a3).
+
 Example: `family = Gamma(link = "log")`
 
-## [Gaussian](https://kaskr.github.io/adcomp/dnorm_8hpp.html)
+## Gaussian
 
@@
 
-## [Lognormal](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54)
+Code defined [within TMB](https://kaskr.github.io/adcomp/dnorm_8hpp.html).
+
+## Lognormal
 
@@
 
+Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.
+
 Example: `family = lognormal(link = "log")`
 
-## [Negative Binomial 1 (NB1)](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/sdmTMB.cpp#L577-L582)
+## Negative Binomial 1 (NB1)
 
@@
 
+Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/sdmTMB.cpp#L577-L582) based on NB2 and borrowed from glmmTMB.
+
 Example: `family = nbinom1(link = "log")`
 
-## [Negative Binomial 2 (NB2)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)
+## Negative Binomial 2 (NB2)
 
@@
 
+Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351).
+
 Example: `family = nbinom2(link = "log")`
 
-## [Poisson](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa1ed15503e1441a381102a8c4c9baaf1)
+## Poisson
 
@@
 
+Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa1ed15503e1441a381102a8c4c9baaf1).
+
 Example: `family = poisson(link = "log")`
 
-## [Student-t](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L37-L45)
+## Student-t
 
@@
 
+Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L37-L45) based on the `dt()` distribution in TMB.
+
 Example: `family = student(link = "log", df = 7)`
 
-## [Tweedie](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga262f3c2d1cf36f322a62d902a608aae0)
+## Tweedie
 
@@
 The [source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) is implemented as in the [cplm](https://CRAN.R-project.org/package=cplm) package [@zhang2013a] and is based on @dunn2005.
+The TMB version is defined [here](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga262f3c2d1cf36f322a62d902a608aae0).
 

### 2021-12-22 doc/appendix-model.Rmd
Edit

@@
     toc: true
-    number_sections: false
+    number_sections: true
 fig_caption: true
@@
 I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
-The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).
+The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and, this case, the intercept should be omitted from the main effects (`formula ~ + 0 + ...` or `formula ~ -1 + ...`).
 

### 2021-12-22 doc/appendix-model.Rmd
ADMB

@@
 
-Like AD Model Builder, TMB allows for parameters to be fit in phases and we include the `multiphase` argument in `sdmTMB::sdmTMBcontrol()` to allow this. 
-For high-dimensional models (many of fixed and random effects), phased estimation may be faster and result in more stable convergence. 
+Like AD Model Builder [@fournier2012ad], TMB allows for parameters to be fit in phases and we include the `multiphase` argument in `sdmTMB::sdmTMBcontrol()` to allow this. 
+For high-dimensional models (many fixed and random effects), phased estimation may be faster and result in more stable convergence. 
 In sdmTMB, phased estimation proceeds by first estimating all fixed-effect parameters contributing to the likelihood (holding random effects constant at initial values). 

### 2021-12-22 doc/appendix-model.Rmd
Cite VAST

@@
 
-* $\mathrm{Var}[y]$ to define the expected variance of the variable $y$, and
+* $\mathrm{Var}[y]$ to define the expected variance of the variable $y$,
 
-* a $^*$ superscript to represent interpolated or projected values as opposed to values at knot locations (e.g., $\omega$ vs. $\omega^*$).
+* a $^*$ superscript to represent interpolated or projected values as opposed to values at knot locations (e.g., $\boldsymbol{\omega}$ vs. $\boldsymbol{\omega}^*$), and
+
+* where possible, notation has been chosen to match VAST [@thorson2019] to maintain consistency (e.g., $\boldsymbol{\omega}$ for spatial fields and $\boldsymbol{\epsilon}_t$ for spatiotemporal fields).
 
@@
 Spatiotemporal random fields are represented by $\boldsymbol{\epsilon}_t$ within sdmTMB. 
-This has been chosen to match the representation in VAST (REF). 
+This has been chosen to match the representation in VAST [@thorson2019]. 
 The marginal standard deviation of $\boldsymbol{\epsilon}_t$ is indicated by `Spatiotemporal SD` in the model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.

### 2021-12-22 doc/paper.Rmd
Edit

@@
 <!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
-We provide examples of the syntax and the reproducibility of results achieved when similar models were built using INLA and sdmTMB (Appendix 4) and VAST and sdmTMB (Appendix 5). 
-We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax. <!-- Change this if we manage to solve the inlabru issue  -->
+We provide examples of the syntax and the reproducibility of results achieved when similar models are fit using INLA (Appendix 4) and VAST (Appendix 5).
+<!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
+<!-- TODO: mention inlabru here -->
 

### 2021-12-22 doc/paper.Rmd
Draft abstract

@@
 abstract: |
-    | \noindent Point 1: set the context for and purpose of the work;
-    | Point 2: indicate the approach and methods;
-    | Point 3: outline the main results;
-    | Point 4: identify the conclusions and the wider implications.
+    | \noindent 1.
+Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs).
+These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models, with an interface familiar to those who have used lme4 or glmmTMB.
+Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB).
+    | 3: We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software.
+We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a user-friendly interface familiar to many R users, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
+    | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
+    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, Template Model Builder
 ---
+
+<!-- Point 1: set the context for and purpose of the work; -->
+<!-- Point 2: indicate the approach and methods; -->
+<!-- Point 3: outline the main results; -->
+<!-- Point 4: identify the conclusions and the wider implications. -->
+
+<!-- This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
+-->
+<!-- These data sometimes include a discrete time dimension, such as yearly samples.  -->
+<!-- Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes and improve ecological management. -->
+<!-- These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+-->
+<!-- We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users. -->
 

### 2021-12-22 doc/paper.Rmd
Edit abstract

@@
 These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
-    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models, with an interface familiar to those who have used lme4 or glmmTMB.
+    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
 Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB).
-    | 3: We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software.
-We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a user-friendly interface familiar to many R users, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
+We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software.
+    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
     | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
@@
 Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
-Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv, for a specific class of spatial and spatiotemporal models [@wood2017a].
+Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv [@wood2017a], for a specific class of spatial and spatiotemporal models.
 One common application is for species distribution models (SDMs), hence the package name.

### 2021-12-22 doc/appendix-validation.Rmd
Substantial paper editing

@@
 ---
-title: "Appendix X: sdmTMB validation"
+title: "Appendix 6: sdmTMB model validation and speed-testing methods"
 output:

### 2021-12-22 doc/paper.Rmd
Substantial paper editing

@@
 author: "|
-  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
-  $^2$Other addresses\n"
+  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], 
+  Eric J. Ward$^2$, 
+  Philina A. English$^1$, 
+  Lewis A. K. Barnett$^3$\n| 
+  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
+  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
+      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
+  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
+      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
 output:
@@
 Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB).
-We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software.
+We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
     | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
@@
 Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
+While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
 Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
-Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
 
-Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
+Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data [@rue2005gmrf; @diggle2007; @cressie2011].
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle2007; @chiles1999].
 
-While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
+Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
 Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
@@
 This can greatly improve computational efficiency [@heaton2017].
-The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
 Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
+See Appendix 1 for a full description of the model.
 Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
@@
 sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
-Geostatistical data simply refers to data observed at specific spatial coordinates. 
-The data can be collected across discrete points in time.
-Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
+Geostatistical data simply refers to data observed at specific spatial coordinates reflecting some underlying spatial process [@rossi1992; @diggle2007]. 
+These data can be collected across discrete points in time.
+Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR) [e.g., @verhoef2018].
 sdmTMB can also fit models with areal data if each polygon has an associated centroid.
@@
 The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
-\@ref(fig:diagram)).
-This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
-See Appendix TODO for a full description of the statistical model.
+\@ref(fig:diagram), Appendix 1).
+This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
 
@@
 The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
-The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim$ 0.13 (REF)---can be shared or estimated separately.
+The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately.
 
@@
 
-By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix TODO).
+By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix 1).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix TODO).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix 1).
 
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
 
@@
 
-Here, `cutoff` defines the minimum allowed distance between points in the units of `X` and `Y` (km).
+Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y` (km).
 Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
@@
 (A) SPDE mesh (lines) combined with the trawl survey observations (points).
-Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
+Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
 The circle area corresponds to density of Pacific Cod for individual trawls.
@@
 Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3):
+We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3).
 
@@
 <!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
-(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection (e.g., via expected log predictive density, ELPD, REF) or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
+(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
 (3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
@@
       "Technically possible but non-trivial",
-      "Technically possible but non-trivial",
-      "Available in a companion package INLAutils"
+      "Technically possible but non-trivial"
     ),
@@
 Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
-spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
+spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
 <!-- TODO: still true? -->
@@
 <!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
-We provide examples of the syntax and the reproducibility of results achieved when similar models are fit using INLA (Appendix 4) and VAST (Appendix 5).
+We provide comparisons of the syntax and the reproducibility of results from models fit using INLA (Appendix 4) and VAST (Appendix 5).
 <!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
@@
 
-We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Appendix 6 TODO, Fig.
-\@ref(fig:timing)).
+We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+\@ref(fig:timing), Appendix 6).
 In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
@@
 With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
-Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern substantially (not shown).
+Results with up to an order of magnitude more data slightly shift the lines up without affecting the overall pattern substantially (not shown).
 Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
+Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
 <!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
@@
 alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
-Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
-Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
+Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
+Speed-wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
 INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.
@@
 
-TODO
-Would not be possible without TMB and INLA.
-Heavily inspired by and in some places code has been adapted from VAST and glmmTMB...
+TODO\
+Would not be possible without TMB and INLA and authors...\
+Heavily inspired by and in some places code has been adapted from VAST and glmmTMB...\
+Friendly/internal reviewers...
 
@@
 
-None?
+TODO: None?
 
@@
 
+Appendix 1: sdmTMB statistical model description\
+Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB\
+Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB\
+Appendix 4: Species distribution model comparison between INLA and sdmTMB\
+Appendix 5: Example of index standardization of fishery-independent survey data with VAST\
+Appendix 6: sdmTMB model validation and speed-testing methods
 <!--
@@
 # Data Availability
+
+TODO\
+Will be on CRAN\
+Paper source code will be on Zenodo\
+<https://github.com/pbs-assess/sdmTMB>\
+<https://github.com/seananderson/sdmTMB-paper>
 

### 2021-12-22 doc/paper.Rmd
Clean .bib

@@
 For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
-Simulation-based residuals [@DHARMa], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
+Simulation-based residuals [@dharma], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
 (5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

### 2021-12-22 doc/clean-bib.sh

@@ -1,0 +1,1 @@
+bibtool -s --print.use.tab=Off --print.line.length=80 -i refs.bib -o refs.bib

### 2021-12-22 doc/appendix-validation.Rmd
Edit appendix

--- 2021-12-14 doc/appendix-validation.Rmd
@@
 ---
-title: "Appendix X: sdmTMB validation"
+title: "Appendix 6: sdmTMB model validation and speed-testing methods"
 output:
   bookdown::pdf_document2:
-    toc: false
+    toc: true
 number_sections: false
@@
 
-```{r}
-library(sdmTMB)
-```
+# Overall description of sdmTMB validation
 
@@
 
-# Penalized smoothers
+## Penalized smoothers
 
@@
 
-## `s(x)`
+### `s(x)`
 
 First, a simple thin plate regression spline with a single variable. We will use simulated data from the mgcv package.
+
+```{r, echo=TRUE}
+library(sdmTMB)
+```
 
@@
 
-## `s(x1, by = x2)`
+### `s(x1, by = x2)`
 
@@
 
-## `s(x, y)`
+### `s(x, y)`
 
@@
 
-# IID random intercepts
+## IID random intercepts
 
@@
 
-# Simulation-testing with various observation families
+## Simulation-testing with various observation families
 
@@
 ```
+
+# Supplemental methods for speed-testing related packages
+
+TODO: Fig. 2 methods description

### 2021-12-22 doc/appendix-binomial-inla.Rmd
Edit inla appendix

@@
 library(INLA)
-# library(inlabru)
-theme_set(ggsidekick::theme_sleek())
+theme_set(theme_light())
 ```
+<!-- # library(inlabru) -->
 
-In this example, we fit a presence-absence species distribution model with fixed effect intercepts for year and spatiotemporal random fields using AR1. 
-This will be fit with a binomial distribution and a logit link. 
-Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
-INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`.
+In this example, we fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1). 
+We will use a binomial distribution and a logit link. 
+We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
+INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:
 
@@
 
-In this example, we will use identical priors for the range and random field SD (aka. sigma) for models built in INLA and with sdmTMB.
-While being described different in each package, the numbers amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).
+In this example, we will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in INLA and sdmTMB.
+The values amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).
 
@@
 
-In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., the which defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now sdmTMB only takes a mean and SD (e.g., `ar1_rho = normal(0, 1)`) so we will not set a prior for either model.
+In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `ar1_rho = normal(0, 1)`).
 
 ```{r}
-h.spec <- list(rho = list(prior = 'pccor1', param = c(0, 0.9)))
+h.spec <- list(rho = list(prior = "pccor1", param = c(0, 0.9)))
 ```
 
-We will start by building an identical meshes using both sdmTMB and INLA.
+We will start by building identical meshes using both sdmTMB and INLA.
 Here, X and Y are coordinates in UTM zone 9.
-These same functions that are used internally in the sdmTMB `make_mesh()` function.
+These same functions are used internally in the sdmTMB `make_mesh()` function.
 
@@
   prior.range = c(range_min, prior_prob),
-    prior.sigma = c(sigma_max, prior_prob)) 
+  prior.sigma = c(sigma_max, prior_prob)
+)
 ```
@@
 ```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in"}
-plot(sdm_mesh);title(main = "sdmTMB mesh");plot(spde$mesh)
+plot(sdm_mesh)
+title(main = "sdmTMB mesh")
+plot(spde$mesh)
 ```
@@
 
-Data preparation for INLA models includes a formulae, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
-TODO: explain further, including `control.inla = list(int.strategy = "eb", strategy = "gaussian")` 
-TODO: We use very weak priors on the fixed effects (`mean = 0, prec = 1/(100*100)`)?
+Data preparation for INLA models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
+We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB (REF).
+We use weak priors on the fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.
 
 ```{r inla}
-tictoc::tic("INLA")
+tictoc::tic("INLA") # start timing
 
 nyear <- length(unique(d$year))
-
-iset <- inla.spde.make.index('i', n.spde = spde$n.spde, n.group = nyear)
-
+iset <- inla.spde.make.index("i", n.spde = spde$n.spde, n.group = nyear)
 A <- inla.spde.make.A(mesh = inla_mesh, loc = loc_xy, group = d$i_year)
-
 sdat <- inla.stack(
@@
   effects = list(iset, year_factor = as.factor(d$year)),
-  tag = 'stdata')
+  tag = "stdata"
+)
 
-formulae <- y ~ 0 + year_factor + f(i, model = spde, group = i.group,
-  control.group = list(model = 'ar1'))
-  # control.group = list(model = 'ar1', hyper = h.spec))
+formula <- y ~ 0 + year_factor + f(i,
+  model = spde, group = i.group,
+  control.group = list(model = "ar1", hyper = h.spec)
+)
 
-m_inla <- inla(formulae,
+m_inla <- inla(formula,
   data = inla.stack.data(sdat),
@@
   control.family = list(link = "logit"),
-  control.fixed = list(expand.factor.strategy = 'inla', mean = 0, prec = 1/(100*100)),
+  control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
   verbose = FALSE,
@@
 )
-
-tictoc::toc()
+tictoc::toc() # stop timing
 
 summary(m_inla)
-# saves fixed effect coefs
 inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
@@
 
-We then fit a very similar model with `sdmTMB()`. 
+We will now fit a nearly identical model with `sdmTMB()` (the AR(1) correlation parameter prior is different).
 We again use very weak priors on the fixed effects.
@@
 tictoc::tic("sdmTMB")
-
 m <- sdmTMB(present ~ 0 + as.factor(year),
@@
   spatial = "off",
-  spatiotemporal = "AR1", 
+  spatiotemporal = "ar1",
   family = binomial(link = "logit"),
@@
     matern_st = pc_matern(
-      range_gt = range_min, range_prob = prior_prob, #P(range > 5) = 1-0.05
-      sigma_lt = sigma_max, sigma_prob = prior_prob #P(sigma < 5) = 1-0.05
+      range_gt = range_min, range_prob = prior_prob,
+      sigma_lt = sigma_max, sigma_prob = prior_prob
     ),
-    b = normal(0, 100)
-  ))
-
+    b = normal(0, 100),
+    ar1_rho = normal(0, 2)
+  )
+)
 tictoc::toc()
@@
 tidy(m, "ran_pars", conf.int = TRUE)
-# saves fixed effect coefs
 sdmtmb_coefs <- tidy(m, effects = "fixed")
@@
 
-With sdmTMB, we can extract parameter simulations from the joint precision matrix using `spread_sims()` or `gather_sims()` functions.
+With sdmTMB, we can extract parameter simulations from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.
 
@@
   levels = unique(post_sdmTMB$`.variable`),
-                  labels = c( "2007", "2009", "2011", "2013", "2015", "2017", 
-                                "Range", "AR1 rho", "Random Field SD")))
+  labels = c(
+    "2007", "2009", "2011", "2013", "2015", "2017",
+    "Range", "AR1 rho", "Random Field SD"
+  )
+))
 ```
 
-We then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA. 
+We can then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA. 
 From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
-TODO: maybe say something about why not using `inla.posterior.sample(result = m_inla, n = 400L)`?
+Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the posterior.
+<!-- # post_inla <- inla.posterior.sample(result = m_inla, n = 400L) -->
 
 ```{r inla-densities}
-# post_inla <- inla.posterior.sample(result = m_inla, n = 400L)
 x <- list()
 x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>%
-  as_tibble() %>% mutate(variable = "as.factor.year.2007") 
+  as_tibble() %>%
+  mutate(variable = "as.factor.year.2007")
 x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>%
-  as_tibble() %>% mutate(variable = "as.factor.year.2009") 
+  as_tibble() %>%
+  mutate(variable = "as.factor.year.2009")
 x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>%
-  as_tibble() %>% mutate(variable = "as.factor.year.2011") 
+  as_tibble() %>%
+  mutate(variable = "as.factor.year.2011")
 x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>%
-  as_tibble() %>% mutate(variable = "as.factor.year.2013") 
+  as_tibble() %>%
+  mutate(variable = "as.factor.year.2013")
 x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>%
-  as_tibble() %>% mutate(variable = "as.factor.year.2015") 
+  as_tibble() %>%
+  mutate(variable = "as.factor.year.2015")
 x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>%
-  as_tibble() %>% mutate(variable = "as.factor.year.2017") 
+  as_tibble() %>%
+  mutate(variable = "as.factor.year.2017")
 x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>%
-  as_tibble() %>% mutate(variable = "range") 
+  as_tibble() %>%
+  mutate(variable = "range")
 x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>%
-  as_tibble() %>% mutate(variable = "sigma_E") 
+  as_tibble() %>%
+  mutate(variable = "sigma_E")
 x[[9]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>%
-  as_tibble() %>% mutate(variable = "ar1_rho") 
+  as_tibble() %>%
+  mutate(variable = "ar1_rho")
 post_inla1 <- do.call(rbind, x)
-
 post_inla2 <- post_inla1 %>% mutate(variable = factor(variable,
   levels = unique(post_inla1$variable),
-                  labels = c( "2007", "2009", "2011", "2013", "2015", "2017", 
-                                "Range", "Random Field SD", "AR1 rho")))
+  labels = c(
+    "2007", "2009", "2011", "2013", "2015", "2017",
+    "Range", "Random Field SD", "AR1 rho"
+  )
+))
 ```
@@
 ggplot(post_inla2) +
-  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), bins = 30, 
-                 fill = "blue", alpha = 0.2) + 
-  geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
+  geom_histogram(
+    data = post_sdmTMB2, aes(.value, after_stat(density)), bins = 30,
+    fill = "blue", alpha = 0.2
+  ) +
+  geom_line(data = post_inla2, aes(x, y), inherit.aes = F) +
   facet_wrap(~variable, scales = "free")
@@
 
-We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year.
+We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year in space.
 
@@
 ```{r maps, cache=FALSE, out.width="6in"}
-ggplot(pred) + geom_raster(aes(X, Y, fill = plogis(est))) + 
-  facet_wrap(~year) + scale_fill_viridis_c() + coord_fixed() 
+ggplot(pred) +
+  geom_raster(aes(X, Y, fill = plogis(est))) +
+  facet_wrap(~year) +
+  scale_fill_viridis_c() +
+  coord_fixed()
 ```
@@
 ```{r inla-pred, out.width="6in"}
-idx <- inla.stack.index(sdat, 'stdata')$data
+idx <- inla.stack.index(sdat, "stdata")$data
 d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
-ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + 
-  facet_wrap(~year) + scale_colour_viridis_c() + coord_fixed() 
+ggplot(d) +
+  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
+  facet_wrap(~year) +
+  scale_colour_viridis_c() +
+  coord_fixed()
 ```

### 2021-12-22 doc/appendix-binomial-inla.Rmd
Copy edit appendices

@@
 ---
-title: "Appendix 4: species distribution model comparison between INLA and sdmTMB"
+title: "Appendix 4: Species distribution model comparison between INLA and sdmTMB"
 output:
@@
 tictoc::tic("INLA") # start timing
-
 nyear <- length(unique(d$year))

### 2021-12-22 doc/appendix-pcod-VAST-tweedie.Rmd
Copy edit appendices

@@
 ---
-title: "Appendix 5: example of index standardization of fishery-independent survey data with VAST"
+title: "Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB"
 output:

### 2021-12-22 doc/appendix-snow-sdmTMB.Rmd
Copy edit appendices

@@
 ---
-title: "Appendix 3: example of a spatially varying effect of climate on citizen science count data with sdmTMB"
+title: "Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB"
 output:
@@
     highlight: tango
+    toc: true
+    number_sections: true
+urlcolor: blue
+highlight: "monochrome"
 bibliography: refs.bib
@@
 library(sf)
-theme_set(ggsidekick::theme_sleek() + theme( # modifications for plotting sf data
+```
+
+```{r theme, echo=FALSE}
+theme_set(theme_light() + theme( # modifications for plotting sf data
   legend.position = "top", legend.justification = c(0, 0), # move legend to top left
@@
   panel.grid.major = element_line(colour = "grey90"), # for lat/lons
-  axis.title = element_blank())) # lat/lon is clear from axis text
+  axis.title = element_blank()
+)) # lat/lon is clear from axis text
 ```
 
-This illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
-We will use counts of Snow Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts [CBCs, @cbc]. 
+This appendix illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
+We will use counts of Snowy Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts [CBCs, @cbc]. 
 Each survey is conducted between December 14 to January 5 in 24.14 kilometre diameter circle. 
@@
 For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
-For these years, we have retrieved mean annual climate indices from the University of East Anglia's Climatic Research Unit ([ https://crudata.uea.ac.uk/cru/data/nao/](https://crudata.uea.ac.uk/cru/data/nao/)).
+For these years, we have retrieved mean annual climate indices from the [University of East Anglia's Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
 Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance.
@@
 ```{r explore, fig.asp = 0.3, cache=FALSE}
-ggplot(snow, aes(year, count, colour = nao)) + geom_jitter() + 
-  scale_colour_viridis_c() + ggsidekick::theme_sleek()
+ggplot(snow, aes(year, count, colour = nao)) +
+  geom_jitter() +
+  scale_colour_viridis_c() +
+  theme_light()
 ```
@@
 Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
-Because there does not appear to be much correlation between years, year will be included as a random factor: `(1|year_f)`. 
-We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "IID"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance.
+Because there does not appear to be much correlation between years, year will be included as a random factor: `(1 | year_f)`. 
+We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "iid"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance.
 The random intercept for year allows the mean estimate for each of these spatiotemporal fields to vary. 
 If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
+Since these models will take a while to fit, in practice, we might set `silent = FALSE` to monitor optimization progress.
 
-```{r fit-poisson, echo=TRUE, eval=FALSE}
-m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
+```{r rds1, echo=FALSE, cache=FALSE}
+rds_saved <- file.exists(here::here("data/snow_w_main_effect_0_150km.rds"))
+```
+
+```{r fit-poisson, echo=TRUE, eval=!rds_saved, warning=FALSE}
+m0 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
   spatial_varying = ~ 0 + nao,
@@
   family = poisson(link = "log"),
-             data = snow, mesh = mesh)
+  data = snow, mesh = mesh
+)
 ```
 
-```{r fit-negb1, echo=TRUE, eval=FALSE}
-m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
+```{r fit-negb1, echo=TRUE, eval=!rds_saved, warning=FALSE}
+m1 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
   spatial_varying = ~ 0 + nao,
@@
   family = nbinom1(link = "log"),
-            data = snow, mesh = mesh)
+  data = snow, mesh = mesh
+)
 ```
 
-```{r fit-negb2, echo=TRUE, eval=FALSE}
-m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
+```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
+m2 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
   spatial_varying = ~ 0 + nao,
@@
   family = nbinom2(link = "log"),
-            data = snow, mesh = mesh)
+  data = snow, mesh = mesh
+)
 ```
 
-```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
+```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
+if (!rds_saved) {
   saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
@@
   saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
+}
 ```
@@
 ```{r aic, echo=TRUE, cache=FALSE}
-AIC(m0,m1,m2)
+AIC(m0, m1, m2)
 ```
@@
 
-```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
-m <- sdmTMB(count ~ 1 + nao + (1|year_f), 
+```{r rds2, echo=FALSE, cache=FALSE}
+rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_150km_reml.rds"))
+```
+
+```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE}
+m <- sdmTMB(count ~ 1 + nao + (1 | year_f),
   spatial_varying = ~ 0 + nao,
-            time = "year", spatial = "on", spatiotemporal = "IID",
+  time = "year", 
+  spatial = "on", 
+  spatiotemporal = "iid",
   family = nbinom2(link = "log"),
   reml = TRUE,
-            data = snow, mesh = mesh)
+  data = snow, mesh = mesh
+)
 ```
@@
 ```{r save-reml-model, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
+if (!rds_saved2) {
   saveRDS(m, here::here("data/snow_w_main_effect_150km_reml.rds"))
+}
 ```
@@
 <!-- \clearpage -->
-We can inspect randomized quantile residuals, although this method is not guaranteed to be accurate.
+We can inspect randomized quantile residuals, although this method has known issues.
 
@@
 ```{r plot-resids, fig.width = 8, cache=FALSE}
-ggplot(pred, aes(X, Y, colour = resid)) + geom_point(size = 0.5) +
-  facet_wrap(~year) + coord_fixed() +
+ggplot(pred, aes(X, Y, colour = resid)) +
+  geom_point(size = 0.5) +
+  facet_wrap(~year) +
+  coord_fixed() +
   scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
@@
 p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
-p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
+p_proj <- p %>%
+  mutate(x = X, y = Y) %>%
+  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
 ```
@@
 lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers) # only large lakes
 ```
@@
 b <- tidy(m)
-
-p_mean <- p %>% group_by(CBCID) %>% summarise(
-  X = mean(X), Y= mean(Y), 
+p_mean <- p %>%
+  group_by(CBCID) %>%
+  summarise(
+    X = mean(X), Y = mean(Y),
     mean_est_count = exp(mean(est)),
@@
 ```{r plot-map, echo=TRUE, eval=T, cache=T}
-ggplot(data = p_proj) + geom_point(data = p_mean, 
-     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
+ggplot(data = p_proj) +
+  geom_point(
+    data = p_mean,
+    aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
+  ) +
   geom_sf(data = coast, colour = "gray50") +
   geom_sf(data = lakes, colour = "gray50", fill = NA) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
-           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
   scale_colour_viridis_c(guide = guide_colourbar(
-    direction = "horizontal", title.position = "top")) +
-  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
-  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
+    direction = "horizontal", title.position = "top"
+  )) +
+  guides(size = "none") +
+  labs(colour = "NAO effect\non Snowy Owl count") +
+  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
 ```
@@
 ggplot(data = filter(p_proj, year > 1978)) +
-  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est)), 
-             size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
-           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
+  geom_point(
+    data = filter(p, year > 1978), aes(X, Y, colour = exp(est)),
+    size = 0.4, alpha = 0.5
+  ) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
+  scale_colour_viridis_c(
+    trans = "sqrt", # makes variation in low numbers visible
     # can also trim the extreme high values
     limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
-    guide = guide_colourbar(title.position = "top")) +
+    guide = guide_colourbar(title.position = "top")
+  ) +
   labs(colour = "Predicted counts") +
@@
 In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
-These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers like coasts.
+These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers such as coasts.
 
@@
 ggplot(data = p_proj) +
-  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
+  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
   geom_sf(data = coast, colour = "gray50") +
   geom_sf(data = lakes, colour = "gray50", fill = NA) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
-           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
   scale_colour_viridis_c(guide = guide_colourbar(
-    direction = "horizontal", title.position = "top")) +
-  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
+    direction = "horizontal", title.position = "top"
+  )) +
+  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
 ```
@@
 ggplot(data = filter(p_proj, year > 1978)) +
-  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st), 
-             size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
-           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
+  geom_point(
+    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
+    size = 0.4, alpha = 0.5
+  ) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
   scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
@@
 
-In this case these only include the overall effect of mean annual NAO and the random intercepts for year.
+In this case, these only include the overall effect of mean annual NAO and the random intercepts for year.
 
@@
 ggplot(data = filter(p_proj, year > 1978)) +
-  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)), 
-             size = 0.4, alpha = 0.5) +
-  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
-           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
-  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
-    guide = guide_colourbar(title.position = "top")) +
+  geom_point(
+    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
+    size = 0.4, alpha = 0.5
+  ) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
+  scale_colour_viridis_c(
+    breaks = c(0.03, 0.05, 0.07),
+    guide = guide_colourbar(title.position = "top")
+  ) +
   theme(axis.text = element_blank(), axis.ticks = element_blank()) +

### 2021-12-22 doc/paper.Rmd
Copy edit appendices

@@
 Appendix 4: Species distribution model comparison between INLA and sdmTMB\
-Appendix 5: Example of index standardization of fishery-independent survey data with VAST\
+Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB\
 Appendix 6: sdmTMB model validation and speed-testing methods

### 2021-12-22 doc/appendix-snow-sdmTMB.Rmd
Edit vignette

@@
 Since these models will take a while to fit, in practice, we might set `silent = FALSE` to monitor optimization progress. 
+However, we will maintain the default `silent = TRUE` to avoid cluttering the document with the output.
 
@@
   family = poisson(link = "log"),
-  data = snow, mesh = mesh
+  data = snow, mesh = mesh,
+  silent = TRUE
 )
@@
   family = nbinom1(link = "log"),
-  data = snow, mesh = mesh
+  data = snow, mesh = mesh,
+  silent = TRUE
 )
@@
   family = nbinom2(link = "log"),
-  data = snow, mesh = mesh
+  data = snow, mesh = mesh,
+  silent = TRUE
 )
@@
   reml = TRUE,
-  data = snow, mesh = mesh
+  data = snow, mesh = mesh,
+  silent = TRUE
 )

### 2021-12-22 doc/appendix-model.Rmd
start model appendix edits

@@
 
-* a $^*$ superscript to represent interpolated or projected values as opposed to values at knot locations (e.g., $\boldsymbol{\omega}$ vs. $\boldsymbol{\omega}^*$), and
+* a $^*$ superscript represents interpolated or projected values as opposed to values at knot locations (e.g., $\boldsymbol{\omega}$ vs. $\boldsymbol{\omega}^*$), and
 
@@
 * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
+<!-- should alpha_g be bold as in the equation above? -->
 * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
@@
 A single sdmTMB model will rarely, if ever, contain all of the above components.
-Next, we will split the model apart to describe the various parts in more detail using $\ldots$ to represent the other optional components.
+Next, we will split the model apart to describe the various parts in more detail using '$\ldots$' to represent the other optional components.
 
@@
 The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed-effects modelling framework.
-This is the same approach as is implemented in the R packages gamm4 (REF) and brms (REF).
+This is the same approach as is implemented in the R packages gamm4 [@wood2020] and brms [@burkner2017].
+<!-- There are too other options for the brms ref, burkner2018 and burkner2021 -->
 

### 2021-12-22 doc/appendix-snow-sdmTMB.Rmd
Edit snowy app.

@@
 ```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
+if (rds_saved) {
   m0 <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
@@
   m2 <- readRDS(here::here("data/snow_w_main_effect_150km.rds"))
+}
 ```
@@
 
-The `nbinom2` distribution appears to have the best fit. 
+According to AIC, the `nbinom2()` (negative binomial, NB2) distribution appears to have the most parsimonious fit. 
 We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model.
@@
 Print the model to inspect the fit:
+
 ```{r check, include = FALSE, cache=FALSE, eval=TRUE}
+if (rds_saved2) {
   m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
-m # loading this twice just prevents some excess messages from cluttering the output when printing below
+}
+# m # loading this twice just prevents some excess messages from cluttering the output when printing below
 ```
@@
 
-View the confidence intervals on both fixed and random parameter estimates using `tidy()`.
+View the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`.
 
@@
 
-<!-- \clearpage -->
 Check the spatiotemporal distribution of the residuals.
@@
 ```
-
-<!-- \clearpage -->
 

### 2021-12-22 doc/appendix-snow-sdmTMB.Rmd
Make snowy app. run

@@
 
-```{r zero-test, echo=TRUE, message=FALSE, cache=FALSE}
+```{r zero-test, echo=TRUE, message=FALSE}
 s_nb2 <- simulate(m, nsim = 500)
@@
 
-```{r plot-resids, fig.width = 8, cache=FALSE}
+```{r plot-resids, fig.width = 8}
 ggplot(pred, aes(X, Y, colour = resid)) +
@@
 
-We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http//www.naturalearthdata.com/download/10m/physical/ne_10m_lakes.zip) and transformed them to the correct projection.
+We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from <http://www.naturalearthdata.com/> and transformed them to the correct projection.
 
 ```{r shapes}
+if (!file.exists(here::here("data/ne_10m_lakes"))) {
+  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
+    "download/10m/physical/ne_10m_lakes.zip")
+  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
+  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+}
 coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
@@
 
-```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE}
 b <- tidy(m)
@@
 
-```{r plot-map, echo=TRUE, eval=T, cache=T}
+```{r plot-map, echo=TRUE, eval=TRUE}
 ggplot(data = p_proj) +
@@
 
-```{r plot-all-effects, echo=TRUE, eval=T, cache=T}
+```{r plot-all-effects, echo=TRUE, eval=TRUE}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=T, cache=T}
+```{r plot-spatial-effects, echo=TRUE, eval=TRUE}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=T}
+```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-```{r plot-fixed-effects, echo=TRUE, eval=T, cache=T}
+```{r plot-fixed-effects, echo=TRUE, eval=TRUE}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2021-12-22 doc/appendix-model.Rmd
Add code to model table

@@
 symbols <- tibble::tribble(
-  ~Symbol, ~Description,
-  "$y$", "Observed response data",
-  "$\\mu$", "Mean",
-  "$\\phi$", "A dispersion parameter for a distribution",
-  "$f$", "Link function",
-  "$f^{-1}$", "Inverse link function",
-  "$\\boldsymbol{\\beta}$", "Parameter vector",
-  "$\\boldsymbol{X}$", "A predictor model matrix",
-  "$O_{\\s, t}$", "An offset variable at point $\\s$ and time $t$",
+  ~Symbol, ~Code, ~Description,
+  "$y$", "\\texttt{y\\_i}", "Observed response data",
+  "$\\mu$", "\\texttt{mu\\_i}", "Mean",
+  "$\\phi$", "\\texttt{phi}", "A dispersion parameter for a distribution",
+  "$f$", "\\texttt{fit\\$family\\$link}", "Link function",
+  "$f^{-1}$", "\\texttt{fit\\$family\\$linkinv}", "Inverse link function",
+  "$\\boldsymbol{\\beta}$", "\\texttt{b\\_j}", "Parameter vector",
+  "$\\boldsymbol{X}$", "\\texttt{X\\_ij}", "A predictor model matrix",
+  "$O_{\\s, t}$", "\\texttt{offset}", "An offset variable at point $\\s$ and time $t$",
 
-  "$\\omega_{\\s}$", "Spatial random field at point $\\s$ (knot)",
-  "$\\omega_{\\s}^*$", "Spatial random field at point $\\s$ (interpolated)",
-  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$ (knot)",
-  "$\\zeta_{\\s, t}^*$", "Spatially varying coefficient random field at point $\\s$ and time $t$ (interpolated)",
-  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
-  "$\\epsilon_{\\s, t}^*$", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
-  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations (knot)",
-  "$\\alpha_g$", "IID random intercept deviation for group $g$",
+  "$\\omega_{\\s}$", "\\texttt{omega\\_s}", "Spatial random field at point $\\s$ (knot)",
+  "$\\omega_{\\s}^*$", "\\texttt{omega\\_s\\_A}", "Spatial random field at point $\\s$ (interpolated)",
+  "$\\zeta_{\\s, t}$", "\\texttt{zeta\\_t}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (knot)",
+  "$\\zeta_{\\s, t}^*$", "\\texttt{zeta\\_t\\_A}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (interpolated)",
+  "$\\epsilon_{\\s, t}$", "\\texttt{epsilon\\_st}", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
+  "$\\epsilon_{\\s, t}^*$", "\\texttt{epsilon\\_st\\_A}", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
+  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR1 or random walk spatiotemporal deviations (knot)",
+  "$\\alpha_g$", "\\texttt{RE}", "IID random intercept deviation for group $g$",
 
-  "$\\boldsymbol{\\Sigma}_\\omega$", "Spatial random field covariance matrix",
-  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
-  "$\\boldsymbol{\\Sigma}_\\epsilon$", "Spatiotemporal random field covariance matrix",
-  "$\\boldsymbol{Q}_\\omega$", "Spatial random field precision matrix",
-  "$\\boldsymbol{Q}_\\zeta$", "Spatially varying coefficient random field precision matrix",
-  "$\\boldsymbol{Q}_\\epsilon$", "Spatiotemporal random field precision matrix",
+  "$\\boldsymbol{\\Sigma}_\\omega$", "\\texttt{-}", "Spatial random field covariance matrix",
+  "$\\boldsymbol{\\Sigma}_\\zeta$", "\\texttt{-}", "Spatially varying coefficient random field covariance matrix",
+  "$\\boldsymbol{\\Sigma}_\\epsilon$", "\\texttt{-}", "Spatiotemporal random field covariance matrix",
+  "$\\boldsymbol{Q}_\\omega$", "\\texttt{Q\\_s}", "Spatial random field precision matrix",
+  "$\\boldsymbol{Q}_\\zeta$", "\\texttt{Q\\_s}", "Spatially varying coefficient random field precision matrix",
+  "$\\boldsymbol{Q}_\\epsilon$", "\\texttt{Q\\_st}", "Spatiotemporal random field precision matrix",
 
-  "$\\sigma_\\alpha^2$", "IID random intercept variance",
-  "$\\sigma_\\epsilon^2$", "Spatiotemporal random field marginal variance",
-  "$\\sigma_\\omega^2$", "Spatial random field marginal variance",
-  "$\\sigma_\\zeta^2$", "Spatially varying coefficient random field marginal variance",
-  "$\\kappa_\\omega$", "Spatial decorrelation rate",
-  "$\\kappa_\\epsilon$", "Spatiotemporal decorrelation rate",
+  "$\\sigma_\\alpha^2$", "\\texttt{sigma\\_G}", "IID random intercept variance",
+  "$\\sigma_\\epsilon^2$", "\\texttt{sigma\\_E}", "Spatiotemporal random field marginal variance",
+  "$\\sigma_\\omega^2$", "\\texttt{sigma\\_O}", "Spatial random field marginal variance",
+  "$\\sigma_\\zeta^2$", "\\texttt{sigma\\_Z}", "Spatially varying coefficient random field marginal variance",
+  "$\\kappa_\\omega$", "\\texttt{kappa(0)}", "Spatial decorrelation rate",
+  "$\\kappa_\\epsilon$", "\\texttt{kappa(1)}", "Spatiotemporal decorrelation rate",
 
-  "$\\rho$", "Correlation between random fields in subsequent time steps",
-  "$\\boldsymbol{A}$", "A sparse projection matrix to bilinearly interpolate between knot and data locations",
-  "$\\boldsymbol{H}$", "A 2-parameter rotation matrix used to define anisotropy"
+  "$\\rho$", "\\texttt{ar1\\_rho}", "Correlation between random fields in subsequent time steps",
+  "$\\boldsymbol{A}$", "\\texttt{A}", "Sparse projection matrix to interpolate between knot and data locations",
+  "$\\boldsymbol{H}$", "\\texttt{H}", "2-parameter rotation matrix used to define anisotropy"
 )
-knitr::kable(symbols, caption = "Symbol notation and descriptions.", escape = FALSE, format = "latex", booktabs = TRUE,
+knitr::kable(symbols, caption = "Symbol notation, code representation (in model output or in model template code), and descriptions.", escape = FALSE, format = "latex", booktabs = TRUE,
    linesep = c(

### 2021-12-22 doc/paper.Rmd
Edit abstract

@@
 Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs).
-These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
-    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
+It is statistically challenging to model such data effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+    | 2: Here, we introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
 Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB).
 We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
-    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
+    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population index standardization).
     | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.

### 2021-12-22 doc/paper.Rmd
Edit opening sentence

@@
     | \noindent 1.
-Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs).
-It is statistically challenging to model such data effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+Analyzing geostatistical data---spatially referenced observations with some spatial process---is increasingly common in ecology for applications such as species distribution models (SDMs).
+It is challenging to model such data effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
     | 2: Here, we introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.

### 2021-12-22 doc/appendix-validation.Rmd
Edit app.

@@
 
-The sdmTMB has a number of validation processes to ensure the code is working as expected.
-This includes unit testing using the testthat package (REF) embedded within the package.
+The sdmTMB package has a number of validation processes to ensure the code is working as expected. 
+This includes unit testing using testthat [@testthat] embedded within the package. 
 This unit testing validates a large number of model configurations against simulated data and comparable models from other packages. 
@@
 
-In addition to the validation discussed here, Appendix X contains a comparison with VAST and Appendix Y contains a comparison with INLA/inlabru.
+In addition to the validation discussed here, Appendix 5 contains a comparison with VAST and Appendix 4 contains a comparison with INLA/inlabru.
 
@@
 
-Here, we compare a number of smoother configurations against mgcv. Throughout the section we will turn off spatial random fields so the models should be identical.
+Here, we compare a number of smoother configurations against mgcv [@wood2017a]. 
+Throughout the section, we will turn off spatial random fields so the models should be comparable.
 
@@
 
-First, a simple thin plate regression spline with a single variable. We will use simulated data from the mgcv package.
+First, a simple thin plate regression spline with a single variable. 
+We will use simulated data from the mgcv package.
 
@@
 
-Next, we compare a model with a continuous `by` argument. This lets the `x2` smoother change depending on the value of `x1`.
+Next, we compare a model with a continuous `by` argument. 
+This lets the `x2` smoother change depending on the value of `x1`.
 
@@
 
-Here we compare a model that includes two IID random intercepts against the same model fit via glmmTMB. Again, we will turn off random fields so the models are comparable.
+Here we compare a model that includes two IID random intercepts against the same model fit via glmmTMB [@brooks2017]. 
+Again, we will turn off random fields so the models are comparable.
 
-First, simulate some data. The variables `g` and `h` will represent groups g and h. g will have 50 groups. h will have 20 groups.
+First, simulate some data. 
+The variables `g` and `h` will represent groups g and h. 
+g will have 50 groups. 
+h will have 20 groups.
 
@@
 
-We can extract the random effect estimates and compare them. We find them to be nearly identical:
+We can extract the random effect estimates and compare them. 
+We find them to be nearly identical:
 
@@
 
-The last plot illustrates the IID random effect deviations themselves with glmmTMB on the x-axis and sdmTMB on the y-axis. The 'g' deviations are black and the 'h' ones are red.
+The last plot illustrates the IID random effect deviations themselves with glmmTMB on the x-axis and sdmTMB on the y-axis. 
+The 'g' deviations are black and the 'h' ones are red.
 

### 2021-12-23 doc/paper.Rmd
Edit abstract

--- 2021-12-22 doc/paper.Rmd
@@
     | \noindent 1.
-Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs).
-It is statistically challenging to model such data effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+Geostatistical data---spatially referenced observations with some underlying spatial process---are increasingly collected in ecology and are used in applications such as species distribution models (SDMs).
+However, appropriate models for these data, such as  generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
     | 2: Here, we introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
-Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB).
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB).
 We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
-    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population index standardization).
+    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB allows for simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population index standardization).
     | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.

### 2021-12-23 doc/appendix-snow-sdmTMB.Rmd
edits to owl appendix

@@
 
+# Introduction
+
 ```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
@@
 library(sf)
-```
-
-```{r theme, echo=FALSE}
-theme_set(theme_light() + theme( # modifications for plotting sf data
-  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
-  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
-  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
-  axis.title = element_blank()
-)) # lat/lon is clear from axis text
 ```
@@
 For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
-For these years, we have retrieved mean annual climate indices from the [University of East Anglia's Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
+For these years, we have retrieved mean annual climate indices from the University of East Anglia's Climatic Research Unit <https://crudata.uea.ac.uk/cru/data/nao/>. 
 Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance.
 Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.
+
+# Preparation of spatial data
 
@@
 
-```{r check, include = FALSE, cache=FALSE, eval=TRUE}
+```{r check, echo=FALSE, include = FALSE, cache=FALSE, eval=TRUE}
 if (rds_saved2) {
   m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
+  m # loading this twice just prevents some excess messages from cluttering the output when printing below
 }
-# m # loading this twice just prevents some excess messages from cluttering the output when printing below
 ```
@@
 We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
-<!-- If our data was zero-inflated ... maybe make suggestions or even link to an example for how to do a delta model in sdmTMB?-->
+If our data was zero-inflated we might needed to use a delta (aka. hurdle) model instead. 
+This is not built in to sdmTMB, but can be achieve using the steps in the vignette [Fitting delta models with sdmTMB](https://https://pbs-assess.github.io/sdmTMB/articles/delta-models.html).
 
@@
 
-Check the spatiotemporal distribution of the residuals.
+Next, we modify a theme to plot spatiotemporal data and check the distribution of these residuals:
+
+```{r theme, cache=FALSE}
+theme_set(theme_bw() + theme(
+  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
+  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
+  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
+  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
+  axis.title = element_blank() # lat/lon is clear from axis text
+)) 
+```
 
@@
 ggplot(pred, aes(X, Y, colour = resid)) +
-  geom_point(size = 0.5) +
+  geom_point(size = 0.25) +
   facet_wrap(~year) +
@@
 p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
-p_proj <- p %>%
-  mutate(x = X, y = Y) %>%
-  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
+p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
 ```
@@
 The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates.
-
 ```{r plot-all-effects, echo=TRUE, eval=TRUE}
-ggplot(data = filter(p_proj, year > 1978)) +
+ggplot(data = p_proj) +
   geom_point(
-    data = filter(p, year > 1978), aes(X, Y, colour = exp(est)),
-    size = 0.4, alpha = 0.5
+    data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5
   ) +
@@
 \clearpage
-## Other components of predicted counts
+## Other components of predictions
 

### 2021-12-29 doc/paper.Rmd
edit figure 2 caption
update owl fig url and minor comment

@@
 The circle area corresponds to density of Pacific Cod for individual trawls.
-The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
-These predictions are made omitting the spatial random field.
 (B) Spatial random field:
@@
 (C) Overall prediction: these estimates represent the combination of all fixed and random effects.
-(D) Conditional effect of depth modelled as a penalized smoother.
-(E) Standardized area-weighted population index derived from a spatiotemporal version of this model.
+(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
+(E) Standardized area-weighted population index derived from a spatiotemporal version of this model. The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
+<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->
 
@@
 ```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="4.1in", fig.align='center'}
-owl <- here::here("figs", "owl_nao_effect.pdf")
+owl <- here::here("figs", "owl-nao-effect-2021-12-22.pdf")
 if (!file.exists(owl)) {
-  download.file("https://www.dropbox.com/s/fz1iw0a0diohyxc/owl_nao_effect.pdf?dl=1", 
+  download.file("https://www.dropbox.com/sh/p5ydljdycaw9x6n/AAAlaKQArhqP9IHvUSxEq_qWa/owl-nao-effect-2021-12-22.pdf?dl=0", 
     destfile = owl)
@@
 ```
+<!-- LB: VAST does allow use of penalized splines, but this requires the splines package and effects package for visualization, e.g.: https://github.com/James-Thorson-NOAA/VAST/wiki/Specify-covariates-and-visualize-responses -->
 

### 2021-12-29 doc/paper.Rmd
Fix dl=0

@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations with some underlying spatial process---are increasingly collected in ecology and are used in applications such as species distribution models (SDMs).
-However, appropriate models for these data, such as  generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
+Geostatistical data---spatially referenced observations with some underlying spatial process---are commonly collected in ecology and are used in applications such as species distribution models (SDMs).
+However, appropriate models for analyzing these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
     | 2: Here, we introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB).
-We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
-    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB allows for simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population index standardization).
+We describe the basic functionality of sdmTMB and the statistical model employed, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
+    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population index standardization).
     | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
@@
 if (!file.exists(owl)) {
-  download.file("https://www.dropbox.com/sh/p5ydljdycaw9x6n/AAAlaKQArhqP9IHvUSxEq_qWa/owl-nao-effect-2021-12-22.pdf?dl=0", 
+  download.file("https://www.dropbox.com/sh/p5ydljdycaw9x6n/AAAlaKQArhqP9IHvUSxEq_qWa/owl-nao-effect-2021-12-22.pdf?dl=1", 
     destfile = owl)

### 2021-12-29 doc/paper.Rmd
minor abstract edits

@@
 if (!file.exists(owl)) {
-  download.file("https://www.dropbox.com/sh/p5ydljdycaw9x6n/AAAlaKQArhqP9IHvUSxEq_qWa/owl-nao-effect-2021-12-22.pdf?dl=1", 
+  download.file("https://www.dropbox.com/sh/p5ydljdycaw9x6n/AAAlaKQArhqP9IHvUSxEq_qWa/owl-nao-effect-2021-12-22.pdf?dl=0", 
     destfile = owl)

### 2021-12-29 doc/paper.Rmd
Force add figs

@@
 dir.create("figs", showWarnings = FALSE)
-f <- here::here("figs", "diagram-2021-12-10.pdf")
-if (!file.exists(f)) {
-  download.file("https://www.dropbox.com/s/lvoem3ahwkftzjy/diagram-2021-12-10.pdf?dl=1", 
-    destfile = f)
-}
+f <- here::here("figs", "diagram.pdf")
 knitr::include_graphics(f)
@@
 ```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", out.width="5.2in", fig.align='center'}
-pcodfig <- here::here("figs", "pcod-fig.pdf")
-if (!file.exists(pcodfig)) {
-  download.file("https://www.dropbox.com/s/s8kl4u49p27qqhy/pcod-fig.pdf?dl=1", 
-    destfile = pcodfig)
-}
+pcodfig <- here::here("figs", "pcod-fig2.pdf")
 knitr::include_graphics(pcodfig)
@@
 ```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="4.1in", fig.align='center'}
-owl <- here::here("figs", "owl-nao-effect-2021-12-22.pdf")
-if (!file.exists(owl)) {
-  download.file("https://www.dropbox.com/sh/p5ydljdycaw9x6n/AAAlaKQArhqP9IHvUSxEq_qWa/owl-nao-effect-2021-12-22.pdf?dl=0", 
-    destfile = owl)
-}
+owl <- here::here("figs", "owl-nao-effect.png")
 knitr::include_graphics(owl)

### 2021-12-29 doc/paper.Rmd
more abstract edits to emphasize utility to broader audience

--- 2021-12-29 doc/paper.Rmd
@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations with some underlying spatial process---are commonly collected in ecology and are used in applications such as species distribution models (SDMs).
-However, appropriate models for analyzing these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
+Geostatistical data---spatially referenced observations with some underlying spatial process---are commonly collected in ecology and have much potential for revealing new ecological processes and improving conservation and management.
+Geostatistical data are used in a diversity of applications, including species distribution models (SDMs); however, appropriate models for analyzing these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
     | 2: Here, we introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
@@
 We describe the basic functionality of sdmTMB and the statistical model employed, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
-    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population index standardization).
+    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
     | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.

### 2021-12-29 doc/paper.Rmd
intro edits

@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations with some underlying spatial process---are commonly collected in ecology and have much potential for revealing new ecological processes and improving conservation and management.
+Geostatistical data---spatially referenced observations with some underlying spatial process---are commonly collected in ecology and have much potential for revealing ecological processes and improving conservation and management, expanding on the time series analyses often used for such efforts by quantifying spatial and spatiotemporal relationships.
 Geostatistical data are used in a diversity of applications, including species distribution models (SDMs); however, appropriate models for analyzing these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
@@
 Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
+While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
+<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?.
+More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit?" -->
 
@@
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle2007; @chiles1999].
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle2007; @chiles1999].
 
@@
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or lack some of the flexibility for the particular class of models discussed here.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or lack some flexibility.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

### 2021-12-29 doc/paper.Rmd
minor model description and pcod example edits

@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+After rapid model exploration with maximum likelihood, one can optionally pass a sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
 
@@
 An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
+It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
 <!-- Since this is not particular to your package, maybe the e.g.
@@
 -->
-Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
+Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.
 
@@
 
-Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y` (km).
+Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
 Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
@@
 
-We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+We could extend this spatial model to be a spatiotemporal model simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured in time.
 Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).
@@
 
-We could generate a population index that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
+We could generate a population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
 \@ref(fig:pcod-fig)E, Appendix 2).

### 2021-12-29 doc/paper.Rmd
owl ex edits

@@
 <!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?.
-More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit?" -->
+More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit?
+-->
 
@@
 Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
+<!-- not familiar with this "irruptive" terminology, please define -->
 The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
@@
 We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao)).
-This result would be consistent with owls closer to the Atlantic coast and those migrating the furthest south being the most effected by NAO.
+This indicates that the owls most effected by NAO were those closer to the Atlantic coast and those migrating the furthest south.
 
@@
 ```
-<!-- LB: VAST does allow use of penalized splines, but this requires the splines package and effects package for visualization, e.g.: https://github.com/James-Thorson-NOAA/VAST/wiki/Specify-covariates-and-visualize-responses -->
+<!-- LB: VAST does allow use of basis splines, but this requires the splines package and effects package for visualization, e.g.: https://github.com/James-Thorson-NOAA/VAST/wiki/Specify-covariates-and-visualize-responses -->
 

### 2021-12-29 doc/paper.Rmd
minor edits to validation and package comparison

@@
 (2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
-(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
+(3) A sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
 (4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
@@
 There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
-sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
 sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
@@
 With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
-Results with up to an order of magnitude more data slightly shift the lines up without affecting the overall pattern substantially (not shown).
+Results with up to an order of magnitude more data slightly shift the lines up without affecting the qualitative pattern (not shown).
 Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
@@
 <!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
+<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->
 

### 2021-12-29 doc/paper.Rmd
minor discussion edits

@@
 How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
-Assuming a given package can fit the model of interest (Table 1), we suggest the primary difference is the user interface.
+Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
 We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
@@
 Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
-These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+These data present statistical challenges to modelling them effectively and efficiently as appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
 We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.

### 2021-12-29 doc/appendix-snow-sdmTMB.Rmd
more owl appendix edit
consider adding cv?

@@
 output:
-  pdf_document:
-    highlight: tango
+  bookdown::pdf_document2:
+    highlight: pygments
     toc: true
@@
 urlcolor: blue
-highlight: "monochrome"
 bibliography: refs.bib
@@
 
-# Introduction
-
-```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
-library(ggplot2)
-library(dplyr)
-library(sdmTMB)
-library(sf)
-```
+# Background
 
@@
 # Preparation of spatial data
+
+```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
+library(ggplot2)
+library(dplyr)
+library(sdmTMB)
+library(sf)
+```
 
@@
 
-The result is a mesh with `r mesh$mesh$n` knots.
+The result is a mesh with `r mesh$mesh$n` (`mesh$mesh$n`) knots.
 
@@
 
+# Model validation
+
 ```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
@@
 
-```{r check, echo=FALSE, include = FALSE, cache=FALSE, eval=TRUE}
+```{r check, echo=FALSE, include = FALSE, eval=TRUE}
 if (rds_saved2) {
   m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
-  m # loading this twice just prevents some excess messages from cluttering the output when printing below
+  m # printing this first just prevents some excess messages from cluttering the output when printing below
 }
@@
 
-```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
+```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE}
 print(m)
@@
 We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
-If our data was zero-inflated we might needed to use a delta (aka. hurdle) model instead. 
-This is not built in to sdmTMB, but can be achieve using the steps in the vignette [Fitting delta models with sdmTMB](https://https://pbs-assess.github.io/sdmTMB/articles/delta-models.html).
+If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead. 
+This is not built directly into sdmTMB, but can be achieved by fitting the two components separately.
+See the steps in the vignette [Fitting delta models with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/delta-models.html).
 
@@
 ```
+
+<!-- # ```{r} -->
+<!-- # library(future) -->
+<!-- # plan(multisession) -->
+<!-- # m_cv <- sdmTMB_cv(count ~ 1 + nao + (1 | year_f), -->
+<!-- #   spatial_varying = ~ 0 + nao, -->
+<!-- #   time = "year",  -->
+<!-- #   spatial = "on",  -->
+<!-- #   spatiotemporal = "iid", -->
+<!-- #   family = nbinom2(link = "log"), -->
+<!-- #   # reml = TRUE, -->
+<!-- #   use_initial_fit = TRUE, -->
+<!-- #   data = snow, mesh = mesh, -->
+<!-- #   k_folds = 10 -->
+<!-- # ) -->
+<!-- # m_cv -->
+<!-- # ``` -->
 

### 2021-12-30 doc/paper.Rmd
Tweak some edits ('an' before 'es'dmTMB)

@@
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
 <!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?.
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can optionally pass a sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
 
@@
 We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao)).
-This indicates that the owls most effected by NAO were those closer to the Atlantic coast and those migrating the furthest south.
+This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
 
@@
 (2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
-(3) A sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
+(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
 (4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.

### 2022-01-05 doc/appendix-model.Rmd
EW edits to model-appendix
Mostly minor notation stuff

@@
 * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
-<!-- should alpha_g be bold as in the equation above? -->
+<!-- should alpha_g be bold as in the equation above? --><!-- EW: yes -- alpha_g should be bolded here -->
 * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
@@
 ### Smoothers
-
+<!-- EW: see this paragraph and below, should package names be enclosed in single quotes, like `sdmTMB()` above? Examples: mgcv, brms, sdmTMB throughout -->
+<!-- EW: unrelated, I never tried to implement factor smooths. I assume they work, but it'd good to check eventually -->
 Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
@@
 ## Spatial random fields
-
-Spatial random fields, $\omega_s$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
+<!-- EW: edited omega in this section to be bolded and subscripted with s, for consistency with above --> 
+Spatial random fields, $\omega_{\bm{s}}$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
 
@@
 \begin{aligned}
-\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_s + \ldots \right),\\
-\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
+\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \boldsymbol{\omega_s} + \ldots \right),\\
+\boldsymbol{\omega_s} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
 \end{aligned}
@@
 $$
-\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
+\boldsymbol{\omega_s} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
 $$
@@
 ### IID spatiotemporal random fields
-
+<-- EW: as above, I changed epsilon to be subscripted with s throughout this section-->
 IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as
@@
 \begin{aligned}
-\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
-\boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
+\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \boldsymbol{\epsilon_{\s,t}} + \ldots \right),\\
+\boldsymbol{\epsilon_{\s,t}} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
 \end{aligned}
@@
 
-where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
+where $\boldsymbol{\epsilon_{\s,t}}$ represent random field deviations at point $\s$ and time $t$.
 The random fields are assumed independent across time slices.
@@
 $$
-\boldsymbol{\epsilon}_t \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
+\boldsymbol{\epsilon_{\s,t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
 $$
@@
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
-\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{\s,t}},  \:
+\boldsymbol{\epsilon_{\s,t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
-\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon_{\s,t-1}},  \:
+\boldsymbol{\epsilon_{\s,t-1}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant spatially for a given time $t$.
-Currently, $\bm{X}_{3,t}$ is restricted to a single covariate, supplied by a one-sided formula to `spatial_varying`.
+Currently, $\bm{X}_{3,t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
 For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.
@@
 ## IID random or multi-level intercepts
-
+<-- EW: make alpha bold here, as above?-->
 $$
@@
 version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
-
+<-- EW: Put lme4 and glmmTMB in '' too?-->
 Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:
@@
 $$
-
+<<- EW: betareg in ''?-->
 where $\mu$ is the mean and $\phi$ is a precision parameter.
@@
 $$
-
+<-- EW: check this, I think the Gamma variance should be mu * phi2 -->
 where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
-The variance is $\mu \cdot \phi$.
+The mean is $\mu$ and variance is $\mu \cdot \phi^2$.
 
@@
 TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
-
+<-- EW: the lognormal definition above includes the bias correction phi2/2, but that's not in the standard implementations in R, etc.
+I think I used those R functions or referenced Wikipedia in writing the equations out.
+So we should pick one or the other.
+I think just changing the definition to not include the bias correction would keep it consistent with R, https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Lognormal.html-->
 Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.

### 2022-01-05 doc/appendix-pcod-sdmTMB.Rmd
Copy edit appendices
Update appendix-pcod-sdmTMB.Rmd
EW edits

@@
 ---
-title: "Appendix 2: example of index standardization of fishery-independent survey data with sdmTMB"
+title: "Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB"
 output:
@@
 The above example used the default `bias_correct = FALSE` in `get_index()`.
-This is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
+The `bias_correct` argument is a feature of TMB for models with random effects, and more detils can be found in the TMB documentation [https://github.com/kaskr/adcomp/wiki](https://github.com/kaskr/adcomp/wiki) or [Thorson & Kristensen (2016)](https://www.sciencedirect.com/science/article/abs/pii/S0165783615301399?via%3Dihub).
+Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
 Setting `bias_correct = TRUE` implements the generic by his correction algorithm that should correct for this, but can take considerable time and memory.

### 2022-01-05 doc/appendix-binomial-inla.Rmd
Update appendix-binomial-inla.Rmd

@@
 # INLA 
-
+<-- EW: missing a citation in REF below -->
 Data preparation for INLA models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.

### 2022-01-05 doc/appendix-binomial-inla.Rmd
Fix inlabru appendix (needed i_year not year)

--- 2021-12-22 doc/appendix-binomial-inla.Rmd
@@
 library(INLA)
+library(inlabru)
 theme_set(theme_light())
 ```
-<!-- # library(inlabru) -->
 
-In this example, we fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1). 
+In this example, we will fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
 We will use a binomial distribution and a logit link.
@@
 
-In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `ar1_rho = normal(0, 1)`).
+In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `sdmTMBpriors(ar1_rho = normal(0, 1))`).
 
 ```{r}
-h.spec <- list(rho = list(prior = "pccor1", param = c(0, 0.9)))
+h_spec <- list(rho = list(prior = "pccor1", param = c(0, 0.9)))
 ```
@@
 ```{r sdm-mesh, echo = TRUE, cache=FALSE}
-sdm_mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 10)
+sdmTMB_mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 10)
 ```
@@
 spde <- inla.spde2.pcmatern(
-  mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
+  mesh = inla_mesh, # an smdTMB mesh could be inserted here with `sdmTMB_mesh$mesh`
   prior.range = c(range_min, prior_prob),
@@
 
+Here, the meshes should be the same:
+
 ```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in"}
-plot(sdm_mesh)
+plot(sdmTMB_mesh)
 title(main = "sdmTMB mesh")
-plot(spde$mesh)
+plot(spde$mesh, main = "")
+title(main = "INLA mesh")
 ```
@@
 Data preparation for INLA models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
-We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB (REF).
-We use weak priors on the fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.
+We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB.
+We use weak priors on the factor fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.
 
@@
   model = spde, group = i.group,
-  control.group = list(model = "ar1", hyper = h.spec)
+  control.group = list(model = "ar1", hyper = h_spec)
 )
@@
   control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
-  verbose = FALSE,
   control.inla = list(int.strategy = "eb", strategy = "gaussian"),
@@
 
-<!-- # inlabru -->
+# inlabru
 
-<!-- *** NOT WORKING YET *** -->
+There are a couple ways to include the fixed effects with `inlabru`. Unfortunately, using factor variables in the standard equation syntax does not work.
+One option is to create a design matrix and add year effects individually to the formula.
+The other way to set up the fixed coefficients with inlabru is to use the `factor_full` (or `factor_contrast`) model (seen below within `fac()`).
+Note that this option causes the location of the factor fixed-effect parameter estimates to end up in `summary.random$fac`.
 
-<!-- There are a couple ways to include the fixed effects with `inlabru`.
-Unfortunately, using factor variables in the standard equation syntax does not work.
-In the first case, we can create a design matrix and add year effects individually to the formula.
--->
+```{r}
+dat <- as.data.frame(d)
+sp::coordinates(dat) <- c("X", "Y")
+dat <- as(dat, "SpatialPointsDataFrame")
+```
 
-<!-- Now we can set up the design matrix for `inlabru`   -->
+```{r}
+tictoc::tic("bru")
+components <- present ~ 0 +
+  f(
+    main = coordinates, model = spde, group = i_year,
+    control.group = list(model = "ar1", hyper = h.spec)
+  ) +
+  fac(main = year, model = "factor_full")
 
-<!-- ```{r echo = F} -->
-<!-- m <- lm(present ~ -1 + as.factor(year), data = d) -->
-<!-- model_mat <- model.matrix(m) -->
-<!-- colnames(model_mat) <- paste0("x", unique(d$year)) -->
-<!-- dat <- cbind(d, model_mat) %>% mutate(x = X, y = Y) -->
+m_bru <- bru(
+  components = components,
+  family = "binomial",
+  data = dat,
+  options =
+    list(
+      bru_verbose = TRUE,
+      control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
+      control.family = list(link = "logit"),
+      control.inla = list(int.strategy = "eb", strategy = "gaussian")
+    )
+)
+tictoc::toc()
+```
 
-<!-- # dat <- dat %>% sf::st_as_sf(coords = c("x", "y")) -->
+```{r}
+summary(m_bru)
 
-<!-- coordinates(dat) <- c("x", "y") -->
-<!-- dat <- as(dat, "SpatialPointsDataFrame") -->
-<!-- ``` -->
+m_inla$summary.hyperpar[, c("mean", "sd")]
+m_bru$summary.hyperpar[, c("mean", "sd")]
 
-<!-- Next we set up the `components` object for the model.
-This includes a formula with fixed or random effects (here only fixed) and the design of the spatial random field.
-The name `spatrf` can be anything, but includes the coordinates, spde model, and how the year effects are shared.
--->
-
-<!-- ```{r eval = F, echo = F} -->
-<!-- components <- present ~ 0 +  -->
-<!--     x2009 + x2011 + x2013 + x2015 + x2017  +  -->
-<!--     # x2009(main=x2009, model="linear") + -->
-<!--     # x2011(main=x2011, model="linear") + -->
-<!--     # x2013(main=x2013, model="linear") + -->
-<!--     # x2015(main=x2015, model="linear") + -->
-<!--     # x2017(main=x2017, model="linear") + -->
-<!--     spatrf(coordinates, model = spde, group = year, ngroup=length(unique(d$year)), -->
-<!--       control.group = list(model = 'ar1', hyper = h.spec)) -->
-
-<!-- m_bru <- bru( -->
-<!--   components = components, -->
-<!--   family = "binomial", -->
-<!--   options =  -->
-<!--    list(bru_verbose = TRUE,  -->
-<!--         control.fixed = list(expand.factor.strategy = 'inla'), -->
-<!--         control.family = list(link = "logit"), -->
-<!--         control.inla = list(int.strategy = "eb", strategy = "gaussian"), -->
-<!--         control.compute = list(config = TRUE) -->
-<!--   ), -->
-<!--   data = dat -->
-<!-- ) -->
-
-<!-- bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")] -->
-<!-- summary(m_bru) -->
-<!-- ``` -->
-
-<!-- The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac` -->
-
-<!-- ```{r } -->
-<!-- tictoc::tic("bru") -->
-
-<!-- components <- present ~ 0 + -->
-<!--   spatrf(main = coordinates, model = spde, group = dat$year, ngroup=length(unique(dat$year)), -->
-<!--       control.group = list(model = 'ar1')) + -->
-<!--       # control.group = list(model = 'ar1', hyper = h.spec)) + -->
-<!--   # fac(main = year,  model = "iid", -->
-<!--   #     hyper = list(prec = list(initial = log(1e-6), -->
-<!--   #                            fixed = TRUE)) -->
-<!--   # ) -->
-<!--   fac(main = year,  model = "factor_full") -->
-
-<!-- m_bru <- bru( -->
-<!--   components = components, -->
-<!--   family = "binomial", -->
-<!--   options =  -->
-<!--    list(bru_verbose = TRUE,  -->
-<!--         control.fixed = list(expand.factor.strategy = 'inla'), -->
-<!--         control.family = list(link = "logit"), -->
-<!--         control.inla = list(int.strategy = "eb", strategy = "gaussian"), -->
-<!--         control.compute = list(config = TRUE) -->
-<!--   ), -->
-<!--   data = dat -->
-<!-- ) -->
-<!-- tictoc::toc() -->
-
-<!-- bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")] -->
-<!-- summary(m_bru) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- m_bru <- bru_rerun(m_bru, options = list()) -->
-<!-- summary(m_bru) -->
-<!-- ``` -->
-
-<!-- Comparing the spatial random effects between `INLA` and `inlabru` shows us that the latter is not estimating well. -->
-
-<!-- ```{r} -->
-<!-- m_inla$summary.hyperpar[, c("mean", "sd")] -->
-<!-- m_bru$summary.hyperpar[, c("mean", "sd")] -->
-<!-- ``` -->
-
-<!-- As a result the fixed effect estimates are also quite different.  -->
-
-<!-- ```{r} -->
-<!-- plot(inla_coefs$mean, bru_coefs$mean) -->
-<!-- abline(a = 0, b = 1) -->
-<!-- ``` -->
+bru_coefs <- m_bru$summary.random$fac[, c("ID", "mean", "sd")]
+m_inla$summary.fixed[,c("mean", "sd")]
+bru_coefs
+```
 
@@
   data = d,
-  mesh = sdm_mesh,
+  mesh = sdmTMB_mesh,
   time = "year",
@@
 tidy(m, "ran_pars", conf.int = TRUE)
-sdmtmb_coefs <- tidy(m, effects = "fixed")
+sdmTMB_coefs <- tidy(m, effects = "fixed")
 ```
@@
 
-Comparing the fixed effects between `sdmTMB` and `inla` shows the estimates are very tightly correlated.
+Comparing the fixed effects between `sdmTMB()` and `inla()` shows the estimates are very tightly correlated.
 
 ```{r , out.width="5.5in"}
-plot(sdmtmb_coefs$estimate, inla_coefs$mean)
+plot(sdmTMB_coefs$estimate, inla_coefs$mean)
 abline(a = 0, b = 1)
@@
 ```{r sdm-sims}
-post_sdmTMB <- gather_sims(m, n_sims = 2000L)
+post_sdmTMB <- gather_sims(m, n_sims = 2000)
 
@@
 Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the posterior.
-<!-- # post_inla <- inla.posterior.sample(result = m_inla, n = 400L) -->
 
@@
 
-<!-- predict.inla? -->
+Or, we can make the prediction with inlabru:
 
-<!-- ```{r} -->
-<!-- pgrid <- inla.mesh.projector(inla_mesh) -->
+```{r inla-pred, out.width="6in"}
+# pxl <- pixels(spde$mesh)
+# pred_bru <- predict(m_bru, pxl)
 
-<!-- predict(m_inla, pgrid) -->
-<!-- ``` -->
+nd_bru <- nd
+sp::coordinates(nd_bru) <- c("X", "Y")
+nd_bru <- as(nd_bru, "SpatialPointsDataFrame")
+pred_bru <- predict(m_bru, nd_bru)
 
-<!-- # ```{r epsilon, cache=FALSE, out.width="6in"} -->
-<!-- # plot_map(pred, "epsilon_st") + -->
-<!-- #   ggtitle("Spatiotemporal random effects only") + -->
-<!-- #   scale_fill_gradient2() -->
-<!-- # ``` -->
-<!-- #  -->
-<!-- # ```{r resids, out.width="3in"} -->
-<!-- # p <- predict(m) -->
-<!-- # p$resid <- residuals(m) -->
-<!-- # qqnorm(p$resid) -->
-<!-- # qqline(p$resid) -->
-<!-- # ``` -->
+# ggplot(d) +
+#   geom_point(aes(X, Y, colour = ??), size = 2) +
+#   facet_wrap(~year) +
+#   scale_colour_viridis_c() +
+#   coord_fixed()
+```

### 2022-01-05 doc/appendix-binomial-inla.Rmd
Tweak inlabru prediction

@@
 ```{r inla-pred, out.width="6in"}
-# pxl <- pixels(spde$mesh)
-# pred_bru <- predict(m_bru, pxl)
+nd_bru <- nd
+nd_bru <- dplyr::left_join(nd_bru, yr_lu)
+sp::coordinates(nd_bru) <- c("X", "Y")
+nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
+pred_bru <- predict(m_bru, data = nd_bru)
 
-nd_bru <- nd
-sp::coordinates(nd_bru) <- c("X", "Y")
-nd_bru <- as(nd_bru, "SpatialPointsDataFrame")
-pred_bru <- predict(m_bru, nd_bru)
-
-# ggplot(d) +
-#   geom_point(aes(X, Y, colour = ??), size = 2) +
-#   facet_wrap(~year) +
-#   scale_colour_viridis_c() +
-#   coord_fixed()
+nrow(pred_bru$Predictor)
 ```

### 2022-01-05 doc/appendix-binomial-inla.Rmd
Fix inlabru prediction

@@
 
-Or, we can make the prediction with inlabru:
+Or, we can make the prediction with inlabru.
+To do that, we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and use the inverse-link transformation `plogis()`:
 
-```{r inla-pred, out.width="6in"}
+```{r inlabru-pred, out.width="6in"}
 nd_bru <- nd
@@
 nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
-pred_bru <- predict(m_bru, data = nd_bru)
-
-nrow(pred_bru$Predictor)
+pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
+ggplot(as.data.frame(pred_bru)) +
+  geom_raster(aes(X, Y, fill = mean)) +
+  facet_wrap(~year) +
+  scale_fill_viridis_c() +
+  coord_fixed()
 ```

### 2022-01-06 doc/appendix-binomial-inla.Rmd
Tweak appendi

@@
 ---
-title: "Appendix 4: Species distribution model comparison between INLA and sdmTMB"
+title: "Appendix 4: Dynamic species distribution model comparison between INLA, inlabru, and sdmTMB"
 output:
@@
     highlight: tango
+    toc: true
+    number_sections: true
 ---
@@
 
-In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `sdmTMBpriors(ar1_rho = normal(0, 1))`).
+In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).
 
@@
 title(main = "sdmTMB mesh")
-plot(spde$mesh, main = "")
+plot(spde$mesh, main = "", asp = 1)
 title(main = "INLA mesh")
@@
 The other way to set up the fixed coefficients with inlabru is to use the `factor_full` (or `factor_contrast`) model (seen below within `fac()`).
-Note that this option causes the location of the factor fixed-effect parameter estimates to end up in `summary.random$fac`.
+This option causes the location of the factor fixed-effect parameter estimates to end up in `summary.random$fac`.
 
@@
     main = coordinates, model = spde, group = i_year,
-    control.group = list(model = "ar1", hyper = h.spec)
+    control.group = list(model = "ar1", hyper = h_spec)
   ) +
@@
 We will now fit a nearly identical model with `sdmTMB()` (the AR(1) correlation parameter prior is different).
-We again use very weak priors on the fixed effects.
+We again use very weak priors on the factor fixed effects.
 
@@
     ar1_rho = normal(0, 2)
-  )
+  ),
+  reml = TRUE # to match INLA "eb" `int.strategy`
 )
 tictoc::toc()
-summary(m)
+```
+
+```{r sdmtmb-coefs}
+print(m)
 tidy(m, "ran_pars", conf.int = TRUE)
-sdmTMB_coefs <- tidy(m, effects = "fixed")
+sdmTMB_coefs <- tidy(m, effects = "fixed", conf.int = TRUE)
 ```
@@
 
-Comparing the fixed effects between `sdmTMB()` and `inla()` shows the estimates are very tightly correlated.
+Comparing the fixed effects between `sdmTMB()` and `inla()` shows the estimates are very similar:
 
-```{r , out.width="5.5in"}
-plot(sdmTMB_coefs$estimate, inla_coefs$mean)
-abline(a = 0, b = 1)
+```{r coef-plot, out.width="5.5in"}
+inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>% 
+  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>% 
+  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")
+
+bru_coefs <- m_bru$summary.random$fac[, c("0.025quant", "mean", "0.975quant")] %>% 
+  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>% 
+  mutate(term = unique(sdmTMB_coefs$term), model = "inlabru")
+
+dplyr::bind_rows(mutate(sdmTMB_coefs, model = "sdmTMB"), inla_coefs, bru_coefs) %>% 
+  ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, colour = model)) +
+  geom_pointrange(position = position_dodge(width = 0.2)) +
+  theme(axis.title.y = element_blank())
 ```
 
-With sdmTMB, we can extract parameter simulations from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.
+With sdmTMB, we can extract simulated draws from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions:
 
-```{r sdm-sims}
+```{r sdmTMB-sims}
+set.seed(1)
 post_sdmTMB <- gather_sims(m, n_sims = 2000)
-
-post_sdmTMB2 <- post_sdmTMB %>% mutate(variable = factor(`.variable`,
+post_sdmTMB <- post_sdmTMB %>% 
+  mutate(variable = 
+      factor(`.variable`,
         levels = unique(post_sdmTMB$`.variable`),
-  labels = c(
-    "2007", "2009", "2011", "2013", "2015", "2017",
-    "Range", "AR1 rho", "Random Field SD"
+        labels = c(seq(2007, 2017, 2), "Range", "AR1 rho", "Random Field SD")
       )
-))
+  )
 ```
@@
 From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
-Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the posterior.
+Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the INLA model posterior.
 
@@
   mutate(variable = "ar1_rho")
-post_inla1 <- do.call(rbind, x)
+post_inla1 <- dplyr::bind_rows(x)
 post_inla2 <- post_inla1 %>% mutate(variable = factor(variable,
@@
   geom_histogram(
-    data = post_sdmTMB2, aes(.value, after_stat(density)), bins = 30,
+    data = post_sdmTMB, aes(.value, after_stat(density)), bins = 30,
     fill = "blue", alpha = 0.2
   ) +
-  geom_line(data = post_inla2, aes(x, y), inherit.aes = F) +
+  geom_line(data = post_inla2, aes(x, y), inherit.aes = FALSE) +
   facet_wrap(~variable, scales = "free")

### 2022-01-06 doc/appendix-binomial-inla.Rmd
Appendix edits

@@
 
+In this example, we will fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
+We will use a binomial distribution and a logit link.
+
+We will start by loading necessary packages:
+
 ```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
@@
 
-In this example, we will fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
-We will use a binomial distribution and a logit link.
 We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
@@
 yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
-d <- dplyr::left_join(d, yr_lu)
+d <- dplyr::left_join(d, yr_lu, by = "year")
 ```
 
-In this example, we will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in INLA and sdmTMB.
+We will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in INLA, inlabru, and sdmTMB.
 The values amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).
@@
 
-In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).
+In INLA, a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).
 
@@
   control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
-  control.inla = list(int.strategy = "eb", strategy = "gaussian"),
-  control.compute = list(config = TRUE)
+  control.inla = list(int.strategy = "eb", strategy = "gaussian")
 )
@@
 From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
-Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the INLA model posterior.
+Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the INLA model posterior assuming we had set `control.compute = list(config = TRUE)`.
 
@@
 ```
+
+In this plot, the blue histograms are from sdmTMB and the density lines are from INLA:
 
@@
 nd_bru <- nd
-nd_bru <- dplyr::left_join(nd_bru, yr_lu)
+nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
 sp::coordinates(nd_bru) <- c("X", "Y")

### 2022-01-06 doc/appendix-model.Rmd
Appendix edits

@@
   bookdown::pdf_document2:
+    highlight: pygments
     toc: true
@@
 * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
-<!-- should alpha_g be bold as in the equation above? --><!-- EW: yes -- alpha_g should be bolded here -->
 * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
@@
 * $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
+
+<!-- should alpha_g be bold as in the equation above? -->
+<!-- EW: yes -- alpha_g should be bolded here -->
 
@@
 ### IID spatiotemporal random fields
-<-- EW: as above, I changed epsilon to be subscripted with s throughout this section-->
+<!-- EW: as above, I changed epsilon to be subscripted with s throughout this section-->
 IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as
@@
 ## IID random or multi-level intercepts
-<-- EW: make alpha bold here, as above?-->
+<!-- EW: make alpha bold here, as above?-->
 $$
@@
 version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
-<-- EW: Put lme4 and glmmTMB in '' too?-->
+<!-- EW: Put lme4 and glmmTMB in '' too?-->
 Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:
@@
 $$
-<<- EW: betareg in ''?-->
+<!-- EW: betareg in ''?-->
 where $\mu$ is the mean and $\phi$ is a precision parameter.
@@
 $$
-<-- EW: check this, I think the Gamma variance should be mu * phi2 -->
+<!-- EW: check this, I think the Gamma variance should be mu * phi2 -->
 where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
@@
 TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
-<-- EW: the lognormal definition above includes the bias correction phi2/2, but that's not in the standard implementations in R, etc.
+<!-- EW: the lognormal definition above includes the bias correction phi2/2, but that's not in the standard implementations in R, etc.
 I think I used those R functions or referenced Wikipedia in writing the equations out.

### 2022-01-06 doc/appendix-pcod-VAST-tweedie.Rmd
Appendix edits

--- 2021-12-15 doc/appendix-pcod-VAST-tweedie.Rmd
@@
 ---
-title: "Appendix 5: example of index standardization of fishery-independent survey data with VAST"
+title: "Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB"
 output:
-  pdf_document:
-    highlight: tango
+  bookdown::pdf_document2:
+    highlight: pygments
+    toc: false
+    number_sections: true
 ---

### 2022-01-06 doc/appendix-pcod-sdmTMB.Rmd
Appendix edits

@@
 output:
-  pdf_document:
-    highlight: tango
+  bookdown::pdf_document2:
+    highlight: pygments
+    toc: true
+    number_sections: true
 ---

### 2022-01-06 doc/appendix-snow-sdmTMB.Rmd
Appendix edits

@@
 
-# Background
+# Introduction
 
@@
 ```
-
-# Model validation
 
@@
 ```
-
-<!-- # ```{r} -->
-<!-- # library(future) -->
-<!-- # plan(multisession) -->
-<!-- # m_cv <- sdmTMB_cv(count ~ 1 + nao + (1 | year_f), -->
-<!-- #   spatial_varying = ~ 0 + nao, -->
-<!-- #   time = "year",  -->
-<!-- #   spatial = "on",  -->
-<!-- #   spatiotemporal = "iid", -->
-<!-- #   family = nbinom2(link = "log"), -->
-<!-- #   # reml = TRUE, -->
-<!-- #   use_initial_fit = TRUE, -->
-<!-- #   data = snow, mesh = mesh, -->
-<!-- #   k_folds = 10 -->
-<!-- # ) -->
-<!-- # m_cv -->
-<!-- # ``` -->
 

### 2022-01-06 doc/appendix-validation.Rmd
Appendix edits

@@
   bookdown::pdf_document2:
+    highlight: pygments
     toc: true
+    number_sections: true
 number_sections: false

### 2022-01-06 doc/appendix-model.Rmd
Edits to model appendix

--- 2021-12-22 doc/appendix-model.Rmd
@@
 
-This appendix uses the following notation conventions:
+This appendix uses the following notation conventions, generally following the guidance in @edwards2019:
 
@@
 
-* the Latin/Roman alphabet for data,
+* the Latin/Roman alphabet for data (except $\boldsymbol{Q}$ and $\boldsymbol{H}$, which are used by convention),
 
-* bold symbols for vectors or matrices,
+* bold symbols for vectors or matrices (e.g., $\boldsymbol{\omega}$ is a vector and $\omega_{\s}$ is the value of $\boldsymbol{\omega}$ at point in space $\s$),
 
@@
 O_{\s,t} +
-\bm{\alpha}_g +
+\alpha_g +
 \bm{X}_{2,\bm{s},t} \gamma_t +
 \bm{X}_{3,t} \zeta_{\bm{s}} +
-\omega_{\bm{s}} +
-\epsilon_{\bm{s},t} \right),
+\omega_{\s} +
+\epsilon_{\s,t} \right),
 \end{aligned}
@@
 * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
-<!-- should alpha_g be bold as in the equation above? -->
 * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
-* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
-* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
-* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
+* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
+* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
+* $\epsilon_{\s,t}$ represents a spatiotemporal component (a random field), $\epsilon_{\s,t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
 
@@
 
-Spatial random fields, $\omega_s$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
+Spatial random fields, $\omega_{\s}$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
 
@@
 \begin{aligned}
-\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_s + \ldots \right),\\
+\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_{\s} + \ldots \right),\\
 \boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
@@
 \mu_{\bm{s},t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
-\boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
+\boldsymbol{\epsilon_{t}} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
 \end{aligned}
@@
 $$
-\boldsymbol{\epsilon}_t \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
+\boldsymbol{\epsilon_{t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
 $$
@@
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
-\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
+\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 \boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
-\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon_{t-1}},  \:
+\boldsymbol{\epsilon_{t-1}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 Note that, in contrast to the AR(1) parametrization, the variance is no longer the steady-state marginal variance.
-
-<!-- * $\alpha_g$, are defined by an lme4-style formula, e.g. `formula = y ~ x + (1 | g)`, where `g` represents a column with grouping factors. -->
-
-<!-- * $\gamma_{t}$, time-varying coefficients, are defined by the `time_varying` formula argument. -->
-
-<!-- * $\zeta_{\bm{s}}$, IID random intercepts, are defined by the `spatial_varying` formula argument. -->
-
-<!-- * $f$ and $f^-1$ functions are defined by the `family` argument. E.g., `family = Gamma(link = "log")`. -->
 
@@
 where $\alpha_g$ is an example optional "random" intercept---an intercept
-with mean zero that varies by level $g$ and is constrained by
-$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
-syntax as in lme4 or glmmTMB. There can be multiple random intercepts. 
+with mean zero that varies by level $g$ and is constrained by $\sigma_\alpha$. 
+This is defined by the `formula` argument via the `(1 | g)` syntax as in lme4 or glmmTMB. 
+There can be multiple random intercepts, despite only showing one above.
 E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.
@@
 version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
-
+<!-- EW: Put lme4 and glmmTMB in '' too?-->
 Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:
@@
 $$
-
+<!-- EW: betareg in ''?-->
 where $\mu$ is the mean and $\phi$ is a precision parameter.
@@
 $$
-
 where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
-The variance is $\mu \cdot \phi$.
+The mean is $\mu$ and variance is $\mu \cdot \phi^2$.
 
@@
 
+<!-- \operatorname{Lognormal} \left( \log \mu - \frac{\phi^2}{2}, \phi^2 \right) -->
+
+sdmTMB uses the bias-corrected lognormal distribution where $\phi$ represents the standard deviation in log-space:
+
 $$
-\operatorname{Lognormal} \left( \log \mu - \frac{\phi^2}{2}, \phi \right)
+\operatorname{Lognormal} \left( \log \mu - \frac{\phi^2}{2}, \phi^2 \right).
 $$
-
-where $\mu$ and $\phi$ are the mean and standard deviation of the natural log of the random variable. 
-If $\boldsymbol{x}$ is lognormally distributed, $\mathbb{E}[\log(\boldsymbol{x})] = \mu$ and $\mathrm{Var}[\log(\boldsymbol{x})] = \phi$.
-This translates to $\mathbb{E}[\boldsymbol{x}] = \exp(\mu + \phi^2 / 2)$ and $\mathrm{Var}[\boldsymbol{x}] = \left(\exp(\phi^2) - 1 \right) \cdot \exp(2 \mu + \phi^2)$.
-
-TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
+Because of the bias correction, $\mathbb{E}[y] = \mu$ and $\mathrm{Var}[\log y] = \phi^2$.
 

### 2022-01-06 doc/paper.Rmd
Edit paper

@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations with some underlying spatial process---are commonly collected in ecology and have much potential for revealing ecological processes and improving conservation and management, expanding on the time series analyses often used for such efforts by quantifying spatial and spatiotemporal relationships.
-Geostatistical data are used in a diversity of applications, including species distribution models (SDMs); however, appropriate models for analyzing these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
-    | 2: Here, we introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models.
+Geostatistical data are spatially referenced observations of some spatial process---they might also be collected across time.
+Such data are common in ecology and can reveal ecological processes and improve conservation and management through applications such as species distribution models (SDMs).
+However, appropriate models to analyze geostatistical data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
+    | 2.
+Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models.
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB).
-We describe the basic functionality of sdmTMB and the statistical model employed, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
-    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
-    | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
-    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, Template Model Builder
+We describe the basic functionality of sdmTMB and the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+    | 3.
+We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
+    | 4. We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
+    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
 ---
@@
 
-<!-- This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
--->
-<!-- These data sometimes include a discrete time dimension, such as yearly samples.  -->
-<!-- Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes and improve ecological management. -->
-<!-- These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
--->
-<!-- We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users. -->
-
-```{r pkg-checks, include=FALSE, echo=FALSE}
-# if (!require("rnaturalearthhires", quietly = TRUE)) {
-#   stop("Please install 'rnaturalearthhires'.\n",
-#   "`remotes::install_github('ropensci/rnaturalearthhires')`")
-# }
-# if (!require("ggsidekick", quietly = TRUE)) {
-#   stop("Please install 'ggsidekick'.\n",
-#   "`remotes::install_github('seananderson/ggsidekick')`")
-# }
-```
-
-```{r, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
-library(here)
-library(ggplot2)
-# theme_set(theme_minimal())
-theme_set(ggsidekick::theme_sleek())
-library(dplyr)
-library(sdmTMB)
-```
+<!-- TODO: emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference
 
@@
 These data can be collected across discrete points in time.
-Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR) [e.g., @verhoef2018].
+Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional (CAR) autoregressive models [e.g., @verhoef2018].
 sdmTMB can also fit models with areal data if each polygon has an associated centroid.
@@
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix 1).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or can be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix 1).
 
@@
 
-<!-- ## Advanced options -->
-
-<!-- While not of interest to all users, sdmTMB also implements several advanced options for model fitting.  -->
-<!-- First, models can include penalized likelihoods via assigning priors to model parameters. -->
-<!-- These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces. -->
-<!-- Following other recent SPDE approaches [@simpson2017; @osgood-zimmerman2021], 'penalized complexity' priors may be particularly useful in assigning priors to the spatial range and variance. -->
-<!-- Second, we allow non-linear effects of covariates to be modeled with penalized regression splines ("P-splines") [@eilers1996], in one or multiple dimensions, using the same `s(x)` syntax familiar to users of `mgcv`.
--->
-<!-- Third, for users interested in Bayesian inference, we allow models to be passed to Stan (cite) via the package tmbstan [@monnahan2018]. -->
-
 # Introductory example with fish survey biomass
@@
 It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
-<!-- Since this is not particular to your package, maybe the e.g.
+<!-- TODO: Since this is not particular to your package, maybe the e.g.
 for how to do this should be moved to the appendix?
@@
 
-<!--
-EW: TODO I think this could use a bit more hand holding — things to highlight are (1) predictions can be made for a new dataframe with any locations, 
-(2) more detail about why you’re predicting on a grid. 
-(3) other uses for predict — e.g. might be helpful to understand marginal effects of covariates for complicated models
--->
-
 We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
@@
 The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
-<!-- The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
--->
-The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)B--D.
 
@@
 
-We could extend this spatial model to be a spatiotemporal model simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured in time.
+The output of `predict()` is a data frame containing overall estimates in link space, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)B--D.
+
+We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured in time.
 Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).
@@
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-The circle area corresponds to density of Pacific Cod for individual trawls.
+The circle area corresponds to density of Pacific cod for individual trawls.
 (B) Spatial random field:
@@
 
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
-<!-- not familiar with this "irruptive" terminology, please define -->
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are
+irruptive
+migrants appearing across mid latitudes of North America in winter.
+<!-- TODO: not familiar with this "irruptive" terminology, please define -->
 The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
@@
 (1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
-<!-- We allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]; however, because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
--->
-<!-- The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. -->
-<!-- The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model. -->
-<!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
 (2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
@@
 Simulation-based residuals [@dharma], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
-(5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
+(5) The `simulate.sdmTMB()` method can simulate from fitted models [e.g., for DHARMa residuals, @DHARMa] and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
 <!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
-<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->
 
@@
 ```
-
-<!--
-Highlights?
-
-- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances)
-- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
-- Spatiotemporal models (RW/AR v IID)
-- Time varying parameters
-- Non-linear effects (smooth splines, threshold/breakpoint functions)
-- Variance estimation and derived quantities (spatial range, area occupied etc)
--->
-
-<!-- \clearpage  -->
 
@@
 d <- d[,c(1, 2, 3, 5, 6, 4, 7, 8, 9)]
-# names(d)[names(d) == "mgcv"] <- "mgcv\\footnote{a}"
 d$Hmsc <- NULL
@@
 ```
-<!-- LB: VAST does allow use of basis splines, but this requires the splines package and effects package for visualization, e.g.: https://github.com/James-Thorson-NOAA/VAST/wiki/Specify-covariates-and-visualize-responses -->
 
@@
 sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
-To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
+To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
 <!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
-<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
 Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
 spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
-<!-- TODO: still true? -->
 There is considerable variability in the available observation likelihoods across packages (Table 1).
-<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
-We provide comparisons of the syntax and the reproducibility of results from models fit using INLA (Appendix 4) and VAST (Appendix 5).
-<!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
-<!-- TODO: mention inlabru here -->
+We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix 4) and VAST (Appendix 5).
+
+(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
+Lines represent means and ribbons ranges across 20 random iterations.
+Note the log x and y axes.
+sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
+VAST should be similar to sdmTMB and so is not shown.
+inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
+mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
+<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
+<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
+<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
 
@@
 
-We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
 \@ref(fig:timing), Appendix 6).
-In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
+In this test, for mesh resolutions under 500 nodes, sdmTMB was faster than inlabru or mgcv.
 inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
@@
 With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
-Results with up to an order of magnitude more data slightly shift the lines up without affecting the qualitative pattern (not shown).
+Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown).
 Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
@@
 
-(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
-Lines represent means and ribbons ranges across 20 random iterations.
-Note the log x and y axes.
-sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
-inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
-mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
-<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
-<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
-<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
-
 # Discussion
-
-<!-- Take home, re-iterate paragraph? -->
-
-<!-- EW: TODO Not sure if Heaton included CAR models, but we might want to talk about them briefly because there’s a lot of packages that include that functionality - including NIMBLE, etc.  -->
 
@@
 -->
-Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
+Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
 VAST is the sole option for fitting some multivariate models;
@@
 
-<!-- Future directions para? -->
-<!-- multivariate, but only if can keep interface simple -->
-<!-- optimized Bayes sampling -->
-<!-- residual diagnostics -->
-<!-- zero inflation -->
 <!-- MVT? time-varying epsilon, non-gaussian fields -->
-
-<!-- limitations -->
-
 <!-- 3d models, continuous time, -->
-<!-- areal models, neighbours or not, must be CAR or SAR, ...
-but if have centroid, could use sdmTMB...
-why not CAR/SAR, spatial process...
-not 1:1 relationship between distance and spatial correlation?
-neibour based only...
--->
-<!-- https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.701.1636&rep=rep1&type=pdf -->
 
@@
 Heavily inspired by and in some places code has been adapted from VAST and glmmTMB...\
-Friendly/internal reviewers...
+Friendly/internal reviewers...\
+P.M. Regular, J. Watson\
+Splines based on mgcv
 
@@
 
--   emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference
-
 -   outline the philosophy and how the functions work together
@@
     -   avoid expensive calculations unless needed
-
--   case studies
-
-    -   an SDM with time-varying depth?
-    -   an index calculation?
-    -   stitching surveys?
-
--   speed comparisons (INLA [inlabru?], VAST, mgcv, glmmTMB without predictive process?)
 
@@
 :::
-
-```{=tex}
-\clearpage
-\beginsupplement
-```
-
-<!--
-
-### Junk to delete
-*EW: These are just placeholders -- not sure what order is best or what to highlight in very limited space*
-
-As a first example, we use a spatial dataset (`parana`) from the `geoR` package (Ribeiro Jr and Diggle 2018) that describes average May-June rainfall recorded at 143 stations in the state of Paraná, Brazil.
-This dataset is also well-known because it is used as an example of spatiotemporal modeling with INLA (cite INLA) book.
-Because the dataset doesn't include covariates and only includes a single year, we can model rainfall using an intercept and spatial component, $g(\mu_{s,t}) = b_{0} + \omega_{s}$.
-We assume the observation model for the rainfall data is normally distributed.
-The first step in fitting a model with `sdmTMB` is constructing the mesh to approximate the spatial process.
-This step can be done with the `make_mesh` function, which accepts the data (coordinates) and at minimum, the minimum allowed distance between points (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with INLA can also be passed in (additional examples of this are included in the SI).
-For example, an initial mesh with a cutoff distance of 0.1km could be made with
-
-```{r eval=FALSE, echo=TRUE}
-rain_spde = make_mesh(data = parana, c("east","north"), cutoff=0.1)
-```
-
-Second, we can fit the model using TMB with the `sdmTMB` function, whose formula syntax should be familiar to most R users,
-
-```{r eval = FALSE, echo=TRUE}
-fit = sdmTMB(data ~ 1, data=parana, spde = rain_spde)
-```
-
-As a second example, we use `sdmTMB` to analyze a 2-year dataset of Eurasian cranes (*Grus grus*) from England, UK (Soriano-Redondo et al.
-2019).
-In the original analysis, Soriano-Redondo et al.
-2019 used log Gaussian Cox process models to jointly model crane presence--absence and the spatiotemporal distribution of wetlands.
-We use this dataset as an example of modeling presence--absence data with sdmTMB, and only model the occurence of cranes given covariates.
-With any multi-year dataset, a common question of interest is often whether observations can be best described by spatial variation, or spatial and spatiotemporal variation.
-With two years of data, we fit two models with `sdmTMB`: a model with a spatial component but no spatiotemporal variation, and a model with only spatiotemporal variation (allowing each year to have a different spatial field).
-We included the same fixed effect covariates in both models: different intercepts by year, and two covariates found to be important by Soriano-Redondo et al.
-2019 (wetland area, perimeter to area ratio).
-As a coarse comparison of data support, we compare the Akaike's Information Criterion (Akaike 1973) from each model.
-The model with a shared spatial field appeared to be more supported ($\Delta AIC$=25.23).
-The sdmTMB package allows for additional model comparisons, including k-fold cross validation with the `sdmTMB_cv` function.
-
-Third more detailed example -- something like `pcod` or a fisheries dataset, with the Tweedie distribution, use of predict() function, and creating indices of abundance?
--->
-
-<!-- CAR/SAR models? and why not if have...-->
-<!-- EOF paper from Jim-->
-<!-- testing that has been done... -->
-<!-- https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.15149 -->
-<!-- https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.15149 -->

### 2022-01-06 doc/paper.Rmd
Fixes so paper builds

@@
 # simplify for paper
-pcod <- dplyr::select(pcod, year, density, depth, X, Y)
+# pcod <- dplyr::select(sdmTMB::pcod, year, density, depth, X, Y)
 ```
@@
 Simulation-based residuals [@dharma], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
-(5) The `simulate.sdmTMB()` method can simulate from fitted models [e.g., for DHARMa residuals, @DHARMa] and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
+(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
 <!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->

### 2022-01-06 doc/paper.Rmd
Fix opening sentence
Edit Abstract

@@
     | \noindent 1.
-Geostatistical data are spatially referenced observations of some spatial process---they might also be collected across time.
-Such data are common in ecology and can reveal ecological processes and improve conservation and management through applications such as species distribution models (SDMs).
+Geostatistical data are spatially referenced observations of some underlying spatial process---possibly collected over time.
+Such data are common in ecology and can reveal ecological processes and improve management through applications such as species distribution models (SDMs).
 However, appropriate models to analyze geostatistical data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
@@
 We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
-    | 4. We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
+    | 4. We hope that user-friendly interfaces such as sdmTMB will open this useful class of models to a wider audience of users for applications within species distribution modelling and beyond.
     | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder

### 2022-01-06 doc/paper.Rmd
Edit abstract

@@
 Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB).
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB) but scales to penalized likelihood or full Bayesian inference.
 We describe the basic functionality of sdmTMB and the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
     | 3.
-We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
-    | 4. We hope that user-friendly interfaces such as sdmTMB will open this useful class of models to a wider audience of users for applications within species distribution modelling and beyond.
+We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
+    | 4. We hope that user-friendly interfaces such as sdmTMB will open this useful class of models to a wider audience for applications within species distribution modelling and beyond.
     | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
@@
 
-<!-- TODO: emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference
+<!-- TODO: emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference-->
 

### 2022-01-06 doc/paper.Rmd
Word
abstract edits

@@
     | \noindent 1.
-Geostatistical data are spatially referenced observations of some underlying spatial process---possibly collected over time.
-Such data are common in ecology and can reveal ecological processes and improve management through applications such as species distribution models (SDMs).
+Geostatistical data are spatially referenced observations influenced by some underlying spatial process---possibly collected over time.
+Such data are common in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
 However, appropriate models to analyze geostatistical data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
@@
 Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB) but scales to penalized likelihood or full Bayesian inference.
-We describe the basic functionality of sdmTMB and the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but scales to penalized likelihood or Bayesian inference.
+We describe the components of sdmTMB's  statistical models, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
     | 3.
-We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
-    | 4. We hope that user-friendly interfaces such as sdmTMB will open this useful class of models to a wider audience for applications within species distribution modelling and beyond.
+We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
+    | 4. We hope that sdmTMB's user-friendly interface will help open this useful class of models to a wider audience for applications within species distribution modelling and beyond.
     | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder

### 2022-01-06 doc/paper.Rmd
Edit abstract

@@
     | \noindent 1.
-Geostatistical data are spatially referenced observations influenced by some underlying spatial process---possibly collected over time.
+Geostatistical data are spatially referenced observations of some spatial process---possibly collected over time.
 Such data are common in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
@@
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but scales to penalized likelihood or Bayesian inference.
-We describe the components of sdmTMB's  statistical models, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+We describe sdmTMB and statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
     | 3.
 We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
-    | 4. We hope that sdmTMB's user-friendly interface will help open this useful class of models to a wider audience for applications within species distribution modelling and beyond.
+    | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
     | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder

### 2022-01-06 doc/paper.Rmd
Edit 1st sentence

@@
     | \noindent 1.
-Geostatistical data are spatially referenced observations of some spatial process---possibly collected over time.
-Such data are common in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
-However, appropriate models to analyze geostatistical data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
+Geostatistical data are common in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
+However, appropriate models to analyze such data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
     | 2.

### 2022-01-06 doc/paper.Rmd
extended etc

@@
     | \noindent 1.
-Geostatistical data are common in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
+Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon--are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
 However, appropriate models to analyze such data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
@@
 Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but scales to penalized likelihood or Bayesian inference.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but can be extended to penalized likelihood or Bayesian inference.
 We describe sdmTMB and statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.

### 2022-01-06 doc/paper.Rmd
Tweak table
combine inla;inlabru

@@
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but can be extended to penalized likelihood or Bayesian inference.
-We describe sdmTMB and statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
     | 3.
@@
 
+d$INLA <- NULL
+names(d)[names(d) == "inlabru"] <- "INLA/inlabru"
+
 knitr::kable(d, format = "latex", caption = "Comparison of functionality between seven R packages that can fit geostatistical GLMMs.",
@@
       "Penalized smoother GAMs",
+      "inlabru but not INLA",
       "Spatiotemporal fields as random walk only",
@@
       "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}",
-      "Technically possible but non-trivial",
+      "Hurdle models possible by fitting components separately",
+      "Hurdle models possible by fitting components separately",
       "Technically possible but non-trivial"
@@
 Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown).
-Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
+Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
 Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.

### 2022-01-06 doc/appendix-model.Rmd
Lewis edits

@@
 * $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
-* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
+* $\omega_{\bm{s}}$ represents a spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
 * $\epsilon_{\s,t}$ represents a spatiotemporal component (a random field), $\epsilon_{\s,t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.
@@
 $$
-s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
+s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln \left(19\right) \cdot \left( x-s50 \right)
      / \left(s95 - s50 \right) } \right] }^{-1},
@@
 $s50$ is a parameter controlling the point at which the function reaches 50% of the maximum ($\psi$), 
-and $s95$ is a parameter controlling the point at which the function reaches 95%. 
+and $s95$ is a parameter controlling the point at which the function reaches 95% of the maximum. 
 The parameter $s50$ is unconstrained but $s95$ is constrained to be larger than $s50$.
@@
 
-Similarly to the spatial random fields, these spatiotemporal random fields (and all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):
+Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):
 
 $$
-\boldsymbol{\epsilon_{t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
+\boldsymbol{\epsilon_{t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right).
 $$
@@
 
+## Time-varying regression parameters
+
+Parameters can be modelled as time varying (a random walk) according to the form
+
+$$
+\begin{aligned}
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} \bm{X}_{2,s,t} + \ldots \right),\\
+  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
+  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
+\end{aligned}
+$$
+
+where $\gamma_t$ is an example optional time-varying regression parameter and $x_{\s,t}$ is the corresponding predictor value.
+This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
+For example, `time_varying ~ 0 + x`.
+The first value is given an implicit uniform prior. 
+I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
+The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and, this case, the intercept should be omitted from the main effects (`formula ~ + 0 + ...` or `formula ~ -1 + ...`).
+
 ## Spatially varying coefficients (SVC)
@@
 
-## Time-varying regression parameters
-
-Parameters can be modelled as time varying (a random walk) according to the form
-
-$$
-\begin{aligned}
-  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} x_{\s,t} + \ldots \right),\\
-  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
-  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
-\end{aligned}
-$$
-
-where $\gamma_t$ is an example optional time-varying regression parameter and $x_{\s,t}$ is the corresponding predictor value.
-This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
-For example, `time_varying ~ 0 + x`.
-The first value is given an implicit uniform prior. 
-I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
-The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and, this case, the intercept should be omitted from the main effects (`formula ~ + 0 + ...` or `formula ~ -1 + ...`).
-
 ## IID random or multi-level intercepts
@@
 There can be multiple random intercepts, despite only showing one above.
-E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.
+E.g., `(1 | g1) + (1 | g2)`, in which case they are assumed independent and uncorrelated from each other.
 
@@
 $$
-where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
+where $N$ is the size or number of trials, and $\mu$ is the probability of success for each trial.
 If $N = 1$, the distribution becomes the Bernoulli distribution.
@@
 version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
-<!-- EW: Put lme4 and glmmTMB in '' too?-->
 Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:
@@
 $$
-<!-- EW: betareg in ''?-->
 where $\mu$ is the mean and $\phi$ is a precision parameter.
@@
 where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
-The Tweedie distribution can be helpful for modelling data that positive and continuous but also contains zeros.
+The Tweedie distribution can be helpful for modelling data that are positive and continuous but also contains zeros.
 
@@
 
-## Matern parameterization
+## Matérn parameterization
 
@@
 There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$.
-The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.
+The elements of $\textbf{H}$ are defined by the parameter vector $\boldsymbol{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.
 
@@
 
-In some cases the spatial domain of interest may be complex and bordered by land or water (e.g., coastlines, islands, lakes). 
+In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes). 
 SPDE models allow for barriers to be included via barrier meshes [@bakka2019]. 

### 2022-01-07 doc/appendix-model.Rmd
Minor edits

@@
   "$\\epsilon_{\\s, t}^*$", "\\texttt{epsilon\\_st\\_A}", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
-  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR1 or random walk spatiotemporal deviations (knot)",
+  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR(1) or random walk spatiotemporal deviations (knot)",
   "$\\alpha_g$", "\\texttt{RE}", "IID random intercept deviation for group $g$",
@@
 
-The complete sdmTMB model can be written as:
+The complete sdmTMB model can be written as
 
@@
 
-Internally, the random fields follow a Gaussian Markov random field (GMRF):
+Internally, the random fields follow a Gaussian Markov random field (GMRF)
 
 $$
-\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
+\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right),
 $$
@@
 
-Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
+Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (independent and identically distributed, `'iid'`; default), AR(1) (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
 These text values are case insensitive.
@@
 
-Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):
+Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$)
 
@@
 
-The random fields are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\zeta$):
+The random fields are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\zeta$)
 
@@
 where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
-The Tweedie distribution can be helpful for modelling data that are positive and continuous but also contains zeros.
+The Tweedie distribution can be helpful for modelling data that are positive and continuous but also contain zeros.
 
@@
 
-The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates from the knot locations to the values at the locations of the observed or predicted data [@lindgren2015]:
+The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates from the knot locations to the values at the locations of the observed or predicted data [@lindgren2015]
 
@@
 The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
-Three non-zero elements on each row define the weight of the neighbouring 3 knot locations for location $s$.
-The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$:
+Three non-zero elements on each row define the weight of the neighbouring 3 knot locations for location $\s$.
+The same bilinear interpolation happens for any spatiotemporal random fields
 

### 2022-01-07 doc/paper.Rmd
Minor edits

@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon--are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
+Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
 However, appropriate models to analyze such data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.

### 2022-01-07 doc/appendix-snow-sdmTMB.Rmd
owl appendix edits; optional cv section currently left out

@@
 
-# Introduction
+# Background
 
@@
 
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
 The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
@@
 
-```{r mesh, echo=TRUE, cache=FALSE, out.width="5.5in"}
+```{r mesh, echo=TRUE, cache=FALSE, out.width="70%"}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-# Model fitting
+# Model fitting with `sdmTMB`
 
@@
 <!-- We will use `normalize = TRUE` to speed up optimization.  -->
-Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.
-
+<!-- Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output. -->
 Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America.
@@
 If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
+It might also be appropriate to include count ID as another random effect (`(1 | CBCID)`) to account for consistent differences in habitat and human observers between count circles. 
+But since its inclusion doesn't change our main result and it slows model fitting, we will rely on the spatial random effects to link observations in the same location for this example.
+
 Since these models will take a while to fit, in practice, we might set `silent = FALSE` to monitor optimization progress. 
@@
 
+# Model validation
+
+<!-- ## Basic inspection -->
 ```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
@@
 According to AIC, the `nbinom2()` (negative binomial, NB2) distribution appears to have the most parsimonious fit. 
-We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model.
+We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and check that the gradients are all small enough.
+If the gradients aren't quite small enough, we can easily run additional optimization steps using the `run_extra_optimization()` function.
 
@@
 
-```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE}
+```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE, cache=FALSE}
 m <- sdmTMB(count ~ 1 + nao + (1 | year_f),
@@
 )
+
+if (max(m$gradients) > 0.01) {
+  m <- run_extra_optimization(m, newton_loops = 1L, nlminb_loops = 0L)
+}
 ```
@@
 
-```{r check, echo=FALSE, include = FALSE, eval=TRUE}
+```{r check, echo=FALSE, include=FALSE, eval=TRUE, cache=FALSE}
 if (rds_saved2) {
@@
 
-```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE}
+```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
 print(m)
@@
 
-```{r tidy}
+```{r tidy, cache=FALSE}
 tidy(m, conf.int = TRUE)
@@
 <!-- \clearpage -->
+
+<!-- ## Checking residuals -->
+
 We can inspect randomized quantile residuals, although this method has known issues.
 
-```{r qqnorm, out.width = "33%"}
+```{r qqnorm, out.width="33%"}
 qqnorm(pred$resid)
@@
 
-```{r stan-qqnorm, out.width = "33%"}
+```{r stan-qqnorm, out.width="33%"}
 qqnorm(r_mcmc)
@@
 
-```{r plot-resids, fig.width = 8}
+```{r plot-resids, fig.width=8}
 ggplot(pred, aes(X, Y, colour = resid)) +
@@
 
+<!-- ## Cross-validation -->
+
+<!-- Cross-validation can also be used to assess model performance. -->
+<!-- Here, we use the `future` package to speed up the running of multiple models and save a smaller version of the output that doesn't include the full model information. -->
+<!-- We use `k_folds = 5`, which amounts to training on 80% of the data and testing with 20% for each model. -->
+<!-- Block cross-validation, which can be done using custom folds defined using the `fold_ids` argument, is generally recommended for spatiotemporal data. -->
+<!-- However, random folds created by `sdmTMB_cv()` are appropriate in this case, because we are only interested in predicting within the spatial and temporal range of our data and are attempting to estimate the spatial variation in the effect of NAO.
+-->
+
+<!-- ```{r cv, eval=FALSE, cache=FALSE} -->
+<!-- library(future) -->
+<!-- plan(multisession) -->
+<!-- options(future.rng.onMisuse = "ignore") -->
+
+<!-- m_cv <- sdmTMB_cv(count ~ 1 + nao + (1 | year_f), -->
+<!--   spatial_varying = ~ 0 + nao, -->
+<!--   time = "year", -->
+<!--   spatial = "on", -->
+<!--   spatiotemporal = "iid", -->
+<!--   family = nbinom2(link = "log"), -->
+<!--   reml = TRUE, -->
+<!--   use_initial_fit = TRUE, -->
+<!--   data = snow, mesh = mesh, -->
+<!--   k_folds = 5 -->
+<!-- ) -->
+<!-- plan(sequential) -->
+<!-- saveRDS(m_cv, file = here::here("data/owlcvallreml.rds")) -->
+<!-- # m_cv$models[[1]] # to retrieve just the model built excluding the first fold -->
+<!-- cv <- m_cv -->
+<!-- cv$models <- NA -->
+<!-- saveRDS(cv, file = here::here("data/owlcvdatreml.rds")) -->
+<!-- ``` -->
+
+<!-- First, we should check that all the models converged and have gradients < 0.01. -->
+
+<!-- ```{r explore-cv, message=FALSE, cache=FALSE} -->
+<!-- cv <- readRDS(here::here("data/owlcvdatreml.rds")) -->
+<!-- cv$converged -->
+<!-- max(cv$max_gradients) -->
+<!-- ``` -->
+
+<!-- If we were conducting model selection, we might contrast the log likelihoods (`cv$sum_loglik`), and the expected log predictive density (`cv$elpd`) for the left out data between different model configurations instead of relying on AIC.
+-->
+<!-- We can also check the prediction accuracy on left-out data by plotting the `cv_predicted` against the observed `count` in log space. -->
+
+<!-- ```{r cv-pred-obs, out.width="70%", cache=FALSE} -->
+<!-- cvdat <- cv$data %>% filter(cv_loglik > -60) # this removes one extreme outlier -->
+<!-- ggplot(cvdat, aes(log(count + 1), log(cv_predicted + 1), colour = -cv_loglik)) + -->
+<!--   geom_abline(slope = 1, colour = "gray") + -->
+<!--   geom_point(size = 2, alpha = 0.25) + -->
+<!--   scale_colour_viridis_c( -->
+<!--     direction = -1, option = "A", begin = 0.2, end = 0.9 -->
+<!--   ) + -->
+<!--   coord_fixed(ylim = c(0, 3.8)) + -->
+<!--   theme_bw() -->
+<!-- ``` -->
+
+<!-- These predictions for left-out locations match the observed counts just as well as those based on the model built using all the data shown below. -->
+<!-- # Does this mean that we are at least not over-fit, despite the spatiotemporal random fields and spatially varying coefficient? -->
+
+<!-- ```{r full-pred-obs, echo=FALSE, out.width="70%"} -->
+<!-- ggplot(pred, aes(log(count + 1), log(exp(est) + 1), colour = resid)) + -->
+<!--   geom_abline(slope = 1, colour = "gray") + -->
+<!--   # geom_point(size = 0.75, alpha=0.5) + facet_wrap(~year) + -->
+<!--   geom_point(size = 2, alpha = 0.25) + -->
+<!--   coord_fixed(ylim = c(0, 3.8)) + -->
+<!--   scale_colour_viridis_c( -->
+<!--     "residuals", -->
+<!--     direction = -1, option = "A", begin = 0.2, end = 0.9 -->
+<!--   ) + -->
+<!--   theme_bw() #+ theme(strip.background = element_rect(fill = NA, colour = NA)) -->
+<!-- ``` -->
+
+<!-- We can also explore the spatiotemporal variation in the predictive accuracy of our model by plotting the `cv_loglik`. -->
+
+<!-- ```{r cv-map, cache = FALSE} -->
+<!-- ggplot(cvdat, aes(X, Y, colour = -cv_loglik)) + -->
+<!--   geom_point(size = 0.25) + -->
+<!--   scale_colour_viridis_c( -->
+<!--     direction = -1, option = "A", begin = 0.2, end = 0.9, -->
+<!--     guide = guide_colourbar(title.position = "top") -->
+<!--   ) + -->
+<!--   coord_fixed() + -->
+<!--   theme(axis.text = element_blank(), axis.ticks = element_blank()) + -->
+<!--   facet_wrap(~year) -->
+<!-- ``` -->
+
+\clearpage
+
 # Plotting spatial predictions
@@
 p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
-p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
+p_proj <- p %>% mutate(x = X, y = Y) %>%
+  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
 ```

### 2022-01-07 doc/paper.Rmd
Word

@@
 Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
-However, appropriate models to analyze such data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
+However, appropriate models to analyze these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
     | 2.

### 2022-01-08 doc/appendix-snow-sdmTMB.Rmd
add maps of upper and lower CI for spatially varying effect

@@
 library(ggplot2)
+library(patchwork)
 library(dplyr)
@@
 
-\clearpage
+<!-- \clearpage -->
 
@@
 We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
-First, we extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.
+Most simply, we can extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.
+However, this method does not give us spatially explicit estimates of uncertainty.
+To do this, we can use the output from the joint precision matrix simulation draws by using the `predict()` function again but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`).
+We also extract the same number of drams for (this time `n_sims = 500`) in the `spread_sims()` function for the main effect of NAO. 
+The simulation draws for both should then be combined before adding the median and lower and upper quantiles from these draws to our prediction data frame.
 
@@
 b <- tidy(m)
+
+n_sims <- 500L
+zeta_sim <- predict(m, sims = n_sims, sims_var = "zeta_s")
+sims <- spread_sims(m, n_sims = n_sims)
+beta <- sims$nao
+combined <- beta + t(zeta_sim)
+
+p$zeta_sim <- apply(t(zeta_sim), 2, median)
+p$nao_effect_sim <- apply(combined, 2, median)
+p$nao_effect_lwr <- apply(combined, 2, quantile, probs = 0.10)
+p$nao_effect_upr <- apply(combined, 2, quantile, probs = 0.90)
+
 p_mean <- p %>%
-  group_by(CBCID) %>%
-  summarise(
+  group_by(CBCID) %>% summarise(
+    year = max(year),
     X = mean(X), Y = mean(Y),
     mean_est_count = exp(mean(est)),
-    nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
+    nao_effect = b[b$term == "nao", 2] + mean(zeta_s),
+    nao_sim = mean(nao_effect_sim),
+    nao_lwr = mean(nao_effect_lwr),
+    nao_upr = mean(nao_effect_upr)
   )
@@
 
-```{r plot-map, echo=TRUE, eval=TRUE}
+```{r plot-map, echo=TRUE, eval=TRUE, out.width="100%"}
 ggplot(data = p_proj) +
@@
   ) +
-  scale_colour_viridis_c(guide = guide_colourbar(
-    direction = "horizontal", title.position = "top"
-  )) +
+  scale_colour_viridis_c(
+    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
+    guide = guide_colourbar(direction = "horizontal", title.position = "top")
+  ) +
   guides(size = "none") +
-  labs(colour = "NAO effect\non Snowy Owl count") +
-  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
+  labs(colour = "Estimated effect in log space") +
+  ggtitle("Map of combined main and spatially varying effect of NAO on Snowy Owl count") +
+  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
 ```
 
-\clearpage
+```{r plot-map-ci, echo=F, eval=TRUE, out.width="100%"}
+p1 <- ggplot(data = p_proj) +
+  geom_point(
+    data = p_mean,
+    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
+  ) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
+  scale_colour_viridis_c(
+    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
+    guide = guide_colourbar(direction = "horizontal", title.position = "top")
+  ) +
+  guides(size = "none", colour = "none") +
+  # labs(colour = "Lower 90% CI") +
+  ggtitle("Lower 80% CI") +
+  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
+
+p2 <- ggplot(data = p_proj) +
+  geom_point(
+    data = p_mean,
+    aes(X, Y, colour = nao_upr, size = mean_est_count), alpha = 0.5
+  ) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  coord_sf(
+    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
+    ylim = c(min(p_proj$Y), max(p_proj$Y))
+  ) +
+  scale_colour_viridis_c(
+    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
+    guide = guide_colourbar(direction = "horizontal", title.position = "top")
+  ) +
+  guides(size = "none", colour = "none") +
+  # labs(colour = "Upper 90% CI") +
+  ggtitle("Upper 80% CI") +
+  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
+
+p1 + p2 + patchwork::plot_layout()
+```
+
+<!-- \clearpage -->
 

### 2022-01-08 doc/appendix-snow-sdmTMB.Rmd
owl appendix heading changes and minor text edits

@@
 
-# Model fitting with `sdmTMB`
+# Model structure
 
@@
 If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
-It might also be appropriate to include count ID as another random effect (`(1 | CBCID)`) to account for consistent differences in habitat and human observers between count circles. 
+It might also be appropriate to include count ID as another random effect `(1 | CBCID)` to account for consistent differences between count circles in habitat and human observers. 
 But since its inclusion doesn't change our main result and it slows model fitting, we will rely on the spatial random effects to link observations in the same location for this example.
@@
 
-<!-- ## Basic inspection -->
+## Basic inspection
 ```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
@@
 
-<!-- ## Checking residuals -->
+## Checking residuals
 

### 2022-01-08 doc/appendix-validation.Rmd
Develop timing appendix; edit paper

--- 2021-12-22 doc/appendix-validation.Rmd
@@
 ---
-title: "Appendix 6: sdmTMB model validation and speed-testing methods"
+title: "Appendix 6: sdmTMB speed testing methods and model validation description"
 output:
   bookdown::pdf_document2:
+    highlight: pygments
     toc: true
+    number_sections: true
 number_sections: false
@@
   comment = "#>",
-  echo = FALSE,
+  echo = TRUE,
   error = FALSE,
@@
   cache.comments = FALSE
+)
+```
+
+# Supplemental methods for speed testing related packages
+
+Here we describe the methods underlying the speed testing in Figure 4.
+
+We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions.
+Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
+The Gaussian random field was parameterized with a range of 0.5 and marginal standard deviation of 0.2.
+The model included an intercept with a value of 0.2 and a normal(0, 1) predictor with an associated coefficient of -0.4.
+The observation error was Gaussian with a standard deviation of 0.3.
+
+We generated a mesh that was consistent across simulated data sets for a given mesh resolution.
+We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
+We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
+In Figure 4, we report on the x-axis the number of mesh nodes ("knots") that result from each of these meshes.
+
+We conducted 20 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.
+
+## Illustration of the simulation components
+
+### Illustration of generating the INLA mesh
+
+```{r pkgs, warning=FALSE, message=FALSE}
+library(INLA)
+library(inlabru)
+library(ggplot2)
+library(sdmTMB)
+library(mgcv)
+library(spaMM)
+```
+
+We will illustrate with a `max.edge` of 0.1:
+
+```{r mesh}
+max_edge <- 0.06
+loc_bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
+segm_bnd <- INLA::inla.mesh.segment(loc_bnd)
+mesh <- INLA::inla.mesh.2d(
+  boundary = segm_bnd,
+  max.edge = c(max_edge, 0.2),
+  offset = c(0.1, 0.05)
+)
+```
+
+This mesh has `r mesh$n` (`mesh$n`) vertices. We can visualize it:
+
+```{r mesh-vis1, out.width="4in"}
+ggplot() +
+  gg(mesh) +
+  theme_light() +
+  coord_equal(expand = FALSE)
+```
+
+Here are the other meshes used from least to most vertices:
+
+```{r mesh-vis, out.width="6.5in", fig.width=10, fig.asp=1.2}
+max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
+g <- list()
+for (i in seq_along(max_edge_vec)) {
+  .mesh <- INLA::inla.mesh.2d(
+    boundary = segm_bnd,
+    max.edge = c(max_edge_vec[i], 0.2),
+    offset = c(0.1, 0.05)
+  )
+  g[[i]] <- ggplot() +
+    gg(.mesh) +
+    theme_light() +
+    coord_equal(expand = FALSE) +
+    ggtitle(paste("Mesh vertices =", .mesh$n))
+}
+cowplot::plot_grid(plotlist = g, ncol = 2)
+```
+
+### Illustration of simulating data
+
+```{r sim}
+set.seed(1)
+
+n_obs <- 1000L
+predictor_dat <- data.frame(
+  X = runif(n_obs), Y = runif(n_obs),
+  a1 = rnorm(n_obs)
+)
+mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)
+
+sim_dat <- sdmTMB_simulate(
+  formula = ~ 1 + a1,
+  data = predictor_dat,
+  mesh = mesh_sdmTMB,
+  family = gaussian(),
+  range = 0.5,
+  phi = 0.3,
+  sigma_O = 0.2,
+  B = c(0.2, -0.4) # B0 = intercept, B1 = a1 slope
+)
+```
+
+Look at the simulated data:
+
+```{r sim-dat-plot, out.width="4.5in"}
+ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
+  geom_point() +
+  scale_color_gradient2() +
+  theme_light()
+```
+
+### Example sdmTMB model fits
+
+```{r sdmTMBfit}
+fit_sdmTMB <- sdmTMB(
+  observed ~ a1,
+  data = sim_dat, mesh = mesh_sdmTMB, family = gaussian(),
+  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2))
+)
+
+# normalize = TRUE:
+fit_sdmTMB2 <- sdmTMB(
+  observed ~ a1,
+  data = sim_dat, mesh = mesh_sdmTMB, family = gaussian(),
+  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)),
+  control = sdmTMBcontrol(normalize = TRUE)
+)
+```
+
+### Example spaMM model fit
+
+```{r spaMMfit}
+spde <- INLA::inla.spde2.pcmatern(
+  mesh = mesh,
+  prior.range = c(0.05, 0.05),
+  prior.sigma = c(2, 0.05)
+)
+
+fit_spaMM <- fitme(
+  observed ~ a1 + IMRF(1 | X + Y, model = spde),
+  family = gaussian(), data = sim_dat
+)
+```
+
+### Example inlabru model fit
+
+```{r inlabrufit}
+# convert to sp SpatialPointsDataFrame first:
+dat_sp <- sp::SpatialPointsDataFrame(
+  cbind(sim_dat$X, sim_dat$Y),
+  proj4string = sp::CRS(
+    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
++ +y_0=0 +datum=NAD83 +units=km +no_defs"
+  ), data = sim_dat
+)
+components <- observed ~ -1 + Intercept(1) + a1 +
+  spatrf(main = coordinates, model = spde)
+like <- like(observed ~ Intercept + a1 + spatrf,
+  family = "gaussian", data = dat_sp
+)
+fit_bru <- bru(
+  like,
+  components = components,
+  options = bru_options(
+    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
+    bru_max_iter = 1, num.threads = "1:1"
+  )
+)
+```
+
+### Example mgcv model fit
+
+```{r mgcvfit}
+# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth():
+# from supplement of:
+# Miller, D.L., Glennie, R. & Seaton, A.E. (2019). Understanding the Stochastic
+# Partial Differential Equation approach to smoothing. Journal of Agricultural,
+# Biological and Environmental Statistics.
+source(here::here("analysis/mgcv_spde_smooth.R"))
+library(mgcv)
+
+fit_bam <- bam(
+  observed ~ a1 + s(X, Y,
+    bs = "spde", k = mesh$n,
+    xt = list(mesh = mesh)
+  ),
+  data = sim_dat,
+  family = gaussian(),
+  method = "fREML",
+  control = gam.control(scalePenalty = FALSE), discrete = TRUE
 )
@@
 This appendix illustrates a comparison of penalized smoothers with mgcv and a comparison of IID random intercepts with glmmTMB.
-Additionally, it illustrates some simulation testing of a spatial random field model with the main observation families within the sdmTMB package.
+<!-- Additionally, it illustrates some simulation testing of a spatial random field model with the main observation families within the sdmTMB package. -->
 
@@
 
-```{r, echo=TRUE}
+```{r lib-load2, echo=TRUE}
 library(sdmTMB)
@@
 
-```{r, echo=TRUE}
+```{r mgcv-eg1, echo=TRUE}
 set.seed(19203)
@@
 
-```{r, echo=TRUE, fig.asp=1}
+```{r plot-smooth, echo=TRUE, fig.asp=1}
 par(mfrow = c(2, 1))
@@
 
-```{r, echo=TRUE}
+```{r continuous-by-compare, echo=TRUE}
 set.seed(19203)
@@
 
-```{r, echo=TRUE}
+```{r bivariate-compare, echo=TRUE}
 m_mgcv <- mgcv::gam(y ~ s(x2, x1), data = dat)
@@
 
-```{r, echo=TRUE}
+```{r iid-sim, echo=TRUE}
 predictor_dat <- data.frame(X = runif(1000), Y = runif(1000))
@@
 sim_dat <- sdmTMB_simulate(
-  formula = ~ 1,
+  formula = ~1,
   data = predictor_dat,
@@
 
-```{r, echo=TRUE}
-m <- sdmTMB(observed ~ 1 + (1 | g) + (1 | h), mesh = mesh, data = sim_dat, 
-  spatial = "off")
+```{r iid-fit, echo=TRUE}
+m <- sdmTMB(observed ~ 1 + (1 | g) + (1 | h),
+  mesh = mesh, data = sim_dat,
+  spatial = "off"
+)
 
@@
 
-```{r, echo=TRUE}
+```{r iid-compare, echo=TRUE}
 sdmTMB_pars <- tidy(m, "ran_pars")
@@
 plot(c(glmmTMB_re$g$`(Intercept)`, glmmTMB_re$h$`(Intercept)`), sdmTMB_re$RE,
-  col = c(rep(1, 50), rep(2, 20)))
+  col = c(rep(1, 50), rep(2, 20))
+)
 abline(0, 1)
@@
 
-## Simulation-testing with various observation families
+<!-- ## Simulation-testing with various observation families -->
 
-```{r}
+```{r, eval=FALSE,include=FALSE}
 knitr::include_graphics(here::here("figs", "sim-test-families-pars.pdf"))
@@
 
-```{r}
+```{r, eval=FALSE,include=FALSE}
 knitr::include_graphics(here::here("figs", "sim-test-families-b1.pdf"))
@@
 
-```{r}
+```{r, eval=FALSE,include=FALSE}
 knitr::include_graphics(here::here("figs", "sim-test-families-phi.pdf"))
@@
 
-```{r}
+```{r, eval=FALSE,include=FALSE}
 knitr::include_graphics(here::here("figs", "sim-test-families-range.pdf"))
@@
 
-```{r}
+```{r, eval=FALSE,include=FALSE}
 knitr::include_graphics(here::here("figs", "sim-test-families-sigmaO.pdf"))
@@
 
-# Supplemental methods for speed-testing related packages
-
-TODO: Fig. 2 methods description
+# References

### 2022-01-08 doc/paper.Rmd
Develop timing appendix; edit paper

@@
 ---
-title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
+title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal random field generalized linear mixed-effects models in R"
 author: "|
@@
   $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
+      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
+    Running headline: sdmTMB: spatial and spatiotemporal GLMMs\n"
 output:
@@
     | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
-    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
+    | Keywords:
+      Gaussian Markov random fields,
+      generalized linear mixed effects models (GLMM),
+      INLA,
+      SPDE,
+      species distribution modelling,
+      R package,
+      spatio-temporal or spatial-temporal,
+      Template Model Builder
 ---
@@
 ```
+<!-- sdmTMB: spatial and spatiotemporal random field generalized linear mixed-effects models in R in a fast, flexible, and user-friendly interface -->
 
@@
 
-TODO\
-Would not be possible without TMB and INLA and authors...\
-Heavily inspired by and in some places code has been adapted from VAST and glmmTMB...\
-Friendly/internal reviewers...\
-P.M. Regular, J. Watson\
-Splines based on mgcv
+sdmTMB would not be possible without the TMB [@kristensen2016] and INLA [@rue2009; @lindgren2011; @lindgren2015] R packages.
+sdmTMB is heavily inspired by and in some places code has been adapted from both the VAST [@thorson2019] and glmmTMB [@brooks2017] R packages.
+Penalized spline support was possible thanks to mgcv [@wood2017a].
+We thank the authors of all these packages.
+We thank P.M. Regular, J. Watson, ..., TODO.
 
@@
 
-TODO: None?
+The authors declare no conflict of interest.
 
@@
 
-TODO
+SA led development of the software and EW contributed key software code. All
+authors contributed software code, guided software development, and
+participated in testing. PE led the Snowy Owl case study. All authors
+contributed to appendix code. SA led the writing of the manuscript. All authors
+contributed critically to the drafts and gave final approval for publication.
 
@@
 Appendix 4: Species distribution model comparison between INLA and sdmTMB\
-Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB\
-Appendix 6: sdmTMB model validation and speed-testing methods
+Appendix 5: Comparison of index standardization of survey data with VAST and sdmTMB\
+Appendix 6: sdmTMB speed testing methods and model validation description
+
 <!--
@@
 
-TODO\
-Will be on CRAN\
-Paper source code will be on Zenodo\
-<https://github.com/pbs-assess/sdmTMB>\
-<https://github.com/seananderson/sdmTMB-paper>
+A development version of sdmTMB is available at <https://github.com/pbs-assess/sdmTMB>.
+sdmTMB will be available on CRAN.
+Paper source code will be available on Zenodo and at <https://github.com/seananderson/sdmTMB-paper> on publication.
 

### 2022-01-08 analysis/timing.R

@@ -327,27 +327,27 @@ to_run <- to_run[sample(seq_len(nrow(to_run)), replace = FALSE), ]
 
 # out <- purrr::pmap_dfr(to_run, sim_fit_time)
 
-plan(multisession, workers = 5L)
+# plan(multisession, workers = 5L)
 
-out <- furrr::future_pmap_dfr(
-  # out <- purrr::pmap_dfr(
+# out <- furrr::future_pmap_dfr(
+out <- purrr::pmap_dfr(
   to_run,
-  sim_fit_time,
-  .progress = TRUE,
-  .env_globals = parent.frame(),
-  .options = furrr::furrr_options(
-    seed = TRUE,
-    globals = c(
-      "simulate_dat", "sim_fit_time",
-      "Predict.matrix.spde.smooth", "smooth.construct.spde.smooth.spec"
-    ),
-    packages = c("mgcv", "inlabru", "INLA", "ggplot2", "dplyr", "spaMM", "sdmTMB")
-  )
+  sim_fit_time
+  # .progress = TRUE,
+  # .env_globals = parent.frame(),
+  # .options = furrr::furrr_options(
+  #   seed = TRUE,
+  #   globals = c(
+  #     "simulate_dat", "sim_fit_time",
+  #     "Predict.matrix.spde.smooth", "smooth.construct.spde.smooth.spec"
+  #   ),
+  #   packages = c("mgcv", "inlabru", "INLA", "ggplot2", "dplyr", "spaMM", "sdmTMB")
+  # )
 )
 
 saveRDS(out, file = "analysis/timing-cache-parallel-spaMM-PE.rds")
 out <- readRDS("analysis/timing-cache-parallel-spaMM-PE.rds")
-plan(sequential)
+# plan(sequential)
 
 # # -------tweedie
 # to_run <- expand.grid(

### 2022-01-08 doc/appendix-validation.Rmd
Update tau_g

@@
 
-sdmTMB_pars$estimate[sdmTMB_pars$term == "tau_G"][1]
+sdmTMB_pars$estimate[sdmTMB_pars$term == "sigma_G"][1]
 sqrt(as.numeric(glmmTMB_varcorr$cond$g))
 
-sdmTMB_pars$estimate[sdmTMB_pars$term == "tau_G"][2]
+sdmTMB_pars$estimate[sdmTMB_pars$term == "sigma_G"][2]
 sqrt(as.numeric(glmmTMB_varcorr$cond$h))

### 2022-01-08 doc/appendix-model.Rmd
Develop timing appendix; edit paper
Fix zeta typo

@@
   "$\\omega_{\\s}^*$", "\\texttt{omega\\_s\\_A}", "Spatial random field at point $\\s$ (interpolated)",
-  "$\\zeta_{\\s, t}$", "\\texttt{zeta\\_t}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (knot)",
-  "$\\zeta_{\\s, t}^*$", "\\texttt{zeta\\_t\\_A}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (interpolated)",
+  "$\\zeta_{\\s}$", "\\texttt{zeta\\_s}", "Spatially varying coefficient random field at point $\\s$ (knot)",
+  "$\\zeta_{\\s}^*$", "\\texttt{zeta\\_s\\_A}", "Spatially varying coefficient random field at point $\\s$ (interpolated)",
   "$\\epsilon_{\\s, t}$", "\\texttt{epsilon\\_st}", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",

### 2022-01-08 doc/paper.Rmd
Title

@@
 ---
-title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal random field generalized linear mixed-effects models in R"
+title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
 author: "|

### 2022-01-08 doc/paper.Rmd
define irruptive migrant

@@
 
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are
-irruptive
-migrants appearing across mid latitudes of North America in winter.
-<!-- TODO: not familiar with this "irruptive" terminology, please define -->
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
 The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
@@
 Users familiar with INLA may find inlabru approachable.
-<!-- ; however, we ran into problems with AR1 spatiotemporal fields in inlabru, so we only provide a comparison with INLA instead (Appendix 4).
--->
 Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.

### 2022-01-08 doc/appendix-binomial-inla.Rmd
refine order and minor edits to inla appendix

@@
 ```
+
+# Background
 
@@
 tictoc::toc() # stop timing
+```
 
+```{r}
 summary(m_inla)
-inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
 ```
@@
 summary(m_bru)
+```
 
-m_inla$summary.hyperpar[, c("mean", "sd")]
-m_bru$summary.hyperpar[, c("mean", "sd")]
-
-bru_coefs <- m_bru$summary.random$fac[, c("ID", "mean", "sd")]
-m_inla$summary.fixed[,c("mean", "sd")]
-bru_coefs
+```{r}
+m_bru$summary.random$fac[, c("ID", "mean", "sd", "0.025quant", "0.975quant")]
 ```
@@
 
+```{r sdmtmb-m}
+print(m)
+```
+
+We can most easily extract the fixed-effect estimates and confidence intervals using smdTMB's `tidy()` function.
+
 ```{r sdmtmb-coefs}
-print(m)
-tidy(m, "ran_pars", conf.int = TRUE)
 sdmTMB_coefs <- tidy(m, effects = "fixed", conf.int = TRUE)
+sdmTMB_coefs
 ```
+
+\clearpage
 
@@
 
-Comparing the fixed effects between `sdmTMB()` and `inla()` shows the estimates are very similar:
+The estimates for the spatial parameters are very similar between the INLA-based models and `sdmTMB()`:
 
-```{r coef-plot, out.width="5.5in"}
+```{r}
+# INLA
+m_inla$summary.hyperpar[, c("mean", "sd", "0.025quant", "0.975quant")]
+
+# inlabru
+m_bru$summary.hyperpar[, c("mean", "sd", "0.025quant", "0.975quant")]
+
+# sdmTMB
+tidy(m, "ran_pars", conf.int = TRUE)
+```
+
+Note that the standard errors for variance terms are not in natural and not log space and so are a very rough approximation and therefore not reported by the `tidy()` function.
+
+Comparing the fixed effects shows the estimates from all three models are also very similar:
+
+```{r coef-plot, fig.asp=0.4, out.width="85%"}
 inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>% 
@@
 We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year in space. 
+The resulting values are in link space and require the appropriate inverse-link transformation, in this case `plogis()`, to produce occurrence probabilities.
 
@@
 
-```{r maps, cache=FALSE, out.width="6in"}
+```{r maps, cache=FALSE, out.width="5.5in"}
 ggplot(pred) +
@@
 
+We can make similar predictions with inlabru using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:
+
+```{r inlabru-pred, out.width="5.5in"}
+nd_bru <- nd
+nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
+sp::coordinates(nd_bru) <- c("X", "Y")
+nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
+pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
+
+ggplot(as.data.frame(pred_bru)) +
+  geom_raster(aes(X, Y, fill = mean)) +
+  facet_wrap(~year) +
+  scale_fill_viridis_c() +
+  coord_fixed()
+```
+
 Similar predictions are more complicated to extract from INLA, but are easy to plot for the original sample locations.
 
-```{r inla-pred, out.width="6in"}
+```{r inla-pred, out.width="5.5in"}
 idx <- inla.stack.index(sdat, "stdata")$data
 d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
+
 ggplot(d) +
@@
 ```
-
-Or, we can make the prediction with inlabru.
-To do that, we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and use the inverse-link transformation `plogis()`:
-
-```{r inlabru-pred, out.width="6in"}
-nd_bru <- nd
-nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
-sp::coordinates(nd_bru) <- c("X", "Y")
-nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
-pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
-ggplot(as.data.frame(pred_bru)) +
-  geom_raster(aes(X, Y, fill = mean)) +
-  facet_wrap(~year) +
-  scale_fill_viridis_c() +
-  coord_fixed()
-```

### 2022-01-08 doc/appendix-binomial-inla.Rmd
more minor edits

@@
 
-We can make similar predictions with inlabru using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:
+We can get similar predictions with inlabru using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:
 
@@
 
-Similar predictions are more complicated to extract from INLA, but are easy to plot for the original sample locations.
+Equivalent predictions are more complicated to extract from INLA, but are easy to plot for the original sample locations.
 

### 2022-01-13 doc/paper.Rmd
minor edits to intro; rework Bayes options para in model description

@@
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or lack some flexibility.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some flexibility.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
@@
 One common application is for species distribution models (SDMs), hence the package name.
-This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
+This paper describes the basic functionality of this R package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
 
@@
 
-```{r diagram, fig.cap="(ref:diagram-fig)"}
+```{r diagram, fig.cap="(ref:diagram-fig)", out.width="1.03\\textwidth"}
 dir.create("figs", showWarnings = FALSE)
@@
 These data can be collected across discrete points in time.
-Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional (CAR) autoregressive models [e.g., @verhoef2018].
+Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models [e.g., @verhoef2018].
 sdmTMB can also fit models with areal data if each polygon has an associated centroid.
-A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).
+A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).
 
@@
 The GLMMs underpinning sdmTMB models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
+In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in mgcv [@wood2017a]).
 The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
@@
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately.
+Derived parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
+<!-- This sentence felt tacked on at the end.
+Maybe it fits better here?
+-->
 
@@
 
-The sdmTMB model is fit by maximum marginal likelihood.
+While the sdmTMB model is fit by maximum marginal likelihood, Bayesian options are also supported.  
 Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
-
-sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+However, after rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+This option can be fully Bayesian, if penalized likelihoods are included by assigning priors to model parameters (`?sdmTMBpriors`).
+These priors can also be useful in cases where maximum likelihood estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
 Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
@@
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
-Derived parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
+<!-- Would this be an appropriate place to say something about the sim options, possibly specifically for getting uncertainty on spatial predictions?
+I don't know how easy that is from the other packages...
+-->
 

### 2022-01-14 doc/paper.Rmd
edits discussed

@@
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some flexibility.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some functionality.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
@@
 Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
-Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv [@wood2017a], for a specific class of spatial and spatiotemporal models.
+Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv [@wood2017a], for a specific class of spatial and spatiotemporal models.
 One common application is for species distribution models (SDMs), hence the package name.
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-However, after rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
+However, after rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
 This option can be fully Bayesian, if penalized likelihoods are included by assigning priors to model parameters (`?sdmTMBpriors`).

### 2022-01-14 doc/appendix-binomial-inla.Rmd
Edits

@@
 We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year in space.
-The resulting values are in link space and require the appropriate inverse-link transformation, in this case `plogis()`, to produce occurrence probabilities.
+For sdmTMB, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:
 
@@
 ```{r maps, cache=FALSE, out.width="5.5in"}
+inverse_link <- m$family$linkinv
 ggplot(pred) +
-  geom_raster(aes(X, Y, fill = plogis(est))) +
+  geom_raster(aes(X, Y, fill = inverse_link(est))) +
   facet_wrap(~year) +

### 2022-01-14 doc/paper.Rmd
Edits

@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs).
+Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
 However, appropriate models to analyze these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
@@
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately.
-Derived parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
-<!-- This sentence felt tacked on at the end.
-Maybe it fits better here?
--->
-
-<!-- - high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?) -->
-
-<!-- Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs. -->
-<!-- GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.
--->
 
@@
 
-While the sdmTMB model is fit by maximum marginal likelihood, Bayesian options are also supported.  
+The sdmTMB model is fit by maximum marginal likelihood.
 Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-However, after rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
-This option can be fully Bayesian, if penalized likelihoods are included by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors can also be useful in cases where maximum likelihood estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.
+
+sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
 Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
@@
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
-<!-- Would this be an appropriate place to say something about the sim options, possibly specifically for getting uncertainty on spatial predictions?
+<!-- TODO: Would this be an appropriate place to say something about the sim options, possibly specifically for getting uncertainty on spatial predictions?
 I don't know how easy that is from the other packages...
@@
 For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
-Simulation-based residuals [@dharma], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
+Simulation-based residuals [@dharma] (`dharma_residuals()`), one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
 (5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

### 2022-01-14 doc/appendix-pcod-sdmTMB.Rmd
experiment with different example headings; minor edits to owl section and both appendices

@@
 The density units are kg/km^2^.
-Here, X and Y are coordinates in UTM zone 9.
+Here, X and Y are UTM zone 9 coordinates in kilometres.
+If our spatial coordinates were not already in an equidistant projection, we would first transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region.
 

### 2022-01-14 doc/appendix-snow-sdmTMB.Rmd
experiment with different example headings; minor edits to owl section and both appendices

@@
 However, this method does not give us spatially explicit estimates of uncertainty.
-To do this, we can use the output from the joint precision matrix simulation draws by using the `predict()` function again but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`).
-We also extract the same number of drams for (this time `n_sims = 500`) in the `spread_sims()` function for the main effect of NAO. 
-The simulation draws for both should then be combined before adding the median and lower and upper quantiles from these draws to our prediction data frame.
+To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`).
+We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `n_sims = 500`). 
+The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws and adding these values to our prediction data frame.
 

### 2022-01-14 doc/paper.Rmd
experiment with different example headings; minor edits to owl section and both appendices

@@
 
-# Introductory example with fish survey biomass
+<!-- # Introductory example with fish survey biomass -->
+# Introductory example using fish survey data
 
@@
 for how to do this should be moved to the appendix?
-I do explain the transformation for the owl data, but it could be added to the pcod appendix too?
+I do explain the transformation for the owl data and added a sentence to the pcod appendix, but maybe still important to include here as well?
 -->
@@
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-The circle area corresponds to density of Pacific cod for individual trawls.
+The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
 (B) Spatial random field:
@@
 
-# Spatially varying coefficient example with citizen science data
+<!-- # Spatially varying coefficient example with citizen science data -->
+# 2. Estimate a spatially varying effect from citizen science data
 
 Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
-The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
+The reasons for this interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
 Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3).
+We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
 
@@
 
-We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao)).
+We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix 3).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

### 2022-01-14 doc/paper.Rmd
Tweak headers; remove marked

@@
 
-<!-- # Introductory example with fish survey biomass -->
 # Introductory example using fish survey data
@@
 
-<!-- # Spatially varying coefficient example with citizen science data -->
-# 2. Estimate a spatially varying effect from citizen science data
+# Example of a spatially varying coefficient with citizen science data
+<!-- # Estimate a spatially varying effect from citizen science data -->
 
@@
 
-We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix 3).
+We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix 3).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

### 2022-01-14 doc/make.R
Add make.R

@@ -1,0 +1,36 @@
+files <- c(
+  "paper",
+  "appendix-model",
+  "appendix-pcod-sdmTMB",
+  "appendix-binomial-inla",
+  "appendix-pcod-VAST-tweedie",
+  "appendix-validation"
+)
+
+# https://github.com/rstudio/rmarkdown/issues/1673
+render_separately <- function(...) callr::r(
+  function(...) rmarkdown::render(..., envir = globalenv()), args = list(...), show = TRUE)
+
+for (i in seq_along(files)) {
+  render_separately(paste0(here::here("doc", files[i]), ".Rmd"))
+}
+
+names_to <- c(
+  "sdmTMB-paper.pdf",
+  "appendix1-model.pdf",
+  "appendix2-index-sdmTMB.pdf",
+  "appendix3-INLA-comparison.pdf",
+  "appendix4-VAST-comparison.pdf",
+  "appendix5-speed-validation.pdf"
+)
+
+loc <- "~/Dropbox/sdmTMB-paper/"
+
+for (i in seq_along(files)) {
+  from <- paste0(here::here("doc", files[i]), ".pdf")
+  to <- paste0(loc, names_to[i])
+  cat(to, "\n")
+  file.copy(from, to, overwrite = TRUE)
+}
+
+system("cd doc; ./clean-bib.sh")

### 2022-01-14 doc/paper.Rmd
Copy edit

@@
 sdmTMB can also fit models with areal data if each polygon has an associated centroid.
-A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).
+A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be estimated.
+<!-- (e.g., spatial variance and range---the distance at which points are effectively independent). -->
 
-The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig.
+The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts, time-varying effects, and spatially varying effects (Fig.
 \@ref(fig:diagram), Appendix 1).
@@
 
-By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix 1).
+By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix 1).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or can be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix 1).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
 
@@
 
-We can then fit a model with spatial random fields via the function `sdmTMB()`.
+We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
 We use a penalized smoother for depth as a main effect via `s()` from the mgcv package and choose a basis dimension of 5.
@@
 
-We could generate a population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
+We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
 \@ref(fig:pcod-fig)E, Appendix 2).
@@
 mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
-With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
+With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
 Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown).
@@
 We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
-Users familiar with INLA may find inlabru approachable.
-Users already familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
+Users familiar with INLA will find inlabru approachable.
+Users familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
 VAST is the sole option for fitting some multivariate models;

### 2022-01-14 doc/appendix-pcod-sdmTMB.Rmd
Small appendix stuff

@@
     number_sections: true
+bibliography: refs.bib
+csl: mee.csl
 ---
@@
 
-For example, a mesh with a cutoff distance of 10 km can be made with
+For example, a mesh with a cutoff distance of 10 km can be made:
 
@@
 
-We can inspect randomized quantile residuals:
+We can inspect simulated residuals from the DHARMa package:
 
-```{r qqnorm, out.width="5in"}
-pred$resid <- residuals(fit)
-qqnorm(pred$resid)
-qqline(pred$resid)
-```
-
-```{r plot-resids}
-ggplot(pred, aes(X, Y, colour = resid)) +
-  geom_point() +
-  facet_wrap(~year) +
-  coord_fixed() +
-  scale_colour_gradient2()
+```{r sim-resids, out.width="3.5in", fig.align='center', warning=FALSE, results='hide'}
+simulate(fit, nsim = 500) %>%
+  dharma_residuals(fit)
 ```
@@
 The above example used the default `bias_correct = FALSE` in `get_index()`.
-The `bias_correct` argument is a feature of TMB for models with random effects, and more detils can be found in the TMB documentation [https://github.com/kaskr/adcomp/wiki](https://github.com/kaskr/adcomp/wiki) or [Thorson & Kristensen (2016)](https://www.sciencedirect.com/science/article/abs/pii/S0165783615301399?via%3Dihub).
+The `bias_correct` argument is a feature of TMB for models with random effects, and more detils can be found in the  [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias].
 Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
-Setting `bias_correct = TRUE` implements the generic by his correction algorithm that should correct for this, but can take considerable time and memory.
+Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but can take considerable time and memory.
 
@@
 This should approximate the output of `get_index(x, bias_correct = TRUE)`.
-However, we note that there may be situations where this approach may fail or be biased.
+However, we note that there may be situations where this approach may fail or be biased and experimentation is ongoing.
 
@@
 ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)
-saveRDS(ind, "sdmTMB-index-sim.rds")
 ggplot(ind, aes(year, est / 1000)) +
@@
 ```
+
+# References

### 2022-01-14 doc/appendix-snow-sdmTMB.Rmd
Small appendix stuff

@@
 bibliography: refs.bib
+csl: mee.csl
 ---

### 2022-01-14 doc/appendix-snow-sdmTMB.Rmd
add text on owl results to appendix

@@
 theme_set(theme_bw() + theme(
+  plot.title = element_text(size = 12),
   legend.position = "top", legend.justification = c(0, 0), # move legend to top left
@@
 lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers) # only large lakes
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)# only large lakes
 ```
@@
 We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `n_sims = 500`). 
-The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws and adding these values to our prediction data frame.
+The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws (in this case we will use 80%) and adding these values to our prediction data frame.
 
 ```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE}
-b <- tidy(m)
+b <- tidy(m, conf.int = TRUE)
 
@@
 
-```{r plot-map, echo=TRUE, eval=TRUE, out.width="100%"}
+```{r}
+nao_effect_lwr95 <- apply(combined, 2, quantile, probs = 0.025)
+```
+
+The main effect of NAO indicates that there is a weak average positive effect of annual mean NAO on overall counts `r round(b[b$term == "nao", 2], 2)` (`b[b$term == "nao", 2]`), while for individual count locations the median estimated effects of NAO range from `r round(min(p_mean$nao_sim), 2)` to `r round(max(p_mean$nao_sim),2)` (`range(p_mean$nao_sim)`).
+Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`).
+All these effects are in link space (`link = "log"`) and could be `exp()` to give the expected multiplicative contribution to the predicted owl counts, however the additive nature of the effect in link space is actually somewhat more intuitive to interpret visually, so this is what we will plot on our maps:
+
+```{r plot-map, echo=TRUE, eval=TRUE, out.width="95%", cache=FALSE}
 ggplot(data = p_proj) +
   geom_point(
-    data = p_mean,
-    aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
+    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
   ) +
@@
   labs(colour = "Estimated effect in log space") +
-  ggtitle("Map of combined main and spatially varying effect of NAO on Snowy Owl count") +
+  ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
   theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
@@
 
-```{r plot-map-ci, echo=F, eval=TRUE, out.width="100%"}
+```{r plot-map-ci, echo=F, eval=TRUE, out.width="100%", cache=FALSE}
 p1 <- ggplot(data = p_proj) +
@@
 
+If we recalculated these at 95% CI the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that for the most conventional level of statistical confidence, we cannot conclude that the effect of NAO at an annual scale differs from 0.
+However, the southeast to northwest gradient in the intensity illustrated in the maps above is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
+This result suggests that exploring the effects of NAO at finer temporal scales (possibly seasonally) and/or with lagged effects could prove fruitful.
+
 <!-- \clearpage -->
@@
 The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates.
-```{r plot-all-effects, echo=TRUE, eval=TRUE}
+
+```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE}
 ggplot(data = p_proj) +
-  geom_point(
-    data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5
-  ) +
+  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
   coord_sf(

### 2022-01-14 doc/appendix-pcod-sdmTMB.Rmd
fix typo

@@
 The above example used the default `bias_correct = FALSE` in `get_index()`.
-The `bias_correct` argument is a feature of TMB for models with random effects, and more detils can be found in the  [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias].
+The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in the [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias].
 Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.

### 2022-01-15 doc/appendix-pcod-sdmTMB.Rmd
remove a duplicate 'first'

@@
 Here, X and Y are UTM zone 9 coordinates in kilometres.
-If our spatial coordinates were not already in an equidistant projection, we would first transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region.
+If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region.
 

### 2022-01-18 doc/appendix-pcod-sdmTMB.Rmd
minor tweaks to appendix figure sizes to make page layouts nicer

@@
 
-```{r plot-mesh, fig.asp=0.9, out.width="4in"}
+```{r plot-mesh, cache=FALSE, fig.align='center', fig.asp=1, out.width="3in"}
 plot(mesh)
@@
 
-```{r sim-resids, out.width="3.5in", fig.align='center', warning=FALSE, results='hide'}
+```{r sim-resids, out.width="4in", warning=FALSE, results='hide'}
 simulate(fit, nsim = 500) %>%
@@
 
-```{r index, out.width="5in"}
+```{r index, fig.asp=0.45, out.width="5in"}
 qcs_grid$area <- 4 # all 2 x 2km
@@
 saveRDS(ind, "sdmTMB-index.rds")
+
 ggplot(ind, aes(year, est / 1000)) +
@@
 
-```{r index-sims, out.width="5in"}
+```{r index-sims, fig.asp=0.45, out.width="5in"}
 ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)
+
 ggplot(ind, aes(year, est / 1000)) +

### 2022-01-18 doc/appendix-validation.Rmd
minor tweaks to appendix figure sizes to make page layouts nicer

@@
   comment = "#>",
+  fig.width = 5,
+  fig.asp = 0.7, 
   echo = TRUE,
@@
 
-```{r pkgs, warning=FALSE, message=FALSE}
+```{r pkgs, warning=FALSE, message=FALSE, cache=FALSE}
 library(INLA)
@@
 
-```{r sim-dat-plot, out.width="4.5in"}
+```{r sim-dat-plot, out.width="5in", cache=FALSE}
 ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
@@
   scale_color_gradient2() +
+  coord_fixed() +
   theme_light()
@@
 
-```{r plot-smooth, echo=TRUE, fig.asp=1}
-par(mfrow = c(2, 1))
+```{r plot-smooth, echo=TRUE, fig.asp=0.6, out.width="5in"}
 plot_smooth(m)
@@
 
-```{r iid-compare, echo=TRUE}
+```{r iid-compare, echo=TRUE, fig.asp=0.7, out.width="6in"}
 sdmTMB_pars <- tidy(m, "ran_pars")

### 2022-01-26 doc/paper.Rmd
Start editing paper.Rmd based on comments

@@
 Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
-Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
+Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
 These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
@@
 
+<!-- TODO: rotate and fill page: -->
+
 ```{r diagram, fig.cap="(ref:diagram-fig)", out.width="1.03\\textwidth"}
@@
 A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be estimated.
+<!-- TODO: Line 83, 84. The parameters of CAR, SAR, ICAR, etc model can be estimated, although they can’t be readily interpreted. Is this what you were referring to? -->
 <!-- (e.g., spatial variance and range---the distance at which points are effectively independent). -->
@@
 For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
-<!-- TODO: Would this be an appropriate place to say something about the sim options, possibly specifically for getting uncertainty on spatial predictions?
-I don't know how easy that is from the other packages...
--->
 
@@
 
-knitr::kable(d, format = "latex", caption = "Comparison of functionality between seven R packages that can fit geostatistical GLMMs.",
+knitr::kable(d, format = "latex", caption = "Comparison of functionality between several R packages that can fit geostatistical GLMMs.",
   booktabs = TRUE,
@@
 There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
-sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
 sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
@@
 inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
-mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
+mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
 <!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
@@
 Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown).
+<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
 Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.

### 2022-01-28 doc/appendix-binomial-inla.Rmd
Edits dealing with Jim and Joe's comments

@@
   cache = TRUE,
-  autodep = TRUE
+  autodep = TRUE,
+  warning = FALSE,
+  message = FALSE
 )
@@
 
-Note that the standard errors for variance terms are not in natural and not log space and so are a very rough approximation and therefore not reported by the `tidy()` function.
+Note that, within sdmTMB, the standard errors for variance terms are not reported in the `tidy()` function as they could be misleading if considered in natural (rather than log) space.
+The confidence intervals are, however, reported.
 

### 2022-01-28 doc/appendix-model.Rmd
Edits dealing with Jim and Joe's comments

@@
 ### Smoothers
-<!-- EW: see this paragraph and below, should package names be enclosed in single quotes, like `sdmTMB()` above? Examples: mgcv, brms, sdmTMB throughout -->
-<!-- EW: unrelated, I never tried to implement factor smooths. I assume they work, but it'd good to check eventually -->
+
 Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
@@
 and $\kappa$ controls the spatial correlation range (decorrelation rate).
-The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
+The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GRF to greatly increase computational efficiency [@lindgren2011].
 Internally, the parameter $\kappa$ and $\tau$ are converted to range and marginal standard deviation $\sigma$ as $\textrm{range} = \sqrt{8} / \kappa$ and $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
@@
 
-## Barrier mesh
+## Incorporating physical barriers into the SPDE
 
 In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes).
-SPDE models allow for barriers to be included via barrier meshes [@bakka2019].
+SPDE models allow for physical barriers to be incorporated into the modelling [@bakka2019].
 With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
@@
 
-The barrier mesh implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019].
+The barrier implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019].
 For example, if the range was estimated at 10 km, `range_fraction = 0.2` would assume that the range was 2 km across the barrier.

### 2022-01-28 doc/appendix-pcod-VAST-tweedie.Rmd
avoid conflict between versions of make_mesh in VAST appendix
Edits dealing with Jim and Joe's comments

@@
 Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
-Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `"input_grid" = input_grid` in `fit_model()`.
+Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `input_grid = input_grid` in `fit_model()`.
 
@@
 ```{r sdmTMB-example, echo = TRUE, cache=TRUE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
+mesh <- sdmTMB::make_mesh(pcod, xy_cols = c("X", "Y"), 
                           mesh = fit$spatial_list$MeshList$isotropic_mesh)

### 2022-01-28 doc/appendix-validation.Rmd
Edits dealing with Jim and Joe's comments

@@
 dat <- mgcv::gamSim(3, n = 800)
-m_mgcv <- mgcv::gam(y ~ s(x2), data = dat)
+m_mgcv <- mgcv::gam(y ~ s(x2), data = dat, method = "REML")
 p_mgcv <- predict(m_mgcv)
@@
 We can see the predictions match nearly exactly.
-Furthermore, we can use the in-development function `sdmTMB::plot_smooth()` to compared the smoothers themselves.
-Note that the sdmTMB version is showing the marginal effect of a prediction including the intercept whereas the mgcv version is showing only the smooth component.
-Nonetheless, here the smoothers look very similar except for these scale on the y axis.
+Furthermore, we can use the in-development function `sdmTMB::plot_smooth()` to compare the smoothers themselves.
+Note that the sdmTMB version shows the marginal effect of a prediction including the intercept whereas the `mgcv::plot.gam()` shows the smooth component only.
+We therefore use `mgcv::predict.gam()` to make a matching plot.
+Below, we add a tiny jitter amount (`0.02`) to the mgcv output just so the two smoothers can both be visualized.
+The blue lines are from mgcv and the black lines are from sdmTMB.
 
@@
 plot_smooth(m)
-plot(m_mgcv)
+
+x2 <- seq(0, 1, length.out = 100)
+p_mgcv <- predict(m_mgcv, newdata = data.frame(x2 = x2), se.fit = TRUE)
+qv <- stats::qnorm(1 - (1 - 0.95) / 2)
+df <- data.frame(
+  x2 = x2,
+  est = p_mgcv$fit, 
+  lwr = p_mgcv$fit - qv * p_mgcv$se.fit, 
+  upr = p_mgcv$fit + qv * p_mgcv$se.fit
+)
+lines(df$x2, df$est + 0.02, col = "blue")
+lines(df$x2, df$lwr + 0.02, lty = 2, col = "blue")
+lines(df$x2, df$upr + 0.02, lty = 2, col = "blue")
 ```
@@
 ```{r bivariate-compare, echo=TRUE}
-m_mgcv <- mgcv::gam(y ~ s(x2, x1), data = dat)
+m_mgcv <- mgcv::gam(y ~ s(x2, x1), data = dat, method = "REML")
 p_mgcv <- predict(m_mgcv)

### 2022-01-28 doc/paper.Rmd
Edits dealing with Jim and Joe's comments

@@
   \renewcommand{\thefigure}{S\arabic{figure}}}
+  \usepackage{lscape}
   \widowpenalty10000
@@
 
-```{r diagram, fig.cap="(ref:diagram-fig)", out.width="1.03\\textwidth"}
+\begin{landscape}
+
+```{r diagram, fig.cap="(ref:diagram-fig)", out.width="9in"}
 dir.create("figs", showWarnings = FALSE)
@@
 ```
+
+\end{landscape}
 
@@
 sdmTMB can also fit models with areal data if each polygon has an associated centroid.
-A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be estimated.
-<!-- TODO: Line 83, 84. The parameters of CAR, SAR, ICAR, etc model can be estimated, although they can’t be readily interpreted. Is this what you were referring to? -->
+A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted [@wall2004].
 <!-- (e.g., spatial variance and range---the distance at which points are effectively independent). -->
 
-The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts, time-varying effects, and spatially varying effects (Fig.
+The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.
 \@ref(fig:diagram), Appendix 1).
@@
 The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
-The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately.
+The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared (`share_range = TRUE`) or estimated separately (`share_range = FALSE`).
 
@@
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
 
@@
 The results from an sdmTMB model can be used to generate various quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity  [`get_cog()`; e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.
 
@@
 We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
-We use a penalized smoother for depth as a main effect via `s()` from the mgcv package and choose a basis dimension of 5.
+We use a penalized smoother for depth as a main effect via `s()` from the mgcv package.
+The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `+ s(x, k = 4)`.
 We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
+<!-- We could instead model catch weight as the response with an offset for log(area). -->
 
@@
 fit <- sdmTMB(
-  density ~ s(depth, k = 5),
+  density ~ s(depth),
   data = pcod,
@@
 ```{}
-#> Spatiotemporal model fit by ML ['sdmTMB']
-#> Formula: density ~ s(depth, k = 5)
+#> Spatial model fit by ML ['sdmTMB']
+#> Formula: density ~ s(depth)
 #> Mesh: mesh
@@
 #>             coef.est coef.se
-#> (Intercept)     2.34    0.23
-#> sdepth        -26.96   21.52
+#> (Intercept)     2.37    0.21
+#> sdepth          6.17   25.17
 #> 
@@
 #>            Std. Dev.
-#> sds(depth)     34.03
+#> sds(depth)     13.93
 #> 
-#> Dispersion parameter: 12.70
+#> Dispersion parameter: 12.69
 #> Tweedie p: 1.58
-#> Matern range: 17.47
+#> Matern range: 16.39
 #> Spatial SD: 1.86
-#> ML criterion at convergence: 6403.637
+#> ML criterion at convergence: 6402.136
 ```
@@
 fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth, k = 5), family = tweedie(link = "log"), data = pcod, mesh = mesh,
-  time = "year", spatial = "off", spatiotemporal = "ar1"
+  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
+  time = "year", spatial = "on", spatiotemporal = "ar1"
 )
@@
 (C) Overall prediction: these estimates represent the combination of all fixed and random effects.
+Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
 (D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
-(E) Standardized area-weighted population index derived from a spatiotemporal version of this model. The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
+(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
 <!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->
@@
 ```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", out.width="5.2in", fig.align='center'}
-pcodfig <- here::here("figs", "pcod-fig2.pdf")
+pcodfig <- here::here("figs", "pcod-fig.pdf")
 knitr::include_graphics(pcodfig)
@@
 d <- as.matrix(d)
-d[d == "1"] <- "\\checkmark"
-d[d == "1*"] <- "\\checkmark\\footnote{a}"
-d[d == "GAMs*"] <- "GAMs\\footnote{a}"
-d[d == "0"] <- "--"
-d[d == "0*"] <- "--\\footnote{a}"
+# d[d == "1"] <- "\\checkmark"
+d <- gsub("^1", "\\\\checkmark", d)
+# d[d == "1*"] <- "\\checkmark\\footnote{a}"
+# d[d == "GAMs*"] <- "GAMs\\footnote{a}"
+# d[d == "0"] <- "--"
+d <- gsub("^0", "--", d)
+# d[d == "0*"] <- "--\\footnote{a}"
 d <- as.data.frame(d)
@@
   kableExtra::footnote(
-    number = c(
-      "Technically possible but non-trivial",
-      "Penalized smoother GAMs",
-      "inlabru but not INLA",
-      "Spatiotemporal fields as random walk only",
-      "SPDE approach as in Miller et al. (2019)",
-      "Does have predictive process knots",
-      "Zero-inflated NB2 only",
-      "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}",
-      "Hurdle models possible by fitting components separately",
-      "Hurdle models possible by fitting components separately",
-      "Technically possible but non-trivial"
+    general = paste(
+      "$^1$Technically possible but non-trivial.",
+      "$^2$Penalized smoother GAMs that determine `wiggliness'.",
+      "$^3$inlabru but not INLA.",
+      "$^4$Spatiotemporal fields as random walk only.",
+      "$^5$SPDE approach as in Miller et al. (2019).",
+      "$^6$Does have predictive process knots.",
+      "$^7$Zero-inflated NB2 only.",
+      "$^8$Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
+      "$^9$Possible as log-linked Poisson GLMM with aggregated data.",
+      "$^{10}$Hurdle models possible by fitting components separately."
     ),
-    escape = FALSE
+    escape = FALSE, threeparttable = TRUE
   )
 ```
+
+<!-- TODO: Line 242.
+Can sdmTMB be run in parallel?
+If so, does the parallelism share memory across the cores or does the memory grow with the number of cores?
+This is a HUGE feature of PARDISO within INLA/inlabru allowing for very large models to be fit on HPCs.
+It might be worth mentioning and adding to Table 1.
+Also Line 242: do you expect the same relative performance between the software packages as more cores are used?
+-->
 
@@
 
-Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, spatial covariance barriers, time-varying spatiotemporal variance, and simulation from the parameter joint precision matrix.
+Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
 Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-gaussian random fields [e.g., @anderson2019].
+The included TMB `.cpp` file also provides a high-quality, bug-free, and tested model template that can be modified to add additional features.
 <!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
@@
 We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend. 
-Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data.
+<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
+<!-- Line 275: I would suggest adding a MAJOR benefit of sdmTMB for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates!
+This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it.
+It saves me a bunch of time.
+If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it.
+-->
 
@@
 We thank the authors of all these packages.
-We thank P.M. Regular, J. Watson, ..., TODO.
+We thank M. Lindmark, P.M. Regular, J.T. Thorson, J. Watson, ..., TODO.
 
@@
 A development version of sdmTMB is available at <https://github.com/pbs-assess/sdmTMB>.
-sdmTMB will be available on CRAN.
-Paper source code will be available on Zenodo and at <https://github.com/seananderson/sdmTMB-paper> on publication.
+sdmTMB will be available on CRAN before publication.
+Paper source code is available at <https://github.com/seananderson/sdmTMB-paper> and will be archived on Zenodo before publication.
 

### 2022-01-28 analysis/pcod-fig.R

@@ -27,16 +27,6 @@ fit <- sdmTMB(
 print(fit)
 p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
 
-# p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
-# se <- apply(p3, 1, sd)
-# p$se <- se
-# g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
-#   geom_raster() +
-#   scale_fill_viridis_c(trans = "log10", option = "C") +
-#   coord_fixed() +
-#   labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
-#   theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
-
 nd <- data.frame(
   depth =
     seq(min(pcod$depth), max(pcod$depth), length.out = 300)
@@ -46,13 +36,14 @@ p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
 fit_spatiotemporal <- sdmTMB(
   density ~ s(depth),
   family = tweedie(link = "log"), data = pcod, mesh = mesh,
-  time = "year", spatial = "off", spatiotemporal = "ar1", silent = FALSE
+  time = "year", spatial = "on", spatiotemporal = "ar1", silent = FALSE
 )
 p_tv <- predict(fit_spatiotemporal,
   newdata = qcs_grid, return_tmb_object = TRUE,
   area = 4
 )
 ind <- get_index(p_tv)
+cog <- get_cog(p_tv, format = "wide")
 
 # just for plotting in UTMs
 pcod$X1000 <- pcod$X * 1000
@@ -66,7 +57,6 @@ lims_y <- c(5366427 + 270000, 6353456 - 513000) + c(-10000, 10000)
 land <- "grey86"
 land_border <- "grey86"
 
-# devtools::install_github("ropensci/rnaturalearthhires") # install if needed
 map_data <- rnaturalearth::ne_countries(
   scale = "large",
   returnclass = "sf", country = "canada"
@@ -84,30 +74,48 @@ bc_coast_proj <- sf::st_transform(bc_coast, crs = utm_zone9)
 
 .lab <- expression("Density" ~ (kg / km^2))
 
-# .lab <- expression(atop("Predicted density", (km/m^2)))
-
+# base plot:
 g_pred <- ggplot(bc_coast_proj) +
[…]

### 2022-01-28 doc/rename-stuff.R

@@ -1,0 +1,26 @@
+files <- c(
+  "paper.pdf",
+  "appendix-model.pdf",
+  "appendix-pcod-sdmTMB.pdf",
+  "appendix-binomial-inla.pdf",
+  "appendix-pcod-VAST-tweedie.pdf",
+  "appendix-validation.pdf"
+)
+
+names <- c(
+  "sdmTMB-paper.pdf",
+  "appendix1-model.pdf",
+  "appendix2-index-sdmTMB.pdf",
+  "appendix3-INLA-comparison.pdf",
+  "appendix4-VAST-comparison.pdf",
+  "appendix5-speed-validation.pdf"
+)
+
+loc <- "~/Dropbox/sdmTMB-paper/"
+
+for (i in seq_along(files)) {
+  from <- here::here("doc", files[i])
+  to <- paste0(loc, names[i])
+  cat(to, "\n")
+  file.copy(from, to, overwrite = TRUE)
+}

### 2022-01-28 doc/appendix-pcod-VAST-tweedie.Rmd
Copy edit/ slightly jitter VAST

@@
 
-We will fit a geostatistical spatiotemporal model with VAST for the purposes of index standardization, following the sdmTMB example.
+We will fit a geostatistical spatiotemporal model with VAST for the purposes of index standardization.
 We will use a data set built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
@@
 To specify the mesh used to approximate the spatial process, which is used in the SPDE calculations, we use the k-means method in VAST.
-Rather than specifying the cutoff distance, meshes in VAST are typically generated by specifying only the number of knots, which we will later pass, along with other model settings to the function make_settings.
-We will use 205 knots, the same number in the mesh created in the other versions of this appendix.
+Rather than specifying the cutoff distance, meshes in VAST are typically generated by specifying only the number of knots, which we will later pass, along with other model settings to the function `make_settings()`.
+We will use 205 knots, the same number in the mesh created in the other appendices.
 
@@
 Settings used for index standardization are specified partially by defining `purpose = "index2"` but we also explicitly provide arguments for these and other key settings here.
-Specifically, we set the elements of `RhoConfig` to 0 to specify that we want the intercepts and spatiotemporal fields to be independent (which is the default).
+Specifically, we set the elements of `RhoConfig` to `0` to specify that we want the intercepts and spatiotemporal fields to be independent, which is the default.
 
@@
 In this case, we want to fit a model with a single component, using the Tweedie distribution.
-To do so, we must turn off estimation of omega and epsilon for 1st linear predictor by entering 0 for those elements in the object we'll use to define the configuration of these fields `FieldConfig`.
+To do so, we must turn off estimation of omega and epsilon for first linear predictor by entering `0` for those elements in the object we will use to define the configuration of these fields in `FieldConfig`.
 
@@
 
-Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
+Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function `fit_model()` and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).
 Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `input_grid = input_grid` in `fit_model()`.
@@
 We can compare the various parameter estimates.
-Note that VAST adds the dispersion parameter (`phi` in sdmTMB) in the mean of the Tweedie distribution, so we need to account for that when comparing main-effect estimates.
+Note that VAST adds the dispersion parameter (`phi` in sdmTMB) into the mean of the Tweedie distribution, so we need to account for that when comparing main-effect estimates.
 
@@
 
-Now, we can compare the indices.
+Now, we can compare the indices. In the ribbon plot below, the lines perfectly covered each other and so we added a small jitter amount to make them both visible.
 
@@
 
+# jitter slightly for visualization:
+jitter_amount <- mean(vast_i$est) * 0.01
+vast_i_gg <- vast_i
+vast_i_gg$est <- vast_i_gg$est + jitter_amount
+vast_i_gg$lwr <- vast_i_gg$lwr + jitter_amount
+vast_i_gg$upr <- vast_i_gg$upr + jitter_amount
+
 sdm_i <- ind_sdmTMB %>% mutate(index = "sdmTMB")
-both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
+both_i <- bind_rows(sdm_i, vast_i_gg) %>% filter(est > 0)
 ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
-  geom_ribbon(alpha = 0.3) +
+  geom_ribbon(fill = NA, lty = 2) +
   geom_line() + 
@@
   coord_cartesian(expand = FALSE) +
-  ylab("Biomass (t)")
+  ylab("Biomass (t)") +
+  theme_light()
 
@@
 
-The results are effectively identical. In the ribbon plot above, the lines perfectly cover each other.
+The results are effectively identical.
 

### 2022-01-28 doc/appendix-validation.Rmd
Change mgcv jitter

@@
 We therefore use `mgcv::predict.gam()` to make a matching plot.
-Below, we add a tiny jitter amount (`0.02`) to the mgcv output just so the two smoothers can both be visualized.
+Below, we add a tiny jitter amount (`0.04`) to the mgcv output just so the two smoothers can both be visualized.
 The blue lines are from mgcv and the black lines are from sdmTMB.
 
-```{r plot-smooth, echo=TRUE, fig.asp=0.6, out.width="5in"}
+```{r plot-smooth, echo=TRUE, fig.asp=0.8, out.width="5in"}
+par(cex = 0.8)
+# sdmTMB:
 plot_smooth(m)
 
+# mgcv:
 x2 <- seq(0, 1, length.out = 100)
@@
 )
-lines(df$x2, df$est + 0.02, col = "blue")
-lines(df$x2, df$lwr + 0.02, lty = 2, col = "blue")
-lines(df$x2, df$upr + 0.02, lty = 2, col = "blue")
+lines(df$x2, df$est + 0.04, col = "blue")
+lines(df$x2, df$lwr + 0.04, lty = 2, col = "blue")
+lines(df$x2, df$upr + 0.04, lty = 2, col = "blue")
 ```

### 2022-01-28 doc/paper.Rmd
Copy edit end of paper
Edit/ move figs/tables around

@@
 
-(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
-An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
-The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
-Notation:
-We refer to design matrices as $\bm{X}$.
-The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
-The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
-All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See Appendix 1 for a full description of the model.
-Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
-
-<!-- TODO: rotate and fill page: -->
-
-\begin{landscape}
-
-```{r diagram, fig.cap="(ref:diagram-fig)", out.width="9in"}
-dir.create("figs", showWarnings = FALSE)
-f <- here::here("figs", "diagram.pdf")
-knitr::include_graphics(f)
-```
-
-\end{landscape}
-
 # Model description
@@
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared (`share_range = TRUE`) or estimated separately (`share_range = FALSE`).
+
+(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
+An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
+The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
+Notation:
+We refer to design matrices as $\bm{X}$.
+The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
+The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
+All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
+See Appendix 1 for a full description of the model.
+Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
+
+\begin{landscape}
+
+```{r diagram, fig.cap="(ref:diagram-fig)", out.width="9in"}
+dir.create("figs", showWarnings = FALSE)
+f <- here::here("figs", "diagram.pdf")
+knitr::include_graphics(f)
+```
+
+\end{landscape}
 
@@
 <!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
-We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method.
+We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
 <!-- (omitted for space). -->
@@
 ```{r,eval=FALSE, echo=TRUE}
-print(fit)
+summary(fit)
 ```
@@
 
-We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured in time.
-Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).
+We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+Here we will keep the spatial field and structure the spatiotemporal random fields as AR(1).
 
@@
 
+# Package comparisons
+
+There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
+sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
+sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
+To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
+<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
+Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
+spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
+There is considerable variability in the available observation likelihoods across packages (Table 1).
+We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix 4) and VAST (Appendix 5).
+
 ```{r compare-table}
@@
 
-# Package comparisons
-
-There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
-sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
-sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
-sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
-To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
-<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
-Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
-spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
-There is considerable variability in the available observation likelihoods across packages (Table 1).
-We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix 4) and VAST (Appendix 5).
+We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+\@ref(fig:timing), Appendix 6).
+In this test, for mesh resolutions under 500 nodes, sdmTMB was faster than inlabru or mgcv.
+inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
+mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
+With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
+<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
+<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
+Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
+Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
+<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
+<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->
 
@@
 sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
-VAST should be similar to sdmTMB and so is not shown.
+VAST is similar to sdmTMB and so is not shown.
 inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
@@
 ```
-
-We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
-\@ref(fig:timing), Appendix 6).
-In this test, for mesh resolutions under 500 nodes, sdmTMB was faster than inlabru or mgcv.
-inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
-mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
-With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
-Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown).
-<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
-Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
-<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
-<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->
 
@@
 We thank the authors of all these packages.
-We thank M. Lindmark, P.M. Regular, J.T. Thorson, J. Watson, ..., TODO.
+We thank M. Lindmark, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
 
@@
 
-SA led development of the software and EW contributed key software code. All
-authors contributed software code, guided software development, and
-participated in testing. PE led the Snowy Owl case study. All authors
-contributed to appendix code. SA led the writing of the manuscript. All authors
-contributed critically to the drafts and gave final approval for publication.
+SA led development of the software and EW contributed key software code.
+All authors contributed software code, guided software development, and
+participated in testing.
+PE led the Snowy Owl case study.
+All authors contributed to appendix code.
+SA led the writing of the manuscript; all authors contributed critically to the drafts and gave final approval for publication.
 

### 2022-01-28 doc/paper.Rmd
Refs, small edits

@@
 Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but can be extended to penalized likelihood or Bayesian inference.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB) but can be extended to penalized likelihood or Bayesian inference.
 We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
@@
     | Keywords:
-      Gaussian Markov random fields,
+      Gaussian Markov random fields (GMRF),
       generalized linear mixed effects models (GLMM),
@@
 These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
-This can greatly improve computational efficiency [@heaton2017].
+This can greatly improve computational efficiency [e.g., @heaton2019].
 The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
@@
 The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
-The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared (`share_range = TRUE`) or estimated separately (`share_range = FALSE`).
+The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).
 
@@
 We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
+The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)`.
 
@@
 We use a penalized smoother for depth as a main effect via `s()` from the mgcv package.
-The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `+ s(x, k = 4)`.
 We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
@@
 
-There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
+There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
 sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
@@
 <!--
-There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
+There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2019]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
 VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.

### 2022-02-02 doc/appendix-snow-sdmTMB.Rmd
Reduce file size, fix white-space in owls

@@
   fig.width = 7,
-  fig.asp = 1,
+  fig.asp = 0.7,
   cache = TRUE,
   autodep = TRUE,
-  cache.comments = FALSE
+  cache.comments = FALSE,
+  dev = "png",
+  dpi = 150,
+  optipng = "-strip all"
 )
+knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 ```
@@
 
-```{r mesh, echo=TRUE, cache=FALSE, out.width="70%"}
+```{r mesh, echo=TRUE, cache=FALSE, out.width="4in"}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
-plot(mesh)
+
+# built-in plot() method:
+# plot(mesh)
+
+# or with ggplot:
+ggplot() + inlabru::gg(mesh$mesh) +
+  geom_point(aes(X, Y), data = snow) +
+  theme_light()
 ```
@@
 Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America.
-Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
+Therefore, the effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
 Because there does not appear to be much correlation between years, year will be included as a random factor: `(1 | year_f)`. 
@@
 
-We can inspect randomized quantile residuals, although this method has known issues.
+We can inspect randomized quantile residuals, although this method has known issues [@thygesen2017a].
 
@@
 
-```{r plot-resids, fig.width=8}
+```{r plot-resids, fig.width=8, dpi=200}
 ggplot(pred, aes(X, Y, colour = resid)) +
@@
 
-```{r plot-map, echo=TRUE, eval=TRUE, out.width="95%", cache=FALSE}
+```{r plot-map, echo=TRUE, eval=TRUE, out.width="80%", cache=FALSE}
 ggplot(data = p_proj) +
@@
 
-```{r plot-map-ci, echo=F, eval=TRUE, out.width="100%", cache=FALSE}
+```{r plot-map-ci, echo=F, eval=TRUE, out.width="85%", cache=FALSE}
 p1 <- ggplot(data = p_proj) +
@@
 
-```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE, dpi=220}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=TRUE}
+```{r plot-spatial-effects, echo=TRUE, eval=TRUE, out.width="5in", dpi=200}
 ggplot(data = p_proj) +
@@
   ) +
-  scale_colour_viridis_c(guide = guide_colourbar(
+  scale_colour_gradient2(guide = guide_colourbar(
     direction = "horizontal", title.position = "top"
@@
 
-```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE}
+```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE, dpi=210}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
   ) +
-  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
+  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
   theme(axis.text = element_blank(), axis.ticks = element_blank()) +
@@
 
-```{r plot-fixed-effects, echo=TRUE, eval=TRUE}
+```{r plot-fixed-effects, echo=TRUE, eval=TRUE, dpi=210}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-02-04 doc/appendix-snow-sdmTMB.Rmd
finally get layout working for owl appendix

--- 2022-01-14 doc/appendix-snow-sdmTMB.Rmd
@@
   comment = "#>",
-  fig.width = 7,
-  fig.asp = 1,
+  fig.asp = 1.5, 
+  out.width="100%",
   cache = TRUE,
@@
 library(dplyr)
+# remotes::install_github("pbs-assess/sdmTMB", dependencies = FALSE)
 library(sdmTMB)
@@
 
-```{r mesh, echo=TRUE, cache=FALSE, out.width="70%"}
+```{r mesh, echo=TRUE, cache=FALSE, out.width="80%"}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-```{r qqnorm, out.width="33%"}
+```{r qqnorm, fig.asp = 1, out.width="33%", cache=FALSE}
 qqnorm(pred$resid)
@@
 
-```{r stan-qqnorm, out.width="33%"}
+```{r stan-qqnorm, fig.asp = 1, out.width="33%", cache=FALSE}
 qqnorm(r_mcmc)
@@
 
-```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE, cache=FALSE, out.width="95%"}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=TRUE}
+```{r plot-spatial-effects, echo=TRUE, eval=TRUE, cache=FALSE, out.width="95%"}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE}
+```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE, cache=FALSE, out.width="95%"}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-```{r plot-fixed-effects, echo=TRUE, eval=TRUE}
+```{r plot-fixed-effects, echo=TRUE, eval=TRUE, cache=FALSE, fig.asp = 1.5, out.width="100%"}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-03-14 doc/paper.Rmd
fix typo in owl caption
JL comments

@@
   $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
+      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
   $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
+      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
     Running headline: sdmTMB: spatial and spatiotemporal GLMMs\n"
@@
 Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
+While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
@@
 The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
-Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].
+Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].
 
@@
 
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
-The reasons for this interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
+The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
@@
 Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
-The effect is multiplicative on owl count per NOA unit.
+The effect is multiplicative on owl count per NAO unit.
 
@@
 Speed-wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
-INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.
+INLA/inlabru, or spaMM for spatial models, scale the best with very high-resolution meshes.
 
@@
 Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
-Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-gaussian random fields [e.g., @anderson2019].
-The included TMB `.cpp` file also provides a high-quality, bug-free, and tested model template that can be modified to add additional features.
+Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-Gaussian random fields [e.g., @anderson2019].
+The included TMB `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
 <!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
@@
 We thank the authors of all these packages.
-We thank M. Lindmark, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
+We thank M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
 

### 2022-03-14 analysis/sim-test-index.R
Add some tests

@@ -162,11 +162,24 @@ bias <- est %>%
   summarise(ratio = mean(ratio, na.rm = TRUE))
 bias
 
+est %>%
+  mutate(re = (est - true)/true) %>%
+  summarise(mare = mean(re, na.rm = TRUE))
+
 # MVN simulation
 bias <- est %>%
   mutate(ratio = est_sim/true) %>%
   summarise(ratio = mean(ratio, na.rm = TRUE))
 bias
+
+est %>%
+  mutate(ratio = est_sim/true) %>%
+  summarise(mean_ratio = mean(ratio, na.rm = TRUE),
+    median_ratio = median(ratio))
+
+est %>%
+  mutate(re = (est_sim - true)/true) %>%
+  summarise(mare = median(re, na.rm = TRUE))
 
 # no bias correction
 bias <- est %>%

### 2022-03-14 doc/make.R
Open dropbox

@@ -36,3 +36,5 @@ for (i in seq_along(files)) {
 }
 
 system("cd doc; ./clean-bib.sh")
+
+system("open ~/Dropbox/sdmTMB-paper/")

### 2022-03-16 doc/paper.Rmd
More SK comments

@@
 We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
-The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)`.
+The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
 
@@
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.
 
@@
 VAST is the sole option for fitting some multivariate models;
-alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
+alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs may find sdmTMB more straightforward to use.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
@@
 <!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
-We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
-We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend.
+<!-- sdmTMB does not currently include  -->
+<!-- We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately. -->
+<!-- We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend. -->
 <!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
@@
 We thank the authors of all these packages.
-We thank M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
+We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
 

### 2022-03-16 doc/appendix-model.Rmd
Edits to model appendix

@@
 
-This appendix uses the following notation conventions, generally following the guidance in @edwards2019:
+This appendix uses the following notation conventions, which generally follows the guidance in @edwards2019:
 
@@
 \mu_{\bm{s},t} &=
-f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} +
+f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
 O_{\s,t} +
 \alpha_g +
-\bm{X}_{2,\bm{s},t} \gamma_t +
-\bm{X}_{3,t} \zeta_{\bm{s}} +
+\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
+\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
 \omega_{\s} +
@@
 A single sdmTMB model will rarely, if ever, contain all of the above components.
-Next, we will split the model apart to describe the various parts in more detail using '$\ldots$' to represent the other optional components.
+Next, we will split the model to describe the various parts in more detail using '$\ldots$' to represent the other optional components.
 
@@
 
-Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
+Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv [@wood2017a] users fitting GAMs (generalized additive models).
 Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
@@
 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g.
-`+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g.
+`+ s(x, k = 4)` (see `?mgcv::choose.k`); and various types of splines may be constructed such as cyclic splines to model seasonality, e.g.
 `+ s(month, bs = "cc", k = 12)`.
@@
 `x` is a covariate.
+The formula can contain a single break-point covariate.
 
@@
 These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.
+The formula can contain a single logistic covariate.
 
@@
 
-Spatial random fields, $\omega_{\s}$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.
+Spatial random fields, $\omega_{\s}$, are included if `spatial = 'on'` (or `TRUE`) and omitted if `spatial = 'off'` (or `FALSE`).
 
@@
 $$
-The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
+The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the printed model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
+The 'O' is for 'omega' ($\omega$).
 
@@
 This has been chosen to match the representation in VAST [@thorson2019].
-The marginal standard deviation of $\boldsymbol{\epsilon}_t$ is indicated by `Spatiotemporal SD` in the model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
+The marginal standard deviation of $\boldsymbol{\epsilon}_t$ is indicated by `Spatiotemporal SD` in the printed model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
+The 'E' is for 'epsilon' ($\epsilon$).
 
@@
 where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
-The random fields are assumed independent across time slices.
+The random fields are assumed independent across time steps.
 
@@
 
-First-order auto regressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next.
+First-order auto regressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time step to the next.
 They are defined as
@@
 
-Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time slice to the next are IID. There defined as
+Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time step to the next are IID. They are defined as
 
@@
 
-Spatially varying coefficient (SVC) models are defined as
+Spatially varying coefficient models are defined as
 
@@
 ## IID random or multi-level intercepts
+
+Multilevel/hierchical intercepts are defined as
 
@@
 TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)).
-Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`.
+Anisotropy can be turned on or off with the logical `anisotropy` argument to `sdmTMB()`.
 There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$.
@@
 The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers.
-See `?sdmTMB::add_barrier_mesh` for examples.
+See `?sdmTMB::add_barrier_mesh` for an example.
 

### 2022-03-16 doc/appendix-model.Rmd
Edits to appendix

@@
 $K_\nu$ represents the modified Bessel function of the second kind,
-and $\kappa$ controls the spatial correlation range (decorrelation rate).
+and $\kappa$ represents the decorrelation rate.
 The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GRF to greatly increase computational efficiency [@lindgren2011].
-Internally, the parameter $\kappa$ and $\tau$ are converted to range and marginal standard deviation $\sigma$ as $\textrm{range} = \sqrt{8} / \kappa$ and $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
+Internally, the parameters $\kappa$ and $\tau$ are converted to range and marginal standard deviation $\sigma$ as $\textrm{range} = \sqrt{8} / \kappa$ and $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
 
-In the case of a spatiotemporal model with both spatial and spatiotemporal fields, if `share_range` is `TRUE` in `sdmTMB()` (the default), then a single $\kappa$ and range are estimated with separate $\sigma_\omega$ and $\sigma_\epsilon$.
-This often makes sense since the data are often only weakly informative about $\kappa$.
+In the case of a spatiotemporal model with both spatial and spatiotemporal fields, if `share_range = TRUE` in `sdmTMB()` (the default), then a single $\kappa$ and range are estimated with separate $\sigma_\omega$ and $\sigma_\epsilon$.
+This often makes sense since data are often only weakly informative about $\kappa$.
 If `share_range = FALSE`, then separate $\kappa_\omega$ and $\kappa_\epsilon$ are estimated.

### 2022-03-17 doc/appendix-pcod-inlabru.Rmd
use correct inla mesh
1->2

@@
 
-This model will attempt to replicate the model fit in Appendix 1 as an example of how index standardization can also be achieved with inlabru. 
+This model will attempt to replicate the model fit in Appendix 2 as an example of how index standardization can also be achieved with inlabru. 
 Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
@@
 spde <- inla.spde2.pcmatern(
-  mesh = mesh$mesh,
+  mesh = inla_mesh,
   prior.range = c(10, 0.05), # P(range < 10) = 0.05

### 2022-03-17 doc/appendix-pcod-sdmTMB.Rmd
Add add_utm_columns()

@@
 Here, X and Y are UTM zone 9 coordinates in kilometres.
-If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region.
+If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region or using the built-in `sdmTMB::add_utm_columns()` function.
 

### 2022-03-17 doc/appendix-binomial-inla.Rmd
Update INLA appendix

@@
 
-There are a couple ways to include the fixed effects with `inlabru`. Unfortunately, using factor variables in the standard equation syntax does not work.
+There are a couple ways to include the factor fixed effects with `inlabru`. Unfortunately, using factor variables in the standard equation syntax does not work.
 One option is to create a design matrix and add year effects individually to the formula.
-The other way to set up the fixed coefficients with inlabru is to use the `factor_full` (or `factor_contrast`) model (seen below within `fac()`).
+The other way to set up the factor fixed coefficients with inlabru is to use the `factor_full` (or `factor_contrast`) model (seen below within `fac()`).
 This option causes the location of the factor fixed-effect parameter estimates to end up in `summary.random$fac`.
@@
 set.seed(1)
-post_sdmTMB <- gather_sims(m, n_sims = 2000)
+post_sdmTMB <- gather_sims(m, nsim = 2000)
 post_sdmTMB <- post_sdmTMB %>% 
@@
 x <- list()
-x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>%
+x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factoryear_factor2007) %>%
   as_tibble() %>%
   mutate(variable = "as.factor.year.2007")
-x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>%
+x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factoryear_factor2009) %>%
   as_tibble() %>%
   mutate(variable = "as.factor.year.2009")
-x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>%
+x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factoryear_factor2011) %>%
   as_tibble() %>%
   mutate(variable = "as.factor.year.2011")
-x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>%
+x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factoryear_factor2013) %>%
   as_tibble() %>%
   mutate(variable = "as.factor.year.2013")
-x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>%
+x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factoryear_factor2015) %>%
   as_tibble() %>%
   mutate(variable = "as.factor.year.2015")
-x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>%
+x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factoryear_factor2017) %>%
   as_tibble() %>%

### 2022-03-17 doc/paper.Rmd
Note add_utm_columns()

@@
 An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
+It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
 <!-- TODO: Since this is not particular to your package, maybe the e.g.

### 2022-03-17 doc/clean-bib.sh
Make combined PDF

@@ -1,1 +1,1 @@
-bibtool -s --print.use.tab=Off --print.line.length=80 -i refs.bib -o refs.bib
+/opt/homebrew/bin/bibtool -s --print.use.tab=Off --print.line.length=80 -i refs.bib -o refs.bib

### 2022-03-17 doc/make.R

@@ -26,7 +26,8 @@ names_to <- c(
   "appendix6-speed-validation.pdf"
 )
 
-loc <- "~/Dropbox/sdmTMB-paper/"
+loc <- "~/Documents/sdmTMB-paper/"
+dir.create(loc, showWarnings = FALSE)
 
 for (i in seq_along(files)) {
   from <- paste0(here::here("doc", files[i]), ".pdf")
@@ -37,4 +38,10 @@ for (i in seq_along(files)) {
 
 system("cd doc; ./clean-bib.sh")
 
-system("open ~/Dropbox/sdmTMB-paper/")
+system("open ~/Documents/sdmTMB-paper/")
+
+f <- paste(names_to, collapse = " ")
+
+# system(paste0("cd ~/Documents/sdmTMB-paper/;'/System/Library/Automator/Combine PDF Pages.action/Contents/Resources/join.py' -o sdmTMB-paper-combined.pdf ", f))
+
+system(paste0("cd ~/Documents/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))

### 2022-03-17 doc/appendix-binomial-inla.Rmd
INLA appendix edits

--- 2022-01-28 doc/appendix-binomial-inla.Rmd
@@
 theme_set(theme_light())
+```
+
+We will limit INLA to using a single thread to match how TMB is set up here. Both could optionally be run with parallel processing:
+
+```{r inla-threads, echo=TRUE}
+INLA::inla.setOption(num.threads = "1:1")
 ```

### 2022-03-23 doc/appendix-model.Rmd
X model matrices

@@
 * $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
-* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
+* $\bm{X}^{\mathrm{main}}$, $\bm{X}^{\mathrm{tvc}}$, and $\bm{X}^{\mathrm{svc}}$ represent design matrices (the superscript identifiers 'main' = main effects, 'tvc' = time varying coefficients, and 'svc' = spatially varying coefficients);
 * $\bm{\beta}$ represents a vector of fixed-effect coefficients;
@@
 \begin{aligned}
-\mu_{\bm{s},t} &= f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} \ldots \right)
+\mu_{\bm{s},t} &= f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} \ldots \right)
 \end{aligned}
@@
 
-Within `sdmTMB()`, $\bm{X}_{1,\bm{s},t} \bm{\beta}$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients.
+Within `sdmTMB()`, $\bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta}$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients.
 This main effect formula can contain optional penalized smoothers or non-linear functions as defined below.
@@
 \begin{aligned}
-  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} \bm{X}_{2,s,t} + \ldots \right),\\
+  \mu_{\s,t} &= f^{-1} \left( \ldots +  \bm{X}^{\mathrm{tvc}}_{\s,t} \gamma_{t} + \ldots \right),\\
   \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
@@
 
-where $\gamma_t$ is an example optional time-varying regression parameter and $x_{\s,t}$ is the corresponding predictor value.
+where $\gamma_t$ is an example optional time-varying regression parameter and $\bm{X}^{\mathrm{tvc}}_{\s,t}$ is the corresponding predictor value.
 This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
@@
 \begin{aligned}
-  \mu_{\s,t} &= f^{-1} \left( \ldots + \bm{X}_{3,t} \zeta_{\bm{s}} + \ldots \right),\\
+  \mu_{\s,t} &= f^{-1} \left( \ldots + \bm{X}^{\mathrm{svc}}_{\s, t} \zeta_{\bm{s}} + \ldots \right),\\
   \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
@@
 where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
-Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant spatially for a given time $t$.
-Currently, $\bm{X}_{3,t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
+Usually, $\bm{X}^{\mathrm{svc}}_{\s, t}$ would represent a predictor that is constant spatially for a given time $t$.
+Currently, $\bm{X}^{\mathrm{svc}}_{t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
 For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.

### 2022-03-23 doc/paper.Rmd
Clean up
Fix citation

@@
   \clubpenalty10000
-editor_options:
-  markdown:
-    wrap: sentence
 abstract: |
@@
 The results from an sdmTMB model can be used to generate various quantities of interest.
-For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity  [`get_cog()`; e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.
+For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.
 

### 2022-03-23 doc/appendix-model.Rmd
bold gamma_t

@@
 \alpha_g +
-\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
+\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
 \bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
@@
 \begin{aligned}
-  \mu_{\s,t} &= f^{-1} \left( \ldots +  \bm{X}^{\mathrm{tvc}}_{\s,t} \gamma_{t} + \ldots \right),\\
+  \mu_{\s,t} &= f^{-1} \left( \ldots +  \bm{X}^{\mathrm{tvc}}_{\s,t} \bm{\gamma_{t}} + \ldots \right),\\
   \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
@@
 
-where $\gamma_t$ is an example optional time-varying regression parameter and $\bm{X}^{\mathrm{tvc}}_{\s,t}$ is the corresponding predictor value.
+where $\bm{\gamma_t}$ is an optional vector of time-varying regression parameters and $\bm{X}^{\mathrm{tvc}}_{\s,t}$ is the corresponding model matrix with covariate values.
 This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
@@
 where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
-Usually, $\bm{X}^{\mathrm{svc}}_{\s, t}$ would represent a predictor that is constant spatially for a given time $t$.
+Usually, $\bm{X}^{\mathrm{svc}}_{\s, t}$ would represent a prediction matrix that is constant spatially for a given time $t$.
 Currently, $\bm{X}^{\mathrm{svc}}_{t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.

### 2022-03-23 doc/make.R
Make in dropbox

@@ -26,7 +26,7 @@ names_to <- c(
   "appendix6-speed-validation.pdf"
 )
 
-loc <- "~/Documents/sdmTMB-paper/"
+loc <- "~/Dropbox/sdmTMB-paper/"
 dir.create(loc, showWarnings = FALSE)
 
 for (i in seq_along(files)) {
@@ -38,10 +38,10 @@ for (i in seq_along(files)) {
 
 system("cd doc; ./clean-bib.sh")
 
-system("open ~/Documents/sdmTMB-paper/")
+system("open ~/Dropbox/sdmTMB-paper/")
 
 f <- paste(names_to, collapse = " ")
 
 # system(paste0("cd ~/Documents/sdmTMB-paper/;'/System/Library/Automator/Combine PDF Pages.action/Contents/Resources/join.py' -o sdmTMB-paper-combined.pdf ", f))
 
-system(paste0("cd ~/Documents/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))
+system(paste0("cd ~/Dropbox/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))

### 2022-03-24 doc/appendix-snow-sdmTMB.Rmd
revert attempt
adjust owl figure

--- 2022-02-02 doc/appendix-snow-sdmTMB.Rmd
@@
 
-```{r plot-map, echo=TRUE, eval=TRUE, out.width="80%", cache=FALSE}
+```{r plot-map, echo=TRUE, eval=TRUE, out.width="90%", cache=FALSE}
 ggplot(data = p_proj) +
@@
 
-```{r plot-map-ci, echo=F, eval=TRUE, out.width="85%", cache=FALSE}
+```{r plot-map-ci, echo=F, eval=TRUE, out.width="90%", cache=FALSE}
 p1 <- ggplot(data = p_proj) +

### 2022-04-07 doc/appendix-pcod-VAST.Rmd
Appendix edits
Update VAST comparison; delta

@@
   ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
-  bias.correct = TRUE,
+  bias.correct = FALSE,
   use_anisotropy = FALSE,
@@
 
-sdmTMB does not have built-in delta models; however, we can fit the two components separately and combine the predictions.
-Here, we will fit the binomial presence-absence model and then a model for the positive catches only with a Gamma observation distribution and a log link.
-We will use the VAST-created mesh so the results are comparable.
+TODO sdmTMB:
 
@@
   mesh = fit$spatial_list$MeshList$isotropic_mesh)
+
 tictoc::tic()
-m1 <- sdmTMB(present ~ 0 + as.factor(year), 
-  data = pcod, mesh = mesh, family = binomial(link = "logit"),
-  time = "year", spatiotemporal = "iid", spatial = "on")
+fit_sdmTMB <- sdmTMB(
+  density ~ 0 + as.factor(year), 
+  data = pcod, mesh = mesh, 
+  family = delta_gamma(),
+  time = "year", silent = FALSE,
+  control = sdmTMBcontrol(newton_loops = 1)
+)
 tictoc::toc()
-m1
-
-dat2 <- subset(pcod, density > 0)
-mesh2 <- make_mesh(dat2, xy_cols = c("X", "Y"),
-  mesh = fit$spatial_list$MeshList$isotropic_mesh)
-tictoc::tic()
-m2 <- sdmTMB(density ~ 0 + as.factor(year), 
-  data = dat2, mesh = mesh2, family = Gamma(link = "log"),
-  time = "year", spatiotemporal = "iid", spatial = "on")
-tictoc::toc()
-m2
+fit_sdmTMB
 ```
@@
 ```{r compare-vast-sdmTMB1, echo = FALSE}
-plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft") {
+plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft", sdmTMB_pars = 1) {
   s <- vast_model$parameter_estimates$SD
@@
   years <- years + 0.05
-  points(years, sdmTMB_est$b_j)
-  segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
-    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
+  points(years, sdmTMB_est$b_j[,sdmTMB_pars])
+  segments(years, sdmTMB_est$b_j[,sdmTMB_pars] - 2 * sdmTMB_sd$b_j[,sdmTMB_pars], 
+    years, sdmTMB_est$b_j[,sdmTMB_pars] + 2 * sdmTMB_sd$b_j[,sdmTMB_pars],
     col = "red")
@@
 ```{r compare-vast-sdmTMB2, echo = TRUE}
-plot_betas(fit, m1, "beta1_ft")
-plot_betas(fit, m2, "beta2_ft")
+par(mfrow = c(2, 1), cex = 0.8, mar = c(1.5, 1, 1, 1), oma = c(2, 3, 1, 1))
+plot_betas(fit, fit_sdmTMB, "beta1_ft", sdmTMB_pars = 1)
+plot_betas(fit, fit_sdmTMB, "beta2_ft", sdmTMB_pars = 2)
 ```
@@
 
-```{r sdmTMB-sim-index, cache=TRUE, echo=TRUE}
+```{r sdmTMB-index, cache=TRUE, echo=TRUE}
 tictoc::tic()
-set.seed(28239)
-p_bin <- predict(m1, newdata = qcs_grid, sims = 500)
-p_pos <- predict(m2, newdata = qcs_grid, sims = 500)
-
-p_bin_prob <- plogis(p_bin)
-p_pos_exp <- exp(p_pos)
-p_combined <- log(p_bin_prob * p_pos_exp)
-ind_combined_sims <- get_index_sims(p_combined, area = rep(4, nrow(p_combined)))
+p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
+  area = rep(4, nrow(qcs_grid))
+)
+ind <- get_index(p, bias_correct = FALSE)
 tictoc::toc()
@@
 library(ggplot2)
-vast_i1 <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
+vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
   mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
@@
   mutate(upr = exp(log(est) + qnorm(0.975) * se))
-sdm_i <- ind_combined_sims %>% mutate(index = "sdmTMB")
+sdm_i <- ind %>% mutate(index = "sdmTMB")
 both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
 ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
-  geom_ribbon(alpha = 0.3) +
-  geom_line() + 
+  geom_ribbon(alpha = 0.1) +
+  geom_line(alpha = 0.8) + 
   ylim(0, max(both_i$upr)) +
@@
 ```{r}
-plot(ind_sdmTMB$est, vast_i$est);abline(0, 1)
-plot(ind_sdmTMB$upr, vast_i$upr);abline(0, 1)
-plot(ind_sdmTMB$lwr, vast_i$lwr);abline(0, 1)
+plot(ind$est, vast_i$est[vast_i$est != 0]);abline(0, 1)
+plot(ind$upr, vast_i$upr[vast_i$est != 0]);abline(0, 1)
+plot(ind$lwr, vast_i$lwr[vast_i$est != 0]);abline(0, 1)
 
-(ind_sdmTMB$est - vast_i$est) / vast_i$est
-(ind_sdmTMB$upr - vast_i$upr) / vast_i$upr
-(ind_sdmTMB$lwr - vast_i$lwr) / vast_i$lwr
+(ind$est - vast_i$est[vast_i$est != 0]) / vast_i$est[vast_i$est != 0]
+(ind$upr - vast_i$upr[vast_i$est != 0]) / vast_i$upr[vast_i$est != 0]
+(ind$lwr - vast_i$lwr[vast_i$est != 0]) / vast_i$lwr[vast_i$est != 0]
 ```

### 2022-04-12 doc/paper.Rmd
Updates to paper with latest sdmTMB

@@
     | 3.
-We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
+We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., delta/hurdle models, penalized smoothers and break-point effects, anisotropy, abundance index standardization).
     | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
@@
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
+Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
 
@@
 We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
+An alternative would be the `delta_gamma()` family to specify a delta/hurdle model [@aitchison1955] or to model catch weight with an offset for log(area swept).
 <!-- We could instead model catch weight as the response with an offset for log(area). -->
@@
 sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
-sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
+sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with various limitations.
 To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
@@
 
-```{r compare-table}
+```{r compare-table, results='asis'}
 library(kableExtra)
@@
 
-We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
 \@ref(fig:timing), Appendix 6).
-In this test, for mesh resolutions under 500 nodes, sdmTMB was faster than inlabru or mgcv.
-inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
+In this test, sdmTMB was the fastest; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
 mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
@@
 (ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
-Lines represent means and ribbons ranges across 20 random iterations.
+Lines represent means and ribbons 95% quantiles across 50 random iterations.
 Note the log x and y axes.
-sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
+<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
 VAST is similar to sdmTMB and so is not shown.
@@
 <!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
-<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
+Note that spaMM only fits spatial, not spatiotemporal, models.
+All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
 <!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
@@
 ```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
-f <- here::here("figs", "timing-spatial.pdf")
+f <- here::here("figs", "timing-spatial-m1-april11.pdf")
 knitr::include_graphics(f)
@@
 VAST is the sole option for fitting some multivariate models;
-alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs may find sdmTMB more straightforward to use.
+alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
-Speed-wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
-INLA/inlabru, or spaMM for spatial models, scale the best with very high-resolution meshes.
+<!-- Speed-wise, the TMB-based packages sdmTMB and VAST were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
+-->
+
+Speed-wise, sdmTMB (and by association VAST) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than INLA/inlabru across the range of mesh complexities considered here.
+These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of models that are generally slower to fit than GLMMs without random fields.
+However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by INLA/inlabru is likely to be considerably faster than passing the same model from sdmTMB/VAST to tmbstan for full Bayesian inference.
+Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores allowing for better scaling with very large models.
 
@@
 Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
-Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-Gaussian random fields [e.g., @anderson2019].
+Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
 The included TMB `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.

### 2022-04-12 doc/make.R

@@ -38,10 +38,11 @@ for (i in seq_along(files)) {
 
 system("cd doc; ./clean-bib.sh")
 
-system("open ~/Dropbox/sdmTMB-paper/")
 
 f <- paste(names_to, collapse = " ")
 
 # system(paste0("cd ~/Documents/sdmTMB-paper/;'/System/Library/Automator/Combine PDF Pages.action/Contents/Resources/join.py' -o sdmTMB-paper-combined.pdf ", f))
 
 system(paste0("cd ~/Dropbox/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))
+
+system("open ~/Dropbox/sdmTMB-paper/")

### 2022-04-12 doc/appendix-snow-sdmTMB.Rmd
Tweak owl appendix

@@
 
-```{r mesh, echo=TRUE, cache=FALSE, out.width="4in"}
+```{r mesh, echo=TRUE, out.width="4in"}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
+We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the DHARMa package.
 If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead. 
@@
 To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`).
-We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `n_sims = 500`). 
+We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `nsim = 500`). 
 The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws (in this case we will use 80%) and adding these values to our prediction data frame.
@@
 
-n_sims <- 500L
-zeta_sim <- predict(m, sims = n_sims, sims_var = "zeta_s")
-sims <- spread_sims(m, n_sims = n_sims)
+nsim <- 500
+zeta_sim <- predict(m, nsim = nsim, sims_var = "zeta_s")
+sims <- spread_sims(m, nsim = nsim)
 beta <- sims$nao
@@
 
-```{r plot-map, echo=TRUE, eval=TRUE, out.width="90%", cache=FALSE}
+```{r plot-map, echo=TRUE, eval=TRUE, out.width="90%", echo=FALSE, cache=FALSE}
 ggplot(data = p_proj) +
@@
 
-```{r plot-map-ci, echo=F, eval=TRUE, out.width="90%", cache=FALSE}
+```{r plot-map-ci, echo=F, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
 p1 <- ggplot(data = p_proj) +

### 2022-04-12 analysis/timing.R
Updating timing test

@@ -307,7 +307,7 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
 to_run <- expand.grid(
   n_obs = c(1000L),
   max.edge = c(0.06, 0.075, 0.1, 0.15, 0.2),
-  iter = seq_len(20L)
+  iter = seq_len(50L)
 )
 
 # # # short test version
@@ -323,13 +323,14 @@ nrow(to_run)
 # test2 <- sim_fit_time(n_obs = 1000L, max.edge = 0.06, family = gaussian(), seed = 1 * 29212, iter = 1)
 
 # scramble for parallel task assignment:
-to_run <- to_run[sample(seq_len(nrow(to_run)), replace = FALSE), ]
+# to_run <- to_run[sample(seq_len(nrow(to_run)), replace = FALSE), ]
 
 # out <- purrr::pmap_dfr(to_run, sim_fit_time)
 
 # plan(multisession, workers = 5L)
 
 # out <- furrr::future_pmap_dfr(
+system.time({
 out <- purrr::pmap_dfr(
   to_run,
   sim_fit_time
@@ -344,9 +345,26 @@ out <- purrr::pmap_dfr(
   #   packages = c("mgcv", "inlabru", "INLA", "ggplot2", "dplyr", "spaMM", "sdmTMB")
   # )
 )
+})
 
-saveRDS(out, file = "analysis/timing-cache-parallel-spaMM-PE.rds")
-out <- readRDS("analysis/timing-cache-parallel-spaMM-PE.rds")
+## # saveRDS(out, file = "analysis/timing-sdmTMB-ad-SA-March31.rds")
+## out2 <- readRDS("analysis/timing-sdmTMB-ad-SA-March31.rds")
+##
+## # saveRDS(out, file = "analysis/timing-cache-parallel-spaMM-SA-March23.rds")
+## out <- readRDS("analysis/timing-cache-parallel-spaMM-SA-March23.rds")
+##
+##
+## out$sdmTMB <- NULL
+## out$sdmTMB_norm <- NULL
+##
+## out3 <- left_join(out, select(out2, sdmTMB, sdmTMB_norm, seed, iter, cutoff, n_obs, mesh_n, max.edge))
+## nrow(out3)
+## out <- out3
+## out <- select(out, sdmTMB, sdmTMB_norm, inlabru_eb:inla_eb_error)
+
+saveRDS(out, file = "analysis/timing-cache-parallel-2022-04-11.rds")
[…]

### 2022-05-03 doc/paper.Rmd
Copy editing

@@
 (2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
-(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`).
-(4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
+(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
+(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
 <!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.
 -->
-For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
-Simulation-based residuals [@dharma] (`dharma_residuals()`), one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
+For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
+A version that uses Markov chain Monte Carlo (MCMC) to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
+Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
 (5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
@@
 \@ref(fig:timing), Appendix 6).
-In this test, sdmTMB was the fastest; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
-mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
-With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
+In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
+mgcv was most affected by mesh resolution.
+<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
+-->
 <!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
@@
 Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
+Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
 <!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
@@
 Speed-wise, sdmTMB (and by association VAST) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than INLA/inlabru across the range of mesh complexities considered here.
-These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of models that are generally slower to fit than GLMMs without random fields.
+These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
 However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by INLA/inlabru is likely to be considerably faster than passing the same model from sdmTMB/VAST to tmbstan for full Bayesian inference.
-Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores allowing for better scaling with very large models.
+Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores.
 
@@
 Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
-These data present statistical challenges to modelling them effectively and efficiently as appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
 We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.

### 2022-05-03 doc/make.R

@@ -39,10 +39,10 @@ for (i in seq_along(files)) {
 system("cd doc; ./clean-bib.sh")
 
 
-f <- paste(names_to, collapse = " ")
+# f <- paste(names_to, collapse = " ")
 
 # system(paste0("cd ~/Documents/sdmTMB-paper/;'/System/Library/Automator/Combine PDF Pages.action/Contents/Resources/join.py' -o sdmTMB-paper-combined.pdf ", f))
 
-system(paste0("cd ~/Dropbox/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))
+# system(paste0("cd ~/Dropbox/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))
 
 system("open ~/Dropbox/sdmTMB-paper/")

### 2022-05-05 doc/paper.Rmd
Add affiliation; keep_tex

@@
 author: "|
-  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
-  Eric J. Ward$^2$,
+  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
+  Eric J. Ward$^3$,
   Philina A. English$^1$,
-  Lewis A. K. Barnett$^3$\n|
+  Lewis A. K. Barnett$^4$\n|
   $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
-  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
+  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
+  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
       National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
-  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
+  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
       National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
@@
       highlight: "monochrome"
+      keep_tex: true
 csl: mee.csl

### 2022-05-05 doc/appendix-pcod-VAST-poisson-link.Rmd
Add draft poisson-link comparison

--- 2022-04-07 doc/appendix-pcod-VAST.Rmd
@@
   RhoConfig = RhoConfig,
-  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
+  ObsModel = c(2, 1), # PG delta; c(10, 2) for Tweedie
   bias.correct = FALSE,
@@
  # create folder for saved output:
-dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)
+dir.create(paste0(here("doc", "appendix-VAST-poisson-link")), showWarnings = FALSE)
 
@@
 pcod <- as.data.frame(pcod) # ensure not a tibble
-f <- here("doc", "appendix-VAST", "vast-cache.rds")
+f <- here("doc", "appendix-VAST-poisson-link", "vast-cache.rds")
 if (!file.exists(f)) {
@@
     input_grid = input_grid,
-    working_dir = paste0(here("doc", "appendix-VAST"), "/")
+    working_dir = paste0(here("doc", "appendix-VAST-poisson-link"), "/")
   )
@@
 
-We can look at parameter estimates and their standard errors. First we see estimates from the binomial component and second we see estimates from the positive Gamma component.
+We can look at parameter estimates and their standard errors.
 
@@
 ```{r sdmTMB-example, echo = TRUE, cache=TRUE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
+mesh <- sdmTMB::make_mesh(pcod, xy_cols = c("X", "Y"),
   mesh = fit$spatial_list$MeshList$isotropic_mesh)
@@
   data = pcod, mesh = mesh, 
-  family = delta_gamma(),
-  time = "year", silent = FALSE,
+  family = delta_poisson_link_gamma(),
+  time = "year", silent = F,
   control = sdmTMBcontrol(newton_loops = 1)
@@
   check_residuals = FALSE,
-  working_dir = paste0(here("doc", "appendix-VAST"), "/")
+  working_dir = paste0(here("doc", "appendix-VAST-poisson-link"), "/")
 )
@@
 ```{r plot-knots, echo = TRUE, out.width="6in"}
-knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
+knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "Data_and_knots.png"))
 ```
@@
 ```{r plot-density, echo = TRUE, out.width="6in"}
-knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
+knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "ln_density-predicted.png"))
 ```
@@
 ```{r plot-index, echo = TRUE, out.width="3in"}
-if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
-  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
+if (file.exists(here("doc", "appendix-VAST-poisson-link", "Index-Biomass.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "Index-Biomass.png"))
 }
-if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
-  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
+if (file.exists(here("doc", "appendix-VAST-poisson-link", "Index.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "Index.png"))
 }
@@
 tictoc::tic()
-p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
-  area = rep(4, nrow(qcs_grid))
-)
-ind <- get_index(p, bias_correct = FALSE)
+p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
+
+d <- p$data
+n_i <- exp(d$est1)
+w_i <- exp(d$est2)
+p_i <- 1 - exp(-n_i)
+r_i <- (n_i * w_i) / p_i
+d_i <- p_i * r_i
+d2_i <- exp(d$est1 + d$est2)
+plot(d_i, d2_i)
+
+library(dplyr)
+nd <- qcs_grid
+# nd <- pcod
+nd$d_i <- d_i
+ind_r <- group_by(nd, year) %>% 
+  summarise(total = sum(d_i))
+ind_r$total * 1
+
+n_i <- exp(r$proj_eta[,1])
+w_i <- exp(r$proj_eta[,2])
+p_i <- 1 - exp(-n_i)
+r_i <- (n_i * w_i) / p_i
+d_i <- p_i * r_i
+d2_i <- exp(d$est1 + d$est2)
+nd$d_i <- d_i
+ind_r_tmb <- group_by(nd, year) %>% 
+  summarise(total = sum(d_i))
+ind_r_tmb$total * 1
+
+ind <- get_index(p, bias_correct = FALSE, area = 1)
 tictoc::toc()
@@
 library(ggplot2)
-vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
+vast_i <- read.csv(here("doc/appendix-VAST-poisson-link/Index.csv")) %>%
   mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 

### 2022-05-05 analysis/inlabru-examples/02_random_fields.R
Cleaning up

@@ -1,403 +1,0 @@
-#' Fitting georeferenced data in 1D and 2D
-#'=========================================================
-
-#' Set things up
-#+results="hide",warning=FALSE,message=FALSE
-library(INLA)
-library(inlabru)
-library(mgcv)
-
-#' Set default INLA values for tutorial session,
-#' and make a shortcut to a nicer colour scale:
-#' 
-#+results="hide",warning=FALSE,message=FALSE
-
-init.tutorial()
-colsc <- function(...) {
-  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11,"RdYlBu")),
-                       limits = range(..., na.rm=TRUE))
-}
-
-#' This is precisely what `init.tutorial` does:
-#+echo=FALSE
-
-print(init.tutorial)
-
-#' `iinla.verbose=TRUE` prints iterated inla progression information
-#'
-#' `int.strategy="eb"` limits INLA to Empirical Bayes so that the
-#' examples run faster; no posterior integration over $\theta$ is performed.
-#' See `?INLA::control.inla` for some more information.
-#' 
-#' The `control.compute` options turn on information needed for posterior sampling, and
-#' tells INLA to compute DIC and WAIC. See `?INLA::control.compute` for more information.
-
-
-
-#' Fitting to 1-dimensional gridded data (counts)
-#'=========================================================
-
-#' Get the data
-#'-----------------------------------
-#+results="hide",warning=FALSE,message=FALSE
-
-data(Poisson2_1D)
-
-#' Put the count data in `cd` (just because '`cd`' is less to type than '`countdata2`'.)
-
-cd <- countdata2
-
-#' Take a look at the count data.
-#' 
-#+warning=FALSE,message=FALSE
-
-cd
-ggplot(cd) + geom_point(aes(x, y = count)) + ylim(0, max(cd$count))
-
-#' _Tip_: `RStudio > Help > Cheatsheets > Data visualisation with ggplot2` is a useful 
-#' reference for `ggplot2` syntax.
-
-
- 
-#' Fit a Generalised Additive Model (GAM)
-#'-----------------------------------
-#' If you're not familiar with GAMs and the syntax of `gam` don't worry, the point of 
-#' this is just to provide something to which we can compare the `inlabru` model fit.
-#' 
-#+results="hide",warning=FALSE,message=FALSE
-
-fit2.gam <- gam(count ~ s(x, k = 10) + offset(log(exposure)), family = poisson(), data = cd)
-
[…]

### 2022-05-05 analysis/sim-test-index-delta.R

@@ -1,364 +1,0 @@
-library(dplyr)
-library(testthat)
-library(ggplot2)
-library(sdmTMB)
-library(future)
-theme_set(ggsidekick::theme_sleek())
-plan(multisession, workers = floor(availableCores() / 2) - 0)
-# options(future.rng.onMisuse = "ignore")
-
-SEED <- 1
-set.seed(SEED)
-N <- 625 # per year
-time_steps <- 20
-sigma_O <- 0
-sigma_E <- 0.6
-phi <- 9
-tweedie_p <- 1.5 # Tweedie p
-.range <- 70
-
-# x <- seq(-1, 1, length.out = 25)
-# y <- seq(-1, 1, length.out = 25)
-x <- seq(340, 580, length.out = 25)
-y <- seq(5500, 5800, length.out = 25)
-loc <- expand.grid(x = x, y = y)
-x <- rep(loc$x, time_steps)
-y <- rep(loc$y, time_steps)
-loc <- data.frame(x = x, y = y)
-spde <- sdmTMB::make_mesh(loc, xy_cols = c("x", "y"), cutoff = 10)
-plot(spde)
-plot(spde$mesh, asp = 1)
-
-sim_test_index <- function(i) {
-  set.seed(i * 1029)
-  betas <- rnorm(time_steps, 1, 1)
-  dat <- data.frame(year = rep(seq_len(time_steps), each = N))
-  s <- sdmTMB_simulate(
-    formula = ~ 0 + as.factor(year),
-    data = dat,
-    mesh = spde,
-    B = betas,
-    phi = phi,
-    range = .range,
-    sigma_O = sigma_O,
-    sigma_E = sigma_E,
-    seed = SEED * i,
-    time = "year",
-    family = tweedie(),
-    tweedie_p = tweedie_p
-  )
-  s <- mutate(s, year = dat$year)
-  s <- mutate(s, x = loc$x)
-  s <- mutate(s, y = loc$y)
-  s_sampled <- s %>% group_by(year) %>% slice_sample(n = 200L)
-
-  if (FALSE) {
-    ggplot(s, aes(x, y, fill = mu)) +
-      geom_raster() +
-      geom_point(aes(size = observed), data = s_sampled, pch = 21, fill = NA) +
-      scale_fill_viridis_c() +
-      scale_size_area() +
-      facet_wrap(vars(year)) +
-      coord_cartesian(expand = FALSE)
-  }
-
-  FieldConfig <- matrix(c(0, "IID", "IID", 0, "IID", "IID"),
[…]

### 2022-05-06 doc/appendix-binomial-inla.Rmd
Double space; update appendices for latest sdmTMB

@@
     number_sections: true
+header-includes:
+  \usepackage[left]{lineno}
+  \usepackage{bm}
+  \usepackage{amssymb}
+  \linenumbers
 ---

### 2022-05-06 doc/appendix-pcod-VAST-tweedie.Rmd
Double space; update appendices for latest sdmTMB

@@
 ```{r sdmTMB-index, cache=TRUE, echo=TRUE}
-pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE, area = 4)
-ind_sdmTMB <- get_index(pred)
+pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
+ind_sdmTMB <- get_index(pred, area = 4)
 ```

### 2022-05-06 doc/appendix-pcod-VAST.Rmd
Double space; update appendices for latest sdmTMB

@@
 tictoc::tic()
-p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
-  area = rep(4, nrow(qcs_grid))
-)
-ind <- get_index(p, bias_correct = FALSE)
+p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
+ind <- get_index(p, bias_correct = FALSE, area = rep(4, nrow(qcs_grid)))
 tictoc::toc()

### 2022-05-06 doc/appendix-pcod-sdmTMB.Rmd
Double space; update appendices for latest sdmTMB

@@
 csl: mee.csl
+header-includes:
+  \usepackage[left]{lineno}
+  \usepackage{bm}
+  \usepackage{amssymb}
+  \linenumbers
 ---
@@
 
+An alternative would be to model catch as the response with an offset (`offset` argument) for log area swept.
+
 ```{r fit-pcod, warning=FALSE, message=FALSE}
@@
 ```{r sim-resids, out.width="4in", warning=FALSE, results='hide'}
+set.seed(123)
 simulate(fit, nsim = 500) %>%
   dharma_residuals(fit)
+```
+
+Or perhaps better, use MCMC residuals:
+
+```{r mcmc-resids, warning=FALSE, results='hide', cache=TRUE}
+set.seed(123)
+r <- residuals(fit,
+  type = "mle-mcmc", 
+  mcmc_iter = 400, mcmc_warmup = 300
+)
+```
+
+```{r mcmc-resids-vis, out.width="4in"}
+qqnorm(r)
+qqline(r)
 ```
@@
 ```{r sims}
-pred_sims <- predict(fit, newdata = qcs_grid, sims = 500)
+pred_sims <- predict(fit, newdata = qcs_grid, nsim = 500)
 dim(pred_sims)
@@
   fit,
-  newdata = qcs_grid, return_tmb_object = TRUE,
-  area = qcs_grid$area
+  newdata = qcs_grid, 
+  return_tmb_object = TRUE
 )
-ind <- get_index(pred2)
-saveRDS(ind, "sdmTMB-index.rds")
+ind <- get_index(pred2, area = qcs_grid$area, bias_correct = TRUE)
 
@@
 
-The above example used the default `bias_correct = FALSE` in `get_index()`.
+The above example uses `bias_correct = TRUE` in `get_index()`.
 The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in the [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias].
 Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
-Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but can take considerable time and memory.
-
-As an alternative, we can use the output from the joint precision matrix simulation draws to calculate our index.
-This should approximate the output of `get_index(x, bias_correct = TRUE)`.
-However, we note that there may be situations where this approach may fail or be biased and experimentation is ongoing.
-
-```{r index-sims, fig.asp=0.45, out.width="5in"}
-ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)
-
-ggplot(ind, aes(year, est / 1000)) +
-  geom_line() +
-  geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000), alpha = 0.4) +
-  ylab("Biomass (t)")
-```
+Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but takes more time and memory than skipping the bias correction.
 

### 2022-05-06 doc/appendix-snow-sdmTMB.Rmd
Double space; update appendices for latest sdmTMB

@@
 ## Basic inspection
+
 ```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
@@
 
-According to AIC, the `nbinom2()` (negative binomial, NB2) distribution appears to have the most parsimonious fit. 
+According to AIC, the `nbinom2()` (negative binomial, NB2) distribution has the most parsimonious fit.
+We could also check residual distributions at this stage.
 We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and check that the gradients are all small enough.
 If the gradients aren't quite small enough, we can easily run additional optimization steps using the `run_extra_optimization()` function.
+
+```{r sanity, echo=TRUE}
+sanity(m2)
+```
 
@@
 )
-
-if (max(m$gradients) > 0.01) {
-  m <- run_extra_optimization(m, newton_loops = 1L, nlminb_loops = 0L)
-}
 ```
@@
 pred <- predict(m)
+set.seed(19208)
 pred$resid <- residuals(m)
@@
 
-More accurate residuals can be generated using tmbstan, but this is a very slow process!
-In this case, it took more than 2.5 hrs.
+More accurate residuals can be generated using tmbstan, but this a slower process.
+These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via tmbstan/Stan.
+To calculate these, we have to use the version of our model fit without REML.
+We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
 
-```{r stan-resids, eval=FALSE}
-stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
-eta <- predict(m, tmbstan_model = stan_fit)
-mu <- m$family$linkinv(as.numeric(eta))
-r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
-saveRDS(r_mcmc, "data/mcmc_residuals.rds")
-```
-
-```{r stan-dat, echo=FALSE, cache=FALSE}
-r_mcmc <- readRDS(here::here("data/mcmc_residuals.rds"))
+```{r stan-resids}
+f <- here::here("data/mcmc_residuals.rds") # for caching
+set.seed(82919)
+if (!file.exists(f)) {
+  mcmc_resids <- residuals(m2, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
+  saveRDS(mcmc_resids, f)
+} else {
+  mcmc_resids <- readRDS(f)
+}
 ```
@@
 ```{r stan-qqnorm, out.width="33%"}
-qqnorm(r_mcmc)
-qqline(r_mcmc)
+qqnorm(mcmc_resids)
+qqline(mcmc_resids)
 ```
@@
 We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the DHARMa package.
-If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead. 
-This is not built directly into sdmTMB, but can be achieved by fitting the two components separately.
-See the steps in the vignette [Fitting delta models with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/delta-models.html).
+If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.
 
@@
 ```
-
-<!-- ## Cross-validation -->
-
-<!-- Cross-validation can also be used to assess model performance. -->
-<!-- Here, we use the `future` package to speed up the running of multiple models and save a smaller version of the output that doesn't include the full model information. -->
-<!-- We use `k_folds = 5`, which amounts to training on 80% of the data and testing with 20% for each model. -->
-<!-- Block cross-validation, which can be done using custom folds defined using the `fold_ids` argument, is generally recommended for spatiotemporal data. -->
-<!-- However, random folds created by `sdmTMB_cv()` are appropriate in this case, because we are only interested in predicting within the spatial and temporal range of our data and are attempting to estimate the spatial variation in the effect of NAO.
--->
-
-<!-- ```{r cv, eval=FALSE, cache=FALSE} -->
-<!-- library(future) -->
-<!-- plan(multisession) -->
-<!-- options(future.rng.onMisuse = "ignore") -->
-
-<!-- m_cv <- sdmTMB_cv(count ~ 1 + nao + (1 | year_f), -->
-<!--   spatial_varying = ~ 0 + nao, -->
-<!--   time = "year", -->
-<!--   spatial = "on", -->
-<!--   spatiotemporal = "iid", -->
-<!--   family = nbinom2(link = "log"), -->
-<!--   reml = TRUE, -->
-<!--   use_initial_fit = TRUE, -->
-<!--   data = snow, mesh = mesh, -->
-<!--   k_folds = 5 -->
-<!-- ) -->
-<!-- plan(sequential) -->
-<!-- saveRDS(m_cv, file = here::here("data/owlcvallreml.rds")) -->
-<!-- # m_cv$models[[1]] # to retrieve just the model built excluding the first fold -->
-<!-- cv <- m_cv -->
-<!-- cv$models <- NA -->
-<!-- saveRDS(cv, file = here::here("data/owlcvdatreml.rds")) -->
-<!-- ``` -->
-
-<!-- First, we should check that all the models converged and have gradients < 0.01. -->
-
-<!-- ```{r explore-cv, message=FALSE, cache=FALSE} -->
-<!-- cv <- readRDS(here::here("data/owlcvdatreml.rds")) -->
-<!-- cv$converged -->
-<!-- max(cv$max_gradients) -->
-<!-- ``` -->
-
-<!-- If we were conducting model selection, we might contrast the log likelihoods (`cv$sum_loglik`), and the expected log predictive density (`cv$elpd`) for the left out data between different model configurations instead of relying on AIC.
--->
-<!-- We can also check the prediction accuracy on left-out data by plotting the `cv_predicted` against the observed `count` in log space. -->
-
-<!-- ```{r cv-pred-obs, out.width="70%", cache=FALSE} -->
-<!-- cvdat <- cv$data %>% filter(cv_loglik > -60) # this removes one extreme outlier -->
-<!-- ggplot(cvdat, aes(log(count + 1), log(cv_predicted + 1), colour = -cv_loglik)) + -->
-<!--   geom_abline(slope = 1, colour = "gray") + -->
-<!--   geom_point(size = 2, alpha = 0.25) + -->
-<!--   scale_colour_viridis_c( -->
-<!--     direction = -1, option = "A", begin = 0.2, end = 0.9 -->
-<!--   ) + -->
-<!--   coord_fixed(ylim = c(0, 3.8)) + -->
-<!--   theme_bw() -->
-<!-- ``` -->
-
-<!-- These predictions for left-out locations match the observed counts just as well as those based on the model built using all the data shown below. -->
-<!-- # Does this mean that we are at least not over-fit, despite the spatiotemporal random fields and spatially varying coefficient? -->
-
-<!-- ```{r full-pred-obs, echo=FALSE, out.width="70%"} -->
-<!-- ggplot(pred, aes(log(count + 1), log(exp(est) + 1), colour = resid)) + -->
-<!--   geom_abline(slope = 1, colour = "gray") + -->
-<!--   # geom_point(size = 0.75, alpha=0.5) + facet_wrap(~year) + -->
-<!--   geom_point(size = 2, alpha = 0.25) + -->
-<!--   coord_fixed(ylim = c(0, 3.8)) + -->
-<!--   scale_colour_viridis_c( -->
-<!--     "residuals", -->
-<!--     direction = -1, option = "A", begin = 0.2, end = 0.9 -->
-<!--   ) + -->
-<!--   theme_bw() #+ theme(strip.background = element_rect(fill = NA, colour = NA)) -->
-<!-- ``` -->
-
-<!-- We can also explore the spatiotemporal variation in the predictive accuracy of our model by plotting the `cv_loglik`. -->
-
-<!-- ```{r cv-map, cache = FALSE} -->
-<!-- ggplot(cvdat, aes(X, Y, colour = -cv_loglik)) + -->
-<!--   geom_point(size = 0.25) + -->
-<!--   scale_colour_viridis_c( -->
-<!--     direction = -1, option = "A", begin = 0.2, end = 0.9, -->
-<!--     guide = guide_colourbar(title.position = "top") -->
-<!--   ) + -->
-<!--   coord_fixed() + -->
-<!--   theme(axis.text = element_blank(), axis.ticks = element_blank()) + -->
-<!--   facet_wrap(~year) -->
-<!-- ``` -->
-
-<!-- \clearpage -->
 
@@
 nsim <- 500
-zeta_sim <- predict(m, nsim = nsim, sims_var = "zeta_s")
+zeta_s <- predict(m, nsim = nsim, sims_var = "zeta_s")
 sims <- spread_sims(m, nsim = nsim)
 beta <- sims$nao
-combined <- beta + t(zeta_sim)
+combined <- beta + t(zeta_s)
 
-p$zeta_sim <- apply(t(zeta_sim), 2, median)
-p$nao_effect_sim <- apply(combined, 2, median)
-p$nao_effect_lwr <- apply(combined, 2, quantile, probs = 0.10)
-p$nao_effect_upr <- apply(combined, 2, quantile, probs = 0.90)
+p$zeta_s <- as.numeric(apply(t(zeta_s), 2, median))
+p$nao_effect_sim <- as.numeric(apply(combined, 2, median))
+p$nao_effect_lwr <- as.numeric(apply(combined, 2, quantile, probs = 0.10))
+p$nao_effect_upr <- as.numeric(apply(combined, 2, quantile, probs = 0.90))
 
@@
     mean_est_count = exp(mean(est)),
-    nao_effect = b[b$term == "nao", 2] + mean(zeta_s),
+    nao_effect = b[b$term == "nao", "estimate"] + mean(zeta_s),
     nao_sim = mean(nao_effect_sim),
@@
 ```{r}
-nao_effect_lwr95 <- apply(combined, 2, quantile, probs = 0.025)
+nao_effect_lwr95 <- as.numeric(apply(combined, 2, quantile, probs = 0.025))
 ```

### 2022-05-06 doc/appendix-validation.Rmd
Double space; update appendices for latest sdmTMB

--- 2022-01-28 doc/appendix-validation.Rmd
@@
   \usepackage{amssymb}
+  \linenumbers
 ---
@@
 
-We conducted 20 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.
+We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.
+The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.1.3 and the default Matrix library. The following package versions were used:
+
+```{r pkg-versions}
+packageVersion("INLA")
+packageVersion("inlabru")
+packageVersion("mgcv")
+packageVersion("spaMM")
+packageVersion("sdmTMB")
+```
 
@@
 This appendix illustrates a comparison of penalized smoothers with mgcv and a comparison of IID random intercepts with glmmTMB.
-<!-- Additionally, it illustrates some simulation testing of a spatial random field model with the main observation families within the sdmTMB package. -->
 
@@
 
-```{r plot-smooth, echo=TRUE, fig.asp=0.6, out.width="5in"}
+```{r plot-smooth, echo=TRUE, fig.asp=0.8, out.width="5in"}
+par(cex = 0.8)
+# sdmTMB:
 plot_smooth(m)
+# also see the more general visreg::visreg()
 
+# mgcv:
 x2 <- seq(0, 1, length.out = 100)

### 2022-05-06 doc/appendix-model.Rmd
Double space; update appendices for latest sdmTMB
SVC not limited to 1 now

@@
 where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
-Usually, $\bm{X}^{\mathrm{svc}}_{\s, t}$ would represent a prediction matrix that is constant spatially for a given time $t$.
-Currently, $\bm{X}^{\mathrm{svc}}_{t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
+Usually, $\bm{X}^{\mathrm{svc}}_{\s, t}$ would represent a prediction matrix that is constant spatially for a given time $t$ as defined by a one-sided formula supplied to `spatial_varying`.
 For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.

### 2022-05-06 doc/appendix-model.Rmd
Update offset

@@
 
-Offset terms can be included with the reserved word `offset` in the formula.
-E.g., `y ~ x + offset`.
+Offset terms can be included through the `offset` argument in `sdmTMB()`.
 These are included in the linear predictor as
@@
 where $O_{\s,t}$ is an offset term---a **log transformed** variable without a coefficient (assuming a log link).
+The offset is *not* included in the prediction.
+Therefore, if `offset` represents a measure of effort, for example, the prediction is for one unit of effort (`log(1) = 0`).
 

### 2022-05-12 doc/appendix-pcod-VAST.Rmd
Update VAST/sdmTMB delta appendix

@@
 settings <- make_settings(
-  n_x = 205, # number of vertices in the SPDE mesh
+  n_x = 185, # number of vertices in the SPDE mesh
   Region = "User",
@@
 pcod$effort <- 1
+if (FALSE) { # to check offset/effort
+  set.seed(1)
+  pcod$effort <- exp(rnorm(nrow(pcod), mean = 0, sd = 0.2))
+}
 pcod <- as.data.frame(pcod) # ensure not a tibble
@@
   data = pcod, mesh = mesh,
+  offset = log(pcod$effort),
   family = delta_gamma(),
@@
   years <- years + 0.05
-  points(years, sdmTMB_est$b_j[,sdmTMB_pars])
-  segments(years, sdmTMB_est$b_j[,sdmTMB_pars] - 2 * sdmTMB_sd$b_j[,sdmTMB_pars], 
-    years, sdmTMB_est$b_j[,sdmTMB_pars] + 2 * sdmTMB_sd$b_j[,sdmTMB_pars],
+  if (sdmTMB_pars == 1) {
+    points(years, sdmTMB_est$b_j)
+    segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
+    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
     col = "red")
+  } else {
+    points(years, sdmTMB_est$b_j2)
+       segments(years, sdmTMB_est$b_j2 - 2 * sdmTMB_sd$b_j2, 
+    years, sdmTMB_est$b_j2 + 2 * sdmTMB_sd$b_j2,
+    col = "red")
+  }
   legend("topright", legend = c("VAST", "sdmTMB"), 

### 2022-06-06 doc/paper.Rmd
Double space; update appendices for latest sdmTMB
Update abstract

--- 2022-05-03 doc/paper.Rmd
@@
 header-includes:
-  \usepackage{setspace}\onehalfspacing
+  \usepackage{setspace}\doublespacing
   \usepackage[left]{lineno}
@@
   \linenumbers
+  \setlength{\parskip}{1.0\baselineskip}
   \newcommand{\beginsupplement}{
@@
     | \noindent 1.
-Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
-However, appropriate models to analyze these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
+Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
     | 2.
-Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB) but can be extended to penalized likelihood or Bayesian inference.
-We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) approach.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.
+We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
     | 3.
-We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., delta/hurdle models, penalized smoothers and break-point effects, anisotropy, abundance index standardization).
-    | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
+The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).
+Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
+Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
+    | 4.
+sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.
+We anticipate its application to the full spectrum of novel ecological problems that involve geostatistical data will prove equally fruitful.
     | Keywords:

### 2022-06-06 doc/paper.Rmd
user-friendly -> accessible

@@
 ---
-title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
+title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
 author: "|

### 2022-07-21 doc/paper.Rmd
Edit?

@@
     | 2.
-Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) approach.
+Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.

### 2022-07-21 doc/paper-peerj.Rmd
Add peerj version

--- 2022-07-21 doc/paper.Rmd
@@
 title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
-author: "|
-  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
-  Eric J. Ward$^3$,
-  Philina A. English$^1$,
-  Lewis A. K. Barnett$^4$\n|
-  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
-  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
-  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
-  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
-    Running headline: sdmTMB: spatial and spatiotemporal GLMMs\n"
+preprint: false
+author:
+  - name: Sean C. Anderson
+    affiliation: 1
+    corresponding: true
+    email: sean.anderson@dfo-mpo.gc.ca
+  - name: Eric J. Ward
+    affiliation: 2
+  - name: Philina A. English
+    affiliation: 1
+  - name: Lewis A. K. Barnett
+    affiliation: 3
+affiliation:
+  - code: 1
+    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
+  - code: 2
+    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
+  - code: 3
+    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
+abstract: >
+  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
+  Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.
+We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
+  The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).
+Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
+Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
+  sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.
+We anticipate its application to the full spectrum of novel ecological problems that involve geostatistical data will prove equally fruitful.
 output:
-    # bookdown::word_document2:
-    bookdown::pdf_document2:
-      toc: false
-      number_sections: false
-      fig_caption: true
-      highlight: "monochrome"
+  bookdown::pdf_book:
+    template: template-peerj.tex
     keep_tex: true
-csl: mee.csl
+    pandoc_args: "--highlight=haddock"
 bibliography: refs.bib
 link-citations: yes
+csl: peerj.csl
 header-includes:
-  \usepackage{setspace}\doublespacing
-  \usepackage[left]{lineno}
   \usepackage{bm}
@@
   \usepackage{tablefootnote}
-  \linenumbers
-  \setlength{\parskip}{1.0\baselineskip}
-  \newcommand{\beginsupplement}{
-  \setcounter{equation}{0}
-  \renewcommand{\theequation}{S.\arabic{equation}}
-  \setcounter{table}{0}
-  \renewcommand{\thetable}{S\arabic{table}}
-  \setcounter{figure}{0}
-  \renewcommand{\thefigure}{S\arabic{figure}}}
-  \usepackage{lscape}
-  \widowpenalty10000
-  \clubpenalty10000
-abstract: |
-    | \noindent 1.
-Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
-    | 2.
-Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.
-We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
-    | 3.
-The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).
-Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
-Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
-    | 4.
-sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.
-We anticipate its application to the full spectrum of novel ecological problems that involve geostatistical data will prove equally fruitful.
-    | Keywords:
-      Gaussian Markov random fields (GMRF),
-      generalized linear mixed effects models (GLMM),
-      INLA,
-      SPDE,
-      species distribution modelling,
-      R package,
-      spatio-temporal or spatial-temporal,
-      Template Model Builder
 ---
@@
 ```
+
 <!-- sdmTMB: spatial and spatiotemporal random field generalized linear mixed-effects models in R in a fast, flexible, and user-friendly interface -->
 
-# Introduction
+<!-- File naming format: 'Supplemental [Item] S[number]' e.g. 'Supplemental Data S1'. -->
+<!-- Use the following style when citing Supplemental files in the text: -->
+<!-- Fig. S1, Table S1, Data S1, Video S1, Article S1, Audio S1. -->
+
+<!-- Introduction -->
+<!-- Materials & Methods -->
+<!-- Results -->
+<!-- Discussion -->
+<!-- Conclusions -->
+<!-- Acknowledgments -->
+
+# Introduction {-}
 
@@
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself, for example, by identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
 <!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?.
@@
 
-# Model description
+# Model description {-}
 
@@
 The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.
-\@ref(fig:diagram), Appendix 1).
+\@ref(fig:diagram), Article S1).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
@@
 The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
+The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays at the same rate in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).
 
-(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
+(ref:diagram-fig) **Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.**
 An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See Appendix 1 for a full description of the model.
+See Article S1 for a full description of the model.
 Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
@@
 
-By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix 1).
+By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Article S1).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Article S1).
 The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.
+After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution with Stan [@carpenter2017] for Bayesian inference.
 
 sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
-These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, to impart prior information, to or achieve regularization.
 Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
@@
 
-# Introductory example using fish survey data
+# Introductory example using fish survey data {-}
 
@@
 
+\noindent
 Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
@@
 <!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
+\noindent
 We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
@@
 fit_spatiotemporal <- sdmTMB(
-  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
-  time = "year", spatial = "on", spatiotemporal = "ar1"
+  density ~ s(depth),
+  family = tweedie(link = "log"),
+  data = pcod,
+  mesh = mesh,
+  time = "year",
+  spatial = "on",
+  spatiotemporal = "ar1"
 )
@@
 We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
-\@ref(fig:pcod-fig)E, Appendix 2).
+\@ref(fig:pcod-fig)E, Article S2).
 
-(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
+(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
 (A) SPDE mesh (lines) combined with the trawl survey observations (points).
@@
 
-# Example of a spatially varying coefficient with citizen science data
+# Example of a spatially varying coefficient with citizen science data  {-}
 <!-- # Estimate a spatially varying effect from citizen science data -->
@@
   spatial_varying = ~ 0 + nao,
-  family = nbinom2(link = "log"), data = snow, mesh = mesh,
-  time = "year", spatial = "on", spatiotemporal = "iid"
+  family = nbinom2(link = "log"),
+  data = snow,
+  mesh = mesh,
+  time = "year",
+  spatial = "on",
+  spatiotemporal = "iid"
 )
@@
 
-We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix 3).
-This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
+We found a weak average positive effect of the annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Article S3).
+This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
 
-(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
+(ref:owl-nao-fig) **Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.**
 Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
@@
 
-# Model validation and selection
+# Model validation and selection  {-}
 
@@
 
-# Package comparisons
+# Package comparisons  {-}
 
-There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
+There are several R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
 sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
@@
 There is considerable variability in the available observation likelihoods across packages (Table 1).
-We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix 4) and VAST (Appendix 5).
+We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Article S4) and VAST (Article S5).
 
@@
 
-knitr::kable(d, format = "latex", caption = "Comparison of functionality between several R packages that can fit geostatistical GLMMs.",
+knitr::kable(d, format = "latex", caption = "\\textbf{Comparison of functionality between several R packages that can fit geostatistical GLMMs.}",
   booktabs = TRUE,
@@
 We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
-\@ref(fig:timing), Appendix 6).
+\@ref(fig:timing), Article S6).
 In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
@@
 
-(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
+(ref:timing-fig) **Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.**
 Lines represent means and ribbons 95% quantiles across 50 random iterations.
@@
 
-# Discussion
+# Discussion {-}
 
@@
 VAST is the sole option for fitting some multivariate models;
-alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
+alternatively, because VAST focuses on multivariate delta models and fisheries applications (Article S5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
@@
 However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by INLA/inlabru is likely to be considerably faster than passing the same model from sdmTMB/VAST to tmbstan for full Bayesian inference.
-Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores.
+<!-- Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores. -->
 
@@
 
-# Acknowledgements
+# Acknowledgements {-}
 
@@
 
-# Conflict of Interest statement
+<!-- A development version of sdmTMB is available at <https://github.com/pbs-assess/sdmTMB>. -->
+<!-- sdmTMB will be available on CRAN before publication. -->
+<!-- Paper source code is available at <https://github.com/seananderson/sdmTMB-paper> and will be archived on Zenodo before publication. -->
 
-The authors declare no conflict of interest.
+<!-- # Author Contributions -->
 
-# Author Contributions
+<!-- SA led development of the software and EW contributed key software code. -->
+<!-- All authors contributed software code, guided software development, and -->
+<!-- participated in testing. -->
+<!-- PE led the Snowy Owl case study. -->
+<!-- All authors contributed to appendix code. -->
+<!-- SA led the writing of the manuscript; all authors contributed critically to the drafts and gave final approval for publication. -->
 
-SA led development of the software and EW contributed key software code.
-All authors contributed software code, guided software development, and
-participated in testing.
-PE led the Snowy Owl case study.
-All authors contributed to appendix code.
-SA led the writing of the manuscript; all authors contributed critically to the drafts and gave final approval for publication.
+# Supporting Information {-}
 
-# Supporting Information
-
-Appendix 1: sdmTMB statistical model description\
-Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB\
-Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB\
-Appendix 4: Species distribution model comparison between INLA and sdmTMB\
-Appendix 5: Comparison of index standardization of survey data with VAST and sdmTMB\
-Appendix 6: sdmTMB speed testing methods and model validation description
+Article S1: sdmTMB statistical model description\
+Article S2: Example of index standardization of fishery-independent survey data with sdmTMB\
+Article S3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB\
+Article S4: Species distribution model comparison between INLA and sdmTMB\
+Article S5: Comparison of index standardization of survey data with VAST and sdmTMB\
+Article S6: sdmTMB speed testing methods and model validation description
 
@@
 
-# Data Availability
+<!-- # Data Availability -->
 
-A development version of sdmTMB is available at <https://github.com/pbs-assess/sdmTMB>.
-sdmTMB will be available on CRAN before publication.
-Paper source code is available at <https://github.com/seananderson/sdmTMB-paper> and will be archived on Zenodo before publication.
+<!-- A development version of sdmTMB is available at <https://github.com/pbs-assess/sdmTMB>. -->
+<!-- sdmTMB will be available on CRAN before publication. -->
+<!-- Paper source code is available at <https://github.com/seananderson/sdmTMB-paper> and will be archived on Zenodo before publication. -->
 
@@
 
-# References
+# References {-}
 

### 2022-07-21 doc/appendix-pcod-VAST.Rmd
Trash old delta text

@@
 We can compare the index we would get using sdmTMB.
-Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models.
-Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws.
-This should be approximately comparable with the bias-corrected index calculated above with VAST.
 

### 2022-07-21 doc/paper-peerj.Rmd
Floating

@@
   sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.
-We anticipate its application to the full spectrum of novel ecological problems that involve geostatistical data will prove equally fruitful.
+We anticipate its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
 output:
@@
 
+(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
+(A) SPDE mesh (lines) combined with the trawl survey observations (points).
+Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
+The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
+(B) Spatial random field:
+these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
+(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
+Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
+(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
+(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
+<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->
+
+```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", out.width="5.2in", fig.align='center', fig.pos="htb"}
+pcodfig <- here::here("figs", "pcod-fig.pdf")
+knitr::include_graphics(pcodfig)
+```
+
 We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
@@
 
-(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
-(A) SPDE mesh (lines) combined with the trawl survey observations (points).
-Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
-(B) Spatial random field:
-these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
-(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
-Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
-(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
-(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
-<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->
-
-```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", out.width="5.2in", fig.align='center'}
-pcodfig <- here::here("figs", "pcod-fig.pdf")
-knitr::include_graphics(pcodfig)
-```
-
 # Example of a spatially varying coefficient with citizen science data  {-}
@@
 
-```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="4.1in", fig.align='center'}
+```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="4.1in", fig.align='center', fig.pos="hb"}
 owl <- here::here("figs", "owl-nao-effect.png")
@@
 We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Article S4) and VAST (Article S5).
+
+(ref:timing-fig) **Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.**
+Lines represent means and ribbons 95% quantiles across 50 random iterations.
+Note the log x and y axes.
+<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
+VAST is similar to sdmTMB and so is not shown.
+inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
+mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
+<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
+Note that spaMM only fits spatial, not spatiotemporal, models.
+All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
+<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
+
+```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center', fig.pos="htb"}
+f <- here::here("figs", "timing-spatial-m1-april11.pdf")
+knitr::include_graphics(f)
+```
 
@@
 
-(ref:timing-fig) **Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.**
-Lines represent means and ribbons 95% quantiles across 50 random iterations.
-Note the log x and y axes.
-<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
-VAST is similar to sdmTMB and so is not shown.
-inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
-mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
-<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
-Note that spaMM only fits spatial, not spatiotemporal, models.
-All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
-<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->
-
-```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
-f <- here::here("figs", "timing-spatial-m1-april11.pdf")
-knitr::include_graphics(f)
-```
+<!-- \clearpage -->
 

### 2022-07-21 doc/paper-peerj.Rmd
Edit abstract

@@
   Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to.
+Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
   Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.
-We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
+We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
   The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).

### 2022-07-22 doc/paper-peerj.Rmd
couple word changes

@@
 This can greatly improve computational efficiency [e.g., @heaton2019].
-The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the R-INLA package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
 Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].
@@
 Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
-Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some functionality.
+Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack key functionality.
 We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

### 2022-08-03 doc/sdmTMB-paper.Rmd
Rename/format

--- 2022-07-22 doc/paper-peerj.Rmd
@@
 ---
-title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
-preprint: false
-author:
-  - name: Sean C. Anderson
-    affiliation: 1
-    corresponding: true
-    email: sean.anderson@dfo-mpo.gc.ca
-  - name: Eric J. Ward
-    affiliation: 2
-  - name: Philina A. English
-    affiliation: 1
-  - name: Lewis A. K. Barnett
-    affiliation: 3
-affiliation:
-  - code: 1
-    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
-  - code: 2
-    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
-  - code: 3
-    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
-abstract: >
-  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to.
-Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
-  Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.
-We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
-  The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).
-Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
-Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
-  sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.
-We anticipate its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
+title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
+author: "|
+  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
+  Eric J. Ward$^3$,
+  Philina A. English$^1$,
+  Lewis A. K. Barnett$^4$\n|
+  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
+  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
+  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
+      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
+  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
+      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n"
 output:
-  bookdown::pdf_book:
-    template: template-peerj.tex
+    bookdown::word_document2:
+    # bookdown::pdf_document2:
+      toc: false
+      number_sections: false
+      fig_caption: true
+      highlight: "monochrome"
       keep_tex: true
-    pandoc_args: "--highlight=haddock"
 bibliography: refs.bib
 link-citations: yes
-csl: peerj.csl
+csl: ecography.csl
 header-includes:
+  \usepackage[document]{ragged2e}
+  \usepackage{setspace}\doublespacing
+  \usepackage[left]{lineno}
   \usepackage{bm}
@@
   \usepackage{tablefootnote}
+  \setlength{\parskip}{1.0\baselineskip}
+  \newcommand{\beginsupplement}{
+  \setcounter{equation}{0}
+  \renewcommand{\theequation}{S.\arabic{equation}}
+  \setcounter{table}{0}
+  \renewcommand{\thetable}{S\arabic{table}}
+  \setcounter{figure}{0}
+  \renewcommand{\thefigure}{S\arabic{figure}}}
+  \usepackage{lscape}
+  \widowpenalty10000
+  \clubpenalty10000
 ---
-
-<!-- Point 1: set the context for and purpose of the work; -->
-<!-- Point 2: indicate the approach and methods; -->
-<!-- Point 3: outline the main results; -->
-<!-- Point 4: identify the conclusions and the wider implications. -->
-
-<!-- TODO: emphasize the niche carved out: a fast, flexible, easy-to-use, predictive-process univariate spatial or spatiotemporal GLMM, scales from maximum likelihood to penalized likelihood to full Bayesian inference-->
 
@@
 
-<!-- File naming format: 'Supplemental [Item] S[number]' e.g. 'Supplemental Data S1'. -->
-<!-- Use the following style when citing Supplemental files in the text: -->
-<!-- Fig. S1, Table S1, Data S1, Video S1, Article S1, Audio S1. -->
+\clearpage
 
-<!-- Introduction -->
-<!-- Materials & Methods -->
-<!-- Results -->
-<!-- Discussion -->
-<!-- Conclusions -->
-<!-- Acknowledgments -->
+\linenumbers
 
-# Introduction {-}
+# sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields
+
+# Abstract
+
+Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. 
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. 
+Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
+Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. 
+We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).
+Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
+Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
+sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. 
+We anticipate that its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
+
+Keywords:
+Gaussian Markov random fields (GMRF),
+generalized linear mixed effects models (GLMM),
+INLA,
+SPDE,
+species distribution modelling,
+R package,
+spatio-temporal or spatial-temporal,
+Template Model Builder
+
+\clearpage
+
+# Background {-}
 
@@
 The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.
-\@ref(fig:diagram), Article S1).
+\@ref(fig:diagram), Appendix S1).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
 Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
-
-<!-- Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time, -->
-<!-- the general form of a GLMM is -->
-<!-- $g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$, -->
-<!-- with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
--->
-<!-- We can partition the random effects $\bm{Z \psi}$ into subcomponents, each of which can optionally be included in the model (Appendix TODO). -->
-<!-- Although, in practice it is rare to include all of these components, the full sdmTMB model is -->
-<!-- $$ -->
-<!-- g(\mu_{\bm{s},t}) = -->
-<!-- \bm{X}_{1,\bm{s},t} \bm{\beta} + -->
-<!-- \bm{\alpha}_g + -->
-<!-- \bm{X}_{2,\bm{s},t} \gamma_t + -->
-<!-- \bm{X}_{3,t} \zeta_{\bm{s}} + -->
-<!-- \omega_{\bm{s}} + -->
-<!-- \epsilon_{\bm{s},t}, -->
-<!-- $$ -->
-<!-- where -->
-
-<!-- * $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices; -->
-<!-- * $\bm{\beta}$ represents a vector of fixed-effect coefficients; -->
-<!-- * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; -->
-<!-- * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$; -->
-<!-- * $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$; -->
-<!-- * $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and -->
-<!-- * $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field). For the IID case, $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$ (Appendix TODO). -->
 
@@
 
-(ref:diagram-fig) **Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.**
+(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
 An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See Article S1 for a full description of the model.
+See Appendix S1 for a full description of the model.
 Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.
@@
 
-By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Article S1).
+By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix S1).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Article S1).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix S1).
 The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
@@
 
-# Introductory example using fish survey data {-}
+# Example using fish survey data {-}
 
@@
 It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
-<!-- TODO: Since this is not particular to your package, maybe the e.g.
-for how to do this should be moved to the appendix?
-I do explain the transformation for the owl data and added a sentence to the pcod appendix, but maybe still important to include here as well?
--->
 Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
@@
 ```
+
+\onehalfspacing
 
@@
 
+\doublespacing
+
 We start by creating a mesh object that contains matrices to apply the SPDE approach.
@@
 
-(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
+(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
 (A) SPDE mesh (lines) combined with the trawl survey observations (points).
@@
 (E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
-<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->
 
@@
 
+\onehalfspacing
+
 ```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
@@
 
+\doublespacing
+
 <!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
@@
 We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
-<!-- (omitted for space). -->
+
+\onehalfspacing
 
@@
 
+\doublespacing
+
 The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
@@
 
+\onehalfspacing
+
 ```{r fit-tv, eval=FALSE, echo=TRUE}
@@
 
+\doublespacing
+
 We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
-\@ref(fig:pcod-fig)E, Article S2).
+\@ref(fig:pcod-fig)E, Appendix S2).
 
 # Example of a spatially varying coefficient with citizen science data  {-}
-<!-- # Estimate a spatially varying effect from citizen science data -->
 
@@
 We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+
+\onehalfspacing
 
@@
 
-We found a weak average positive effect of the annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Article S3).
+\doublespacing
+
+We found a weak average positive effect of the annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix S3).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
 
-(ref:owl-nao-fig) **Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.**
+(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
 Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
@@
 There is considerable variability in the available observation likelihoods across packages (Table 1).
-We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Article S4) and VAST (Article S5).
+We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix S4) and VAST (Appendix S5).
 
-(ref:timing-fig) **Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.**
+(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
 Lines represent means and ribbons 95% quantiles across 50 random iterations.
@@
 We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
-\@ref(fig:timing), Article S6).
+\@ref(fig:timing), Appendix S6).
 In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
@@
 VAST is the sole option for fitting some multivariate models;
-alternatively, because VAST focuses on multivariate delta models and fisheries applications (Article S5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
+alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix S5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
@@
 However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by INLA/inlabru is likely to be considerably faster than passing the same model from sdmTMB/VAST to tmbstan for full Bayesian inference.
-<!-- Furthermore, although both INLA/inlabru and sdmTMB can use parallel processing, the PARDISO library [@pardiso-7.2c] within INLA/inlabru allows memory to be shared across cores. -->
-
-<!-- However, with the SPDE, mgcv is typically slower than sdmTMB for similar models without optimized matrix libraries and lacks certain  -->
-
-<!--
-There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2019]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1).
-VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.
-Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists.
-For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
-From a fisheries perspective, sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
-Unlike VAST, however, sdmTMB does not currently extend to multivariate applications or support delta/hurdle models.
-
-inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
-For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig.
-\@ref(fig:timing), @osgood-zimmerman2021].
-One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
-Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
-On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.
-
-mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields [@miller2019]; however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
-First, the smoothers typically used with mgcv (e.g., thin plate splines), while capable of making equivalent predictions [@miller2019; @pedersen2019], do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance estimates).
-Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
-Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
-Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.
--->
-
-<!-- MVT? time-varying epsilon, non-gaussian fields -->
-<!-- 3d models, continuous time, -->
 
@@
 The included TMB `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
-<!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
-<!-- sdmTMB does not currently include  -->
-<!-- We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately. -->
-<!-- We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend. -->
-<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
-<!-- Line 275: I would suggest adding a MAJOR benefit of sdmTMB for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates!
-This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it.
-It saves me a bunch of time.
-If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it.
--->
 
@@
 We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.
+
+<!-- Significance statement  -->
+<!-- Authors must provide a statement outlining why Ecography is the best outlet for publication.
+This should include a description of why the submitted work makes a significant contribution to and is of general interest for understanding spatial and temporal patterns in ecology.
+You must provide a clear statement of how the manuscript builds on both recent work by yourself or your coauthors cited in this manuscript (published during the past 12 months, or in press, or submitted) and other published work.
+-->
+
+<!-- Data archiving statement  -->
+<!-- For articles published in Ecography, it is required that authors deposit data supporting their accepted papers in public archives of their choice (see section on Data sharing and repositories below).
+Authors must confirm that they deposit their data in a public repository and indicate the repository of their choice.
+-->
+
+<!-- Conflict of interest statement  -->
+<!-- Ecography requires that all authors disclose any potential sources of conflict of interest.
+Any interest or relationship, financial or otherwise, that might be perceived as influencing an author’s objectivity is considered a potential source of conflict of interest.
+For more information, please read our ethics and policies page.
+-->
+
+<!-- Ethics statement  -->
+<!-- Please provide an ethics statement with relevant fieldwork details (approvals, licences, permissions) to be included in the published article.
+The details of any museum specimens used (e.g.
+the specimen numbers and the institutions holding these) must be provided either in the manuscript or the supplementary files.
+For more information about preparing this section please visit our ethics and policies page.
+-->
+
+<!-- Funding statement  -->
+<!-- If applicable, please provide a ‘funding statement’ that can be included in the published article.
+List the sources of funding for each author.
+Here you may also list service providers that did provide facilities or equipment.
+-->
+
+<!-- Acknowledgements  -->
+<!-- Those who contributed to the work but do not meet our authorship criteria may be listed in the Acknowledgements.
+Authors are responsible for ensuring that anyone named in the acknowledgements agrees to be named.
+-->
 
@@
 
-Article S1: sdmTMB statistical model description\
-Article S2: Example of index standardization of fishery-independent survey data with sdmTMB\
-Article S3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB\
-Article S4: Species distribution model comparison between INLA and sdmTMB\
-Article S5: Comparison of index standardization of survey data with VAST and sdmTMB\
-Article S6: sdmTMB speed testing methods and model validation description
+Appendix S1: sdmTMB statistical model description\
+Appendix S2: Example of index standardization of fishery-independent survey data with sdmTMB\
+Appendix S3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB\
+Appendix S4: Species distribution model comparison between INLA and sdmTMB\
+Appendix S5: Comparison of index standardization of survey data with VAST and sdmTMB\
+Appendix S6: sdmTMB speed testing methods and model validation description
 
@@
 
-# References {-}
+# References
+
+\vspace{8mm}
 

### 2022-08-03 doc/appendix-pcod-VAST-poisson-link.Rmd
Fixes to unused appendix

--- 2022-04-07 doc/appendix-pcod-VAST.Rmd
@@
   RhoConfig = RhoConfig,
-  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
+  ObsModel = c(2, 1), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
   bias.correct = FALSE,
@@
  # create folder for saved output:
-dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)
+dir.create(paste0(here("doc", "appendix-VAST-pg")), showWarnings = FALSE)
 
@@
 pcod$effort <- 1
+if (FALSE) { # to check offset/effort
+  set.seed(1)
+  pcod$effort <- exp(rnorm(nrow(pcod), mean = 0, sd = 0.2))
+}
 pcod <- as.data.frame(pcod) # ensure not a tibble
-f <- here("doc", "appendix-VAST", "vast-cache.rds")
+f <- here("doc", "appendix-VAST-pg", "vast-cache.rds")
 if (!file.exists(f)) {
@@
     input_grid = input_grid,
-    working_dir = paste0(here("doc", "appendix-VAST"), "/")
+    working_dir = paste0(here("doc", "appendix-VAST-pg"), "/")
   )
@@
   data = pcod, mesh = mesh,
-  family = delta_gamma(),
+  offset = log(pcod$effort),
+  family = delta_poisson_link_gamma(),
   time = "year", silent = FALSE,
@@
   check_residuals = FALSE,
-  working_dir = paste0(here("doc", "appendix-VAST"), "/")
+  working_dir = paste0(here("doc", "appendix-VAST-pg"), "/")
 )
@@
 ```{r plot-knots, echo = TRUE, out.width="6in"}
-knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
+knitr::include_graphics(here("doc", "appendix-VAST-pg", "Data_and_knots.png"))
 ```
@@
 ```{r plot-density, echo = TRUE, out.width="6in"}
-knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
+knitr::include_graphics(here("doc", "appendix-VAST-pg", "ln_density-predicted.png"))
 ```
@@
 ```{r plot-index, echo = TRUE, out.width="3in"}
-if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
-  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
+if (file.exists(here("doc", "appendix-VAST-pg", "Index-Biomass.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST-pg", "Index-Biomass.png"))
 }
-if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
-  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
+if (file.exists(here("doc", "appendix-VAST-pg", "Index.png"))) {
+  knitr::include_graphics(here("doc", "appendix-VAST-pg", "Index.png"))
 }
@@
 library(ggplot2)
-vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
+vast_i <- read.csv(here("doc/appendix-VAST-pg/Index.csv")) %>%
   mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 

### 2022-08-08 doc/make.R
Rename file

@@ -1,5 +1,5 @@
 files <- c(
-  "paper",
+  "sdmTMB-paper",
   "appendix-model",
   "appendix-pcod-sdmTMB",
   "appendix-snow-sdmTMB",

### 2022-08-08 doc/sdmTMB-paper.Rmd
Copy editing

@@
 
+\setcounter{page}{1}
+
 # sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields
@@
 Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
-Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) based approach.
 Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. 
 We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
-The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models).
+The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru).
 Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
@@
 sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. 
-We anticipate that its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
+We anticipate that its application to the full spectrum of problems involving geostatistical data will prove equally fruitful.
 
@@
 Gaussian Markov random fields (GMRF),
-generalized linear mixed effects models (GLMM),
-INLA,
 SPDE,
@@
 An alternative would be the `delta_gamma()` family to specify a delta/hurdle model [@aitchison1955] or to model catch weight with an offset for log(area swept).
+We could allow for spatial anisotropy by adding `anisotropy = TRUE` [@lindgren2011; @thorson2015ices].
 <!-- We could instead model catch weight as the response with an offset for log(area). -->
@@
 
-We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
+We could extend the spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
 Here we will keep the spatial field and structure the spatiotemporal random fields as AR(1).
@@
 
-<!-- Significance statement  -->
-<!-- Authors must provide a statement outlining why Ecography is the best outlet for publication.
-This should include a description of why the submitted work makes a significant contribution to and is of general interest for understanding spatial and temporal patterns in ecology.
-You must provide a clear statement of how the manuscript builds on both recent work by yourself or your coauthors cited in this manuscript (published during the past 12 months, or in press, or submitted) and other published work.
--->
-
-<!-- Data archiving statement  -->
-<!-- For articles published in Ecography, it is required that authors deposit data supporting their accepted papers in public archives of their choice (see section on Data sharing and repositories below).
-Authors must confirm that they deposit their data in a public repository and indicate the repository of their choice.
--->
-
-<!-- Conflict of interest statement  -->
-<!-- Ecography requires that all authors disclose any potential sources of conflict of interest.
-Any interest or relationship, financial or otherwise, that might be perceived as influencing an author’s objectivity is considered a potential source of conflict of interest.
-For more information, please read our ethics and policies page.
--->
-
-<!-- Ethics statement  -->
-<!-- Please provide an ethics statement with relevant fieldwork details (approvals, licences, permissions) to be included in the published article.
-The details of any museum specimens used (e.g.
-the specimen numbers and the institutions holding these) must be provided either in the manuscript or the supplementary files.
-For more information about preparing this section please visit our ethics and policies page.
--->
-
-<!-- Funding statement  -->
-<!-- If applicable, please provide a ‘funding statement’ that can be included in the published article.
-List the sources of funding for each author.
-Here you may also list service providers that did provide facilities or equipment.
--->
-
-<!-- Acknowledgements  -->
-<!-- Those who contributed to the work but do not meet our authorship criteria may be listed in the Acknowledgements.
-Authors are responsible for ensuring that anyone named in the acknowledgements agrees to be named.
--->
-
 # Acknowledgements {-}
@@
 We thank the authors of all these packages.
-We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
+
+*Individual acknowledgements omitted from this submitted main text file*
+
+<!-- sdmTMB would not be possible without the TMB [@kristensen2016] and INLA [@rue2009; @lindgren2011; @lindgren2015] R packages. -->
+<!-- sdmTMB is heavily inspired by and in some places code has been adapted from both the VAST [@thorson2019] and glmmTMB [@brooks2017] R packages. -->
+<!-- Penalized spline support was possible thanks to mgcv [@wood2017a]. -->
+<!-- We thank the authors of all these packages. -->
+<!-- We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript. -->
 

### 2022-08-08 doc/sdmTMB-PeerJ.Rmd
Reformat

--- 2022-08-08 doc/sdmTMB-paper.Rmd
@@
 title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
-author: "|
-  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
-  Eric J. Ward$^3$,
-  Philina A. English$^1$,
-  Lewis A. K. Barnett$^4$\n|
-  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
-  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
-  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
-  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
-      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n"
+preprint: false
+author:
+  - name: Sean C. Anderson
+    affiliation: 1
+    corresponding: true
+    email: sean.anderson@dfo-mpo.gc.ca
+  - name: Eric J. Ward
+    affiliation: 2
+  - name: Philina A. English
+    affiliation: 1
+  - name: Lewis A. K. Barnett
+    affiliation: 3
+affiliation:
+  - code: 1
+    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
+  - code: 2
+    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
+  - code: 3
+    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
 output:
-    bookdown::word_document2:
-    # bookdown::pdf_document2:
+    bookdown::pdf_book:
+      template: template-peerj.tex
+      pandoc_args: "--highlight=haddock"
       toc: false
       number_sections: false
-      fig_caption: true
-      highlight: "monochrome"
       keep_tex: true
+abstract: >
+  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement.
+Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
+Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) based approach.
+Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.
+We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
+The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru).
+Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
+Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
+sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.
+We anticipate that its application to the full spectrum of problems involving geostatistical data will prove equally fruitful.
 bibliography: refs.bib
 link-citations: yes
-csl: ecography.csl
+csl: peerj.csl
 header-includes:
-  \usepackage[document]{ragged2e}
-  \usepackage{setspace}\doublespacing
-  \usepackage[left]{lineno}
   \usepackage{bm}
@@
   \usepackage{tablefootnote}
-  \setlength{\parskip}{1.0\baselineskip}
-  \newcommand{\beginsupplement}{
-  \setcounter{equation}{0}
-  \renewcommand{\theequation}{S.\arabic{equation}}
-  \setcounter{table}{0}
-  \renewcommand{\thetable}{S\arabic{table}}
-  \setcounter{figure}{0}
-  \renewcommand{\thefigure}{S\arabic{figure}}}
-  \usepackage{lscape}
   \widowpenalty10000
@@
 
-\clearpage
+<!-- \clearpage -->
 
-\linenumbers
+<!-- \linenumbers -->
 
-\setcounter{page}{1}
+<!-- Keywords: -->
+<!-- Gaussian Markov random fields (GMRF), -->
+<!-- SPDE, -->
+<!-- species distribution modelling, -->
+<!-- R package, -->
+<!-- spatio-temporal or spatial-temporal, -->
+<!-- Template Model Builder -->
 
-# sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields
-
-# Abstract
-
-Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. 
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. 
-Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
-Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) based approach.
-Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. 
-We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
-The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru).
-Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.
-Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
-sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. 
-We anticipate that its application to the full spectrum of problems involving geostatistical data will prove equally fruitful.
-
-Keywords:
-Gaussian Markov random fields (GMRF),
-SPDE,
-species distribution modelling,
-R package,
-spatio-temporal or spatial-temporal,
-Template Model Builder
-
-\clearpage
+<!-- \clearpage -->
 
@@
 
-\onehalfspacing
-
 ```{r, echo=TRUE, eval=FALSE}
@@
 ```
-
-\doublespacing
 
@@
 
-\onehalfspacing
-
 ```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
@@
 
-\doublespacing
-
 <!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
@@
 We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
-
-\onehalfspacing
 
@@
 
-\doublespacing
-
 The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
@@
 
-\onehalfspacing
-
 ```{r fit-tv, eval=FALSE, echo=TRUE}
@@
 
-\doublespacing
-
 We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
@@
 We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
-
-\onehalfspacing
 

### 2022-10-31 doc/paper-jss/paper-jss.Rmd
Start JSS

---
documentclass: jss
author:
  - name: FirstName LastName
    orcid: 0000-0000-0000-0000
    affiliation: University/Company
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{name@company.com}
    url: http://rstudio.com
  - name: Second Author
    orcid: 0000-0000-0000-0000
    affiliation: 'Affiliation \AND'
    # To add another line, use \AND at the end of the previous one as above
  - name: Third Author
    orcid: 0000-0000-0000-0000
    address: |
      | Department of Statistics and Mathematics,
      | Faculty of Biosciences,
      | Universitat Autònoma de Barcelona
    affiliation: |
      | Universitat Autònoma
      | de Barcelona
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across disciplines and available in increasingly large volumes. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less interpretable ways. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of `stats::glm()`, \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are also possible. We describe the statistical model, illustrate the package through two case studies, and compare the functionality, speed, and interface to related software. The case studies illustrate \pkg{sdmTMB}'s flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Modelling of these common but computationally complex relationships is further aided by built-in handling of delta (hurdle) models, break-point effects, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that \pkg{sdmTMB} is 4--16 times faster than \pkg{INLA} or \pkg{inlabru}. \pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling. We anticipate its application to the full spectrum geostatistical data will prove equally fruitful.
keywords:
  # at least one keyword must be supplied
  formatted: [test]
  plain:     [test]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data [@rue2005gmrf; @diggle2007; @cressie2011].
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle2007; @chiles1999].

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \pkg{INLA} to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as \pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as \pkg{lme4} [@bates2015] and \pkg{glmmTMB} [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as \pkg{VAST} [@thorson2019] and \pkg{inlabru} [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \pkg{INLA} to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application in the field of ecology is species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

# Model description

\pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates reflecting some underlying spatial process [@rossi1992; @diggle2007].
These data can be collected across discrete points in time.
Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models [e.g., @verhoef2018].
\pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted [@wall2004] (e.g., spatial variance and range---the distance at which points are effectively independent).

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix 1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix 1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

The \pkg{sdmTMB} model is fit by maximum marginal likelihood.
Internally, a \pkg{TMB} [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in \proglang{R} [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an \pkg{sdmTMB} model to the \proglang{R} package \pkg{tmbstan} [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to model parameters (`?\pkg{sdmTMB}priors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

The results from an \pkg{sdmTMB} model can be used to generate various quantities of interest.
For example, \pkg{sdmTMB} provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.

\begin{landscape}
\begin{figure}[p]
{\centering \includegraphics[width=9in]{../../figs/diagram}}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix 1 for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} [@wood2017a], \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

# Introductory example using fish survey data

An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

```{r, include=FALSE}
# simplify for paper
# pcod <- dplyr::select(sdmTMB::pcod, year, density, depth, X, Y)
```

```{r sdmTMB-lib, eval=TRUE}
library(sdmTMB)
```

```{r, echo=TRUE, eval=FALSE}
library(sdmTMB)
head(pcod)
```

```{}
#>    year density depth     X     Y
#>   <int>   <dbl> <dbl> <dbl> <dbl>
#> 1  2003   113.    201  446. 5793.
#> 2  2003    41.7   212  446. 5800.
#> 3  2003     0     220  449. 5802.
```

We start by creating a mesh object that contains matrices to apply the SPDE approach.

```{r, echo=TRUE, eval=TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).
<!-- We could instead model catch weight as the response with an offset for log(area). -->

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = "on"
)
```

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

```{r pcod-eg1-summary,eval=TRUE, echo=TRUE}
summary(fit)
```

The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} (REF) or \pkg{broom.mixed} (REF) packages:

```{r pcod-eg1-tidy-fe, eval=TRUE, echo=TRUE}
tidy(fit, conf.int = TRUE)
```

```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

And we can run some basic checks on our model with the `sanity()` function:

```{r pcod-eg1-sanity}
sanity(fit)
```

We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
Here, we will predict on a 2x2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.

```{r pcod-eg1-suppress-depth2, eval=TRUE, echo=FALSE}
qcs_grid$depth_scaled <- NULL # for simplicity in paper
qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
```

```{r pcod-eg1-predict, echo=TRUE, eval=TRUE, cache=TRUE}
p <- predict(fit, newdata = qcs_grid)
```

The output of `predict()` is a data frame containing overall estimates in link space, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig.\ \ref{fig:pcod-fig}B--D.

\begin{figure}
{\centering \includegraphics[width=5.2in]{../../figs/pcod-fig}}
\caption[]{Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada. 
(A) SPDE mesh (lines) combined with the trawl survey observations (points). Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. The circle area corresponds to biomass density of Pacific cod caught on individual trawls. 
(B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect. (C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (\texttt{get\_cog()}) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother. 
These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}). The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig}
\end{figure}

We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
Here we will keep the spatial field and structure the spatiotemporal random fields as AR(1).

```{r time1, cache=TRUE}
t1 <- Sys.time()
```

```{r fit-tv, eval=TRUE, echo=TRUE, cache=TRUE, results='hide'}
fit_spatiotemporal <- sdmTMB(
  density ~ 0 + s(depth, k = 5), 
  time_varying = ~ 1,
  family = tweedie(link = "log"), 
  data = pcod,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  time_varying_type = "ar1",
  anisotropy = TRUE,
  silent = FALSE # print fitting progress
)
```

```{r time2, cache=TRUE}
t2 <- Sys.time()
```

```{r fit-tv-print, eval=TRUE, echo=TRUE}
summary(fit_spatiotemporal)
```

This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop. 
We can inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix 2).

# Example of a spatially varying coefficient with citizen science data

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.

```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owls <- sdmTMB(
  count ~ nao + (1 | year_factor),
  spatial_varying = ~ 0 + nao,
  family = nbinom2(link = "log"), 
  data = snow, 
  mesh = mesh,
  time = "year", 
  spatial = "on", 
  spatiotemporal = "iid"
)
```

We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}, Appendix 3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

```{r owl-nao, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit.", out.width="4.1in", fig.align='center'}
owl <- here::here("figs", "owl-nao-effect.png")
knitr::include_graphics(owl)
```

# Model validation and selection

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
<!-- Simulations can also be passed to the \proglang{R} package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->

<!-- We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary \pkg{TMB} object (Fig. \@ref(fig:pcod-fig)E). `get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations. -->

```{r, eval=FALSE, echo=FALSE}
p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
ind <- get_index(p)
```

# Package comparisons

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, \pkg{VAST} is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table 1).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} or \pkg{inlabru} (Appendix 4) and \pkg{VAST} (Appendix 5).

```{r compare-table, results='asis', eval=FALSE}
library(kableExtra)
suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE, comment = "#")))
d <- as.matrix(d)
# d[d == "1"] <- "\\checkmark"
d <- gsub("^1", "\\\\checkmark", d)
# d[d == "1*"] <- "\\checkmark\\footnote{a}"
# d[d == "GAMs*"] <- "GAMs\\footnote{a}"
# d[d == "0"] <- "--"
d <- gsub("^0", "--", d)
# d[d == "0*"] <- "--\\footnote{a}"
d <- as.data.frame(d)
names(d)[1] <- ""
row.names(d) <- NULL
d <- d[,c(1, 2, 3, 5, 6, 4, 7, 8, 9)]
d$Hmsc <- NULL

d$INLA <- NULL
names(d)[names(d) == "inlabru"] <- "INLA/inlabru"

knitr::kable(d, format = "latex", caption = "Comparison of functionality between several R packages that can fit geostatistical GLMMs.",
  booktabs = TRUE,
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>%
  kableExtra::footnote(
    general = paste(
      "$^1$Technically possible but non-trivial.",
      "$^2$Penalized smoother GAMs that determine `wiggliness'.",
      "$^3$inlabru but not INLA.",
      "$^4$Spatiotemporal fields as random walk only.",
      "$^5$SPDE approach as in Miller et al. (2019).",
      "$^6$Does have predictive process knots.",
      "$^7$Zero-inflated NB2 only.",
      "$^8$Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
      "$^9$Possible as log-linked Poisson GLMM with aggregated data.",
      "$^{10}$Hurdle models possible by fitting components separately."
    ),
    escape = FALSE, threeparttable = TRUE
  )
```

\begin{table}

\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.}
\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \pkg{sdmTMB} & \pkg{VAST} & \pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta model & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item $^1$Technically possible but non-trivial. $^2$Penalized smoother GAMs that determine `wiggliness'. $^3$\pkg{inlabru} but not \pkg{INLA}. $^4$Spatiotemporal fields as random walk only. $^5$SPDE approach as in Miller et al. (2019). $^6$Does have predictive process knots. $^7$Zero-inflated NB2 only. $^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}. $^9$Possible as log-linked Poisson GLMM with aggregated data. $^{10}$Hurdle models possible by fitting components separately.
\end{tablenotes}
\end{threeparttable}
\end{table}

<!-- TODO: Line 242. Can \pkg{sdmTMB} be run in parallel? If so, does the parallelism share memory across the cores or does the memory grow with the number of cores? This is a HUGE feature of PARDISO within \pkg{INLA}/\pkg{inlabru} allowing for very large models to be fit on HPCs. It might be worth mentioning and adding to Table 1. Also Line 242: do you expect the same relative performance between the software packages as more cores are used? -->

We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

\begin{figure}
{\centering \includegraphics[width=3.5in]{../../figs/timing-spatial-m1-april11}}
\caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions. Lines represent means and ribbons 95% quantiles across 50 random iterations. Note the log x and y axes. \pkg{VAST} is similar to \pkg{sdmTMB} and so is not shown. \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation. \pkg{mgcv} used \texttt{bam()}, \texttt{method = 'fREML'}, and discretized covariates [@miller2019]. Note that spaMM only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}\label{fig:timing}
\end{figure}

# Discussion

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

Speed-wise, \pkg{sdmTMB} (and by association \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
Furthermore, although both \pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores.

<!-- However, with the SPDE, \pkg{mgcv} is typically slower than \pkg{sdmTMB} for similar models without optimized matrix libraries and lacks certain  -->

<!--
There are many \proglang{R} packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2019]; however, the three most closely related are \pkg{VAST}, \pkg{inlabru}, and \pkg{mgcv} (Table 1).
\pkg{VAST} is also built on \pkg{TMB} and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.
Interface-wise, \pkg{VAST} is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make \pkg{sdmTMB} more broadly usable by applied ecologists.
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
From a fisheries perspective, \pkg{sdmTMB} achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike \pkg{VAST}, however, \pkg{sdmTMB} does not currently extend to multivariate applications or support delta/hurdle models.

\pkg{inlabru} is similar to \pkg{sdmTMB} in design, but differs fundamentally in that fitting is achieved via \pkg{INLA} and approximate Bayesian computation.
For some models, TMB/\pkg{sdmTMB} can be considerably faster than \pkg{INLA}/\pkg{inlabru}, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig. \@ref(fig:timing), @osgood-zimmerman2021].
One major capability included in \pkg{sdmTMB} but not \pkg{INLA}/\pkg{inlabru} is anisotropy.
Furthermore, we think the \pkg{sdmTMB} interface will be more familiar than \pkg{inlabru} to users of `glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB}.
On the other hand, \pkg{inlabru} fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with \pkg{INLA}.

\pkg{mgcv} can fit many similar models to \pkg{sdmTMB} either with various flavours of splines or even with the SPDE approximation to random fields [@miller2019]; however, there are some notable advantages of \pkg{sdmTMB} for the specific class of models discussed here.
First, the smoothers typically used with \pkg{mgcv} (e.g., thin plate splines), while capable of making equivalent predictions [@miller2019; @pedersen2019], do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance estimates).
Second, the implementation of the Tweedie distribution in \pkg{mgcv} requires the power parameter to be fixed in a mixed effects model.
Third, \pkg{mgcv} does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
Despite these \pkg{sdmTMB} advantages for the models discussed here, \pkg{mgcv} is an immensely powerful modelling package that extends well beyond spatial random field models.
-->

<!-- MVT? time-varying epsilon, non-gaussian fields -->
<!-- 3d models, continuous time, -->

Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
The included \pkg{TMB} `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
<!-- Not sure what kind of refs you want here?  \pkg{INLA}'s non-gaussian random fields are mentioned here @bakka2018review-->
<!-- \pkg{sdmTMB} does not currently include  -->
<!-- We may introduce explicit delta models, but many such models can already be fit with \pkg{sdmTMB} by modelling the two components separately. -->
<!-- We may introduce capability for multivariate response data, but \pkg{VAST} already focuses on such models with a \pkg{TMB} backend. -->
<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
<!-- Line 275: I would suggest adding a MAJOR benefit of \pkg{sdmTMB} for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates! This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it. It saves me a bunch of time. If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it. -->

Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users.

# Acknowledgements

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
Penalized spline support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.

### 2022-11-02 doc/paper-jss/inla-comparison.Rmd
Preping JSS

```{r setup-inla-comparison, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  warning = FALSE,
  message = FALSE
)
```

\section{Comparison between INLA, inlabru, and sdmTMB}

In this example, we will fit a presence-absence dynamic species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
We will use a binomial distribution and a logit link.

We will start by loading necessary packages:

```{r packages-inla-comparison, message=FALSE, warning=FALSE, cache=FALSE, echo=TRUE}
library("ggplot2")
library("dplyr")
library("sdmTMB")
library("INLA")
library("inlabru")
theme_set(theme_light())
```

We will limit \pkg{INLA} to using a single thread to match how \pkg{TMB} is set up here. Both could optionally be run with parallel processing:

```{r inla-threads, echo=TRUE}
INLA::inla.setOption(num.threads = "1:1")
```

We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
\pkg{INLA} requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:

```{r pcod-inla-comparison-data, echo=TRUE}
d <- pcod %>% filter(year >= 2007)
yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
d <- dplyr::left_join(d, yr_lu, by = "year")
```

We will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in \pkg{INLA}, \pkg{inlabru}, and \pkg{sdmTMB}.
The values amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).

```{r set-priors, echo = TRUE, cache=FALSE}
range_min <- 5
sigma_max <- 5
prior_prob <- 0.05
```

In INLA, a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).

```{r, echo=TRUE}
h_spec <- list(rho = list(prior = "pccor1", param = c(0, 0.9)))
```

We will start by building identical meshes using both sdmTMB and INLA.
Here, X and Y are coordinates in UTM zone 9.
These same functions are used internally in the sdmTMB `make_mesh()` function.

```{r sdm-mesh, echo = TRUE, cache=FALSE}
sdmTMB_mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 10)
```

```{r inla-mesh, echo = TRUE, cache=FALSE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, # an smdTMB mesh could be inserted here with `sdmTMB_mesh$mesh`
  prior.range = c(range_min, prior_prob),
  prior.sigma = c(sigma_max, prior_prob)
)
```

Here, the meshes should be the same.

```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in",echo=FALSE, eval=FALSE}
plot(sdmTMB_mesh)
title(main = "sdmTMB mesh")
plot(spde$mesh, main = "", asp = 1)
title(main = "INLA mesh")
```

## INLA version

Data preparation for \pkg{INLA} models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB.
We use weak priors on the factor fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.

```{r inla, echo=TRUE}
tictoc::tic("INLA") # start timing
nyear <- length(unique(d$year))
iset <- inla.spde.make.index("i", n.spde = spde$n.spde, n.group = nyear)
A <- inla.spde.make.A(mesh = inla_mesh, loc = loc_xy, group = d$i_year)
sdat <- inla.stack(
  data = list(y = d$present),
  A = list(A, 1),
  effects = list(iset, year_factor = as.factor(d$year)),
  tag = "stdata"
)

formula <- y ~ 0 + year_factor + f(i,
  model = spde, group = i.group,
  control.group = list(model = "ar1", hyper = h_spec)
)

m_inla <- inla(formula,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(
    expand.factor.strategy = "inla",
    mean = 0, prec = 1 / (100 * 100)
  ),
  control.inla = list(int.strategy = "eb", strategy = "gaussian")
)
tictoc::toc() # stop timing
```

We could use `summary()` to inspect the model, but will omit the output to save space.

```{r, echo=FALSE, eval=FALSE}
summary(m_inla)
```

## inlabru version

There are a couple ways to include the factor fixed effects with `inlabru`---using factor variables in the standard equation syntax does not work.
One option is to create a design matrix and add year effects individually to the formula.
The other way to set up the factor fixed coefficients with inlabru is to use the `factor_full` (or `factor_contrast`) model (seen below within `fac()`).
This option causes the location of the factor fixed-effect parameter estimates to end up in `summary.random$fac`.

```{r, echo=TRUE}
dat <- as.data.frame(d)
sp::coordinates(dat) <- c("X", "Y")
dat <- as(dat, "SpatialPointsDataFrame")
```

```{r, echo=TRUE}
tictoc::tic("bru")
components <- present ~ 0 +
  f(
    main = coordinates, model = spde, group = i_year,
    control.group = list(model = "ar1", hyper = h_spec)
  ) +
  fac(main = year, model = "factor_full")

m_bru <- bru(
  components = components,
  family = "binomial",
  data = dat,
  options =
    list(
      bru_verbose = TRUE,
      control.fixed = list(
        expand.factor.strategy = "inla",
        mean = 0, prec = 1 / (100 * 100)
      ),
      control.family = list(link = "logit"),
      control.inla = list(int.strategy = "eb", strategy = "gaussian")
    )
)
tictoc::toc()
```

We will again suppress the output of `summary()` to save space.

```{r, echo=TRUE, eval=FALSE}
summary(m_bru)
```

```{r, echo=TRUE}
m_bru$summary.random$fac[, c("ID", "mean", "sd", "0.025quant", "0.975quant")]
```

## sdmTMB version

We will now fit a nearly identical model with `sdmTMB()` (the AR(1) correlation parameter prior is different).
We again use very weak priors on the factor fixed effects.

```{r sdmtmb, warning=FALSE, message=FALSE, echo=TRUE}
tictoc::tic("sdmTMB")
m <- sdmTMB(present ~ 0 + as.factor(year),
  data = d,
  mesh = sdmTMB_mesh,
  time = "year",
  spatial = "off",
  spatiotemporal = "ar1",
  family = binomial(link = "logit"),
  priors = sdmTMBpriors(
    matern_st = pc_matern(
      range_gt = range_min, range_prob = prior_prob,
      sigma_lt = sigma_max, sigma_prob = prior_prob
    ),
    b = normal(0, 100),
    ar1_rho = normal(0, 2)
  ),
  reml = TRUE # to match INLA "eb" `int.strategy`
)
tictoc::toc()
```

```{r sdmtmb-m, echo=TRUE}
print(m)
```

We can most easily extract the fixed-effect estimates and confidence intervals using \pkg{smdTMB}'s `tidy()` function.

```{r sdmtmb-coefs, echo=TRUE}
sdmTMB_coefs <- tidy(m, effects = "fixed", conf.int = TRUE)
sdmTMB_coefs
```

## Comparison

The estimates for the spatial parameters are very similar between the INLA-based models and `sdmTMB()`:

```{r comparison, echo=TRUE}
# INLA
m_inla$summary.hyperpar[, c("mean", "sd", "0.025quant", "0.975quant")]

# inlabru
m_bru$summary.hyperpar[, c("mean", "sd", "0.025quant", "0.975quant")]

# sdmTMB
tidy(m, "ran_pars", conf.int = TRUE)
```

Note that, within \pkg{sdmTMB}, the standard errors for variance terms are not reported in the `tidy()` function as they could be misleading if considered in natural (rather than log) space.
The confidence intervals are, however, reported.

Comparing the fixed effects shows the estimates from all three models are also very similar.

```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>%
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

bru_coefs <- m_bru$summary.random$fac[, c("0.025quant", "mean", "0.975quant")] %>%
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>%
  mutate(term = unique(sdmTMB_coefs$term), model = "inlabru")

dplyr::bind_rows(mutate(sdmTMB_coefs, model = "sdmTMB"), inla_coefs, bru_coefs) %>%
  ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, colour = model)) +
  geom_pointrange(position = position_dodge(width = 0.2)) +
  theme(axis.title.y = element_blank())
```

With sdmTMB, we can extract simulated draws from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.

```{r sdmTMB-sims, echo=FALSE}
set.seed(1)
post_sdmTMB <- gather_sims(m, nsim = 2000)
post_sdmTMB <- post_sdmTMB %>%
  mutate(
    variable =
      factor(`.variable`,
        levels = unique(post_sdmTMB$`.variable`),
        labels = c(seq(2007, 2017, 2), "Range", "AR1 rho", "Random Field SD")
      )
  )
```

We can then compare the distribution of these \pkg{sdmTMB} simulations to the marginal posterior densities from \pkg{INLA}
From \pkg{INLA}, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the \pkg{sdmTMB} `gather_sims()` function.
Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the \pkg{INLA} model posterior assuming we had set\
`control.compute` ` = list(config = TRUE)`.

```{r inla-densities, echo=FALSE}
x <- list()
x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2007")
x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2009")
x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2011")
x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2013")
x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2015")
x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2017")
x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>%
  as_tibble() %>%
  mutate(variable = "range")
x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>%
  as_tibble() %>%
  mutate(variable = "sigma_E")
x[[9]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>%
  as_tibble() %>%
  mutate(variable = "ar1_rho")
post_inla1 <- dplyr::bind_rows(x)
post_inla2 <- post_inla1 %>% mutate(variable = factor(variable,
  levels = unique(post_inla1$variable),
  labels = c(
    "2007", "2009", "2011", "2013", "2015", "2017",
    "Range", "Random Field SD", "AR1 rho"
  )
))
```

In this plot, the blue histograms are from \pkg{sdmTMB} and the density lines are from \pkg{INLA}:

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) +
  geom_histogram(
    data = post_sdmTMB, aes(.value, after_stat(density)), bins = 30,
    fill = "blue", alpha = 0.2
  ) +
  geom_line(data = post_inla2, aes(x, y), inherit.aes = FALSE) +
  facet_wrap(~variable, scales = "free")
```

## Annual distribution maps

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space.
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:

```{r sdm-predictions}
nd <- filter(qcs_grid, year >= 2007)
pred <- predict(m, newdata = nd)
```

```{r maps, cache=FALSE, out.width="5.5in"}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:

```{r inlabru-pred, out.width="5.5in"}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

ggplot(as.data.frame(pred_bru)) +
  geom_raster(aes(X, Y, fill = mean)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

Equivalent predictions are more complicated to extract from \pkg{INLA}, but are easy to plot for the original sample locations.

```{r inla-pred, out.width="5.5in"}
idx <- inla.stack.index(sdat, "stdata")$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]

ggplot(d) +
  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
  facet_wrap(~year) +
  scale_colour_viridis_c() +
  coord_fixed()
```

### 2022-11-02 doc/paper-jss/model-description.Rmd
Preping JSS

--- 2022-05-06 doc/appendix-model.Rmd
@@
----
-title: "Appendix 1: sdmTMB statistical model description"
-output:
-  bookdown::pdf_document2:
-    highlight: pygments
-    toc: true
-    number_sections: true
-fig_caption: true
-urlcolor: blue
-highlight: "monochrome"
-csl: mee.csl
-bibliography: refs.bib
-header-includes:
-  \usepackage[left]{lineno}
-  \usepackage{bm}
-  \usepackage{amssymb}
-  \linenumbers
----
-
-```{r setup, echo=FALSE, cache=FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  echo = FALSE,
-  error = FALSE,
-  message = FALSE,
-  warning = FALSE
-)
-```
+# Model
 
@@
 
-\clearpage
-
-# Notation conventions
+## Notation conventions
 
@@
 
-```{r indices}
-library(magrittr)
-subscripts <- tibble::tribble(
-  ~Symbol, ~Description,
-  "$\\s$", "Index for space; a vector of x and y coordinates",
-  "$t$", "Index for time",
-  "$g$", "Group"
-)
-knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex", booktabs = TRUE) %>%
-  kableExtra::kable_styling(latex_options = "hold_position")
-```
+\begin{table}[!h]
+\caption{Subscript notation}
+\centering
+\begin{tabular}[t]{ll}
+\toprule
+Symbol & Description\\
+\midrule
+$\s$ & Index for space; a vector of x and y coordinates\\
+$t$ & Index for time\\
+$g$ & Group\\
+\bottomrule
+\end{tabular}
+\end{table}
 
-```{r symbols, results='asis'}
-symbols <- tibble::tribble(
-  ~Symbol, ~Code, ~Description,
-  "$y$", "\\texttt{y\\_i}", "Observed response data",
-  "$\\mu$", "\\texttt{mu\\_i}", "Mean",
-  "$\\phi$", "\\texttt{phi}", "A dispersion parameter for a distribution",
-  "$f$", "\\texttt{fit\\$family\\$link}", "Link function",
-  "$f^{-1}$", "\\texttt{fit\\$family\\$linkinv}", "Inverse link function",
-  "$\\boldsymbol{\\beta}$", "\\texttt{b\\_j}", "Parameter vector",
-  "$\\boldsymbol{X}$", "\\texttt{X\\_ij}", "A predictor model matrix",
-  "$O_{\\s, t}$", "\\texttt{offset}", "An offset variable at point $\\s$ and time $t$",
-
-  "$\\omega_{\\s}$", "\\texttt{omega\\_s}", "Spatial random field at point $\\s$ (knot)",
-  "$\\omega_{\\s}^*$", "\\texttt{omega\\_s\\_A}", "Spatial random field at point $\\s$ (interpolated)",
-  "$\\zeta_{\\s}$", "\\texttt{zeta\\_s}", "Spatially varying coefficient random field at point $\\s$ (knot)",
-  "$\\zeta_{\\s}^*$", "\\texttt{zeta\\_s\\_A}", "Spatially varying coefficient random field at point $\\s$ (interpolated)",
-  "$\\epsilon_{\\s, t}$", "\\texttt{epsilon\\_st}", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
-  "$\\epsilon_{\\s, t}^*$", "\\texttt{epsilon\\_st\\_A}", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
-  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR(1) or random walk spatiotemporal deviations (knot)",
-  "$\\alpha_g$", "\\texttt{RE}", "IID random intercept deviation for group $g$",
-
-  "$\\boldsymbol{\\Sigma}_\\omega$", "\\texttt{-}", "Spatial random field covariance matrix",
-  "$\\boldsymbol{\\Sigma}_\\zeta$", "\\texttt{-}", "Spatially varying coefficient random field covariance matrix",
-  "$\\boldsymbol{\\Sigma}_\\epsilon$", "\\texttt{-}", "Spatiotemporal random field covariance matrix",
-  "$\\boldsymbol{Q}_\\omega$", "\\texttt{Q\\_s}", "Spatial random field precision matrix",
-  "$\\boldsymbol{Q}_\\zeta$", "\\texttt{Q\\_s}", "Spatially varying coefficient random field precision matrix",
-  "$\\boldsymbol{Q}_\\epsilon$", "\\texttt{Q\\_st}", "Spatiotemporal random field precision matrix",
-
-  "$\\sigma_\\alpha^2$", "\\texttt{sigma\\_G}", "IID random intercept variance",
-  "$\\sigma_\\epsilon^2$", "\\texttt{sigma\\_E}", "Spatiotemporal random field marginal variance",
-  "$\\sigma_\\omega^2$", "\\texttt{sigma\\_O}", "Spatial random field marginal variance",
-  "$\\sigma_\\zeta^2$", "\\texttt{sigma\\_Z}", "Spatially varying coefficient random field marginal variance",
-  "$\\kappa_\\omega$", "\\texttt{kappa(0)}", "Spatial decorrelation rate",
-  "$\\kappa_\\epsilon$", "\\texttt{kappa(1)}", "Spatiotemporal decorrelation rate",
-
-  "$\\rho$", "\\texttt{ar1\\_rho}", "Correlation between random fields in subsequent time steps",
-  "$\\boldsymbol{A}$", "\\texttt{A}", "Sparse projection matrix to interpolate between knot and data locations",
-  "$\\boldsymbol{H}$", "\\texttt{H}", "2-parameter rotation matrix used to define anisotropy"
-)
-knitr::kable(symbols, caption = "Symbol notation, code representation (in model output or in model template code), and descriptions.", escape = FALSE, format = "latex", booktabs = TRUE,
-   linesep = c(
-    rep('', 7), # y, mean etc.
-    '\\addlinespace',
-    rep('', 7), # fields
-    '\\addlinespace',
-    rep('', 5), # covariance stuff
-    '\\addlinespace',
-    rep('', 5), # more covariance stuff
-    '\\addlinespace',
-    rep('', 99) # end
-  ))
-```
+\begin{table}
+\caption{Symbol notation, code representation (in model output or in model template code), and descriptions.}
+\centering
+\begin{tabular}[t]{lll}
+\toprule
+Symbol & Code & Description\\
+\midrule
+$y$ & \texttt{y\_i} & Observed response data\\
+$\mu$ & \texttt{mu\_i} & Mean\\
+$\phi$ & \texttt{phi} & A dispersion parameter for a distribution\\
+$f$ & \texttt{fit\$family\$link} & Link function\\
+$f^{-1}$ & \texttt{fit\$family\$linkinv} & Inverse link function\\
+$\boldsymbol{\beta}$ & \texttt{b\_j} & Parameter vector\\
+$\boldsymbol{X}$ & \texttt{X\_ij} & A predictor model matrix\\
+$O_{\s, t}$ & \texttt{offset} & An offset variable at point $\s$ and time $t$\\
+\addlinespace
+$\omega_{\s}$ & \texttt{omega\_s} & Spatial random field at point $\s$ (knot)\\
+$\omega_{\s}^*$ & \texttt{omega\_s\_A} & Spatial random field at point $\s$ (interpolated)\\
+$\zeta_{\s}$ & \texttt{zeta\_s} & Spatially varying coefficient random field at point $\s$ (knot)\\
+$\zeta_{\s}^*$ & \texttt{zeta\_s\_A} & Spatially varying coefficient random field at point $\s$ (interpolated)\\
+$\epsilon_{\s, t}$ & \texttt{epsilon\_st} & Spatiotemporal random field at point $\s$ and time $t$ (knot)\\
+$\epsilon_{\s, t}^*$ & \texttt{epsilon\_st\_A} & Spatiotemporal random field at point $\s$ and time $t$ (interpolated)\\
+$\delta_{\s,t}$ & \texttt{b\_t} & AR(1) or random walk spatiotemporal deviations (knot)\\
+$\alpha_g$ & \texttt{RE} & IID random intercept deviation for group $g$\\
+\addlinespace
+$\boldsymbol{\Sigma}_\omega$ & \texttt{-} & Spatial random field covariance matrix\\
+$\boldsymbol{\Sigma}_\zeta$ & \texttt{-} & Spatially varying coefficient random field covariance matrix\\
+$\boldsymbol{\Sigma}_\epsilon$ & \texttt{-} & Spatiotemporal random field covariance matrix\\
+$\boldsymbol{Q}_\omega$ & \texttt{Q\_s} & Spatial random field precision matrix\\
+$\boldsymbol{Q}_\zeta$ & \texttt{Q\_s} & Spatially varying coefficient random field precision matrix\\
+$\boldsymbol{Q}_\epsilon$ & \texttt{Q\_st} & Spatiotemporal random field precision matrix\\
+\addlinespace
+$\sigma_\alpha^2$ & \texttt{sigma\_G} & IID random intercept variance\\
+$\sigma_\epsilon^2$ & \texttt{sigma\_E} & Spatiotemporal random field marginal variance\\
+$\sigma_\omega^2$ & \texttt{sigma\_O} & Spatial random field marginal variance\\
+$\sigma_\zeta^2$ & \texttt{sigma\_Z} & Spatially varying coefficient random field marginal variance\\
+$\kappa_\omega$ & \texttt{kappa(0)} & Spatial decorrelation rate\\
+$\kappa_\epsilon$ & \texttt{kappa(1)} & Spatiotemporal decorrelation rate\\
+\addlinespace
+$\rho$ & \texttt{ar1\_rho} & Correlation between random fields in subsequent time steps\\
+$\boldsymbol{A}$ & \texttt{A} & Sparse projection matrix to interpolate between knot and data locations\\
+$\boldsymbol{H}$ & \texttt{H} & 2-parameter rotation matrix used to define anisotropy\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-# sdmTMB model structure
+## sdmTMB model structure
 
@@
 
-## Main effects
+### Main effects
 
@@
 
-### Smoothers
+#### Smoothers
 
@@
 
-### Linear break-point threshold models
+#### Linear break-point threshold models
 
@@
 
-### Logistic threshold models
+#### Logistic threshold models
 
@@
 
-## Spatial random fields
+### Spatial random fields
 
@@
 
-## Spatiotemporal random fields
+### Spatiotemporal random fields
 
@@
 
-### IID spatiotemporal random fields
+#### IID spatiotemporal random fields
 
@@
 
-### AR(1) spatiotemporal random fields
+#### AR(1) spatiotemporal random fields
 
@@
 
-### Random walk spatiotemporal random fields (RW)
+#### Random walk spatiotemporal random fields (RW)
 
@@
 
-## Time-varying regression parameters
+### Time-varying regression parameters
 
@@
 
-## Spatially varying coefficients (SVC)
+### Spatially varying coefficients (SVC)
 
@@
 
-## IID random or multi-level intercepts
+### IID random or multi-level intercepts
 
@@
 
-## Offset terms
+### Offset terms
 
@@
 
-# Observation model families
+## Observation model families
 
@@
 
-## Binomial
+### Binomial
 
@@
 
-## Beta
+### Beta
 
@@
 
-## Gamma
+### Gamma
 
@@
 
-## Gaussian
+### Gaussian
 
@@
 
-## Lognormal
+### Lognormal
 
@@
 
-Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.
+Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the \pkg{TMB} `dnorm()` normal density.
 
@@
 
-## Negative Binomial 1 (NB1)
+### Negative Binomial 1 (NB1)
 
@@
 
-## Negative Binomial 2 (NB2)
+### Negative Binomial 2 (NB2)
 
@@
 
-## Poisson
+### Poisson
 
@@
 
-## Student-t
+### Student-t
 
@@
 
-## Tweedie
+### Tweedie
 
@@
 
-# Gaussian random fields
+## Gaussian random fields
 
-## Matérn parameterization
+### Matérn parameterization
 
@@
 
-## Projection $\boldsymbol{A}$ matrix
+### Projection A matrix
 
@@
 
-## Anisoptropy
+### Anisotropy
 
@@
 
-## Incorporating physical barriers into the SPDE
+### Incorporating physical barriers into the SPDE
 
@@
 
-# Optimization
+## Optimization
 
-## Optimization details
+### Optimization details
 
@@
 
-## Assessing convergence
+### Assessing convergence
 
@@
 If maximum gradients are already relatively small, they can often be reduced further with additional optimization calls beginning at the previous best parameter vector as described above with `sdmTMB::run_extra_optimization()`.
-
-# References

### 2022-11-02 doc/paper-jss/owls.Rmd
Preping JSS

--- 2022-05-06 doc/appendix-snow-sdmTMB.Rmd
@@
----
-title: "Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB"
-output:
-  bookdown::pdf_document2:
-    highlight: pygments
-    toc: true
-    number_sections: true
-urlcolor: blue
-bibliography: refs.bib
-csl: mee.csl
-header-includes:
-  \usepackage[left]{lineno}
-  \usepackage{bm}
-  \usepackage{amssymb}
-  \linenumbers
----
-
-```{r setup, include = FALSE, cache=FALSE}
+```{r setup-owls, include = FALSE, cache=FALSE}
 knitr::opts_chunk$set(
@@
 
-# Background
+# Example: spatially varying effect on owl counts
+
+## Background
 
@@
 
-# Preparation of spatial data
+## Preparation of spatial data
 
-```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
-library(ggplot2)
-library(patchwork)
-library(dplyr)
-library(sdmTMB)
-library(sf)
+```{r packages-owls, message=FALSE, warning=FALSE, cache=FALSE}
+library("ggplot2")
+library("patchwork")
+library("dplyr")
+library("sf")
+library("sdmTMB")
+```
+
+```{r rnaturalearthdata, echo=FALSE}
+if (!require("rnaturalearthdata", quietly = TRUE)) {
+  stop(
+    "Please install 'rnaturalearthdata'.\n",
+    "`remotes::install_github('ropensci/rnaturalearthdata')`"
+  )
+}
 ```
@@
 snow <- readRDS(here::here("data/SNOW_data.rds"))
-glimpse(snow)
+```
+
+Our data frame is called `snow`:
+
+```{r head-snow, echo=TRUE}
+head(snow, n = 3)
 ```
@@
 
-```{r mesh, echo=TRUE, out.width="4in"}
+```{r mesh-owls, echo=TRUE, out.width="4in"}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO.
+<!-- Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO.
 The dataframe also contains the Southern Oscillation Index (SOI), which is related to El Niño cycles, but we will only consider NAO in this example because any effect of SOI was even weaker.
+-->
 
-```{r explore, fig.asp = 0.3, cache=FALSE}
-ggplot(snow, aes(year, count, colour = nao)) +
-  geom_jitter() +
-  scale_colour_viridis_c() +
-  theme_light()
-```
+## Model structure
 
-# Model structure
-
-First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
-<!-- We will use `normalize = TRUE` to speed up optimization.  -->
-<!-- Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output. -->
 Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America.
@@
 If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
-It might also be appropriate to include count ID as another random effect `(1 | CBCID)` to account for consistent differences between count circles in habitat and human observers. 
-But since its inclusion doesn't change our main result and it slows model fitting, we will rely on the spatial random effects to link observations in the same location for this example.
+<!-- It might also be appropriate to include count ID as another random effect `(1 | CBCID)` to account for consistent differences between count circles in habitat and human observers.  -->
+<!-- But since its inclusion does not change our main result and it slows model fitting, we will rely on the spatial random effects to link observations in the same location for this example. -->
+<!-- Since these models will take a while to fit we will set `silent = FALSE` to monitor optimization progress.  -->
 
-Since these models will take a while to fit, in practice, we might set `silent = FALSE` to monitor optimization progress. 
-However, we will maintain the default `silent = TRUE` to avoid cluttering the document with the output.
+We initially fit this model with NB1 and NB2 negative binomial observation error and Poisson error.
+According to AIC (`AIC()`), the `nbinom2()` (NB2) distribution has the most parsimonious fit.
+We will therefore proceed with that model.
 
@@
 rds_saved <- file.exists(here::here("data/snow_w_main_effect_0_150km.rds"))
+rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_0_150km2.rds"))
 ```
 
-```{r fit-poisson, echo=TRUE, eval=!rds_saved, warning=FALSE}
-m0 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ 0 + nao,
+```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
+fit_owl <- sdmTMB(count ~ 1 + nao + (1 | year_f),
+  spatial_varying = ~ nao,
   time = "year",
+  data = snow, 
+  family = nbinom2(link = "log"),
   spatial = "on",
-  spatiotemporal = "IID",
-  family = poisson(link = "log"),
-  data = snow, mesh = mesh,
-  silent = TRUE
+  spatiotemporal = "iid",
+  mesh = mesh,
+  reml = TRUE,
+  silent = FALSE
 )
@@
 
-```{r fit-negb1, echo=TRUE, eval=!rds_saved, warning=FALSE}
-m1 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ 0 + nao,
-  time = "year",
-  spatial = "on",
-  spatiotemporal = "IID",
-  family = nbinom1(link = "log"),
-  data = snow, mesh = mesh,
-  silent = TRUE
-)
+```{r cache-negb2, echo=FALSE, eval=!rds_saved}
+saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km.rds"))
 ```
 
-```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
-m2 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ 0 + nao,
-  time = "year",
-  spatial = "on",
-  spatiotemporal = "IID",
-  family = nbinom2(link = "log"),
-  data = snow, mesh = mesh,
-  silent = TRUE
-)
+```{r read-negb2, echo=FALSE, eval=rds_saved}
+fit_owl <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
 ```
 
-# Model validation
+## Model validation
 
-## Basic inspection
+We will run the `sanity()` function to apply some basic checks.
 
-```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
-if (!rds_saved) {
-  saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
-  saveRDS(m1, here::here("data/snow_w_main_effect_1_150km.rds"))
-  saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
-}
+```{r sanity, echo=TRUE}
+sanity(fit_owl)
 ```
 
-```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
-if (rds_saved) {
-  m0 <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
-  m1 <- readRDS(here::here("data/snow_w_main_effect_1_150km.rds"))
-  m2 <- readRDS(here::here("data/snow_w_main_effect_150km.rds"))
-}
+```
+#>   Non-linear minimizer suggests successful convergence
+#>   Hessian matrix is positive definite
+#>   No extreme or very small eigen values detected
+#> x `ln_tau_E` gradient > 0.001
+#>   See `?run_extra_optimization()`
+#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
+#> 
+#> x `ln_phi` gradient > 0.001
+#>   See `?run_extra_optimization()`
+#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
+#> 
+#>   No fixed-effect standard errors are NA
+#>   No fixed-effect standard errors look unreasonably large
+#>   No sigma parameters are < 0.01
+#>   No sigma parameters are > 100
+#>   Range parameter doesn't look unreasonably large
 ```
 
-```{r aic, echo=TRUE, cache=FALSE}
-AIC(m0, m1, m2)
+We note that some log-likelihood gradients are slightly large for some fixed effects.
+
+```{r}
+fit_owl$gradients
 ```
 
-According to AIC, the `nbinom2()` (negative binomial, NB2) distribution has the most parsimonious fit.
-We could also check residual distributions at this stage.
-We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and check that the gradients are all small enough.
-If the gradients aren't quite small enough, we can easily run additional optimization steps using the `run_extra_optimization()` function.
+We can run additional optimization steps using the `run_extra_optimization()` function.
 
-```{r sanity, echo=TRUE}
-sanity(m2)
+```{r owl-newton, results='hide', eval=!rds_saved2}
+fit_owl2 <- run_extra_optimization(fit_owl, nlminb_loops = 0, newton_loops = 1)
 ```
 
-```{r rds2, echo=FALSE, cache=FALSE}
-rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_150km_reml.rds"))
+```{r cache-negb2-2, echo=FALSE, eval=!rds_saved2}
+saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km2.rds"))
 ```
 
-```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE, cache=FALSE}
-m <- sdmTMB(count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ 0 + nao,
-  time = "year", 
-  spatial = "on", 
-  spatiotemporal = "iid",
-  family = nbinom2(link = "log"),
-  reml = TRUE,
-  data = snow, mesh = mesh,
-  silent = TRUE
-)
+```{r read-negb2-2, echo=FALSE, eval=rds_saved2}
+fit_owl2 <- readRDS(here::here("data/snow_w_main_effect_0_150km2.rds"))
 ```
 
-```{r save-reml-model, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
-if (!rds_saved2) {
-  saveRDS(m, here::here("data/snow_w_main_effect_150km_reml.rds"))
-}
+And confirm that the gradients are smaller:
+
+```{r sanity2, echo=TRUE}
+sanity(fit_owl2)
+```
+
+```
+#> Non-linear minimizer suggests successful convergence
+#> Hessian matrix is positive definite
+#> No extreme or very small eigen values detected
+#> No gradients with respect to fixed effects are >= 0.001
+#> No fixed-effect standard errors are NA
+#> No fixed-effect standard errors look unreasonably large
+#> No sigma parameters are < 0.01
+#> No sigma parameters are > 100
+#> Range parameter doesn't look unreasonably large
 ```
@@
 
-```{r check, echo=FALSE, include=FALSE, eval=TRUE, cache=FALSE}
-if (rds_saved2) {
-  m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
-  m # printing this first just prevents some excess messages from cluttering the output when printing below
-}
-```
-
-```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
-print(m)
+```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
+print(fit_owl2)
 ```
@@
 
-```{r tidy, cache=FALSE}
-tidy(m, conf.int = TRUE)
-tidy(m, effects = "ran_pars", conf.int = TRUE)
+```{r tidy-owls, cache=FALSE}
+tidy(fit_owl2, conf.int = TRUE)
+tidy(fit_owl2, effects = "ran_pars", conf.int = TRUE)
 ```
@@
 
-```{r predict, message=FALSE}
-pred <- predict(m)
+```{r predict-owls, message=FALSE}
+pred <- predict(fit_owl2)
 set.seed(19208)
-pred$resid <- residuals(m)
+pred$resid <- residuals(fit_owl2)
 ```
-
-<!-- \clearpage -->
 
@@
 
-```{r qqnorm, out.width="33%"}
+```{r qqnorm-owls, out.width="60%"}
 qqnorm(pred$resid)
@@
 
-More accurate residuals can be generated using tmbstan, but this a slower process.
-These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via tmbstan/Stan.
+More accurate residuals can be generated using \pkg{tmbstan}, but this a slower process.
+These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
 To calculate these, we have to use the version of our model fit without REML.
 We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
+For the purposes of this building this appendix quickly, we will not run this code.
 
-```{r stan-resids}
-f <- here::here("data/mcmc_residuals.rds") # for caching
-set.seed(82919)
-if (!file.exists(f)) {
-  mcmc_resids <- residuals(m2, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
-  saveRDS(mcmc_resids, f)
-} else {
-  mcmc_resids <- readRDS(f)
-}
+```{r stan-resids-owls, eval=FALSE}
+mcmc_resids <- residuals(fit_owl, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
 ```
 
-```{r stan-qqnorm, out.width="33%"}
-qqnorm(mcmc_resids)
-qqline(mcmc_resids)
-```
-
-We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the DHARMa package.
+We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the \pkg{DHARMa} package.
 If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.
 
-```{r zero-test, echo=TRUE, message=FALSE}
-s_nb2 <- simulate(m, nsim = 500)
-pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
+```{r zero-test-owls, echo=TRUE, message=FALSE}
+s_nb2 <- simulate(fit_owl2, nsim = 500)
+pred_fixed <- fit_owl2$family$linkinv(pred$est_non_rf)
 r_nb2 <- DHARMa::createDHARMa(
   simulatedResponse = s_nb2,
-  observedResponse = m$data$count,
+  observedResponse = fit_owl2$data$count,
   fittedPredictedResponse = pred_fixed
@@
 
-Next, we modify a theme to plot spatiotemporal data and check the distribution of these residuals:
+We do not find evidence of zero inflation.
 
-```{r theme, cache=FALSE}
+```{r theme-owls, cache=FALSE, echo=FALSE}
 theme_set(theme_bw() + theme(
@@
 
-```{r plot-resids, fig.width=8, dpi=200}
-ggplot(pred, aes(X, Y, colour = resid)) +
-  geom_point(size = 0.25) +
-  facet_wrap(~year) +
-  coord_fixed() +
-  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
-  theme(axis.text = element_blank(), axis.ticks = element_blank())
-```
+## Plotting spatial predictions
 
-# Plotting spatial predictions
-
-To produce pretty maps of these predictions, we back transformed the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.
+To produce maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an \pkg{sf} object so that the plot axes will be in latitude and longitude.
 
@@
 ```
-
-We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from <http://www.naturalearthdata.com/> and transformed them to the correct projection.
+{
+Next we retrieve continental coastlines from the package \pkg{rnaturalearth} and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.
 
@@
 We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
-Most simply, we can extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.
+Most simply, we can extract the overall coefficient using the `b <- tidy(fit_owl2)` and add it to the spatially varying coefficients `zeta_s`.
 However, this method does not give us spatially explicit estimates of uncertainty.
-To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`).
-We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `nsim = 500`). 
+To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `nsim = 200`) and what component we want estimated (`sims_var = "zeta_s"`).
+We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `nsim = 200`). 
 The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws (in this case we will use 80%) and adding these values to our prediction data frame.
 
-```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE}
-b <- tidy(m, conf.int = TRUE)
+```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
+b <- tidy(fit_owl2, conf.int = TRUE)
 
-nsim <- 500
-zeta_s <- predict(m, nsim = nsim, sims_var = "zeta_s")
-sims <- spread_sims(m, nsim = nsim)
+nsim <- 200
+zeta_s <- predict(fit_owl2, nsim = nsim, sims_var = "zeta_s")
+sims <- spread_sims(fit_owl2, nsim = nsim)
 beta <- sims$nao
@@
 
-```{r}
+```{r, echo=FALSE}
 nao_effect_lwr95 <- as.numeric(apply(combined, 2, quantile, probs = 0.025))
@@
 Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`).
-All these effects are in link space (`link = "log"`) and could be `exp()` to give the expected multiplicative contribution to the predicted owl counts, however the additive nature of the effect in link space is actually somewhat more intuitive to interpret visually, so this is what we will plot on our maps:
+All these effects are in link space (`link = "log"`) and could be exponentiated to give the expected multiplicative contribution to the predicted owl counts; however, the additive nature of the effect in link space is more intuitive to interpret visually, so this is what we will plot on our maps:
 
-```{r plot-map, echo=TRUE, eval=TRUE, out.width="90%", echo=FALSE, cache=FALSE}
+```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%", echo=FALSE}
 ggplot(data = p_proj) +
@@
 
-```{r plot-map-ci, echo=F, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
+```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
 p1 <- ggplot(data = p_proj) +
@@
 
-If we recalculated these at 95% CI the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that for the most conventional level of statistical confidence, we cannot conclude that the effect of NAO at an annual scale differs from 0.
+If we recalculated these at 95% confidence intervals levels the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that for the most conventional level of statistical confidence, we cannot conclude that the effect of NAO at an annual scale differs from 0.
 However, the southeast to northwest gradient in the intensity illustrated in the maps above is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
 This result suggests that exploring the effects of NAO at finer temporal scales (possibly seasonally) and/or with lagged effects could prove fruitful.
-
-<!-- \clearpage -->
 
@@
 
-```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE, dpi=220}
+```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, cache=FALSE, dpi=220}
 ggplot(data = p_proj) +
@@
 
-<!-- Note that for all of these annually varying predictions, we are showing only the most recent 25 years.  -->
-
-\clearpage
 ## Other components of predictions
@@
 
-```{r plot-spatial-effects, echo=TRUE, eval=TRUE, out.width="5in", dpi=200}
+```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200}
 ggplot(data = p_proj) +
@@
 
-\clearpage
-
 ### 2. Spatiotemporal random effects
@@
 These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
-For owls, this is most likely to be things like landcover changes, prey abundance cycles, and snow depth.
+For owls, this is most likely to be factors like landcover changes, prey abundance cycles, and snow depth.
 
-```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE, dpi=210}
+```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-<!-- Note that for these annually varying predictions, we are showing only the most recent 25 years.  -->
-
-\clearpage
-
 ### 3. Just fixed effects
@@
 
-```{r plot-fixed-effects, echo=TRUE, eval=TRUE, dpi=210}
+```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 ```
-
-\clearpage
-# References
-
-::: {#refs}
-:::

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Preping JSS

@@
 author:
-  - name: FirstName LastName
-    orcid: 0000-0000-0000-0000
-    affiliation: University/Company
-    # use this syntax to add text on several lines
+  - name: Sean C. Anderson
+    orcid: 0000-0001-9563-1937
+    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
     address: |
-      | First line
-      | Second line
-    email: \email{name@company.com}
-    url: http://rstudio.com
-  - name: Second Author
-    orcid: 0000-0000-0000-0000
-    affiliation: 'Affiliation \AND'
-    # To add another line, use \AND at the end of the previous one as above
-  - name: Third Author
-    orcid: 0000-0000-0000-0000
-    address: |
-      | Department of Statistics and Mathematics,
-      | Faculty of Biosciences,
-      | Universitat Autònoma de Barcelona
-    affiliation: |
-      | Universitat Autònoma
-      | de Barcelona
-    # use a different affiliation in adress field (differently formated here)
-    affiliation2: Universitat Autònoma de Barcelona
+      | Pacific Biological Station, Nanaimo, BC, Canada
+    email: \email{sean.anderson@dfo-mpo.gc.ca}
+  - name: Eric J. Ward
+    orcid: 0000-0002-4359-0296
+    affiliation: 'Northwest Fisheries Science Center, National Marine Fisheries Service \AND'
+  - name: Philina A. English
+    orcid: 0000-0003-2992-6782
+    address: ''
+    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
+  - name: Lewis A. K. Barnett
+    orcid: 0000-0002-9381-8375
+    affiliation: 'Alaska Fisheries Science Center, National Marine Fisheries Service'
 bibliography: refs.bib
 title:
-  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
+  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
   # If you use tex in the formatted title, also supply version without
@@
 \begin{figure}[p]
-{\centering \includegraphics[width=9in]{../../figs/diagram}}
+\centering 
+\includegraphics[width=9in]{../../figs/diagram}
 \caption[]{
@@
 
-And we can run some basic checks on our model with the `sanity()` function:
+We can run some basic checks on our model with the `sanity()` function:
 
-```{r pcod-eg1-sanity}
+```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE}
 sanity(fit)
+```
+
+```
+#> Non-linear minimizer suggests successful convergence
+#> Hessian matrix is positive definite
+#> No extreme or very small eigen values detected
+#> No gradients with respect to fixed effects are >= 0.001
+#> No fixed-effect standard errors are NA
+#> No fixed-effect standard errors look unreasonably large
+#> No sigma parameters are < 0.01
+#> No sigma parameters are > 100
+#> Range parameter doesn't look unreasonably large
 ```
@@
 \begin{figure}
-{\centering \includegraphics[width=5.2in]{../../figs/pcod-fig}}
+\centering
+\includegraphics[width=5.2in]{../../figs/pcod-fig}
 \caption[]{Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada. 
@@
 
-We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
-Here we will keep the spatial field and structure the spatiotemporal random fields as AR(1).
+We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
+Here we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as an AR(1) process (`time_varying_type = "ar1"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
 
@@
   anisotropy = TRUE,
-  silent = FALSE # print fitting progress
+  silent = FALSE
 )
@@
 This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop. 
-We can inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
+We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
 We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix 2).
@@
 \begin{figure}
-{\centering \includegraphics[width=3.5in]{../../figs/timing-spatial-m1-april11}}
+\centering
+\includegraphics[width=3.5in]{../../figs/timing-spatial-m1-april11}
 \caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
@@
 We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
+
+\clearpage
+
+\appendix
+
+```{r child="model-description.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```
+
+```{r child="pcod.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```
+
+```{r child="owls.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```
+
+```{r child="inla-comparison.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```
+
+```{r child="timing.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```

### 2022-11-02 doc/paper-jss/pcod.Rmd
Preping JSS

--- 2022-05-06 doc/appendix-pcod-sdmTMB.Rmd
@@
----
-title: "Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB"
-output:
-  bookdown::pdf_document2:
-    highlight: pygments
-    toc: true
-    number_sections: true
-bibliography: refs.bib
-csl: mee.csl
-header-includes:
-  \usepackage[left]{lineno}
-  \usepackage{bm}
-  \usepackage{amssymb}
-  \linenumbers
----
+# Example: spatiotemporal modelling of Pacific cod
 
-```{r setup, include = FALSE, cache=FALSE}
+```{r setup-pcod, include = FALSE, cache=FALSE}
 knitr::opts_chunk$set(
@@
   autodep = TRUE,
-  cache.comments = FALSE
+  cache.comments = FALSE,
+  echo = TRUE
 )
@@
 The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in.
+Custom built meshes with \pkg{INLA} (including meshes with barriers limiting covariance across polygons) can also be passed in.
 See examples included in `?make_mesh`.
@@
 
-```{r mesh, echo = TRUE, cache=FALSE}
+```{r mesh, echo = TRUE, cache=TRUE}
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
@@
 
-```{r plot-mesh, cache=FALSE, fig.align='center', fig.asp=1, out.width="3in"}
+```{r plot-mesh, cache=TRUE, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE}
 plot(mesh)
@@
 
-Next we will fit a GLMM (generalized linear mixed effects model).
+Next we will fit a GLMM.
 For this example, we will include `0 + as.factor(year)` so that there is a factor predictor that represents the mean estimate for each time slice.
-Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.
+Setting `silent = FALSE` would gives us optimization details as the model fits.
 
@@
 
-```{r fit-pcod, warning=FALSE, message=FALSE}
-fit <- sdmTMB(density ~ 0 + as.factor(year),
+```{r fit-pcod, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, results='hide'}
+fit <- sdmTMB(
+  density ~ 0 + as.factor(year),
   data = pcod,
-  time = "year", mesh = mesh, family = tweedie(link = "log")
+  time = "year", 
+  mesh = mesh,
+  family = tweedie(link = "log"),
+  silent = FALSE
 )
@@
 
-```{r print-fit}
+```{r print-fit, echo=TRUE}
 print(fit)
@@
 
-```{r tidy}
+```{r pcod-tidy, echo=TRUE}
 tidy(fit, conf.int = TRUE)
@@
 
-```{r predict}
+```{r pcod-predict, cache=TRUE, echo=TRUE}
 pred <- predict(fit)
@@
 
-```{r sim-resids, out.width="4in", warning=FALSE, results='hide'}
+```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE}
 set.seed(123)
@@
   type = "mle-mcmc", 
-  mcmc_iter = 400, mcmc_warmup = 300
+  mcmc_iter = 200, mcmc_warmup = 100
 )
@@
 
-```{r sims}
-pred_sims <- predict(fit, newdata = qcs_grid, nsim = 500)
+```{r pcod-sims, cache=TRUE}
+pred_sims <- predict(fit, newdata = qcs_grid, nsim = 200)
 dim(pred_sims)
@@
 
-```{r sims-cv-plot}
+```{r sims-cv-plot, cache=TRUE}
 qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
@@
 
-We can calculate our standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE`.
-This returns the TMB objective function, which is required to pass the output to `get_index()`.
+We can calculate a standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE`.
+This returns the \pkg{TMB} objective function, which is required to pass the output to `get_index()`.
 
-```{r index, fig.asp=0.45, out.width="5in"}
+```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE}
 qcs_grid$area <- 4 # all 2 x 2km
@@
 The above example uses `bias_correct = TRUE` in `get_index()`.
-The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in the [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias].
-Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
+The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in @thorson2016bias.
+Leaving `bias_correct = FALSE` is fine for experimentation, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
 Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but takes more time and memory than skipping the bias correction.
-
-# References

### 2022-11-02 doc/paper-jss/timing.Rmd
Preping JSS

# Speed testing related packages

Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions.
Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
The Gaussian random field was parameterized with a range of 0.5 and marginal standard deviation of 0.2.
The model included an intercept with a value of 0.2 and a normal(0, 1) predictor with an associated coefficient of -0.4.
The observation error was Gaussian with a standard deviation of 0.3.

We generated a mesh that was consistent across simulated data sets for a given mesh resolution.
We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure 4, we report on the x-axis the number of mesh nodes ("knots") that result from each of these meshes.

We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.
The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.1.3 and the default Matrix library. The following package versions were used:

```{r pkg-versions, echo=TRUE}
packageVersion("INLA")
packageVersion("inlabru")
packageVersion("mgcv")
packageVersion("spaMM")
packageVersion("sdmTMB")
```

## Illustration of generating the INLA mesh

```{r pkgs, warning=FALSE, message=FALSE, cache=FALSE, echo=FALSE}
library("INLA")
library("inlabru")
library("ggplot2")
library("sdmTMB")
library("mgcv")
library("spaMM")
```

We will illustrate with a `max.edge` of 0.06:

```{r mesh-timing, echo=TRUE}
max_edge <- 0.06
loc_bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
segm_bnd <- INLA::inla.mesh.segment(loc_bnd)
mesh <- INLA::inla.mesh.2d(
  boundary = segm_bnd,
  max.edge = c(max_edge, 0.2),
  offset = c(0.1, 0.05)
)
```

This mesh has `r mesh$n` (`mesh$n`) vertices.

Here are the other meshes used from least to most vertices:

```{r mesh-vis-timing, out.width="6.5in", fig.width=10, fig.asp=1.2, echo=FALSE}
max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
g <- list()
for (i in seq_along(max_edge_vec)) {
  .mesh <- INLA::inla.mesh.2d(
    boundary = segm_bnd,
    max.edge = c(max_edge_vec[i], 0.2),
    offset = c(0.1, 0.05)
  )
  g[[i]] <- ggplot() +
    gg(.mesh) +
    theme_light() +
    coord_equal(expand = FALSE) +
    ggtitle(paste("Mesh vertices =", .mesh$n))
}
cowplot::plot_grid(plotlist = g, ncol = 2)
```

## Illustration of simulating data

```{r sim-timing, echo=TRUE}
set.seed(1)

n_obs <- 1000L
predictor_dat <- data.frame(
  X = runif(n_obs), Y = runif(n_obs),
  a1 = rnorm(n_obs)
)
mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)

sim_dat <- sdmTMB_simulate(
  formula = ~ 1 + a1,
  data = predictor_dat,
  mesh = mesh_sdmTMB,
  family = gaussian(),
  range = 0.5,
  phi = 0.3,
  sigma_O = 0.2,
  B = c(0.2, -0.4) # B0 = intercept, B1 = a1 slope
)
```

Look at the simulated data:

```{r sim-dat-plot-timing, out.width="5in", cache=FALSE, echo=FALSE}
ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
  geom_point() +
  scale_color_gradient2() +
  coord_fixed() +
  theme_light()
```

## Example sdmTMB model fit

```{r sdmTMBfit-timing, echo=TRUE}
fit_sdmTMB <- sdmTMB(
  observed ~ a1,
  data = sim_dat, mesh = mesh_sdmTMB, family = gaussian(),
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2))
)
```

## Example spaMM model fit

```{r spaMMfit-timing, echo=TRUE, cache=TRUE}
spde <- INLA::inla.spde2.pcmatern(
  mesh = mesh,
  prior.range = c(0.05, 0.05),
  prior.sigma = c(2, 0.05)
)

fit_spaMM <- fitme(
  observed ~ a1 + IMRF(1 | X + Y, model = spde),
  family = gaussian(), data = sim_dat
)
```

## Example inlabru model fit

```{r inlabrufit-timing, echo=TRUE, cache=TRUE}
# convert to sp SpatialPointsDataFrame first:
dat_sp <- sp::SpatialPointsDataFrame(
  cbind(sim_dat$X, sim_dat$Y),
  proj4string = sp::CRS(
    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
+ +y_0=0 +datum=NAD83 +units=km +no_defs"
  ), data = sim_dat
)
components <- observed ~ -1 + Intercept(1) + a1 +
  spatrf(main = coordinates, model = spde)
like <- like(observed ~ Intercept + a1 + spatrf,
  family = "gaussian", data = dat_sp
)
fit_bru <- bru(
  like,
  components = components,
  options = bru_options(
    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
    bru_max_iter = 1, num.threads = "1:1"
  )
)
```

## Example mgcv model fit

```{r mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE, cache=TRUE}
# -------------------------------------------------------------------------
# Code in this chunk is from:
#
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
smooth.construct.spde.smooth.spec <- function(object, data, knots) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  if (is.null(object$xt)) {
    if (dim == 1) {
      t <- seq(min(x), max(x), len = object$bs.dim)
      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
    } else {
      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
    }
  } else {
    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
    mesh <- object$xt$mesh
  }
  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
  inlamats <- INLA::inla.mesh.fem(mesh)
  object$S <- list()
  object$S[[1]] <- as.matrix(inlamats$c1)
  object$S[[2]] <- 2 * as.matrix(inlamats$g1)
  object$S[[3]] <- as.matrix(inlamats$g2)
  object$L <- matrix(c(2, 2, 2, 4, 2, 0), ncol = 2)
  object$rank <- rep(object$bs.dim, 3)
  object$null.space.dim <- 0
  object$mesh <- mesh
  object$df <- ncol(object$X)
  class(object) <- "spde.smooth"
  return(object)
}
Predict.matrix.spde.smooth <- function(object, data) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  Xp <- INLA::inla.spde.make.A(object$mesh, x)
  return(as.matrix(Xp))
}
# End of code from
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
# -------------------------------------------------------------------------
```

```{r mgcvfit-timing, echo=TRUE, eval=FALSE}
# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth():
# from supplement of:
# Miller, D.L., Glennie, R. & Seaton, A.E. (2019). Understanding the Stochastic
# Partial Differential Equation approach to smoothing. Journal of Agricultural,
# Biological and Environmental Statistics.
# https://doi.org/10.1007/s13253-019-00377-z
fit_bam <- bam(
  observed ~ a1 + s(X, Y,
    bs = "spde", k = mesh$n,
    xt = list(mesh = mesh)
  ),
  data = sim_dat,
  family = gaussian(),
  method = "fREML",
  control = gam.control(scalePenalty = FALSE), discrete = TRUE
)
```

### 2022-11-02 analysis/mgcv_spde_smooth.R

@@ -1,36 +1,26 @@
-# MATERN SPDE MODEL IN MGCV
-# These functions define the Matern SPDE model as a basis-penalty smoother
-# using mgcv. In mgcv, a new smooth can be specified by creating two functions:
-# 1. A smooth.construct.xxx.smooth.spex function defines a smooth named "xxx", allowing
-#    the user to use this smooth in mgcv with the term s(..., bs = "xxx").
-# 2. A Predict.matrix.xxx.smooth function that defines how to compute the smooth
-#    over a user-defined set of values.
-
-# The INLA package is used to handle the mesh and to
-# compute the finite element matrices, but not for model
-# fitting.
-require(INLA)
-
-# Setup the SPDE Matern smooth in mgcv
-# See ?smooth.construct in mgcv for details of the input and output
-# Special note: the xt argument in s(..., bs = "spde", xt = ...) can be used
-# to specify a mesh, if NULL, a mesh with regular knots is constructed.
-smooth.construct.spde.smooth.spec <- function(object, data, knots){
-  # observation locations
+# -------------------------------------------------------------------------
+# Code in this chunk is from:
+#
+# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
+# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
+# https://doi.org/10.1007/s13253-019-00377-z
+#
+# Re-used here under a Creative Commons Attribution 4.0 International License
+# http://creativecommons.org/licenses/by/4.0/
+smooth.construct.spde.smooth.spec <- function(object, data, knots) {
   dim <- length(object$term)
   if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
   if (dim == 1) {
     x <- data[[object$term]]
   } else {
     x <- matrix(0, nr = length(data[[1]]), nc = 2)
-    x[,1] <- data[[object$term[1]]]
-    x[,2] <- data[[object$term[2]]]
+    x[, 1] <- data[[object$term[1]]]
+    x[, 2] <- data[[object$term[2]]]
   }
-  # setup mesh or accept user mesh
   if (is.null(object$xt)) {
     if (dim == 1) {
-      t <- seq(min(x), max(x), len=object$bs.dim)
-      mesh <- inla.mesh.1d(loc=t, degree=2, boundary="free")
[…]

### 2022-11-02 analysis/pcod-fig.R

@@ -1,22 +1,60 @@
-library(sdmTMB)
-library(dplyr)
-library(ggplot2)
-library(sf)
-library(patchwork)
+library("sdmTMB")
+library("dplyr")
+library("ggplot2")
+library("sf")
+library("patchwork")
+dir.create("figs", showWarnings = FALSE)
 
 if (!require("rnaturalearthhires", quietly = TRUE)) {
-  stop("Please install 'rnaturalearthhires'.\n",
-  "`remotes::install_github('ropensci/rnaturalearthhires')`")
-}
-if (!require("ggsidekick", quietly = TRUE)) {
-  stop("Please install 'ggsidekick'.\n",
-  "`remotes::install_github('seananderson/ggsidekick')`")
+  stop(
+    "Please install 'rnaturalearthhires'.\n",
+    "`remotes::install_github('ropensci/rnaturalearthhires')`"
+  )
 }
 
-theme_set(ggsidekick::theme_sleek())
+# from remotes::install_github('seananderson/ggsidekick')
+theme_sleek <- function(base_size = 11, base_family = "") {
+  half_line <- base_size / 2
+  theme_light(base_size = base_size, base_family = base_family) +
+    theme(
+      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+      axis.ticks.length = grid::unit(half_line / 2.2, "pt"), strip.background = element_rect(
+        fill = NA,
+        colour = NA
+      ), strip.text.x = element_text(colour = "grey30"),
+      strip.text.y = element_text(colour = "grey30"), axis.text = element_text(colour = "grey30"),
+      axis.title = element_text(colour = "grey30"), legend.title = element_text(
+        colour = "grey30",
+        size = rel(0.9)
+      ), panel.border = element_rect(
+        fill = NA,
+        colour = "grey70", size = 1
+      ), legend.key.size = grid::unit(
+        0.9,
+        "lines"
+      ), legend.text = element_text(
+        size = rel(0.7),
+        colour = "grey30"
+      ), legend.key = element_rect(
+        colour = NA,
+        fill = NA
+      ), legend.background = element_rect(
+        colour = NA,
[…]

### 2022-11-02 analysis/timing.R

@@ -1,19 +1,83 @@
-# Timing figure
-library(sdmTMB)
-library(dplyr)
-library(ggplot2)
-library(INLA)
-library(inlabru)
-library(mgcv)
-library(spaMM)
+# Timing figure -----------------------------------------------------------
+
+library("sdmTMB")
+library("dplyr")
+library("ggplot2")
+library("INLA")
+library("inlabru")
+library("mgcv")
+library("spaMM")
+dir.create("figs", showWarnings = FALSE)
+dir.create("analysis", showWarnings = FALSE)
 
 INLA::inla.setOption(num.threads = "1:1")
-source(here::here("analysis/mgcv_spde_smooth.R"))
 
-# if (Sys.info()[["user"]] == "seananderson") {
-# INLA::inla.setOption("pardiso.license", "~/Dropbox/licenses/pardiso.lic")
-# INLA::inla.setOption(inla.mode="experimental")
-# }
+# source(here::here("analysis/mgcv_spde_smooth.R"))
+# -------------------------------------------------------------------------
+# Code in this chunk is from:
+#
+# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
+# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
+# https://doi.org/10.1007/s13253-019-00377-z
+#
+# Re-used here under a Creative Commons Attribution 4.0 International License
+# http://creativecommons.org/licenses/by/4.0/
+smooth.construct.spde.smooth.spec <- function(object, data, knots) {
+  dim <- length(object$term)
+  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
+  if (dim == 1) {
+    x <- data[[object$term]]
+  } else {
+    x <- matrix(0, nr = length(data[[1]]), nc = 2)
+    x[, 1] <- data[[object$term[1]]]
+    x[, 2] <- data[[object$term[2]]]
+  }
+  if (is.null(object$xt)) {
+    if (dim == 1) {
+      t <- seq(min(x), max(x), len = object$bs.dim)
+      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
+    } else {
[…]

### 2022-11-02 doc/paper-jss/model-description.Rmd
Fixes

@@
 
-We include tables of all major indices (Table \@ref(tab:indices)) and symbols (Table \@ref(tab:symbols)).
+We include tables of all major indices (Table \ref{tab:indices}) and symbols (Table \ref{tab:symbols}).
 
@@
 \bottomrule
+\label{tab:indices}
 \end{tabular}
@@
 $\omega_{\s}^*$ & \texttt{omega\_s\_A} & Spatial random field at point $\s$ (interpolated)\\
-$\zeta_{\s}$ & \texttt{zeta\_s} & Spatially varying coefficient random field at point $\s$ (knot)\\
-$\zeta_{\s}^*$ & \texttt{zeta\_s\_A} & Spatially varying coefficient random field at point $\s$ (interpolated)\\
+$\zeta_{\s}$ & \texttt{zeta\_s} & Spatially varying coefficient (SVC) random field at point $\s$ (knot)\\
+$\zeta_{\s}^*$ & \texttt{zeta\_s\_A} & SVC random field at point $\s$ (interpolated)\\
 $\epsilon_{\s, t}$ & \texttt{epsilon\_st} & Spatiotemporal random field at point $\s$ and time $t$ (knot)\\
@@
 $\rho$ & \texttt{ar1\_rho} & Correlation between random fields in subsequent time steps\\
-$\boldsymbol{A}$ & \texttt{A} & Sparse projection matrix to interpolate between knot and data locations\\
+$\boldsymbol{A}$ & \texttt{A} & Sparse projection matrix (from knot to data locations)\\
 $\boldsymbol{H}$ & \texttt{H} & 2-parameter rotation matrix used to define anisotropy\\
 \bottomrule
+\label{tab:symbols}
 \end{tabular}

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Edits

@@
 abstract: >
-  Geostatistical data are common across disciplines and available in increasingly large volumes.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less interpretable ways.
-Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of `stats::glm()`, \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
-SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are also possible.
+  Geostatistical data are pervasive and available in increasingly large volumes.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement.
+Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are possible.
 We describe the statistical model, illustrate the package through two case studies, and compare the functionality, speed, and interface to related software.
 The case studies illustrate \pkg{sdmTMB}'s flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
-Modelling of these common but computationally complex relationships is further aided by built-in handling of delta (hurdle) models, break-point effects, and diagnostic tools including out-of-sample cross validation.
+Additional functionality includes delta (hurdle) models, break-point effects, and out-of-sample cross validation.
 Finally, we demonstrate that \pkg{sdmTMB} is 4--16 times faster than \pkg{INLA} or \pkg{inlabru}.
-\pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling.
-We anticipate its application to the full spectrum geostatistical data will prove equally fruitful.
+\pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling (SDMs).
+We anticipate its application to the full spectrum of geostatistical data will prove equally fruitful.
 keywords:
@@
 
-Ecological data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
+Data are often collected in space or in space repeatedly over time.
+While such data are a rich source of information about mechanistic processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].
 <!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?.
@@
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
-One common application in the field of ecology is species distribution models (SDMs), hence the package name.
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: sdmTMB).
 This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

### 2022-11-02 doc/paper-jss/pcod.Rmd
Edits

@@
+\clearpage
+
 # Example: spatiotemporal modelling of Pacific cod
@@
 ```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
-library(ggplot2)
-library(dplyr)
-library(sdmTMB)
+library("ggplot2")
+library("dplyr")
+library("sdmTMB")
 theme_set(theme_minimal())
@@
 
-We will fit a geostatistical spatiotemporal model with sdmTMB for the purposes of index standardization.
+We will fit a geostatistical spatiotemporal model with \pkg{sdmTMB} for the purposes of index standardization.
 We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
@@
 Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year.
-There is a grid built into the sdmTMB package for Queen Charlotte Sound (named `qcs_grid`).
+There is a grid built into the \pkg{sdmTMB} package for Queen Charlotte Sound (named `qcs_grid`).
 Our prediction grid also needs to have all the covariates that we used in the model above.

### 2022-11-02 doc/sdmTMB-PeerJ.Rmd
Old stuff

@@
 
-(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
+(ref:diagram-fig) **Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.**
 An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
@@
 
-(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
+(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
 (A) SPDE mesh (lines) combined with the trawl survey observations (points).
@@
 
-(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
+(ref:owl-nao-fig) **Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.**
 Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).

### 2022-11-02 doc/paper-jss/owls.Rmd
Edits
More caching, edit abstract

@@
+\clearpage
+
 ```{r setup-owls, include = FALSE, cache=FALSE}
@@
 
-```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
+```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE}
 print(fit_owl2)
@@
 
-```{r tidy-owls, cache=FALSE}
+```{r tidy-owls}
 tidy(fit_owl2, conf.int = TRUE)
@@
 
-```{r theme-owls, cache=FALSE, echo=FALSE}
+```{r theme-owls, echo=FALSE}
 theme_set(theme_bw() + theme(
@@
 
-```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
+```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3}
 p1 <- ggplot(data = p_proj) +
@@
 
-```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, cache=FALSE, dpi=220}
+```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220}
 ggplot(data = p_proj) +

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
More caching, edit abstract

@@
     email: \email{sean.anderson@dfo-mpo.gc.ca}
-  - name: Eric J. Ward
-    orcid: 0000-0002-4359-0296
-    affiliation: 'Northwest Fisheries Science Center, National Marine Fisheries Service \AND'
-  - name: Philina A. English
-    orcid: 0000-0003-2992-6782
-    address: ''
-    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
-  - name: Lewis A. K. Barnett
-    orcid: 0000-0002-9381-8375
-    affiliation: 'Alaska Fisheries Science Center, National Marine Fisheries Service'
 bibliography: refs.bib
@@
 Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
-SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are possible.
-We describe the statistical model, illustrate the package through two case studies, and compare the functionality, speed, and interface to related software.
-The case studies illustrate \pkg{sdmTMB}'s flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
+SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or Bayesian inference via \pkg{tmbstan} and \pkg{rstan}).
+We describe the model and use case studies to illustrate sdmTMB's flexibility by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
 Additional functionality includes delta (hurdle) models, break-point effects, and out-of-sample cross validation.
-Finally, we demonstrate that \pkg{sdmTMB} is 4--16 times faster than \pkg{INLA} or \pkg{inlabru}.
-\pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling (SDMs).
-We anticipate its application to the full spectrum of geostatistical data will prove equally fruitful.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB is 4–16 times faster than INLA.
+We anticipate sdmTMB's accessible interface will help open this useful class of models to the full spectrum geostatistical analysts.
 keywords:

### 2022-11-02 doc/paper-jss/inla-comparison.Rmd
Edits
Keywords, 1st page, caching

@@
+\clearpage
+
 ```{r setup-inla-comparison, include = FALSE, cache=FALSE}
@@
 
-```{r set-priors, echo = TRUE, cache=FALSE}
+```{r set-priors, echo = TRUE}
 range_min <- 5
@@
 
-```{r sdm-mesh, echo = TRUE, cache=FALSE}
+```{r sdm-mesh, echo = TRUE}
 sdmTMB_mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 10)
@@
 
-```{r inla-mesh, echo = TRUE, cache=FALSE}
+```{r inla-mesh, echo = TRUE}
 loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
@@
 
-```{r maps, cache=FALSE, out.width="5.5in"}
+```{r maps, out.width="5.5in"}
 inverse_link <- m$family$linkinv

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Keywords, 1st page, caching

@@
     orcid: 0000-0001-9563-1937
-    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
+    affiliation: 'Pacific Biological Station'
     address: |
@@
     email: \email{sean.anderson@dfo-mpo.gc.ca}
+  - name: Eric J. Ward
+    orcid: 0000-0002-4359-0296
+    affiliation: 'Northwest Fisheries Science Center \AND'
+  - name: Philina A. English
+    orcid: 0000-0003-2992-6782
+    address: ''
+    affiliation: 'Pacific Biological Station'
+  - name: Lewis A. K. Barnett
+    orcid: 0000-0002-9381-8375
+    affiliation: 'Alaska Fisheries Science Center'
 bibliography: refs.bib
@@
   # at least one keyword must be supplied
-  formatted: [test]
-  plain:     [test]
+  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
+  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
 preamble: >
@@
 
+\clearpage 
 # Introduction

### 2022-11-02 doc/paper-jss/timing.Rmd
Keywords, 1st page, caching

@@
 
-```{r sim-dat-plot-timing, out.width="5in", cache=FALSE, echo=FALSE}
+```{r sim-dat-plot-timing, out.width="5in", echo=FALSE}
 ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Ref apps

@@
 
-The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix 1).
+The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
@@
 
-By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix 1).
+By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
-The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
+The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
 
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See Appendix 1 for a full description of the model.
+See Appendix \ref{app:model} for a full description of the model.
 Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} [@wood2017a], \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
@@
 An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).
-<!-- We could instead model catch weight as the response with an offset for log(area). -->
 
@@
 
-<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
 We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
-<!-- (omitted for space). -->
 
@@
 We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
-Here we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as an AR(1) process (`time_varying_type = "ar1"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
+Here we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
 
@@
   spatiotemporal = "iid",
-  time_varying_type = "ar1",
+  time_varying_type = "rw",
   anisotropy = TRUE,
@@
 
-This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop. 
+<!-- This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop.  -->
 We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
-We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix 2).
+We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
 
@@
 
-We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}, Appendix 3).
+We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
+We provide an extended version of this example in Appendix \ref{app:owls}.
 
@@
 We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
-\@ref(fig:timing), Appendix 6).
+\@ref(fig:timing), Appendix \ref{app:speed}).
 In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
@@
 How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
-Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
+Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
 We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
@@
 \pkg{VAST} is the sole option for fitting some multivariate models;
-alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
+alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).

### 2022-11-02 doc/paper-jss/owls.Rmd
Ref apps
saved optimized file instead of original

@@
 
-# Example: spatially varying effect on owl counts
+# Example: spatially varying effect on owl counts {#app:owls}
 
@@
 ```{r cache-negb2-2, echo=FALSE, eval=!rds_saved2}
-saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km2.rds"))
+saveRDS(fit_owl2, here::here("data/snow_w_main_effect_0_150km2.rds"))
 ```

### 2022-11-02 doc/paper-jss/inla-comparison.Rmd
Ref apps
Work with old inla

@@
 
-\section{Comparison between INLA, inlabru, and sdmTMB}
+# Comparison between INLA, inlabru, and sdmTMB {#app:inla}
 
@@
 ```{r inla-densities, echo=FALSE}
+old_inla <- any(grepl(
+  "year_factoryear_factor",
+  names(m_inla$marginals.fixed)
+))
+yrs <- seq(2007, 2017, 2)
+if (old_inla) {
+  yf <- "year_factoryear_factor"
+} else {
+  yf <- "year_factor"
+}
 x <- list()
-x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>%
+for (i in seq_along(yrs)) {
+  x[[i]] <- inla.smarginal(m_inla$marginals.fixed[[paste0(yf, yrs[i])]]) %>%
     as_tibble() %>%
-  mutate(variable = "as.factor.year.2007")
-x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>%
-  as_tibble() %>%
-  mutate(variable = "as.factor.year.2009")
-x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>%
-  as_tibble() %>%
-  mutate(variable = "as.factor.year.2011")
-x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>%
-  as_tibble() %>%
-  mutate(variable = "as.factor.year.2013")
-x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>%
-  as_tibble() %>%
-  mutate(variable = "as.factor.year.2015")
-x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>%
-  as_tibble() %>%
-  mutate(variable = "as.factor.year.2017")
+    mutate(variable = paste0("as.factor.year.", yrs[i]))
+}
 x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>%
@@
   levels = unique(post_inla1$variable),
-  labels = c(
-    "2007", "2009", "2011", "2013", "2015", "2017",
-    "Range", "Random Field SD", "AR1 rho"
-  )
+  labels = c(yrs, "Range", "Random Field SD", "AR1 rho")
 ))

### 2022-11-02 doc/paper-jss/owls.Rmd
Checkmarks and x's

@@
 ```
-#>   Non-linear minimizer suggests successful convergence
-#>   Hessian matrix is positive definite
-#>   No extreme or very small eigen values detected
-#> x `ln_tau_E` gradient > 0.001
+#> ✓ Non-linear minimizer suggests successful convergence
+#> ✓ Hessian matrix is positive definite
+#> ✓ No extreme or very small eigen values detected
+#> ✗ `ln_tau_E` gradient > 0.001
 #>   See `?run_extra_optimization()`
@@
 #> 
-#> x `ln_phi` gradient > 0.001
+#> ✗ `ln_phi` gradient > 0.001
 #>   See `?run_extra_optimization()`
@@
 #> 
-#>   No fixed-effect standard errors are NA
-#>   No fixed-effect standard errors look unreasonably large
-#>   No sigma parameters are < 0.01
-#>   No sigma parameters are > 100
-#>   Range parameter doesn't look unreasonably large
+#> ✓ No fixed-effect standard errors are NA
+#> ✓ No fixed-effect standard errors look unreasonably large
+#> ✓ No sigma parameters are < 0.01
+#> ✓ No sigma parameters are > 100
+#> ✓ Range parameter doesn't look unreasonably large
 ```
@@
 ```
-#> Non-linear minimizer suggests successful convergence
-#> Hessian matrix is positive definite
-#> No extreme or very small eigen values detected
-#> No gradients with respect to fixed effects are >= 0.001
-#> No fixed-effect standard errors are NA
-#> No fixed-effect standard errors look unreasonably large
-#> No sigma parameters are < 0.01
-#> No sigma parameters are > 100
-#> Range parameter doesn't look unreasonably large
+#> ✓ Non-linear minimizer suggests successful convergence
+#> ✓ Hessian matrix is positive definite
+#> ✓ No extreme or very small eigen values detected
+#> ✓ No gradients with respect to fixed effects are >= 0.001
+#> ✓ No fixed-effect standard errors are NA
+#> ✓ No fixed-effect standard errors look unreasonably large
+#> ✓ No sigma parameters are < 0.01
+#> ✓ No sigma parameters are > 100
+#> ✓ Range parameter doesn't look unreasonably large
 ```

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Checkmarks and x's

@@
   \usepackage{booktabs}
+  \usepackage{pifont}
+  \usepackage{newunicodechar}
+  \newunicodechar{✓}{\ding{51}}
+  \newunicodechar{✗}{\ding{55}}
 output: rticles::jss_article

### 2022-11-02 doc/paper-jss/owls.Rmd
Editing

--- 2022-11-02 doc/paper-jss/owls.Rmd
@@
 ```
-#>   Non-linear minimizer suggests successful convergence
-#>   Hessian matrix is positive definite
-#>   No extreme or very small eigen values detected
-#> x `ln_tau_E` gradient > 0.001
+#> ✓ Non-linear minimizer suggests successful convergence
+#> ✓ Hessian matrix is positive definite
+#> ✓ No extreme or very small eigenvalues detected
+#> ✗ `ln_tau_E` gradient > 0.001
 #>   See `?run_extra_optimization()`
@@
 ```{r owl-newton, results='hide', eval=!rds_saved2}
-fit_owl2 <- run_extra_optimization(fit_owl, nlminb_loops = 0, newton_loops = 1)
+fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
 ```
@@
 ```
-#> Non-linear minimizer suggests successful convergence
-#> Hessian matrix is positive definite
-#> No extreme or very small eigen values detected
-#> No gradients with respect to fixed effects are >= 0.001
-#> No fixed-effect standard errors are NA
-#> No fixed-effect standard errors look unreasonably large
-#> No sigma parameters are < 0.01
-#> No sigma parameters are > 100
-#> Range parameter doesn't look unreasonably large
+#> ✓ Non-linear minimizer suggests successful convergence
+#> ✓ Hessian matrix is positive definite
+#> ✓ No extreme or very small eigenvalues detected
+#> ✓ No gradients with respect to fixed effects are >= 0.001
+#> ✓ No fixed-effect standard errors are NA
+#> ✓ No fixed-effect standard errors look unreasonably large
+#> ✓ No sigma parameters are < 0.01
+#> ✓ No sigma parameters are > 100
+#> ✓ Range parameter doesn't look unreasonably large
 ```
@@
 
-We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the \pkg{DHARMa} package.
+We can also test for zero-inflation using the `simulate.sdmTMB()` function.
 If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.
@@
 ```{r zero-test-owls, echo=TRUE, message=FALSE}
-s_nb2 <- simulate(fit_owl2, nsim = 500)
+s_nb2 <- simulate(fit_owl2, nsim = 400)
+```
+
+The proportion of zeros looks approximately the same in the simulated and real data.
+
+```{r zero-test-owls2, echo=TRUE}
+mean(s_nb2 == 0)
+mean(snow$count == 0)
+```
+
+And we can more formally test this with functions from the \pkg{DHARMa} package.
+
+```{r zero-test-owls3, echo=TRUE}
 pred_fixed <- fit_owl2$family$linkinv(pred$est_non_rf)
@@
     mean_est_count = exp(mean(est)),
-    nao_effect = b[b$term == "nao", "estimate"] + mean(zeta_s),
+    nao_effect = b$estimate[b$term == "nao"] + mean(zeta_s),
     nao_sim = mean(nao_effect_sim),

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Editing

@@
   collapse = TRUE,
-  comment = "#>"
+  comment = "#>",
+  prompt = FALSE
+  # R.options = list(prompt = " ", continue = " ")
 )
@@
 See Appendix \ref{app:model} for a full description of the model.
-Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} [@wood2017a], \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
+Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
 \end{figure}
@@
 
-An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
 Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.
 
@@
 
-```{r sdmTMB-lib, eval=TRUE}
+```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
 library(sdmTMB)
+# simplify for paper:
+pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |> 
+  tibble::as_tibble()
 ```
 
-```{r, echo=TRUE, eval=FALSE}
+```{r sdmTMB-lib, eval=TRUE, echo=TRUE}
 library(sdmTMB)
-head(pcod)
+head(pcod, n = 3)
 ```
 
-```{}
-#>    year density depth     X     Y
-#>   <int>   <dbl> <dbl> <dbl> <dbl>
-#> 1  2003   113.    201  446. 5793.
-#> 2  2003    41.7   212  446. 5800.
-#> 3  2003     0     220  449. 5802.
+<!-- An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. -->
+Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `add_utm_columns()`].
+Here we use `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
+By default, the function will guess our UTM zone and create new columns `X` and `Y`.
+
+```{r pcod-utms-eval, echo=FALSE, eval=TRUE}
+pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 ```
 
-We start by creating a mesh object that contains matrices to apply the SPDE approach.
+```{r pcod-utms-echo, echo=TRUE, eval=FALSE}
+pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+#> Detected UTM zone 9N; CRS = 32609.
+#> Visit https://epsg.io/32609 to verify.
+```
+
+```{r pcod-head2, echo=TRUE}
+head(pcod, n = 3)
+```
+
+We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach.
 
@@
 Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
-Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to `make_mesh()`.
-We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).
+`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()`.
+Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
+We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \ref{fig:pcod-fig}A).
 
@@
 Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
-As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} (REF) or \pkg{broom.mixed} (REF) packages:
+The smoother summary follows the format used in the \pkg{brms} package [@brms].
+As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} [@broom] or \pkg{broom.mixed} [@broommixed] packages:
 
@@
 #> ✓ Hessian matrix is positive definite
-#> ✓ No extreme or very small eigen values detected
+#> ✓ No extreme or very small eigenvalues detected
 #> ✓ No gradients with respect to fixed effects are >= 0.001
@@
 
+```{r qcs-head, echo=TRUE}
+head(qcs_grid, n = 3)
+```
+
 ```{r pcod-eg1-predict, echo=TRUE, eval=TRUE, cache=TRUE}
@@
 
-The output of `predict()` is a data frame containing overall estimates in link space, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig.\ \ref{fig:pcod-fig}B--D.
+The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`), and the individual random field components (here `omega_s`---the spatial field).
+
+```{r pcod-p-tibble}
+p <- tibble::as_tibble(p)
+```
+
+```{r pcod-head-predict, echo=TRUE}
+head(p, n = 3)
+```
+
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \ref{fig:pcod-fig}B--D.
 
@@
 We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
-We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
+We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
+\ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
 
@@
 We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
-\@ref(fig:timing), Appendix \ref{app:speed}).
+\ref{fig:timing}, Appendix \ref{app:speed}).
 In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.

### 2022-11-02 doc/paper-jss/inla-comparison.Rmd
Streamline chunk settings

@@
 \clearpage
-
-```{r setup-inla-comparison, include = FALSE, cache=FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  fig.width = 7,
-  fig.asp = 0.618,
-  cache = TRUE,
-  autodep = TRUE,
-  warning = FALSE,
-  message = FALSE
-)
-```
 
@@
 ```{r inla-pred, out.width="5.5in"}
-idx <- inla.stack.index(sdat, "stdata")$data
-d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
-
-ggplot(d) +
-  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
-  facet_wrap(~year) +
-  scale_colour_viridis_c() +
-  coord_fixed()
+# idx <- inla.stack.index(sdat, "stdata")$data
+# d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
+# 
+# ggplot(d) +
+#   geom_point(aes(X, Y, colour = inla_pred), size = 2) +
+#   facet_wrap(~year) +
+#   scale_colour_viridis_c() +
+  # coord_fixed()
 ```

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Streamline chunk settings

@@
   comment = "#>",
-  prompt = FALSE
-  # R.options = list(prompt = " ", continue = " ")
+  prompt = FALSE,
+  cache = TRUE,
+  autodep = TRUE,
+  fig.width = 7,
+  fig.asp = 0.618,
+  cache.comments = FALSE,
+  dev = "png",
+  dpi = 180
+  # optipng = "-strip all"
+  # R.options = list(prompt = "R> ", continue = "> ")
 )
+# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 ```
@@
 
-\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to model parameters (`?\pkg{sdmTMB}priors`).
+\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
@@
 The results from an \pkg{sdmTMB} model can be used to generate various quantities of interest.
-For example, \pkg{sdmTMB} provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.
+For example, for the field of ecology, \pkg{sdmTMB} provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.
 
@@
 # simplify for paper:
+pcod_original <- pcod # save it for later
 pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |> 
@@
 ```{r pcod-eg1-suppress-depth2, eval=TRUE, echo=FALSE}
+qcs_grid_original <- qcs_grid
 qcs_grid$depth_scaled <- NULL # for simplicity in paper
@@
 
+```{r pcod-restore, echo=FALSE}
+pcod <- pcod_original
+qcs_grid <- qcs_grid_original
+```
+
 # Example of a spatially varying coefficient with citizen science data
@@
   count ~ nao + (1 | year_factor),
-  spatial_varying = ~ 0 + nao,
+  spatial_varying = ~ nao,
   family = nbinom2(link = "log"), 
@@
 
-Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
-These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
-We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users.
+<!-- Spatially and spatiotemporally explicit data are increasingly collected and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
+-->
+<!-- These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
+-->
+<!-- We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users. -->
 

### 2022-11-02 doc/paper-jss/pcod.Rmd
Ref apps
Streamline chunk settings

@@
 
-# Example: spatiotemporal modelling of Pacific cod
+# Example: spatiotemporal modelling of Pacific cod {#app:pcod}
 
@@
 knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  fig.width = 7,
-  fig.asp = 0.618,
-  cache = TRUE,
-  autodep = TRUE,
-  cache.comments = FALSE,
   echo = TRUE

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Table 1 links

@@
 Packages such as \pkg{VAST} [@thorson2019] and \pkg{inlabru} [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
-We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
+We provide a more detailed comparison of related software packages in Table \ref{tab:functionality} and the Discussion.
 
@@
 There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
-\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
+\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
 \pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
@@
 \pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
-There is considerable variability in the available observation likelihoods across packages (Table 1).
+There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
 We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} or \pkg{inlabru} (Appendix 4) and \pkg{VAST} (Appendix 5).
@@
 \begin{tabular}[t]{lllllll}
+\label{tab:functionality}
 \toprule
@@
 How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
-Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
+Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
 We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
@@
 \pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
-Penalized spline support was possible thanks to \pkg{mgcv} [@wood2017a].
+Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].
 We thank the authors of all these packages.

### 2022-11-03 doc/paper-jss/paper-jss.Rmd
Fix Table label

@@
 \begin{tabular}[t]{lllllll}
-\label{tab:functionality}
 \toprule
@@
 \end{threeparttable}
+\label{tab:functionality}
 \end{table}
@@
 
-```{r child="model-description.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```{r child="model-description.Rmd", eval=TRUE}
 ```
 
-```{r child="pcod.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```{r child="pcod.Rmd", eval=TRUE}
 ```
 
-```{r child="owls.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```{r child="owls.Rmd", eval=TRUE}
 ```
 
-```{r child="inla-comparison.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```{r child="inla-comparison.Rmd", eval=TRUE}
 ```
 
-```{r child="timing.Rmd", echo=FALSE, cache=FALSE, eval=TRUE}
+```{r child="timing.Rmd", eval=TRUE}
 ```

### 2022-11-03 doc/paper-jss/owls.Rmd
Streamline chunk settings
Added figure caption in the owl appendix

@@
 ```{r setup-owls, include = FALSE, cache=FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  fig.width = 7,
-  fig.asp = 0.7,
-  cache = TRUE,
-  autodep = TRUE,
-  cache.comments = FALSE,
-  dev = "png",
-  dpi = 150,
-  optipng = "-strip all"
-)
-knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+theme_set(theme_bw() + theme(
+  plot.title = element_text(size = 12),
+  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
+  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
+  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
+  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
+  axis.title = element_blank() # lat/lon is clear from axis text
+)) 
 ```
@@
 
-First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
+First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations (Fig.\ \ref{fig:mesh-owls}).
 The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
@@
 
-```{r mesh-owls, echo=TRUE, out.width="4in"}
+```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="Mesh used for model is based on X and Y that use an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-We can inspect randomized quantile residuals, although this method has known issues [@thygesen2017a].
+We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues [@thygesen2017a].
 
-```{r qqnorm-owls, out.width="60%"}
+```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='h', fig.align='center'}
 qqnorm(pred$resid)
@@
 We do not find evidence of zero inflation.
-
-```{r theme-owls, echo=FALSE}
-theme_set(theme_bw() + theme(
-  plot.title = element_text(size = 12),
-  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
-  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
-  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
-  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
-  axis.title = element_blank() # lat/lon is clear from axis text
-)) 
-```
 
@@
 Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`).
-All these effects are in link space (`link = "log"`) and could be exponentiated to give the expected multiplicative contribution to the predicted owl counts; however, the additive nature of the effect in link space is more intuitive to interpret visually, so this is what we will plot on our maps:
+All these effects are in link space (`link = "log"`) and could be exponentiated to give the expected multiplicative contribution to the predicted owl counts; however, the additive nature of the effect in link space is more intuitive to interpret visually, so this is what we will plot on our maps (Fig.\ \ref{fig:plot-map-ci}).
 
-```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%", echo=FALSE}
+```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%"}
 ggplot(data = p_proj) +
@@
   labs(colour = "Estimated effect in log space") +
-  ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
+  # ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
   theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
@@
 
-```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3}
+```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3, fig.cap="Map of combined main and spatially varying effects of NAO on Snowy Owl count with confidence intervals.
+In the west, the lower values span zero implying a no evidence of an effect at this confidence level, while in the southeast the lowest estimates are still positive.
+Note that the values are in log space and thus additive.
+Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='h', fig.align='center'}
 p1 <- ggplot(data = p_proj) +
@@
 
-These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
+For each year in the model, we can generate predictions of winter owl counts that incorporate all fixed effects and random effects (Fig.\ \ref{fig:plot-all-effects-owls}). 
 The mean estimated count is `r round(mean(exp(p$est)), 2)`, which is essentially equal to the observed mean of `r round(mean(snow$count), 2)`.
@@
 
-```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220}
+```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
 ggplot(data = p_proj) +
@@
 
-We can look at the spatial random effects (`omega_s`) that represent consistent deviations in space that are not accounted for by our fixed effects. 
+We can look at the spatial random effects (`omega_s`) that represent consistent deviations in space that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects-owls}). 
 In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
@@
 
-```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200}
+```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200, fig.cap="Spatial random effects that apply across all years.", fig.pos='h', fig.align='center'}
 ggplot(data = p_proj) +
@@
 
-Finally, we can look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
+We can also look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations (Fig.\ \ref{fig:plot-spatiotemporal-effects-owls}).
 These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
@@
 
-```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
+```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-In this case, these only include the overall effect of mean annual NAO and the random intercepts for year.
+Finally, the non-random field based elements in the model are represented by `est_non_rf` variable in the prediction dataframe, and in this case only include the overall effect of mean annual NAO and the random intercepts for year (Fig.\ \ref{fig:plot-fixed-effects-owls}).
 
-```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
+```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The facet colours represent the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-11-03 doc/paper-jss/paper-jss.Rmd
Added figure caption in the owl appendix

@@
   fig.asp = 0.618,
+  fig.pos = "ht",
   cache.comments = FALSE,

### 2022-11-03 doc/paper-jss/owls.Rmd
draft pcod captions

@@
 
-```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="Mesh used for model is based on X and Y that use an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
+```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The facet colours represent the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
+```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The colour of each year represents the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-11-03 doc/paper-jss/pcod.Rmd
draft pcod captions

@@
 
-For example, a mesh with a cutoff distance of 10 km can be made:
+For example, a mesh with a cutoff distance of 10 km can be made (Fig.\ \ref{fig:mesh}).
 
@@
 
-```{r plot-mesh, cache=TRUE, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE}
+```{r plot-mesh, cache=TRUE, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="This mesh uses a cutoff distance of 10 km.", fig.pos='h', fig.align='center'}
 plot(mesh)
@@
 
-We can inspect simulated residuals from the DHARMa package:
+We can inspect simulated residuals from the DHARMa package (Fig.\ \ref{fig:sim-resids}).
 
-```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE}
+```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE, fig.cap="Simulated residuals from the DHARMa package.", fig.pos='h', fig.align='center'}
 set.seed(123)
@@
 
-Or perhaps better, use MCMC residuals:
+Or perhaps better, use MCMC residuals (Fig.\ \ref{fig:mcmc-resids}).
 
@@
 
-```{r mcmc-resids-vis, out.width="4in"}
+```{r mcmc-resids-vis, out.width="4in", fig.cap="MCMC residuals.", fig.pos='h', fig.align='center'}
 qqnorm(r)
@@
 
-There are four kinds of predictions that we get out of the model. First we will show the predictions that incorporate all fixed effects and random effects:
+There are four kinds of predictions that we get out of the model. First we will show the predictions that incorporate all fixed effects and random effects (Fig.\ \ref{fig:plot-all-effects}).
 
-```{r plot-all-effects}
+```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='h', fig.align='center'}
 plot_map(pred_qcs, "exp(est)") +
-  scale_fill_viridis_c(trans = "log10") +
-  ggtitle("Prediction (fixed effects + all random effects)")
+  # ggtitle("Prediction (fixed effects + all random effects)") +
+  scale_fill_viridis_c(trans = "log10") 
 ```
 
-We can also look at just the fixed effects, here `year`:
+We can also look at just the fixed effects, here `year` (Fig.\ \ref{fig:plot-fix-defects}).
 
-```{r plot-fix-defects}
+```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='h', fig.align='center'}
 plot_map(pred_qcs, "exp(est_non_rf)") +
-  ggtitle("Prediction (fixed effects only)") +
+  # ggtitle("Prediction (fixed effects only)") +
   scale_fill_viridis_c(trans = "sqrt")
@@
 
-We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects.
+We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects}).
 In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.
 
-```{r plot-spatial-effects}
+```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='h', fig.align='center'}
 plot_map(pred_qcs, "omega_s") +
-  ggtitle("Spatial random effects only") +
+  # ggtitle("Spatial random effects only") +
   scale_fill_gradient2()
@@
 
-Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations.
+Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations (Fig.\ \ref{fig:plot-spatiotemporal-effects}).
 These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model.
 
-```{r plot-spatiotemporal-effects}
+```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='h', fig.align='center'}
 plot_map(pred_qcs, "epsilon_st") +
-  ggtitle("Spatiotemporal random effects only") +
+  # ggtitle("Spatiotemporal random effects only") +
   scale_fill_gradient2()
@@
 
-This returns a matrix with a row for each row of the `newdata` data frame and a column for each replicate draw. We can use this matrix to calculate the standard deviation or coefficient of variation:
+This returns a matrix with a row for each row of the `newdata` data frame and a column for each replicate draw.
+We can use this matrix to calculate the standard deviation or coefficient of variation (Fig.\ \ref{fig:sims-cv-plot}).
 
-```{r sims-cv-plot, cache=TRUE}
+```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='h', fig.align='center'}
 qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
@@
 
-We can calculate a standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE`.
+We can calculate a standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE` (Fig.\ \ref{fig:pcod-app-index}).
 This returns the \pkg{TMB} objective function, which is required to pass the output to `get_index()`.
 
-```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE}
+```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='h', fig.align='center'}
 qcs_grid$area <- 4 # all 2 x 2km

### 2022-11-03 doc/paper-jss/pcod.Rmd
fix fig refs

@@
 
-For example, a mesh with a cutoff distance of 10 km can be made (Fig.\ \ref{fig:mesh}).
+For example, a mesh with a cutoff distance of 10 km can be made (Fig.\ \ref{fig:plot-mesh}).
 
@@
 
-Or perhaps better, use MCMC residuals (Fig.\ \ref{fig:mcmc-resids}).
+Or perhaps better, use MCMC residuals (Fig.\ \ref{fig:mcmc-resids-vis}).
 

### 2022-11-03 doc/paper-jss/owls.Rmd
owl facet plot tweaks

@@
 
-```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
+```{r plot-all-effects-owls, out.width="85%", echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
 ggplot(data = p_proj) +
@@
   labs(colour = "Predicted counts") +
-  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
+  theme(axis.text = element_blank(), 
+  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
+  axis.ticks = element_blank()) +
   facet_wrap(~year_f)
@@
 
-```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
+```{r plot-spatiotemporal-effects-owls, out.width="85%", echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
   scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
-  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
+  theme(axis.text = element_blank(), 
+  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
+  axis.ticks = element_blank()) +
   facet_wrap(~year_f)
@@
 
-```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The colour of each year represents the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
+```{r plot-fixed-effects-owls, out.width="85%", echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements in this model.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
   ) +
-  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
+  theme(axis.text = element_blank(), 
+  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
+  axis.ticks = element_blank()) +
   facet_wrap(~year_f)

### 2022-11-03 doc/paper-jss/inla-comparison.Rmd
figure captions for inla appendix

@@
 
-Comparing the fixed effects shows the estimates from all three models are also very similar.
+Comparing the fixed effects shows the estimates from all three models are also very similar (Fig.\ \ref{fig:coef-plot}).
 
-```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE}
+```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='h', fig.align='center'}
 inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
@@
 We can then compare the distribution of these \pkg{sdmTMB} simulations to the marginal posterior densities from \pkg{INLA}
-From \pkg{INLA}, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the \pkg{sdmTMB} `gather_sims()` function.
+From \pkg{INLA} (Fig.\ \ref{fig:compare-plot}). 
+We do this by extracting the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the \pkg{sdmTMB} `gather_sims()` function.
 Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the \pkg{INLA} model posterior assuming we had set\
@@
 
-In this plot, the blue histograms are from \pkg{sdmTMB} and the density lines are from \pkg{INLA}:
+<!-- In this plot, the blue histograms are from \pkg{sdmTMB} and the density lines are from \pkg{INLA}: -->
 
-```{r compare-plot, out.width="6in"}
+```{r compare-plot, out.width="6in", fig.cap="Blue histograms represent simulated draws from the joint precision matrix of the sdmTMB model compared with the marginal posterior densities from the INLA model.", fig.align='center'}
 ggplot(post_inla2) +
@@
 
-We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space.
+We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
 For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:
@@
 
-```{r maps, out.width="5.5in"}
+```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the sdmTMB model.", fig.align='center'}
 inverse_link <- m$family$linkinv
@@
 
-We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:
+We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation (Fig.\ \ref{fig:inlabru-pred}).
 
-```{r inlabru-pred, out.width="5.5in"}
+```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the inlabru model.", fig.align='center'}
 nd_bru <- nd

### 2022-11-04 doc/paper-jss/paper-jss.Rmd
Fix citation
fix latex in fig 4 caption

@@
 $^4$Spatiotemporal fields as random walk only.
-$^5$SPDE approach as in Miller et al.
-(2019).
+$^5$SPDE approach as in \citet{miller2019}.
 $^6$Does have predictive process knots.
@@
 \caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
-Lines represent means and ribbons 95% quantiles across 50 random iterations.
+Lines represent means and ribbons 95\% quantiles across 50 random iterations.
 Note the log x and y axes.
 \pkg{VAST} is similar to \pkg{sdmTMB} and so is not shown.
-\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
-\pkg{mgcv} used \texttt{bam()}, \texttt{method = 'fREML'}, and discretized covariates [@miller2019].
+\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
+\pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates [@miller2019].
 Note that spaMM only fits spatial, not spatiotemporal, models.

### 2022-11-04 doc/paper-jss/timing.Rmd
Ref apps
captions for timing figures

@@
-# Speed testing related packages
+\clearpage
+# Speed testing related packages {#app:speed}
 
 Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
-We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions.
+
+We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
+We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
+We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
+In Figure 4, we report on the x-axis the number of mesh nodes ("knots") that result from each of these meshes.
+
+We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions (Fig.\ \ref{fig:sim-dat-plot-timing}).
 Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
@@
 
-We generated a mesh that was consistent across simulated data sets for a given mesh resolution.
-We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
-We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
-In Figure 4, we report on the x-axis the number of mesh nodes ("knots") that result from each of these meshes.
-
-We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.
+We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Fig.\ \ref{fig:timing}).
 The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.1.3 and the default Matrix library. The following package versions were used:
@@
 
-Here are the other meshes used from least to most vertices:
+<!-- Here are the other meshes used from least to most vertices: -->
 
-```{r mesh-vis-timing, out.width="6.5in", fig.width=10, fig.asp=1.2, echo=FALSE}
+```{r mesh-vis-timing, fig.width=7, fig.asp=1.1, echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='h', fig.align='center'}
 max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
@@
 ```
+
+\clearpage
 
@@
 
-Look at the simulated data:
+<!-- Look at the simulated data: -->
 
-```{r sim-dat-plot-timing, out.width="5in", echo=FALSE}
+```{r sim-dat-plot-timing, out.width="5in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='h', fig.align='center'}
 ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
@@
 ```
+
+\clearpage

### 2022-11-04 doc/paper-jss/inla-comparison.Rmd
captions for timing figures
Edits

@@
 spde <- inla.spde2.pcmatern(
-  mesh = inla_mesh, # an smdTMB mesh could be inserted here with `sdmTMB_mesh$mesh`
+  mesh = inla_mesh, 
   prior.range = c(range_min, prior_prob),

### 2022-11-04 doc/paper-jss/owls.Rmd
Edits

--- 2022-11-03 doc/paper-jss/owls.Rmd
@@
 
-We can run additional optimization steps using the `run_extra_optimization()` function.
+We can run additional optimization steps with a Newton optimizer using the `run_extra_optimization()` function to further reduce these gradients.
 
@@
 
-And confirm that the gradients are smaller:
+We can confirm that the gradients are smaller.
 
@@
 
-View the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`.
+View the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`:
 
@@
 ```
+
+`sigma_O` is the spatial random field SD, `sigma_E` is the spatiotemporal random field SD, `sigma_Z` is the spatially varying coefficient random field SD, and `sigma_G` is the random intercept SD.
 
@@
 
-We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues [@thygesen2017a].
+We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues when based on random effects derived with the Laplace approximation [@thygesen2017a].
 
@@
 
-More accurate residuals can be generated using \pkg{tmbstan}, but this a slower process.
+More accurate residuals can be generated using MCMC and \pkg{tmbstan}, but this a slower process.
 These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
-To calculate these, we have to use the version of our model fit without REML.
+To calculate these, we have to use a version of our model fit without REML, which we can quickly obtain with the `update()` method.
 We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
@@
 ```{r stan-resids-owls, eval=FALSE}
+fit_ml <- update(fit_owl, reml = FALSE)
 mcmc_resids <- residuals(fit_owl, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
@@
 
-We can also test for zero-inflation using the `simulate.sdmTMB()` function.
+We can test for zero-inflation using the `simulate.sdmTMB()` function.
 If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.
@@
 
-And we can more formally test this with functions from the \pkg{DHARMa} package.
+We can more formally test this with functions from the \pkg{DHARMa} package.
 
@@
 
-```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
+```{r plot-all-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
+```{r plot-spatiotemporal-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The colour of each year represents the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
+```{r plot-fixed-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='h', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-11-04 doc/paper-jss/timing.Rmd
Edits

@@
   observed ~ a1,
-  data = sim_dat, mesh = mesh_sdmTMB, family = gaussian(),
-  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2))
+  data = sim_dat, 
+  mesh = mesh_sdmTMB, 
+  family = gaussian(),
+  priors = sdmTMBpriors(
+    matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)
+  )
 )
@@
   observed ~ a1 + IMRF(1 | X + Y, model = spde),
-  family = gaussian(), data = sim_dat
+  family = gaussian(), 
+  data = sim_dat
 )
@@
   method = "fREML",
-  control = gam.control(scalePenalty = FALSE), discrete = TRUE
+  control = gam.control(scalePenalty = FALSE), 
+  discrete = TRUE
 )

### 2022-11-04 doc/paper-jss/owls.Rmd
Edits

@@
 
-```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
+```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='ht', fig.align='center'}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='h', fig.align='center'}
+```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='ht', fig.align='center'}
 qqnorm(pred$resid)
@@
 Note that the values are in log space and thus additive.
-Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='h', fig.align='center'}
+Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='ht', fig.align='center'}
 p1 <- ggplot(data = p_proj) +
@@
 
-```{r plot-all-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
+```{r plot-all-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200, fig.cap="Spatial random effects that apply across all years.", fig.pos='h', fig.align='center'}
+```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200, fig.cap="Spatial random effects that apply across all years.", fig.pos='ht', fig.align='center'}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatiotemporal-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
+```{r plot-spatiotemporal-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-```{r plot-fixed-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='h', fig.align='center'}
+```{r plot-fixed-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-11-04 doc/paper-jss/pcod.Rmd
Edits

@@
 
-```{r mesh, echo = TRUE, cache=TRUE}
+```{r plot-mesh, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="Delaunay triangulation mesh.", fig.pos='ht', fig.align='center'}
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
-```
-
-```{r plot-mesh, cache=TRUE, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="This mesh uses a cutoff distance of 10 km.", fig.pos='h', fig.align='center'}
 plot(mesh)
@@
 ```{r print-fit, echo=TRUE}
-print(fit)
+fit
 ```
@@
 
-```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE, fig.cap="Simulated residuals from the DHARMa package.", fig.pos='h', fig.align='center'}
+```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE, fig.cap="Simulated residuals from the DHARMa package.", fig.pos='ht', fig.align='center'}
 set.seed(123)
@@
 
-```{r mcmc-resids-vis, out.width="4in", fig.cap="MCMC residuals.", fig.pos='h', fig.align='center'}
+```{r mcmc-resids-vis, out.width="4in", fig.cap="MCMC residuals.", fig.pos='ht', fig.align='center'}
 qqnorm(r)
@@
 
-```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='h', fig.align='center'}
+```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, "exp(est)") +
@@
 
-```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='h', fig.align='center'}
+```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, "exp(est_non_rf)") +
@@
 
-```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='h', fig.align='center'}
+```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, "omega_s") +
@@
 
-```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='h', fig.align='center'}
+```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, "epsilon_st") +
@@
 
-```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='h', fig.align='center'}
+```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='ht', fig.align='center'}
 qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
@@
 
-```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='h', fig.align='center'}
+```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='ht', fig.align='center'}
 qcs_grid$area <- 4 # all 2 x 2km

### 2022-11-04 doc/paper-jss/owls.Rmd
style new resid chunk

--- 2022-11-04 doc/paper-jss/owls.Rmd
@@
 fit_ml <- update(fit_owl, reml = FALSE)
-mcmc_resids <- residuals(fit_owl, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
+mcmc_resids <- residuals(fit_owl,
+  type = "mle-mcmc",
+  mcmc_iter = 101,
+  mcmc_warmup = 100
+)
 ```

### 2022-11-04 doc/paper-jss/inla-comparison.Rmd

@@
 
-```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='h', fig.align='center'}
+```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='ht', fig.align='center'}
 inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%

### 2022-11-04 doc/paper-jss/paper-jss.Rmd
Edits

@@
 abstract: >
-  Geostatistical data are pervasive and available in increasingly large volumes.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement.
+  Geostatistical data are pervasive and available in increasingly large datasets; however, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement.
 Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
 SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or Bayesian inference via \pkg{tmbstan} and \pkg{rstan}).
-We describe the model and use case studies to illustrate sdmTMB's flexibility by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
-Additional functionality includes delta (hurdle) models, break-point effects, and out-of-sample cross validation.
-Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB is 4–16 times faster than INLA.
-We anticipate sdmTMB's accessible interface will help open this useful class of models to the full spectrum geostatistical analysts.
+We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
+Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than INLA.
+We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
 keywords:
@@
 
-```{r, setup, include=FALSE}
+```{r main-setup, include=FALSE}
 options(prompt = 'R> ', continue = '+ ')
@@
   fig.pos = "ht",
-  cache.comments = FALSE,
+  cache.comments = TRUE,
   dev = "png",
@@
 Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].
-<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?.
-More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit?
--->
 
@@
 
-```{r, include=FALSE}
-# simplify for paper
-# pcod <- dplyr::select(sdmTMB::pcod, year, density, depth, X, Y)
-```
-
 ```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
@@
 
-We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
+We can get a summary of the fit with the `print()` or `summary()` methods.
 
@@
 
-<!-- This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop.  -->
 We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
@@
 (5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
-<!-- Simulations can also be passed to the \proglang{R} package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
-
-<!-- We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary \pkg{TMB} object (Fig.
-\@ref(fig:pcod-fig)E).
-`get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations.
--->
-
-```{r, eval=FALSE, echo=FALSE}
-p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
-ind <- get_index(p)
-```
 
@@
 To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
-<!-- However, \pkg{VAST} is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
 Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.

### 2022-11-04 doc/paper-jss/timing.Rmd

@@
 
-<!-- Here are the other meshes used from least to most vertices: -->
-
-```{r mesh-vis-timing, fig.width=7, fig.asp=1.1, echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='h', fig.align='center'}
+```{r mesh-vis-timing, fig.width=7, fig.asp=1.1, echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'}
 max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
@@
 set.seed(1)
-
-n_obs <- 1000L
+n_obs <- 1000
 predictor_dat <- data.frame(
@@
 
-```{r sim-dat-plot-timing, out.width="5in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='h', fig.align='center'}
+```{r sim-dat-plot-timing, out.width="5in", echo=FALSE, fig.cap="Example simulated dataset with a spatial random field.", fig.pos='ht', fig.align='center'}
 ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
@@
 ```{r mgcvfit-timing, echo=TRUE, eval=FALSE}
-# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth():
+# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth()
 # from supplement of:
@@
 # https://doi.org/10.1007/s13253-019-00377-z
+
 fit_bam <- bam(

### 2022-11-04 doc/paper-jss/paper-jss.Rmd
Edit

@@
 Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
-SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or Bayesian inference via \pkg{tmbstan} and \pkg{rstan}).
+SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
 We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
 Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
-Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than INLA.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}.
 We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
@@
   # at least one keyword must be supplied
-  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
-  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
+  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
+  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
 preamble: >

### 2022-11-04 doc/paper-jss/paper-jss.Rmd
Edit

@@
     address: |
-      | Pacific Biological Station, Nanaimo, BC, Canada
+      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
     email: \email{sean.anderson@dfo-mpo.gc.ca}
@@
     affiliation: 'Northwest Fisheries Science Center \AND'
+    address: |
+      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
   - name: Philina A. English
     orcid: 0000-0003-2992-6782
-    address: ''
     affiliation: 'Pacific Biological Station'
+    address: |
+      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
   - name: Lewis A. K. Barnett
@@
     affiliation: 'Alaska Fisheries Science Center'
+    address: |
+      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
 bibliography: refs.bib
@@
 abstract: >
-  Geostatistical data are pervasive and available in increasingly large datasets; however, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement.
+  Geostatistical data are common across scientific fields and available in increasingly large datasets.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement for many users.
 Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
@@
 
-# Example of a spatially varying coefficient with citizen science data
+# Spatially varying coefficient example with citizen science data
 

### 2022-11-05 doc/paper-jss/owls.Rmd
Edits

@@
 \clearpage
+
+```{r owl-knitr-setup, include=FALSE}
+knitr::opts_chunk$set(
+  dpi = 125,
+  dev = "png",
+  optipng = "-strip all"
+)
+knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+```
 
@@
 
-```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='ht', fig.align='center'}
+```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='ht', fig.align='center', dev='png', dpi=100}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='ht', fig.align='center'}
+```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='ht', fig.align='center', dpi=90, dev='png'}
 qqnorm(pred$resid)

### 2022-11-05 doc/paper-jss/paper-jss.Rmd
Edits

@@
   \newunicodechar{✗}{\ding{55}}
+  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
 output: rticles::jss_article
@@
   cache.comments = TRUE,
-  dev = "png",
-  dpi = 180
-  # optipng = "-strip all"
+  dev = "pdf",
+  dpi = 140,
+  optipng = "-strip all"
   # R.options = list(prompt = "R> ", continue = "> ")
 )
-# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 ```
@@
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
-We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} or \pkg{inlabru} (Appendix 4) and \pkg{VAST} (Appendix 5).
+We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
+<!-- and \pkg{VAST} (Appendix 5). -->
 
@@
 \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
-\pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates [@miller2019].
+\pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates \citep{miller2019}.
 Note that spaMM only fits spatial, not spatiotemporal, models.
@@
 
-Speed-wise, \pkg{sdmTMB} (and by association \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
+Speed-wise, \pkg{sdmTMB} (and due to the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.

### 2022-11-11 doc/paper-jss/owls.Rmd
minor line edit

@@
 lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)# only large lakes
+# keep only large lakes
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
 ```

### 2022-11-14 doc/paper-jss/model-description.Rmd
Edits
Ref apps
Edit

@@
-# Model
+\clearpage
+
+# Model description {#app:model}
 
@@
 
-Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$)
+Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$).
 

### 2022-11-14 doc/paper-jss/paper-jss.Rmd
Edti

@@
   \DeclareGraphicsExtensions{.pdf,.png,.jpg}
+  \widowpenalty10000
+  \clubpenalty10000
 output: rticles::jss_article
@@
 Data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information about mechanistic processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
+While such data are a rich source of information [e.g., @legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
@@
 This can greatly improve computational efficiency [e.g., @heaton2019].
-The SPDE approach is a solution popularized via the \pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \pkg{INLA} to create input matrices [e.g., @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}\-pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
 Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].
 
-Systems for specifying statistical models that can include the SPDE, such as \pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
+Systems for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
 For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
@@
 
-Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \pkg{INLA} to set up SPDE matrices.
+Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices.
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
@@
 The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
+The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance is directionally dependent) [@haskard2007].
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).
@@
 We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
-The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
+The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].
 
@@
 Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
-After rapid model exploration with maximum likelihood, one can optionally pass an \pkg{sdmTMB} model to the \proglang{R} package \pkg{tmbstan} [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.
+After rapid model exploration with maximum likelihood, one can optionally pass an \pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} [@carpenter2017; @rstan] via \pkg{tmbstan} [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.
 
@@
 <!-- An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. -->
-Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `add_utm_columns()`].
-Here we use `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
+Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
+Here we use the helper function `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
 By default, the function will guess our UTM zone and create new columns `X` and `Y`.
@@
 `make_mesh()` is a wrapper function for `INLA::inla.mesh.create()`.
-Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
+Alternatively, we could have created any mesh via the \proglang{R}-\pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
 We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \ref{fig:pcod-fig}A).
@@
 Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
-The smoother summary follows the format used in the \pkg{brms} package [@brms].
+The smoother summary follows the format used in the \pkg{brms} package [@burkner2017].
 As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} [@broom] or \pkg{broom.mixed} [@broommixed] packages:
@@
 We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
-Here, we will predict on a 2x2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
+Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
 The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
@@
 
-The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`), and the individual random field components (here `omega_s`---the spatial field).
+The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`) and the individual random field components (here `omega_s`---the spatial field).
 
@@
 
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \ref{fig:pcod-fig}B--D.
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig.\ \ref{fig:pcod-fig}B--D.
 
@@
 We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
-We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.
-\ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
+We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
 
@@
 Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
-(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
+(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model. 
+Models can be fit to these simulated data to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
 
@@
 There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
-\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
-\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
-\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
+\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
+\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
+\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
 To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
@@
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
-We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
+We provide comparisons of the syntax and the reproducibility of results from models fit using \proglang{R}-\pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
 <!-- and \pkg{VAST} (Appendix 5). -->
@@
 \toprule
- & \pkg{sdmTMB} & \pkg{VAST} & \pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
+ & \pkg{sdmTMB} & \pkg{VAST} & \proglang{R}-\pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
 \midrule
@@
 $^2$Penalized smoother GAMs that determine `wiggliness'.
-$^3$\pkg{inlabru} but not \pkg{INLA}.
+$^3$\pkg{inlabru} but not \proglang{R}-\pkg{INLA}.
 $^4$Spatiotemporal fields as random walk only.
@@
 If so, does the parallelism share memory across the cores or does the memory grow with the number of cores?
-This is a HUGE feature of PARDISO within \pkg{INLA}/\pkg{inlabru} allowing for very large models to be fit on HPCs.
+This is a HUGE feature of PARDISO within \proglang{R}-\pkg{INLA}/\pkg{inlabru} allowing for very large models to be fit on HPCs.
 It might be worth mentioning and adding to Table 1.
@@
 
-We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
+We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
 \ref{fig:timing}, Appendix \ref{app:speed}).
-In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
+In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
 \pkg{mgcv} was most affected by mesh resolution.
@@
 Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
-We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
-Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
-Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
+We think users familiar with `stats::glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
+Users familiar with \proglang{R}-\pkg{INLA} will find \pkg{inlabru} approachable.
+Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
 \pkg{VAST} is the sole option for fitting some multivariate models;
@@
 
-Speed-wise, \pkg{sdmTMB} (and due to the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
+Speed-wise, \pkg{sdmTMB} (and due to the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
-Furthermore, although both \pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores.
+However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
+<!-- Furthermore, although both \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \proglang{R}-\pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores.
+-->
 
@@
 
-\pkg{inlabru} is similar to \pkg{sdmTMB} in design, but differs fundamentally in that fitting is achieved via \pkg{INLA} and approximate Bayesian computation.
-For some models, TMB/\pkg{sdmTMB} can be considerably faster than \pkg{INLA}/\pkg{inlabru}, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig.
+\pkg{inlabru} is similar to \pkg{sdmTMB} in design, but differs fundamentally in that fitting is achieved via \proglang{R}-\pkg{INLA} and approximate Bayesian computation.
+For some models, TMB/\pkg{sdmTMB} can be considerably faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru}, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig.
 \@ref(fig:timing), @osgood-zimmerman2021].
-One major capability included in \pkg{sdmTMB} but not \pkg{INLA}/\pkg{inlabru} is anisotropy.
+One major capability included in \pkg{sdmTMB} but not \proglang{R}-\pkg{INLA}/\pkg{inlabru} is anisotropy.
 Furthermore, we think the \pkg{sdmTMB} interface will be more familiar than \pkg{inlabru} to users of `glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB}.
-On the other hand, \pkg{inlabru} fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with \pkg{INLA}.
+On the other hand, \pkg{inlabru} fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with \proglang{R}-\pkg{INLA}.
 
@@
 The included \pkg{TMB} `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
-<!-- Not sure what kind of refs you want here?  \pkg{INLA}'s non-gaussian random fields are mentioned here @bakka2018review-->
+<!-- Not sure what kind of refs you want here?  \proglang{R}-\pkg{INLA}'s non-gaussian random fields are mentioned here @bakka2018review-->
 <!-- \pkg{sdmTMB} does not currently include  -->
@@
 
-\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
+\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \proglang{R}-\pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
 \pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.

### 2022-11-14 doc/paper-jss/paper-jss.Rmd
More ref fixes

@@
 Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
-Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
+Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2022].
 These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.

### 2022-11-14 doc/paper-jss/owls.Rmd
Small fixes to lake data import

--- 2022-11-05 doc/paper-jss/owls.Rmd
@@
 ```
-{
+
 Next we retrieve continental coastlines from the package \pkg{rnaturalearth} and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.
 
-```{r shapes}
+```{r shapes, echo=FALSE}
 if (!file.exists(here::here("data/ne_10m_lakes"))) {
@@
 }
+```
+
+```{r shapes-read, echo=TRUE}
 coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
@@
 lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)# only large lakes
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
 ```

### 2022-11-15 doc/paper-jss/paper-jss.Rmd
Edits

@@
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].
+Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself [e.g., @thorson2019d; @barnett2021].
 
@@
 This can greatly improve computational efficiency [e.g., @heaton2019].
-The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}\-pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
 Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

### 2022-11-15 doc/paper-jss/owls.Rmd
Reduce overall file size

@@
 
-```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%"}
+```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="65%", dpi=140, dev="png"}
 ggplot(data = p_proj) +
@@
 
-```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3, fig.cap="Map of combined main and spatially varying effects of NAO on Snowy Owl count with confidence intervals.
+```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="80%",fig.asp=0.3, fig.cap="Map of combined main and spatially varying effects of NAO on Snowy Owl count with confidence intervals.
 In the west, the lower values span zero implying a no evidence of an effect at this confidence level, while in the southeast the lowest estimates are still positive.
 Note that the values are in log space and thus additive.
-Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='ht', fig.align='center'}
+Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='ht', fig.align='center', dpi=140, dev="png"}
 p1 <- ggplot(data = p_proj) +
@@
 
-```{r plot-all-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
+```{r plot-all-effects-owls, out.width="90%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=140, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='ht', fig.align='center', dev="png"}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200, fig.cap="Spatial random effects that apply across all years.", fig.pos='ht', fig.align='center'}
+```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="3in", dpi=130, fig.cap="Spatial random effects that apply across all years.", fig.pos='ht', fig.align='center', dev = "png",}
 ggplot(data = p_proj) +
@@
 
-```{r plot-spatiotemporal-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center'}
+```{r plot-spatiotemporal-effects-owls, out.width="75%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=120, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center', dev="png"}
 ggplot(data = filter(p_proj, year > 1978)) +
@@
 
-Finally, the non-random field based elements in the model are represented by `est_non_rf` variable in the prediction dataframe, and in this case only include the overall effect of mean annual NAO and the random intercepts for year (Fig.\ \ref{fig:plot-fixed-effects-owls}).
+Finally, the non-random field based elements in the model are represented by `est_non_rf` variable in the prediction data frame, and in this case only include the overall effect of mean annual NAO and the random intercepts for year (Fig.\ \ref{fig:plot-fixed-effects-owls}).
 
-```{r plot-fixed-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
+```{r plot-fixed-effects-owls, out.width="75%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=120, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='ht', fig.align='center', dev="png"}
 ggplot(data = filter(p_proj, year > 1978)) +

### 2022-11-15 doc/paper-jss/paper-jss.Rmd
Reduce overall file size

@@
 
-The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
+The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}; see the model description vignette in `browseVignettes("sdmTMB")` or [rendered on the web](https://pbs-assess.github.io/sdmTMB/articles/)).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
@@
 
-```{r child="model-description.Rmd", eval=TRUE}
+```{r child="model-description.Rmd", eval=FALSE}
 ```

### 2022-11-15 doc/paper-jss/paper-jss.Rmd
Link to vignette

--- 2022-11-15 doc/paper-jss/paper-jss.Rmd
@@
 
-The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
+The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
 Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
+See the the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
 
@@
 
-By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
+By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}).
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`).
 The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See Appendix \ref{app:model} for a full description of the model.
+See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
 Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}

### 2022-11-16 doc/paper-jss/paper-jss.Rmd
Copy edits

@@
 Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`).
+We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
 The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].
@@
 Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
-`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()`.
+`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` and `cutoff` is passed through to it.
 Alternatively, we could have created any mesh via the \proglang{R}-\pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.

### 2023-01-24 doc/appendix-pcod-VAST-poisson-link.Rmd
Updating paper

@@
 # get coordinates in geographic coordinates from UTM projection, after changing units to m
-qcs_grid_ll <- qcs_grid
+qcs_grid_ll <- replicate(qcs_grid, "year", unique(pcod$year))
 qcs_grid_ll$Y <- qcs_grid_ll$Y * 1000
@@
 ```{r sdmTMB-index, cache=TRUE, echo=TRUE}
+survey_grid <- replicate_df(qcs_grid, "year", unique(pcod$year))
 tictoc::tic()
-p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
-ind <- get_index(p, bias_correct = FALSE, area = rep(4, nrow(qcs_grid)))
+p <- predict(fit_sdmTMB, newdata = survey_grid, return_tmb_object = TRUE)
+ind <- get_index(p, bias_correct = FALSE, area = 4)
 tictoc::toc()

### 2023-01-24 doc/appendix-pcod-VAST-tweedie.Rmd
Print sdmTMB model
Updating paper

@@
   time = "year", spatiotemporal = "iid", spatial = "on")
+```
+
+```{r sdmTMB-example-print, echo = TRUE}
+fit_sdmTMB
 ```
@@
 ```{r sdmTMB-index, cache=TRUE, echo=TRUE}
-pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
+survey_grid <- replicate_df(qcs_grid, "year", unique(pcod$year))
+pred <- predict(fit_sdmTMB, newdata = survey_grid, return_tmb_object = TRUE)
 ind_sdmTMB <- get_index(pred, area = 4)

### 2023-01-24 doc/paper-jss/inla-comparison.Rmd
Edits
Updating paper

@@
 \clearpage
+
+```{r inla-knitr-setup, include=FALSE}
+knitr::opts_chunk$set(
+  dev = "pdf"
+)
+```
 
@@
 ```{r sdm-predictions}
-nd <- filter(qcs_grid, year >= 2007)
+nd <- replicate_df(qcs_grid, time_name = "year", time_values = unique(d$year))
 pred <- predict(m, newdata = nd)

### 2023-01-24 doc/paper-jss/owls.Rmd
Updating paper

@@
 
-```{r rds1, echo=FALSE, cache=FALSE}
-rds_saved <- file.exists(here::here("data/snow_w_main_effect_0_150km.rds"))
-rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_0_150km2.rds"))
-```
-
-```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
+```{r fit-negb2, echo=TRUE, warning=FALSE}
 fit_owl <- sdmTMB(count ~ 1 + nao + (1 | year_f),
@@
 )
-```
-
-```{r cache-negb2, echo=FALSE, eval=!rds_saved}
-saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km.rds"))
-```
-
-```{r read-negb2, echo=FALSE, eval=rds_saved}
-fit_owl <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
 ```
@@
 
-```{r owl-newton, results='hide', eval=!rds_saved2}
+```{r owl-newton, results='hide'}
 fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
-```
-
-```{r cache-negb2-2, echo=FALSE, eval=!rds_saved2}
-saveRDS(fit_owl2, here::here("data/snow_w_main_effect_0_150km2.rds"))
-```
-
-```{r read-negb2-2, echo=FALSE, eval=rds_saved2}
-fit_owl2 <- readRDS(here::here("data/snow_w_main_effect_0_150km2.rds"))
 ```

### 2023-01-24 doc/paper-jss/paper-jss.Rmd
Updating paper

@@
 
-Speed-wise, \pkg{sdmTMB} (and due to the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
+With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
@@
 Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
-Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
-The included \pkg{TMB} `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
-<!-- Not sure what kind of refs you want here?  \proglang{R}-\pkg{INLA}'s non-gaussian random fields are mentioned here @bakka2018review-->
-<!-- \pkg{sdmTMB} does not currently include  -->
-<!-- We may introduce explicit delta models, but many such models can already be fit with \pkg{sdmTMB} by modelling the two components separately. -->
-<!-- We may introduce capability for multivariate response data, but \pkg{VAST} already focuses on such models with a \pkg{TMB} backend. -->
-<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
-<!-- Line 275: I would suggest adding a MAJOR benefit of \pkg{sdmTMB} for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates!
-This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it.
-It saves me a bunch of time.
-If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it.
--->
-
-<!-- Spatially and spatiotemporally explicit data are increasingly collected and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
--->
-<!-- These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
--->
-<!-- We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users. -->
+Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous-time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
+The included \pkg{TMB} `.cpp` file provides a high-quality and tested model template that can be modified to add additional features.
 
@@
 We thank the authors of all these packages.
-We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
+We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
 

### 2023-01-24 doc/paper-jss/pcod.Rmd
Updating paper

@@
 In this case, that is a `year` column as well as `X` and `Y`.
+We will replicate the internal spatial grid over all years (calling it `survey_grid`) and predict on it.
 
 ```{r predict-newdata}
-pred_qcs <- predict(fit, newdata = qcs_grid)
+survey_grid <- replicate_df(
+  qcs_grid, 
+  time_name = "year", 
+  time_values = unique(pcod$year)
+)
+
+pred_qcs <- predict(fit, newdata = survey_grid)
 ```
@@
 ```{r pcod-sims, cache=TRUE}
-pred_sims <- predict(fit, newdata = qcs_grid, nsim = 200)
+pred_sims <- predict(fit, newdata = survey_grid, nsim = 200)
 dim(pred_sims)
@@
 ```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='ht', fig.align='center'}
-qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
-ggplot(qcs_grid, aes(X, Y, fill = cv)) +
+survey_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
+ggplot(survey_grid, aes(X, Y, fill = cv)) +
   geom_raster() +
@@
 ```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='ht', fig.align='center'}
-qcs_grid$area <- 4 # all 2 x 2km
+survey_grid$area <- 4 # all 2 x 2km
 pred2 <- predict(
   fit,
-  newdata = qcs_grid, 
+  newdata = survey_grid, 
   return_tmb_object = TRUE
 )
-ind <- get_index(pred2, area = qcs_grid$area, bias_correct = TRUE)
+ind <- get_index(pred2, area = survey_grid$area, bias_correct = TRUE)
 

### 2023-01-24 doc/paper-jss/timing.Rmd
Edits
Updating paper

@@
 \clearpage
+
 # Speed testing related packages {#app:speed}
@@
 ```{r sim-timing, echo=TRUE}
-set.seed(1)
+set.seed(123)
 n_obs <- 1000
@@
 
-```{r mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE, cache=TRUE}
+```{r mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE}
 # -------------------------------------------------------------------------
@@
 
-```{r mgcvfit-timing, echo=TRUE, eval=FALSE}
+```{r mgcvfit-timing, echo=TRUE, eval=TRUE}
 # define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth()

### 2023-01-24 analysis/pcod-fig.R

@@ -63,7 +63,7 @@ fit <- sdmTMB(
   spatial = "on", silent = FALSE
 )
 print(fit)
-p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
+p <- predict(fit, newdata = qcs_grid)
 
 nd <- data.frame(
   depth =
@@ -77,8 +77,9 @@ fit_spatiotemporal <- sdmTMB(
   family = tweedie(link = "log"), data = pcod, mesh = mesh,
   time = "year", spatial = "on", spatiotemporal = "iid", silent = FALSE
 )
+survey_grid <- replicate_df(qcs_grid, "year", unique(pcod$year))
 p_tv <- predict(fit_spatiotemporal,
-  newdata = qcs_grid, return_tmb_object = TRUE
+  newdata = survey_grid, return_tmb_object = TRUE
 )
 ind <- get_index(p_tv, area = 4, bias_correct = TRUE)
 cog <- get_cog(p_tv, format = "wide", area = 4)

### 2023-01-24 doc/paper-jss/owls.Rmd
Prep for paper submission

@@
 
-```{r fit-negb2, echo=TRUE, warning=FALSE}
+```{r fit-negb2, echo=TRUE, warning=FALSE, cache=TRUE, results='hide'}
 fit_owl <- sdmTMB(count ~ 1 + nao + (1 | year_f),
@@
 
-```{r owl-newton, results='hide'}
+```{r owl-newton, results='hide', cache=TRUE}
 fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
@@
 
-```{r predict-owls, message=FALSE}
+```{r predict-owls, message=FALSE, cache=TRUE}
 pred <- predict(fit_owl2)

### 2023-01-24 doc/paper-jss/paper-jss.Rmd
Prep for paper submission

@@
 
+\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>:
+
+```{r sdmTMB-install, eval=FALSE, echo=TRUE}
+install.packages("sdmTMB", dependencies = TRUE)
+```
+
 Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
@@
 ```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
-library(sdmTMB)
+library("sdmTMB")
 # simplify for paper:
@@
 ```{r sdmTMB-lib, eval=TRUE, echo=TRUE}
-library(sdmTMB)
+library("sdmTMB")
 head(pcod, n = 3)
@@
 
-```{r child="pcod.Rmd", eval=TRUE}
+```{r child="pcod.Rmd", eval=TRUE, cache=TRUE}
 ```
 
-```{r child="owls.Rmd", eval=TRUE}
+```{r child="owls.Rmd", eval=TRUE, cache=TRUE}
 ```
 
-```{r child="inla-comparison.Rmd", eval=TRUE}
+```{r child="inla-comparison.Rmd", eval=TRUE, cache=TRUE}
 ```
 
-```{r child="timing.Rmd", eval=TRUE}
+```{r child="timing.Rmd", eval=TRUE, cache=TRUE}
 ```

### 2023-01-24 doc/paper-jss/make-reprex-jss.R

@@ -1,0 +1,34 @@
+knitr::purl("paper-jss.Rmd", documentation = 1L)
+
+# fix this:
+
+
+if (!file.exists(here::here("data/ne_10m_lakes"))) {
+  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
+    "download/10m/physical/ne_10m_lakes.zip")
+  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
+  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+}
+
+
+## ----shapes-read, echo=TRUE-------------------------------------------------------------------------------------------------
+coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
+  sf::st_transform(crs = Albers)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
+
+
+
+# trash this:
+
+## ----compare-table, results='asis', eval=FALS
+# ....
+
+# put this in root folder:
+
+snow <- readRDS(here::here("data/SNOW_data.rds"))
+
+
+# look for all here::here
+
+# add timing onto bottom

### 2023-01-24 doc/paper-jss/inla-comparison.Rmd
Copy editing

@@
 We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
-\pkg{INLA} requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:
+Since \pkg{INLA} requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:
 
@@
 
-We will start by building identical meshes using both sdmTMB and INLA.
+We will start by building identical meshes using both \pkg{sdmTMB} and \pkg{INLA}.
 Here, X and Y are coordinates in UTM zone 9.
-These same functions are used internally in the sdmTMB `make_mesh()` function.
+These same functions are used internally in the \pkg{sdmTMB} `make_mesh()` function.
 
@@
 
-Here, the meshes should be the same.
+These meshes should be the same.
 
@@
 Data preparation for \pkg{INLA} models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
-We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB.
+We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with \pkg{TMB}.
 We use weak priors on the factor fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.
@@
 
-With sdmTMB, we can extract simulated draws from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.
+With \pkg{sdmTMB}, we can extract simulated draws from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.
 
@@
 
-```{r compare-plot, out.width="6in", fig.cap="Blue histograms represent simulated draws from the joint precision matrix of the sdmTMB model compared with the marginal posterior densities from the INLA model.", fig.align='center'}
+```{r compare-plot, out.width="6in", fig.cap="Blue histograms represent simulated draws from the joint precision matrix of the \\pkg{sdmTMB} model compared with the marginal posterior densities from the \\pkg{INLA} model.", fig.align='center'}
 ggplot(post_inla2) +
@@
 We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
-For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:
+For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`---the inverse logit here), to produce occurrence probabilities:
 
@@
 
-We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation (Fig.\ \ref{fig:inlabru-pred}).
+We can get similar predictions with \pkg{inlabru} using a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation (Fig.\ \ref{fig:inlabru-pred}).
 
@@
 ```{r inla-pred, out.width="5.5in"}
-# idx <- inla.stack.index(sdat, "stdata")$data
-# d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
-# 
-# ggplot(d) +
-#   geom_point(aes(X, Y, colour = inla_pred), size = 2) +
-#   facet_wrap(~year) +
-#   scale_colour_viridis_c() +
-#   coord_fixed()
+idx <- inla.stack.index(sdat, "stdata")$data
+d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
+
+ggplot(d) +
+  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
+  facet_wrap(~year) +
+  scale_colour_viridis_c() +
+  coord_fixed()
 ```

### 2023-01-24 doc/paper-jss/paper-jss.Rmd
Copy editing

@@
 SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
-We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations).
+We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
 Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
-Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
 We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
@@
   # at least one keyword must be supplied
-  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
-  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
+  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
+  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
 preamble: >
@@
 
-Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data [@rue2005gmrf; @diggle2007; @cressie2011].
+Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects are a class of models appropriate for these data [@rue2005gmrf; @diggle2007; @cressie2011].
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
@@
 This can greatly improve computational efficiency [e.g., @heaton2019].
-The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
 Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].
@@
 Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices.
-Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
-One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: sdmTMB).
+Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a] for a specific class of spatial and spatiotemporal models.
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}).
 This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
@@
 \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
-Geostatistical data simply refers to data observed at specific spatial coordinates reflecting some underlying spatial process [@rossi1992; @diggle2007].
+Geostatistical data refers to data observed at specific spatial coordinates reflecting some underlying spatial process [@rossi1992; @diggle2007].
 These data can be collected across discrete points in time.
-Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models [e.g., @verhoef2018].
+Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models [e.g., @verhoef2018];
 \pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
@@
 The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
+In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} [@liaw2002], \pkg{MaxEnt} [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
 The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
-The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance is directionally dependent) [@haskard2007].
+The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) [@haskard2007].
 The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).
@@
 By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}).
-Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
+Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
 We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
@@
 
-We can get a summary of the fit with the `print()` or `summary()` methods.
+We can get a summary of the fit with the `print()` method:
 
 ```{r pcod-eg1-summary,eval=TRUE, echo=TRUE}
-summary(fit)
+fit
 ```
@@
 The smoother summary follows the format used in the \pkg{brms} package [@burkner2017].
-As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} [@broom] or \pkg{broom.mixed} [@broommixed] packages:
+As noted in the print method, we can extract a tidy data frame similar to the format in the \pkg{broom} [@broom] and \pkg{broom.mixed} [@broommixed] packages:
 
@@
 ```
+
+The latter `tidy()` call returns all variance components of the model and omits the standard errors since these are calculated in log space.
 
@@
 We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
-Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
+Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially.
 The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
@@
 
-The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`) and the individual random field components (here `omega_s`---the spatial field).
+The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (`est_non_rf`; here, intercept and depth), estimates from the random field components combined (`est_rf`), and estimates from the individual random field components (here, `omega_s`---the spatial field).
 
@@
 
-We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
-Here we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
+We could extend this to a spatiotemporal model simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
+Here, we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
+
+\clearpage
 
@@
   time_varying = ~ 1,
+  time_varying_type = "rw",
   family = tweedie(link = "log"),
@@
   spatiotemporal = "iid",
-  time_varying_type = "rw",
   anisotropy = TRUE,
@@
 ```{r fit-tv-print, eval=TRUE, echo=TRUE}
-summary(fit_spatiotemporal)
+fit_spatiotemporal
 ```
@@
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
-Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
+We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
 
@@
 
+\clearpage
+
 ```{r owl-nao, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
@@
 For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
-A version that uses Markov chain Monte Carlo (MCMC) to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
+A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
 Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
@@
 
+\clearpage
+
 # Package comparisons
@@
 There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
-\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
+\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
+In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} since it is a prominent package that can fit related predictive-process models without the SPDE.
 \pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
@@
 Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
-\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
+\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
-We provide comparisons of the syntax and the reproducibility of results from models fit using \proglang{R}-\pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
+We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix \ref{app:inla}.
 <!-- and \pkg{VAST} (Appendix 5). -->
@@
 
-\begin{figure}
+\begin{figure}[htb]
 \centering
-\includegraphics[width=3.5in]{../../figs/timing-spatial-m1-april11}
+\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-01-24-xkcd.pdf}
 \caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
 Lines represent means and ribbons 95\% quantiles across 50 random iterations.
-Note the log x and y axes.
-\pkg{VAST} is similar to \pkg{sdmTMB} and so is not shown.
+Note the log-distributed x axis.
+\pkg{VAST} should be similar to \pkg{sdmTMB} and so is not shown.
 \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
 \pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates \citep{miller2019}.
-Note that spaMM only fits spatial, not spatiotemporal, models.
+Note that \pkg{spaMM} only fits spatial, not spatiotemporal, models.
 All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}\label{fig:timing}
@@
 \pkg{VAST} is the sole option for fitting some multivariate models;
-alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
+alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
@@
 
-With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
+With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 6--24 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
@@
 
-Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
+Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model [@bakka2019], time-varying spatiotemporal covariance parameters [@ward2022], and simulation from the parameter joint precision matrix.
 Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous-time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
-The included \pkg{TMB} `.cpp` file provides a high-quality and tested model template that can be modified to add additional features.
+The included \pkg{TMB} `.cpp` file provides a tested model template that can be modified to add additional features.
 
@@
 We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
+Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.
 

### 2023-01-24 doc/paper-jss/pcod.Rmd
Copy editing

@@
 
-We will fit a geostatistical spatiotemporal model with \pkg{sdmTMB} for the purposes of index standardization.
+We will fit a geostatistical spatiotemporal model with \pkg{sdmTMB} for the purposes of area-weighted biomass index standardization.
 We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
 The density units are kg/km^2^.
-Here, X and Y are UTM zone 9 coordinates in kilometres.
+Here, X and Y are UTM zone 9 coordinates in kilometers.
 If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region or using the built-in `sdmTMB::add_utm_columns()` function.
@@
 plot_map(pred_qcs, "exp(est)") +
-  # ggtitle("Prediction (fixed effects + all random effects)") +
   scale_fill_viridis_c(trans = "log10") 
@@
 plot_map(pred_qcs, "exp(est_non_rf)") +
-  # ggtitle("Prediction (fixed effects only)") +
   scale_fill_viridis_c(trans = "sqrt")
@@
 We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects}).
-In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.
+In other words, these deviations represent consistent biotic and abiotic factors that are affecting spatial correlation in biomass density but are not accounted for in the model.
 
@@
 plot_map(pred_qcs, "omega_s") +
-  # ggtitle("Spatial random effects only") +
   scale_fill_gradient2()
@@
 plot_map(pred_qcs, "epsilon_st") +
-  # ggtitle("Spatiotemporal random effects only") +
   scale_fill_gradient2()
@@
 Leaving `bias_correct = FALSE` is fine for experimentation, but will be biased due to the nonlinear (`exp()`) transformation of the random effects.
-Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but takes more time and memory than skipping the bias correction.
+Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this [@thorson2016bias], but takes more time and memory than skipping the bias correction.

### 2023-01-24 doc/paper-jss/timing.Rmd
Copy editing

@@
 Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
-
 We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
@@
 We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
-In Figure 4, we report on the x-axis the number of mesh nodes ("knots") that result from each of these meshes.
+In Figure 4, we report on the x-axis the number of mesh nodes ("knots" or "vertices") that result from each of these meshes.
 
@@
 We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Fig.\ \ref{fig:timing}).
-The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.1.3 and the default Matrix library. The following package versions were used:
+The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library. The following package versions were used:
 

### 2023-01-24 analysis/timing.R
Tweak timing fig

@@ -294,15 +294,17 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
 to_run <- expand.grid(
   n_obs = c(1000L),
   max.edge = c(0.06, 0.1, 0.2),
-  iter = seq_len(1L)
+  iter = seq_len(3L)
 )
 
-# full run:
-# to_run <- expand.grid(
-#   n_obs = c(1000L),
-#   max.edge = c(0.06, 0.075, 0.1, 0.15, 0.2),
-#   iter = seq_len(50L)
-# )
+# full (slow) run to match paper:
+if (FALSE) {
+  to_run <- expand.grid(
+    n_obs = c(1000L),
+    max.edge = c(0.06, 0.075, 0.1, 0.15, 0.2),
+    iter = seq_len(50L)
+  )
+}
 
 to_run$seed <- to_run$iter * 29212
 nrow(to_run)
@@ -315,8 +317,8 @@ system.time({
   )
 })
 
-saveRDS(out, file = "analysis/timing-cache-parallel-2022-11-01.rds")
-out <- readRDS("analysis/timing-cache-parallel-2022-11-01.rds")
+saveRDS(out, file = "analysis/timing-cache-parallel-2023-01-24.rds")
+out <- readRDS("analysis/timing-cache-parallel-2023-01-24.rds")
 
 out_long <- tidyr::pivot_longer(
   out,
@@ -357,15 +359,46 @@ leg <- tribble(
   "mgcv::bam\n(discretize = TRUE) SPDE", 249, 20.5
 )
 
+# log-space version:
+
+# g <- out_long_sum %>%
+#   ggplot(aes(mean_mesh_n, time, colour = model_clean)) +
+#   geom_ribbon(aes(ymin = lwr, ymax = upr, fill = model_clean),
+#     alpha = 0.2,
+#     colour = NA
+#   ) +
+#   geom_line(lwd = 0.7) +
+#   # scale_y_log10() +
+#   scale_x_log10(breaks = c(250, 500, 1000)) +
+#   ggsidekick::theme_sleek() +
+#   theme(panel.grid.major = element_line(colour = "grey90")) +
+#   theme(legend.position = c(0.4, 0.85)) +
+#   geom_text(aes(label = model_clean), data = leg, hjust = 0) +
+#   scale_fill_manual(values = colour_vect) +
+#   scale_colour_manual(values = colour_vect) +
[…]

### 2023-01-24 doc/paper-jss/make-reprex-jss.R
Add to 'spin' notes

@@ -31,4 +31,27 @@ snow <- readRDS(here::here("data/SNOW_data.rds"))
 
 # look for all here::here
 
-# add timing onto bottom
+
+
+knitr::spin("pcod-fig.R")
+knitr::spin("timing.R")
+
+
+
+
+# add timing script onto bottom
+
+# go change:
+saveRDS(out, file = "analysis/timing-cache-parallel-2022-11-01.rds")
+out <- readRDS("analysis/timing-cache-parallel-2022-11-01.rds")
+
+# at bottom of added timing
+
+# add sessionInfo()  at end!
+
+# remove all child chunks
+# remove knitr::opts lines except first?
+# remove   optipng = "-strip all"
+# remove knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+
+# add pcod-fig.R stuff to bottom

### 2023-01-24 doc/paper-jss/paper-jss.R
Commit main .R reprex code

@@ -1,0 +1,1763 @@
+## ----main-setup, include=FALSE----------------------------------------------------------------------------------------------
+options(prompt = 'R> ', continue = '+ ')
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  collapse = TRUE,
+  comment = "#>",
+  prompt = FALSE,
+  cache = TRUE,
+  autodep = TRUE,
+  fig.width = 7,
+  fig.asp = 0.618,
+  fig.pos = "ht",
+  cache.comments = TRUE,
+  dev = "pdf",
+  dpi = 140
+  # optipng = "-strip all"
+  # R.options = list(prompt = "R> ", continue = "> ")
+)
+# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+
+
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-------------------------------------------------------------------------------------
+library(sdmTMB)
+# simplify for paper:
+pcod_original <- pcod # save it for later
+pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
+  tibble::as_tibble()
+
+
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE---------------------------------------------------------------------------------------
+library(sdmTMB)
+head(pcod, n = 3)
+
+
+## ----pcod-utms-eval, echo=FALSE, eval=TRUE----------------------------------------------------------------------------------
+pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+
+
+## ----pcod-utms-echo, echo=TRUE, eval=FALSE----------------------------------------------------------------------------------
+#> pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+#> #> Detected UTM zone 9N; CRS = 32609.
+#> #> Visit https://epsg.io/32609 to verify.
+
+
+## ----pcod-head2, echo=TRUE--------------------------------------------------------------------------------------------------
+head(pcod, n = 3)
+
+
+## ---- echo=TRUE, eval=TRUE, cache=TRUE--------------------------------------------------------------------------------------
+mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
+
+
+## ----pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE-------------------------------------------
+fit <- sdmTMB(
[…]

### 2023-01-24 analysis/pcod-fig.R
Add sessionInfo()

@@ -215,3 +215,5 @@ g <- wrap_plots(g_mesh, g_omega, g_pred, g3) + plot_layout(design = layout) +
 
 print(g)
 ggsave("figs/pcod-fig.pdf", width = 9, height = 7)
+
+sessionInfo()

### 2023-01-24 doc/paper-jss/make-reprex-jss.R
Auto update paper-jss.R

@@ -1,3 +1,5 @@
+setwd("doc/paper-jss/")
+
 knitr::purl("paper-jss.Rmd", documentation = 1L)
 
 # fix this:

### 2023-01-24 doc/paper-jss/paper-jss.R

@@ -1,4 +1,4 @@
-## ----main-setup, include=FALSE----------------------------------------------------------------------------------------------
+## ----main-setup, include=FALSE--------------------------------
 options(prompt = 'R> ', continue = '+ ')
 knitr::opts_chunk$set(
   echo = FALSE,
@@ -14,45 +14,49 @@ knitr::opts_chunk$set(
   fig.pos = "ht",
   cache.comments = TRUE,
   dev = "pdf",
-  dpi = 140
-  # optipng = "-strip all"
+  dpi = 140,
+  optipng = "-strip all"
   # R.options = list(prompt = "R> ", continue = "> ")
 )
-# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 
 
-## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-------------------------------------------------------------------------------------
-library(sdmTMB)
+## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------------
+## install.packages("sdmTMB", dependencies = TRUE)
+
+
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-----------------------
+library("sdmTMB")
 # simplify for paper:
 pcod_original <- pcod # save it for later
 pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
   tibble::as_tibble()
 
 
-## ----sdmTMB-lib, eval=TRUE, echo=TRUE---------------------------------------------------------------------------------------
-library(sdmTMB)
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE-------------------------
+library("sdmTMB")
 head(pcod, n = 3)
 
 
-## ----pcod-utms-eval, echo=FALSE, eval=TRUE----------------------------------------------------------------------------------
+## ----pcod-utms-eval, echo=FALSE, eval=TRUE--------------------
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 
 
-## ----pcod-utms-echo, echo=TRUE, eval=FALSE----------------------------------------------------------------------------------
-#> pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
-#> #> Detected UTM zone 9N; CRS = 32609.
-#> #> Visit https://epsg.io/32609 to verify.
[…]

### 2023-01-25 doc/paper-jss/owls.Rmd
Develop reprex R code

@@
 
-This appendix illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
+This appendix illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using \pkg{sdmTMB}. 
 We will use counts of Snowy Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts [CBCs, @cbc]. 
-Each survey is conducted between December 14 to January 5 in 24.14 kilometre diameter circle. 
+Each survey is conducted between December 14 to January 5 in a 24.14 kilometer (15 mile) diameter circle. 
 Circle locations are chosen by local volunteers and therefore reflect both locations of higher human densities and known bird watching locations.
@@
 
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
+Snowy Owls breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
 The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
@@
 
-We initially fit this model with NB1 and NB2 negative binomial observation error and Poisson error.
-According to AIC (`AIC()`), the `nbinom2()` (NB2) distribution has the most parsimonious fit.
+We initially fit this model with NB1 and NB2 [@hilbe2011] negative binomial observation error and Poisson error.
+According to AIC (`AIC()`), the `nbinom2()` family has the most parsimonious fit.
 We will therefore proceed with that model.
@@
 
-We note that some log-likelihood gradients are slightly large for some fixed effects.
+We note that the magnitude of some log-likelihood gradients is slightly large for some fixed effects.
 
@@
 
-We can run additional optimization steps with a Newton optimizer using the `run_extra_optimization()` function to further reduce these gradients.
+We can run additional optimization steps with a Newton optimizer using the\
+`run_extra_optimization()` function to further reduce these gradients.
 
@@
 
-View the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`:
+We can view the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`:
 
@@
 
-We can then generate predictions and residuals for the original data locations:
+We can then generate predictions at the original data locations:
 
@@
 pred <- predict(fit_owl2)
-set.seed(19208)
-pred$resid <- residuals(fit_owl2)
+
 ```
@@
 We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues when based on random effects derived with the Laplace approximation [@thygesen2017a].
+
+```{r owls-resids}
+set.seed(19208)
+pred$resid <- residuals(fit_owl2)
+```
 
@@
 More accurate residuals can be generated using MCMC and \pkg{tmbstan}, but this a slower process.
-These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
+The MCMC-based residuals are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
 To calculate these, we have to use a version of our model fit without REML, which we can quickly obtain with the `update()` method.
 We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
-For the purposes of this building this appendix quickly, we will not run this code.
+For the purposes of building this appendix quickly, we will not run this code.
 
@@
 ```{r shapes, echo=FALSE}
-if (!file.exists(here::here("data/ne_10m_lakes"))) {
+if (!file.exists(here::here("data/ne_10m_lakes.rds"))) {
   zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
@@
   download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
-  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes.rds"))
 }
@@
   sf::st_transform(crs = Albers)
-lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes.rds"), quiet = TRUE)
 lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
@@
 To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `nsim = 200`) and what component we want estimated (`sims_var = "zeta_s"`).
-We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `nsim = 200`). 
+We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function. 
 The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws (in this case we will use 80%) and adding these values to our prediction data frame.
@@
 
-If we recalculated these at 95% confidence intervals levels the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that for the most conventional level of statistical confidence, we cannot conclude that the effect of NAO at an annual scale differs from 0.
+If we recalculated these at 95% confidence intervals levels the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 0.
 However, the southeast to northwest gradient in the intensity illustrated in the maps above is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

### 2023-01-25 doc/paper-jss/make-reprex-jss.R

@@ -1,59 +1,103 @@
+setwd(here::here())
 setwd("doc/paper-jss/")
 
 knitr::purl("paper-jss.Rmd", documentation = 1L)
 
-# fix this:
+system("mv paper-jss.R reprex/paper-jss.R")
+
+d <- readLines("reprex/paper-jss.R")
+d[grepl("## ----child=", d)]
+d <- d[!grepl("## ----child=", d)]
 
 
-if (!file.exists(here::here("data/ne_10m_lakes"))) {
-  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
-    "download/10m/physical/ne_10m_lakes.zip")
-  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
-  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+do_replace <- function(find, replacement) {
+  cat("found:\n")
+  i <- grep(find, d)
+  print(d[i])
+  d[grepl(find, d)] <- replacement
+  cat("after replacement:\n")
+  print(d[i])
+  d
 }
 
+d[grep("dpi = 140", d)]
+d <- do_replace("dpi = 140", "  dpi = 140")
 
-## ----shapes-read, echo=TRUE-------------------------------------------------------------------------------------------------
-coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
-  sf::st_transform(crs = Albers)
-lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
+d[grep(" cache = TRUE,", d)]
+d <- do_replace(" cache = TRUE,", " # cache = TRUE,")
+d <- do_replace(" autodep = TRUE,", " # autodep = TRUE,")
 
+d[grep("-strip all", d)]
+d <- do_replace("-strip all", "  # optipng = \"-strip all\"")
 
+d[grep("optipng = knitr", d)]
+d <- do_replace("optipng = knitr", "# knitr::knit_hooks$set(optipng = knitr::hook_optipng)")
 
-# trash this:
+i <- grep("## ----compare-table", d)
+j <- grep("## ----setup-pcod", d)
+d <- d[-seq(i, j-2)]
 
[…]

### 2023-01-25 doc/paper-jss/reprex/paper-code.R

@@ -1,4 +1,4 @@
-## ----main-setup, include=FALSE--------------------------------
+## ----main-setup, include=FALSE----------------------------------------------------------------------------
 options(prompt = 'R> ', continue = '+ ')
 knitr::opts_chunk$set(
   echo = FALSE,
@@ -7,25 +7,25 @@ knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>",
   prompt = FALSE,
-  cache = TRUE,
-  autodep = TRUE,
+ # cache = TRUE,
+ # autodep = TRUE,
   fig.width = 7,
   fig.asp = 0.618,
   fig.pos = "ht",
   cache.comments = TRUE,
   dev = "pdf",
-  dpi = 140,
-  optipng = "-strip all"
+  dpi = 140
+  # optipng = "-strip all"
   # R.options = list(prompt = "R> ", continue = "> ")
 )
-knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE----------------------------------------------------------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-----------------------
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-------------------------------------------------------------------
 library("sdmTMB")
 # simplify for paper:
 pcod_original <- pcod # save it for later
@@ -33,30 +33,30 @@ pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
   tibble::as_tibble()
 
 
-## ----sdmTMB-lib, eval=TRUE, echo=TRUE-------------------------
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE---------------------------------------------------------------------
 library("sdmTMB")
 head(pcod, n = 3)
 
 
-## ----pcod-utms-eval, echo=FALSE, eval=TRUE--------------------
+## ----pcod-utms-eval, echo=FALSE, eval=TRUE----------------------------------------------------------------
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 
 
-## ----pcod-utms-echo, echo=TRUE, eval=FALSE--------------------
[…]

### 2023-01-25 doc/paper-jss/reprex/pcod-fig.R

@@ -1,0 +1,219 @@
+library("sdmTMB")
+library("dplyr")
+library("ggplot2")
+library("sf")
+library("patchwork")
+dir.create("figs", showWarnings = FALSE)
+
+if (!require("rnaturalearthhires", quietly = TRUE)) {
+  stop(
+    "Please install 'rnaturalearthhires'.\n",
+    "`remotes::install_github('ropensci/rnaturalearthhires')`"
+  )
+}
+
+# from remotes::install_github('seananderson/ggsidekick')
+theme_sleek <- function(base_size = 11, base_family = "") {
+  half_line <- base_size / 2
+  theme_light(base_size = base_size, base_family = base_family) +
+    theme(
+      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+      axis.ticks.length = grid::unit(half_line / 2.2, "pt"), strip.background = element_rect(
+        fill = NA,
+        colour = NA
+      ), strip.text.x = element_text(colour = "grey30"),
+      strip.text.y = element_text(colour = "grey30"), axis.text = element_text(colour = "grey30"),
+      axis.title = element_text(colour = "grey30"), legend.title = element_text(
+        colour = "grey30",
+        size = rel(0.9)
+      ), panel.border = element_rect(
+        fill = NA,
+        colour = "grey70", size = 1
+      ), legend.key.size = grid::unit(
+        0.9,
+        "lines"
+      ), legend.text = element_text(
+        size = rel(0.7),
+        colour = "grey30"
+      ), legend.key = element_rect(
+        colour = NA,
+        fill = NA
+      ), legend.background = element_rect(
+        colour = NA,
+        fill = NA
+      ), plot.title = element_text(
+        colour = "grey30",
+        size = rel(1)
+      ), plot.subtitle = element_text(
+        colour = "grey30",
+        size = rel(0.85)
+      )
+    )
+}
+theme_set(theme_sleek())
+
+# Pacific cod example (Figure 2) ------------------------------------------
+
[…]

### 2023-01-25 doc/paper-jss/reprex/timing.R

@@ -1,0 +1,425 @@
+# Timing figure -----------------------------------------------------------
+
+library("sdmTMB")
+library("dplyr")
+library("ggplot2")
+library("INLA")
+library("inlabru")
+library("mgcv")
+library("spaMM")
+dir.create("figs", showWarnings = FALSE)
+dir.create("analysis", showWarnings = FALSE)
+
+INLA::inla.setOption(num.threads = "1:1")
+
+# source(here::here("analysis/mgcv_spde_smooth.R"))
+# -------------------------------------------------------------------------
+# Code in this chunk is from:
+#
+# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
+# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
+# https://doi.org/10.1007/s13253-019-00377-z
+#
+# Re-used here under a Creative Commons Attribution 4.0 International License
+# http://creativecommons.org/licenses/by/4.0/
+smooth.construct.spde.smooth.spec <- function(object, data, knots) {
+  dim <- length(object$term)
+  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
+  if (dim == 1) {
+    x <- data[[object$term]]
+  } else {
+    x <- matrix(0, nr = length(data[[1]]), nc = 2)
+    x[, 1] <- data[[object$term[1]]]
+    x[, 2] <- data[[object$term[2]]]
+  }
+  if (is.null(object$xt)) {
+    if (dim == 1) {
+      t <- seq(min(x), max(x), len = object$bs.dim)
+      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
+    } else {
+      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
+    }
+  } else {
+    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
+    mesh <- object$xt$mesh
+  }
+  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
+  inlamats <- INLA::inla.mesh.fem(mesh)
+  object$S <- list()
[…]

### 2023-01-25 doc/paper-jss/owls.Rmd
Tweaks to reprex

@@
 ```{r shapes, echo=FALSE}
-if (!file.exists(here::here("data/ne_10m_lakes.rds"))) {
+if (!file.exists(here::here("data/ne_10m_lakes"))) {
   zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
@@
   download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
-  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes.rds"))
+  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
 }
@@
   sf::st_transform(crs = Albers)
-lakes <- sf::st_read(here::here("data/ne_10m_lakes.rds"), quiet = TRUE)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
 lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)

### 2023-01-25 doc/paper-jss/make-reprex-jss.R

@@ -55,14 +55,14 @@ d[grep("data/ne_10m_lakes", d)]
 d <- gsub("data/ne_10m_lakes", "ne_10m_lakes", d)
 
 d[grep("lakes <- sf::st_read", d)]
-d <- do_replace("lakes <- sf::st_read", "lakes <- sf::st_read(\"ne_10m_lakes.rds\", quiet = TRUE)")
+d <- do_replace("lakes <- sf::st_read", "lakes <- sf::st_read(\"ne_10m_lakes\", quiet = TRUE)")
 
-d[grep("file.exists\\(here::here\\(\"ne_10m_lakes\\.rds\"", d)]
-d <- do_replace("file.exists\\(here::here\\(\"ne_10m_lakes\\.rds\"", "if (!file.exists(\"ne_10m_lakes.rds\")) {")
+d[grep("file.exists\\(here::here\\(\"ne_10m_lakes\\\"", d)]
+d <- do_replace("file.exists\\(here::here\\(\"ne_10m_lakes\\\"", "if (!file.exists(\"ne_10m_lakes\")) {")
 
 d[grep("here::here\\(\"ne_10m_lakes.zip\"\\)", d)]
 d <- gsub("here::here\\(\"ne_10m_lakes.zip\"\\)", "\"ne_10m_lakes.zip\"", d)
-d <- gsub("here::here\\(\"ne_10m_lakes.rds\"\\)", "\"ne_10m_lakes.rds\"", d)
+d <- gsub("here::here\\(\"ne_10m_lakes\"\\)", "\"ne_10m_lakes\"", d)
 
 d <- c(d, "sessionInfo()")
 
@@ -85,13 +85,15 @@ system("cp ~/src/sdmTMB-paper/data/SNOW_data.rds ~/src/sdmTMB-paper/doc/paper-js
 system("rm reprex/paper-jss.R")
 
 setwd("reprex")
-knitr::spin("paper-code.R")
+system("R -e 'knitr::spin(\"paper-code.R\")'")
 
 d <- readLines(here::here("analysis/timing.R"))
 writeLines(d, "timing.R")
+system("R -e 'knitr::spin(\"timing.R\")'")
 
 d <- readLines(here::here("analysis/pcod-fig.R"))
 writeLines(d, "pcod-fig.R")
+system("R -e 'knitr::spin(\"pcod-fig.R\")'")
 
 setwd(here::here())
 

### 2023-01-25 doc/paper-jss/reprex/paper-code.R

@@ -436,11 +436,11 @@ p_proj <- p %>% mutate(x = X, y = Y) %>%
 
 
 ## ----shapes, echo=FALSE-----------------------------------------------------------------------------------
-if (!file.exists("ne_10m_lakes.rds")) {
+if (!file.exists("ne_10m_lakes")) {
   zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
     "download/10m/physical/ne_10m_lakes.zip")
   download.file(zip_file, destfile = "ne_10m_lakes.zip")
-  unzip("ne_10m_lakes.zip", exdir = "ne_10m_lakes.rds")
+  unzip("ne_10m_lakes.zip", exdir = "ne_10m_lakes")
 }
 
 

### 2023-01-25 doc/paper-jss/make-reprex-jss.R
Update reprex

@@ -1,4 +1,4 @@
-RUN_SPIN <- FALSE
+RUN_SPIN <- TRUE
 
 setwd(here::here())
 setwd("doc/paper-jss/")
@@ -80,28 +80,21 @@ d <- c(d[seq(1, i-1)], "\n# Pacific Cod appendix -------------------------------
 i <- grep("## ----inla-knitr-setup", d)
 d <- c(d[seq(1, i-1)], "\n# INLA comparison appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])
 
-writeLines(d, "reprex/paper-code.R")
+writeLines(d, "reprex/replication-code-main.R")
 
 system("cp ~/src/sdmTMB-paper/data/SNOW_data.rds ~/src/sdmTMB-paper/doc/paper-jss/reprex/snow-data.rds")
 
 system("rm reprex/paper-jss.R")
 
 setwd("reprex")
-if (RUN_SPIN) system("R -e 'knitr::spin(\"paper-code.R\")'")
+if (RUN_SPIN) system("R -e 'knitr::spin(\"replication-code-main.R\")'")
 
 d <- readLines(here::here("analysis/timing.R"))
-writeLines(d, "timing.R")
-if (RUN_SPIN) system("R -e 'knitr::spin(\"timing.R\")'")
+writeLines(d, "replication-code-timing.R")
+if (RUN_SPIN) system("R -e 'knitr::spin(\"replication-code-timing.R\")'")
 
 d <- readLines(here::here("analysis/pcod-fig.R"))
-writeLines(d, "pcod-fig.R")
-if (RUN_SPIN) system("R -e 'knitr::spin(\"pcod-fig.R\")'")
+writeLines(d, "replication-code-pcod.R")
+if (RUN_SPIN) system("R -e 'knitr::spin(\"replication-code-pcod.R\")'")
 
 setwd(here::here())
-
-# # go change:
-# saveRDS(out, file = "analysis/timing-cache-parallel-2022-11-01.rds")
-# out <- readRDS("analysis/timing-cache-parallel-2022-11-01.rds")
-#
-#
-# # add pcod-fig.R stuff to bottom

### 2023-01-25 doc/paper-jss/reprex/replication-code-main.R

@@ -1,4 +1,4 @@
-## ----main-setup, include=FALSE----------------------------------------------------------------------------
+## ----main-setup, include=FALSE-----------------------------------------------------------------
 options(prompt = 'R> ', continue = '+ ')
 knitr::opts_chunk$set(
   echo = FALSE,
@@ -21,11 +21,11 @@ knitr::opts_chunk$set(
 # knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE----------------------------------------------------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE-----------------------------------------------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-------------------------------------------------------------------
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE--------------------------------------------------------
 library("sdmTMB")
 # simplify for paper:
 pcod_original <- pcod # save it for later
@@ -33,30 +33,30 @@ pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
   tibble::as_tibble()
 
 
-## ----sdmTMB-lib, eval=TRUE, echo=TRUE---------------------------------------------------------------------
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE----------------------------------------------------------
 library("sdmTMB")
 head(pcod, n = 3)
 
 
-## ----pcod-utms-eval, echo=FALSE, eval=TRUE----------------------------------------------------------------
+## ----pcod-utms-eval, echo=FALSE, eval=TRUE-----------------------------------------------------
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 
 
-## ----pcod-utms-echo, echo=TRUE, eval=FALSE----------------------------------------------------------------
+## ----pcod-utms-echo, echo=TRUE, eval=FALSE-----------------------------------------------------
 ## pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 ## #> Detected UTM zone 9N; CRS = 32609.
 ## #> Visit https://epsg.io/32609 to verify.
 
 
-## ----pcod-head2, echo=TRUE--------------------------------------------------------------------------------
+## ----pcod-head2, echo=TRUE---------------------------------------------------------------------
 head(pcod, n = 3)
 
 
[…]

### 2023-01-25 doc/paper-jss/sdmTMB-cover.tex
Add cover

\documentclass[letterpaper]{letter}
\usepackage[margin=1.50in]{geometry}
\geometry{letterpaper}
\usepackage{graphicx}
\addtolength{\topmargin}{-30pt}
\addtolength{\textheight}{55pt}

\usepackage{textcomp}
% \usepackage[sb]{libertine}
% \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
% \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
% \usepackage[cal=boondoxo]{mathalfa} % mathcal
% \useosf % osf for text, not math
% \usepackage[supstfm=libertinesups,%
%   supscaled=1.2,%
%   raised=-.13em]{superiors}

\providecommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\let\proglang=\textsf

\usepackage[dvipsnames]{xcolor}
\definecolor{niceblue}{HTML}{236899}
\newcommand{\rev}[1]{{\color{niceblue} #1}}
\usepackage{hyperref}
\hypersetup{
    colorlinks = true,
    linkcolor = black,
    filecolor = black,
    urlcolor = blue,
    citecolor = blue,
    linkbordercolor = white
}

% \signature{Sean Anderson}
\address{
  % Sean C. Anderson\\
  % Research Scientist \\
  % Fisheries and Oceans Canada \\
  % Pacific Biological Station \\
  % Nanaimo, BC, Canada \\
  % sean.anderson@dfo-mpo.gc.ca
}

\begin{document}
\begin{letter}{}
\pagestyle{empty}

\opening{Dear Editorial Board,}

Thank you for considering our manuscript for publication in the Journal of Statistical Software. Our manuscript introduces the \proglang{R} package \pkg{sdmTMB} for fitting a broad class of generalized linear mixed effect models (GLMMs) that include spatial or spatiotemporal random fields. Such models are useful across a wide array of disciplines.

We believe that \pkg{sdmTMB} represents a major advance in terms of speed, functionality, and accessibility for fitting SPDE (stochastic partial differential equation)-based GLMMs to spatial or spatiotemporal data.

Speed: The package combines the SPDE approximation with estimation in the \pkg{TMB} \proglang{R} package for computational efficiency. Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages, the next fastest related software.

Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors, threshold/breakpoint models) and spatial anisotropy (directionally dependent spatial correlation; lacking in \pkg{INLA} and \pkg{inlabru}). Spatial anisotropy is critical in many applied contexts such as when modelling species distributions near any geographic boundary or along an environmental gradient.

Accessibility: \pkg{sdmTMB} is a much more usable interface than alternatives for applied users, who may be familiar with \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv}. In our experience, such applied users struggle to code these types of models in \pkg{INLA}; \pkg{inlabru} improves the experience but still requires an understanding of \pkg{INLA} to fit many useful but complex structures. Furthermore, \pkg{INLA} and \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors. \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, while by default allowing users the simplicity and speed of maximum likelihood.

As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{16 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and workshops that in total have been attended by $>$200 participants.

% We think the readers of JSS would be the ideal audience for our manuscf

JSS would be the ideal outlet to provide a citable resource for \pkg{sdmTMB} and to introduce the package to a wider audience. \pkg{sdmTMB} is an open-source \proglang{R} package \href{https://CRAN.R-project.org/package=sdmTMB}{published on CRAN}. Our code is fully reproducible and included with our submission. We have posted a preprint version of our manuscript \href{https://doi.org/10.1101/2022.03.24.485545}{on bioRxiv}. We look forward to constructive feedback from your reviewers and editors.

\vspace{3mm}
\closing{Sincerely,\\
\fromsig{\includegraphics[scale=0.42]{/Users/seananderson/Dropbox/sig.pdf}}
}

\end{letter}
\end{document}

### 2023-01-25 doc/paper-jss/sdmTMB-cover.tex
Update sdmTMB-cover.tex
Feel free to take or leave these -- mostly trying to shorten / focus

@@
 
-We believe that \pkg{sdmTMB} represents a major advance in terms of speed, functionality, and accessibility for fitting SPDE (stochastic partial differential equation)-based GLMMs to spatial or spatiotemporal data.
+We believe that \pkg{sdmTMB} represents a major advance in terms of functionality, accessibility, and speed for fitting the broad class of SPDE (stochastic partial differential equation)-based GLMMs to spatial or spatiotemporal data.
+
+Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors as in \pkg{INLA}, threshold/breakpoint models), spatial anisotropy (directionally dependent spatial correlation), and methods for model comparison (cross-validation, stacking).
+
+Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particular for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages.
+Existing packages, such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors.
+While \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, by default estimation is done using the simplicity and speed of maximum likelihood.
 
@@
 
-Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors, threshold/breakpoint models) and spatial anisotropy (directionally dependent spatial correlation; lacking in \pkg{INLA} and \pkg{inlabru}).
-Spatial anisotropy is critical in many applied contexts such as when modelling species distributions near any geographic boundary or along an environmental gradient.
-
-Accessibility: \pkg{sdmTMB} is a much more usable interface than alternatives for applied users, who may be familiar with \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv}.
-In our experience, such applied users struggle to code these types of models in \pkg{INLA}; \pkg{inlabru} improves the experience but still requires an understanding of \pkg{INLA} to fit many useful but complex structures.
-Furthermore, \pkg{INLA} and \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors.
-\pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, while by default allowing users the simplicity and speed of maximum likelihood.
-
-As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{16 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and workshops that in total have been attended by $>$200 participants.
+As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{16 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and international workshops that in total have been attended by $>$200 participants.
 

### 2023-01-25 doc/paper-jss/sdmTMB-cover.tex
Edits

@@
 
-Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors as in \pkg{INLA}, threshold/breakpoint models), spatial anisotropy (directionally dependent spatial correlation), and methods for model comparison (cross-validation, stacking).
+Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond standard random fields (penalized smoothers as in \pkg{mgcv}, priors as in \pkg{INLA}, threshold/breakpoint models), spatial anisotropy (directionally dependent spatial correlation), and methods for model comparison (cross-validation, stacking).
 
-Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particular for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages.
+Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particularly for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages.
 Existing packages, such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors.
@@
 Speed: The package combines the SPDE approximation with estimation in the \pkg{TMB} \proglang{R} package for computational efficiency.
-Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages, the next fastest related software.
+Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages, the software with the most similar capabilities for this class of models.
 

### 2023-02-02 doc/paper-jss/pcod.Rmd
Update for sdmTMBextra

@@
 
-We can inspect simulated residuals from the DHARMa package (Fig.\ \ref{fig:sim-resids}).
+We can inspect randomized quantile residuals calculated with predictions that used the Laplace approximation:
 
-```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE, fig.cap="Simulated residuals from the DHARMa package.", fig.pos='ht', fig.align='center'}
+```{r}
 set.seed(123)
-simulate(fit, nsim = 500) %>%
-  dharma_residuals(fit)
+r <- residuals(fit)
 ```
@@
 set.seed(123)
-r <- residuals(fit,
-  type = "mle-mcmc", 
-  mcmc_iter = 200, mcmc_warmup = 100
-)
+post <- sdmTMBextra::predict_mle_mcmc(fit, mcmc_iter = 200, mcmc_warmup = 100)
+r <- residuals(fit, mcmc_samples = post, type = "mle-mcmc")
 ```

### 2023-02-02 doc/paper-jss/make-reprex-jss.R
Remove stuff suppressing spin figs

@@ -80,6 +80,13 @@ d <- c(d[seq(1, i-1)], "\n# Pacific Cod appendix -------------------------------
 i <- grep("## ----inla-knitr-setup", d)
 d <- c(d[seq(1, i-1)], "\n# INLA comparison appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])
 
+# these make figs not show up in html:
+d <- gsub('dev = \\"pdf\\"', 'dev = \\"png\\"', d)
+d <- gsub('out.width=\\"[0-9]+in\\", ', "", d)
+
+# check:
+d[grepl("out.width=\\'[0-9]+in\\'", d)]
+
 writeLines(d, "reprex/replication-code-main.R")
 
 system("cp ~/src/sdmTMB-paper/data/SNOW_data.rds ~/src/sdmTMB-paper/doc/paper-jss/reprex/snow-data.rds")

### 2023-02-07 doc/paper-jss/owls.Rmd
Add CBC data note

@@
 We will use counts of Snowy Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts [CBCs, @cbc].
+
+All requests for use of Christmas Bird Count data must be [submitted to Audubon](https://survey123.arcgis.com/share/7dc33b4fff77468a8bba855291f86527) and must adhere to the [Audubon Policy Regarding use of Christmas Bird Count Data](https://www.audubon.org/content/policy-regarding-use-christmas-bird-count-data).
+Questions can be directed to `cbcadmin@audubon.org`.
+
 Each survey is conducted between December 14 to January 5 in a 24.14 kilometer (15 mile) diameter circle. 

### 2023-02-14 doc/paper-jss/owls.Rmd
Remove MCMC sdmTMBextra code

@@
 The MCMC-based residuals are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
-To calculate these, we have to use a version of our model fit without REML, which we can quickly obtain with the `update()` method.
-We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
-For the purposes of building this appendix quickly, we will not run this code.
+To calculate these, we would have to use a version of our model fit without REML, which we can quickly obtain with the `update()` method.
 
@@
 fit_ml <- update(fit_owl, reml = FALSE)
-mcmc_resids <- residuals(fit_owl,
-  type = "mle-mcmc",
-  mcmc_iter = 101,
-  mcmc_warmup = 100
-)
 ```
+
+Then, see the [residuals vignette](https://pbs-assess.github.io/sdmTMB/articles/) for examples of how MCMC residuals can be estimated with the \pkg{sdmTMBextra} package.
 

### 2023-02-14 doc/paper-jss/pcod.Rmd
Remove MCMC sdmTMBextra code

@@
 
-We can inspect randomized quantile residuals calculated with predictions that used the Laplace approximation:
+We can inspect randomized quantile residuals calculated with predictions that used the Laplace approximation (Fig.\ \ref{fig:laplace-resids-vis}):
 
-```{r}
+```{r laplace-resids-vis}
 set.seed(123)
 r <- residuals(fit)
-```
-
-Or perhaps better, use MCMC residuals (Fig.\ \ref{fig:mcmc-resids-vis}).
-
-```{r mcmc-resids, warning=FALSE, results='hide', cache=TRUE}
-set.seed(123)
-post <- sdmTMBextra::predict_mle_mcmc(fit, mcmc_iter = 200, mcmc_warmup = 100)
-r <- residuals(fit, mcmc_samples = post, type = "mle-mcmc")
-```
-
-```{r mcmc-resids-vis, out.width="4in", fig.cap="MCMC residuals.", fig.pos='ht', fig.align='center'}
 qqnorm(r)
@@
 ```
+
+These residuals can suggest model fit issues even with properly specified models due to the Laplace approximation [@thygesen2017a].
+We could instead use MCMC residuals, which more accurate but slower.
+These are currently available in the \pkg{sdmTMBextra} package at <https://github.com/pbs-assess/sdmTMBextra> and are demonstrated in the [residuals vignette](https://pbs-assess.github.io/sdmTMB/articles/).
 

### 2023-02-14 analysis/pcod-fig.R

@@ -6,10 +6,13 @@ library("patchwork")
 dir.create("figs", showWarnings = FALSE)
 
 if (!require("rnaturalearthhires", quietly = TRUE)) {
-  stop(
-    "Please install 'rnaturalearthhires'.\n",
+  message(
+    "Please install 'rnaturalearthhires' to use a high-resolution map.\n",
     "`remotes::install_github('ropensci/rnaturalearthhires')`"
   )
+  rnaturalearthhires_installed <- FALSE
+} else {
+  rnaturalearthhires_installed <- TRUE
 }
 
 # from remotes::install_github('seananderson/ggsidekick')
@@ -28,7 +31,7 @@ theme_sleek <- function(base_size = 11, base_family = "") {
         size = rel(0.9)
       ), panel.border = element_rect(
         fill = NA,
-        colour = "grey70", size = 1
+        colour = "grey70", linewidth = 1
       ), legend.key.size = grid::unit(
         0.9,
         "lines"
@@ -95,8 +98,9 @@ lims_x <- c(230957.7 + 105000, 1157991 - 570000) + c(-10000, 10000)
 lims_y <- c(5366427 + 270000, 6353456 - 513000) + c(-10000, 10000)
 land <- "grey86"
 land_border <- "grey86"
+
 map_data <- rnaturalearth::ne_countries(
-  scale = "large",
+  scale = if (rnaturalearthhires_installed) "large" else "medium",
   returnclass = "sf", country = "canada"
 )
 

### 2023-02-14 doc/paper-jss/make-reprex-jss.R

@@ -2,6 +2,7 @@ RUN_SPIN <- TRUE
 
 setwd(here::here())
 setwd("doc/paper-jss/")
+dir.create("reprex", showWarnings = FALSE)
 
 knitr::purl("paper-jss.Rmd", documentation = 1L)
 
@@ -69,7 +70,6 @@ d <- gsub("here::here\\(\"ne_10m_lakes\"\\)", "\"ne_10m_lakes\"", d)
 d <- c(d, "sessionInfo()")
 
 d[grepl("here::", d)]
-
 
 i <- grep("## ----setup-owls", d)
 d <- c(d[seq(1, i-1)], "\n# Snowy Owl appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])

### 2023-02-14 doc/paper-jss/reprex/replication-code-main.R

@@ -1,4 +1,4 @@
-## ----main-setup, include=FALSE-----------------------------------------
+## ----main-setup, include=FALSE------------------------------------------------------------------
 options(prompt = 'R> ', continue = '+ ')
 knitr::opts_chunk$set(
   echo = FALSE,
@@ -21,11 +21,11 @@ knitr::opts_chunk$set(
 # knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE-----------------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE------------------------------------------------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----sdmTMB-lib0, eval=TRUE, echo=FALSE--------------------------------
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE---------------------------------------------------------
 library("sdmTMB")
 # simplify for paper:
 pcod_original <- pcod # save it for later
@@ -33,30 +33,30 @@ pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
   tibble::as_tibble()
 
 
-## ----sdmTMB-lib, eval=TRUE, echo=TRUE----------------------------------
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE-----------------------------------------------------------
 library("sdmTMB")
 head(pcod, n = 3)
 
 
-## ----pcod-utms-eval, echo=FALSE, eval=TRUE-----------------------------
+## ----pcod-utms-eval, echo=FALSE, eval=TRUE------------------------------------------------------
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 
 
-## ----pcod-utms-echo, echo=TRUE, eval=FALSE-----------------------------
+## ----pcod-utms-echo, echo=TRUE, eval=FALSE------------------------------------------------------
 ## pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 ## #> Detected UTM zone 9N; CRS = 32609.
 ## #> Visit https://epsg.io/32609 to verify.
 
 
-## ----pcod-head2, echo=TRUE---------------------------------------------
+## ----pcod-head2, echo=TRUE----------------------------------------------------------------------
 head(pcod, n = 3)
 
 
-## ---- echo=TRUE, eval=TRUE, cache=TRUE---------------------------------
[…]

### 2023-02-14 doc/paper-jss/sdmTMB-cover.tex
Update cover

@@
 \documentclass[letterpaper]{letter}
-\usepackage[margin=1.50in]{geometry}
+\usepackage[margin=1.55in]{geometry}
 \geometry{letterpaper}
@@
 Thank you for considering our manuscript for publication in the Journal of Statistical Software.
+This is a revised version of manuscript 4939 after receiving the following editor comments with an encouragement to resubmit:
+
+\begin{quote}
+\input{editor-comments0}
+\end{quote}
+
+We apologize for not better dealing with the GitHub-only package \pkg{sdmTMBextra}.
+Indeed, this was originally a part of the main package.
+We had just split this functionality out of the main package because of valgrind/gcc-ASAN check errors on CRAN that could not be replicated or resolved.
+This also had the benefit of substantially reducing the number of package dependencies for the main CRAN package.
+We have now omitted the corresponding code in the appendix and referenced the vignette on residuals for further details.
+
+Our original cover letter follows:
+
 Our manuscript introduces the \proglang{R} package \pkg{sdmTMB} for fitting a broad class of generalized linear mixed effect models (GLMMs) that include spatial or spatiotemporal random fields.
@@
 
-Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particularly for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages.
-Existing packages, such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors.
+Accessibility: \pkg{sdmTMB} provides a more usable interface than alternatives, particularly for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages.
+Furthermore, existing packages such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors.
 While \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, by default estimation is done using the simplicity and speed of maximum likelihood.
@@
 Speed: The package combines the SPDE approximation with estimation in the \pkg{TMB} \proglang{R} package for computational efficiency.
-Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages, the software with the most similar capabilities for this class of models.
+Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages---the software with the most similar capabilities for this class of models.
 
@@
 We have posted a preprint version of our manuscript \href{https://doi.org/10.1101/2022.03.24.485545}{on bioRxiv}.
+We have included a list of potential reviewers on the following page should it be helpful.
 We look forward to constructive feedback from your reviewers and editors.
@@
 
+\vspace{6mm}
+
+Suggested reviewers:
+
+\input{suggested-reviewers}
+
 \end{letter}

### 2023-02-14 doc/make.R
Clean up

@@ -1,48 +1,0 @@
-files <- c(
-  "sdmTMB-paper",
-  "appendix-model",
-  "appendix-pcod-sdmTMB",
-  "appendix-snow-sdmTMB",
-  "appendix-binomial-inla",
-  "appendix-pcod-VAST-tweedie",
-  "appendix-validation"
-)
-
-# https://github.com/rstudio/rmarkdown/issues/1673
-render_separately <- function(...) callr::r(
-  function(...) rmarkdown::render(..., envir = globalenv()), args = list(...), show = TRUE)
-
-for (i in seq_along(files)) {
-  render_separately(paste0(here::here("doc", files[i]), ".Rmd"))
-}
-
-names_to <- c(
-  "sdmTMB-paper.pdf",
-  "appendix1-model.pdf",
-  "appendix2-index-sdmTMB.pdf",
-  "appendix3-owl-sdmTMB.pdf",
-  "appendix4-INLA-comparison.pdf",
-  "appendix5-VAST-comparison.pdf",
-  "appendix6-speed-validation.pdf"
-)
-
-loc <- "~/Dropbox/sdmTMB-paper/"
-dir.create(loc, showWarnings = FALSE)
-
-for (i in seq_along(files)) {
-  from <- paste0(here::here("doc", files[i]), ".pdf")
-  to <- paste0(loc, names_to[i])
-  cat(to, "\n")
-  file.copy(from, to, overwrite = TRUE)
-}
-
-system("cd doc; ./clean-bib.sh")
-
-
-# f <- paste(names_to, collapse = " ")
-
-# system(paste0("cd ~/Documents/sdmTMB-paper/;'/System/Library/Automator/Combine PDF Pages.action/Contents/Resources/join.py' -o sdmTMB-paper-combined.pdf ", f))
-
-# system(paste0("cd ~/Dropbox/sdmTMB-paper/;/opt/homebrew/bin/gs -q -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=sdmTMB-paper-combined.pdf ", f))
-
-system("open ~/Dropbox/sdmTMB-paper/")

### 2023-03-10 analysis/timing.R
Add theme_sleek #13

@@ -12,6 +12,46 @@ dir.create("analysis", showWarnings = FALSE)
 
 INLA::inla.setOption(num.threads = "1:1")
 
+# from remotes::install_github('seananderson/ggsidekick')
+theme_sleek <- function(base_size = 11, base_family = "") {
+  half_line <- base_size / 2
+  theme_light(base_size = base_size, base_family = base_family) +
+    theme(
+      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+      axis.ticks.length = grid::unit(half_line / 2.2, "pt"), strip.background = element_rect(
+        fill = NA,
+        colour = NA
+      ), strip.text.x = element_text(colour = "grey30"),
+      strip.text.y = element_text(colour = "grey30"), axis.text = element_text(colour = "grey30"),
+      axis.title = element_text(colour = "grey30"), legend.title = element_text(
+        colour = "grey30",
+        size = rel(0.9)
+      ), panel.border = element_rect(
+        fill = NA,
+        colour = "grey70", linewidth = 1
+      ), legend.key.size = grid::unit(
+        0.9,
+        "lines"
+      ), legend.text = element_text(
+        size = rel(0.7),
+        colour = "grey30"
+      ), legend.key = element_rect(
+        colour = NA,
+        fill = NA
+      ), legend.background = element_rect(
+        colour = NA,
+        fill = NA
+      ), plot.title = element_text(
+        colour = "grey30",
+        size = rel(1)
+      ), plot.subtitle = element_text(
+        colour = "grey30",
+        size = rel(0.85)
+      )
+    )
+}
+theme_set(theme_sleek())
+
 # source(here::here("analysis/mgcv_spde_smooth.R"))
 # -------------------------------------------------------------------------
 # Code in this chunk is from:
@@ -359,31 +399,6 @@ leg <- tribble(
   "mgcv::bam\n(discretize = TRUE) SPDE", 249, 20.5
 )
 
-# log-space version:
-
-# g <- out_long_sum %>%
-#   ggplot(aes(mean_mesh_n, time, colour = model_clean)) +
-#   geom_ribbon(aes(ymin = lwr, ymax = upr, fill = model_clean),
-#     alpha = 0.2,
-#     colour = NA
-#   ) +
[…]

### 2023-03-10 doc/paper-jss/pcod.Rmd
Remove aes_string #13

@@
 plot_map <- function(dat, column) {
-  ggplot(dat, aes_string("X", "Y", fill = column)) +
+  ggplot(dat, aes(X, Y, fill = {{ column }})) +
     geom_raster() +
@@
 ```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
-plot_map(pred_qcs, "exp(est)") +
+plot_map(pred_qcs, exp(est)) +
   scale_fill_viridis_c(trans = "log10") 
@@
 ```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
-plot_map(pred_qcs, "exp(est_non_rf)") +
+plot_map(pred_qcs, exp(est_non_rf)) +
   scale_fill_viridis_c(trans = "sqrt")
@@
 ```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='ht', fig.align='center'}
-plot_map(pred_qcs, "omega_s") +
+plot_map(pred_qcs, omega_s) +
   scale_fill_gradient2()
@@
 ```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='ht', fig.align='center'}
-plot_map(pred_qcs, "epsilon_st") +
+plot_map(pred_qcs, epsilon_st) +
   scale_fill_gradient2()

### 2023-03-10 doc/paper-jss/paper-jss.Rmd
Update for sdmTMBextra
Drop GLMM abbrev in abstract #13

@@
   Geostatistical data are common across scientific fields and available in increasingly large datasets.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement for many users.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users.
 Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
@@
 A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
-Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
+Simulation-based residuals [@dharma] are also possible.
 (5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model. 

### 2023-03-11 doc/paper-jss/paper-jss.Rmd
Formatting #13

@@
   \usepackage{newunicodechar}
+  \usepackage{gensymb}
   \newunicodechar{✓}{\ding{51}}
   \newunicodechar{✗}{\ding{55}}
+  \newunicodechar{˚}{\degree}
   \DeclareGraphicsExtensions{.pdf,.png,.jpg}
@@
 ```{r main-setup, include=FALSE}
-options(prompt = 'R> ', continue = '+ ')
 knitr::opts_chunk$set(
@@
   comment = "#>",
-  prompt = FALSE,
+  # prompt = FALSE,
   cache = TRUE,
@@
   optipng = "-strip all"
-  # R.options = list(prompt = "R> ", continue = "> ")
+  # R.options = list(prompt = "R> ", continue = "+ ")
 )
 knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
 ```
@@
 
-```{r pcod-utms-eval, echo=FALSE, eval=TRUE}
+```{r pcod-utms-eval, echo=TRUE, eval=TRUE}
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
@@
 
-```{r pcod-utms-echo, echo=TRUE, eval=FALSE}
-pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+```
 #> Detected UTM zone 9N; CRS = 32609.
@@
 
-\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.}
 \centering
@@
 \end{tabular}
-\begin{tablenotes}
-\item \textit{Note: }
-\item $^1$Technically possible but non-trivial.
+\end{threeparttable}
+\label{tab:functionality}
+
+\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.
+Notes: 
+$^1$Technically possible but non-trivial.
 $^2$Penalized smoother GAMs that determine `wiggliness'.
@@
 $^{10}$Hurdle models possible by fitting components separately.
-\end{tablenotes}
-\end{threeparttable}
-\label{tab:functionality}
+}
 \end{table}

### 2023-03-11 doc/paper-jss/inla-comparison.Rmd
more \pkg #13

@@
 
-# Comparison between INLA, inlabru, and sdmTMB {#app:inla}
+# Comparison between \pkg{INLA}, \pkg{inlabru}, and \pkg{sdmTMB} {#app:inla}
 
@@
 
-In INLA, a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).
+In INLA, a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, \pkg{sdmTMB} only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).
 
@@
 
-## INLA version
+## \pkg{INLA} version
 
@@
 
-## inlabru version
+## \pkg{inlabru} version
 
@@
 
-## sdmTMB version
+## \pkg{sdmTMB} version
 
@@
 
-```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the sdmTMB model.", fig.align='center'}
+```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \pkg{sdmTMB} model.", fig.align='center'}
 inverse_link <- m$family$linkinv
@@
 
-```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the inlabru model.", fig.align='center'}
+```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \pkg{inlabru} model.", fig.align='center'}
 nd_bru <- nd

### 2023-03-11 doc/paper-jss/inla-comparison.Rmd
Make sections work with TeX #13

@@
 
-# Comparison between \pkg{INLA}, \pkg{inlabru}, and \pkg{sdmTMB} {#app:inla}
+\section[Comparison between INLA, inlabru, and sdmTMB]{Comparison between \pkg{INLA}, \pkg{inlabru}, and \pkg{sdmTMB}}\label{app:inla}
 
@@
 
-## \pkg{INLA} version
+\subsection[INLA version]{\pkg{INLA} version}
 
@@
 
-## \pkg{inlabru} version
+\subsection[inlabru version]{\pkg{inlabru} version}
 
@@
 
-## \pkg{sdmTMB} version
+\subsection[sdmTMB version]{\pkg{sdmTMB} version}
 
@@
 
-```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \pkg{sdmTMB} model.", fig.align='center'}
+```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{sdmTMB} model.", fig.align='center'}
 inverse_link <- m$family$linkinv
@@
 
-```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \pkg{inlabru} model.", fig.align='center'}
+```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{inlabru} model.", fig.align='center'}
 nd_bru <- nd

### 2023-03-11 doc/paper-jss/paper-jss.Rmd
Make sections work with TeX #13

@@
 This can greatly improve computational efficiency [e.g., @heaton2019].
-The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
+The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
 Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].
@@
 
-```{r child="model-description.Rmd", eval=FALSE}
-```
-
 ```{r child="pcod.Rmd", eval=TRUE, cache=TRUE}

### 2023-03-11 doc/paper-jss/timing.Rmd
Make sections work with TeX #13

@@
 
-## Illustration of generating the INLA mesh
+## Illustration of generating the \textbf{INLA} mesh
 
@@
 
-## Example sdmTMB model fit
+\subsection[Example sdmTMB model fit]{Example \pkg{sdmTMB} model fit}
 
@@
 
-## Example spaMM model fit
+\subsection[Example spaMM model fit]{Example \pkg{spaMM} model fit}
 
@@
 
-## Example inlabru model fit
+\subsection[Example inlabru model fit]{Example \pkg{inlabru} model fit}
 
@@
 
-## Example mgcv model fit
+\subsection[Example mgcv model fit]{Example \pkg{mgcv} model fit}
 

### 2023-03-11 analysis/timing.R
Newton loops off; update cache

@@ -187,8 +187,12 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
 
   cat("sdmTMB\n")
   out <- system.time({
-    fit <- sdmTMB(observed ~ a1,
-      data = sim_dat, mesh = mesh, family = family,
+    fit <- sdmTMB(
+      observed ~ a1,
+      data = sim_dat,
+      mesh = mesh,
+      family = family,
+      control = sdmTMBcontrol(newton_loops = 0L),
       priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2))
     )
   })
@@ -357,8 +361,8 @@ system.time({
   )
 })
 
-saveRDS(out, file = "analysis/timing-cache-parallel-2023-01-24.rds")
-out <- readRDS("analysis/timing-cache-parallel-2023-01-24.rds")
+saveRDS(out, file = "analysis/timing-cache-parallel-2023-03-10.rds")
+out <- readRDS("analysis/timing-cache-parallel-2023-03-10.rds")
 
 out_long <- tidyr::pivot_longer(
   out,

### 2023-03-11 doc/paper-jss/owls.Rmd
No need for extra optim now

@@
 #> ✓ No extreme or very small eigenvalues detected
-#> ✗ `ln_tau_E` gradient > 0.001
-#>   See `?run_extra_optimization()`
-#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
-#> 
-#> ✗ `ln_phi` gradient > 0.001
-#>   See `?run_extra_optimization()`
-#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
-#> 
-#> ✓ No fixed-effect standard errors are NA
-#> ✓ No fixed-effect standard errors look unreasonably large
-#> ✓ No sigma parameters are < 0.01
-#> ✓ No sigma parameters are > 100
-#> ✓ Range parameter doesn't look unreasonably large
-```
-
-We note that the magnitude of some log-likelihood gradients is slightly large for some fixed effects.
-
-```{r}
-fit_owl$gradients
-```
-
-We can run additional optimization steps with a Newton optimizer using the\
-`run_extra_optimization()` function to further reduce these gradients.
-
-```{r owl-newton, results='hide', cache=TRUE}
-fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
-```
-
-We can confirm that the gradients are smaller.
-
-```{r sanity2, echo=TRUE}
-sanity(fit_owl2)
-```
-
-```
-#> ✓ Non-linear minimizer suggests successful convergence
-#> ✓ Hessian matrix is positive definite
-#> ✓ No extreme or very small eigenvalues detected
 #> ✓ No gradients with respect to fixed effects are >= 0.001
@@
 ```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE}
-print(fit_owl2)
+print(fit_owl)
 ```
@@
 ```{r tidy-owls}
-tidy(fit_owl2, conf.int = TRUE)
-tidy(fit_owl2, effects = "ran_pars", conf.int = TRUE)
+tidy(fit_owl, conf.int = TRUE)
+tidy(fit_owl, effects = "ran_pars", conf.int = TRUE)
 ```
@@
 ```{r predict-owls, message=FALSE, cache=TRUE}
-pred <- predict(fit_owl2)
+pred <- predict(fit_owl)
 
@@
 set.seed(19208)
-pred$resid <- residuals(fit_owl2)
+pred$resid <- residuals(fit_owl)
 ```
@@
 ```{r zero-test-owls, echo=TRUE, message=FALSE}
-s_nb2 <- simulate(fit_owl2, nsim = 400)
+s_nb2 <- simulate(fit_owl, nsim = 400)
 ```
@@
 ```{r zero-test-owls3, echo=TRUE}
-pred_fixed <- fit_owl2$family$linkinv(pred$est_non_rf)
+pred_fixed <- fit_owl$family$linkinv(pred$est_non_rf)
 r_nb2 <- DHARMa::createDHARMa(
   simulatedResponse = s_nb2,
-  observedResponse = fit_owl2$data$count,
+  observedResponse = fit_owl$data$count,
   fittedPredictedResponse = pred_fixed
@@
 We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
-Most simply, we can extract the overall coefficient using the `b <- tidy(fit_owl2)` and add it to the spatially varying coefficients `zeta_s`.
+Most simply, we can extract the overall coefficient using the `b <- tidy(fit_owl)` and add it to the spatially varying coefficients `zeta_s`.
 However, this method does not give us spatially explicit estimates of uncertainty.
@@
 ```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
-b <- tidy(fit_owl2, conf.int = TRUE)
+b <- tidy(fit_owl, conf.int = TRUE)
 
 nsim <- 200
-zeta_s <- predict(fit_owl2, nsim = nsim, sims_var = "zeta_s")
-sims <- spread_sims(fit_owl2, nsim = nsim)
+zeta_s <- predict(fit_owl, nsim = nsim, sims_var = "zeta_s")
+sims <- spread_sims(fit_owl, nsim = nsim)
 beta <- sims$nao

### 2023-03-11 doc/paper-jss/reprex/replication-code-main.R
Update reprex scripts

@@ -1,12 +1,11 @@
-## ----main-setup, include=FALSE------------------------------------------------------------------
-options(prompt = 'R> ', continue = '+ ')
+## ----main-setup, include=FALSE----------------------------------------
 knitr::opts_chunk$set(
   echo = FALSE,
   message = FALSE,
   warning = FALSE,
   collapse = TRUE,
   comment = "#>",
-  prompt = FALSE,
+  # prompt = FALSE,
  # cache = TRUE,
  # autodep = TRUE,
   fig.width = 7,
@@ -16,16 +15,17 @@ knitr::opts_chunk$set(
   dev = "png",
   dpi = 140
   # optipng = "-strip all"
-  # R.options = list(prompt = "R> ", continue = "> ")
+  # R.options = list(prompt = "R> ", continue = "+ ")
 )
 # knitr::knit_hooks$set(optipng = knitr::hook_optipng)
+options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE------------------------------------------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE----------------------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----sdmTMB-lib0, eval=TRUE, echo=FALSE---------------------------------------------------------
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-------------------------------
 library("sdmTMB")
 # simplify for paper:
 pcod_original <- pcod # save it for later
@@ -33,30 +33,24 @@ pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
   tibble::as_tibble()
 
 
-## ----sdmTMB-lib, eval=TRUE, echo=TRUE-----------------------------------------------------------
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE---------------------------------
 library("sdmTMB")
 head(pcod, n = 3)
 
 
-## ----pcod-utms-eval, echo=FALSE, eval=TRUE------------------------------------------------------
+## ----pcod-utms-eval, echo=TRUE, eval=TRUE-----------------------------
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 
 
-## ----pcod-utms-echo, echo=TRUE, eval=FALSE------------------------------------------------------
-## pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
-## #> Detected UTM zone 9N; CRS = 32609.
[…]

### 2023-03-11 doc/paper-jss/reprex/replication-code-timing.R

@@ -12,6 +12,46 @@ dir.create("analysis", showWarnings = FALSE)
 
 INLA::inla.setOption(num.threads = "1:1")
 
+# from remotes::install_github('seananderson/ggsidekick')
+theme_sleek <- function(base_size = 11, base_family = "") {
+  half_line <- base_size / 2
+  theme_light(base_size = base_size, base_family = base_family) +
+    theme(
+      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+      axis.ticks.length = grid::unit(half_line / 2.2, "pt"), strip.background = element_rect(
+        fill = NA,
+        colour = NA
+      ), strip.text.x = element_text(colour = "grey30"),
+      strip.text.y = element_text(colour = "grey30"), axis.text = element_text(colour = "grey30"),
+      axis.title = element_text(colour = "grey30"), legend.title = element_text(
+        colour = "grey30",
+        size = rel(0.9)
+      ), panel.border = element_rect(
+        fill = NA,
+        colour = "grey70", linewidth = 1
+      ), legend.key.size = grid::unit(
+        0.9,
+        "lines"
+      ), legend.text = element_text(
+        size = rel(0.7),
+        colour = "grey30"
+      ), legend.key = element_rect(
+        colour = NA,
+        fill = NA
+      ), legend.background = element_rect(
+        colour = NA,
+        fill = NA
+      ), plot.title = element_text(
+        colour = "grey30",
+        size = rel(1)
+      ), plot.subtitle = element_text(
+        colour = "grey30",
+        size = rel(0.85)
+      )
+    )
+}
+theme_set(theme_sleek())
+
 # source(here::here("analysis/mgcv_spde_smooth.R"))
 # -------------------------------------------------------------------------
 # Code in this chunk is from:
@@ -147,8 +187,12 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
 
   cat("sdmTMB\n")
   out <- system.time({
-    fit <- sdmTMB(observed ~ a1,
-      data = sim_dat, mesh = mesh, family = family,
+    fit <- sdmTMB(
+      observed ~ a1,
+      data = sim_dat,
[…]

### 2023-03-13 doc/paper-jss/inla-comparison.Rmd
Copy edit

@@
 
-In this example, we will fit a presence-absence dynamic species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
+In this example, we will fit a presence-absence (technically, encounter vs.
+lack of encounter) dynamic species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
 We will use a binomial distribution and a logit link.
@@
 
+```{r sdmtmb-sanity-comparison, echo=TRUE}
+sanity(m)
+```
+
 ```{r sdmtmb-m, echo=TRUE}
@@
 
-```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='ht', fig.align='center'}
+```{r coef-plot, fig.asp=0.25, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='ht', fig.align='center'}
 inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
@@
 
-We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
-For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`---the inverse logit here), to produce occurrence probabilities:
+We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map encounter probabilities in each year in space (Fig.\ \ref{fig:maps}).
+For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`---the inverse logit here), to produce encounter probabilities:
 
@@
 
-```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{sdmTMB} model.", fig.align='center'}
+```{r maps, out.width="5.5in", fig.cap="Predicted encounter probabilities in time and space from the \\pkg{sdmTMB} model.", fig.align='center'}
 inverse_link <- m$family$linkinv
@@
 
-```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{inlabru} model.", fig.align='center'}
+```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted encounter probabilities in time and space from the \\pkg{inlabru} model.", fig.align='center'}
 nd_bru <- nd

### 2023-03-13 doc/paper-jss/owls.Rmd
Copy edit

@@
 
-First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations (Fig.\ \ref{fig:mesh-owls}).
+First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
 The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
-This mesh uses a cutoff distance of 1.5 units (or 150 km).
+This mesh uses a cutoff distance of 1.5 units (or 150 km) (Fig.\ \ref{fig:mesh-owls}).
 
@@
 
-The result is a mesh with `r mesh$mesh$n` (`mesh$mesh$n`) knots.
+The result is a mesh with `r mesh$mesh$n` (`mesh$mesh$n`) vertices.
 
@@
 
-We can more formally test this with functions from the \pkg{DHARMa} package.
+We can more formally test this with functions from the \pkg{DHARMa} [@dharma] package.
 
@@
 
-To produce maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an \pkg{sf} object so that the plot axes will be in latitude and longitude.
+To produce maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an \pkg{sf} [@pebesma2018] object so that the plot axes will be in latitude and longitude.
 
@@
 
-Next we retrieve continental coastlines from the package \pkg{rnaturalearth} and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.
+Next we retrieve continental coastlines from the package \pkg{rnaturalearth} [@rnaturalearth] and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.
 

### 2023-03-13 doc/paper-jss/paper-jss.Rmd
Copy edit

@@
 \pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \proglang{R}-\pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
-\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
+\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages (as described in the DESCRIPTION and inst/COPYRIGHTS files).
 Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].

### 2023-03-13 doc/paper-jss/pcod.Rmd
Cap #13
Copy edit

@@
 )
-```
-
-```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
-library("ggplot2")
-library("dplyr")
-library("sdmTMB")
-theme_set(theme_minimal())
 ```
@@
 
-```{r plot-mesh, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="Delaunay triangulation mesh.", fig.pos='ht', fig.align='center'}
+```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
+library("ggplot2")
+library("dplyr")
+library("sdmTMB")
+```
+
+```{r theme-minimal}
+theme_set(theme_minimal())
+```
+
+```{r plot-mesh, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="Delaunay triangulation mesh. Dots represent observations of Pacific Cod.", fig.pos='ht', fig.align='center'}
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
@@
 
-```{r laplace-resids-vis}
+```{r laplace-resids-vis, fig.cap="Randomized quantile residuals."}
 set.seed(123)
@@
 These residuals can suggest model fit issues even with properly specified models due to the Laplace approximation [@thygesen2017a].
-We could instead use MCMC residuals, which more accurate but slower.
+We could instead use MCMC residuals, which are more accurate but slower.
 These are currently available in the \pkg{sdmTMBextra} package at <https://github.com/pbs-assess/sdmTMBextra> and are demonstrated in the [residuals vignette](https://pbs-assess.github.io/sdmTMB/articles/).
@@
 
-```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
+```{r plot-fix-defects, fig.cap="Predictions based on only the non-random-field elements---a mean effect for each year.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, exp(est_non_rf)) +
@@
 
-```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='ht', fig.align='center'}
+```{r plot-spatial-effects, fig.cap="Spatial random effects only. This random field is constant through time.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, omega_s) +
@@
 
-```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='ht', fig.align='center'}
+```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center'}
 plot_map(pred_qcs, epsilon_st) +
@@
 
-```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='ht', fig.align='center'}
+```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.
+The line represents the estimated mean and the ribbon represents a 95\\% confidence interval.", fig.pos='ht', fig.align='center'}
 survey_grid$area <- 4 # all 2 x 2km
@@
   geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000), alpha = 0.4) +
-  ylab("Biomass (t)")
+  ylab("Biomass (t)") + xlab("Year")
 ```

### 2023-03-13 doc/paper-jss/timing.Rmd
More \pkg #13
Copy edit

@@
 
-Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
+Here, we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
 We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
@@
 
-## Illustration of generating the \textbf{INLA} mesh
+\subsection[Illustration of generating the INLA mesh]{Illustration of generating the \pkg{INLA} mesh}
 
@@
 
+First define `smooth.construct.spde.smooth.spec()` and `Predict.matrix.spde.smooth()`
+from the supplement of @miller2019, then:
+
 ```{r mgcvfit-timing, echo=TRUE, eval=TRUE}
-# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth()
-# from supplement of:
-# Miller, D.L., Glennie, R. & Seaton, A.E. (2019). Understanding the Stochastic
-# Partial Differential Equation approach to smoothing. Journal of Agricultural,
-# Biological and Environmental Statistics.
-# https://doi.org/10.1007/s13253-019-00377-z
-
 fit_bam <- bam(

### 2023-03-13 doc/paper-jss/paper-jss.Rmd
Copy editing; mostly ref formatting

@@
   \usepackage{bm}
-  \usepackage{lscape}
+  \usepackage{pdflscape}
   \usepackage{tablefootnote}
@@
 See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
-Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
+Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{$\sim$ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
 \end{figure}
@@
 
-\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>:
+\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>
 
@@
 
-Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
+Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package  [@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed for the `make_mesh()` function below.
+
+Here, we illustrate a spatial model fit to Pacific Cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
 Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.
+Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific Cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.
 
@@
 We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
-We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
+We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros.
 An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).
@@
 \includegraphics[width=5.2in]{../../figs/pcod-fig}
-\caption[]{Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
+\caption[]{Output from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada.
 (A) SPDE mesh (lines) combined with the trawl survey observations (points).
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
+The circle area corresponds to biomass density of Pacific Cod caught on individual trawls.
 (B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
@@
 For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
-A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
+A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue [@rufener2021] is recommended but slower (`?residuals.sdmTMB`).
 Simulation-based residuals [@dharma] are also possible.
@@
 \pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
-In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} since it is a prominent package that can fit related predictive-process models without the SPDE.
+In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} [@finley2007; @finley2015] since it is a prominent package that can fit related predictive-process models without the SPDE.
 \pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
@@
 Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
+Results with optimized math libraries on one core (openBLAS: @openblas and PARDISO: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
 <!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

### 2023-03-13 doc/paper-jss/paper-jss.Rmd
Copy edit

@@
 We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
-The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].
+\pkg{sdmTMB} can use many `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].
 
@@
 See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
-Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{$\sim$ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
+Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde\ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
 \end{figure}
@@
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
-We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix \ref{app:inla}.
+We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix&nbsp;\ref{app:inla}.
 <!-- and \pkg{VAST} (Appendix 5). -->
@@
 Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries on one core (openBLAS: @openblas and PARDISO: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
+Results with optimized math libraries on one core (\pkg{openBLAS}: @openblas and \pkg{PARDISO}: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
 <!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
@@
 How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
-Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
+Assuming a given package can fit the model of interest (Table  &nbsp;\ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
 We think users familiar with `stats::glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.

### 2023-03-20 analysis/timing.R
Update timing fig

@@ -361,8 +361,8 @@ system.time({
   )
 })
 
-saveRDS(out, file = "analysis/timing-cache-parallel-2023-03-10.rds")
-out <- readRDS("analysis/timing-cache-parallel-2023-03-10.rds")
+saveRDS(out, file = "analysis/timing-cache-parallel-2023-03-15.rds")
+out <- readRDS("analysis/timing-cache-parallel-2023-03-15.rds")
 
 out_long <- tidyr::pivot_longer(
   out,
@@ -431,7 +431,7 @@ g <- out_long_sum %>%
 print(g)
 
 .width <- 5
-ggsave("figs/timing-spatial-2023-01-24-xkcd.pdf", width = .width, height = .width / 1.3)
+ggsave("figs/timing-spatial-2023-03-16-xkcd.pdf", width = .width, height = .width / 1.3)
 
 cat("inlabru to sdmTMB timing ratios\n")
 x <- filter(

### 2023-03-20 doc/paper-jss/paper-jss.Rmd
Bring in new timing fig; edit timing text

@@
 \pkg{mgcv} was most affected by mesh resolution.
-<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
--->
-<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
-<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
 Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
 Results with optimized math libraries on one core (\pkg{openBLAS}: @openblas and \pkg{PARDISO}: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
-<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
-<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->
 
@@
 \centering
-\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-01-24-xkcd.pdf}
+\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
 \caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
@@
 
-With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 6--24 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
+With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \proglang{R}-\pkg{INLA}/\pkg{inlabru}.
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.

### 2023-03-20 doc/paper-jss/reprex/replication-code-main.R
Rebuild reprex R files

@@ -1,4 +1,4 @@
-## ----main-setup, include=FALSE----------------------------------------
+## ----main-setup, include=FALSE--------------------------
 knitr::opts_chunk$set(
   echo = FALSE,
   message = FALSE,
@@ -21,11 +21,11 @@ knitr::opts_chunk$set(
 options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE----------------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-------------------------------
+## ----sdmTMB-lib0, eval=TRUE, echo=FALSE-----------------
 library("sdmTMB")
 # simplify for paper:
 pcod_original <- pcod # save it for later
@@ -33,20 +33,20 @@ pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
   tibble::as_tibble()
 
 
-## ----sdmTMB-lib, eval=TRUE, echo=TRUE---------------------------------
+## ----sdmTMB-lib, eval=TRUE, echo=TRUE-------------------
 library("sdmTMB")
 head(pcod, n = 3)
 
 
-## ----pcod-utms-eval, echo=TRUE, eval=TRUE-----------------------------
+## ----pcod-utms-eval, echo=TRUE, eval=TRUE---------------
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 
 
-## ----pcod-head2, echo=TRUE--------------------------------------------
+## ----pcod-head2, echo=TRUE------------------------------
 head(pcod, n = 3)
 
 
-## ---- echo=TRUE, eval=TRUE, cache=TRUE--------------------------------
+## ---- echo=TRUE, eval=TRUE, cache=TRUE------------------
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
 
 
@@ -60,49 +60,49 @@ fit <- sdmTMB(
 )
 
 
-## ----pcod-eg1-summary,eval=TRUE, echo=TRUE----------------------------
+## ----pcod-eg1-summary,eval=TRUE, echo=TRUE--------------
 fit
 
 
[…]

### 2023-03-20 doc/paper-jss/reprex/replication-code-timing.R

@@ -361,8 +361,8 @@ system.time({
   )
 })
 
-saveRDS(out, file = "analysis/timing-cache-parallel-2023-01-24.rds")
-out <- readRDS("analysis/timing-cache-parallel-2023-01-24.rds")
+saveRDS(out, file = "analysis/timing-cache-parallel-2023-03-15.rds")
+out <- readRDS("analysis/timing-cache-parallel-2023-03-15.rds")
 
 out_long <- tidyr::pivot_longer(
   out,
@@ -431,7 +431,7 @@ g <- out_long_sum %>%
 print(g)
 
 .width <- 5
-ggsave("figs/timing-spatial-2023-01-24-xkcd.pdf", width = .width, height = .width / 1.3)
+ggsave("figs/timing-spatial-2023-03-16-xkcd.pdf", width = .width, height = .width / 1.3)
 
 cat("inlabru to sdmTMB timing ratios\n")
 x <- filter(

### 2023-08-04 doc/paper-jss/paper-jss.Rmd
In progress revamp of examples Requires very latest sdmTMB

@@
 Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
-Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
 We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
@@
 Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
-See the the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
+See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
 
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
+See the \pkg{sdmTMB} \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
 Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde\ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
@@
 # Introductory example using fish survey data
+
+## Illustrating the need for spatial random fields
+
+We begin with a simple species distribution model of encounter probability of
+Pacific Cod (TODO spp) from a trawl survey conducted off the west
+coast of Vancouver Island, British Columbia, Canada. We will fit a logistic
+regression of encounter probability with and without spatial random fields to
+illustrate the importance of accounting for spatial correlation.
+
+```{r libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
+library(sdmTMB)
+library(dplyr)
+library(ggplot2)
+```
+
+```{r libs-extras}
+theme_set(theme_light())
+options(ggplot2.continuous.fill = "viridis")
+```
+
+```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
+plot(mesh_pcod)
+```
+
+```{r}
+fit_bin_rf <- sdmTMB(
+  present ~ poly(log(depth), 2),
+  data = dat, mesh = mesh, spatial = "on",
+  family = binomial(link = "logit")
+)
+fit_bin <- update(fit_bin_rf, spatial = "off")
+```
+
+```{r dog-bin-summary, echo=TRUE}
+summary(fit_bin_rf)
+```
+
+Describe what we see here
+
+We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals.
+
+```{r}
+# set.seed(1)
+# pcod$resid_bin <- residuals(fit_bin)
+# set.seed(1)
+# pcod$resid_bin_rf <- residuals(fit_bin_rf)
+# g1 <- ggplot(pcod, aes(X, Y, colour = resid_bin)) + geom_point() +
+#   scale_color_gradient2() +
+#   theme(legend.position = "bottom")
+# g2 <- ggplot(pcod, aes(X, Y, colour = resid_bin_rf)) + geom_point() +
+#   scale_color_gradient2() +
+#   theme(legend.position = "bottom")
+# cowplot::plot_grid(g1, g2, ncol = 2)
+```
+
+```{r, echo=FALSE, results='hide'}
+test_autocor <- function(obj) {
+  set.seed(1)
+  s <- simulate(obj, nsim = 500)
+  pr <- predict(obj, type = "response")$est
+  r <- DHARMa::createDHARMa(
+    simulatedResponse = s,
+    observedResponse = pcod$present,
+    fittedPredictedResponse = pr
+  )
+  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
+}
+test_autocor(fit_bin)
+test_autocor(fit_bin_rf)
+```
+
+We can also see that the AIC is considerably lower:
+
+```{r}
+AIC(fit_bin_rf, fit_bin)
+```
+
+Further testing could be done with cross validation via `sdmTMB_cv()`.
+
+```{r}
+cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
+  data = dat, mesh = mesh, spatial = "on",
+  family = binomial(link = "logit"), k_folds = 10
+)
+cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
+  data = dat, mesh = mesh, spatial = "off",
+  family = binomial(link = "logit"), k_folds = 10
+)
+```
+
+Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields:
+
+```{r}
+cv_bin_rf$sum_loglik
+cv_bin_rf$sum_loglik
+```
+
+The standard errors on our fixed effects have increased:
+
+```{r}
+tidy(fit_bin_rf, conf.int = T)
+tidy(fit_bin, conf.int = T)
+```
+
+We now have an effect of depth and a spatial random field, which combined in link space result in our overall prediction.
+Note the difference between only the fixed effects (left) and including the spatial random field (right).
+
+```{r}
+p <- predict(fit_bin_rf, newdata = qcs_grid)
+plot_spatial_map <- function(dat, column, title) {
+  ggplot(dat, aes(X, Y, fill = {{ column }})) +
+    geom_raster() +
+    coord_fixed() +
+    theme(legend.position = "bottom") +
+    ggtitle(title)
+}
+g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
+g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
+g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
+cowplot::plot_grid(g1, g2, g3, ncol = 3)
+```
+
+# Spatiotemporal species distribution modeling with dogfish
+
+```{r dog-head-dat, echo=TRUE}
+dat <- select(dogfish, longitude, latitude, year, catch_weight, area_swept, depth)
+head(dat)
+```
+
+```{r dog-utms, echo=TRUE}
+dat <- add_utm_columns(dat, c("longitude", "latitude"), units = "km", utm_crs = 32609)
+dat$log_depth <- log(dat$depth)
+```
+
+```{r dog-mesh, echo=TRUE}
+mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
+plot(mesh)
+```
+
+```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+fit_tw <- sdmTMB(
+  catch_weight ~ s(log_depth),
+  data = dat,
+  mesh = mesh,
+  offset = log(dat$area_swept),
+  time = "year",
+  time_varying = ~ 1,
+  time_varying_type = "ar1",
+  spatial = "on",
+  spatiotemporal = "iid",
+  anisotropy = TRUE,
+  share_range = TRUE,
+  silent = FALSE,
+  family = tweedie()
+)
+```
+
+We can use `update()` to refit our model with four additional families:
+delta-gamma, delta-lognormal, delta-Poisson-link-gamma, and
+delta-Poisson-link-lognormal.
+
+```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+fit_dg <- update(fit_tw, family = delta_gamma(), spatiotemporal = list("off", "iid"))
+fit_dl <- update(fit_dg, family = delta_lognormal())
+fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
+fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
+```
+
+We can then compare the fits via AIC:
+
+```{r dog-aic, echo=TRUE}
+AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
+  mutate(delta_AIC = AIC - min(AIC)) |> 
+  arrange(delta_AIC)
+```
+
+We find that the Poisson-link-lognormal model is favoured with AIC. We next
+test two additional model formulations: making the spatial correlation
+isotropic (the default) instead of anisotropic and structuring the
+spatiotemporal random fields as AR(1).
+
+```{r dog-ar1, results='hide', echo=TRUE}
+fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
+fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
+```
+
+```{r dog-aci2, echo=TRUE}
+AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
+```
+
+We find that the anisotropic AR(1) is favoured. On inspecting the model output
+with `summary()`, we notice that the spatial random field SD has become very
+small once AR(1) spatiotemporal fields were added. We therefore consider
+a model where these are turned off in the second component of the delta model:
+
+```{r dog-ar1-2, echo=TRUE, results="hide"}
+fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
+```
+
+```{r dog-aic3, echo=TRUE}
+AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
+```
+
+```{r dog-print, echo=TRUE}
+fit <- fit_dpl_ar1_only
+sanity(fit)
+summary(fit)
+```
+
+```{r dog-aniso, echo=TRUE}
+plot_anisotropy(fit_dpl)
+```
+
+```{r dog-grid, echo=TRUE}
+head(wcvi_grid, n = 3)
+grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
+grid$log_depth <- log(grid$depth)
+head(grid, n = 3)
+```
+
+```{r dog-pred1, echo=TRUE}
+pred <- predict(fit, newdata = grid, type = "response")
+```
+
+```{r dog-pred2, echo=TRUE}
+names(pred)
+```
+
+```{r plot-map, echo=TRUE}
+plot_map <- function(dat, column) {
+  ggplot(dat, aes(X, Y, fill = {{ column }})) +
+    geom_raster() +
+    facet_wrap(vars(year)) +
+    coord_fixed()
+}
+```
+
+```{r dog-plot1, echo=TRUE}
+pred |> filter(year %in% c(2004, 2014, 2022)) |> 
+  plot_map(est) +
+  labs(fill = "Density") +
+  scale_fill_viridis_c(trans = "log10")
+```
+
+```{r dog-plot2, echo=TRUE}
+pred |> filter(year %in% 2004) |> 
+  plot_map(omega_s2) +
+  scale_fill_gradient2()
+```
+
+```{r dog-plot3, echo=TRUE}
+pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(epsilon_st2) +
+  scale_fill_gradient2()
+```
+
+```{r dog-plot4, echo=TRUE}
+pred |> filter(year %in% 2004) |> 
+  plot_map(est_non_rf1) +
+  scale_fill_viridis_c(trans = "log10")
+```
+
+```{r dog-depth, echo=TRUE}
+nd <- data.frame(
+  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
+  year = max(dat$year)
+)
+pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA, model = NA)
+ggplot(pred_depth, aes(
+  exp(log_depth), exp(est), 
+  ymin = exp(est - 2 * est_se), 
+  ymax = exp(est + 2 * est_se))) +
+  geom_ribbon(fill = "grey90") +
+  geom_line() +
+  labs(x = "Depth (m)", y = "Density")
+```
+
+```{r dog-index, echo=TRUE}
+pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
+ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
+```
+
+```{r dog-index-plot, echo=TRUE}
+ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
+  geom_pointrange() +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
+  labs(y = "Biomass", x = "Year")
+```
+
+# Spatially varying coefficient example with citizen science data
+
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
+The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
+The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
+Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
+We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+
+We can write the model as TODO
+
+```{r owl-data}
+snow <- readRDS(here::here("data/SNOW_data.rds"))
+```
+
+```{r owl-fit, echo=TRUE, results = "hide"}
+mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
+fit_owl <- sdmTMB(
+  count ~ 1 + nao + (1 | year_f),
+  spatial_varying = ~ nao,
+  time = "year",
+  data = snow, 
+  family = nbinom2(link = "log"),
+  spatial = "on",
+  spatiotemporal = "iid",
+  mesh = mesh_snow,
+  reml = TRUE,
+  silent = FALSE
+)
+```
+
+```{r owl-sanity, eval=FALSE}
+sanity(fit_owl)
+```
+
+```{r owl-print, echo=TRUE}
+summary(fit_owl)
+```
+
+We can extract the SVC effect with TODO
+
+```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
+nsim <- 200
+zeta_s <- predict(fit_owl, newdata = snow, nsim = nsim, sims_var = "zeta_s")
+sims <- spread_sims(fit_owl, nsim = nsim)
+combined <- sims$nao + t(zeta_s)
+snow$nao_effect <- apply(exp(combined), 2, median)
+```
+
+We can make a basic plot using the following code. A more elaborate version is shown in Figure \@ref(fig:owl-plot-fancy).
+
+```{r owl-plot-basic, echo=TRUE, eval=FALSE}
+ggplot(snow, aes(X, Y)) +
+  geom_point(aes(colour = nao_effect)) +
+  scale_colour_viridis_c()
+```
+
+Overall, we found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}).
+This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
+We provide an extended version of this example in Appendix \ref{app:owls}.
+
+```{r shapes, echo=FALSE}
+if (!file.exists(here::here("data/ne_10m_lakes"))) {
+  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
+    "download/10m/physical/ne_10m_lakes.zip")
+  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
+  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+}
+```
+
+```{r shapes-read, echo=FALSE}
+Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
+coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
+  sf::st_transform(crs = Albers)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
+```
+
+```{r owl-proj2}
+# project to get nice axis limits 
+snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
+snow2 <- snow2 |> mutate(x = X, y = Y) |> 
+  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
+```
+
+```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Map of spatially varying coefficient effects of NAO on Snowy Owl count.
+In the west, the lower values are near 1 implying no effect, whereas in the southeast the effect becomes positive.
+Points represent all count locations."}
+nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
+  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
+ggplot(data = nao_effect_df) +
+  geom_point(aes(X, Y, colour = nao_effect)) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  coord_sf(
+    xlim = c(min(snow2$X), max(snow2$X)),
+    ylim = c(min(snow2$Y), max(snow2$Y))
+  ) +
+   scale_colour_viridis_c(
+    guide = guide_colourbar(direction = "horizontal", title.position = "left")
+  ) +
+  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
+    theme(legend.position = "bottom", axis.title = element_blank())
+```
+
+\clearpage
 
@@
 
-```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
-library("sdmTMB")
-# simplify for paper:
+```{r sdmTMB-lib0, eval=FALSE, echo=FALSE}
+# library("sdmTMB")
+# # simplify for paper:
 pcod_original <- pcod # save it for later
-pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
-  tibble::as_tibble()
+# pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
+#   tibble::as_tibble()
+
 ```
 
-```{r sdmTMB-lib, eval=TRUE, echo=TRUE}
+```{r sdmTMB-lib, eval=FALSE, echo=TRUE}
 library("sdmTMB")
-head(pcod, n = 3)
+dplyr::select(pcod, year, density, depth, lon, lat) |>
+  head(n = 3)
 ```
@@
 
-```{r pcod-utms-eval, echo=TRUE, eval=TRUE}
+```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
@@
 
-```{r pcod-head2, echo=TRUE}
+```{r pcod-head2, echo=TRUE, eval=FALSE}
 head(pcod, n = 3)
@@
 
-```{r, echo=TRUE, eval=TRUE, cache=TRUE}
+```{r, echo=TRUE, eval=FALSE, cache=TRUE}
 mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
@@
 
-```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
+```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE, cache=TRUE}
 fit <- sdmTMB(
@@
 
-```{r pcod-eg1-summary,eval=TRUE, echo=TRUE}
+```{r pcod-eg1-summary,eval=FALSE, echo=TRUE}
 fit
@@
 
-```{r pcod-eg1-tidy-fe, eval=TRUE, echo=TRUE}
+```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
 tidy(fit, conf.int = TRUE)
@@
 
-```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
+```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
 tidy(fit, effects = "ran_pars", conf.int = TRUE)
@@
 
-```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE}
+```{r pcod-eg1-sanity, eval=FALSE, echo=TRUE}
 sanity(fit)
@@
 
-```{r pcod-eg1-suppress-depth2, eval=TRUE, echo=FALSE}
+```{r pcod-eg1-suppress-depth2, eval=FALSE, echo=FALSE}
 qcs_grid_original <- qcs_grid
@@
 
-```{r qcs-head, echo=TRUE}
+```{r qcs-head, echo=FALSE}
 head(qcs_grid, n = 3)
@@
 
-```{r pcod-eg1-predict, echo=TRUE, eval=TRUE, cache=TRUE}
+```{r pcod-eg1-predict, echo=FALSE, eval=TRUE, cache=TRUE}
 p <- predict(fit, newdata = qcs_grid)
@@
 
-```{r pcod-p-tibble}
+```{r pcod-p-tibble, eval=FALSE}
 p <- tibble::as_tibble(p)
@@
 
-```{r pcod-head-predict, echo=TRUE}
+```{r pcod-head-predict, echo=TRUE, eval=FALSE}
 head(p, n = 3)
@@
 
-```{r time1, cache=TRUE}
+```{r time1, cache=TRUE, eval=FALSE}
 t1 <- Sys.time()
@@
 
-```{r fit-tv, eval=TRUE, echo=TRUE, cache=TRUE, results='hide'}
+```{r fit-tv, eval=FALSE, echo=TRUE, cache=TRUE, results='hide'}
 fit_spatiotemporal <- sdmTMB(
@@
 
-```{r time2, cache=TRUE}
+```{r time2, cache=TRUE, eval=FALSE}
 t2 <- Sys.time()
@@
 
-```{r fit-tv-print, eval=TRUE, echo=TRUE}
+```{r fit-tv-print, eval=FALSE, echo=TRUE}
 fit_spatiotemporal
@@
 
-```{r pcod-restore, echo=FALSE}
+```{r pcod-restore, echo=FALSE, eval=FALSE}
 pcod <- pcod_original
@@
 
-```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
+```{r owl-fit-old, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
 mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 
-```{r child="pcod.Rmd", eval=TRUE, cache=TRUE}
+```{r child="pcod.Rmd", eval=FALSE, cache=TRUE}
 ```
 
-```{r child="owls.Rmd", eval=TRUE, cache=TRUE}
+```{r child="owls.Rmd", eval=FALSE, cache=TRUE}
 ```
 
-```{r child="inla-comparison.Rmd", eval=TRUE, cache=TRUE}
+```{r child="inla-comparison.Rmd", eval=FALSE, cache=TRUE}
 ```
 
-```{r child="timing.Rmd", eval=TRUE, cache=TRUE}
+```{r child="timing.Rmd", eval=FALSE, cache=TRUE}
 ```

### 2023-08-31 doc/paper-jss/paper-jss.Rmd
Init comments on fish models

@@
 Pacific Cod (TODO spp) from a trawl survey conducted off the west
-coast of Vancouver Island, British Columbia, Canada. We will fit a logistic
+coast of Vancouver Island, British Columbia, Canada.
+This survey consists of spatially random sampling in each year, resulting in a georeferenced dataset.
+For compatibility with other packages, spatial random fields in \pkg{sdmTMB} can be toggled on and off.
+We will fit a logistic
 regression of encounter probability with and without spatial random fields to
-illustrate the importance of accounting for spatial correlation.
+illustrate the importance of accounting for spatial correlation. In addition to the spatial random field, we include an intercept and a quadratic effect of depth on the probability of occurrence.
 
@@
 
-We can also see that the AIC is considerably lower:
+We can also see that the AIC of the model with spatial random fields is considerably lower:
 
@@
 # Spatiotemporal species distribution modeling with dogfish
+
+<!-- Need brief description of dogfish data and model here?? -->
+
+As a second example, we will construct a model of catch rates of North Pacific spiny dogfish from the same survey.
+This example extends the spatial model described previously by also including (1) spatiotemporal fields allowing the distribution in each year to be unique (modelled as an autoregressive process, allowing hotspots to persist through time), (2) a time varying intercept to allow each year to have a different mean catch rate, and (3) a smooth effect of ln(depth), allowing catch rates to vary non-linearly with depth.
+As catch rates are both skewed and zero inflated, we specify the response family as a Tweedie distribution [@tweedie1984].
 
@@
 
-We can use `update()` to refit our model with four additional families:
+We can use `update()` to refit our model quickly without specifying all of the arguments.
+This is highlighted by exploring models with four additional families, each using a delta- or hurdle distribution:
 delta-gamma, delta-lognormal, delta-Poisson-link-gamma, and
@@
 
+<!-- Need description of anisotropy and why it's important here or somewhere else -->
 We find that the anisotropic AR(1) is favoured. On inspecting the model output
@@
 
+<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
 ```{r dog-grid, echo=TRUE}

### 2023-08-31 doc/paper-jss/paper-jss.Rmd
example of paragraph on modular design etc

@@
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a] for a specific class of spatial and spatiotemporal models.
+Many features of \pkg{sdmTMB} may be found in other software (Table 1) but to date no package has included all in a single package.
 One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}).
@@
 
-The results from an \pkg{sdmTMB} model can be used to generate various quantities of interest.
-For example, for the field of ecology, \pkg{sdmTMB} provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.
+The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both modular and familiar to users of widely used \proglang{R} packages (e.g.
+\pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
+Data may be simulated with or without a fitted model (`simulate.sdmTMB()`, `sdmTMB_simulate()` respectively).
+Estimation is done via maximum marginal likelihood with the `sdmTMB()` function (or `sdmTMB_cv()` for cross-validation), and the resulting fitted model can be used to make predictions (`predict.sdmTMB()`) or calculate residuals (`residuals.sdmTMB()`), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+Common quantities of interest may include summarizing shifts in the center of gravity (`get_cog()`) [e.g., @thorson2016cog] or in the field of ecology, tracking total population densities through time (`get_index()`).
 

### 2023-09-08 doc/paper-jss/paper-jss.Rmd
Revising updated examples

@@
 
-Systems for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
+Software packages for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
 For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
@@
 
-The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}).
+The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Figure \ref{fig:diagram}).
 This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
@@
 
-By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}).
+By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure \ref{fig:diagram}).
 Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
@@
 
+We can create a 'mesh' for the SPDE calculations with `make_mesh()`.
+
 ```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
@@
 
+We will fit a model with and without spatial random fields by toggling the `spatial` argument from `"on"` to `"off`. The `update()` S3 method refits the same model updating any specified arguments.
+
 ```{r}
@@
   present ~ poly(log(depth), 2),
-  data = dat, mesh = mesh, spatial = "on",
+  data = pcod, mesh = mesh_pcod, 
+  spatial = "on",
   family = binomial(link = "logit")
@@
 
+We can get a summary of our model fit by printing it.
+
 ```{r dog-bin-summary, echo=TRUE}
-summary(fit_bin_rf)
+fit_bin_rf
 ```
@@
 
-Further testing could be done with cross validation via `sdmTMB_cv()`.
+We could do additional testing with cross validation via `sdmTMB_cv()`.
+Here we will do 10-fold cross validation with the folds constructed randomly.
+Using the `fold_ids` argument, we could supply our own folds and conduct spatially blocked cross validation, for example.
 
@@
 cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
-  data = dat, mesh = mesh, spatial = "on",
+  data = pcod, mesh = mesh_pcod, spatial = "on",
   family = binomial(link = "logit"), k_folds = 10
@@
 cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
-  data = dat, mesh = mesh, spatial = "off",
+  data = pcod, mesh = mesh_pcod, spatial = "off",
   family = binomial(link = "logit"), k_folds = 10
@@
 mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
+```
+
+```{r dog-mesh2, eval=FALSE}
 plot(mesh)
@@
 test two additional model formulations: making the spatial correlation
-isotropic (the default) instead of anisotropic and structuring the
+isotropic (the default) instead of anisotropic and structuring the positive model component
 spatiotemporal random fields as AR(1).
@@
 We find that the anisotropic AR(1) is favoured. On inspecting the model output
-with `summary()`, we notice that the spatial random field SD has become very
+with `summary()`, we notice that the spatial random field SD in the positive model component has become very
 small once AR(1) spatiotemporal fields were added. We therefore consider
@@
 
+Our prediction data frame has..., which we can easily plot with ggplot2 or similar. We suppress that code in this paper for brevity.
+
 ```{r dog-pred2, echo=TRUE}
@@
 
-```{r plot-map, echo=TRUE}
+```{r plot-map, echo=FALSE}
 plot_map <- function(dat, column) {
@@
     facet_wrap(vars(year)) +
-    coord_fixed()
+    coord_fixed() +
+    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
 }
@@
 
-```{r dog-plot1, echo=TRUE}
-pred |> filter(year %in% c(2004, 2014, 2022)) |> 
+```{r dog-plot1, echo=FALSE}
+g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
   plot_map(est) +
-  labs(fill = "Density") +
+  # labs(fill = "Density") +
   scale_fill_viridis_c(trans = "log10")
@@
 
-```{r dog-plot2, echo=TRUE}
-pred |> filter(year %in% 2004) |> 
-  plot_map(omega_s2) +
+```{r dog-plot2, echo=FALSE}
+g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(omega_s1) +
   scale_fill_gradient2()
@@
 
-```{r dog-plot3, echo=TRUE}
-pred |> filter(year %in% c(2004, 2022)) |> 
+```{r dog-plot3, echo=FALSE}
+g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
   plot_map(epsilon_st2) +
@@
 
-```{r dog-plot4, echo=TRUE}
-pred |> filter(year %in% 2004) |> 
+```{r dog-plot4, echo=FALSE}
+g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
   plot_map(est_non_rf1) +
   scale_fill_viridis_c(trans = "log10")
+```
+
+```{r dog-wvci-pred, fig.asp = 0.66, fig.cap="Example prediction components..."}
+cowplot::plot_grid(
+  g_nonrf,
+  g_omega,
+  g_eps,
+  g_est,
+  ncol = 2L
+)
 ```
@@
 pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA, model = NA)
+```
+
+```{r dog-depth-plot, echo=FALSE}
 ggplot(pred_depth, aes(
@@
 
+The SVC effect at any point in space is the combination of the fixed effect and spatial random effect for `nao`. Here we illustrate
+
 We can extract the SVC effect with TODO
@@
 
-We can make a basic plot using the following code. A more elaborate version is shown in Figure \@ref(fig:owl-plot-fancy).
+We can make a basic plot using the following code. A more elaborate version is shown in Figure \ref(fig:owl-plot-fancy).
 
@@
 
-Overall, we found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}).
+Overall, we found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure \ref{fig:owl-nao}).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
@@
 Alternatively, we could have created any mesh via the \proglang{R}-\pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
-We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \ref{fig:pcod-fig}A).
+We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Figure \ref{fig:pcod-fig}A).
 
@@
 
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig.\ \ref{fig:pcod-fig}B--D.
+We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Figure \ref{fig:pcod-fig}B--D.
 
@@
 We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
-We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
+We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Figure \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
 
@@
 
-We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}).
+We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure \ref{fig:owl-nao}).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
@@
 
-We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig.
-\ref{fig:timing}, Appendix \ref{app:speed}).
+We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure \ref{fig:timing}, Appendix \ref{app:speed}).
 In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
@@
 \pkg{inlabru} is similar to \pkg{sdmTMB} in design, but differs fundamentally in that fitting is achieved via \proglang{R}-\pkg{INLA} and approximate Bayesian computation.
-For some models, TMB/\pkg{sdmTMB} can be considerably faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru}, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig.
-\@ref(fig:timing), @osgood-zimmerman2021].
+For some models, TMB/\pkg{sdmTMB} can be considerably faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru}, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Figure \@ref(fig:timing), @osgood-zimmerman2021].
 One major capability included in \pkg{sdmTMB} but not \proglang{R}-\pkg{INLA}/\pkg{inlabru} is anisotropy.
@@
 
+<!-- TODO: RTMB -->
+<!-- TODO: multivariate -->
+
 # Acknowledgements
@@
 Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.
+<!-- TODO: thank reviewers -->
 

### 2023-09-08 doc/paper-jss/owls.Rmd
owls tweak

@@
   geom_point(
-    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
+    data = p, aes(X, Y, colour = nao_effect_sim, size = exp(est)), alpha = 0.5
   ) +

### 2023-09-08 doc/paper-jss/sdmTMB-cover.tex
Update cover letter

@@
 \documentclass[letterpaper]{letter}
-\usepackage[margin=1.55in]{geometry}
+\usepackage[margin=1.43in]{geometry}
 \geometry{letterpaper}
 \usepackage{graphicx}
-\addtolength{\topmargin}{-30pt}
-\addtolength{\textheight}{55pt}
+\addtolength{\topmargin}{-20pt}
+\addtolength{\textheight}{75pt}
 
@@
 Thank you for considering our manuscript for publication in the Journal of Statistical Software.
-This is a revised version of manuscript 4939 after receiving the following editor comments with an encouragement to resubmit:
-
-\begin{quote}
-\input{editor-comments0}
-\end{quote}
-
-We apologize for not better dealing with the GitHub-only package \pkg{sdmTMBextra}.
-Indeed, this was originally a part of the main package.
-We had just split this functionality out of the main package because of valgrind/gcc-ASAN check errors on CRAN that could not be replicated or resolved.
-This also had the benefit of substantially reducing the number of package dependencies for the main CRAN package.
-We have now omitted the corresponding code in the appendix and referenced the vignette on residuals for further details.
+We have made all the requested editorial changes as described in the file \texttt{formatting-response.pdf}.
 
@@
 Speed: The package combines the SPDE approximation with estimation in the \pkg{TMB} \proglang{R} package for computational efficiency.
-Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages---the software with the most similar capabilities for this class of models.
+Across our comparisons, \pkg{sdmTMB} was 7--30 times faster than the \pkg{INLA} or \pkg{inlabru} packages---the software with the most similar capabilities for this class of models.
 
-As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{16 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and international workshops that in total have been attended by $>$200 participants.
+As evidence of its utility, there are already \href{https://github.com/pbs-assess/sdmTMB/wiki/Publications-using-sdmTMB}{19 published preprints, papers, or reports} using \pkg{sdmTMB}, six funded research grants, an ever-growing number of ongoing research projects, and international workshops that in total have been attended by $>$200 participants.
 
@@
 
-\vspace{3mm}
+\vspace{2mm}
 \closing{Sincerely,\\
-\fromsig{\includegraphics[scale=0.42]{/Users/seananderson/Dropbox/sig.pdf}}
+\fromsig{\includegraphics[scale=0.3]{/Users/seananderson/Dropbox/sig.pdf}}
 }

### 2023-09-08 doc/paper-jss/paper-jss.Rmd
Comment out old examples

@@
 ```{r sdmTMB-install, eval=FALSE, echo=TRUE}
-install.packages("sdmTMB", dependencies = TRUE)
+# install.packages("sdmTMB", dependencies = TRUE)
 ```
@@
 ```{r sdmTMB-lib0, eval=FALSE, echo=FALSE}
+# # library("sdmTMB")
+# # # simplify for paper:
+# pcod_original <- pcod # save it for later
+# # pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
+# #   tibble::as_tibble()
+```
+
+```{r sdmTMB-lib, eval=FALSE, echo=TRUE}
 # library("sdmTMB")
-# # simplify for paper:
-pcod_original <- pcod # save it for later
-# pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
-#   tibble::as_tibble()
-
-```
-
-```{r sdmTMB-lib, eval=FALSE, echo=TRUE}
-library("sdmTMB")
-dplyr::select(pcod, year, density, depth, lon, lat) |>
-  head(n = 3)
+# dplyr::select(pcod, year, density, depth, lon, lat) |>
+#   head(n = 3)
 ```
@@
 ```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
-pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+# pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
 ```
@@
 ```{r pcod-head2, echo=TRUE, eval=FALSE}
-head(pcod, n = 3)
+# head(pcod, n = 3)
 ```
@@
 ```{r, echo=TRUE, eval=FALSE, cache=TRUE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
+# mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
 ```
@@
 ```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE, cache=TRUE}
-fit <- sdmTMB(
-  density ~ s(depth),
-  data = pcod,
-  family = tweedie(link = "log"),
-  mesh = mesh,
-  spatial = "on"
-)
+# fit <- sdmTMB(
+#   density ~ s(depth),
+#   data = pcod,
+#   family = tweedie(link = "log"),
+#   mesh = mesh,
+#   spatial = "on"
+# )
 ```
@@
 ```{r pcod-eg1-summary,eval=FALSE, echo=TRUE}
-fit
+# fit
 ```
@@
 ```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
-tidy(fit, conf.int = TRUE)
+# tidy(fit, conf.int = TRUE)
 ```
@@
 ```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
-tidy(fit, effects = "ran_pars", conf.int = TRUE)
+# tidy(fit, effects = "ran_pars", conf.int = TRUE)
 ```
@@
 ```{r pcod-eg1-sanity, eval=FALSE, echo=TRUE}
-sanity(fit)
+# sanity(fit)
 ```
@@
 ```{r pcod-eg1-suppress-depth2, eval=FALSE, echo=FALSE}
-qcs_grid_original <- qcs_grid
-qcs_grid$depth_scaled <- NULL # for simplicity in paper
-qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
+# qcs_grid_original <- qcs_grid
+# qcs_grid$depth_scaled <- NULL # for simplicity in paper
+# qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
 ```
@@
 ```{r qcs-head, echo=FALSE}
-head(qcs_grid, n = 3)
+# head(qcs_grid, n = 3)
 ```
@@
 ```{r pcod-eg1-predict, echo=FALSE, eval=TRUE, cache=TRUE}
-p <- predict(fit, newdata = qcs_grid)
+# p <- predict(fit, newdata = qcs_grid)
 ```
@@
 ```{r pcod-p-tibble, eval=FALSE}
-p <- tibble::as_tibble(p)
+# p <- tibble::as_tibble(p)
 ```
@@
 ```{r pcod-head-predict, echo=TRUE, eval=FALSE}
-head(p, n = 3)
+# head(p, n = 3)
 ```
@@
 ```{r time1, cache=TRUE, eval=FALSE}
-t1 <- Sys.time()
+# t1 <- Sys.time()
 ```
@@
 ```{r fit-tv, eval=FALSE, echo=TRUE, cache=TRUE, results='hide'}
-fit_spatiotemporal <- sdmTMB(
-  density ~ 0 + s(depth, k = 5),
-  time_varying = ~ 1,
-  time_varying_type = "rw",
-  family = tweedie(link = "log"),
-  data = pcod,
-  mesh = mesh,
-  time = "year",
-  spatial = "on",
-  spatiotemporal = "iid",
-  anisotropy = TRUE,
-  silent = FALSE
-)
+# fit_spatiotemporal <- sdmTMB(
+#   density ~ 0 + s(depth, k = 5),
+#   time_varying = ~ 1,
+#   time_varying_type = "rw",
+#   family = tweedie(link = "log"),
+#   data = pcod,
+#   mesh = mesh,
+#   time = "year",
+#   spatial = "on",
+#   spatiotemporal = "iid",
+#   anisotropy = TRUE,
+#   silent = FALSE
+# )
 ```
@@
 ```{r time2, cache=TRUE, eval=FALSE}
-t2 <- Sys.time()
+# t2 <- Sys.time()
 ```
@@
 ```{r fit-tv-print, eval=FALSE, echo=TRUE}
-fit_spatiotemporal
+# fit_spatiotemporal
 ```
@@
 ```{r pcod-restore, echo=FALSE, eval=FALSE}
-pcod <- pcod_original
-qcs_grid <- qcs_grid_original
+# pcod <- pcod_original
+# qcs_grid <- qcs_grid_original
 ```
@@
 ```{r owl-fit-old, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
-mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
-fit_owls <- sdmTMB(
-  count ~ nao + (1 | year_factor),
-  spatial_varying = ~ nao,
-  family = nbinom2(link = "log"),
-  data = snow,
-  mesh = mesh,
-  time = "year",
-  spatial = "on",
-  spatiotemporal = "iid"
-)
+# mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
+# fit_owls <- sdmTMB(
+#   count ~ nao + (1 | year_factor),
+#   spatial_varying = ~ nao,
+#   family = nbinom2(link = "log"),
+#   data = snow,
+#   mesh = mesh,
+#   time = "year",
+#   spatial = "on",
+#   spatiotemporal = "iid"
+# )
 ```

### 2023-09-08 doc/paper-jss/paper-jss.Rmd
Overhauling examples

--- 2023-09-08 doc/paper-jss/paper-jss.Rmd
@@
   \usepackage{bm}
-  \usepackage{pdflscape}
+  \usepackage{lscape}
   \usepackage{tablefootnote}
@@
 
-# Introductory example using fish survey data
-
-## Illustrating the need for spatial random fields
+# Installation
+
+\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive
+Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>
+
+```{r sdmTMB-install, eval=FALSE, echo=TRUE}
+install.packages("sdmTMB", dependencies = TRUE)
+```
+
+Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package
+[@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed
+for the `make_mesh()` function below. Soon, we intend to switch to using the
+\pkg{fmesher} package to eliminate this dependency.
+
+Alternatively, the development version can be installed with:
+
+```{r remotes, eval=FALSE, echo=TRUE}
+# install.packages("remotes")
+remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
+```
+
+Additional utilities are maintained in the \pkg{sdmTMBextra} package at
+<https://github.com/pbs-assess/sdmTMBextra>.
+
+# Species distribution modelling with spatial random fields
 
 We begin with a simple species distribution model of encounter probability of
-Pacific Cod (TODO spp) from a trawl survey conducted off the west
-coast of Vancouver Island, British Columbia, Canada.
-This survey consists of spatially random sampling in each year, resulting in a georeferenced dataset.
-For compatibility with other packages, spatial random fields in \pkg{sdmTMB} can be toggled on and off.
-We will fit a logistic
-regression of encounter probability with and without spatial random fields to
-illustrate the importance of accounting for spatial correlation. In addition to the spatial random field, we include an intercept and a quadratic effect of depth on the probability of occurrence.
+Pacific Cod (*Gadus macrocephalus*) from a trawl survey conducted in Queen
+Charlotte Sound, British Columbia, Canada. The purpose of our example is to
+illustrate the need for spatial random fields. This survey conducts
+depth-stratified random sampling biennially, resulting in a georeferenced
+dataset. The dataset includes columns for year, latitude, longitude, Universal
+Transverse Mercator (UTM) zone 9 coordinates, bottom depth, and encounter vs.
+non-encounter.
 
@@
 
-We can create a 'mesh' for the SPDE calculations with `make_mesh()`.
+```{r pcod-head, echo=TRUE}
+select(pcod, year, lat, lon, X, Y, depth, present) |> 
+  head(n = 3)
+```
+
+An \pkg{sdmTMB} model requires a data frame that contains a response column,
+columns for any predictors, and columns for spatial coordinates.
+Usually it makes sense to convert the spatial coordinates to an equidistant
+projection such as the Universal Transverse Mercator (UTM) to ensure that
+distance remains constant throughout the study region [e.g., using
+`sf::st_transform()`, @pebesma2018]. Here we use the helper function
+`add_utm_columns()` to add UTM coordinates with km units (so our estimated
+spatial range parameter is not too big or small). By default, the function will
+guess the UTM zone and create new columns `X` and `Y`. Since our example
+data set already has these UTM columns, we can skip running this code.
+
+```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
+pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+```
+
+We then create a mesh object that contains triangulation and projection
+matrices needed to apply the SPDE approach. Here, `cutoff` defines the minimum
+allowed distance between mesh vertices in the units of `X` and `Y`.
+`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` and `cutoff`
+is passed through to it. Soon, \pkg{sdmTMB} will switch to using the
+\pkg{fmesher} package for this purpose, which is hosted on CRAN. Alternatively,
+we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and
+supplied it to the `mesh` argument in `make_mesh()`. We can inspect our mesh
+object with the associated plotting method.
 
@@
 
-We will fit a model with and without spatial random fields by toggling the `spatial` argument from `"on"` to `"off`. The `update()` S3 method refits the same model updating any specified arguments.
-
-```{r}
+We can then fit our model with `sdmTMB()`. We will fit a logistic regression of
+encounter probability with and without spatial random fields (by toggling the
+`spatial` argument) to illustrate the importance of accounting for spatial
+correlation. In addition to the spatial random field, we include an intercept
+and a quadratic effect of depth on the probability of encounter. The `update()`
+S3 method refits the same model updating any specified arguments.
+
+Our random field model can be written as TODO.
+
+```{r, echo=TRUE}
 fit_bin_rf <- sdmTMB(
@@
 
-We can get a summary of our model fit by printing it.
-
-```{r dog-bin-summary, echo=TRUE}
-fit_bin_rf
-```
-
-Describe what we see here
-
-We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals.
+We can run some basic checks on our model with the `sanity()` function:
+
+```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
+sanity(fit)
+```
+
+```
+#> ✓ Non-linear minimizer suggests successful convergence
+#> ✓ Hessian matrix is positive definite
+#> ✓ No extreme or very small eigenvalues detected
+#> ✓ No gradients with respect to fixed effects are >= 0.001
+#> ✓ No fixed-effect standard errors are NA
+#> ✓ No fixed-effect standard errors look unreasonably large
+#> ✓ No sigma parameters are < 0.01
+#> ✓ No sigma parameters are > 100
+#> ✓ Range parameter doesn't look unreasonably large
+```
+
+Here, this does not flag any issues. `sanity()` is checking that the `nlminb()`
+optimizer reported successful convergence, that the Hessian matrix is positive
+definite, that no extreme (XX) or small (XX) eigenvalues are detected, that no
+absolute log likelihood gradients with respect two fixed defects are >= 0.001,
+that all fixed effects have reported standard errors and that they do not look
+unreasonably large (XX by default), that random field marginal standard
+deviations are not unexpectedly small or large, and that the random field
+Matérn range parameter does not look unreasonably large.
+
+We can get a summary of our model fit:
+
+```{r pcod-bin-summary, echo=TRUE}
+summary(fit_bin_rf)
+```
+
+The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
+We also see the formula, mesh, fitted data, and family. Next we see any
+estimated main effects, the Matérn range distance, the spatial random field
+standard deviation, and the negative log likelihood at convergence.
+
+<!-- The -->
+<!-- smoother summary follows the format used in the \pkg{brms} package -->
+<!-- [@burkner2017]. As noted in the print method, we can extract a tidy data frame -->
+<!-- similar to the format in the \pkg{broom} [@broom] and \pkg{broom.mixed} -->
+<!-- [@broommixed] packages: -->
 
@@
 }
-test_autocor(fit_bin)
-test_autocor(fit_bin_rf)
-```
-
-We can also see that the AIC of the model with spatial random fields is considerably lower:
-
-```{r}
+(t_no_rf <- test_autocor(fit_bin))
+(t_rf <- test_autocor(fit_bin_rf))
+p_rf <- round(t_rf$p.value, 2)
+```
+
+We can test for spatial autocorrelation with a Moran's I test or with a visual
+inspection of the residuals calculated from the `residuals()` S3 method. We
+omit the Moran's I code for brevity but include it in the code supplement. The
+p-value for spatial autocorrelation is low for the model without a random field
+(p < 0.01) but large for the model with a random field (p = `r p_rf`)
+suggesting significant residual correlation in the simpler model that is
+alleviated by including the random field.
+
+We can also see that the AIC of the model with spatial random fields is
+considerably lower:
+
+```{r pcod-aic, eval=TRUE, echo=TRUE, results='markup'}
 AIC(fit_bin_rf, fit_bin)
@@
 
-We could do additional testing with cross validation via `sdmTMB_cv()`.
-Here we will do 10-fold cross validation with the folds constructed randomly.
-Using the `fold_ids` argument, we could supply our own folds and conduct spatially blocked cross validation, for example.
-
-```{r}
+We could do additional testing with cross validation via `sdmTMB_cv()`. Here we
+will do 10-fold cross validation with the folds constructed randomly. Using the
+`fold_ids` argument, we could supply our own folds and conduct spatially
+blocked cross validation.
+
+```{r pcod-cv, eval=TRUE, echo=TRUE}
 cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
   data = pcod, mesh = mesh_pcod, spatial = "on",
-  family = binomial(link = "logit"), k_folds = 10
+  family = binomial(), k_folds = 10
 )
-cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
+cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
   data = pcod, mesh = mesh_pcod, spatial = "off",
-  family = binomial(link = "logit"), k_folds = 10
+  family = binomial(), k_folds = 10
 )
@@
 
-Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields:
-
-```{r}
+Indeed, the log likelihood of left-out data is considerably larger for the
+model that includes random fields:
+
+```{r pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'}
 cv_bin_rf$sum_loglik
-cv_bin_rf$sum_loglik
-```
-
-The standard errors on our fixed effects have increased:
-
-```{r}
-tidy(fit_bin_rf, conf.int = T)
-tidy(fit_bin, conf.int = T)
-```
-
-We now have an effect of depth and a spatial random field, which combined in link space result in our overall prediction.
-Note the difference between only the fixed effects (left) and including the spatial random field (right).
-
-```{r}
+cv_bin$sum_loglik
+```
+
+The standard errors on our fixed effects have increased with the spatial random
+field:
+
+```{r pcod-tidy, eval=TRUE, echo=TRUE, results='markup'}
+tidy(fit_bin_rf, conf.int = TRUE)
+tidy(fit_bin, conf.int = TRUE)
+```
+
+\sloppy
+To visualize our model, we can make predictions with the `predict()` method
+(`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on
+a new data frame with any locations and values for the predictor columns. Here,
+we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire
+region of interest so we can visualize the predictions spatially. The grid
+contains spatial covariate columns and all predictors used in the model set at
+values for which we want to predict. The output of `predict()` is a data frame
+containing overall estimates in link space (`est`), estimates from the
+non-random-field components (`est_non_rf`; here, intercept and depth), and
+estimates from the individual random field components (here, `omega_s`---the
+spatial field).
+
+```{r pcod-predict, echo=TRUE, results='markup'}
 p <- predict(fit_bin_rf, newdata = qcs_grid)
+select(p, X, Y, depth, est, est_non_rf, omega_s) |>
+  as_tibble() |> head(n = 2)
+```
+
+TODO MOVE TO CAPTION? We have an effect of depth and a spatial random field,
+which combined in link space result in our overall prediction. Note the
+difference between only the fixed effects (left) and including the spatial
+random field (right). TODO DESCRIBE PLOTTING AND/OR INCLUDE
+
+```{r pcod-predict-maps, fig.width=11, fig.asp=0.3, out.width="6in", fig.cap="TODO"}
 plot_spatial_map <- function(dat, column, title) {
@@
 
-As a second example, we will construct a model of catch rates of North Pacific spiny dogfish from the same survey.
-This example extends the spatial model described previously by also including (1) spatiotemporal fields allowing the distribution in each year to be unique (modelled as an autoregressive process, allowing hotspots to persist through time), (2) a time varying intercept to allow each year to have a different mean catch rate, and (3) a smooth effect of ln(depth), allowing catch rates to vary non-linearly with depth.
-As catch rates are both skewed and zero inflated, we specify the response family as a Tweedie distribution [@tweedie1984].
+As a second example, we will construct a spatiotemporal model of catch rates of
+Pacific Spiny Dogfish (*Squalus suckleyi*) from a trawl survey off the west
+coast of Vancouver Island, Canada. This example extends the spatial model
+described above by also including (1) spatiotemporal fields allowing the
+distribution in each year to be unique, (2) a time-varying intercept as an
+AR(1) process to allow each year to have a different mean catch rate, (3)
+a smooth effect of depth, allowing catch rates to vary non-linearly with
+depth, and (4) spatial anisotropy allowing spatial correlation to be directionally
+dependent. Since catch rates are positive, continuous, and contain zeros, we begin
+by specifying the response family as a Tweedie distribution [@tweedie1984] with
+a log link. We then compare alternative families, spatiotemporal random field
+structures, and the inclusion of anisotropy to illustrate the flexibility of
+\pkg{sdmTMB}.
+
+<!-- We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`. -->
+<!-- We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package. -->
+<!-- We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros. -->
+<!-- An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument). -->
+
+The dataset includes spatial coordinates, year, dogfish catch weight in kg,
+area swept in km, and bottom depth:
 
 ```{r dog-head-dat, echo=TRUE}
-dat <- select(dogfish, longitude, latitude, year, catch_weight, area_swept, depth)
-head(dat)
-```
+dat <- select(dogfish, longitude, latitude, year, 
+  catch_weight, area_swept, depth)
+head(dat, n = 3)
+```
+
+We add UTM zone 9 columns, create a log depth column for convenience, and
+create a basic mesh:
 
 ```{r dog-utms, echo=TRUE}
-dat <- add_utm_columns(dat, c("longitude", "latitude"), units = "km", utm_crs = 32609)
+dat <- add_utm_columns(dat, c("longitude", "latitude"), 
+  units = "km", utm_crs = 32609)
 dat$log_depth <- log(dat$depth)
-```
-
-```{r dog-mesh, echo=TRUE}
 mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@@
 ```
+
+We can then specify our model. We include an offset for the effort variable log
+area swept such that we are effectively modelling density and our predictions
+will be for an area swept of 1 km^2^.
+
+Our model can be written as TODO
 
@@
   anisotropy = TRUE,
-  share_range = TRUE,
   silent = FALSE,
@@
 
-We can use `update()` to refit our model quickly without specifying all of the arguments.
-This is highlighted by exploring models with four additional families, each using a delta- or hurdle distribution:
-delta-gamma, delta-lognormal, delta-Poisson-link-gamma, and
-delta-Poisson-link-lognormal.
+We can explore four alternative families that may better represent the data.
+Each alternative family uses a delta or hurdle distribution: delta-gamma and
+delta-lognormal (REF), and delta-Poisson-link-gamma and
+delta-Poisson-link-lognormal (REF). Note that we choose to omit the
+spatiotemporal random fields from the first model component 
+(`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
+collapsing to zero indicating minimal changes in encounter spatial distribution
+from year to year.
+
+TODO EQUATIONS
 
 ```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
-fit_dg <- update(fit_tw, family = delta_gamma(), spatiotemporal = list("off", "iid"))
+fit_dg <- update(fit_tw, family = delta_gamma(), 
+  spatiotemporal = list("off", "iid"))
 fit_dl <- update(fit_dg, family = delta_lognormal())
@@
 
-We can then compare the fits via AIC:
+We can then compare the models via AIC:
 
@@
 test two additional model formulations: making the spatial correlation
-isotropic (the default) instead of anisotropic and structuring the positive model component
-spatiotemporal random fields as AR(1).
+isotropic (the default) instead of anisotropic, and structuring the positive
+model component spatiotemporal random fields as AR(1) to allow spatiotemporal
+hotspots to persist.
+
+The AR(1) fields can be represented as TODO.
 
@@
 
-<!-- Need description of anisotropy and why it's important here or somewhere else -->
-We find that the anisotropic AR(1) is favoured. On inspecting the model output
-with `summary()`, we notice that the spatial random field SD in the positive model component has become very
-small once AR(1) spatiotemporal fields were added. We therefore consider
-a model where these are turned off in the second component of the delta model:
+<!-- Need description of anisotropy and why it's important here or somewhere
+else -->
+
+We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
+is important here given the elongated shape of the continental shelf with
+a rapid transition to deeper water. On inspecting the model output with
+`summary()`, we notice that the spatial random field SD in the positive model
+component has become very small once AR(1) spatiotemporal fields were added. We
+therefore consider a model where these are turned off in the second component
+of the delta model:
 
@@
 ```
+
+We save our chosen model to the object `fit` to simplify subsequent code, run
+the `sanity()` check (suppressed for space), and inspect `summary()`:
 
@@
 
-```{r dog-aniso, echo=TRUE}
+The output is considerably more complex compared to our binomial spatial model.
+We now have two model components, which are shown one after the other. Starting
+with the binomial component, we have output from the smoother, which includes
+a linear component (`slog_depth`) and the standard deviation on the smoother
+weights (`sds(log_depth)`). The smoother summary follows the format used in the
+\pkg{brms} package [@burkner2017]. Next, we have the time-varying intercepts
+and information on our anisotropic spatial correlation. We then have the second
+model component (lognormal) with a similar summary structure but with the
+addition of a dispersion parameter for the lognormal, the AR(1) correlation of
+the spatiotemporal random fields, and a spatiotemporal random field marginal
+standard deviation. We can use `plot_anisotropy()` to visually inspect the
+anisotropy (Figure \ref{fig:dog-aniso}).
+
+```{r dog-aniso, echo=TRUE, fig.cap="TODO", out.width="4in"}
 plot_anisotropy(fit_dpl)
 ```
+
+Similarly to the first example, we can visualize model predictions on a grid
+covering the area of interest. Because this is a spatiotemporal model, we first
+need to replicate our grid for each year we will predict on. Since this is
+a common operation, we include the convenience function `replicate_df()` to
+replicate a data frame. We then ensure our data frame contains all the
+predictors used in the model (here `log_depth`).
 
@@
 ```{r dog-grid, echo=TRUE}
-head(wcvi_grid, n = 3)
 grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
 grid$log_depth <- log(grid$depth)
-head(grid, n = 3)
+head(grid, n = 2)
 ```
@@
 
-Our prediction data frame has..., which we can easily plot with ggplot2 or similar. We suppress that code in this paper for brevity.
-
 ```{r dog-pred2, echo=TRUE}
@@
 ```
+
+Our prediction data frame is similar to the binomial spatial model, but
+includes columns for the two delta model components (labelled with suffixes `1`
+and `2`) and adds an `epsilon_st` column for spatiotemporal random effects. We can
+easily generate plots from this data frame with ggplot2 or similar. We suppress
+that code in this paper for brevity.
 
@@
 
-```{r dog-wvci-pred, fig.asp = 0.66, fig.cap="Example prediction components..."}
+```{r dog-wvci-pred, fig.asp = 0.5, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
 cowplot::plot_grid(
@@
 
-```{r dog-depth, echo=TRUE}
+We can visualize the conditional effect of the bottom depth smoother by
+predicting across a sequence of depths and holding other variables at reference
+values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
+to include both delta model components (`model = NA`), omit the random fields
+(`re_form = NA`), and return standard errors (`se_fit = TRUE`).
+
+```{r dog-depth, echo=TRUE, out.width="5in"}
 nd <- data.frame(
@@
 )
-pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA, model = NA)
-```
-
-```{r dog-depth-plot, echo=FALSE}
+pred_depth <- predict(
+  fit, newdata = nd, 
+  model = NA, re_form = NA, se_fit = TRUE
+)
+```
+
+```{r dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
 ggplot(pred_depth, aes(
@@
   geom_line() +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
   labs(x = "Depth (m)", y = "Density")
 ```
+
+We can generate an area-weighted population index (e.g., a relative or absolute
+index of abundance or biomass) that is independent of sampling locations by
+predicting from the model on a grid covering the area of interest and summing
+the predicted biomass with the `get_index()` function (Figure
+\ref{fig:dog-index-plot}). We supply the grid cell area (4 km^2^) to the `area`
+argument and specify `bias_correct = TRUE` to enable a generic epsilon-bias
+correction needed due to the non-linear transformation of the random effects
+(REF).
 
@@
 
-```{r dog-index-plot, echo=TRUE}
+```{r dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
 ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
@@
 
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
-The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
-The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
-Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+In our final example, we demonstrate a model with spatially varying coefficient
+(SVC) effects and illustrate combining uncertainty from parameters by working
+with draws from the joint parameter precision matrix. Spatially varying
+coefficients ... TODO REFS. In addition, we illustrate using restricted maximum
+likelihood (REML), which integrates over the fixed effects as well as the
+random effects.
+
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive
+migrants, meaning that they appear across the mid-latitudes of North America in
+much greater numbers in some winters than others. The reasons for this
+interannual variation in the number of individuals migrating south are not well
+understood but seem to be related to high abundances of food during the
+breeding season and therefore sharp increases in breeding ground population
+densities [@robillard2016]. The North Atlantic Oscillation Index (NAO) has been
+linked to productivity of both owls and their prey in Europe [@millon2014].
+Because both productivity and the choice of wintering location could be
+influenced by climate, we modelled a spatially varying effect of annual mean
+NAO index on winter abundance across the southern boundary of their winter
+distribution. We fit counts observed in North America during annual Christmas
+Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution,
+random intercepts for year, spatial and spatiotemporal random fields, and
+a spatially varying coefficient associated with the NAO.
 
@@
 
-The SVC effect at any point in space is the combination of the fixed effect and spatial random effect for `nao`. Here we illustrate
-
-We can extract the SVC effect with TODO
+The spatially varying coefficient effect at any point in space is the
+combination of the main effect and spatial random effect for `nao`. Here we
+illustrate deriving uncertainty on their combination by simulating from the
+parameter precision matrix assuming the joint parameter covariance matrix in
+multivariate normal. We do this by specifying a non-null number of simulation
+draws to `nsim` in `predict.sdmTMB()`. By default, `nsim > 0` will return
+a matrix of draws from the overall prediction. Here, we we instead specify that
+we want to return draws from the `zeta_s` random field, which is the spatially
+varying random field. This returns a matrix where each row matches a row of
+`newdata` and each column is a simulation draw. We then use the function
+`spread_sims()` to draw 200 simulations from the parameters themselves, combine
+the main effect draws (`sims$nao`) with the random field values `zeta_s`, and
+calculate the median for each row of data so we can plot them.
 
 ```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
-nsim <- 200
-zeta_s <- predict(fit_owl, newdata = snow, nsim = nsim, sims_var = "zeta_s")
-sims <- spread_sims(fit_owl, nsim = nsim)
+zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
+dim(zeta_s)
+sims <- spread_sims(fit_owl, nsim = 200)
 combined <- sims$nao + t(zeta_s)
@@
 
-We can make a basic plot using the following code. A more elaborate version is shown in Figure \ref(fig:owl-plot-fancy).
+We can make a basic plot using the following code. A more elaborate version is
+shown in Figure \ref{fig:owl-plot-fancy}.
 
 ```{r owl-plot-basic, echo=TRUE, eval=FALSE}
-ggplot(snow, aes(X, Y)) +
-  geom_point(aes(colour = nao_effect)) +
-  scale_colour_viridis_c()
-```
-
-Overall, we found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure \ref{fig:owl-nao}).
-This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
-We provide an extended version of this example in Appendix \ref{app:owls}.
+ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
+```
+
+Overall, we find a weak average positive effect of annual mean NAO on overall
+counts, but a southeast to northwest gradient in the intensity of the effect
+(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
+to the Atlantic coast and those migrating the furthest south being the most
+affected by NAO.
+
+<!-- TODO keep this? -->
+<!-- We provide an extended version of this example in Appendix \ref{app:owls}. -->
 
@@
 
-```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Map of spatially varying coefficient effects of NAO on Snowy Owl count.
-In the west, the lower values are near 1 implying no effect, whereas in the southeast the effect becomes positive.
-Points represent all count locations."}
+```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
+Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
+The effect is multiplicative on owl count per NAO unit.
+In the west, the lower values are near 1 implying no effect, whereas in the southeast the effect becomes positive."}
 nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
@@
 
-\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>
-
-```{r sdmTMB-install, eval=FALSE, echo=TRUE}
-install.packages("sdmTMB", dependencies = TRUE)
-```
-
-Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package  [@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed for the `make_mesh()` function below.
-
-Here, we illustrate a spatial model fit to Pacific Cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
-Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
-Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific Cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.
-
-```{r sdmTMB-lib0, eval=FALSE, echo=FALSE}
-# library("sdmTMB")
-# # simplify for paper:
-pcod_original <- pcod # save it for later
-# pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
-#   tibble::as_tibble()
-
-```
-
-```{r sdmTMB-lib, eval=FALSE, echo=TRUE}
-library("sdmTMB")
-dplyr::select(pcod, year, density, depth, lon, lat) |>
-  head(n = 3)
-```
-
-<!-- An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. -->
-Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
-Here we use the helper function `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
-By default, the function will guess our UTM zone and create new columns `X` and `Y`.
-
-```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
-pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
-```
-
-```
-#> Detected UTM zone 9N; CRS = 32609.
-#> Visit https://epsg.io/32609 to verify.
-```
-
-```{r pcod-head2, echo=TRUE, eval=FALSE}
-head(pcod, n = 3)
-```
-
-We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach.
-
-```{r, echo=TRUE, eval=FALSE, cache=TRUE}
-mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
-```
-
-Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
-`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` and `cutoff` is passed through to it.
-Alternatively, we could have created any mesh via the \proglang{R}-\pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
-We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Figure \ref{fig:pcod-fig}A).
-
-We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
-We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
-We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros.
-An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).
-
-```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE, cache=TRUE}
-fit <- sdmTMB(
-  density ~ s(depth),
-  data = pcod,
-  family = tweedie(link = "log"),
-  mesh = mesh,
-  spatial = "on"
-)
-```
-
-We can get a summary of the fit with the `print()` method:
-
-```{r pcod-eg1-summary,eval=FALSE, echo=TRUE}
-fit
-```
-
-The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
-We also see the formula, mesh, fitted data, and family.
-Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
-The smoother summary follows the format used in the \pkg{brms} package [@burkner2017].
-As noted in the print method, we can extract a tidy data frame similar to the format in the \pkg{broom} [@broom] and \pkg{broom.mixed} [@broommixed] packages:
-
 ```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
-tidy(fit, conf.int = TRUE)
+# tidy(fit, conf.int = TRUE)
 ```
@@
 ```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
-tidy(fit, effects = "ran_pars", conf.int = TRUE)
-```
-
-The latter `tidy()` call returns all variance components of the model and omits the standard errors since these are calculated in log space.
-
-We can run some basic checks on our model with the `sanity()` function:
-
-```{r pcod-eg1-sanity, eval=FALSE, echo=TRUE}
-sanity(fit)
-```
-
-```
-#> ✓ Non-linear minimizer suggests successful convergence
-#> ✓ Hessian matrix is positive definite
-#> ✓ No extreme or very small eigenvalues detected
-#> ✓ No gradients with respect to fixed effects are >= 0.001
-#> ✓ No fixed-effect standard errors are NA
-#> ✓ No fixed-effect standard errors look unreasonably large
-#> ✓ No sigma parameters are < 0.01
-#> ✓ No sigma parameters are > 100
-#> ✓ Range parameter doesn't look unreasonably large
-```
-
-We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
-Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially.
-The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
-
-```{r pcod-eg1-suppress-depth2, eval=FALSE, echo=FALSE}
-qcs_grid_original <- qcs_grid
-qcs_grid$depth_scaled <- NULL # for simplicity in paper
-qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
-```
-
-```{r qcs-head, echo=FALSE}
-head(qcs_grid, n = 3)
-```
-
-```{r pcod-eg1-predict, echo=FALSE, eval=TRUE, cache=TRUE}
-p <- predict(fit, newdata = qcs_grid)
-```
-
-The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (`est_non_rf`; here, intercept and depth), estimates from the random field components combined (`est_rf`), and estimates from the individual random field components (here, `omega_s`---the spatial field).
-
-```{r pcod-p-tibble, eval=FALSE}
-p <- tibble::as_tibble(p)
-```
-
-```{r pcod-head-predict, echo=TRUE, eval=FALSE}
-head(p, n = 3)
-```
-
-We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Figure \ref{fig:pcod-fig}B--D.
-
-\begin{figure}
-\centering
-\includegraphics[width=5.2in]{../../figs/pcod-fig}
-\caption[]{Output from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada.
-(A) SPDE mesh (lines) combined with the trawl survey observations (points).
+# tidy(fit, effects = "ran_pars", conf.int = TRUE)
+```
+
+<!-- The latter `tidy()` call returns all variance components of the model and omits the standard errors since these are calculated in log space. -->
+
+<!-- We can run some basic checks on our model with the `sanity()` function: -->
+
+<!-- We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
+-->
+<!-- Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially. -->
+<!-- The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict. -->
+
+<!-- \begin{figure} -->
+<!-- \centering -->
+<!-- \includegraphics[width=5.2in]{../../figs/pcod-fig} -->
+<!-- \caption[]{Output from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada. -->
+<!-- (A) SPDE mesh (lines) combined with the trawl survey observations (points).
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
 The circle area corresponds to biomass density of Pacific Cod caught on individual trawls.
-(B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
+-->
+<!-- (B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
 (C) Overall prediction: these estimates represent the combination of all fixed and random effects.
 Crosses illustrate center of gravity (\texttt{get\_cog()}) from a spatiotemporal version; lighter crosses are more recent.
-(D) Conditional effect of depth modelled as a penalized smoother.
-These predictions are made omitting the spatial random field.
-(E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}).
-The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig}
-\end{figure}
-
-We could extend this to a spatiotemporal model simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
-Here, we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
+-->
+<!-- (D) Conditional effect of depth modelled as a penalized smoother. -->
+<!-- These predictions are made omitting the spatial random field. -->
+<!-- (E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}).
+The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig} -->
+<!-- \end{figure} -->
+
+<!-- We could extend this to a spatiotemporal model simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
+-->
+<!-- Here, we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
+-->
 
 \clearpage
-
-```{r time1, cache=TRUE, eval=FALSE}
-t1 <- Sys.time()
-```
-
-```{r fit-tv, eval=FALSE, echo=TRUE, cache=TRUE, results='hide'}
-fit_spatiotemporal <- sdmTMB(
-  density ~ 0 + s(depth, k = 5),
-  time_varying = ~ 1,
-  time_varying_type = "rw",
-  family = tweedie(link = "log"),
-  data = pcod,
-  mesh = mesh,
-  time = "year",
-  spatial = "on",
-  spatiotemporal = "iid",
-  anisotropy = TRUE,
-  silent = FALSE
-)
-```
-
-```{r time2, cache=TRUE, eval=FALSE}
-t2 <- Sys.time()
-```
-
-```{r fit-tv-print, eval=FALSE, echo=TRUE}
-fit_spatiotemporal
-```
-
-We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
-We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Figure \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).
-
-```{r pcod-restore, echo=FALSE, eval=FALSE}
-pcod <- pcod_original
-qcs_grid <- qcs_grid_original
-```
-
-# Spatially varying coefficient example with citizen science data
-
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
-The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
-The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
-Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
-We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
-
-```{r owl-fit-old, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
-mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
-fit_owls <- sdmTMB(
-  count ~ nao + (1 | year_factor),
-  spatial_varying = ~ nao,
-  family = nbinom2(link = "log"),
-  data = snow,
-  mesh = mesh,
-  time = "year",
-  spatial = "on",
-  spatiotemporal = "iid"
-)
-```
-
-We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure \ref{fig:owl-nao}).
-This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
-We provide an extended version of this example in Appendix \ref{app:owls}.
-
-\clearpage
-
-```{r owl-nao, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
-Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
-The effect is multiplicative on owl count per NAO unit.", out.width="4.1in", fig.align='center'}
-owl <- here::here("figs", "owl-nao-effect.png")
-knitr::include_graphics(owl)
-```
 

### 2023-09-09 doc/paper-jss/paper-jss.Rmd
Start adding eqns

@@
 Charlotte Sound, British Columbia, Canada. The purpose of our example is to
-illustrate the need for spatial random fields. This survey conducts
-depth-stratified random sampling biennially, resulting in a georeferenced
-dataset. The dataset includes columns for year, latitude, longitude, Universal
-Transverse Mercator (UTM) zone 9 coordinates, bottom depth, and encounter vs.
-non-encounter.
+illustrate the need for spatial random fields. This survey is conducted by
+Fisheries and Oceans Canada (REF) and follows a depth-stratified random
+sampling design, resulting in a georeferenced dataset. The data frame `pcod` is
+available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
+longitude, Universal Transverse Mercator (UTM) coordinates, bottom
+depth, and encounter (`present = 1`) vs. non-encounter (`present = 0`).
 
@@
 ```{r pcod-head, echo=TRUE}
-select(pcod, year, lat, lon, X, Y, depth, present) |> 
+select(pcod, lat, lon, X, Y, depth, present) |> 
   head(n = 3)
@@
 An \pkg{sdmTMB} model requires a data frame that contains a response column,
-columns for any predictors, and columns for spatial coordinates.
-Usually it makes sense to convert the spatial coordinates to an equidistant
-projection such as the Universal Transverse Mercator (UTM) to ensure that
-distance remains constant throughout the study region [e.g., using
-`sf::st_transform()`, @pebesma2018]. Here we use the helper function
-`add_utm_columns()` to add UTM coordinates with km units (so our estimated
-spatial range parameter is not too big or small). By default, the function will
-guess the UTM zone and create new columns `X` and `Y`. Since our example
-data set already has these UTM columns, we can skip running this code.
+columns for any predictors, and columns for spatial coordinates. Usually it
+makes sense to convert the spatial coordinates to an equidistant projection
+such as UTMs to ensure that distance remains constant throughout the study
+region [e.g., using `sf::st_transform()`, @pebesma2018]. Here we use the helper
+function `add_utm_columns()` to add UTM coordinates with km units (so our
+estimated spatial range parameter is not too big or small). By default, the
+function will guess the UTM zone and create new columns `X` and `Y`. Since our
+example data already has these UTM columns, we can skip running this code.
 
@@
 
+\sloppy
 We then create a mesh object that contains triangulation and projection
@@
 allowed distance between mesh vertices in the units of `X` and `Y`.
-`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` and `cutoff`
-is passed through to it. Soon, \pkg{sdmTMB} will switch to using the
-\pkg{fmesher} package for this purpose, which is hosted on CRAN. Alternatively,
-we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and
-supplied it to the `mesh` argument in `make_mesh()`. We can inspect our mesh
-object with the associated plotting method.
+`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` or
+`fmesher::fm_rcdt_2d_inla()` and `cutoff` is passed through to these functions.
+Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or
+\pkg{fmesher} and supplied it to the `mesh` argument in `make_mesh()`. We can
+inspect our mesh object with the associated plotting method.
 
@@
 
-Our random field model can be written as TODO.
+Our random field model can be written as
+
+$$
+\begin{aligned}
+y_{\boldsymbol{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\boldsymbol{s}}\right),\\
+\mu_{\boldsymbol{s}} &= \operatorname{logit}^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}} \boldsymbol{\beta} +
+\omega_{\boldsymbol{s}}
+\right),
+\end{aligned}
+$$
+
+where TODO.
+
+$$
+\omega_{\boldsymbol{s}} \sim \mathrm{MVN}(\boldsymbol{0},\boldsymbol{\Sigma}_\omega)
+$$
 
@@
 
+$$
+\begin{aligned}
+\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
+\mu_{\boldsymbol{s},t} &=
+f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
+O_{\boldsymbol{s},t} +
+\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \gamma_t +
+\omega_{\boldsymbol{s}} +
+\epsilon_{\boldsymbol{s},t} 
+\right),
+\end{aligned}
+$$
+
+where TODO.
+
+and TODO
+
+$$
+\begin{aligned}
+  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
+  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
+\end{aligned}
+$$
+
 ```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
@@
 
-TODO EQUATIONS
+TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.
 
@@
 
-We find that the Poisson-link-lognormal model is favoured with AIC. We next
-test two additional model formulations: making the spatial correlation
+We find that the Poisson-link delta-lognormal model is favoured with AIC. We
+next test two additional model formulations: making the spatial correlation
 isotropic (the default) instead of anisotropic, and structuring the positive
 model component spatiotemporal random fields as AR(1) to allow spatiotemporal
-hotspots to persist.
+hotspots to persist from year to year.
 
 The AR(1) fields can be represented as TODO.
+
+$$
+\begin{aligned}
+\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
+\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
+\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\end{aligned}
+$$
+
+where TODO.
 
@@
 We can write the model as TODO
+
+$$
+\begin{aligned}
+\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
+\mu_{\boldsymbol{s},t} &=
+f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
+\alpha_g +
+\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} +
+\omega_{\boldsymbol{s}} +
+\epsilon_{\boldsymbol{s},t} \right),
+\end{aligned}
+$$
+
+where TODO.
 

### 2023-09-09 doc/paper-jss/paper-jss.Rmd
Edits

@@
 
+TODO ADD EQUATIONS
+
 \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
@@
 
-# Installation
-
-\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive
-Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>
-
-```{r sdmTMB-install, eval=FALSE, echo=TRUE}
-install.packages("sdmTMB", dependencies = TRUE)
-```
-
-Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package
-[@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed
-for the `make_mesh()` function below. Soon, we intend to switch to using the
-\pkg{fmesher} package to eliminate this dependency.
-
-Alternatively, the development version can be installed with:
-
-```{r remotes, eval=FALSE, echo=TRUE}
-# install.packages("remotes")
-remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
-```
-
-Additional utilities are maintained in the \pkg{sdmTMBextra} package at
-<https://github.com/pbs-assess/sdmTMBextra>.
-
-# Species distribution modelling with spatial random fields
-
-We begin with a simple species distribution model of encounter probability of
-Pacific Cod (*Gadus macrocephalus*) from a trawl survey conducted in Queen
-Charlotte Sound, British Columbia, Canada. The purpose of our example is to
-illustrate the need for spatial random fields. This survey is conducted by
-Fisheries and Oceans Canada (REF) and follows a depth-stratified random
-sampling design, resulting in a georeferenced dataset. The data frame `pcod` is
-available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
-longitude, Universal Transverse Mercator (UTM) coordinates, bottom
-depth, and encounter (`present = 1`) vs. non-encounter (`present = 0`).
-
-```{r libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
-library(sdmTMB)
-library(dplyr)
-library(ggplot2)
-```
-
-```{r libs-extras}
-theme_set(theme_light())
-options(ggplot2.continuous.fill = "viridis")
-```
-
-```{r pcod-head, echo=TRUE}
-select(pcod, lat, lon, X, Y, depth, present) |> 
-  head(n = 3)
-```
-
-An \pkg{sdmTMB} model requires a data frame that contains a response column,
-columns for any predictors, and columns for spatial coordinates. Usually it
-makes sense to convert the spatial coordinates to an equidistant projection
-such as UTMs to ensure that distance remains constant throughout the study
-region [e.g., using `sf::st_transform()`, @pebesma2018]. Here we use the helper
-function `add_utm_columns()` to add UTM coordinates with km units (so our
-estimated spatial range parameter is not too big or small). By default, the
-function will guess the UTM zone and create new columns `X` and `Y`. Since our
-example data already has these UTM columns, we can skip running this code.
-
-```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
-pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
-```
-
-\sloppy
-We then create a mesh object that contains triangulation and projection
-matrices needed to apply the SPDE approach. Here, `cutoff` defines the minimum
-allowed distance between mesh vertices in the units of `X` and `Y`.
-`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` or
-`fmesher::fm_rcdt_2d_inla()` and `cutoff` is passed through to these functions.
-Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or
-\pkg{fmesher} and supplied it to the `mesh` argument in `make_mesh()`. We can
-inspect our mesh object with the associated plotting method.
-
-```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
-mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
-plot(mesh_pcod)
-```
-
-We can then fit our model with `sdmTMB()`. We will fit a logistic regression of
-encounter probability with and without spatial random fields (by toggling the
-`spatial` argument) to illustrate the importance of accounting for spatial
-correlation. In addition to the spatial random field, we include an intercept
-and a quadratic effect of depth on the probability of encounter. The `update()`
-S3 method refits the same model updating any specified arguments.
-
-Our random field model can be written as
-
-$$
-\begin{aligned}
-y_{\boldsymbol{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\boldsymbol{s}}\right),\\
-\mu_{\boldsymbol{s}} &= \operatorname{logit}^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}} \boldsymbol{\beta} +
-\omega_{\boldsymbol{s}}
-\right),
-\end{aligned}
-$$
-
-where TODO.
-
-$$
-\omega_{\boldsymbol{s}} \sim \mathrm{MVN}(\boldsymbol{0},\boldsymbol{\Sigma}_\omega)
-$$
-
-```{r, echo=TRUE}
-fit_bin_rf <- sdmTMB(
-  present ~ poly(log(depth), 2),
-  data = pcod, mesh = mesh_pcod, 
-  spatial = "on",
-  family = binomial(link = "logit")
-)
-fit_bin <- update(fit_bin_rf, spatial = "off")
-```
-
-We can run some basic checks on our model with the `sanity()` function:
-
-```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
-sanity(fit)
-```
-
-```
-#> ✓ Non-linear minimizer suggests successful convergence
-#> ✓ Hessian matrix is positive definite
-#> ✓ No extreme or very small eigenvalues detected
-#> ✓ No gradients with respect to fixed effects are >= 0.001
-#> ✓ No fixed-effect standard errors are NA
-#> ✓ No fixed-effect standard errors look unreasonably large
-#> ✓ No sigma parameters are < 0.01
-#> ✓ No sigma parameters are > 100
-#> ✓ Range parameter doesn't look unreasonably large
-```
-
-Here, this does not flag any issues. `sanity()` is checking that the `nlminb()`
-optimizer reported successful convergence, that the Hessian matrix is positive
-definite, that no extreme (XX) or small (XX) eigenvalues are detected, that no
-absolute log likelihood gradients with respect two fixed defects are >= 0.001,
-that all fixed effects have reported standard errors and that they do not look
-unreasonably large (XX by default), that random field marginal standard
-deviations are not unexpectedly small or large, and that the random field
-Matérn range parameter does not look unreasonably large.
-
-We can get a summary of our model fit:
-
-```{r pcod-bin-summary, echo=TRUE}
-summary(fit_bin_rf)
-```
-
-The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
-We also see the formula, mesh, fitted data, and family. Next we see any
-estimated main effects, the Matérn range distance, the spatial random field
-standard deviation, and the negative log likelihood at convergence.
-
-<!-- The -->
-<!-- smoother summary follows the format used in the \pkg{brms} package -->
-<!-- [@burkner2017]. As noted in the print method, we can extract a tidy data frame -->
-<!-- similar to the format in the \pkg{broom} [@broom] and \pkg{broom.mixed} -->
-<!-- [@broommixed] packages: -->
-
-```{r}
-# set.seed(1)
-# pcod$resid_bin <- residuals(fit_bin)
-# set.seed(1)
-# pcod$resid_bin_rf <- residuals(fit_bin_rf)
-# g1 <- ggplot(pcod, aes(X, Y, colour = resid_bin)) + geom_point() +
-#   scale_color_gradient2() +
-#   theme(legend.position = "bottom")
-# g2 <- ggplot(pcod, aes(X, Y, colour = resid_bin_rf)) + geom_point() +
-#   scale_color_gradient2() +
-#   theme(legend.position = "bottom")
-# cowplot::plot_grid(g1, g2, ncol = 2)
-```
-
-```{r, echo=FALSE, results='hide'}
-test_autocor <- function(obj) {
-  set.seed(1)
-  s <- simulate(obj, nsim = 500)
-  pr <- predict(obj, type = "response")$est
-  r <- DHARMa::createDHARMa(
-    simulatedResponse = s,
-    observedResponse = pcod$present,
-    fittedPredictedResponse = pr
-  )
-  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
-}
-(t_no_rf <- test_autocor(fit_bin))
-(t_rf <- test_autocor(fit_bin_rf))
-p_rf <- round(t_rf$p.value, 2)
-```
-
-We can test for spatial autocorrelation with a Moran's I test or with a visual
-inspection of the residuals calculated from the `residuals()` S3 method. We
-omit the Moran's I code for brevity but include it in the code supplement. The
-p-value for spatial autocorrelation is low for the model without a random field
-(p < 0.01) but large for the model with a random field (p = `r p_rf`)
-suggesting significant residual correlation in the simpler model that is
-alleviated by including the random field.
-
-We can also see that the AIC of the model with spatial random fields is
-considerably lower:
-
-```{r pcod-aic, eval=TRUE, echo=TRUE, results='markup'}
-AIC(fit_bin_rf, fit_bin)
-```
-
-We could do additional testing with cross validation via `sdmTMB_cv()`. Here we
-will do 10-fold cross validation with the folds constructed randomly. Using the
-`fold_ids` argument, we could supply our own folds and conduct spatially
-blocked cross validation.
-
-```{r pcod-cv, eval=TRUE, echo=TRUE}
-cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
-  data = pcod, mesh = mesh_pcod, spatial = "on",
-  family = binomial(), k_folds = 10
-)
-cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
-  data = pcod, mesh = mesh_pcod, spatial = "off",
-  family = binomial(), k_folds = 10
-)
-```
-
-Indeed, the log likelihood of left-out data is considerably larger for the
-model that includes random fields:
-
-```{r pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'}
-cv_bin_rf$sum_loglik
-cv_bin$sum_loglik
-```
-
-The standard errors on our fixed effects have increased with the spatial random
-field:
-
-```{r pcod-tidy, eval=TRUE, echo=TRUE, results='markup'}
-tidy(fit_bin_rf, conf.int = TRUE)
-tidy(fit_bin, conf.int = TRUE)
-```
-
-\sloppy
-To visualize our model, we can make predictions with the `predict()` method
-(`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on
-a new data frame with any locations and values for the predictor columns. Here,
-we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire
-region of interest so we can visualize the predictions spatially. The grid
-contains spatial covariate columns and all predictors used in the model set at
-values for which we want to predict. The output of `predict()` is a data frame
-containing overall estimates in link space (`est`), estimates from the
-non-random-field components (`est_non_rf`; here, intercept and depth), and
-estimates from the individual random field components (here, `omega_s`---the
-spatial field).
-
-```{r pcod-predict, echo=TRUE, results='markup'}
-p <- predict(fit_bin_rf, newdata = qcs_grid)
-select(p, X, Y, depth, est, est_non_rf, omega_s) |>
-  as_tibble() |> head(n = 2)
-```
-
-TODO MOVE TO CAPTION? We have an effect of depth and a spatial random field,
-which combined in link space result in our overall prediction. Note the
-difference between only the fixed effects (left) and including the spatial
-random field (right). TODO DESCRIBE PLOTTING AND/OR INCLUDE
-
-```{r pcod-predict-maps, fig.width=11, fig.asp=0.3, out.width="6in", fig.cap="TODO"}
-plot_spatial_map <- function(dat, column, title) {
-  ggplot(dat, aes(X, Y, fill = {{ column }})) +
-    geom_raster() +
-    coord_fixed() +
-    theme(legend.position = "bottom") +
-    ggtitle(title)
-}
-g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
-g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
-g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
-cowplot::plot_grid(g1, g2, g3, ncol = 3)
-```
-
-# Spatiotemporal species distribution modeling with dogfish
-
-<!-- Need brief description of dogfish data and model here?? -->
-
-As a second example, we will construct a spatiotemporal model of catch rates of
-Pacific Spiny Dogfish (*Squalus suckleyi*) from a trawl survey off the west
-coast of Vancouver Island, Canada. This example extends the spatial model
-described above by also including (1) spatiotemporal fields allowing the
-distribution in each year to be unique, (2) a time-varying intercept as an
-AR(1) process to allow each year to have a different mean catch rate, (3)
-a smooth effect of depth, allowing catch rates to vary non-linearly with
-depth, and (4) spatial anisotropy allowing spatial correlation to be directionally
-dependent. Since catch rates are positive, continuous, and contain zeros, we begin
-by specifying the response family as a Tweedie distribution [@tweedie1984] with
-a log link. We then compare alternative families, spatiotemporal random field
-structures, and the inclusion of anisotropy to illustrate the flexibility of
-\pkg{sdmTMB}.
-
-<!-- We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`. -->
-<!-- We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package. -->
-<!-- We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros. -->
-<!-- An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument). -->
-
-The dataset includes spatial coordinates, year, dogfish catch weight in kg,
-area swept in km, and bottom depth:
-
-```{r dog-head-dat, echo=TRUE}
-dat <- select(dogfish, longitude, latitude, year, 
-  catch_weight, area_swept, depth)
-head(dat, n = 3)
-```
-
-We add UTM zone 9 columns, create a log depth column for convenience, and
-create a basic mesh:
-
-```{r dog-utms, echo=TRUE}
-dat <- add_utm_columns(dat, c("longitude", "latitude"), 
-  units = "km", utm_crs = 32609)
-dat$log_depth <- log(dat$depth)
-mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
-```
-
-```{r dog-mesh2, eval=FALSE}
-plot(mesh)
-```
-
-We can then specify our model. We include an offset for the effort variable log
-area swept such that we are effectively modelling density and our predictions
-will be for an area swept of 1 km^2^.
-
-Our model can be written as TODO
-
-$$
-\begin{aligned}
-\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
-\mu_{\boldsymbol{s},t} &=
-f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
-O_{\boldsymbol{s},t} +
-\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \gamma_t +
-\omega_{\boldsymbol{s}} +
-\epsilon_{\boldsymbol{s},t} 
-\right),
-\end{aligned}
-$$
-
-where TODO.
-
-and TODO
-
-$$
-\begin{aligned}
-  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
-  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
-\end{aligned}
-$$
-
-```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
-fit_tw <- sdmTMB(
-  catch_weight ~ s(log_depth),
-  data = dat,
-  mesh = mesh,
-  offset = log(dat$area_swept),
-  time = "year",
-  time_varying = ~ 1,
-  time_varying_type = "ar1",
-  spatial = "on",
-  spatiotemporal = "iid",
-  anisotropy = TRUE,
-  silent = FALSE,
-  family = tweedie()
-)
-```
-
-We can explore four alternative families that may better represent the data.
-Each alternative family uses a delta or hurdle distribution: delta-gamma and
-delta-lognormal (REF), and delta-Poisson-link-gamma and
-delta-Poisson-link-lognormal (REF). Note that we choose to omit the
-spatiotemporal random fields from the first model component 
-(`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
-collapsing to zero indicating minimal changes in encounter spatial distribution
-from year to year.
-
-TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.
-
-```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
-fit_dg <- update(fit_tw, family = delta_gamma(), 
-  spatiotemporal = list("off", "iid"))
-fit_dl <- update(fit_dg, family = delta_lognormal())
-fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
-fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
-```
-
-We can then compare the models via AIC:
-
-```{r dog-aic, echo=TRUE}
-AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
-  mutate(delta_AIC = AIC - min(AIC)) |> 
-  arrange(delta_AIC)
-```
-
-We find that the Poisson-link delta-lognormal model is favoured with AIC. We
-next test two additional model formulations: making the spatial correlation
-isotropic (the default) instead of anisotropic, and structuring the positive
-model component spatiotemporal random fields as AR(1) to allow spatiotemporal
-hotspots to persist from year to year.
-
-The AR(1) fields can be represented as TODO.
-
-$$
-\begin{aligned}
-\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
-\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
-\end{aligned}
-$$
-
-where TODO.
-
-```{r dog-ar1, results='hide', echo=TRUE}
-fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
-fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
-```
-
-```{r dog-aci2, echo=TRUE}
-AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
-```
-
-<!-- Need description of anisotropy and why it's important here or somewhere
-else -->
-
-We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
-is important here given the elongated shape of the continental shelf with
-a rapid transition to deeper water. On inspecting the model output with
-`summary()`, we notice that the spatial random field SD in the positive model
-component has become very small once AR(1) spatiotemporal fields were added. We
-therefore consider a model where these are turned off in the second component
-of the delta model:
-
-```{r dog-ar1-2, echo=TRUE, results="hide"}
-fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
-```
-
-```{r dog-aic3, echo=TRUE}
-AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
-```
-
-We save our chosen model to the object `fit` to simplify subsequent code, run
-the `sanity()` check (suppressed for space), and inspect `summary()`:
-
-```{r dog-print, echo=TRUE}
-fit <- fit_dpl_ar1_only
-sanity(fit)
-summary(fit)
-```
-
-The output is considerably more complex compared to our binomial spatial model.
-We now have two model components, which are shown one after the other. Starting
-with the binomial component, we have output from the smoother, which includes
-a linear component (`slog_depth`) and the standard deviation on the smoother
-weights (`sds(log_depth)`). The smoother summary follows the format used in the
-\pkg{brms} package [@burkner2017]. Next, we have the time-varying intercepts
-and information on our anisotropic spatial correlation. We then have the second
-model component (lognormal) with a similar summary structure but with the
-addition of a dispersion parameter for the lognormal, the AR(1) correlation of
-the spatiotemporal random fields, and a spatiotemporal random field marginal
-standard deviation. We can use `plot_anisotropy()` to visually inspect the
-anisotropy (Figure \ref{fig:dog-aniso}).
-
-```{r dog-aniso, echo=TRUE, fig.cap="TODO", out.width="4in"}
-plot_anisotropy(fit_dpl)
-```
-
-Similarly to the first example, we can visualize model predictions on a grid
-covering the area of interest. Because this is a spatiotemporal model, we first
-need to replicate our grid for each year we will predict on. Since this is
-a common operation, we include the convenience function `replicate_df()` to
-replicate a data frame. We then ensure our data frame contains all the
-predictors used in the model (here `log_depth`).
-
-<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
-```{r dog-grid, echo=TRUE}
-grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
-grid$log_depth <- log(grid$depth)
-head(grid, n = 2)
-```
-
-```{r dog-pred1, echo=TRUE}
-pred <- predict(fit, newdata = grid, type = "response")
-```
-
-```{r dog-pred2, echo=TRUE}
-names(pred)
-```
-
-Our prediction data frame is similar to the binomial spatial model, but
-includes columns for the two delta model components (labelled with suffixes `1`
-and `2`) and adds an `epsilon_st` column for spatiotemporal random effects. We can
-easily generate plots from this data frame with ggplot2 or similar. We suppress
-that code in this paper for brevity.
-
-```{r plot-map, echo=FALSE}
-plot_map <- function(dat, column) {
-  ggplot(dat, aes(X, Y, fill = {{ column }})) +
-    geom_raster() +
-    facet_wrap(vars(year)) +
-    coord_fixed() +
-    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
-}
-```
-
-```{r dog-plot1, echo=FALSE}
-g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(est) +
-  # labs(fill = "Density") +
-  scale_fill_viridis_c(trans = "log10")
-```
-
-```{r dog-plot2, echo=FALSE}
-g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(omega_s1) +
-  scale_fill_gradient2()
-```
-
-```{r dog-plot3, echo=FALSE}
-g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(epsilon_st2) +
-  scale_fill_gradient2()
-```
-
-```{r dog-plot4, echo=FALSE}
-g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(est_non_rf1) +
-  scale_fill_viridis_c(trans = "log10")
-```
-
-```{r dog-wvci-pred, fig.asp = 0.5, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
-cowplot::plot_grid(
-  g_nonrf,
-  g_omega,
-  g_eps,
-  g_est,
-  ncol = 2L
-)
-```
-
-We can visualize the conditional effect of the bottom depth smoother by
-predicting across a sequence of depths and holding other variables at reference
-values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
-to include both delta model components (`model = NA`), omit the random fields
-(`re_form = NA`), and return standard errors (`se_fit = TRUE`).
-
-```{r dog-depth, echo=TRUE, out.width="5in"}
-nd <- data.frame(
-  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
-  year = max(dat$year)
-)
-pred_depth <- predict(
-  fit, newdata = nd, 
-  model = NA, re_form = NA, se_fit = TRUE
-)
-```
-
-```{r dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
-ggplot(pred_depth, aes(
-  exp(log_depth), exp(est), 
-  ymin = exp(est - 2 * est_se), 
-  ymax = exp(est + 2 * est_se))) +
-  geom_ribbon(fill = "grey90") +
-  geom_line() +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
-  labs(x = "Depth (m)", y = "Density")
-```
-
-We can generate an area-weighted population index (e.g., a relative or absolute
-index of abundance or biomass) that is independent of sampling locations by
-predicting from the model on a grid covering the area of interest and summing
-the predicted biomass with the `get_index()` function (Figure
-\ref{fig:dog-index-plot}). We supply the grid cell area (4 km^2^) to the `area`
-argument and specify `bias_correct = TRUE` to enable a generic epsilon-bias
-correction needed due to the non-linear transformation of the random effects
-(REF).
-
-```{r dog-index, echo=TRUE}
-pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
-ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
-```
-
-```{r dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
-ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
-  geom_pointrange() +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
-  labs(y = "Biomass", x = "Year")
-```
-
-# Spatially varying coefficient example with citizen science data
-
-In our final example, we demonstrate a model with spatially varying coefficient
-(SVC) effects and illustrate combining uncertainty from parameters by working
-with draws from the joint parameter precision matrix. Spatially varying
-coefficients ... TODO REFS. In addition, we illustrate using restricted maximum
-likelihood (REML), which integrates over the fixed effects as well as the
-random effects.
-
-Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive
-migrants, meaning that they appear across the mid-latitudes of North America in
-much greater numbers in some winters than others. The reasons for this
-interannual variation in the number of individuals migrating south are not well
-understood but seem to be related to high abundances of food during the
-breeding season and therefore sharp increases in breeding ground population
-densities [@robillard2016]. The North Atlantic Oscillation Index (NAO) has been
-linked to productivity of both owls and their prey in Europe [@millon2014].
-Because both productivity and the choice of wintering location could be
-influenced by climate, we modelled a spatially varying effect of annual mean
-NAO index on winter abundance across the southern boundary of their winter
-distribution. We fit counts observed in North America during annual Christmas
-Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution,
-random intercepts for year, spatial and spatiotemporal random fields, and
-a spatially varying coefficient associated with the NAO.
-
-We can write the model as TODO
-
-$$
-\begin{aligned}
-\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
-\mu_{\boldsymbol{s},t} &=
-f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
-\alpha_g +
-\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} +
-\omega_{\boldsymbol{s}} +
-\epsilon_{\boldsymbol{s},t} \right),
-\end{aligned}
-$$
-
-where TODO.
-
-```{r owl-data}
-snow <- readRDS(here::here("data/SNOW_data.rds"))
-```
-
-```{r owl-fit, echo=TRUE, results = "hide"}
-mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
-fit_owl <- sdmTMB(
-  count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ nao,
-  time = "year",
-  data = snow, 
-  family = nbinom2(link = "log"),
-  spatial = "on",
-  spatiotemporal = "iid",
-  mesh = mesh_snow,
-  reml = TRUE,
-  silent = FALSE
-)
-```
-
-```{r owl-sanity, eval=FALSE}
-sanity(fit_owl)
-```
-
-```{r owl-print, echo=TRUE}
-summary(fit_owl)
-```
-
-The spatially varying coefficient effect at any point in space is the
-combination of the main effect and spatial random effect for `nao`. Here we
-illustrate deriving uncertainty on their combination by simulating from the
-parameter precision matrix assuming the joint parameter covariance matrix in
-multivariate normal. We do this by specifying a non-null number of simulation
-draws to `nsim` in `predict.sdmTMB()`. By default, `nsim > 0` will return
-a matrix of draws from the overall prediction. Here, we we instead specify that
-we want to return draws from the `zeta_s` random field, which is the spatially
-varying random field. This returns a matrix where each row matches a row of
-`newdata` and each column is a simulation draw. We then use the function
-`spread_sims()` to draw 200 simulations from the parameters themselves, combine
-the main effect draws (`sims$nao`) with the random field values `zeta_s`, and
-calculate the median for each row of data so we can plot them.
-
-```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
-zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
-dim(zeta_s)
-sims <- spread_sims(fit_owl, nsim = 200)
-combined <- sims$nao + t(zeta_s)
-snow$nao_effect <- apply(exp(combined), 2, median)
-```
-
-We can make a basic plot using the following code. A more elaborate version is
-shown in Figure \ref{fig:owl-plot-fancy}.
-
-```{r owl-plot-basic, echo=TRUE, eval=FALSE}
-ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
-```
-
-Overall, we find a weak average positive effect of annual mean NAO on overall
-counts, but a southeast to northwest gradient in the intensity of the effect
-(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
-to the Atlantic coast and those migrating the furthest south being the most
-affected by NAO.
-
-<!-- TODO keep this? -->
-<!-- We provide an extended version of this example in Appendix \ref{app:owls}. -->
-
-```{r shapes, echo=FALSE}
-if (!file.exists(here::here("data/ne_10m_lakes"))) {
-  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
-    "download/10m/physical/ne_10m_lakes.zip")
-  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
-  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
-}
-```
-
-```{r shapes-read, echo=FALSE}
-Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
-coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
-  sf::st_transform(crs = Albers)
-lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
-```
-
-```{r owl-proj2}
-# project to get nice axis limits 
-snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
-snow2 <- snow2 |> mutate(x = X, y = Y) |> 
-  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
-```
-
-```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
-Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
-The effect is multiplicative on owl count per NAO unit.
-In the west, the lower values are near 1 implying no effect, whereas in the southeast the effect becomes positive."}
-nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
-  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
-ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect)) +
-  geom_sf(data = coast, colour = "gray50") +
-  geom_sf(data = lakes, colour = "gray50", fill = NA) +
-  coord_sf(
-    xlim = c(min(snow2$X), max(snow2$X)),
-    ylim = c(min(snow2$Y), max(snow2$Y))
-  ) +
-   scale_colour_viridis_c(
-    guide = guide_colourbar(direction = "horizontal", title.position = "left")
-  ) +
-  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
-    theme(legend.position = "bottom", axis.title = element_blank())
-```
-
-\clearpage
-
-```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
-# tidy(fit, conf.int = TRUE)
-```
-
-```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
-# tidy(fit, effects = "ran_pars", conf.int = TRUE)
-```
-
-<!-- The latter `tidy()` call returns all variance components of the model and omits the standard errors since these are calculated in log space. -->
-
-<!-- We can run some basic checks on our model with the `sanity()` function: -->
-
-<!-- We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
--->
-<!-- Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially. -->
-<!-- The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict. -->
-
-<!-- \begin{figure} -->
-<!-- \centering -->
-<!-- \includegraphics[width=5.2in]{../../figs/pcod-fig} -->
-<!-- \caption[]{Output from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada. -->
-<!-- (A) SPDE mesh (lines) combined with the trawl survey observations (points).
-Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-The circle area corresponds to biomass density of Pacific Cod caught on individual trawls.
--->
-<!-- (B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
-(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
-Crosses illustrate center of gravity (\texttt{get\_cog()}) from a spatiotemporal version; lighter crosses are more recent.
--->
-<!-- (D) Conditional effect of depth modelled as a penalized smoother. -->
-<!-- These predictions are made omitting the spatial random field. -->
-<!-- (E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}).
-The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig} -->
-<!-- \end{figure} -->
-
-<!-- We could extend this to a spatiotemporal model simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
--->
-<!-- Here, we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
--->
-
-\clearpage
-
 # Model validation and selection
@@
 # Package comparisons
+
+TODO: BEFORE OR AFTER EXAMPLES
 
@@
 
+# Installation
+
+\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive
+Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>
+
+```{r sdmTMB-install, eval=FALSE, echo=TRUE}
+install.packages("sdmTMB", dependencies = TRUE)
+```
+
+Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package
+[@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed
+for the `make_mesh()` function below. Soon, we intend to switch to using the
+\pkg{fmesher} package to eliminate this dependency.
+
+Alternatively, the development version can be installed with:
+
+```{r remotes, eval=FALSE, echo=TRUE}
+# install.packages("remotes")
+remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
+```
+
+Additional utilities are maintained in the \pkg{sdmTMBextra} package at
+<https://github.com/pbs-assess/sdmTMBextra>.
+
+# Species distribution modelling with spatial random fields
+
+We begin with a simple species distribution model of encounter probability of
+Pacific Cod (*Gadus macrocephalus*) from a trawl survey conducted in Queen
+Charlotte Sound, British Columbia, Canada. The purpose of our example is to
+illustrate the need for spatial random fields. This survey is conducted by
+Fisheries and Oceans Canada (REF) and follows a depth-stratified random
+sampling design, resulting in a georeferenced dataset. The data frame `pcod` is
+available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
+longitude, Universal Transverse Mercator (UTM) coordinates, bottom
+depth, and encounter (`present = 1`) vs. non-encounter (`present = 0`).
+
+```{r libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
+library(sdmTMB)
+library(dplyr)
+library(ggplot2)
+```
+
+```{r libs-extras}
+theme_set(theme_light())
+options(ggplot2.continuous.fill = "viridis")
+```
+
+```{r pcod-head, echo=TRUE}
+select(pcod, lat, lon, X, Y, depth, present) |> 
+  head(n = 3)
+```
+
+An \pkg{sdmTMB} model requires a data frame that contains a response column,
+columns for any predictors, and columns for spatial coordinates. Usually it
+makes sense to convert the spatial coordinates to an equidistant projection
+such as UTMs to ensure that distance remains constant throughout the study
+region [e.g., using `sf::st_transform()`, @pebesma2018]. Here we use the helper
+function `add_utm_columns()` to add UTM coordinates with km units (so our
+estimated spatial range parameter is not too big or small). By default, the
+function will guess the UTM zone and create new columns `X` and `Y`. Since our
+example data already has these UTM columns, we can skip running this code.
+
+```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
+pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+```
+
+\sloppy
+We then create a mesh object that contains triangulation and projection
+matrices needed to apply the SPDE approach. Here, `cutoff` defines the minimum
+allowed distance between mesh vertices in the units of `X` and `Y`.
+`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` or
+`fmesher::fm_rcdt_2d_inla()` and `cutoff` is passed through to these functions.
+Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or
+\pkg{fmesher} and supplied it to the `mesh` argument in `make_mesh()`. We can
+inspect our mesh object with the associated plotting method.
+
+```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
+plot(mesh_pcod)
+```
+
+We can then fit our model with `sdmTMB()`. We will fit a logistic regression of
+encounter probability with and without spatial random fields (by toggling the
+`spatial` argument) to illustrate the importance of accounting for spatial
+correlation. In addition to the spatial random field, we include an intercept
+and a quadratic effect of depth on the probability of encounter. The `update()`
+S3 method refits the same model updating any specified arguments.
+
+Our random field model can be written as
+
+$$
+\begin{aligned}
+y_{\boldsymbol{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\boldsymbol{s}}\right),\\
+\mu_{\boldsymbol{s}} &= \operatorname{logit}^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}} \boldsymbol{\beta} +
+\omega_{\boldsymbol{s}}
+\right),
+\end{aligned}
+$$
+
+where TODO.
+
+$$
+\omega_{\boldsymbol{s}} \sim \mathrm{MVN}(\boldsymbol{0},\boldsymbol{\Sigma}_\omega)
+$$
+
+```{r, echo=TRUE}
+fit_bin_rf <- sdmTMB(
+  present ~ poly(log(depth), 2),
+  data = pcod, mesh = mesh_pcod, 
+  spatial = "on",
+  family = binomial(link = "logit")
+)
+fit_bin <- update(fit_bin_rf, spatial = "off")
+```
+
+We can run some basic checks on our model with the `sanity()` function:
+
+```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
+sanity(fit)
+```
+
+```
+#> ✓ Non-linear minimizer suggests successful convergence
+#> ✓ Hessian matrix is positive definite
+#> ✓ No extreme or very small eigenvalues detected
+#> ✓ No gradients with respect to fixed effects are >= 0.001
+#> ✓ No fixed-effect standard errors are NA
+#> ✓ No fixed-effect standard errors look unreasonably large
+#> ✓ No sigma parameters are < 0.01
+#> ✓ No sigma parameters are > 100
+#> ✓ Range parameter doesn't look unreasonably large
+```
+
+Here, this does not flag any issues. `sanity()` is checking that the `nlminb()`
+optimizer reported successful convergence, that the Hessian matrix is positive
+definite, that no extreme (XX) or small (XX) eigenvalues are detected, that no
+absolute log likelihood gradients with respect two fixed defects are >= 0.001,
+that all fixed effects have reported standard errors and that they do not look
+unreasonably large (XX by default), that random field marginal standard
+deviations are not unexpectedly small or large, and that the random field
+Matérn range parameter does not look unreasonably large.
+
+We can get a summary of our model fit:
+
+```{r pcod-bin-summary, echo=TRUE}
+summary(fit_bin_rf)
+```
+
+The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
+We also see the formula, mesh, fitted data, and family. Next we see any
+estimated main effects, the Matérn range distance, the spatial random field
+standard deviation, and the negative log likelihood at convergence.
+
+<!-- The -->
+<!-- smoother summary follows the format used in the \pkg{brms} package -->
+<!-- [@burkner2017]. As noted in the print method, we can extract a tidy data frame -->
+<!-- similar to the format in the \pkg{broom} [@broom] and \pkg{broom.mixed} -->
+<!-- [@broommixed] packages: -->
+
+```{r}
+# set.seed(1)
+# pcod$resid_bin <- residuals(fit_bin)
+# set.seed(1)
+# pcod$resid_bin_rf <- residuals(fit_bin_rf)
+# g1 <- ggplot(pcod, aes(X, Y, colour = resid_bin)) + geom_point() +
+#   scale_color_gradient2() +
+#   theme(legend.position = "bottom")
+# g2 <- ggplot(pcod, aes(X, Y, colour = resid_bin_rf)) + geom_point() +
+#   scale_color_gradient2() +
+#   theme(legend.position = "bottom")
+# cowplot::plot_grid(g1, g2, ncol = 2)
+```
+
+```{r, echo=FALSE, results='hide'}
+test_autocor <- function(obj) {
+  set.seed(1)
+  s <- simulate(obj, nsim = 500)
+  pr <- predict(obj, type = "response")$est
+  r <- DHARMa::createDHARMa(
+    simulatedResponse = s,
+    observedResponse = pcod$present,
+    fittedPredictedResponse = pr
+  )
+  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
+}
+(t_no_rf <- test_autocor(fit_bin))
+(t_rf <- test_autocor(fit_bin_rf))
+p_rf <- round(t_rf$p.value, 2)
+```
+
+We can test for spatial autocorrelation with a Moran's I test or with a visual
+inspection of the residuals calculated from the `residuals()` S3 method. We
+omit the Moran's I code for brevity but include it in the code supplement. The
+p-value for spatial autocorrelation is low for the model without a random field
+(p < 0.01) but large for the model with a random field (p = `r p_rf`)
+suggesting significant residual correlation in the simpler model that is
+alleviated by including the random field.
+
+We can also see that the AIC of the model with spatial random fields is
+considerably lower:
+
+```{r pcod-aic, eval=TRUE, echo=TRUE, results='markup'}
+AIC(fit_bin_rf, fit_bin)
+```
+
+We could do additional testing with cross validation via `sdmTMB_cv()`. Here we
+will do 10-fold cross validation with the folds constructed randomly. Using the
+`fold_ids` argument, we could supply our own folds and conduct spatially
+blocked cross validation.
+
+```{r pcod-cv, eval=TRUE, echo=TRUE}
+cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
+  data = pcod, mesh = mesh_pcod, spatial = "on",
+  family = binomial(), k_folds = 10
+)
+cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
+  data = pcod, mesh = mesh_pcod, spatial = "off",
+  family = binomial(), k_folds = 10
+)
+```
+
+Indeed, the log likelihood of left-out data is considerably larger for the
+model that includes random fields:
+
+```{r pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'}
+cv_bin_rf$sum_loglik
+cv_bin$sum_loglik
+```
+
+The standard errors on our fixed effects have increased with the spatial random
+field:
+
+```{r pcod-tidy, eval=TRUE, echo=TRUE, results='markup'}
+tidy(fit_bin_rf, conf.int = TRUE)
+tidy(fit_bin, conf.int = TRUE)
+```
+
+We can also use `tidy()` to get a data frame of our random field parameters,
+where `sigma_O` is $\sigma_O$ in the above equations: the standard deviation of
+the random field.
+
+```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
+tidy(fit, effects = "ran_pars", conf.int = TRUE)
+```
+
+\sloppy
+To visualize our model, we can make predictions with the `predict()` method
+(`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on
+a new data frame with any locations and values for the predictor columns. Here,
+we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire
+region of interest so we can visualize the predictions spatially. The grid
+contains spatial covariate columns and all predictors used in the model set at
+values for which we want to predict. The output of `predict()` is a data frame
+containing overall estimates in link space (`est`), estimates from the
+non-random-field components (`est_non_rf`; here, intercept and depth), and
+estimates from the individual random field components (here, `omega_s`---the
+spatial field).
+
+```{r pcod-predict, echo=TRUE, results='markup'}
+p <- predict(fit_bin_rf, newdata = qcs_grid)
+select(p, X, Y, depth, est, est_non_rf, omega_s) |>
+  as_tibble() |> head(n = 2)
+```
+
+TODO MOVE TO CAPTION? We have an effect of depth and a spatial random field,
+which combined in link space result in our overall prediction. Note the
+difference between only the fixed effects (left) and including the spatial
+random field (right). TODO DESCRIBE PLOTTING AND/OR INCLUDE
+
+```{r pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="TODO"}
+plot_spatial_map <- function(dat, column, title) {
+  ggplot(dat, aes(X, Y, fill = {{ column }})) +
+    geom_raster() +
+    coord_fixed() +
+    theme(legend.position= "bottom") +
+    ggtitle(title) +
+    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
+}
+g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
+g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
+g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
+cowplot::plot_grid(g1, g2, g3, ncol = 3)
+```
+
+# Spatiotemporal species distribution modeling with dogfish
+
+<!-- Need brief description of dogfish data and model here?? -->
+
+As a second example, we will construct a spatiotemporal model of catch rates of
+Pacific Spiny Dogfish (*Squalus suckleyi*) from a trawl survey off the west
+coast of Vancouver Island, Canada. This example extends the spatial model
+described above by also including (1) spatiotemporal fields allowing the
+distribution in each year to be unique, (2) a time-varying intercept as an
+AR(1) process to allow each year to have a different mean catch rate, (3)
+a smooth effect of depth, allowing catch rates to vary non-linearly with
+depth, and (4) spatial anisotropy allowing spatial correlation to be directionally
+dependent. Since catch rates are positive, continuous, and contain zeros, we begin
+by specifying the response family as a Tweedie distribution [@tweedie1984] with
+a log link. We then compare alternative families, spatiotemporal random field
+structures, and the inclusion of anisotropy to illustrate the flexibility of
+\pkg{sdmTMB}.
+
+<!-- We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`. -->
+<!-- We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package. -->
+<!-- We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros. -->
+<!-- An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument). -->
+
+The dataset includes spatial coordinates, year, dogfish catch weight in kg,
+area swept in km, and bottom depth:
+
+```{r dog-head-dat, echo=TRUE}
+dat <- select(dogfish, longitude, latitude, year, 
+  catch_weight, area_swept, depth)
+head(dat, n = 3)
+```
+
+We add UTM zone 9 columns, create a log depth column for convenience, and
+create a basic mesh:
+
+```{r dog-utms, echo=TRUE}
+dat <- add_utm_columns(dat, c("longitude", "latitude"), 
+  units = "km", utm_crs = 32609)
+dat$log_depth <- log(dat$depth)
+mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
+```
+
+```{r dog-mesh2, eval=FALSE}
+plot(mesh)
+```
+
+We can then specify our model. We include an offset for the effort variable log
+area swept such that we are effectively modelling density and our predictions
+will be for an area swept of 1 km^2^.
+
+Our model can be written as TODO
+
+$$
+\begin{aligned}
+\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
+\mu_{\boldsymbol{s},t} &=
+f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
+O_{\boldsymbol{s},t} +
+\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \gamma_t +
+\omega_{\boldsymbol{s}} +
+\epsilon_{\boldsymbol{s},t} 
+\right),
+\end{aligned}
+$$
+
+where TODO.
+
+and TODO
+
+$$
+\begin{aligned}
+  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
+  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
+\end{aligned}
+$$
+
+```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+fit_tw <- sdmTMB(
+  catch_weight ~ s(log_depth),
+  data = dat,
+  mesh = mesh,
+  offset = log(dat$area_swept),
+  time = "year",
+  time_varying = ~ 1,
+  time_varying_type = "ar1",
+  spatial = "on",
+  spatiotemporal = "iid",
+  anisotropy = TRUE,
+  silent = FALSE,
+  family = tweedie()
+)
+```
+
+We can explore four alternative families that may better represent the data.
+Each alternative family uses a delta or hurdle distribution: delta-gamma and
+delta-lognormal (REF), and delta-Poisson-link-gamma and
+delta-Poisson-link-lognormal (REF). Note that we choose to omit the
+spatiotemporal random fields from the first model component 
+(`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
+collapsing to zero indicating minimal changes in encounter spatial distribution
+from year to year.
+
+TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.
+
+```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+fit_dg <- update(fit_tw, family = delta_gamma(), 
+  spatiotemporal = list("off", "iid"))
+fit_dl <- update(fit_dg, family = delta_lognormal())
+fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
+fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
+```
+
+We can then compare the models via AIC:
+
+```{r dog-aic, echo=TRUE}
+AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
+  mutate(delta_AIC = AIC - min(AIC)) |> 
+  arrange(delta_AIC)
+```
+
+We find that the Poisson-link delta-lognormal model is favoured with AIC. We
+next test two additional model formulations: making the spatial correlation
+isotropic (the default) instead of anisotropic, and structuring the positive
+model component spatiotemporal random fields as AR(1) to allow spatiotemporal
+hotspots to persist from year to year.
+
+The AR(1) fields can be represented as TODO.
+
+$$
+\begin{aligned}
+\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
+\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
+\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\end{aligned}
+$$
+
+where TODO.
+
+```{r dog-ar1, results='hide', echo=TRUE}
+fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
+fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
+```
+
+```{r dog-aci2, echo=TRUE}
+AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
+```
+
+<!-- Need description of anisotropy and why it's important here or somewhere
+else -->
+
+We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
+is important here given the elongated shape of the continental shelf with
+a rapid transition to deeper water. On inspecting the model output with
+`summary()`, we notice that the spatial random field SD in the positive model
+component has become very small once AR(1) spatiotemporal fields were added. We
+therefore consider a model where these are turned off in the second component
+of the delta model:
+
+```{r dog-ar1-2, echo=TRUE, results="hide"}
+fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
+```
+
+```{r dog-aic3, echo=TRUE}
+AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
+```
+
+We save our chosen model to the object `fit` to simplify subsequent code, run
+the `sanity()` check (suppressed for space), and inspect `summary()`:
+
+```{r dog-print, echo=TRUE}
+fit <- fit_dpl_ar1_only
+sanity(fit)
+summary(fit)
+```
+
+The output is considerably more complex compared to our binomial spatial model.
+We now have two model components, which are shown one after the other. Starting
+with the binomial component, we have output from the smoother, which includes
+a linear component (`slog_depth`) and the standard deviation on the smoother
+weights (`sds(log_depth)`). The smoother summary follows the format used in the
+\pkg{brms} package [@burkner2017]. Next, we have the time-varying intercepts
+and information on our anisotropic spatial correlation. We then have the second
+model component (lognormal) with a similar summary structure but with the
+addition of a dispersion parameter for the lognormal, the AR(1) correlation of
+the spatiotemporal random fields, and a spatiotemporal random field marginal
+standard deviation. We can use `plot_anisotropy()` to visually inspect the
+anisotropy (Figure \ref{fig:dog-aniso}).
+
+```{r dog-aniso, echo=TRUE, fig.cap="TODO", out.width="4in"}
+plot_anisotropy(fit_dpl)
+```
+
+Similarly to the first example, we can visualize model predictions on a grid
+covering the area of interest. Because this is a spatiotemporal model, we first
+need to replicate our grid for each year we will predict on. Since this is
+a common operation, we include the convenience function `replicate_df()` to
+replicate a data frame. We then ensure our data frame contains all the
+predictors used in the model (here `log_depth`).
+
+<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
+```{r dog-grid, echo=TRUE}
+grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
+grid$log_depth <- log(grid$depth)
+head(grid, n = 2)
+```
+
+```{r dog-pred1, echo=TRUE}
+pred <- predict(fit, newdata = grid, type = "response")
+```
+
+```{r dog-pred2, echo=TRUE}
+names(pred)
+```
+
+Our prediction data frame is similar to the binomial spatial model, but
+includes columns for the two delta model components (labelled with suffixes `1`
+and `2`) and adds an `epsilon_st` column for spatiotemporal random effects. We can
+easily generate plots from this data frame with ggplot2 or similar. We suppress
+that code in this paper for brevity.
+
+```{r plot-map, echo=FALSE}
+plot_map <- function(dat, column) {
+  ggplot(dat, aes(X, Y, fill = {{ column }})) +
+    geom_raster() +
+    facet_wrap(vars(year)) +
+    coord_fixed() +
+    theme(legend.position= "bottom") +
+    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
+}
+```
+
+```{r dog-plot1, echo=FALSE}
+g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(est) +
+  # labs(fill = "Density") +
+  scale_fill_viridis_c(trans = "log10")
+```
+
+```{r dog-plot2, echo=FALSE}
+g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(omega_s1) +
+  scale_fill_gradient2()
+```
+
+```{r dog-plot3, echo=FALSE}
+g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(epsilon_st2) +
+  scale_fill_gradient2()
+```
+
+```{r dog-plot4, echo=FALSE}
+g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(est_non_rf1) +
+  scale_fill_viridis_c(trans = "log10")
+```
+
+```{r dog-wvci-pred, fig.asp = 0.7, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
+cowplot::plot_grid(
+  g_nonrf,
+  g_omega,
+  g_eps,
+  g_est,
+  ncol = 2L
+)
+```
+
+We can visualize the conditional effect of the bottom depth smoother by
+predicting across a sequence of depths and holding other variables at reference
+values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
+to include both delta model components (`model = NA`), omit the random fields
+(`re_form = NA`), and return standard errors (`se_fit = TRUE`).
+
+```{r dog-depth, echo=TRUE, out.width="5in"}
+nd <- data.frame(
+  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
+  year = max(dat$year)
+)
+pred_depth <- predict(
+  fit, newdata = nd, 
+  model = NA, re_form = NA, se_fit = TRUE
+)
+```
+
+```{r dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
+ggplot(pred_depth, aes(
+  exp(log_depth), exp(est), 
+  ymin = exp(est - 2 * est_se), 
+  ymax = exp(est + 2 * est_se))) +
+  geom_ribbon(fill = "grey90") +
+  geom_line() +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
+  labs(x = "Depth (m)", y = "Density")
+```
+
+We can generate an area-weighted population index (e.g., a relative or absolute
+index of abundance or biomass) that is independent of sampling locations by
+predicting from the model on a grid covering the area of interest and summing
+the predicted biomass with the `get_index()` function (Figure
+\ref{fig:dog-index-plot}). We supply the grid cell area (4 km^2^) to the `area`
+argument and specify `bias_correct = TRUE` to enable a generic epsilon-bias
+correction needed due to the non-linear transformation of the random effects
+(REF).
+
+```{r dog-index, echo=TRUE}
+pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
+ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
+```
+
+```{r dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
+ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
+  geom_pointrange() +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
+  labs(y = "Biomass", x = "Year")
+```
+
+# Spatially varying coefficient example with citizen science data
+
+In our final example, we demonstrate a model with spatially varying coefficient
+(SVC) effects and illustrate combining uncertainty from parameters by working
+with draws from the joint parameter precision matrix. Spatially varying
+coefficients ... TODO REFS. In addition, we illustrate using restricted maximum
+likelihood (REML), which integrates over the fixed effects as well as the
+random effects.
+
+Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive
+migrants, meaning that they appear across the mid-latitudes of North America in
+much greater numbers in some winters than others. The reasons for this
+interannual variation in the number of individuals migrating south are not well
+understood but seem to be related to high abundances of food during the
+breeding season and therefore sharp increases in breeding ground population
+densities [@robillard2016]. The North Atlantic Oscillation Index (NAO) has been
+linked to productivity of both owls and their prey in Europe [@millon2014].
+Because both productivity and the choice of wintering location could be
+influenced by climate, we modelled a spatially varying effect of annual mean
+NAO index on winter abundance across the southern boundary of their winter
+distribution. We fit counts observed in North America during annual Christmas
+Bird Counts [@cbc] using a negative binomial [NB2, @hilbe2011] distribution,
+random intercepts for year, spatial and spatiotemporal random fields, and
+a spatially varying coefficient associated with the NAO.
+
+We can write the model as TODO
+
+$$
+\begin{aligned}
+\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
+\mu_{\boldsymbol{s},t} &=
+f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
+\alpha_g +
+\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} +
+\omega_{\boldsymbol{s}} +
+\epsilon_{\boldsymbol{s},t} \right),
+\end{aligned}
+$$
+
+where TODO.
+
+```{r owl-data}
+snow <- readRDS(here::here("data/SNOW_data.rds"))
+```
+
+```{r owl-fit, echo=TRUE, results = "hide"}
+mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
+fit_owl <- sdmTMB(
+  count ~ 1 + nao + (1 | year_f),
+  spatial_varying = ~ nao,
+  time = "year",
+  data = snow, 
+  family = nbinom2(link = "log"),
+  spatial = "on",
+  spatiotemporal = "iid",
+  mesh = mesh_snow,
+  reml = TRUE,
+  silent = FALSE
+)
+```
+
+```{r owl-sanity, eval=FALSE}
+sanity(fit_owl)
+```
+
+```{r owl-print, echo=TRUE}
+summary(fit_owl)
+```
+
+The spatially varying coefficient effect at any point in space is the
+combination of the main effect and spatial random effect for `nao`. We
+illustrate deriving uncertainty on their combination by simulating from the
+parameter precision matrix assuming the joint parameter covariance matrix in
+multivariate normal. We do this by specifying a non-null number of simulation
+draws to `nsim` in `predict.sdmTMB()`. By default, `nsim > 0` will return
+a matrix of draws from the overall prediction. Here, we instead specify that we
+want to return draws from the `zeta_s` random field, which is the spatially
+varying random field. This returns a matrix where each row matches a row of
+`newdata` and each column is a simulation draw. We then use the function
+`spread_sims()` to draw 200 simulations from the parameters themselves, combine
+the main effect draws (`sims$nao`) with the random field values `zeta_s`, and
+calculate the median for each row of data so we can plot them.
+
+```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
+zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
+dim(zeta_s)
+sims <- spread_sims(fit_owl, nsim = 200)
+combined <- sims$nao + t(zeta_s)
+snow$nao_effect <- apply(exp(combined), 2, median)
+```
+
+We can make a basic plot using the following code. A more elaborate version is
+shown in Figure \ref{fig:owl-plot-fancy}.
+
+```{r owl-plot-basic, echo=TRUE, eval=FALSE}
+ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
+```
+
+Overall, we find a weak average positive effect of annual mean NAO on overall
+counts, but a southeast to northwest gradient in the intensity of the effect
+(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
+to the Atlantic coast and those migrating the furthest south being the most
+affected by NAO.
+
+<!-- TODO keep this? -->
+<!-- We provide an extended version of this example in Appendix \ref{app:owls}. -->
+
+```{r shapes, echo=FALSE}
+if (!file.exists(here::here("data/ne_10m_lakes"))) {
+  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
+    "download/10m/physical/ne_10m_lakes.zip")
+  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
+  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+}
+```
+
+```{r shapes-read, echo=FALSE}
+Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
+coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
+  sf::st_transform(crs = Albers)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
+lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
+```
+
+```{r owl-proj2}
+# project to get nice axis limits 
+snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
+snow2 <- snow2 |> mutate(x = X, y = Y) |> 
+  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
+```
+
+```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
+Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
+The effect is multiplicative on owl count per NAO unit.
+In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5}
+nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
+  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
+ggplot(data = nao_effect_df) +
+  geom_point(aes(X, Y, colour = nao_effect)) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  coord_sf(
+    xlim = c(min(snow2$X), max(snow2$X)),
+    ylim = c(min(snow2$Y), max(snow2$Y))
+  ) +
+   scale_colour_viridis_c(
+    guide = guide_colourbar(direction = "horizontal", title.position = "left")
+  ) +
+  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
+    theme(legend.position = "bottom", axis.title = element_blank())
+```
+
+\clearpage
+
+<!-- The latter `tidy()` call returns all variance components of the model and omits the standard errors since these are calculated in log space. -->
+
+<!-- We can run some basic checks on our model with the `sanity()` function: -->
+
+<!-- We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
+-->
+<!-- Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially. -->
+<!-- The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict. -->
+
+<!-- \begin{figure} -->
+<!-- \centering -->
+<!-- \includegraphics[width=5.2in]{../../figs/pcod-fig} -->
+<!-- \caption[]{Output from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada. -->
+<!-- (A) SPDE mesh (lines) combined with the trawl survey observations (points).
+Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
+The circle area corresponds to biomass density of Pacific Cod caught on individual trawls.
+-->
+<!-- (B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
+(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
+Crosses illustrate center of gravity (\texttt{get\_cog()}) from a spatiotemporal version; lighter crosses are more recent.
+-->
+<!-- (D) Conditional effect of depth modelled as a penalized smoother. -->
+<!-- These predictions are made omitting the spatial random field. -->
+<!-- (E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}).
+The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig} -->
+<!-- \end{figure} -->
+
+<!-- We could extend this to a spatiotemporal model simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
+-->
+<!-- Here, we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.
+-->
+
+\clearpage
+
 # Discussion

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
- add main equation - remove ELPD description (could add back in a Draper ref for log-scoring) - add description for 1st example

@@
 
-TODO ADD EQUATIONS
-
 \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
@@
 
-The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Figure \ref{fig:diagram}).
-This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
+The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects $\boldsymbol{\beta}$ and offset $O_{\boldsymbol{s},t}$, IID (independent and identically distributed) random intercepts $\alpha_g$, spatial intercept random fields $\omega_{\boldsymbol{s}}$, spatiotemporal intercept random fields $\epsilon_{\boldsymbol{s},t}$, time-varying effects $\boldsymbol{\gamma_t}$, and spatially varying effects $\zeta_{\boldsymbol{s}}$ (Figure \ref{fig:diagram}).
+These components are discussed in more detail via our examples below.
+
+$$
+\begin{aligned}
+\mu_{\boldsymbol{s},t} &=
+f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
+O_{\boldsymbol{s},t} +
+\alpha_g +
+\omega_{\boldsymbol{s}} +
+\epsilon_{\boldsymbol{s},t} + 
+\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \boldsymbol{\gamma_t} +
+\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} \right),
+\end{aligned}
+$$
+This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link function $f^{-1}$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
 Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
@@
 (1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
-(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
+(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
 (3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
@@
 
-where TODO.
+where $\boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}}$ contains main effect covariates (intercept, quadratic effects of depth), $\boldsymbol{\beta}$ represents estimated main effect coefficients, and $\omega_{\boldsymbol{s}}$ represents the estimated spatial field,
 
@@
 distribution in each year to be unique, (2) a time-varying intercept as an
-AR(1) process to allow each year to have a different mean catch rate, (3)
+AR(1) process to allow year effects to vary but remain autocorrelated, (3)
 a smooth effect of depth, allowing catch rates to vary non-linearly with

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
More description

@@
 
-where TODO.
-
-and TODO
+where $\boldsymbol{\beta}$ are coefficients associated with the main effects, $O_{\boldsymbol{s},t}$ represents the offset (here, log area swept), $\gamma_t$ represents the time-varying coefficients, $\omega_{\boldsymbol{s}}$ is the spatial field (equivalent to a spatial intercept) and $\epsilon_{\boldsymbol{s},t} $ represents spatiotemporal fields.
+
+The temporally varying intercepts $\gamma_t$ are modeled as a stationary AR(1) process,
 

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
\boldsymbol{} -> \bm{}

--- 2023-09-12 doc/paper-jss/paper-jss.Rmd
@@
 
-The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects $\boldsymbol{\beta}$ and offset $O_{\boldsymbol{s},t}$, IID (independent and identically distributed) random intercepts $\alpha_g$, spatial intercept random fields $\omega_{\boldsymbol{s}}$, spatiotemporal intercept random fields $\epsilon_{\boldsymbol{s},t}$, time-varying effects $\boldsymbol{\gamma_t}$, and spatially varying effects $\zeta_{\boldsymbol{s}}$ (Figure \ref{fig:diagram}).
+The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects $\bm{\beta}$ and offset $O_{\bm{s},t}$, IID (independent and identically distributed) random intercepts $\alpha_g$, spatial intercept random fields $\omega_{\bm{s}}$, spatiotemporal intercept random fields $\epsilon_{\bm{s},t}$, time-varying effects $\bm{\gamma_t}$, and spatially varying effects $\zeta_{\bm{s}}$ (Figure \ref{fig:diagram}).
 These components are discussed in more detail via our examples below.
@@
 \begin{aligned}
-\mu_{\boldsymbol{s},t} &=
-f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
-O_{\boldsymbol{s},t} +
+\mu_{\bm{s},t} &=
+f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
+O_{\bm{s},t} +
 \alpha_g +
-\omega_{\boldsymbol{s}} +
-\epsilon_{\boldsymbol{s},t} + 
-\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \boldsymbol{\gamma_t} +
-\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} \right),
+\omega_{\bm{s}} +
+\epsilon_{\bm{s},t} + 
+\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
+\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} \right),
 \end{aligned}
@@
 \begin{aligned}
-y_{\boldsymbol{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\boldsymbol{s}}\right),\\
-\mu_{\boldsymbol{s}} &= \operatorname{logit}^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}} \boldsymbol{\beta} +
-\omega_{\boldsymbol{s}}
+y_{\bm{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\bm{s}}\right),\\
+\mu_{\bm{s}} &= \operatorname{logit}^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s}} \bm{\beta} +
+\omega_{\bm{s}}
 \right),
@@
 
-where $\boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}}$ contains main effect covariates (intercept, quadratic effects of depth), $\boldsymbol{\beta}$ represents estimated main effect coefficients, and $\omega_{\boldsymbol{s}}$ represents the estimated spatial field,
-
-$$
-\omega_{\boldsymbol{s}} \sim \mathrm{MVN}(\boldsymbol{0},\boldsymbol{\Sigma}_\omega)
+where $\bm{X}^{\mathrm{main}}_{\bm{s}}$ contains main effect covariates (intercept, quadratic effects of depth), $\bm{\beta}$ represents estimated main effect coefficients, and $\omega_{\bm{s}}$ represents the estimated spatial field,
+
+$$
+\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)
 $$
@@
 \begin{aligned}
-\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
-\mu_{\boldsymbol{s},t} &=
-f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
-O_{\boldsymbol{s},t} +
-\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \gamma_t +
-\omega_{\boldsymbol{s}} +
-\epsilon_{\boldsymbol{s},t} 
+\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
+\mu_{\bm{s},t} &=
+f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
+O_{\bm{s},t} +
+\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
+\omega_{\bm{s}} +
+\epsilon_{\bm{s},t} 
 \right),
@@
 
-where TODO.
-
-and TODO
+where $\bm{\beta}$ are coefficients associated with the main effects, $O_{\bm{s},t}$ represents the offset (here, log area swept), $\gamma_t$ represents the time-varying coefficients, $\omega_{\bm{s}}$ is the spatial field (equivalent to a spatial intercept) and $\epsilon_{\bm{s},t}$ represents spatiotemporal fields.
+
+The temporally varying intercepts $\gamma_t$ are modeled as a stationary AR(1) process,
 
@@
 \begin{aligned}
-\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
-\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
-\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
+\bm{\delta}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
+\bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
+\bm{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 \begin{aligned}
-\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
-\mu_{\boldsymbol{s},t} &=
-f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
+\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
+\mu_{\bm{s},t} &=
+f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
 \alpha_g +
-\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} +
-\omega_{\boldsymbol{s}} +
-\epsilon_{\boldsymbol{s},t} \right),
+\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
+\omega_{\bm{s}} +
+\epsilon_{\bm{s},t} \right),
 \end{aligned}

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
Copy editing

@@
 \mu_{\bm{s},t} &=
-f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
+g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
 O_{\bm{s},t} +
@@
 $$
-This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link function $f^{-1}$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
+This process component is combined with an observation error family (e.g., Gaussian, gamma, binomial, Tweedie) and link function $g$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
 Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
-See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
+See the model description vignette for an expanded description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
 
@@
 The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) [@haskard2007].
-The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).
+The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared among random fields or estimated separately (`share_range` argument).
 
@@
 Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
-We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
-\pkg{sdmTMB} can use many `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].
+We include additional flexibility in specifying the linear main effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
+\pkg{sdmTMB} can use many `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers `s(x, y)`, smoothers varying by continuous or categorical variables `s(x1, by = x2)`, cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].
 
@@
 Data may be simulated with or without a fitted model (`simulate.sdmTMB()`, `sdmTMB_simulate()` respectively).
-Estimation is done via maximum marginal likelihood with the `sdmTMB()` function (or `sdmTMB_cv()` for cross-validation), and the resulting fitted model can be used to make predictions (`predict.sdmTMB()`) or calculate residuals (`residuals.sdmTMB()`), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+Estimation is done via maximum marginal likelihood with the `sdmTMB()` function, and the resulting fitted model can be used to make predictions (`predict.sdmTMB()`) or calculate residuals (`residuals.sdmTMB()`), derive additional quantities of interest, and simulate Bayesian posterior distributions.
 Common quantities of interest may include summarizing shifts in the center of gravity (`get_cog()`) [e.g., @thorson2016cog] or in the field of ecology, tracking total population densities through time (`get_index()`).
+<!-- TODO: describe this in more detail and how it could apply beyond ecology; maybe a derived variables subsection? -->
 
@@
 
-```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="TODO"}
 mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
@@
 inspection of the residuals calculated from the `residuals()` S3 method. We
-omit the Moran's I code for brevity but include it in the code supplement. The
-p-value for spatial autocorrelation is low for the model without a random field
-(p < 0.01) but large for the model with a random field (p = `r p_rf`)
-suggesting significant residual correlation in the simpler model that is
-alleviated by including the random field.
+omit code for Moran's I calculation for brevity but include an example in the
+code supplement. The p-value for spatial autocorrelation is low for the model
+without a random field (p < 0.01) but large for the model with a random field
+(p = `r p_rf`) suggesting significant residual correlation in the simpler model
+that is alleviated by including the random field.
 
@@
 where `sigma_O` is $\sigma_O$ in the above equations: the standard deviation of
-the random field.
+the spatial random field $\bm{\omega}$.
 
@@
 \mu_{\bm{s},t} &=
-f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
+\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
 O_{\bm{s},t} +
@@
 delta-Poisson-link-lognormal (REF). Note that we choose to omit the
-spatiotemporal random fields from the first model component 
+spatiotemporal random fields from the binomial model component 
 (`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
@@
 \begin{aligned}
-\bm{\delta}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
+\bm{\delta}_{t=1} &\sim \operatorname{MVN} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
 \bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
-\bm{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
+\bm{\epsilon_{t}} \sim \operatorname{MVN} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 
-Our prediction data frame is similar to the binomial spatial model, but
-includes columns for the two delta model components (labelled with suffixes `1`
-and `2`) and adds an `epsilon_st` column for spatiotemporal random effects. We can
-easily generate plots from this data frame with ggplot2 or similar. We suppress
-that code in this paper for brevity.
+Our prediction data frame is similar to the binomial spatial model, but includes
+columns for the two delta model components (labelled with suffixes `1` and `2`)
+and adds an `epsilon_st` column for spatiotemporal random effects. We can easily
+generate plots from this data frame (e.g., with `ggplot2::geom_raster()`). We
+suppress that code in this paper for brevity.
 
@@
 (REF).
+<!-- TODO: and the Laplace? -->
 
@@
 \mu_{\bm{s},t} &=
-f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
+\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
 \alpha_g +
@@
 where TODO.
+TODO: get NB2 above and similar in previous examples
 

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
Fixes

@@
 ```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
-sanity(fit)
+sanity(fit_bin_rf)
 ```
@@
 a log link. We then compare alternative families, spatiotemporal random field
-structures, and the inclusion of anisotropy to illustrate the flexibility of
+structures, and the exclusion of anisotropy to illustrate the flexibility of
 \pkg{sdmTMB}.
@@
 dat <- select(dogfish, longitude, latitude, year, 
-  catch_weight, area_swept, depth)
+  catch_weight, area_swept, depth) |> as_tibble()
 head(dat, n = 3)

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
Minor edits

@@
 Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
-Poisson-link delta model & \checkmark & \checkmark & \checkmark & -- & -- & --\\
+Poisson-link delta models & \checkmark & \checkmark & \checkmark & -- & -- & --\\
 Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
@@
 
+```{r setoptions, echo=FALSE}
+options(
+  pillar.print_max = 3,
+  pillar.print_min = 3,
+  pillar.advice = FALSE,
+  pillar.width = 72
+)
+```
+
 ```{r libs-extras}
@@
 select(pcod, lat, lon, X, Y, depth, present) |> 
-  head(n = 3)
+  head()
 ```
@@
 
-```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="TODO"}
+```{r dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="TODO"}
 mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
@@
 ```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
-tidy(fit, effects = "ran_pars", conf.int = TRUE)
+tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
 ```
@@
 ```{r dog-head-dat, echo=TRUE}
-dat <- select(dogfish, longitude, latitude, year, 
+dat <- select(dogfish, lon = longitude, lat = latitude, year, 
   catch_weight, area_swept, depth) |> as_tibble()
-head(dat, n = 3)
+head(dat)
 ```
@@
 ```{r dog-utms, echo=TRUE}
-dat <- add_utm_columns(dat, c("longitude", "latitude"), 
+dat <- add_utm_columns(dat, c("lon", "lat"), 
   units = "km", utm_crs = 32609)

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
Writing; largely captions

@@
   pillar.advice = FALSE,
-  pillar.width = 72
+  pillar.width = 80
 )
+options(width = 80)
 ```
@@
 \pkg{fmesher} and supplied it to the `mesh` argument in `make_mesh()`. We can
-inspect our mesh object with the associated plotting method.
-
-```{r dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="TODO"}
+inspect our mesh object with the associated plotting method. Our mesh has `r mesh_pcod$mesh$n` (`mesh_pcod$mesh$n`) vertices.
+
+```{r dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
+The locations where lines intersect are referred to as ``vertices''.
+Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
+This is a simple mesh created with the convenience function \\code{make\\_mesh()}.
+A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5}
 mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
@@
 optimizer reported successful convergence, that the Hessian matrix is positive
-definite, that no extreme (XX) or small (XX) eigenvalues are detected, that no
-absolute log likelihood gradients with respect two fixed defects are >= 0.001,
-that all fixed effects have reported standard errors and that they do not look
-unreasonably large (XX by default), that random field marginal standard
-deviations are not unexpectedly small or large, and that the random field
-Matérn range parameter does not look unreasonably large.
+definite, that no extreme or small eigenvalues are detected, that no
+absolute log likelihood gradients with respect to fixed defects are >= 0.001,
+that all fixed effects have reported standard errors that do not look
+unreasonably large (< 100 by default), that random field marginal standard
+deviations are not unexpectedly small or large (> 0.01 and < 100), and that the random field
+Matérn range parameter does not look unreasonably large (< 1.5 times the largest edge of a bounding box around the observations).
 
@@
 inspection of the residuals calculated from the `residuals()` S3 method. We
-omit code for Moran's I calculation for brevity but include an example in the
+omit example code for Moran's I calculation for brevity but include it in the
 code supplement. The p-value for spatial autocorrelation is low for the model
@@
 We can also use `tidy()` to get a data frame of our random field parameters,
-where `sigma_O` is $\sigma_O$ in the above equations: the standard deviation of
+where `sigma_O` is $\sigma_O$ in the above equations---the standard deviation of
 the spatial random field $\bm{\omega}$.
@@
 
-TODO MOVE TO CAPTION? We have an effect of depth and a spatial random field,
-which combined in link space result in our overall prediction. Note the
-difference between only the fixed effects (left) and including the spatial
-random field (right). TODO DESCRIBE PLOTTING AND/OR INCLUDE
-
-```{r pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="TODO"}
+```{r pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod.
+Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link space, and (c) the overall prediction, which here is the combination of panels a and b.
+Note the difference between only the fixed effects (a) and including the spatial random field (c)."}
 plot_spatial_map <- function(dat, column, title) {
@@
 }
-g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
-g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
-g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
+g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
+g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
+g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
 cowplot::plot_grid(g1, g2, g3, ncol = 3)
@@
 
-```{r dog-aniso, echo=TRUE, fig.cap="TODO", out.width="4in"}
-plot_anisotropy(fit_dpl)
+(ref:dog-aniso)
+
+```{r dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
+Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled.
+The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in"}
+plot_anisotropy(fit)
 ```
@@
 ```{r dog-plot1, echo=FALSE}
+g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(est_non_rf1) +
+  scale_fill_viridis_c(trans = "log10") +
+  ggtitle("(a) Non-random-field components; first delta model")
+```
+
+```{r dog-plot2, echo=FALSE}
+g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(omega_s1) +
+  scale_fill_gradient2() +
+  ggtitle("(b) Spatial random field; first delta model")
+```
+
+```{r dog-plot3, echo=FALSE}
+g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
+  plot_map(epsilon_st2) +
+  scale_fill_gradient2() +
+  ggtitle("(c) Spatiotemporal random field; second delta model")
+```
+
+```{r dog-plot4, echo=FALSE}
 g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
@@
   # labs(fill = "Density") +
-  scale_fill_viridis_c(trans = "log10")
-```
-
-```{r dog-plot2, echo=FALSE}
-g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(omega_s1) +
-  scale_fill_gradient2()
-```
-
-```{r dog-plot3, echo=FALSE}
-g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(epsilon_st2) +
-  scale_fill_gradient2()
-```
-
-```{r dog-plot4, echo=FALSE}
-g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
-  plot_map(est_non_rf1) +
-  scale_fill_viridis_c(trans = "log10")
-```
-
-```{r dog-wvci-pred, fig.asp = 0.7, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
+  scale_fill_viridis_c(trans = "log10") +
+  ggtitle("(d) Overall prediction")
+```
+
+```{r dog-wvci-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density.
+Throughout, two example years are shown.
+\\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects.
+The spatial random field is constant through time and represents TODO.
+The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process.
+They represent TODO.", fig.width=9, out.width="6in"}
 cowplot::plot_grid(

### 2023-09-14 doc/paper-jss/paper-jss.Rmd
more edits to captions, poisson-link ref

@@
 Estimation is done via maximum marginal likelihood with the `sdmTMB()` function, and the resulting fitted model can be used to make predictions (`predict.sdmTMB()`) or calculate residuals (`residuals.sdmTMB()`), derive additional quantities of interest, and simulate Bayesian posterior distributions.
-Common quantities of interest may include summarizing shifts in the center of gravity (`get_cog()`) [e.g., @thorson2016cog] or in the field of ecology, tracking total population densities through time (`get_index()`).
-<!-- TODO: describe this in more detail and how it could apply beyond ecology; maybe a derived variables subsection? -->
+Common quantities of interest may include summarizing shifts in the center of gravity (`get_cog()`) [e.g., @thorson2016cog], or tracking total densities through time (`get_index()`).
+The `get_index()` function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+<!-- TODO: describe this in more detail and how it could apply beyond ecology; maybe a derived variables subsection?
+Expanded this sentence a bit -- COGs are pretty common in other fields, but I don't know about indices -- removed `ecology` from the description-->
 
@@
 delta-lognormal (REF), and delta-Poisson-link-gamma and
-delta-Poisson-link-lognormal (REF). Note that we choose to omit the
+delta-Poisson-link-lognormal (REF).
+The Poisson-link models depart from the conventional delta models in treating the probability of occurrence at a given location ($p_i$) as a Poisson process rather than a Bernoulli one, $p_i = 1 - exp(a_i * n_i)$, where $n_i$ is the total number of individuals available to be encountered, and $a_i$ is a measure of sampling effort [@thorson2018_poissonlink].
+Note that we choose to omit the
 spatiotemporal random fields from the binomial model component 
@@
 
-TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.
+<!-- TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE. -->
 
@@
 
-where TODO.
+where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$.
 
@@
 \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects.
-The spatial random field is constant through time and represents TODO.
+The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g.
+habitat).
 The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process.
-They represent TODO.", fig.width=9, out.width="6in"}
+They represent temporal variability in the spatial patterning of Pacific Dogfish (resulting from movement, local changes in population densities, etc).", fig.width=9, out.width="6in"}
 cowplot::plot_grid(
@@
 
-We can write the model as TODO
+We can write the model as
 
@@
 
-where TODO.
-TODO: get NB2 above and similar in previous examples
+where three spatial fields are now estimated: the spatial intercept ($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and spatially varying NAO coefficients ($\zeta_{\bm{s}}$).
 
@@
 
+There are a number of planned additions to the sdmTMB model structure.
+A subset of features to be added in the near future includes (1) multivariate models to allow for simultaneous modeling of data from multiple species, (2) integration with the RTMB package so that the model code base will be written in R rather than C++, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB} [@brooks2017], and (4) estimation of random slopes.
+
 <!-- TODO: RTMB -->

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Switch to .Rnw, lots of editing/writing

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

\usepackage[utf8]{inputenc}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
Sean C. Anderson~\orcidlink{0000-0001-9563-1937}\\Pacific Biological
Station \And Eric J. Ward~\orcidlink{0000-0002-4359-0296}\\Northwest
Fisheries Science Center \AND Philina A.
English~\orcidlink{0000-0003-2992-6782}\\Pacific Biological
Station \And Lewis A. K. Barnett~\orcidlink{0000-0002-9381-8375}\\Alaska
Fisheries Science Center
}
\title{\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and
User-Friendly Generalized Linear Mixed Effects Models with Spatial and
Spatiotemporal Random Fields}

\Plainauthor{Sean C. Anderson, Eric J. Ward, Philina A. English, Lewis
A. K. Barnett}
\Plaintitle{sdmTMB: An R Package for Fast, Flexible, and Accessible
Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal
Random Fields}
\Shorttitle{\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with
\pkg{TMB}}

\Abstract{Geostatistical data are common across scientific fields and available
  in increasingly large datasets. However, appropriate models to analyse these
  data, such as generalised linear mixed effects models with Gaussian random
  fields, are computationally intensive and challenging to implement for many
  users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends
  the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or
  \pkg{mgcv} to include spatial and spatiotemporal latent fields using
  a predictive-process SPDE-(stochastic partial differential equation) based
  approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and
  estimation is conducted via maximum marginal likelihood with Template Model
  Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and
  \pkg{rstan}. We describe the model and explore case studies that illustrate
  \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary
  processes (time-varying and spatially varying coefficients), and anisotropy
  (directionally dependent spatial correlation). Additional functionality
  includes hurdle models, break-point effects, and out-of-sample cross
  validation. Finally, we compare the functionality, speed, and interfaces of
  related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude
  faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible
interface will help open this useful class of models to a wider field of
geostatistical analysts.}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.

\Keywords{Gaussian Markov random fields (GMRF), generalized linear mixed effects
models (GLMM), INLA, SPDE, species distribution modelling, R package,
spatio-temporal, spatial-temporal, Template Model Builder}

\Plainkeywords{Gaussian Markov random fields (GMRF), generalized linear mixed
effects models (GLMM), INLA, SPDE, species distribution modelling, R package,
spatio-temporal, spatial-temporal, Template Model Builder}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).

\Address{
    Sean C. Anderson\\
    Pacific Biological Station\\
    Fisheries and Oceans Canada, Nanaimo, BC, Canada\\
  E-mail: \email{sean.anderson@dfo-mpo.gc.ca}\\

      Eric J. Ward\\
    Northwest Fisheries Science Center\\
    National Oceanic and Atmospheric Administration, National Marine
Fisheries Service, Seattle, WA, USA\\

      Philina A. English\\
    Pacific Biological Station\\
    Fisheries and Oceans Canada, Nanaimo, BC, Canada\\

      Lewis A. K. Barnett\\
    Alaska Fisheries Science Center\\
    National Marine Fisheries Service, National Oceanic and Atmospheric
Administration, Seattle, WA, USA\\

  }

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amsmath} \usepackage{amssymb} \usepackage{bm} \usepackage{lscape}
\usepackage{tablefootnote} \usepackage{threeparttable} \usepackage{booktabs}
\usepackage{pifont} \usepackage{newunicodechar} \usepackage{gensymb}

\newunicodechar{✓}{\ding{51}} \newunicodechar{✗}{\ding{55}}
\newunicodechar{˚}{\degree} \DeclareGraphicsExtensions{.pdf,.png,.jpg}

\widowpenalty10000 \clubpenalty10000

\begin{document}

<<preliminaries, echo=FALSE, results='hide', include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(engine = 'R', tidy = FALSE)
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
@

<<main-setup, include=FALSE, cache=FALSE>>=
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all",
  highlight = FALSE
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
opts_chunk$set(prompt = TRUE)
options(replace.assign = TRUE, width = 72, prompt = "R> ")
knitr::render_sweave()
@

\clearpage

\section{Introduction}\label{introduction}

Data are often collected in space or in space repeatedly over time. While such
data are a rich source of information \citep[e.g.,][]{legendre1989a, rossi1992,
tilman1997spatial}, they are challenging to properly model---data closer in
space and time are usually more similar to each other than data farther apart
due to measured and unmeasured variables \citep{cressie1993, diggle2007,
cressie2011}. While measured variables can be accounted for with predictors in
a model (e.g., measuring and modelling temperature effects on species
abundance), unmeasured variables (e.g., everything influencing species abundance
but not explicitly modelled) can cause residual spatial correlation. Accounting
for this residual correlation is important because doing so allows for valid
statistical inference \citep{legendre1989a, dormann2007}, can improve
predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself
\citep[e.g.,][]{thorson2019d, barnett2021}.

Geostatistical generalized linear mixed effects models (GLMMs) with spatially
correlated random effects are a class of models appropriate for these data
\citep{rue2005gmrf, diggle2007, cressie2011}. Similarly to how random intercepts
can account for correlation among groups, spatial or spatiotemporal random
effects can account for unmeasured variables causing observations to be
correlated in space or space and time. A common approach to modelling these
spatial effects is with Gaussian random fields (GRFs), where the random effects
describing the spatial patterning are assumed to be drawn from a multivariate
normal (MVN) distribution, constrained by some covariance function such as the
exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.

Such models quickly become computationally limiting due to the need to invert
large matrices to keep track of covariation among data
\citep[e.g.,][]{rue2005gmrf, latimer2009}. Many solutions have been proposed,
such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic
partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011},
and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}. These
approaches reduce the scale of the covariance estimation problem while providing
a means to evaluate the data likelihood, thereby allowing fitting via Bayesian
\citep{gelfand2017} or maximum likelihood methods. This can greatly improve
computational efficiency \citep[e.g.,][]{heaton2019}. The SPDE approach is
a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009,
lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template
Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to
create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
Details are beyond the scope of this paper and are not necessary to use the
software discussed here, but the idea is that the solution to a specific SPDE is
a GRF with a Matérn covariance function and this `trick' enables one to
efficiently fit approximations to GRFs to large spatial datasets
\citep{lindgren2011}.

Software packages for specifying statistical models that can include the SPDE,
such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are
challenging to use for many applied researchers. For example, \pkg{TMB} requires
the user to program in a \proglang{C++} template and it can be slow to
experiment with multiple models when writing bespoke model code. Packages such
as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users
quickly iterate and explore statistical models---focusing on evaluating fit and
comparing models---but do not have built-in SPDE functionality. Packages such as
\pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful
user interfaces to fit spatial models that use the SPDE, but they either lack
a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB},
or lack some functionality. We provide a more detailed comparison of related
software packages in Table \ref{tab:functionality} and the Discussion.

Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements
geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model
fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices. Our aim is not to
replace the above-mentioned statistical packages, but to provide a fast,
flexible, and user-friendly interface that is familiar to users of \pkg{lme4},
\pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial
and spatiotemporal models. Many features of \pkg{sdmTMB} may be found in other
software (Table 1) but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models
(SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}). This
paper describes the basic functionality of this \proglang{R} package and its
underlying statistical model, illustrates its use through two case studies, and
concludes with a comparison to related software.

\section{Model description}\label{model-description}

\pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data refers to data observed at specific spatial coordinates
reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
data can be collected across discrete points in time. Areal data (data
aggregated to polygon or grid level) may be analyzed using other spatial models,
including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018};
\pkg{sdmTMB} can also fit models with areal data if each polygon has an
associated centroid. A benefit of the geostatistical approach over CAR or
similar models is that the parameters describing spatial covariance can be more
easily interpreted \citep{wall2004} (e.g., spatial variance and range---the
distance at which points are effectively independent).

The process component of an \pkg{sdmTMB} model can be formed by any combination
of main (``fixed'') effects \(\bm{\beta}\) and offset \(O_{\bm{s},t}\), IID
(independent and identically distributed) random intercepts \(\alpha_g\),
spatial intercept random fields \(\omega_{\bm{s}}\), spatiotemporal intercept
random fields \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\),
and spatially varying effects \(\zeta_{\bm{s}}\) (Figure \ref{fig:diagram}).
These components are discussed in more detail via our examples below.

\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\alpha_g +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} \right),
\end{aligned}
\]
This process component is combined with an observation error family (e.g.,
Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
families can be combined into a two-part ``delta'' or ``hurdle'' model
\citep{aitchison1955} to model the zero vs.~non-zero observations separately
from the positive observations. See the model description vignette for an
expanded description of the statistical model
(\texttt{browseVignettes("sdmTMB")} or see the
\href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they
estimate interpretable parameters of a spatial covariance function: parameters
defining the magnitude of spatial variation and the rate of correlation decay
with distance. In contrast, semi- and non-parametric approaches do not estimate
spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002},
\pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv}
\citep{wood2017a}). The random fields in \pkg{sdmTMB} are structured as MVN
constrained by a Matérn covariance function \citep{matern1960}. The Matérn can
accommodate a range of shapes and can be both isotropic (correlation decays the
same in all directions) or anisotropic (correlation is directionally dependent)
\citep{haskard2007}. The Matérn standard deviations are estimated separately for
the various fields and the range---the distance at which spatial correlation
decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields
or estimated separately (\texttt{share\_range} argument).

By default, if spatiotemporal fields are included, they are assumed IID;
however, additional options allow them to be modelled as a random walk or
first-order autoregressive, AR(1), process (Figure \ref{fig:diagram}). Turning
off both spatial and spatiotemporal effects allows for comparison with standard
non-spatial GLMs or GLMMs. We include additional flexibility in specifying the
linear main effect matrix: covariates can be modelled as penalized smooth
functions (generalized additive models, GAMs) using the same \texttt{s()} syntax
as in \pkg{mgcv} \citep{wood2017a} (thereby allowing automatic selection of
smoother `wiggliness'), or can be modelled with threshold shapes: hockey-stick
models, \texttt{breakpt()}, \citep{barrowman2000} or logistic functions,
\texttt{logistic()}. \pkg{sdmTMB} can use many \texttt{mgcv::s()} and
\texttt{mgcv::t2()} smoothers including bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by =
x2)}, cyclical smoothers \texttt{s(x, bs = "cc")}, and smoothers with
specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.

The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
marginal log likelihood and its gradient, and the negative log likelihood is
minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
\proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
maximize the log likelihood conditional on the estimated fixed effects and are
integrated over via the Laplace approximation \citep{kristensen2016}. After
rapid model exploration with maximum likelihood, one can optionally pass an
\pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
distribution for Bayesian inference.

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to
model parameters (\texttt{?sdmTMBpriors}). These priors may be useful in cases
where estimation is difficult because of identifiability issues or relatively
flat likelihood surfaces, or to impart prior information or achieve
regularization. Following other recent SPDE implementations in \pkg{TMB}
\citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors
\citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the
spatial range and variance parameters.

The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both
modular and familiar to users of widely used \proglang{R} packages
(e.g.~\pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or
without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()}
respectively). Estimation is done via maximum marginal likelihood with the
\texttt{sdmTMB()} function, and the resulting fitted model can be used to make
predictions (\texttt{predict.sdmTMB()}) or calculate residuals
(\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and
simulate Bayesian posterior distributions. Common quantities of interest may
include summarizing shifts in the center of gravity (\texttt{get\_cog()})
\citep[e.g.,][]{thorson2016cog} or in the field of ecology, tracking total
population densities through time (\texttt{get\_index()}).

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

\section{Model validation and selection}\label{model-validation-and-selection}

Validation and selection of state-space models is challenging, particularly when
using the Laplace approximation \citep{thygesen2017a}. We provide several
approaches to assist this process: (1) The Akaike Information Criterion
\citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
has well-documented biases with mixed-effects models \citep{liang2008}. (2)
Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
with user-specified or randomly chosen folds for model selection or to evaluate
goodness of fit according to user-calculated criteria (e.g., mean squared error,
area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
\pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
approximation, or perform posterior predictive checks (see
\texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
randomized quantile \citep{dunn1996} or probability integral transform (PIT)
\citep{smith1985} residuals. For state-space models, these residuals have known
statistical issues with the Laplace approximation \citep{thygesen2017a} but are
quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
of the random effects to avoid this issue \citep{rufener2021} is recommended but
slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
simulating data without starting from a fitted model. Models can be fit to these
simulated data to ensure identifiability, evaluate bias and precision in
parameter estimation, or evaluate the consequences of model misspecification.

\clearpage

\section{Installation}\label{installation}

\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R}
Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=sdmTMB}

<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
install.packages("sdmTMB", dependencies = TRUE)
@

Importantly, \texttt{dependencies = TRUE} includes the \proglang{R}-INLA
package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is
needed for the \texttt{make\_mesh()} function below. Soon, we intend to switch
to using the \pkg{fmesher} package to eliminate this dependency.

Alternatively, the development version can be installed with:

<<remotes, eval=FALSE, echo=TRUE>>=
# install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
@

Additional utilities are maintained in the \pkg{sdmTMBextra} package at
\url{https://github.com/pbs-assess/sdmTMBextra}.

\section{Example: spatial species distribution modelling}\label{pcod}

We begin with a simple species distribution model of encounter probability of
Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen
Charlotte Sound, British Columbia, Canada. The purpose of our example is to
illustrate the need for spatial random fields. This survey is conducted by
Fisheries and Oceans Canada (REF) and follows a depth-stratified random sampling
design, resulting in a georeferenced dataset. The data frame \texttt{pcod} is
available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and
encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of
Pacific Cod for a given survey sample.

<<libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE>>=
library(sdmTMB)
library(dplyr)
library(ggplot2)
@

<<setoptions, echo=FALSE>>=
options(
  pillar.print_max = 3,
  pillar.print_min = 3,
  pillar.advice = FALSE,
  pillar.width = 80
)
options(width = 80)
@

<<libs-extras>>=
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
@

<<pcod-head, echo=TRUE>>=
select(pcod, lat, lon, X, Y, depth, present) |>
  head()
@

\subsection{Adding UTM columns}

An \pkg{sdmTMB} model requires a data frame that contains a response column,
columns for any predictors, and columns for spatial coordinates. Usually it
makes sense to convert the spatial coordinates to an equidistant projection such
as UTMs to ensure that distance remains constant throughout the study region
\citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}. Here we use
the helper function \texttt{add\_utm\_columns()} to add UTM coordinates with km
units (so our estimated spatial range parameter is not too big or small). By
default, the function will guess the UTM zone and create new columns \texttt{X}
and \texttt{Y}. Since our example data already has these UTM columns, we can
skip running this code.

<<pcod-utms-eval, echo=TRUE, eval=FALSE>>=
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
@

\subsection{SPDE mesh creation}

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach. The argument \texttt{cutoff} defines
the minimum allowed distance between mesh vertices in the units of \texttt{X}
and \texttt{Y}. \texttt{make\_mesh()} is a wrapper function for
\texttt{INLA::inla.mesh.create()} or \texttt{fmesher::fm\_rcdt\_2d\_inla()} and
\texttt{cutoff} is passed through to these functions. Alternatively, we could
have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied
it to the \texttt{mesh} argument in \texttt{make\_mesh()}. We recommend
constructing such a mesh for applied analyses to take advantage of
additional functionality such as coarser mesh triangles in a border around the
data and control over maximum triangle length and minimum triangle angles. We
can inspect mesh object with the associated plotting method. Our mesh has
238 (\texttt{mesh\_pcod\$mesh\$n}) vertices.

<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. This is a simple mesh created with the convenience function \\code{make\\_mesh()}. A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
@

\subsection{Fitting the model}

We will fit a logistic regression of encounter probability with and without
spatial random fields to illustrate the importance of accounting for spatial
correlation. In addition to the spatial random field, we include an intercept
and a quadratic effect of depth on the probability of encounter. Our random
field model can be written as
\[
\begin{aligned}
y_{\bm{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\bm{s}}\right),\\
\mu_{\bm{s}} &= \operatorname{logit}^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s}} \bm{\beta} +
\omega_{\bm{s}}
\right),
\end{aligned}
\]
where \(\bm{X}^{\mathrm{main}}_{\bm{s}}\) represents a matrix of main effect
covariates (intercept, quadratic effects of depth), \(\bm{\beta}\) represents
a vector of estimated main effect coefficients, and \(\omega_{\bm{s}}\)
represents the estimated spatial field, \(\omega_{\bm{s}} \sim
\mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)\).

We can implement this model with \texttt{sdmTMB()}:

<<pcod-fit, echo=TRUE>>=
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod, 
  mesh = mesh_pcod,
  spatial = "on",
  family = binomial(link = "logit")
)
@

We can also fit a version that omits the spatial random field by setting
\texttt{spatial = "off"}. We will use the \texttt{update()} S3 method to refit
the model updating any specified arguments:

<<pcod-fit-off, echo=TRUE>>=
fit_bin <- update(fit_bin_rf, spatial = "off")
@

\subsection{Inspecting the model}

We can run some basic checks on our model with the \texttt{sanity()} function:

<<pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE>>=
sanity(fit_bin_rf)
@

\begin{verbatim}
#> ✓ Non-linear minimizer suggests successful convergence
#> ✓ Hessian matrix is positive definite
#> ✓ No extreme or very small eigenvalues detected
#> ✓ No gradients with respect to fixed effects are >= 0.001
#> ✓ No fixed-effect standard errors are NA
#> ✓ No fixed-effect standard errors look unreasonably large
#> ✓ No sigma parameters are < 0.01
#> ✓ No sigma parameters are > 100
#> ✓ Range parameter doesn't look unreasonably large
\end{verbatim}

Here, this does not flag any issues. \texttt{sanity()} is checking that the
\texttt{nlminb()} optimizer reported successful convergence, that the Hessian
matrix is positive definite, that no extreme or small eigenvalues are detected,
that no absolute log likelihood gradients with respect to fixed defects are
$\le$ 0.001, that all fixed effects have reported standard errors that do not
look unreasonably large ($<$ 100 by default), that random field marginal
standard deviations are not unexpectedly small or large ($>$
0.01 and $<$ 100), and that the random field Matérn range parameter does not
  look unreasonably large ($<$ 1.5 times the largest edge of a bounding box
  around the observations).

We can get a summary of our model fit:

<<pcod-bin-summary, echo=TRUE>>=
summary(fit_bin_rf)
@

The output indicates our model was fit by maximum (marginal) likelihood
(\texttt{ML}). We also see the formula, mesh, fitted data, and family. Next we
see any estimated main effects, the Matérn range distance, the spatial random
field standard deviation, and the negative log likelihood at convergence.

We can use the \texttt{tidy()} function (formatted to match the \pkg{broom}
package TODO) to obtain a data frame with parameter estimates (standard methods
such as \texttt{fixef()} and \texttt{vcov()} are also available). The standard
errors on our fixed effects have increased with the spatial random field:

<<pcod-tidy, eval=TRUE, echo=TRUE, results='markup'>>=
tidy(fit_bin_rf, conf.int = TRUE)
tidy(fit_bin, conf.int = TRUE)
@

By setting \texttt{effects = "ran\_pars"}, \texttt{tidy()} will return random
field parameters, where \texttt{sigma\_O} is \(\sigma_O\) in the above
equations---the standard deviation of the spatial random field \(\bm{\omega}\).

<<pcod-eg1-tidy-re, eval=TRUE, echo=TRUE>>=
tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
@

\subsection{Checking the effect of including a random field}

We can test for spatial autocorrelation with a Moran's I test or with a visual
inspection of the residuals calculated from the \texttt{residuals()} S3 method.
We omit example code for Moran's I calculation for brevity but include it in the
code supplement. The p-value for spatial autocorrelation is low for the model
without a random field (p $<$ 0.01) but large for the model with
a random field (p = 0.84) suggesting significant residual correlation in the
simpler model that is alleviated by including the random field.

<<morans, echo=FALSE, results='hide'>>=
test_autocor <- function(obj) {
  set.seed(1)
  s <- simulate(obj, nsim = 500)
  pr <- predict(obj, type = "response")$est
  r <- DHARMa::createDHARMa(
    simulatedResponse = s,
    observedResponse = pcod$present,
    fittedPredictedResponse = pr
  )
  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
}
(t_no_rf <- test_autocor(fit_bin))
(t_rf <- test_autocor(fit_bin_rf))
p_rf <- round(t_rf$p.value, 2)
@

We can also see that the AIC of the model with spatial random fields is
considerably lower:

<<pcod-aic, eval=TRUE, echo=TRUE, results='markup'>>=
AIC(fit_bin_rf, fit_bin)
@

\subsection{Comparing models with cross validation}

We can do additional model comparison with cross validation via
\texttt{sdmTMB\_cv()}. Here we will do 10-fold cross validation with the folds
constructed randomly. Using the \texttt{fold\_ids} argument, we could supply our
own folds and conduct spatially blocked cross validation. If we set a parallel
plan with the \pkg{future} (REF TODO) package, our folds will be fit in
parallel.

<<pcod-cv-future, eval=FALSE, echo=TRUE>>=
library(future)
plan(multisession)
@

<<pcod-cv, eval=TRUE, echo=TRUE>>=
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "on",
  family = binomial(), k_folds = 10
)
cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "off",
  family = binomial(), k_folds = 10
)
@

Indeed, the log likelihood of left-out data is considerably larger for the model
that includes random fields:

<<pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'>>=
cv_bin_rf$sum_loglik
cv_bin$sum_loglik
@

\subsection{Visualizing model predictions}

\sloppy To visualize our model, we can make predictions with the
\texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the
\texttt{newdata} argument to predict on a new data frame with any locations and
values for the predictor columns. Here, we will predict on a 2 \(\times\) 2 km
grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can
visualize the predictions spatially. The grid contains spatial covariate columns
and all predictors used in the model set at values for which we want to predict.
The output of \texttt{predict()} is a data frame containing overall estimates in
link space (\texttt{est}), estimates from the non-random-field components
(\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the
individual random field components (here, \texttt{omega\_s}---the spatial
field).

<<pcod-predict, echo=TRUE, results='markup'>>=
p <- predict(fit_bin_rf, newdata = qcs_grid)
select(p, X, Y, depth, est, est_non_rf, omega_s) |>
  as_tibble() |> head(n = 2)
@

<<pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod. Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link (logit) space, and (c) the overall prediction, which here is the combination of panels a and b. Note the difference between predictions from depth alone (a) and predictions including a spatial random field (c).">>=
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position= "bottom") +
    ggtitle(title) +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
@

\section{Example: spatiotemporal species distribution modeling}\label{dogfish}

As a second example, we will construct a spatiotemporal model of catch rates of
Pacific Spiny Dogfish (\emph{Squalus suckleyi}) from a trawl survey off the west
coast of Vancouver Island, Canada. This example extends the spatial model
described above by also including (1) spatiotemporal fields allowing the
distribution in each year to be unique, (2) a time-varying intercept as an AR(1)
process to allow year effects to vary but remain autocorrelated, (3) a smooth
effect of depth, allowing catch rates to vary non-linearly with depth, and (4)
spatial anisotropy allowing spatial correlation to be directionally dependent.
Since catch rates are positive, continuous, and contain zeros, we begin by
specifying the response family as a Tweedie distribution \citep{tweedie1984}
with a log link. We then compare alternative families, spatiotemporal random
field structures, and the exclusion of anisotropy to illustrate the flexibility
of \pkg{sdmTMB}.

The dataset includes spatial coordinates, year, dogfish catch weight in kg, area
swept in km, and bottom depth:

<<dog-head-dat, echo=TRUE>>=
dat <- select(dogfish, lon = longitude, lat = latitude, year, 
  catch_weight, area_swept, depth) |> as_tibble()
head(dat)
@

We add UTM zone 9 columns, create a log depth column for convenience, and create
a basic mesh:

<<dog-utms, echo=TRUE>>=
dat <- add_utm_columns(dat, c("lon", "lat"), 
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@

<<dog-mesh2, eval=FALSE>>=
plot(mesh)
@

We can then specify our model. We include an offset for the effort variable log
area swept such that we are effectively modelling density and our predictions
will be for an area swept of 1 km\textsuperscript{2}.

Our model can be written as TODO

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),
\end{aligned}
\]

where \(\bm{\beta}\) are coefficients associated with the main effects,
\(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\)
represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial
field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents
spatiotemporal fields.

The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1)
process,

\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
\end{aligned}
\]

<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE,
  family = tweedie()
)
@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal (REF), and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal (REF). Note that we choose to omit the
spatiotemporal random fields from the binomial model component
(\texttt{spatiotemporal = list("off", "iid")}) because initial fits indicated
this random field was collapsing to zero indicating minimal changes in encounter
spatial distribution from year to year.

% TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.

<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_dg <- update(fit_tw, family = delta_gamma(), 
  spatiotemporal = list("off", "iid"))
fit_dl <- update(fit_dg, family = delta_lognormal())
fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
@

We can then compare the models via AIC:

<<dog-aic, echo=TRUE>>=
AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
  mutate(delta_AIC = AIC - min(AIC)) |> 
  arrange(delta_AIC)
@

We find that the Poisson-link delta-lognormal model is favoured with AIC. We
next test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic, and structuring the positive
model component spatiotemporal random fields as AR(1) to allow spatiotemporal
hotspots to persist from year to year.

The AR(1) fields can be represented as TODO.

\[
\begin{aligned}
\bm{\delta}_{t=1} &\sim \operatorname{MVN} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
\bm{\epsilon_{t}} \sim \operatorname{MVN} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]

where TODO.

<<dog-ar1, results='hide', echo=TRUE>>=
fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
@

<<dog-aci2, echo=TRUE>>=
AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
@

% TODO: Need description of anisotropy and why it's important here or somewhere else

We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
is important here given the elongated shape of the continental shelf with
a rapid transition to deeper water. On inspecting the model output with
\texttt{summary()}, we notice that the spatial random field SD in the positive
model component has become very small once AR(1) spatiotemporal fields were
added. We therefore consider a model where these are turned off in the second
component of the delta model:

<<dog-ar1-2, echo=TRUE, results="hide">>=
fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
@

<<dog-aic3, echo=TRUE>>=
AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
@

We save our chosen model to the object \texttt{fit} to simplify subsequent code,
run the \texttt{sanity()} check (suppressed for space), and inspect
\texttt{summary()}:

<<dog-print, echo=TRUE>>=
fit <- fit_dpl_ar1_only
sanity(fit)
summary(fit)
@

The output is more complex compared to our binomial spatial model. We now have
two model components, which are shown one after the other. Starting with the
binomial component, we have output from the smoother, which includes a linear
component (\texttt{slog\_depth}) and the standard deviation on the smoother
weights (\texttt{sds(log\_depth)}). The smoother summary follows the format used
in the \pkg{brms} package \citep{burkner2017}. Next, we have the time-varying
intercepts and information on our anisotropic spatial correlation. We then have
the second model component (lognormal) with a similar summary structure but with
the addition of a dispersion parameter for the lognormal, the AR(1) correlation
of the spatiotemporal random fields, and a spatiotemporal random field marginal
standard deviation. We can use \texttt{plot\_anisotropy()} to visually inspect
the anisotropy (Figure \ref{fig:dog-aniso}).

<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in">>=
plot_anisotropy(fit)
@

Similarly to the first example, we can visualize model predictions on a grid
covering the area of interest. Because this is a spatiotemporal model, we first
need to replicate our grid for each year we will predict on. Since this is
a common operation, we include the convenience function \texttt{replicate\_df()}
to replicate a data frame. We then ensure our data frame contains all the
predictors used in the model (here \texttt{log\_depth}).

% TODO Propose moving this to a short paragraph `Prediction` to highlight modularity

<<dog-grid, echo=TRUE>>=
grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
grid$log_depth <- log(grid$depth)
head(grid, n = 2)
@

<<dog-pred1, echo=TRUE>>=
pred <- predict(fit, newdata = grid, type = "response")
@

<<dog-pred2, echo=TRUE>>=
names(pred)
@

Our prediction data frame is similar to the binomial spatial model, but includes
columns for the two delta model components (labelled with suffixes \texttt{1}
and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
random effects. We can easily generate plots from this data frame (e.g., with
\texttt{ggplot2::geom\_raster()}). We suppress that code in this paper for
brevity.

<<plot-map, echo=FALSE>>=
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(vars(year)) +
    coord_fixed() +
    theme(legend.position= "bottom") +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
@

<<dog-plot1, echo=FALSE>>=
g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(est_non_rf1) +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("(a) Non-random-field components; first delta model")
@

<<dog-plot2, echo=FALSE>>=
g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(omega_s1) +
  scale_fill_gradient2() +
  ggtitle("(b) Spatial random field; first delta model")
@

<<dog-plot3, echo=FALSE>>=
g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(epsilon_st2) +
  scale_fill_gradient2() +
  ggtitle("(c) Spatiotemporal random field; second delta model")
@

<<dog-plot4, echo=FALSE>>=
g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(est) +
  # labs(fill = "Density") +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("(d) Overall prediction")
@

<<dog-wvci-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time and represents TODO. The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent TODO.", fig.width=9, out.width="6in">>=
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
@

We can visualize the conditional effect of the bottom depth smoother by
predicting across a sequence of depths and holding other variables at reference
values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
to include both delta model components (\texttt{model = NA}), omit the random
fields (\texttt{re\_form = NA}), and return standard errors (\texttt{se\_fit\
= TRUE}).

<<dog-depth, echo=TRUE, out.width="5in">>=
nd <- data.frame(
  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
  year = max(dat$year)
)
pred_depth <- predict(
  fit, newdata = nd, 
  model = NA, re_form = NA, se_fit = TRUE
)
@

<<dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
@

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the \texttt{get\_index()} function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4
km\textsuperscript{2}) to the \texttt{area} argument and specify
\texttt{bias\_correct = TRUE} to enable a generic epsilon-bias correction
needed due to the non-linear transformation of the random effects (REF).

% TODO and the Laplace?

<<dog-index, echo=TRUE>>=
pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
@

<<dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
@

\clearpage

\section{Example: spatially varying coefficients} \label{svc-owls}

In our final example, we demonstrate a model with spatially varying coefficient
(SVC) effects and illustrate combining uncertainty from parameters by working
with draws from the joint parameter precision matrix. Spatially varying
coefficients \ldots{} TODO REFS. In addition, we illustrate using restricted
maximum likelihood (REML), which integrates over the fixed effects as well as
the random effects.

Snowy Owls (\emph{Bubo scandiacus}) breed on the arctic tundra and are irruptive
migrants, meaning that they appear across the mid-latitudes of North America in
much greater numbers in some winters than others. The reasons for this
interannual variation in the number of individuals migrating south are not well
understood but seem to be related to high abundances of food during the breeding
season and therefore sharp increases in breeding ground population densities
\citep{robillard2016}. The North Atlantic Oscillation Index (NAO) has been
linked to productivity of both owls and their prey in Europe \citep{millon2014}.
Because both productivity and the choice of wintering location could be
influenced by climate, we modelled a spatially varying effect of annual mean NAO
index on winter abundance across the southern boundary of their winter
distribution. We fit counts observed in North America during annual Christmas
Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011}
distribution, random intercepts for year, spatial and spatiotemporal random
fields, and a spatially varying coefficient associated with the NAO.

We can write the model as TODO

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\alpha_g +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
\]

where TODO. TODO: get NB2 above and similar in previous examples

<<owl-data>>=
snow <- readRDS(here::here("data/SNOW_data.rds"))
@

<<owl-fit, echo=TRUE, results = "hide">>=
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh_snow,
  reml = TRUE,
  silent = FALSE
)
@

<<owl-sanity, eval=FALSE>>=
sanity(fit_owl)
@

<<owl-print, echo=TRUE>>=
summary(fit_owl)
@

The spatially varying coefficient effect at any point in space is the
combination of the main effect and spatial random effect for \texttt{nao}. We
illustrate deriving uncertainty on their combination by simulating from the
parameter precision matrix assuming the joint parameter covariance matrix in
multivariate normal. We do this by specifying a non-null number of simulation
draws to \texttt{nsim} in \texttt{predict.sdmTMB()}. By default, \texttt{nsim >
0} will return a matrix of draws from the overall prediction. Here, we instead
   specify that we want to return draws from the \texttt{zeta\_s} random field,
   which is the spatially varying random field. This returns a matrix where each
   row matches a row of \texttt{newdata} and each column is a simulation draw.
   We then use the function \texttt{spread\_sims()} to draw 200 simulations from
   the parameters themselves, combine the main effect draws (\texttt{sims\$nao})
   with the random field values \texttt{zeta\_s}, and calculate the median for
   each row of data so we can plot them.

<<p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
dim(zeta_s)
sims <- spread_sims(fit_owl, nsim = 200)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- apply(exp(combined), 2, median)
@

We can make a basic plot using the following code. A more elaborate
version is shown in Figure \ref{fig:owl-plot-fancy}.

<<owl-plot-basic, echo=TRUE, eval=FALSE>>=
ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
@

Overall, we find a weak average positive effect of annual mean NAO on overall
counts, but a southeast to northwest gradient in the intensity of the effect
(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
to the Atlantic coast and those migrating the furthest south being the most
affected by NAO.

% TODO keep this? We provide an extended version of this example in Appendix \ref{app:owls}.

<<shapes, echo=FALSE>>=
if (!file.exists(here::here("data/ne_10m_lakes"))) {
  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
    "download/10m/physical/ne_10m_lakes.zip")
  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
}
@

<<shapes-read, echo=FALSE>>=
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
@

<<owl-proj2>>=
# project to get nice axis limits 
snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
snow2 <- snow2 |> mutate(x = X, y = Y) |> 
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5>>=
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
@

\clearpage

\section{Package comparisons}\label{package-comparisons}

TODO: BEFORE OR AFTER EXAMPLES

There are many \proglang{R} packages capable of fitting geostatistical spatial
or spatiotemporal models \citep[e.g.,][]{heaton2019}. \pkg{sdmTMB}, \pkg{VAST},
\proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are
the most closely related, as they all provide a user interface to SPDE-based GRF
models. In our software comparison (Table \ref{tab:functionality}), we also
include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and
\pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package
that can fit related predictive-process models without the SPDE. \pkg{sdmTMB},
\pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas
\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to
isotropic covariance. \pkg{sdmTMB} and \pkg{mgcv} focus on univariate response
data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM},
and \pkg{spBayes} extend to multivariate responses with various limitations. To
our knowledge, \pkg{VAST} is the only package to implement spatial
\citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and
spatial empirical orthogonal function (EOF) regression \citep{thorson2020}. Of
these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold
(e.g., hockey-stick) covariate relationships. \pkg{spaMM} is limited to
a spatial random field (i.e., does not fit spatiotemporal fields) and
\pkg{spBayes} implements spatiotemporal fields, but only as a random walk. There
is considerable variability in the available observation likelihoods across
packages (Table \ref{tab:functionality}). We provide comparisons of the syntax
and the reproducibility of results from models fit using \pkg{sdmTMB},
\proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.

\begin{table}

\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \pkg{sdmTMB} & \pkg{VAST} & \proglang{R}-\pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta models & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\end{threeparttable}
\label{tab:functionality}

\caption{Comparison of functionality between several \proglang{R} packages that
  can fit geostatistical GLMMs.
Notes:
$^1$Technically possible but non-trivial.
$^2$Penalized smoother GAMs that determine `wiggliness'.
$^3$\pkg{inlabru} but not \proglang{R}-\pkg{INLA}.
$^4$Spatiotemporal fields as random walk only.
$^5$SPDE approach as in \citet{miller2019}.
$^6$Does have predictive process knots.
$^7$Zero-inflated NB2 only.
$^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}.
$^9$Possible as log-linked Poisson GLMM with aggregated data.
$^{10}$Hurdle models possible by fitting components separately.
}
\end{table}

We ran a simple speed comparison between \pkg{sdmTMB},
\proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE
spatial random field model to 1000 data points with Gaussian error across
a range of mesh resolutions (Figure \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however,
\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh
resolution than \pkg{sdmTMB}. \pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \proglang{R} algebra libraries;
all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (\pkg{openBLAS}:
\citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in
a \textasciitilde10\% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and
a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.

\begin{figure}[htb]
\centering
\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
\caption[]{Comparison of time to fit an SPDE spatial random field model with
  1000 observations, an intercept and one predictor, Gaussian error, and
  a sequence of SPDE resolutions. Lines represent means and ribbons 95\%
  quantiles across 50 random iterations. Note the log-distributed x axis.
  \pkg{VAST} should be similar to \pkg{sdmTMB} and so is not shown.
  \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian
  approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()}
  formulation. \pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and
  discretized covariates \citep{miller2019}. Note that \pkg{spaMM} only fits
  spatial, not spatiotemporal, models. All platforms were restricted to one core
and could be faster with parallel computation or optimized algebra
libraries.}\label{fig:timing}
\end{figure}

\section{Discussion}\label{discussion}

How does one choose among the related packages mentioned in this paper to fit
SPDE-based geostatistical GLMMs? Assuming a given package can fit the model of
interest (Table~\ref{tab:functionality}), we suggest the primary differences are
the user interface (Appendix \ref{app:inla}) and speed. We think users familiar
with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find
\pkg{sdmTMB} most approachable. Users familiar with \proglang{R}-\pkg{INLA} will
find \pkg{inlabru} approachable. Users familiar with \pkg{mgcv} can adapt
\pkg{mgcv} to fit similar models with custom code \citep{miller2019} and
\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose
modelling packages. \pkg{VAST} is the sole option for fitting some multivariate
models; alternatively, because \pkg{VAST} focuses on multivariate delta models
and fisheries applications, users fitting ``simple'' univariate
spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB}
more straightforward. Users looking for calculation, with uncertainty, of
derived variables such as area-weighted population indexes, may favour
\pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived
with other packages).

With respect to computational speed, \pkg{sdmTMB} (and due to having the same
\pkg{TMB} backend, \pkg{VAST}) were fastest with a spatial model up to at least
\textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165
vertices) to 30-times (at 125 vertices) speed increase over
\proglang{R}-\pkg{INLA}/\pkg{inlabru}. These speed increases can allow for more
rapid and thorough model exploration and experimentation with a class of
computationally intensive models. However, for users ultimately interested in
Bayesian inference, the approximate Bayesian inference offered by
\proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than
passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full
Bayesian inference.

Additional functionality in \pkg{sdmTMB} not already mentioned includes
interpolating across missing time slices and forecasting, the barrier SPDE model
\citep{bakka2019}, time-varying spatiotemporal covariance parameters
\citep{ward2022}, and simulation from the parameter joint precision matrix.
Future development may include additional zero-inflated models, improvements to
Bayesian sampling efficiency \citep[e.g.,][]{monnahan2021}, continuous-time
models \citep[e.g.,][]{blangiardo2015book}, multivariate responses, and
non-Gaussian random fields \citep[e.g.,][]{anderson2019}. The included \pkg{TMB}
\texttt{.cpp} file provides a tested model template that can be modified to add
additional features.

\section{Acknowledgements}\label{acknowledgements}

\pkg{sdmTMB} would not be possible without the \pkg{TMB} \citep{kristensen2016}
and \proglang{R}-\pkg{INLA} \citep{rue2009, lindgren2011, lindgren2015}
\proglang{R} packages. \pkg{sdmTMB} is heavily inspired by and in some places
code has been adapted from both the \pkg{VAST} \citep{thorson2019} and
\pkg{glmmTMB} \citep{brooks2017} \proglang{R} packages (as described in the
DESCRIPTION and inst/COPYRIGHTS files). Smoother support was possible thanks to
\pkg{mgcv} \citep{wood2017a}. We thank the authors of all these packages. We
thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T.
Thorson, and J. Watson for helpful comments that substantially improved the
manuscript. Christmas Bird Count data were provided by National Audubon Society
and through the generous efforts of Bird Studies Canada and countless volunteers
across the Western Hemisphere.

\clearpage

% \appendix

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\bibliography{refs.bib}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\end{document}

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Bring over Eric's last edits on the Rmd

@@
 
-The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both
-modular and familiar to users of widely used \proglang{R} packages
-(e.g.~\pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or
-without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()}
-respectively). Estimation is done via maximum marginal likelihood with the
-\texttt{sdmTMB()} function, and the resulting fitted model can be used to make
-predictions (\texttt{predict.sdmTMB()}) or calculate residuals
-(\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and
-simulate Bayesian posterior distributions. Common quantities of interest may
-include summarizing shifts in the center of gravity (\texttt{get\_cog()})
-\citep[e.g.,][]{thorson2016cog} or in the field of ecology, tracking total
-population densities through time (\texttt{get\_index()}).
+The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both modular and familiar to users of widely used \proglang{R} packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
+Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively).
+Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
+The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+% TODO: describe this in more detail and how it could apply beyond ecology; maybe a derived variables subsection?
+Expanded this sentence a bit -- COGs are pretty common in other fields, but I don't know about indices -- removed `ecology` from the description
 
@@
 Each alternative family uses a delta or hurdle distribution: delta-gamma and
-delta-lognormal (REF), and delta-Poisson-link-gamma and
-delta-Poisson-link-lognormal (REF). Note that we choose to omit the
+delta-lognormal \citep{aitchison1955}, and delta-Poisson-link-gamma and
+delta-Poisson-link-lognormal \citep{thorson2018poisson}. The Poisson-link models
+depart from the conventional delta models in treating the probability of
+occurrence at a given location ($p_i$) as a Poisson process rather than
+a Bernoulli one, $p_i = 1 - exp(a_i * n_i)$, where $n_i$ is the total number of
+individuals available to be encountered, and $a_i$ is a measure of sampling
+effort \citep{thorson2018poisson}. Note that we choose to omit the
 spatiotemporal random fields from the binomial model component
@@
 
-where TODO.
+where $\rho$ represents the estimated autoregressive parameter allowing the
+spatial field at time $t$ to be correlated with the spatial field at time $t-1$.
 
@@
 
-<<dog-wvci-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density.
+<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density.
 Throughout, two example years are shown.
 \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects.
-The spatial random field is constant through time and represents TODO.
+The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g.
+habitat).
 The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process.
-They represent TODO.", fig.width=9, out.width="6in">>=
+They represent temporal variability in the spatial patterning of Pacific Dogfish (resulting from movement, local changes in population densities, etc).", fig.width=9, out.width="6in">>=
 cowplot::plot_grid(
@@
 
-where TODO. TODO: get NB2 above and similar in previous examples
+where three spatial fields are now estimated: the spatial intercept ($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and spatially varying NAO coefficients ($\zeta_{\bm{s}}$).
 
@@
 
+There are several planned additions to the \pkg{sdmTMB} model structure.
+A subset of features to be added in the near future includes (1) multivariate
+models to allow for simultaneous modelling of data from multiple species, (2)
+integration with the \pkg{RTMB} package so that the model code base will be
+written in \proglang{R} rather than \proglang{C++}, (3) inclusion of
+a dispersion formula similar to \pkg{glmmTMB}, and (4) estimation of random
+slopes.
+% TODO random slopes looks gnarly to do generically?
+
 \section{Acknowledgements}\label{acknowledgements}

### 2023-09-15 doc/paper-jss/paper-jss.Rmd
Update bib and old Rmd

@@
 
-```{r main-setup, include=FALSE}
+<<main-setup, include=FALSE}
 knitr::opts_chunk$set(
@@
 
-```{r compare-table, results='asis', eval=FALSE}
+<<compare-table, results='asis', eval=FALSE>>=
 library(kableExtra)
@@
   )
-```
+@
 
@@
 
-```{r sdmTMB-install, eval=FALSE, echo=TRUE}
+<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
 install.packages("sdmTMB", dependencies = TRUE)
-```
+@
 
@@
 
-```{r remotes, eval=FALSE, echo=TRUE}
+<<remotes, eval=FALSE, echo=TRUE>>=
 # install.packages("remotes")
 remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
-```
+@
 
@@
 
-```{r libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
+<<libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE>>=
 library(sdmTMB)
@@
 library(ggplot2)
-```
-
-```{r setoptions, echo=FALSE}
+@
+
+<<setoptions, echo=FALSE>>=
 options(
@@
 options(width = 80)
-```
-
-```{r libs-extras}
+@
+
+<<libs-extras>>=
 theme_set(theme_light())
 options(ggplot2.continuous.fill = "viridis")
-```
-
-```{r pcod-head, echo=TRUE}
+@
+
+<<pcod-head, echo=TRUE>>=
 select(pcod, lat, lon, X, Y, depth, present) |> 
   head()
-```
+@
 
@@
 
-```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
+<<pcod-utms-eval, echo=TRUE, eval=FALSE>>=
 pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
-```
+@
 
@@
 
-```{r dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
+<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
 The locations where lines intersect are referred to as ``vertices''.
@@
 This is a simple mesh created with the convenience function \\code{make\\_mesh()}.
-A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5}
+A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
 mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
 plot(mesh_pcod)
-```
+@
 
@@
 
-```{r, echo=TRUE}
+<<pcod-fit, echo=TRUE>>=
 fit_bin_rf <- sdmTMB(
@@
 fit_bin <- update(fit_bin_rf, spatial = "off")
-```
+@
 
@@
 
-```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
+<<pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE>>=
 sanity(fit_bin_rf)
-```
+@
 
@@
 
-```{r pcod-bin-summary, echo=TRUE}
+<<pcod-bin-summary, echo=TRUE>>=
 summary(fit_bin_rf)
-```
+@
 
@@
 
-```{r pcod-aic, eval=TRUE, echo=TRUE, results='markup'}
+<<pcod-aic, eval=TRUE, echo=TRUE, results='markup'>>=
 AIC(fit_bin_rf, fit_bin)
-```
+@
 
@@
 
-```{r pcod-cv, eval=TRUE, echo=TRUE}
+<<pcod-cv, eval=TRUE, echo=TRUE>>=
 cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
@@
 )
-```
+@
 
@@
 
-```{r pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'}
+<<pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'>>=
 cv_bin_rf$sum_loglik
 cv_bin$sum_loglik
-```
+@
 
@@
 
-```{r pcod-tidy, eval=TRUE, echo=TRUE, results='markup'}
+<<pcod-tidy, eval=TRUE, echo=TRUE, results='markup'>>=
 tidy(fit_bin_rf, conf.int = TRUE)
 tidy(fit_bin, conf.int = TRUE)
-```
+@
 
@@
 
-```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
+<<pcod-eg1-tidy-re, eval=TRUE, echo=TRUE>>=
 tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
-```
+@
 
@@
 
-```{r pcod-predict, echo=TRUE, results='markup'}
+<<pcod-predict, echo=TRUE, results='markup'>>=
 p <- predict(fit_bin_rf, newdata = qcs_grid)
@@
   as_tibble() |> head(n = 2)
-```
-
-```{r pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod.
+@
+
+<<pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod.
 Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link space, and (c) the overall prediction, which here is the combination of panels a and b.
-Note the difference between only the fixed effects (a) and including the spatial random field (c)."}
+Note the difference between only the fixed effects (a) and including the spatial random field (c).">>=
 plot_spatial_map <- function(dat, column, title) {
@@
 cowplot::plot_grid(g1, g2, g3, ncol = 3)
-```
+@
 
@@
 
-```{r dog-head-dat, echo=TRUE}
+<<dog-head-dat, echo=TRUE>>=
 dat <- select(dogfish, lon = longitude, lat = latitude, year, 
@@
 head(dat)
-```
+@
 
@@
 
-```{r dog-utms, echo=TRUE}
+<<dog-utms, echo=TRUE>>=
 dat <- add_utm_columns(dat, c("lon", "lat"), 
@@
 mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
-```
-
-```{r dog-mesh2, eval=FALSE}
+@
+
+<<dog-mesh2, eval=FALSE>>=
 plot(mesh)
-```
+@
 
@@
 
-```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
 fit_tw <- sdmTMB(
@@
 
-```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
+<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
 fit_dg <- update(fit_tw, family = delta_gamma(), 
@@
 fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
-```
+@
 
@@
 
-```{r dog-aic, echo=TRUE}
+<<dog-aic, echo=TRUE>>=
 AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
@@
   arrange(delta_AIC)
-```
+@
 
@@
 
-```{r dog-ar1, results='hide', echo=TRUE}
+<<dog-ar1, results='hide', echo=TRUE>>=
 fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
 fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
-```
-
-```{r dog-aci2, echo=TRUE}
+@
+
+<<dog-aci2, echo=TRUE>>=
 AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
-```
+@
 
@@
 
-```{r dog-ar1-2, echo=TRUE, results="hide"}
+<<dog-ar1-2, echo=TRUE, results="hide">>=
 fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
-```
-
-```{r dog-aic3, echo=TRUE}
+@
+
+<<dog-aic3, echo=TRUE>>=
 AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
-```
+@
 
@@
 
-```{r dog-print, echo=TRUE}
+<<dog-print, echo=TRUE>>=
 fit <- fit_dpl_ar1_only
@@
 summary(fit)
-```
+@
 
@@
 
-```{r dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
+<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
 Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled.
-The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in"}
+The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in">>=
 plot_anisotropy(fit)
-```
+@
 
@@
 <!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
-```{r dog-grid, echo=TRUE}
+<<dog-grid, echo=TRUE>>=
 grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
@@
 head(grid, n = 2)
-```
-
-```{r dog-pred1, echo=TRUE}
+@
+
+<<dog-pred1, echo=TRUE>>=
 pred <- predict(fit, newdata = grid, type = "response")
-```
-
-```{r dog-pred2, echo=TRUE}
+@
+
+<<dog-pred2, echo=TRUE>>=
 names(pred)
-```
+@
 
@@
 
-```{r plot-map, echo=FALSE}
+<<plot-map, echo=FALSE>>=
 plot_map <- function(dat, column) {
@@
 }
-```
-
-```{r dog-plot1, echo=FALSE}
+@
+
+<<dog-plot1, echo=FALSE>>=
 g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
@@
   ggtitle("(a) Non-random-field components; first delta model")
-```
-
-```{r dog-plot2, echo=FALSE}
+@
+
+<<dog-plot2, echo=FALSE>>=
 g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
@@
   ggtitle("(b) Spatial random field; first delta model")
-```
-
-```{r dog-plot3, echo=FALSE}
+@
+
+<<dog-plot3, echo=FALSE>>=
 g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
@@
   ggtitle("(c) Spatiotemporal random field; second delta model")
-```
-
-```{r dog-plot4, echo=FALSE}
+@
+
+<<dog-plot4, echo=FALSE>>=
 g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
@@
   ggtitle("(d) Overall prediction")
-```
+@
 
@@
 
-```{r dog-depth, echo=TRUE, out.width="5in"}
+<<dog-depth, echo=TRUE, out.width="5in">>=
 nd <- data.frame(
@@
 )
-```
-
-```{r dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
+@
+
+<<dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
 ggplot(pred_depth, aes(
@@
   labs(x = "Depth (m)", y = "Density")
-```
+@
 
@@
 
-```{r dog-index, echo=TRUE}
+<<dog-index, echo=TRUE>>=
 pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
 ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
-```
-
-```{r dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
+@
+
+<<dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
 ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
@@
   labs(y = "Biomass", x = "Year")
-```
+@
 
@@
 
-```{r owl-data}
+<<owl-data>>=
 snow <- readRDS(here::here("data/SNOW_data.rds"))
-```
-
-```{r owl-fit, echo=TRUE, results = "hide"}
+@
+
+<<owl-fit, echo=TRUE, results = "hide">>=
 mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
@@
 )
-```
-
-```{r owl-sanity, eval=FALSE}
+@
+
+<<owl-sanity, eval=FALSE>>=
 sanity(fit_owl)
-```
-
-```{r owl-print, echo=TRUE}
+@
+
+<<owl-print, echo=TRUE>>=
 summary(fit_owl)
-```
+@
 
@@
 
-```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
+<<p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
 zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
@@
 snow$nao_effect <- apply(exp(combined), 2, median)
-```
+@
 
@@
 
-```{r owl-plot-basic, echo=TRUE, eval=FALSE}
+<<owl-plot-basic, echo=TRUE, eval=FALSE>>=
 ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
-```
+@
 
@@
 
-```{r shapes, echo=FALSE}
+<<shapes, echo=FALSE>>=
 if (!file.exists(here::here("data/ne_10m_lakes"))) {
@@
 }
-```
-
-```{r shapes-read, echo=FALSE}
+@
+
+<<shapes-read, echo=FALSE>>=
 Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
@@
 lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
-```
-
-```{r owl-proj2}
+@
+
+<<owl-proj2>>=
 # project to get nice axis limits 
@@
   sf::st_as_sf(coords = c("x", "y"), crs = Albers)
-```
-
-```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
+@
+
+<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
 Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
 The effect is multiplicative on owl count per NAO unit.
-In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5}
+In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5>>=
 nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
@@
     theme(legend.position = "bottom", axis.title = element_blank())
-```
+@
 
@@
 
-```{r child="pcod.Rmd", eval=FALSE, cache=TRUE}
-```
-
-```{r child="owls.Rmd", eval=FALSE, cache=TRUE}
-```
-
-```{r child="inla-comparison.Rmd", eval=FALSE, cache=TRUE}
-```
-
-```{r child="timing.Rmd", eval=FALSE, cache=TRUE}
-```
+<<child="pcod.Rmd", eval=FALSE, cache=TRUE>>=
+@
+
+<<child="owls.Rmd", eval=FALSE, cache=TRUE>>=
+@
+
+<<child="inla-comparison.Rmd", eval=FALSE, cache=TRUE>>=
+@
+
+<<child="timing.Rmd", eval=FALSE, cache=TRUE>>=
+@

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 it to the \texttt{mesh} argument in \texttt{make\_mesh()}. We recommend
-constructing such a mesh for applied analyses to take advantage of
-additional functionality such as coarser mesh triangles in a border around the
-data and control over maximum triangle length and minimum triangle angles. We
-can inspect mesh object with the associated plotting method. Our mesh has
-238 (\texttt{mesh\_pcod\$mesh\$n}) vertices.
+constructing such a mesh for applied analyses to take advantage of additional
+functionality such as coarser mesh triangles in a border around the data and
+control over maximum triangle length and minimum triangle angles. We can inspect
+mesh object with the associated plotting method
+(Figure~\ref{fig:dog-binomial-mesh}). Our mesh has 238
+(\texttt{mesh\_pcod\$mesh\$n}) vertices.
 
@@
 This is a simple mesh created with the convenience function \\code{make\\_mesh()}.
-A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
+A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
 mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
@@
 \texttt{spatial = "off"}. We will use the \texttt{update()} S3 method to refit
-the model updating any specified arguments:
+the model while updating any specified arguments:
 
@@
 
-Here, this does not flag any issues. \texttt{sanity()} is checking that the
+This does not flag any issues. \texttt{sanity()} is checking that the
 \texttt{nlminb()} optimizer reported successful convergence, that the Hessian
@@
 We can also see that the AIC of the model with spatial random fields is
-considerably lower:
+lower:
 
@@
 own folds and conduct spatially blocked cross validation. If we set a parallel
-plan with the \pkg{future} (REF TODO) package, our folds will be fit in
-parallel.
+plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will
+be fit in parallel.
 
@@
 
-\subsection{Visualizing model predictions}
+\subsection{Making predictions on new data}
 

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Editing

@@
 random fields \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\),
-and spatially varying effects \(\zeta_{\bm{s}}\) (Figure \ref{fig:diagram}).
+and spatially varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}).
 These components are discussed in more detail via our examples below.
@@
 however, additional options allow them to be modelled as a random walk or
-first-order autoregressive, AR(1), process (Figure \ref{fig:diagram}). Turning
+first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}). Turning
 off both spatial and spatiotemporal effects allows for comparison with standard
@@
 <<pcod-head, echo=TRUE>>=
-select(pcod, lat, lon, X, Y, depth, present) |>
-  head()
+select(pcod, lat, lon, X, Y, depth, present)
 @
@@
 
+\subsection{Adding UTM columns and creating a mesh}
+
 The dataset includes spatial coordinates, year, dogfish catch weight in kg, area
@@
 dat <- select(dogfish, lon = longitude, lat = latitude, year, 
-  catch_weight, area_swept, depth) |> as_tibble()
-head(dat)
+  catch_weight, area_swept, depth)
+dat
 @
@@
 @
+
+\subsection{Fitting the model}
 
@@
 
+\subsection{Exploring delta-model alternative families}
+
 We can explore four alternative families that may better represent the data.
@@
 delta-Poisson-link-lognormal \citep{thorson2018poisson}. The Poisson-link models
-depart from the conventional delta models in treating the probability of
-occurrence at a given location ($p_i$) as a Poisson process rather than
-a Bernoulli one, $p_i = 1 - exp(a_i * n_i)$, where $n_i$ is the total number of
-individuals available to be encountered, and $a_i$ is a measure of sampling
-effort \citep{thorson2018poisson}. Note that we choose to omit the
-spatiotemporal random fields from the binomial model component
-(\texttt{spatiotemporal = list("off", "iid")}) because initial fits indicated
-this random field was collapsing to zero indicating minimal changes in encounter
-spatial distribution from year to year.
+depart from conventional delta models in treating the probability of occurrence
+at a given location ($p_i$) as a Poisson process rather than a Bernoulli one,
+$p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals
+available to be encountered, and $a_i$ is a measure of sampling effort
+\citep{thorson2018poisson}. Note that we choose to omit the spatiotemporal
+random fields from the binomial model component (\texttt{spatiotemporal
+= list("off", "iid")}) because initial fits indicated this random field was
+collapsing to zero indicating minimal changes in encounter spatial distribution
+from year to year.
 
@@
 
-We find that the Poisson-link delta-lognormal model is favoured with AIC. We
-next test two additional model formulations: making the spatial correlation
+We find that the Poisson-link delta-lognormal model is favoured with AIC.
+
+\subsection{Adding AR(1) random fields and comparing isotropic correlation}
+
+We next test two additional model formulations: making the spatial correlation
 isotropic (the default) instead of anisotropic, and structuring the positive
@@
 
-The AR(1) fields can be represented as TODO.
+The AR(1) fields can be represented as:
 
@@
 is important here given the elongated shape of the continental shelf with
-a rapid transition to deeper water. On inspecting the model output with
+a rapid transition to deeper water. We can use \texttt{plot\_anisotropy()} to
+visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
+
+<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
+Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled.
+The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
+plot_anisotropy(fit)
+@
+
+On inspecting the model output with
 \texttt{summary()}, we notice that the spatial random field SD in the positive
@@
 added. We therefore consider a model where these are turned off in the second
-component of the delta model:
+component of the delta model and choose this model.
 
@@
 
+\subsection{Inspecting the model}
+
 We save our chosen model to the object \texttt{fit} to simplify subsequent code,
-run the \texttt{sanity()} check (suppressed for space), and inspect
+run the \texttt{sanity()} check (suppressed for brevity), and inspect
 \texttt{summary()}:
@@
 of the spatiotemporal random fields, and a spatiotemporal random field marginal
-standard deviation. We can use \texttt{plot\_anisotropy()} to visually inspect
-the anisotropy (Figure \ref{fig:dog-aniso}).
-
-<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
-Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled.
-The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in">>=
-plot_anisotropy(fit)
-@
+standard deviation.
+
+\subsection{Visualizing model predictions}
 
@@
 and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
-random effects. We can easily generate plots from this data frame (e.g., with
-\texttt{ggplot2::geom\_raster()}). We suppress that code in this paper for
+random effects. We can easily generate plots from this data frame using code
+similar to in our spatial example with Pacific Cod. We suppress that code for
 brevity.
@@
 The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process.
-They represent temporal variability in the spatial patterning of Pacific Dogfish (resulting from movement, local changes in population densities, etc).", fig.width=9, out.width="6in">>=
+They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).", fig.width=9, out.width="6in">>=
 cowplot::plot_grid(
@@
 predicting across a sequence of depths and holding other variables at reference
-values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
+values (Figure~\ref{fig:dog-depth-plot}). Here, we pick the last year, specify
 to include both delta model components (\texttt{model = NA}), omit the random
@@
 
-<<dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
+<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish density.
+The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively.
+Other fixed effects are held at constant values and the random fields are omitted from the prediction.", out.width="4in">>=
 ggplot(pred_depth, aes(
@@
 
+\subsection{Calculating an area-weighted index}
+
 We can generate an area-weighted population index (e.g., a relative or absolute
@@
 km\textsuperscript{2}) to the \texttt{area} argument and specify
-\texttt{bias\_correct = TRUE} to enable a generic epsilon-bias correction
-needed due to the non-linear transformation of the random effects (REF).
+\texttt{bias\_correct = TRUE} to enable a bias correction needed due to the
+non-linear transformation of the random effects \citep{thorson2016bias}.
 
@@
 
-<<dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
+<<dog-index-plot, echo=FALSE, fig.cap="Area-weighted index of relative biomass over time for Pacific Spiny Dogfish.
+Dots and line segments represent means and 95\\% confidence intervals.", out.width="4in">>=
 ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
@@
 index on winter abundance across the southern boundary of their winter
-distribution. We fit counts observed in North America during annual Christmas
+distribution.
+
+Our data are read in from supplementary data and contain columns for UTM spatial coordinates, year, year as a factor, and the annual NAO value:
+
+<<owl-data>>=
+snow <- readRDS(here::here("data/SNOW_data.rds"))
+@
+
+<<owl-data-head, echo=TRUE>>=
+select(snow, X, Y, year, year_f, nao)
+@
+
+\subsection{Fitting the model}
+
+We fit counts observed in North America during annual Christmas
 Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011}
@@
 \begin{aligned}
-\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
+y_{\bm{s},t} &\sim \operatorname{NB2} 
+  \left(
+  \mu_{\bm{s},t}, 
+  \phi 
+  \right),\\
 \mu_{\bm{s},t} &=
 \exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
-\alpha_g +
+\alpha_t +
 \bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
@@
 
-where three spatial fields are now estimated: the spatial intercept ($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and spatially varying NAO coefficients ($\zeta_{\bm{s}}$).
-
-<<owl-data>>=
-snow <- readRDS(here::here("data/SNOW_data.rds"))
-@
+where three types of random fields are now estimated: the spatial intercept
+($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and
+spatially varying NAO coefficients ($\zeta_{\bm{s}}$). The NB2 distribution is
+specified with a mean $\mu_{\bm{s},t}$ and size parameter $\phi$. The
+observation variance scales quadratically with the mean: $\mathrm{Var}[y] = \mu
++ \mu^2 / \phi$ \citep{hilbe2011}. The $\alpha_t$ represent IID random
+intercepts by year: $\alpha_t \sim \operatorname{Normal} \left(0,
+\sigma_\alpha^2 \right)$.
 
@@
   data = snow, 
+  mesh = mesh_snow,
   family = nbinom2(link = "log"),
@@
   spatiotemporal = "iid",
-  mesh = mesh_snow,
   reml = TRUE,
@@
 
+\subsection{Inspecting the model}
+
 <<owl-sanity, eval=FALSE>>=
@@
 @
+
+In addition to the output seen for other models, we now have a section for
+random intercepts and a standard deviation for the spatially varying coefficient
+random field. Given our model specification, all random fields are sharing
+a single Matérn range.
+
+\subsection{Extracting the spatially varying coefficient effects}
 
@@
 We can make a basic plot using the following code. A more elaborate
-version is shown in Figure \ref{fig:owl-plot-fancy}.
+version is shown in Figure~\ref{fig:owl-plot-fancy}.
 
@@
 counts, but a southeast to northwest gradient in the intensity of the effect
-(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
+(Figure~\ref{fig:owl-plot-fancy}). This result is consistent with owls closest
 to the Atlantic coast and those migrating the furthest south being the most
@@
 spatial random field model to 1000 data points with Gaussian error across
-a range of mesh resolutions (Figure \ref{fig:timing}, Appendix \ref{app:speed}).
+a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
 In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however,

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Edits + adding timing appendix

@@
 
+% TODO rename this? move this? add additional sections above!?
+
 \section{Model validation and selection}\label{model-validation-and-selection}
@@
 
-\clearpage
-
 \section{Installation}\label{installation}
@@
 to the Atlantic coast and those migrating the furthest south being the most
-affected by NAO.
+affected by the NAO.
 
@@
 
+\newpage
+
+\clearpage
+
+\appendix
+
+\section{Speed testing related packages}\label{app:speed}
+
+Here, we describe the methods underlying the speed testing in
+Figure~\ref{fig:timing}. We generated a mesh that was consistent across
+simulated data sets for a given mesh resolution
+(Figure~\ref{fig:mesh-vis-timing}). We did this by setting the \texttt{max.edge}
+argument, which controls the largest allowed triangle edge length. We tested
+values of \texttt{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2. In
+Figure~\ref{fig:timing}, we report on the x-axis the number of mesh nodes
+(``vertices'') that result from each of these meshes.
+
+We simulated 1000 spatial observations with both x and y coordinates from
+uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}). Each
+iteration generated unique coordinates, predictor data, Gaussian random field
+values, and observation error. The Gaussian random field was parameterized with
+a range of 0.5 and marginal standard deviation of 0.2. The model included an
+intercept with a value of 0.2 and a normal(0, 1) predictor with an associated
+coefficient of -0.4. The observation error was Gaussian with a standard
+deviation of 0.3.
+
+We conducted 50 simulation iterations per model and mesh resolution and show
+results for the mean, lower, and upper values (Figure~\ref{fig:timing}). The
+models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core
+CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library. The following
+package versions were used:
+
+<<pkg-versions, echo=TRUE>>=
+packageVersion("INLA")
+packageVersion("inlabru")
+packageVersion("mgcv")
+packageVersion("spaMM")
+packageVersion("sdmTMB")
+@
+
+\subsection[Illustration of generating the INLA mesh]{Illustration of generating the \pkg{INLA} mesh}
+
+<<pkgs, warning=FALSE, message=FALSE, cache=FALSE, echo=FALSE>>=
+library("INLA")
+library("inlabru")
+library("ggplot2")
+library("sdmTMB")
+library("mgcv")
+library("spaMM")
+@
+
+We will illustrate with a \texttt{max.edge} of 0.06:
+
+<<mesh-timing, echo=TRUE>>=
+max_edge <- 0.06
+loc_bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
+segm_bnd <- INLA::inla.mesh.segment(loc_bnd)
+mesh <- INLA::inla.mesh.2d(
+  boundary = segm_bnd,
+  max.edge = c(max_edge, 0.2),
+  offset = c(0.1, 0.05)
+)
+@
+
+<<get-vertices, echo=FALSE>>=
+vertices <- mesh$n
+@
+
+This mesh has \Sexpr{vertices} (\texttt{mesh\$n}) vertices.
+
+<<mesh-vis-timing, fig.width=7, fig.asp=1.1, echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
+max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
+g <- list()
+for (i in seq_along(max_edge_vec)) {
+  .mesh <- INLA::inla.mesh.2d(
+    boundary = segm_bnd,
+    max.edge = c(max_edge_vec[i], 0.2),
+    offset = c(0.1, 0.05)
+  )
+  g[[i]] <- ggplot() +
+    gg(.mesh) +
+    theme_light() +
+    coord_equal(expand = FALSE) +
+    ggtitle(paste("Mesh vertices =", .mesh$n))
+}
+cowplot::plot_grid(plotlist = g, ncol = 2)
+@
+
+\subsection{Illustration of simulating data}
+
+<<sim-timing, echo=TRUE>>=
+set.seed(123)
+n_obs <- 1000
+predictor_dat <- data.frame(
+  X = runif(n_obs), Y = runif(n_obs),
+  a1 = rnorm(n_obs)
+)
+mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)
+
+sim_dat <- sdmTMB_simulate(
+  formula = ~ 1 + a1,
+  data = predictor_dat,
+  mesh = mesh_sdmTMB,
+  family = gaussian(),
+  range = 0.5,
+  phi = 0.3,
+  sigma_O = 0.2,
+  B = c(0.2, -0.4) # B0 = intercept, B1 = a1 slope
+)
+@
+
+% Look at the simulated data:
+
+<<sim-dat-plot-timing, out.width="5in", echo=FALSE, fig.cap="Example simulated dataset with a spatial random field.", fig.pos='ht', fig.align='center'>>=
+ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
+  geom_point() +
+  scale_color_gradient2() +
+  coord_fixed() +
+  theme_light()
+@
+
+\subsection[Example sdmTMB model fit]{Example \pkg{sdmTMB} model fit}
+
+<<sdmTMBfit-timing, echo=TRUE>>=
+fit_sdmTMB <- sdmTMB(
+  observed ~ a1,
+  data = sim_dat,
+  mesh = mesh_sdmTMB,
+  family = gaussian(),
+  priors = sdmTMBpriors(
+    matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)
+  )
+)
+@
+
+\subsection[Example spaMM model fit]{Example \pkg{spaMM} model fit}
+
+<<spaMMfit-timing, echo=TRUE, cache=TRUE>>=
+spde <- INLA::inla.spde2.pcmatern(
+  mesh = mesh,
+  prior.range = c(0.05, 0.05),
+  prior.sigma = c(2, 0.05)
+)
+
+fit_spaMM <- fitme(
+  observed ~ a1 + IMRF(1 | X + Y, model = spde),
+  family = gaussian(),
+  data = sim_dat
+)
+@
+
+\subsection[Example inlabru model fit]{Example \pkg{inlabru} model fit}
+
+<<inlabrufit-timing, echo=TRUE, cache=TRUE>>=
+# convert to sp SpatialPointsDataFrame first:
+dat_sp <- sp::SpatialPointsDataFrame(
+  cbind(sim_dat$X, sim_dat$Y),
+  proj4string = sp::CRS(
+    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
++ +y_0=0 +datum=NAD83 +units=km +no_defs"
+  ), data = sim_dat
+)
+components <- observed ~ -1 + Intercept(1) + a1 +
+  spatrf(main = coordinates, model = spde)
+like <- like(observed ~ Intercept + a1 + spatrf,
+  family = "gaussian", data = dat_sp
+)
+fit_bru <- bru(
+  like,
+  components = components,
+  options = bru_options(
+    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
+    bru_max_iter = 1, num.threads = "1:1"
+  )
+)
+@
+
+\subsection[Example mgcv model fit]{Example \pkg{mgcv} model fit}
+
+<<mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE>>=
+# -------------------------------------------------------------------------
+# Code in this chunk is from:
+#
+# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
+# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
+# https://doi.org/10.1007/s13253-019-00377-z
+#
+# Re-used here under a Creative Commons Attribution 4.0 International License
+# http://creativecommons.org/licenses/by/4.0/
+smooth.construct.spde.smooth.spec <- function(object, data, knots) {
+  dim <- length(object$term)
+  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
+  if (dim == 1) {
+    x <- data[[object$term]]
+  } else {
+    x <- matrix(0, nr = length(data[[1]]), nc = 2)
+    x[, 1] <- data[[object$term[1]]]
+    x[, 2] <- data[[object$term[2]]]
+  }
+  if (is.null(object$xt)) {
+    if (dim == 1) {
+      t <- seq(min(x), max(x), len = object$bs.dim)
+      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
+    } else {
+      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
+    }
+  } else {
+    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
+    mesh <- object$xt$mesh
+  }
+  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
+  inlamats <- INLA::inla.mesh.fem(mesh)
+  object$S <- list()
+  object$S[[1]] <- as.matrix(inlamats$c1)
+  object$S[[2]] <- 2 * as.matrix(inlamats$g1)
+  object$S[[3]] <- as.matrix(inlamats$g2)
+  object$L <- matrix(c(2, 2, 2, 4, 2, 0), ncol = 2)
+  object$rank <- rep(object$bs.dim, 3)
+  object$null.space.dim <- 0
+  object$mesh <- mesh
+  object$df <- ncol(object$X)
+  class(object) <- "spde.smooth"
+  return(object)
+}
+Predict.matrix.spde.smooth <- function(object, data) {
+  dim <- length(object$term)
+  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
+  if (dim == 1) {
+    x <- data[[object$term]]
+  } else {
+    x <- matrix(0, nr = length(data[[1]]), nc = 2)
+    x[, 1] <- data[[object$term[1]]]
+    x[, 2] <- data[[object$term[2]]]
+  }
+  Xp <- INLA::inla.spde.make.A(object$mesh, x)
+  return(as.matrix(Xp))
+}
+# End of code from Miller et al. 2020.
+# -------------------------------------------------------------------------
+@
+
+First define \texttt{smooth.construct.spde.smooth.spec()} and
+\texttt{Predict.matrix.spde.smooth()} from the supplement of \citet{miller2019},
+then:
+
+<<mgcvfit-timing, echo=TRUE, eval=TRUE>>=
+class(mesh) <- "inla.mesh"
+fit_bam <- bam(
+  observed ~ a1 + s(X, Y,
+    bs = "spde", k = mesh$n,
+    xt = list(mesh = mesh)
+  ),
+  data = sim_dat,
+  family = gaussian(),
+  method = "fREML",
+  control = gam.control(scalePenalty = FALSE),
+  discrete = TRUE
+)
+@
+
+\clearpage
+
 \end{document}

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 
-\begin{landscape}
+% \begin{landscape}
 \begin{figure}[p]
 \centering
-\includegraphics[width=9in]{../../figs/diagram}
+\includegraphics[width=\textwidth]{../../figs/diagram}
 \caption[]{
@@
 \end{figure}
-\end{landscape}
+% \end{landscape}
 
@@
 
-We find that the Poisson-link delta-lognormal model is favoured with AIC.
+We find that the Poisson-link delta-lognormal model is favoured by AIC.
 
@@
 models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core
-CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library. The following
-package versions were used:
-
-<<pkg-versions, echo=TRUE>>=
+CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library.
+
+% The following package versions were used:
+
+<<pkg-versions, echo=FALSE, eval=FALSE>>=
 packageVersion("INLA")

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Edits for snow

@@
 intercepts by year: $\alpha_t \sim \operatorname{Normal} \left(0,
-\sigma_\alpha^2 \right)$.
+\sigma_\alpha^2 \right)$. We can then fit this model:
 
@@
 The spatially varying coefficient effect at any point in space is the
-combination of the main effect and spatial random effect for \texttt{nao}. We
-illustrate deriving uncertainty on their combination by simulating from the
-parameter precision matrix assuming the joint parameter covariance matrix in
-multivariate normal. We do this by specifying a non-null number of simulation
-draws to \texttt{nsim} in \texttt{predict.sdmTMB()}. By default, \texttt{nsim >
-0} will return a matrix of draws from the overall prediction. Here, we instead
-   specify that we want to return draws from the \texttt{zeta\_s} random field,
-   which is the spatially varying random field. This returns a matrix where each
-   row matches a row of \texttt{newdata} and each column is a simulation draw.
+combination of the main effect and spatially varying random effect for
+\texttt{nao}. Mean estimates of the spatially varying random effect are
+available in the output of \texttt{predict.sdmTMB()} in a column starting with
+\texttt{zeta\_s} (here, \texttt{zeta\_s\_nao}). However, we wish to combine the
+fixed and random effects with uncertainty on the combined predictions. We
+illustrate a way of accomplishing this by simulating from the parameter
+precision matrix assuming the joint parameter covariance matrix is multivariate
+normal. We do this by specifying a non-null number of simulation draws to
+\texttt{nsim} in \texttt{predict.sdmTMB()}. By default, \texttt{nsim > 0} will
+return a matrix of draws from the overall prediction. Here, we instead specify
+that we want to return draws from the \texttt{zeta\_s} ($\zeta_{\bm{s}}$) random
+field, which is the spatially varying random field. This returns a matrix where
+each row matches a row of \texttt{newdata} and each column is a simulation draw.
 We then use the function \texttt{spread\_sims()} to draw 200 simulations from
 the parameters themselves, combine the main effect draws (\texttt{sims\$nao})
-   with the random field values \texttt{zeta\_s}, and calculate the median for
-   each row of data so we can plot them.
+with the random field values \texttt{zeta\_s}, and calculate the median for each
+column of data so we can plot them.
 
@@
 The effect is multiplicative on owl count per NAO unit.
-In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5>>=
+In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5.75>>=
 nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
@@
 \section{Package comparisons}\label{package-comparisons}
-
-TODO: BEFORE OR AFTER EXAMPLES
 

### 2023-09-18 doc/paper-jss/sdmTMB-paper.Rnw
SVCs
Edits

@@
 
-<<libs-extras>>=
+<<libs-extras, cache=FALSE, echo=FALSE>>=
 theme_set(theme_light())
@@
 
-We can use the \texttt{tidy()} function (formatted to match the \pkg{broom}
-package TODO) to obtain a data frame with parameter estimates (standard methods
-such as \texttt{fixef()} and \texttt{vcov()} are also available). The standard
-errors on our fixed effects have increased with the spatial random field:
+We can use the \texttt{tidy()} function \citep[formatted to match the
+\pkg{broom} package;][]{broom} to obtain a data frame with parameter estimates
+(standard methods such as \texttt{fixef()}, \texttt{confint()}, and
+\texttt{vcov()} are also available). The standard errors on our fixed effects
+have increased with the spatial random field:
 
@@
 We can do additional model comparison with cross validation via
-\texttt{sdmTMB\_cv()}. Here we will do 10-fold cross validation with the folds
+\texttt{sdmTMB\_cv()}. We will do 10-fold cross validation with the folds
 constructed randomly. Using the \texttt{fold\_ids} argument, we could supply our
@@
 individual random field components (here, \texttt{omega\_s}---the spatial
-field).
+field). We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2}
+package (Figure~\ref{fig:pcod-predict-maps}).
 
@@
 Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link (logit) space, and (c) the overall prediction, which here is the combination of panels a and b.
+The spatial random field represents spatially correlated latent effects not accounted for by the fixed effects.
 Note the difference between predictions from depth alone (a) and predictions including a spatial random field (c).">>=
@@
   \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
-  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
+  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
 \end{aligned}
 \]
+
+where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\)
+at time $t-1$ and $t$.
+
+We can fit this model as:
 
@@
   mesh = mesh,
+  family = tweedie(),
   offset = log(dat$area_swept),
@@
   anisotropy = TRUE,
-  silent = FALSE,
-  family = tweedie()
+  silent = FALSE
 )
@@
 
-We find that the Poisson-link delta-lognormal model is favoured by AIC.
+We find that the Poisson-link delta-lognormal model is favoured by AIC. In an
+applied situation we would consider comparing models with cross validation and
+inspecting the distribution of residuals.
 
@@
 need to replicate our grid for each year we will predict on. Since this is
-a common operation, we include the convenience function \texttt{replicate\_df()}
-to replicate a data frame. We then ensure our data frame contains all the
+a common operation, we include the function \texttt{replicate\_df()} to
+replicate a data frame. We then ensure our data frame contains all the
 predictors used in the model (here \texttt{log\_depth}).
@@
 and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
-random effects. We can easily generate plots from this data frame using code
-similar to in our spatial example with Pacific Cod. We suppress that code for
+random effects. We can easily generate plots from this data frame using
+\pkg{ggplot2} code with \texttt{geom\_raster()} similar to our spatial example
+with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}). We suppress that code for
 brevity.
@@
 with draws from the joint parameter precision matrix. Spatially varying
-coefficients \ldots{} TODO REFS. In addition, we illustrate using restricted
-maximum likelihood (REML), which integrates over the fixed effects as well as
-the random effects.
+coefficient models \citep{hastie1993} are a class of models in which
+coefficients are allowed to vary spatially constrained by some smooth function.
+In addition, we illustrate using restricted maximum likelihood (REML), which
+integrates over the fixed effects as well as the random effects.
 

### 2023-09-18 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 needed for the \texttt{make\_mesh()} function below. Soon, we intend to switch
-to using the \pkg{fmesher} package to eliminate this dependency.
+to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.
 
@@
 models to allow for simultaneous modelling of data from multiple species, (2)
-integration with the \pkg{RTMB} package so that the model code base will be
-written in \proglang{R} rather than \proglang{C++}, (3) inclusion of
+integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base
+will be written in \proglang{R} rather than \proglang{C++}, (3) inclusion of
 a dispersion formula similar to \pkg{glmmTMB}, and (4) estimation of random

### 2023-09-19 doc/paper-jss/sdmTMB-paper.Rnw
Developing model section

@@
 
-\pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
-Geostatistical data refers to data observed at specific spatial coordinates
-reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
-data can be collected across discrete points in time. Areal data (data
-aggregated to polygon or grid level) may be analyzed using other spatial models,
-including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018};
-\pkg{sdmTMB} can also fit models with areal data if each polygon has an
-associated centroid. A benefit of the geostatistical approach over CAR or
-similar models is that the parameters describing spatial covariance can be more
-easily interpreted \citep{wall2004} (e.g., spatial variance and range---the
-distance at which points are effectively independent).
-
-The process component of an \pkg{sdmTMB} model can be formed by any combination
-of main (``fixed'') effects \(\bm{\beta}\) and offset \(O_{\bm{s},t}\), IID
-(independent and identically distributed) random intercepts \(\alpha_g\),
-spatial intercept random fields \(\omega_{\bm{s}}\), spatiotemporal intercept
-random fields \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\),
-and spatially varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}).
-These components are discussed in more detail via our examples below.
-
+\subsection{A spatial Gaussian random field GLMM}
+
+A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
+
+\[
+\begin{aligned}
+\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
+\mu_{\bm{s}} &=
+g^{-1} \left( \eta_{\bm{s}} \right),\\
+\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
+\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
+\end{aligned}
+\]
+
+where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
+That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
+In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial GRF $\omega_{\bm{s}}$.
+This spatial GRF represents the effect of latent spatial variables that are not otherwise accounted for in the model.
+Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
+More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the prediction in link space.
+
+The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from a multivariate normal distribution with covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that defines the rate that spatial covariance decays with
+distance.
+Various covariance functions are possible, but a popular and flexible function is the Mat\'ern (REF) (FIGURE TODO).
+The Mat\'ern function describes the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$.
+If we define $\lVert d \rVert$ as the Euclidean distance between $s_j$ and $s_k$, we can define the covariance as
+\[
+\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right),
+\]
+where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
+function, $K_\nu$ represents the modified Bessel function of the second kind,
+and $\kappa$ represents the spatial decorrelation rate. The parameter $\nu$
+controls the smoothness of the covariance function, which is fixed at $\nu = 1$, to take advantage of the SPDE \citep{lindgren2011}.
+Internally, the parameters $\kappa$ and $\tau$ are converted to correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and to a marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
+The range is the distance that correlation decays to $\approx 0.13$.
+
+The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
+\[
+\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+\left(\kappa \lVert d \rVert \bm{H} \right)^\nu K_\nu \left( \kappa \lVert d \rVert \bm{H} \right).
+\]
+Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.
+Again, the smoothness parameter $\nu$ is fixed at 1.
+
+In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally expensive; methods of working directly with the sparse precision matrix ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient.
+We introduce the SPDE approach to working with the precision matrix in the software section, but for now, note that we can instead represent our spatial random field as
+\[
+\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
+\]
+where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
+
+%
+% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
+%
+% The approach involves finding an SPDE w
+
+% Details are beyond the scope of this paper and are not necessary to use the
+% software discussed here, but the idea is that the solution to a specific SPDE is
+% a GRF with a Matérn covariance function and this `trick' enables one to
+% efficiently fit approximations to GRFs to large spatial datasets
+% \citep{lindgren2011}.
+
+% Miller et al:
+% The SPDE approach provides a method to approximate Q without the common substantial computational burden.
+The SPDE is an equation to be solved.
+Solutions to this equation are stochastic processes whose covariance structure is chosen to satisfy the relationship the SPDE specifies.
+The SPDE approach involves finding an SPDE whose solutions have the covariance structure, and implied precision matrix, that is desired for f .
+Lindgren et al.
+(2011) show how to find an approximate solution to the SPDE by representing f as a sum of basis functions multiplied by coefficients; this provides a computationally efficient way to compute Q: Lindgren et al.
+(2011) show that the coefficients of these basis functions form a Gaussian Markov random field, for which methods for fast computation of precision matrices already exist (Rue and Held 2005).
+These computations make it possible to fit models quickly using integrated nested Laplace approximations (INLA; Rue et al.
+2009).
+Rue et al.
+(2017) is a comprehensive review of INLA that defines the class of latent
+
+\subsection{Adding spatiotemporal random fields}
+
+We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\epsilon_{\bm{s},t}$
+
+\[
+\begin{aligned}
+\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
+\mu_{\bm{s},t} &=
+g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
+\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
+\boldsymbol{\epsilon_{t}} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
+\end{aligned}
+\]
+
+\subsection{Adding additional model components}
+
+In practice, the above models can become considerably more complex by, for example, structuring the spatiotemporal random fields as autoregressive, letting coefficients vary though time, letting coefficients vary through space as random fields (i.e., SVC REF), or including offsets (REF).
+All these components are additive in link space.
+Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a maximally complex model could be written as
 \[
@@
 g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
+\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
+\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
+\alpha_g +
 O_{\bm{s},t} +
-\alpha_g +
 \omega_{\bm{s}} +
-\epsilon_{\bm{s},t} +
-\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
-\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} \right),
+\epsilon_{\bm{s},t} \right).
 \end{aligned}
 \]
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$, $O$ represents an offset variable (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before.
+We demonstrate all of these model components in the examples later in this paper.
+
+\subsection{Delta models}
+
+So far we have described models with one linear predictor and one family.
+Frequently, observations are better represented with two-part delta or hurdle models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values.
+Here, we describe two types of delta models.
+We drop the space and time subscripts for simplicity and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.
+
+\subsubsection{Traditional delta models}
+
+Using $p$ to denote the probability of a non-zero observation and $r$ to represent the expected rate for positive data, we can construct linear predictors in link space
+\[
+\begin{aligned}
+\operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
+\log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
+\end{aligned}
+\]
+We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood distribution (e.g., lognormal or gamma with $\phi$ representing a generic dispersion parameter).
+We use $\mathrm{I}(.)$ to denote an indicator function, which is $1$ if $y > 0$ and 0 if $y = 0$.
+\[
+\begin{aligned}
+\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
+y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
+\end{aligned}
+\]
+
+The prediction for a new data point $y$ is then $p \cdot r$.
+
+\subsubsection{Poisson-link delta models}
+
+An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
+The primary advantage is that both linear components use a log link so the effects are additive in log space and can be combined.
+In this case, there are two linear predictors, but they represent a theoretical group number $n$ and a theoretical weight per group $w$ \citep{thorson2018poisson}.
+\[
+\begin{aligned}
+\log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
+\log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
+\end{aligned}
+\]
+Note that the first linear predictor has a log link, \emph{not} a logit link and the linear predictor predicts group number $n$, not positive observation probability $p$.
+These get transformed \citep{thorson2018poisson} via
+\[
+\begin{aligned}
+p &= 1 - \exp(- \exp( \log n)),\\
+r &= \frac{n w}{p}.
+\end{aligned}
+\]
+The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} but the group number $n$ also enters into the expected positive rate $r$.
+These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation distribution as before
+\[
+\begin{aligned}
+\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
+y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
+\end{aligned}
+\]
+
+\section{break}
+
+\pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
+Geostatistical data refers to data observed at specific spatial coordinates
+reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
+data can be collected across discrete points in time. The process component of
+an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
+effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
+identically distributed) random intercepts \(\alpha_g\), spatial intercept
+random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
+\(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
+varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
+are discussed in more detail via our examples below.
+
 This process component is combined with an observation error family (e.g.,
@@
 or estimated separately (\texttt{share\_range} argument).
+
+TODO INTEGRATE THIS SOMEWHERE: Areal data (data aggregated to polygon or grid
+level) may be analyzed using other spatial models, including conditional (CAR)
+autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit
+models with areal data if each polygon has an associated centroid. A benefit of
+the geostatistical approach over CAR or similar models is that the parameters
+describing spatial covariance can be more easily interpreted \citep{wall2004}
+(e.g., spatial variance and range---the distance at which points are effectively
+independent).
 
@@
 
+\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to
+model parameters (\texttt{?sdmTMBpriors}). These priors may be useful in cases
+where estimation is difficult because of identifiability issues or relatively
+flat likelihood surfaces, or to impart prior information or achieve
+regularization. Following other recent SPDE implementations in \pkg{TMB}
+\citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors
+\citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the
+spatial range and variance parameters.
+
+\section{Software}
+
+INLA/fmesher - TMB - R combination - describe here.
++ tmbstan
+
+Describe modular design philosopy
+
+The design goal of...
+
+flexible, modular, familiar, make SPDE accessible and fast
+
+combine functionality commonly needed together
+
 The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
@@
 
-\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to
-model parameters (\texttt{?sdmTMBpriors}). These priors may be useful in cases
-where estimation is difficult because of identifiability issues or relatively
-flat likelihood surfaces, or to impart prior information or achieve
-regularization. Following other recent SPDE implementations in \pkg{TMB}
-\citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors
-\citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the
-spatial range and variance parameters.
-
-The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both modular and familiar to users of widely used \proglang{R} packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
-Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively).
-Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
-Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
-The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-% TODO: describe this in more detail and how it could apply beyond ecology; maybe a derived variables subsection?
-Expanded this sentence a bit -- COGs are pretty common in other fields, but I don't know about indices -- removed `ecology` from the description
+The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both
+modular and familiar to users of widely used \proglang{R} packages (e.g.,
+\pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or without
+a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()}
+respectively). Estimation is done via maximum marginal likelihood with the
+\texttt{sdmTMB()} function, and the resulting fitted model can be used to make
+predictions (\texttt{predict.sdmTMB()}) or calculate residuals
+(\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and
+simulate Bayesian posterior distributions. Common quantities of interest may
+include summarizing shifts in the center of gravity (\texttt{get\_cog()})
+\citep[e.g.,][]{thorson2016cog}, or tracking total densities through time
+(\texttt{get\_index()}). The \texttt{get\_index()} function generates a weighted
+average of densities for the prediction grid, integrating across space; for
+prediction grids with equally sized cells, each cell will contribute equally to
+the total average, however cells may be of unequal size.
+% TODO: describe this in more detail and how it could apply beyond ecology;
+% maybe a derived variables subsection? Expanded this sentence a bit -- COGs are
+% pretty common in other fields, but I don't know about indices -- removed
+% `ecology` from the description
 
@@
 The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
-plot_anisotropy(fit)
+plot_anisotropy(fit_dpl)
 @

### 2023-09-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
-More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the prediction in link space.
+More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the expectation in link space.
 
@@
 Various covariance functions are possible, but a popular and flexible function is the Mat\'ern (REF) (FIGURE TODO).
-The Mat\'ern function describes the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$.
-If we define $\lVert d \rVert$ as the Euclidean distance between $s_j$ and $s_k$, we can define the covariance as
+If we define $\lVert d \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
 \[
-\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
-\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right),
+\Phi\left( \bm{s}_j, \bm{s}_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right).
 \]
-where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
+The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
 function, $K_\nu$ represents the modified Bessel function of the second kind,
-and $\kappa$ represents the spatial decorrelation rate. The parameter $\nu$
-controls the smoothness of the covariance function, which is fixed at $\nu = 1$, to take advantage of the SPDE \citep{lindgren2011}.
-Internally, the parameters $\kappa$ and $\tau$ are converted to correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and to a marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
+and $\kappa$ represents the spatial decorrelation rate.
+The parameter $\nu$
+controls the smoothness of the covariance function, but can be fixed at $\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}.
+The parameters $\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
 The range is the distance that correlation decays to $\approx 0.13$.
@@
 Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.
-Again, the smoothness parameter $\nu$ is fixed at 1.
 
@@
 
-We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\epsilon_{\bm{s},t}$
+We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
 
@@
 g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
-\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
-\boldsymbol{\epsilon_{t}} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
+\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
+\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
 \end{aligned}
@@
 
-\subsection{Adding additional model components}
-
-In practice, the above models can become considerably more complex by, for example, structuring the spatiotemporal random fields as autoregressive, letting coefficients vary though time, letting coefficients vary through space as random fields (i.e., SVC REF), or including offsets (REF).
+In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in examples below).
+The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.
+
+\subsection{Additional model components}
+
+In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or including offsets \citep[][p.~206]{mccullagh1989}.
 All these components are additive in link space.
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$, $O$ represents an offset variable (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before.
-We demonstrate all of these model components in the examples later in this paper.
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
+We demonstrate these model components in the examples later in this paper.
 
@@
 with other packages).
+
+TODO add paragraph here on uniting functionality not available in any one package?
+
+% TODO: B9: I kept thinking about potential overlaps with spatial / spatiotemporal econometrics.
+Maybe a short note on how those methods relate to spatial econometrics (e.g.
+Elhorst, 2012; Lee & Yu, 2010) may broaden the interest in the paper.
 

### 2023-09-19 doc/paper-jss/sdmTMB-paper.Rnw
Editing models

@@
 (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}). This
-paper describes the basic functionality of this \proglang{R} package and its
-underlying statistical model, illustrates its use through two case studies, and
+paper describes the underlying statistical models, explains how \pkg{sdmTMB} is designed and summarizes its functionality, illustrates its use through three case studies, and
 concludes with a comparison to related software.
@@
 
-In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or including offsets \citep[][p.~206]{mccullagh1989}.
+In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors.
 All these components are additive in link space.
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
 We demonstrate these model components in the examples later in this paper.
@@
 
-So far we have described models with one linear predictor and one family.
-Frequently, observations are better represented with two-part delta or hurdle models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values.
+So far we have described models with one linear predictor and one family (common terminology in \proglang{R} packages for the combination of an observation likelihood and link).
+Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values \citep{aitchison1955}.
 Here, we describe two types of delta models.
@@
 
-The prediction for a new data point $y$ is then $p \cdot r$.
+The expectation for a new data point $y$ is then $p \cdot r$.
 
@@
 The primary advantage is that both linear components use a log link so the effects are additive in log space and can be combined.
-In this case, there are two linear predictors, but they represent a theoretical group number $n$ and a theoretical weight per group $w$ \citep{thorson2018poisson}.
+There are still two linear predictors, but they represent a theoretical group number $n$ and a theoretical weight per group $w$ \citep{thorson2018poisson}.
 \[
@@
 \]
-Note that the first linear predictor has a log link, \emph{not} a logit link and the linear predictor predicts group number $n$, not positive observation probability $p$.
+Note that the first linear predictor has a log link, \emph{not} a logit link and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
 These get transformed \citep{thorson2018poisson} via
@@
 \]
-The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} but the group number $n$ also enters into the expected positive rate $r$.
+The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
 These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation distribution as before

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Wrapping; complex mesh; new fig 1 Needs my local sdmTMB for mesh!

@@
   fig.pos = "ht",
+  fig.align = "center",
   cache.comments = TRUE,
@@
 knitr::render_sweave()
+@
+
+<<libraries, echo=FALSE, warning=FALSE, message=FALSE>>=
+library(ggplot2)
+library(dplyr)
+library(sdmTMB)
 @
@@
 (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}). This
-paper describes the underlying statistical models, explains how \pkg{sdmTMB} is designed and summarizes its functionality, illustrates its use through three case studies, and
-concludes with a comparison to related software.
+paper describes the underlying statistical models, explains how \pkg{sdmTMB} is
+designed and summarizes its functionality, illustrates its use through three
+case studies, and concludes with a comparison to related software.
 
@@
 A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
-
 \[
@@
 \eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
-\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
+\boldsymbol{\omega} &\sim \operatorname{MVNormal}
+  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
 \end{aligned}
 \]
-
-where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
-That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
-In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial GRF $\omega_{\bm{s}}$.
-This spatial GRF represents the effect of latent spatial variables that are not otherwise accounted for in the model.
-Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
-More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the expectation in link space.
-
-The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from a multivariate normal distribution with covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that defines the rate that spatial covariance decays with
-distance.
-Various covariance functions are possible, but a popular and flexible function is the Mat\'ern (REF) (FIGURE TODO).
-If we define $\lVert d \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
+where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in
+space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse
+link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that
+linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$
+multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial
+GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial
+variables that are not otherwise accounted for in the model. Alternatively,
+$\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
+More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
+value of zero that is added to the expectation in link space.
+
+The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
+a multivariate normal distribution with covariance matrix
+$\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that
+defines the rate that spatial covariance decays with distance. Various
+covariance functions are possible, but a popular and flexible function is the
+Mat\'ern (REF) (FIGURE TODO). If we define $\lVert h \rVert$ as the Euclidean
+distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
+Mat\'ern covariance $\Phi$ as
 \[
-\Phi\left( \bm{s}_j, \bm{s}_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
-\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right).
+\Phi\left( \bm{s}_j, \bm{s}_k \right) =
+  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
+  \left( \kappa \lVert h \rVert \right).
 \]
-The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
-function, $K_\nu$ represents the modified Bessel function of the second kind,
-and $\kappa$ represents the spatial decorrelation rate.
-The parameter $\nu$
-controls the smoothness of the covariance function, but can be fixed at $\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}.
-The parameters $\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
+The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the
+Gamma function, $K_\nu$ represents the modified Bessel function of the second
+kind, and $\kappa$ represents the spatial decorrelation rate. The parameter
+$\nu$ controls the smoothness of the covariance function, but can be fixed at
+$\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}. The parameters
+$\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
+as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as
+$\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
 The range is the distance that correlation decays to $\approx 0.13$.
 
-The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
+<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="TODO">>=
+predictor_dat <- expand.grid(
+  x = seq(0, 1, length.out = 100),
+  y = seq(0, 1, length.out = 100),
+  year = seq_len(6)
+)
+sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
+s1 <- sdmTMB_simulate(
+  formula = ~1,
+  data = predictor_dat,
+  mesh = sim_mesh,
+  range = 0.2,
+  phi = 0.1,
+  sigma_O = 0.2,
+  seed = 1,
+  B = 0
+)
+sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
+  geom_raster(show.legend = FALSE) +
+  scale_fill_viridis_c(option = "C") +
+  coord_fixed(expand = FALSE) +
+  theme_light() +
+  # theme(axis.title = element_blank()) +
+  ggtitle("Range = 0.2") +
+  labs(x = "X", y = "Y")
+
+s2 <- sdmTMB_simulate(
+  formula = ~1,
+  data = predictor_dat,
+  mesh = sim_mesh,
+  range = 0.6,
+  phi = 0.1,
+  sigma_O = 0.2,
+  seed = 1,
+  B = 0
+)
+sim_g2 <- ggplot(s2, aes(x, y, fill = mu)) +
+  geom_raster(show.legend = FALSE) +
+  scale_fill_viridis_c(option = "C") +
+  coord_fixed(expand = FALSE) +
+  theme_light() +
+  # theme(axis.title = element_blank()) +
+  ggtitle("Range = 0.6") +
+  labs(x = "X", y = "Y")
+
+x <- seq(0, 1, length.out = 200)
+r <- seq(0.2, 1, 0.2)
+r <- c(0.2, 0.6)
+df <- data.frame(
+  x = rep(x, length(r)),
+  range = rep(r, each = length(x))
+)
+matern <- function(h, sigma = 1, kappa, nu = 1) {
+  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
+    ((kappa * abs(h))^nu) *
+    besselK(kappa * abs(h), nu)
+  ret
+  ret[x == 0] <- sigma^2
+  ret
+}
+blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
+df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
+sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
+  geom_line() +
+  theme_light() +
+  xlab("Distance") +
+  ylab("Correlation") +
+  labs(colour = "Range") +
+  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
+  scale_colour_manual(values = blues) +
+  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
+  scale_x_continuous(breaks = r) +
+  theme(legend.position = c(0.7, 0.7)) +
+  ggtitle("Matérn correlation function")
+
+cowplot::plot_grid(sim_g3, sim_g1, sim_g2, align = "h", ncol = 3L)
+@
+
+The assumption that correlation decays equally in all directions can be relaxed
+to allow for geometric anisotropy
 \[
-\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
-\left(\kappa \lVert d \rVert \bm{H} \right)^\nu K_\nu \left( \kappa \lVert d \rVert \bm{H} \right).
+\Phi\left( s_j, s_k \right) =
+  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
+  \left( \kappa \lVert h \rVert \bm{H} \right).
 \]
-Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.
-
-In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally expensive; methods of working directly with the sparse precision matrix ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient.
-We introduce the SPDE approach to working with the precision matrix in the software section, but for now, note that we can instead represent our spatial random field as
+Here, we add a linear transformation matrix $\bm{H}$ with two estimated
+parameters governing the major axis direction of geometric anisotropy and the
+ratio of the major and minor axes \citep{lindgren2011, thorson2015ices,
+thorson2018comp}.
+
+In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally
+expensive and methods of working directly with the sparse precision matrix
+($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
+the SPDE approach to working with the precision matrix in the software section,
+but for now, note that we can instead represent our spatial random field as
 \[
-\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
+\boldsymbol{\omega} \sim
+  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
 \]
 where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
-
 %
@@
 
-% Miller et al:
-% The SPDE approach provides a method to approximate Q without the common substantial computational burden.
-The SPDE is an equation to be solved.
-Solutions to this equation are stochastic processes whose covariance structure is chosen to satisfy the relationship the SPDE specifies.
-The SPDE approach involves finding an SPDE whose solutions have the covariance structure, and implied precision matrix, that is desired for f .
-Lindgren et al.
-(2011) show how to find an approximate solution to the SPDE by representing f as a sum of basis functions multiplied by coefficients; this provides a computationally efficient way to compute Q: Lindgren et al.
-(2011) show that the coefficients of these basis functions form a Gaussian Markov random field, for which methods for fast computation of precision matrices already exist (Rue and Held 2005).
-These computations make it possible to fit models quickly using integrated nested Laplace approximations (INLA; Rue et al.
-2009).
-Rue et al.
-(2017) is a comprehensive review of INLA that defines the class of latent
+% Miller et al: The SPDE approach provides a method to approximate Q without the
+% common substantial computational burden. The SPDE is an equation to be solved.
+% Solutions to this equation are stochastic processes whose covariance structure
+% is chosen to satisfy the relationship the SPDE specifies. The SPDE approach
+% involves finding an SPDE whose solutions have the covariance structure, and
+% implied precision matrix, that is desired for f . Lindgren et al. (2011) show
+% how to find an approximate solution to the SPDE by representing f as a sum of
+% basis functions multiplied by coefficients; this provides a computationally
+% efficient way to compute Q: Lindgren et al. (2011) show that the coefficients
+% of these basis functions form a Gaussian Markov random field, for which
+% methods for fast computation of precision matrices already exist (Rue and Held
+% 2005). These computations make it possible to fit models quickly using
+% integrated nested Laplace approximations (INLA; Rue et al. 2009). Rue et al.
+% (2017) is a comprehensive review of INLA that defines the class of latent
 
@@
 
-We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
-
+We can extend our model to accommodate spatiotemporal data by adding Gaussian
+random fields for each time step $t$, $\bm{\epsilon}_{t}$
 \[
@@
 \]
-
-In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in examples below).
-The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.
+In this case we assume the spatiotemporal random fields are independent at each
+time step, but we could alternatively assume they are structured as a random
+walk or autoregressive process (demonstrated in examples below). The
+spatiotemporal random fields represent latent variables causing spatial
+correlation that changes with each time step.
 
@@
 
-In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors.
-All these components are additive in link space.
-Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a maximally complex model could be written as
+In practice, the above models can become considerably more complex by, for
+example, letting coefficients vary though time, letting coefficients vary
+through space \citep{hastie1993} as random fields, or adding random intercepts
+by grouping factors. All these components are additive in link space. Adopting
+the notation ``main'' for main effects, ``tvc'' for time-varying coefficients,
+and ``svc'' for spatially varying coefficients, a maximally complex model could
+be written as
 \[
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
-We demonstrate these model components in the examples later in this paper.
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector
+of coefficients that are allowed to vary through time as random walks or AR(1)
+processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following
+a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim
+\operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an
+offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and
+$\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal
+intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate
+these model components in the examples later in this paper.
 
@@
 
-So far we have described models with one linear predictor and one family (common terminology in \proglang{R} packages for the combination of an observation likelihood and link).
-Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values \citep{aitchison1955}.
-Here, we describe two types of delta models.
-We drop the space and time subscripts for simplicity and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.
+So far we have described models with one linear predictor and one family (common
+terminology in \proglang{R} packages for the combination of an observation
+likelihood and link). Frequently, data are better represented with two-part
+``delta'' or ``hurdle'' models, which include linear predictors and observation
+distributions for two processes: zero vs.\ non-zero values and positive values
+\citep{aitchison1955}. Here, we describe two types of delta models. We drop the
+space and time subscripts for simplicity and use $\bm{X_1} \bm{\beta_1}$ and
+$\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors
+for the first and second model components.
 
@@
 
-Using $p$ to denote the probability of a non-zero observation and $r$ to represent the expected rate for positive data, we can construct linear predictors in link space
+Using $p$ to denote the probability of a non-zero observation and $r$ to
+represent the expected rate for positive data, we can construct linear
+predictors in link space
 \[
@@
 \]
-We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood distribution (e.g., lognormal or gamma with $\phi$ representing a generic dispersion parameter).
-We use $\mathrm{I}(.)$ to denote an indicator function, which is $1$ if $y > 0$ and 0 if $y = 0$.
+We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood
+distribution (e.g., lognormal or gamma with $\phi$ representing a generic
+dispersion parameter). We use $\mathrm{I}(.)$ to denote an indicator function,
+which is $1$ if $y > 0$ and 0 if $y = 0$.
 \[
@@
 
-An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
-The primary advantage is that both linear components use a log link so the effects are additive in log space and can be combined.
-There are still two linear predictors, but they represent a theoretical group number $n$ and a theoretical weight per group $w$ \citep{thorson2018poisson}.
+An additional delta-model is possible that has several advantages over the
+logit-log delta model \citep{thorson2018poisson}. The primary advantage is that
+both linear components use a log link so the effects are additive in log space
+and can be combined. There are still two linear predictors, but they represent
+a theoretical group number $n$ and a theoretical weight per group $w$
+\citep{thorson2018poisson}.
 \[
@@
 \]
-Note that the first linear predictor has a log link, \emph{not} a logit link and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
-These get transformed \citep{thorson2018poisson} via
+Note that the first linear predictor has a log link, \emph{not} a logit link and
+the linear predictor predicts group number $n$, \emph{not} positive observation
+probability $p$. These get transformed \citep{thorson2018poisson} via
 \[
@@
 \]
-The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
-These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation distribution as before
+The first part is the complementary-log-log inverse link
+\citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but
+the group number $n$ also enters into the expected positive rate $r$. These
+probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and
+positive observation distribution as before
 \[
@@
 \sloppy We then create a mesh object that contains triangulation and projection
-matrices needed to apply the SPDE approach. The argument \texttt{cutoff} defines
-the minimum allowed distance between mesh vertices in the units of \texttt{X}
-and \texttt{Y}. \texttt{make\_mesh()} is a wrapper function for
-\texttt{INLA::inla.mesh.create()} or \texttt{fmesher::fm\_rcdt\_2d\_inla()} and
-\texttt{cutoff} is passed through to these functions. Alternatively, we could
-have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied
-it to the \texttt{mesh} argument in \texttt{make\_mesh()}. We recommend
-constructing such a mesh for applied analyses to take advantage of additional
-functionality such as coarser mesh triangles in a border around the data and
-control over maximum triangle length and minimum triangle angles. We can inspect
-mesh object with the associated plotting method
-(Figure~\ref{fig:dog-binomial-mesh}). Our mesh has 238
-(\texttt{mesh\_pcod\$mesh\$n}) vertices.
-
-<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
+matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
+\texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher}
+package. The argument \texttt{cutoff} defines the minimum allowed distance
+between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We can
+also specify additional arguments, in this case passed to
+\texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 20 km
+and 40 km for the inner and outer mesh, and an offset border width of 5 km and
+40 km for the inner and outer meshes. Alternatively, we could have created any
+mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supply it to the
+\texttt{mesh} argument in \texttt{make\_mesh()}. We can inspect mesh object with
+the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}). Our mesh
+has 238 (\texttt{mesh\_pcod\$mesh\$n}) vertices. Guidelines for what makes
+a good mesh are described in TODO.
+
+<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
+mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
+@
+
+<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
 The locations where lines intersect are referred to as ``vertices''.
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-This is a simple mesh created with the convenience function \\code{make\\_mesh()}.
 A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
-mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
-plot(mesh_pcod)
+mesh_pcod2 <- make_mesh(
+  pcod, c("X", "Y"),
+  fmesher_func = fmesher::fm_mesh_2d_inla,
+  cutoff = 8,
+  convex = 10,
+  max.edge = c(20, 40),
+  offset = c(5, 40)
+)
+plot(mesh_pcod2)
 @
@@
   data = pcod,
-  mesh = mesh_pcod,
+  mesh = mesh_pcod2,
   spatial = "on",
@@
 dat$log_depth <- log(dat$depth)
+
 mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@@
 plot(mesh)
+mesh$mesh$n
 @

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Edit

@@
 \eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
-\boldsymbol{\omega} &\sim \operatorname{MVNormal}
-  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
 \end{aligned}
@@
 More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
-value of zero that is added to the expectation in link space.
+value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
 
 The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
-a multivariate normal distribution with covariance matrix
-$\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that
-defines the rate that spatial covariance decays with distance. Various
-covariance functions are possible, but a popular and flexible function is the
-Mat\'ern (REF) (FIGURE TODO). If we define $\lVert h \rVert$ as the Euclidean
+a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some
+function that defines the rate that spatial covariance decays with
+distance,
+\[
+\boldsymbol{\omega} \sim \operatorname{MVNormal}
+  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
+\]
+Various covariance functions are possible, but a popular and flexible one is the
+Mat\'ern \citep{matern1960} (Figure~\ref{fig:matern-range}). If we define $\lVert h \rVert$ as the Euclidean
 distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
@@
 $\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
-as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as
-$\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
-The range is the distance that correlation decays to $\approx 0.13$.
+as $\sqrt{8} / \kappa$ and (2) marginal standard deviation as
+$1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
+The range is the distance that correlation decays to $\approx 0.13$ (Figure~\ref{fig:matern-range}).
 

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
-the SPDE approach to working with the precision matrix in the software section,
+the SPDE approach to working with the precision matrix in section \ref{sec:software},
 but for now, note that we can instead represent our spatial random field as
@@
 
-\subsection{Delta models}
+\subsection{Delta models}\label{sec:delta}
 
@@
 
-\section{Software}
+\section{Software}\label{sec:software}
 
@@
 package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is
-needed for the \texttt{make\_mesh()} function below. Soon, we intend to switch
-to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.
+required to generate matrices for the SPDE.
+%Soon, we intend to switch to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.
 
@@
 illustrate the need for spatial random fields. This survey is conducted by
-Fisheries and Oceans Canada (REF) and follows a depth-stratified random sampling
+Fisheries and Oceans Canada and follows a depth-stratified random sampling
 design, resulting in a georeferenced dataset. The data frame \texttt{pcod} is
@@
 package. The argument \texttt{cutoff} defines the minimum allowed distance
-between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We can
-also specify additional arguments, in this case passed to
-\texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 20 km
-and 40 km for the inner and outer mesh, and an offset border width of 5 km and
-40 km for the inner and outer meshes. Alternatively, we could have created any
-mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supply it to the
-\texttt{mesh} argument in \texttt{make\_mesh()}. We can inspect mesh object with
-the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}). Our mesh
-has 238 (\texttt{mesh\_pcod\$mesh\$n}) vertices. Guidelines for what makes
-a good mesh are described in TODO.
+between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We could create a basic mesh just specifying this:
 
@@
 @
+
+We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
+Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
+For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
+Recommendations... The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).
+
+We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
+Our mesh has 563 (\texttt{mesh\_pcod2\$mesh\$n}) vertices.
+Mesh complexity has the single largest influence on the speed of fitting these models...
+
+% https://haakonbakkagit.github.io/btopic104.html
+% https://becarioprecario.bitbucket.io/spde-gitbook/index.html
 
@@
   cutoff = 8,
-  convex = 10,
-  max.edge = c(20, 40),
-  offset = c(5, 40)
+  max.edge = c(10, 40),
+  offset = c(10, 40)
 )
@@
 By setting \texttt{effects = "ran\_pars"}, \texttt{tidy()} will return random
-field parameters, where \texttt{sigma\_O} is \(\sigma_O\) in the above
-equations---the standard deviation of the spatial random field \(\bm{\omega}\).
+field parameters, where \texttt{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').
 
@@
 
-We can then specify our model. We include an offset for the effort variable log
-area swept such that we are effectively modelling density and our predictions
-will be for an area swept of 1 km\textsuperscript{2}.
-
-Our model can be written as TODO
+We can then specify our model.
+We include an offset for the effort variable log area swept such that we are effectively modelling density and our predictions will be for an area swept of 1 km\textsuperscript{2}.
+
+Our model can be written as
 
@@
 \]
-
-where \(\bm{\beta}\) are coefficients associated with the main effects,
-\(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\)
-represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial
-field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents
-spatiotemporal fields.
+where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents spatiotemporal fields.
 
@@
 process,
-
 \[
@@
 \]
-
 where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\)
@@
 We can explore four alternative families that may better represent the data.
-Each alternative family uses a delta or hurdle distribution: delta-gamma and
-delta-lognormal \citep{aitchison1955}, and delta-Poisson-link-gamma and
-delta-Poisson-link-lognormal \citep{thorson2018poisson}. The Poisson-link models
-depart from conventional delta models in treating the probability of occurrence
-at a given location ($p_i$) as a Poisson process rather than a Bernoulli one,
-$p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals
-available to be encountered, and $a_i$ is a measure of sampling effort
-\citep{thorson2018poisson}. Note that we choose to omit the spatiotemporal
-random fields from the binomial model component (\texttt{spatiotemporal
-= list("off", "iid")}) because initial fits indicated this random field was
-collapsing to zero indicating minimal changes in encounter spatial distribution
-from year to year.
+Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
+% The Poisson-link models depart from conventional delta models in treating the probability of occurrence at a given location ($p_i$) as a Poisson process rather than a Bernoulli one, $p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals available to be encountered, and $a_i$ is a measure of sampling effort \citep{thorson2018poisson}.
+Note that we choose to omit the spatiotemporal random fields from the binomial model component (\texttt{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
 

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Add a few ideas on discussion - combine future ideas paragraphs into 1

@@
 
+TODO add paragraph here on applications to other fields, beyond ecology and fisheries? Specifically thinkgin about:
+- spatial econometrics (e.g. Elhorst, 2012; Lee & Yu, 2010)
+- disease and medicine (e.g. Wakefield work)
+- agriculture (https://link.springer.com/article/10.1007/s00122-019-03424-y)
+
 TODO add paragraph here on uniting functionality not available in any one package?
@@
 \citep{ward2022}, and simulation from the parameter joint precision matrix.
-Future development may include additional zero-inflated models, improvements to
-Bayesian sampling efficiency \citep[e.g.,][]{monnahan2021}, continuous-time
-models \citep[e.g.,][]{blangiardo2015book}, multivariate responses, and
-non-Gaussian random fields \citep[e.g.,][]{anderson2019}. The included \pkg{TMB}
-\texttt{.cpp} file provides a tested model template that can be modified to add
-additional features.
-
-There are several planned additions to the \pkg{sdmTMB} model structure.
+
+There are several planned future additions to the \pkg{sdmTMB} model structure.
 A subset of features to be added in the near future includes (1) multivariate
@@
 will be written in \proglang{R} rather than \proglang{C++}, (3) inclusion of
-a dispersion formula similar to \pkg{glmmTMB}, and (4) estimation of random
-slopes.
-% TODO random slopes looks gnarly to do generically?
+a dispersion formula similar to \pkg{glmmTMB}, (4) continuous-time
+models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses. The included \pkg{TMB}
+\texttt{.cpp} file is flexible and provides a tested model template that can be modified to add
+additional features.
+% TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
+% removed random slopes -- another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750
 

### 2023-09-21 doc/paper-jss/sdmTMB-paper.Rnw
Updates

@@
 \newcommand{\fct}[1]{\code{#1()}}
+
+\newcommand{\sdmTMB}{\pkg{sdmTMB}}
+\newcommand{\INLA}{\proglang{R}-\pkg{INLA}}
+\newcommand{\TMB}{\pkg{TMB}}
+\newcommand{\s}{\bm{s}}
+\newcommand{\R}{\proglang{R}}
 
@@
 
-\Abstract{Geostatistical data are common across scientific fields and available
-  in increasingly large datasets. However, appropriate models to analyse these
-  data, such as generalised linear mixed effects models with Gaussian random
-  fields, are computationally intensive and challenging to implement for many
-  users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends
-  the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or
-  \pkg{mgcv} to include spatial and spatiotemporal latent fields using
-  a predictive-process SPDE-(stochastic partial differential equation) based
-  approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and
-  estimation is conducted via maximum marginal likelihood with Template Model
-  Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and
-  \pkg{rstan}. We describe the model and explore case studies that illustrate
-  \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary
-  processes (time-varying and spatially varying coefficients), and anisotropy
-  (directionally dependent spatial correlation). Additional functionality
-  includes hurdle models, break-point effects, and out-of-sample cross
-  validation. Finally, we compare the functionality, speed, and interfaces of
-  related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude
-  faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible
-interface will help open this useful class of models to a wider field of
-geostatistical analysts.}
+\Abstract{
+Geostatistical data are common across scientific fields and available in increasingly large datasets.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users.
+Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
+We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
+Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
+We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
+}
 
@@
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
 \usepackage{amsmath} \usepackage{amssymb} \usepackage{bm} \usepackage{lscape}
@@
 
-\widowpenalty10000 \clubpenalty10000
+% \widowpenalty10000
+% \clubpenalty10000
 
@@
 
-Data are often collected in space or in space repeatedly over time. While such
-data are a rich source of information \citep[e.g.,][]{legendre1989a, rossi1992,
-tilman1997spatial}, they are challenging to properly model---data closer in
-space and time are usually more similar to each other than data farther apart
-due to measured and unmeasured variables \citep{cressie1993, diggle2007,
-cressie2011}. While measured variables can be accounted for with predictors in
-a model (e.g., measuring and modelling temperature effects on species
-abundance), unmeasured variables (e.g., everything influencing species abundance
-but not explicitly modelled) can cause residual spatial correlation. Accounting
-for this residual correlation is important because doing so allows for valid
-statistical inference \citep{legendre1989a, dormann2007}, can improve
-predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself
-\citep[e.g.,][]{thorson2019d, barnett2021}.
-
-Geostatistical generalized linear mixed effects models (GLMMs) with spatially
-correlated random effects are a class of models appropriate for these data
-\citep{rue2005gmrf, diggle2007, cressie2011}. Similarly to how random intercepts
-can account for correlation among groups, spatial or spatiotemporal random
-effects can account for unmeasured variables causing observations to be
-correlated in space or space and time. A common approach to modelling these
-spatial effects is with Gaussian random fields (GRFs), where the random effects
-describing the spatial patterning are assumed to be drawn from a multivariate
-normal (MVN) distribution, constrained by some covariance function such as the
-exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
-
-Such models quickly become computationally limiting due to the need to invert
-large matrices to keep track of covariation among data
-\citep[e.g.,][]{rue2005gmrf, latimer2009}. Many solutions have been proposed,
-such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic
-partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011},
-and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}. These
-approaches reduce the scale of the covariance estimation problem while providing
-a means to evaluate the data likelihood, thereby allowing fitting via Bayesian
-\citep{gelfand2017} or maximum likelihood methods. This can greatly improve
-computational efficiency \citep[e.g.,][]{heaton2019}. The SPDE approach is
-a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009,
-lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template
-Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to
-create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
-Details are beyond the scope of this paper and are not necessary to use the
-software discussed here, but the idea is that the solution to a specific SPDE is
-a GRF with a Matérn covariance function and this `trick' enables one to
-efficiently fit approximations to GRFs to large spatial datasets
-\citep{lindgren2011}.
-
-Software packages for specifying statistical models that can include the SPDE,
-such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are
-challenging to use for many applied researchers. For example, \pkg{TMB} requires
-the user to program in a \proglang{C++} template and it can be slow to
-experiment with multiple models when writing bespoke model code. Packages such
-as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users
-quickly iterate and explore statistical models---focusing on evaluating fit and
-comparing models---but do not have built-in SPDE functionality. Packages such as
-\pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful
-user interfaces to fit spatial models that use the SPDE, but they either lack
-a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB},
-or lack some functionality. We provide a more detailed comparison of related
-software packages in Table \ref{tab:functionality} and the Discussion.
-
-Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements
-geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model
-fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices. Our aim is not to
-replace the above-mentioned statistical packages, but to provide a fast,
-flexible, and user-friendly interface that is familiar to users of \pkg{lme4},
-\pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial
-and spatiotemporal models. Many features of \pkg{sdmTMB} may be found in other
-software (Table 1) but to date no package has included all in a single package.
-One common application in the field of ecology is species distribution models
-(SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}). This
-paper describes the underlying statistical models, explains how \pkg{sdmTMB} is
-designed and summarizes its functionality, illustrates its use through three
-case studies, and concludes with a comparison to related software.
+Data are often collected in space or in space repeatedly over time.
+While such data are a rich source of information \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial}, they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
+While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
+Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.
+
+Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects are a class of models appropriate for these data \citep{rue2005gmrf, diggle2007, cressie2011}.
+Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
+
+Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data \citep[e.g.,][]{rue2005gmrf, latimer2009}.
+Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
+These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
+This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
+The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
+Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.
+
+Software packages for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
+For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
+Packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
+Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
+We provide a more detailed comparison of related software packages in Table \ref{tab:functionality} and the Discussion.
+
+Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices.
+Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
+Many features of \pkg{sdmTMB} may be found in other software (Table 1) but to date no package has included all in a single package.
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}).
+This paper describes the underlying statistical models, explains how \pkg{sdmTMB} is designed and summarizes its functionality, illustrates its use through three case studies, and concludes with a comparison to related software.
 
@@
 
-A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
+A basic GLMM with spatial Gaussian random fields can be written as
 \[
@@
 \]
-where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in
-space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse
-link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that
-linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$
-multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial
-GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial
-variables that are not otherwise accounted for in the model. Alternatively,
-$\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
-More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
-value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
-
-The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
-a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some
-function that defines the rate that spatial covariance decays with
-distance,
+where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
+That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
+In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
+This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
+Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
+More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}ab).
+
+The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some function that defines the rate at which spatial covariance decays with distance,
 \[
@@
 \]
-Various covariance functions are possible, but a popular and flexible one is the
-Mat\'ern \citep{matern1960} (Figure~\ref{fig:matern-range}). If we define $\lVert h \rVert$ as the Euclidean
-distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
-Mat\'ern covariance $\Phi$ as
+Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
+If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
 \[
 \Phi\left( \bm{s}_j, \bm{s}_k \right) =
-  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
   \left(\kappa \lVert h \rVert \right)^\nu K_\nu
@@
 \]
-The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the
-Gamma function, $K_\nu$ represents the modified Bessel function of the second
-kind, and $\kappa$ represents the spatial decorrelation rate. The parameter
-$\nu$ controls the smoothness of the covariance function, but can be fixed at
-$\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}. The parameters
-$\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
-as $\sqrt{8} / \kappa$ and (2) marginal standard deviation as
-$1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
-The range is the distance that correlation decays to $\approx 0.13$ (Figure~\ref{fig:matern-range}).
-
-<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="TODO">>=
+The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggliness''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
+The parameter $\nu$ controls the smoothness of the covariance function.
+In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
+A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
+A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\sqrt{8} / \kappa$), which has the empirically derived property that correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
+
+\[
+\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right)
+\]
+
+% Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .
+
+% ($1 / \sqrt{4 \pi \tau^2 \kappa^2} $).
+
+<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values.
+The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent).
+Panel a shows a shorter range than panel b, which results in a ``wigglier'' surface.
+Panel c shows the Mat\\'ern function for these two range values.
+The dashed horizontal line shows a correlation of 0.13.">>=
 predictor_dat <- expand.grid(
@@
   # theme(axis.title = element_blank()) +
-  ggtitle("Range = 0.2") +
+  ggtitle("(a) Range = 0.2") +
   labs(x = "X", y = "Y")
@@
   # theme(axis.title = element_blank()) +
-  ggtitle("Range = 0.6") +
+  ggtitle("(b) Range = 0.6") +
   labs(x = "X", y = "Y")
@@
   theme(legend.position = c(0.7, 0.7)) +
-  ggtitle("Matérn correlation function")
-
-cowplot::plot_grid(sim_g3, sim_g1, sim_g2, align = "h", ncol = 3L)
-@
-
-The assumption that correlation decays equally in all directions can be relaxed
-to allow for geometric anisotropy
+  ggtitle("(c) Matérn correlation function")
+
+cowplot::plot_grid(sim_g1, sim_g2, sim_g3, align = "h", ncol = 3L)
+@
+
+The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
 \[
 \Phi\left( s_j, s_k \right) =
-  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
+  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
   \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
@@
 \]
-Here, we add a linear transformation matrix $\bm{H}$ with two estimated
-parameters governing the major axis direction of geometric anisotropy and the
-ratio of the major and minor axes \citep{lindgren2011, thorson2015ices,
-thorson2018comp}.
-
-In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally
-expensive and methods of working directly with the sparse precision matrix
-($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
-the SPDE approach to working with the precision matrix in section \ref{sec:software},
-but for now, note that we can instead represent our spatial random field as
+Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.
+
+In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally expensive and methods for working directly with the sparse precision matrix ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are more efficient.
+We introduce the SPDE approach to working with the precision matrix in Section \ref{sec:software}, but for now, note that we can instead represent our spatial random field as
 \[
@@
 
-We can extend our model to accommodate spatiotemporal data by adding Gaussian
-random fields for each time step $t$, $\bm{\epsilon}_{t}$
+We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
 \[
@@
 \]
-In this case we assume the spatiotemporal random fields are independent at each
-time step, but we could alternatively assume they are structured as a random
-walk or autoregressive process (demonstrated in examples below). The
-spatiotemporal random fields represent latent variables causing spatial
-correlation that changes with each time step.
+In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in Section~\ref{sec:dogar1}).
+The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.
 
@@
 
-In practice, the above models can become considerably more complex by, for
-example, letting coefficients vary though time, letting coefficients vary
-through space \citep{hastie1993} as random fields, or adding random intercepts
-by grouping factors. All these components are additive in link space. Adopting
-the notation ``main'' for main effects, ``tvc'' for time-varying coefficients,
-and ``svc'' for spatially varying coefficients, a maximally complex model could
-be written as
+In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors.
+All these components are additive in link space.
+Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
 \[
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector
-of coefficients that are allowed to vary through time as random walks or AR(1)
-processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following
-a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim
-\operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an
-offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and
-$\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal
-intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate
-these model components in the examples later in this paper.
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
+We demonstrate these model components in the examples below.
 
@@
 
-So far we have described models with one linear predictor and one family (common
-terminology in \proglang{R} packages for the combination of an observation
-likelihood and link). Frequently, data are better represented with two-part
-``delta'' or ``hurdle'' models, which include linear predictors and observation
-distributions for two processes: zero vs.\ non-zero values and positive values
-\citep{aitchison1955}. Here, we describe two types of delta models. We drop the
-space and time subscripts for simplicity and use $\bm{X_1} \bm{\beta_1}$ and
-$\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors
-for the first and second model components.
+So far we have described models with one linear predictor and one family (common terminology in \proglang{R} packages for the combination of an observation likelihood and link).
+Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}.
+Here, we describe two types of delta models, dropping the space and time subscripts for simplicity.
+% and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.
 
@@
 
-Using $p$ to denote the probability of a non-zero observation and $r$ to
-represent the expected rate for positive data, we can construct linear
-predictors in link space
+Using $p$ to denote the probability of a non-zero observation and $r$ to represent the expected rate for positive data, we can construct two linear predictors ($\eta_1$ and $\eta_2$) in link space
 \[
 \begin{aligned}
-\operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
-\log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
+% \operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
+% \log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
+\operatorname{logit} (p) &= \eta_1,\\
+\log (r) &= \eta_2.
 \end{aligned}
 \]
-We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood
-distribution (e.g., lognormal or gamma with $\phi$ representing a generic
-dispersion parameter). We use $\mathrm{I}(.)$ to denote an indicator function,
-which is $1$ if $y > 0$ and 0 if $y = 0$.
+We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood distribution (e.g., lognormal or gamma with $\phi$ representing a generic dispersion parameter).
+We use $\mathrm{I}(.)$ to denote an indicator function, which is $1$ if $y > 0$ and 0 if $y = 0$.
 \[
@@
 
-The expectation for a new data point $y$ is then $p \cdot r$.
+The expectation for a new data point is then the probability of a non-zero multiplied by the positive rate: $p \cdot r$.
 
@@
 
-An additional delta-model is possible that has several advantages over the
-logit-log delta model \citep{thorson2018poisson}. The primary advantage is that
-both linear components use a log link so the effects are additive in log space
-and can be combined. There are still two linear predictors, but they represent
-a theoretical group number $n$ and a theoretical weight per group $w$
-\citep{thorson2018poisson}.
+An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
+The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space.
+In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
 \[
 \begin{aligned}
-\log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
-\log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
+% \log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
+% \log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
+\log (n) &= \eta_1,\\
+\log (w) &= \eta_2.
 \end{aligned}
 \]
-Note that the first linear predictor has a log link, \emph{not} a logit link and
-the linear predictor predicts group number $n$, \emph{not} positive observation
-probability $p$. These get transformed \citep{thorson2018poisson} via
+Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
+These get transformed \citep{thorson2018poisson} via
 \[
@@
 \]
-The first part is the complementary-log-log inverse link
-\citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but
-the group number $n$ also enters into the expected positive rate $r$. These
-probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and
-positive observation distribution as before
+The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
+These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
 \[
@@
 \section{Software}\label{sec:software}
+
+The design goal of \sdmTMB\ was to make...
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=4in]{flowchart}
+\caption[]{
+TODO
+}\label{fig:flowchart}
+\end{figure}
+
+\sdmTMB\ relies on several well-established \R\ packages (FIG?).
+\sdmTMB\ uses \pkg{fmesher} and \INLA\ to generate input matrices for SPDE calculations \citep[$\bm{C}$, $\bm{G}$, and $\bm{K}$ in][]{lindgren2011}.
+The SPDE calculations themselves are performed by functions within \pkg{TMB} (\texttt{R\_inla} namespace).
+Data structures required to implement penalized smoothers are constructed with \pkg{mgcv}'s \texttt{smooth2random()} function similarly to \pkg{gamm4} and \pkg{brms}.
+The statistical model itself is written as a \proglang{C++} template for \TMB.
+\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior distribution for Bayesian inference.
+
+Is modular in nature...
 
@@
 in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
-'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes
+`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
 a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
@@
 
-We find that the Poisson-link delta-lognormal model is favoured by AIC. In an
-applied situation we would consider comparing models with cross validation and
-inspecting the distribution of residuals.
-
-\subsection{Adding AR(1) random fields and comparing isotropic correlation}
+We find that the Poisson-link delta-lognormal model is favoured by AIC. In an applied situation we would inspect the distribution of the residuals and consider comparing models with cross validation.
+
+\subsection{Adding AR(1) random fields and comparing isotropic correlation}\label{sec:dogar1}
 
@@
 The AR(1) fields can be represented as:
-
 \[
 \begin{aligned}
-\bm{\delta}_{t=1} &\sim \operatorname{MVN} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
-\bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
-\bm{\epsilon_{t}} \sim \operatorname{MVN} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
+\bm{\epsilon}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
+\bm{\epsilon}_{t>1} &= \rho \bm{\epsilon}_{t-1} + \sqrt{1 - \rho^2} \bm{\delta}_{t},  \:
+\bm{\delta}_{t} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
 \end{aligned}
 \]
-
-where $\rho$ represents the estimated autoregressive parameter allowing the
-spatial field at time $t$ to be correlated with the spatial field at time $t-1$.
+where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields by year.
 
@@
 
-The output is more complex compared to our binomial spatial model. We now have
-two model components, which are shown one after the other. Starting with the
-binomial component, we have output from the smoother, which includes a linear
-component (\texttt{slog\_depth}) and the standard deviation on the smoother
-weights (\texttt{sds(log\_depth)}). The smoother summary follows the format used
-in the \pkg{brms} package \citep{burkner2017}. Next, we have the time-varying
-intercepts and information on our anisotropic spatial correlation. We then have
-the second model component (lognormal) with a similar summary structure but with
-the addition of a dispersion parameter for the lognormal, the AR(1) correlation
-of the spatiotemporal random fields, and a spatiotemporal random field marginal
-standard deviation.
+The output is more complex compared to our binomial spatial model.
+We now have two model components, which are shown one after the other.
+Starting with the binomial component, we have output from the smoother, which includes a linear component (\texttt{slog\_depth}) and the standard deviation on the smoother weights (\texttt{sds(log\_depth)}).
+The smoother summary follows the format used in the \pkg{brms} package \citep{burkner2017}.
+Next, we have the time-varying intercepts and information on our anisotropic spatial correlation.
+We then have the second model component (lognormal) with a similar summary structure but with the addition of a dispersion parameter, the AR(1) correlation of the spatiotemporal random fields, and a spatiotemporal random field marginal standard deviation.
 
@@
 TODO add paragraph here on applications to other fields, beyond ecology and fisheries? Specifically thinkgin about:
-- spatial econometrics (e.g. Elhorst, 2012; Lee & Yu, 2010)
+- spatial econometrics (e.g. Elhorst, 2012; Lee \& Yu, 2010)
 - disease and medicine (e.g. Wakefield work)

### 2023-09-22 doc/paper-jss/sdmTMB-paper.Rnw
Adding SPDE stuff etc.

@@
 \]
+
+\subsection{The Mat\'ern covariance function}
+
 Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
@@
 
-\[
-\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right)
-\]
-
 % Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .
-
-% ($1 / \sqrt{4 \pi \tau^2 \kappa^2} $).
 
@@
 
+\subsection{Geometric anisotropy}
+
 The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
@@
 
-In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally expensive and methods for working directly with the sparse precision matrix ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are more efficient.
-We introduce the SPDE approach to working with the precision matrix in Section \ref{sec:software}, but for now, note that we can instead represent our spatial random field as
+\subsection{The SPDE approach}
+
+In practice, working directly with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
+and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
+One such approach is the SPDE approach.
+A full description of the SPDE approach is beyond the scope of this paper.
+Instead, we refer to the following literature:
+\citet{lindgren2011} develops the approach.
+\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
+The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
+\citet{miller2019} contains an introduction to the SPDE and illustrates its equivalence to penalized smoothing approaches.
+\citet{lindgren2022ten} provides a recent review of the approach and its applications over the last decade.
+
+For a user of \sdmTMB, the following are the important elements to understand.
+First, the SPDE approach links Mat\'ern Gaussian random fields (GRFs) to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
+This means that GRF models can be computationally fit as GMRFs.
+By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
+Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
+Commonly, this is referred to as a ``mesh''.
+The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy of the SPDE approach (and involve a tradeoff between computational burden and accuracy) \citep{lindgren2011}.
+Third, the Mat\'ern precision matrix calculations involve estimating the decorrelation rate parameter ($\kappa$), a parameter that scales the precision matrix ($\tau$) and three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
 \[
-\boldsymbol{\omega} \sim
-  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
+\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right).
 \]
-where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
-%
+Based on an estimate of $\tau$, we can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.
+
+%GMRF
+%is an SPDE whose solution exhibits the desired covariance structure and associated precision matrix.
+%The approach outlined in \citet{lindgren2011} streamlines the computation of Q and he demonstrates that
+%Can then take advantage of methods for rapid precision matrix computation \citep{rue2005}
+
+%The SPDE approach offers an efficient means to estimate Q without the typical computational overhead.
+It revolves around solving an equation known as the SPDE.
+The solutions to this equation are stochastic processes, and their covariance structure is tailored to meet the specifications outlined by the SPDE.
+To employ the SPDE approach effectively for f, the goal is to identify an SPDE whose solutions exhibit the desired covariance structure and implied precision matrix.
+In the work of Lindgren et al.
+(2011), a practical solution to the SPDE is presented by representing f as a combination of basis functions multiplied by coefficients.
+This approach significantly streamlines the computation of Q.
+Moreover, Lindgren et al.
+(2011) demonstrate that these coefficients form a Gaussian Markov random field, benefiting from existing methods for swift precision matrix computation (Rue and Held 2005).
+These computations pave the way for rapid model fitting using integrated nested Laplace approximations (INLA; Rue et al.
+2009).
+
+%The SPDE approach links
+
+%A user should know that...
+%A GMRF has the advantage that it can be fit modelling the precision matrix, which is computationally much more efficient than working with the covariance matrix.
+
+%Because the parameters and input matrices are revelant to understanding sdmTMB internals, we note that the precision
+
+%a way to approximate Q
+%SPDE = an equation to be solved
+
+%We introduce the SPDE approach to working with the precision matrix in Section \ref{sec:software}, but for now, note that we can instead represent our spatial random field as
+%\[
+%\boldsymbol{\omega} \sim
+%  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
+%\]
+%where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
+%%
 % The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
@@
 
-% Miller et al: The SPDE approach provides a method to approximate Q without the
-% common substantial computational burden. The SPDE is an equation to be solved.
-% Solutions to this equation are stochastic processes whose covariance structure
-% is chosen to satisfy the relationship the SPDE specifies. The SPDE approach
-% involves finding an SPDE whose solutions have the covariance structure, and
-% implied precision matrix, that is desired for f . Lindgren et al. (2011) show
-% how to find an approximate solution to the SPDE by representing f as a sum of
-% basis functions multiplied by coefficients; this provides a computationally
-% efficient way to compute Q: Lindgren et al. (2011) show that the coefficients
-% of these basis functions form a Gaussian Markov random field, for which
-% methods for fast computation of precision matrices already exist (Rue and Held
-% 2005). These computations make it possible to fit models quickly using
-% integrated nested Laplace approximations (INLA; Rue et al. 2009). Rue et al.
-% (2017) is a comprehensive review of INLA that defines the class of latent
+% "piece-wise linear basis functions defined by a triangulation of the domain of interest"
+
+% Miller et al: The SPDE approach provides a method to approximate Q without the common substantial computational burden.
+The SPDE is an equation to be solved.
+Solutions to this equation are stochastic processes whose covariance structure is chosen to satisfy the relationship the SPDE specifies.
+The SPDE approach involves finding an SPDE whose solutions have the covariance structure, and implied precision matrix, that is desired for f .
+Lindgren et al.
+(2011) show how to find an approximate solution to the SPDE by representing f as a sum of basis functions multiplied by coefficients; this provides a computationally efficient way to compute Q: Lindgren et al.
+(2011) show that the coefficients of these basis functions form a Gaussian Markov random field, for which methods for fast computation of precision matrices already exist (Rue and Held 2005).
+These computations make it possible to fit models quickly using integrated nested Laplace approximations (INLA; Rue et al.
+2009).
 

### 2023-09-22 doc/paper-jss/sdmTMB-paper.Rnw
Editing for clarity

@@
 For a user of \sdmTMB, the following are the important elements to understand.
-First, the SPDE approach links Mat\'ern Gaussian random fields (GRFs) to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
+First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
 This means that GRF models can be computationally fit as GMRFs.
@@
 Commonly, this is referred to as a ``mesh''.
-The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy of the SPDE approach (and involve a tradeoff between computational burden and accuracy) \citep{lindgren2011}.
-Third, the Mat\'ern precision matrix calculations involve estimating the decorrelation rate parameter ($\kappa$), a parameter that scales the precision matrix ($\tau$) and three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
+The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
+Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an additional parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
 \[
-\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right).
+\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
 \]
-Based on an estimate of $\tau$, we can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.
+where $\kappa$ is the Mat\'ern decorrelation rate as before.
+We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.
 

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Software section work

@@
 \]
-Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.
+Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices, thorson2018comp}.
 
@@
 
-%GMRF
-%is an SPDE whose solution exhibits the desired covariance structure and associated precision matrix.
-%The approach outlined in \citet{lindgren2011} streamlines the computation of Q and he demonstrates that
-%Can then take advantage of methods for rapid precision matrix computation \citep{rue2005}
-
-%The SPDE approach offers an efficient means to estimate Q without the typical computational overhead.
-It revolves around solving an equation known as the SPDE.
-The solutions to this equation are stochastic processes, and their covariance structure is tailored to meet the specifications outlined by the SPDE.
-To employ the SPDE approach effectively for f, the goal is to identify an SPDE whose solutions exhibit the desired covariance structure and implied precision matrix.
-In the work of Lindgren et al.
-(2011), a practical solution to the SPDE is presented by representing f as a combination of basis functions multiplied by coefficients.
-This approach significantly streamlines the computation of Q.
-Moreover, Lindgren et al.
-(2011) demonstrate that these coefficients form a Gaussian Markov random field, benefiting from existing methods for swift precision matrix computation (Rue and Held 2005).
-These computations pave the way for rapid model fitting using integrated nested Laplace approximations (INLA; Rue et al.
-2009).
-
-%The SPDE approach links
-
-%A user should know that...
-%A GMRF has the advantage that it can be fit modelling the precision matrix, which is computationally much more efficient than working with the covariance matrix.
-
-%Because the parameters and input matrices are revelant to understanding sdmTMB internals, we note that the precision
-
-%a way to approximate Q
-%SPDE = an equation to be solved
-
-%We introduce the SPDE approach to working with the precision matrix in Section \ref{sec:software}, but for now, note that we can instead represent our spatial random field as
-%\[
-%\boldsymbol{\omega} \sim
-%  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
-%\]
-%where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
-%%
-% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
-%
-% The approach involves finding an SPDE w
-
-% Details are beyond the scope of this paper and are not necessary to use the
-% software discussed here, but the idea is that the solution to a specific SPDE is
-% a GRF with a Matérn covariance function and this `trick' enables one to
-% efficiently fit approximations to GRFs to large spatial datasets
-% \citep{lindgren2011}.
-
-% "piece-wise linear basis functions defined by a triangulation of the domain of interest"
-
-% Miller et al: The SPDE approach provides a method to approximate Q without the common substantial computational burden.
-The SPDE is an equation to be solved.
-Solutions to this equation are stochastic processes whose covariance structure is chosen to satisfy the relationship the SPDE specifies.
-The SPDE approach involves finding an SPDE whose solutions have the covariance structure, and implied precision matrix, that is desired for f .
-Lindgren et al.
-(2011) show how to find an approximate solution to the SPDE by representing f as a sum of basis functions multiplied by coefficients; this provides a computationally efficient way to compute Q: Lindgren et al.
-(2011) show that the coefficients of these basis functions form a Gaussian Markov random field, for which methods for fast computation of precision matrices already exist (Rue and Held 2005).
-These computations make it possible to fit models quickly using integrated nested Laplace approximations (INLA; Rue et al.
-2009).
-
 \subsection{Adding spatiotemporal random fields}
@@
 
-In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors.
+In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
 All these components are additive in link space.
@@
 We demonstrate these model components in the examples below.
+% TODO: add MVN(0, sigma)s
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=\textwidth]{../../figs/diagram}
+\caption[]{
+Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
+notation, and example code. An \pkg{sdmTMB} model can be built from any
+combination of the process components (first six rows) plus an observation
+component (last row). The examples are from an SDM context, but the model can be
+fit to any spatially referenced point data. Notation: We refer to design
+matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
+coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
+symbols represent standard deviations and covariance matrices, respectively. All
+other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
+$\bm{\omega}$ refer to a vector of main effects and spatial random field
+deviations, respectively). See the \pkg{sdmTMB}
+\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
+for a full description of the model. Note that \code{s()} denotes a smoother as
+in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
+`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
+a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
+an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
+\end{figure}
 
@@
 
-So far we have described models with one linear predictor and one family (common terminology in \proglang{R} packages for the combination of an observation likelihood and link).
+So far we have described models with one linear predictor and one family (common terminology in \R\ packages for the combination of an observation likelihood and link).
 Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}.
@@
 
-\section{break}
-
-\pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
-Geostatistical data refers to data observed at specific spatial coordinates
-reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
-data can be collected across discrete points in time. The process component of
-an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
-effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
-identically distributed) random intercepts \(\alpha_g\), spatial intercept
-random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
-\(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
-varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
-are discussed in more detail via our examples below.
-
-This process component is combined with an observation error family (e.g.,
-Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
-log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
-families can be combined into a two-part ``delta'' or ``hurdle'' model
-\citep{aitchison1955} to model the zero vs.~non-zero observations separately
-from the positive observations. See the model description vignette for an
-expanded description of the statistical model
-(\texttt{browseVignettes("sdmTMB")} or see the
-\href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
-
-The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they
-estimate interpretable parameters of a spatial covariance function: parameters
-defining the magnitude of spatial variation and the rate of correlation decay
-with distance. In contrast, semi- and non-parametric approaches do not estimate
-spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002},
-\pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv}
-\citep{wood2017a}). The random fields in \pkg{sdmTMB} are structured as MVN
-constrained by a Matérn covariance function \citep{matern1960}. The Matérn can
-accommodate a range of shapes and can be both isotropic (correlation decays the
-same in all directions) or anisotropic (correlation is directionally dependent)
-\citep{haskard2007}. The Matérn standard deviations are estimated separately for
-the various fields and the range---the distance at which spatial correlation
-decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields
-or estimated separately (\texttt{share\_range} argument).
-
-TODO INTEGRATE THIS SOMEWHERE: Areal data (data aggregated to polygon or grid
-level) may be analyzed using other spatial models, including conditional (CAR)
-autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit
-models with areal data if each polygon has an associated centroid. A benefit of
-the geostatistical approach over CAR or similar models is that the parameters
-describing spatial covariance can be more easily interpreted \citep{wall2004}
-(e.g., spatial variance and range---the distance at which points are effectively
-independent).
-
-By default, if spatiotemporal fields are included, they are assumed IID;
-however, additional options allow them to be modelled as a random walk or
-first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}). Turning
-off both spatial and spatiotemporal effects allows for comparison with standard
-non-spatial GLMs or GLMMs. We include additional flexibility in specifying the
-linear main effect matrix: covariates can be modelled as penalized smooth
-functions (generalized additive models, GAMs) using the same \texttt{s()} syntax
-as in \pkg{mgcv} \citep{wood2017a} (thereby allowing automatic selection of
-smoother `wiggliness'), or can be modelled with threshold shapes: hockey-stick
-models, \texttt{breakpt()}, \citep{barrowman2000} or logistic functions,
-\texttt{logistic()}. \pkg{sdmTMB} can use many \texttt{mgcv::s()} and
-\texttt{mgcv::t2()} smoothers including bivarate smoothers \texttt{s(x, y)},
-smoothers varying by continuous or categorical variables \texttt{s(x1, by =
-x2)}, cyclical smoothers \texttt{s(x, bs = "cc")}, and smoothers with
-specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
-
-\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to
-model parameters (\texttt{?sdmTMBpriors}). These priors may be useful in cases
-where estimation is difficult because of identifiability issues or relatively
-flat likelihood surfaces, or to impart prior information or achieve
-regularization. Following other recent SPDE implementations in \pkg{TMB}
-\citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors
-\citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the
-spatial range and variance parameters.
+% \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
+% Geostatistical data refers to data observed at specific spatial coordinates
+% reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
+% data can be collected across discrete points in time. The process component of
+% an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
+% effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
+% identically distributed) random intercepts \(\alpha_g\), spatial intercept
+% random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
+% \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
+% varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
+% are discussed in more detail via our examples below.
+
+% This process component is combined with an observation error family (e.g.,
+% Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
+% log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
+% families can be combined into a two-part ``delta'' or ``hurdle'' model
+% \citep{aitchison1955} to model the zero vs.~non-zero observations separately
+% from the positive observations. See the model description vignette for an
+% expanded description of the statistical model
+% (\texttt{browseVignettes("sdmTMB")} or see the
+% \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
+
+% By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}).
+% Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
 
@@
 
-The design goal of \sdmTMB\ was to make...
-
-\begin{figure}[p]
+The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
+The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable far beyond ecology.
+
+\sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
+\sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
+The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
+Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+The statistical model itself is written as a \proglang{C++} template for \TMB.
+\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
+
+\begin{figure}[htbp]
 \centering
-\includegraphics[width=4in]{flowchart}
+\includegraphics[width=3.8in]{flowchart}
 \caption[]{
@@
 
-\sdmTMB\ relies on several well-established \R\ packages (FIG?).
-\sdmTMB\ uses \pkg{fmesher} and \INLA\ to generate input matrices for SPDE calculations \citep[$\bm{C}$, $\bm{G}$, and $\bm{K}$ in][]{lindgren2011}.
-The SPDE calculations themselves are performed by functions within \pkg{TMB} (\texttt{R\_inla} namespace).
-Data structures required to implement penalized smoothers are constructed with \pkg{mgcv}'s \texttt{smooth2random()} function similarly to \pkg{gamm4} and \pkg{brms}.
-The statistical model itself is written as a \proglang{C++} template for \TMB.
-\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
-After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior distribution for Bayesian inference.
-
-Is modular in nature...
-
-INLA/fmesher - TMB - R combination - describe here.
-+ tmbstan
-
-Describe modular design philosopy
-
-The design goal of...
-
-flexible, modular, familiar, make SPDE accessible and fast
-
-combine functionality commonly needed together
-
-The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
-a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
-marginal log likelihood and its gradient, and the negative log likelihood is
-minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
-\proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
-maximize the log likelihood conditional on the estimated fixed effects and are
-integrated over via the Laplace approximation \citep{kristensen2016}. After
-rapid model exploration with maximum likelihood, one can optionally pass an
-\pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
-rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
-distribution for Bayesian inference.
-
-The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both
-modular and familiar to users of widely used \proglang{R} packages (e.g.,
-\pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or without
-a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()}
-respectively). Estimation is done via maximum marginal likelihood with the
-\texttt{sdmTMB()} function, and the resulting fitted model can be used to make
-predictions (\texttt{predict.sdmTMB()}) or calculate residuals
-(\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and
-simulate Bayesian posterior distributions. Common quantities of interest may
-include summarizing shifts in the center of gravity (\texttt{get\_cog()})
-\citep[e.g.,][]{thorson2016cog}, or tracking total densities through time
-(\texttt{get\_index()}). The \texttt{get\_index()} function generates a weighted
-average of densities for the prediction grid, integrating across space; for
-prediction grids with equally sized cells, each cell will contribute equally to
-the total average, however cells may be of unequal size.
+The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). 
+The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()}. 
+\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} and \INLA's triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
+Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
+Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}).
+The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
+A fitted model summary can be viewed with \texttt{print.sdmTMB()} or \texttt{summary.sdmTMB()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
+\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
+Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
+Prediction on fitted or new data is accomplished with \texttt{predict.sdmTMB()}.
+\texttt{residuals.sdmTMB()} calculates various types of residuals.
+New observations can be simulated from a fitted model with \texttt{simulate.sdmTMB()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
+\texttt{sdmTMB\_cv()} facilitates cross validation.
+Derived quantities can be calculated on the fitted model.
+% TODO: more COG refs
+Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
+% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+% Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively). 
+% Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+% Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
+% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+The \sdmTMB\ formula syntax (\texttt{formula}) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
+In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and smoothers.
+Penalized smoothers (i.e., for generalized additive models, GAMs) use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
+Supported functionality includes bivarate smoothers \texttt{s(x, y)},
+smoothers varying by continuous or categorical variables \texttt{s(x1, by = x2)},
+cyclical smoothers \texttt{s(x, bs = "cc")},
+and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
+In addition to the main formula, \texttt{sdmTMB()} accepts \emph{one-sided} formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).
+
+The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
+whether to return predictions on the link or response scale (type),
+whether to return standard errors (\texttt{se\_fit},
+can be slow if conditioned on random fields),
+whether to condition on the random fields (\texttt{re\_form}),
+whether to condition on the random intercepts (\texttt{re\_form\_iid}),
+which delta model linear predictor to use (\texttt{model}),
+whether to take draws from the joint parameter precision matrix (\texttt{nsim}),
+and whether to use MCMC samples from a \pkg{tmbstan} model fit (\texttt{mcmc\_samples}).
+
+\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}). 
+These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization. 
+Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
+These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
+If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for change of variables (priors applied to parameters that are internally transformed) (REF).
+
+% Describe modular design philosopy
+
+% The design goal of...
+
+% flexible, modular, familiar, make SPDE accessible and fast
+
+% combine functionality commonly needed together
+
+% The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
+% a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
+% marginal log likelihood and its gradient, and the negative log likelihood is
+% minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
+% \proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
+% maximize the log likelihood conditional on the estimated fixed effects and are
+% integrated over via the Laplace approximation \citep{kristensen2016}. After
+% rapid model exploration with maximum likelihood, one can optionally pass an
+% \pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
+% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
+% distribution for Bayesian inference.
+
 % TODO: describe this in more detail and how it could apply beyond ecology;
@@
 
-% \begin{landscape}
-\begin{figure}[p]
-\centering
-\includegraphics[width=\textwidth]{../../figs/diagram}
-\caption[]{
-Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
-notation, and example code. An \pkg{sdmTMB} model can be built from any
-combination of the process components (first six rows) plus an observation
-component (last row). The examples are from an SDM context, but the model can be
-fit to any spatially referenced point data. Notation: We refer to design
-matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
-coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
-symbols represent standard deviations and covariance matrices, respectively. All
-other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
-$\bm{\omega}$ refer to a vector of main effects and spatial random field
-deviations, respectively). See the \pkg{sdmTMB}
-\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
-for a full description of the model. Note that \code{s()} denotes a smoother as
-in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
-`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
-a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
-an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
-\end{figure}
-% \end{landscape}
-
 % TODO rename this? move this? add additional sections above!?
 
-\section{Model validation and selection}\label{model-validation-and-selection}
-
-Validation and selection of state-space models is challenging, particularly when
-using the Laplace approximation \citep{thygesen2017a}. We provide several
-approaches to assist this process: (1) The Akaike Information Criterion
-\citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
-has well-documented biases with mixed-effects models \citep{liang2008}. (2)
-Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
-with user-specified or randomly chosen folds for model selection or to evaluate
-goodness of fit according to user-calculated criteria (e.g., mean squared error,
-area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
-\pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
-with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
-approximation, or perform posterior predictive checks (see
-\texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
-randomized quantile \citep{dunn1996} or probability integral transform (PIT)
-\citep{smith1985} residuals. For state-space models, these residuals have known
-statistical issues with the Laplace approximation \citep{thygesen2017a} but are
-quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
-of the random effects to avoid this issue \citep{rufener2021} is recommended but
-slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
-are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
-fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
-simulating data without starting from a fitted model. Models can be fit to these
-simulated data to ensure identifiability, evaluate bias and precision in
-parameter estimation, or evaluate the consequences of model misspecification.
-
-\section{Installation}\label{installation}
+% \section{Model validation and selection}\label{model-validation-and-selection}
+
+% Validation and selection of state-space models is challenging, particularly when
+% using the Laplace approximation \citep{thygesen2017a}. We provide several
+% approaches to assist this process: (1) The Akaike Information Criterion
+% \citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
+% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
+% Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
+% with user-specified or randomly chosen folds for model selection or to evaluate
+% goodness of fit according to user-calculated criteria (e.g., mean squared error,
+% area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
+% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
+% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
+% approximation, or perform posterior predictive checks (see
+% \texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
+% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
+% \citep{smith1985} residuals. For state-space models, these residuals have known
+% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
+% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
+% of the random effects to avoid this issue \citep{rufener2021} is recommended but
+% slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
+% are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
+% fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
+% simulating data without starting from a fitted model. Models can be fit to these
+% simulated data to ensure identifiability, evaluate bias and precision in
+% parameter estimation, or evaluate the consequences of model misspecification.
+
+\subsection{Installation}\label{installation}
 
@@
 
-Additional utilities are maintained in the \pkg{sdmTMBextra} package at
-\url{https://github.com/pbs-assess/sdmTMBextra}.
+Additional utilities are maintained in the \pkg{sdmTMBextra} package at\\
+\url{https://github.com/pbs-assess/sdmTMBextra}
 
@@
 matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
-\texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher}
-package. The argument \texttt{cutoff} defines the minimum allowed distance
-between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We could create a basic mesh just specifying this:
+% \texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package. 
+The argument \texttt{cutoff} defines the minimum allowed distance
+between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). 
+We could create a basic mesh just specifying this:
 
@@
 We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
-Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
+% Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}. 
 For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
-Recommendations... The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).
+The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).
 
@@
 Our mesh has 563 (\texttt{mesh\_pcod2\$mesh\$n}) vertices.
-Mesh complexity has the single largest influence on the speed of fitting these models...
+Mesh complexity has the single largest influence on the speed of fitting these models.
 
@@
 \mu_{\bm{s}} &= \operatorname{logit}^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s}} \bm{\beta} +
-\omega_{\bm{s}}
-\right),
+\omega_{\bm{s}} \right),\\
+\boldsymbol{\omega} &\sim \operatorname{MVNormal}
+  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
 \end{aligned}
@@
 a vector of estimated main effect coefficients, and \(\omega_{\bm{s}}\)
-represents the estimated spatial field, \(\omega_{\bm{s}} \sim
-\mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)\).
+represents the estimated spatial random field.
 
@@
 
-We can use the \texttt{tidy()} function \citep[formatted to match the
-\pkg{broom} package;][]{broom} to obtain a data frame with parameter estimates
-(standard methods such as \texttt{fixef()}, \texttt{confint()}, and
-\texttt{vcov()} are also available). The standard errors on our fixed effects
-have increased with the spatial random field:
+We can use the \texttt{tidy()} function to obtain a data frame with parameter estimates (standard methods such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()} are also available). 
+The standard errors on our fixed effects have increased with the spatial random field:
 
@@
 Our model can be written as
-
 \[
@@
 \epsilon_{\bm{s},t}
-\right),
+\right),\\
+\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
+\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
 \end{aligned}
@@
 where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents spatiotemporal fields.
-
-The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1)
-process,
+The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
 \[
@@
 \]
-where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\)
-at time $t-1$ and $t$.
+where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$.
 
@@
 \]
-where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields by year.
+where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields each year.
 
@@
 We can write the model as
-
 \[
@@
 \omega_{\bm{s}} +
-\epsilon_{\bm{s},t} \right),
+\epsilon_{\bm{s},t} \right),\\
+\alpha_t &\sim \operatorname{Normal} (0, \sigma_\alpha^2 ),\\
+\bm{\zeta} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\zeta \right),\\
+\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
+\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
 \end{aligned}
 \]
-
-where three types of random fields are now estimated: the spatial intercept
-($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and
-spatially varying NAO coefficients ($\zeta_{\bm{s}}$). The NB2 distribution is
-specified with a mean $\mu_{\bm{s},t}$ and size parameter $\phi$. The
-observation variance scales quadratically with the mean: $\mathrm{Var}[y] = \mu
-+ \mu^2 / \phi$ \citep{hilbe2011}. The $\alpha_t$ represent IID random
-intercepts by year: $\alpha_t \sim \operatorname{Normal} \left(0,
-\sigma_\alpha^2 \right)$. We can then fit this model:
+where three types of random fields are now included:
+spatially varying NAO coefficients ($\zeta_{\bm{s}}$),
+a spatial intercept ($\omega_{\bm{s}}$), and
+spatiotemporal variation ($\epsilon_{\bm{s},t}$). 
+The NB2 distribution is specified with a mean $\mu_{\bm{s},t}$ and size parameter $\phi$. 
+The observation variance scales quadratically with the mean: $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ \citep{hilbe2011}. 
+The $\alpha_t$ represent IID random intercepts by year. 
+We can then fit this model:
 
@@
 
-How does one choose among the related packages mentioned in this paper to fit
-SPDE-based geostatistical GLMMs? Assuming a given package can fit the model of
-interest (Table~\ref{tab:functionality}), we suggest the primary differences are
-the user interface (Appendix \ref{app:inla}) and speed. We think users familiar
-with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find
-\pkg{sdmTMB} most approachable. Users familiar with \proglang{R}-\pkg{INLA} will
-find \pkg{inlabru} approachable. Users familiar with \pkg{mgcv} can adapt
-\pkg{mgcv} to fit similar models with custom code \citep{miller2019} and
-\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose
-modelling packages. \pkg{VAST} is the sole option for fitting some multivariate
-models; alternatively, because \pkg{VAST} focuses on multivariate delta models
-and fisheries applications, users fitting ``simple'' univariate
-spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB}
-more straightforward. Users looking for calculation, with uncertainty, of
-derived variables such as area-weighted population indexes, may favour
-\pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived
-with other packages).
+How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs? 
+Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed. 
+We think users familiar with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable. 
+Users familiar with \proglang{R}-\pkg{INLA} will find \pkg{inlabru} approachable. 
+Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
+\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward.
+Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
 
@@
 
-With respect to computational speed, \pkg{sdmTMB} (and due to having the same
-\pkg{TMB} backend, \pkg{VAST}) were fastest with a spatial model up to at least
-\textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165
-vertices) to 30-times (at 125 vertices) speed increase over
-\proglang{R}-\pkg{INLA}/\pkg{inlabru}. These speed increases can allow for more
-rapid and thorough model exploration and experimentation with a class of
-computationally intensive models. However, for users ultimately interested in
-Bayesian inference, the approximate Bayesian inference offered by
-\proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than
-passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full
-Bayesian inference.
-
-Additional functionality in \pkg{sdmTMB} not already mentioned includes
-interpolating across missing time slices and forecasting, the barrier SPDE model
-\citep{bakka2019}, time-varying spatiotemporal covariance parameters
-\citep{ward2022}, and simulation from the parameter joint precision matrix.
+% TODO: figure out where to put this paragraph
+The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
+The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1960}. 
+The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
+The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\texttt{share\_range} argument).
+
+% TODO: figure out where to put this paragraph
+Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
+A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004} (e.g., spatial variance and range---the distance at which points are effectively independent).
+
+With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \proglang{R}-\pkg{INLA}/\pkg{inlabru}.
+These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
+However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
+
+Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and simulation from the parameter joint precision matrix.
 
 There are several planned future additions to the \pkg{sdmTMB} model structure.
-A subset of features to be added in the near future includes (1) multivariate
-models to allow for simultaneous modelling of data from multiple species, (2)
-integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base
-will be written in \proglang{R} rather than \proglang{C++}, (3) inclusion of
-a dispersion formula similar to \pkg{glmmTMB}, (4) continuous-time
-models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses. The included \pkg{TMB}
-\texttt{.cpp} file is flexible and provides a tested model template that can be modified to add
-additional features.
+A subset of features to be added in the near future includes (1) multivariate models to allow for simultaneous modelling of data from multiple species, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \proglang{R} rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB}, (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
+The included \pkg{TMB} \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features. 
 % TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
+% TODO: decide on using GRF throughout
 
@@
 The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
-Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.
+% Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.
 
@@
 For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
-Packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
+Packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE or spatiotemporal functionality.
 Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
-We provide a more detailed comparison of related software packages in Table \ref{tab:functionality} and the Discussion.
-
-Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices.
+
+% We provide a more detailed comparison of related software packages in Table~\ref{tab:functionality} and the Discussion (Section~\ref{discussion}).
+
+Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
-Many features of \pkg{sdmTMB} may be found in other software (Table 1) but to date no package has included all in a single package.
-One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}).
-This paper describes the underlying statistical models, explains how \pkg{sdmTMB} is designed and summarizes its functionality, illustrates its use through three case studies, and concludes with a comparison to related software.
+Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
+
+This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion and comparison to related software (Section~\ref{discussion}).
+
+% TODO: there is now some duplication between these two paragraphs!?
 
@@
 \centering
-\includegraphics[width=3.8in]{flowchart}
+\includegraphics[width=3.4in]{flowchart}
 \caption[]{
-TODO
+  Description of the model fitting procedure in \pkg{sdmTMB}.
 }\label{fig:flowchart}
 \end{figure}
+
+% TODO: () on end of functions!? see flowchart vs. rest
+% TODO: blah.sdmTMB() for S3 methods?
 
@@
 These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
-If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for change of variables (priors applied to parameters that are internally transformed) (REF).
+If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
 
@@
 
-Importantly, \texttt{dependencies = TRUE} includes the \proglang{R}-INLA
-package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is
-required to generate matrices for the SPDE.
+Importantly, \texttt{dependencies = TRUE} includes the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is required to generate matrices for the SPDE.
 %Soon, we intend to switch to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.
@@
 
-There are many \proglang{R} packages capable of fitting geostatistical spatial
-or spatiotemporal models \citep[e.g.,][]{heaton2019}. \pkg{sdmTMB}, \pkg{VAST},
-\proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are
-the most closely related, as they all provide a user interface to SPDE-based GRF
-models. In our software comparison (Table \ref{tab:functionality}), we also
-include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and
-\pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package
-that can fit related predictive-process models without the SPDE. \pkg{sdmTMB},
-\pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas
-\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to
-isotropic covariance. \pkg{sdmTMB} and \pkg{mgcv} focus on univariate response
-data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM},
-and \pkg{spBayes} extend to multivariate responses with various limitations. To
-our knowledge, \pkg{VAST} is the only package to implement spatial
-\citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and
-spatial empirical orthogonal function (EOF) regression \citep{thorson2020}. Of
-these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold
-(e.g., hockey-stick) covariate relationships. \pkg{spaMM} is limited to
-a spatial random field (i.e., does not fit spatiotemporal fields) and
-\pkg{spBayes} implements spatiotemporal fields, but only as a random walk. There
-is considerable variability in the available observation likelihoods across
-packages (Table \ref{tab:functionality}). We provide comparisons of the syntax
-and the reproducibility of results from models fit using \pkg{sdmTMB},
-\proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.
+There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}. 
+\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
+In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
+\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance. 
+\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
+To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
+Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships. 
+\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk. 
+There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}). 
+% We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.
 
@@
 How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs? 
-Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed. 
+Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed. 
 We think users familiar with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable. 
@@
 In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
-The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1960}. 
+The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1986}. 
 The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
@@
 
-<<mesh-vis-timing, fig.width=7, fig.asp=1.1, echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
+<<mesh-vis-timing, fig.width=7, fig.asp=1.1, out.width="4in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
 max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
@@
 
-<<sim-dat-plot-timing, out.width="5in", echo=FALSE, fig.cap="Example simulated dataset with a spatial random field.", fig.pos='ht', fig.align='center'>>=
+<<sim-dat-plot-timing, out.width="4in", echo=FALSE, fig.cap="Example simulated dataset with a spatial random field.", fig.pos='ht', fig.align='center'>>=
 ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 
-% \widowpenalty10000
-% \clubpenalty10000
+% \widowpenalty1000
+% \clubpenalty1000
 
@@
 
-\begin{figure}[p]
+\begin{figure}[htbp]
 \centering
@@
 Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
-Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}).
+Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).
 The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
@@
 The effect is multiplicative on owl count per NAO unit.
-In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5.75>>=
+In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, fig.pos="htbp", out.width="4in">>=
 nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
@@
 
-\clearpage
+\newpage
 
@@
 
-\clearpage
+% \clearpage
 
@@
 
-<<mesh-vis-timing, fig.width=7, fig.asp=1.1, out.width="4in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
+<<mesh-vis-timing, fig.width=7, fig.asp=1.1, out.width="4.3in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
 max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 
-This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion and comparison to related software (Section~\ref{discussion}).
+This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of TODO  (Section~\ref{discussion}).
 
@@
 
-\pkg{sdmTMB} would not be possible without the \pkg{TMB} \citep{kristensen2016}
-and \proglang{R}-\pkg{INLA} \citep{rue2009, lindgren2011, lindgren2015}
-\proglang{R} packages. \pkg{sdmTMB} is heavily inspired by and in some places
-code has been adapted from both the \pkg{VAST} \citep{thorson2019} and
-\pkg{glmmTMB} \citep{brooks2017} \proglang{R} packages (as described in the
-DESCRIPTION and inst/COPYRIGHTS files). Smoother support was possible thanks to
-\pkg{mgcv} \citep{wood2017a}. We thank the authors of all these packages. We
-thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T.
-Thorson, and J. Watson for helpful comments that substantially improved the
-manuscript. Christmas Bird Count data were provided by National Audubon Society
-and through the generous efforts of Bird Studies Canada and countless volunteers
-across the Western Hemisphere.
+\pkg{sdmTMB} would not be possible without the \pkg{TMB} \citep{kristensen2016} and \proglang{R}-\pkg{INLA} \citep{rue2009, lindgren2011, lindgren2015} \proglang{R} packages. 
+\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} \citep{thorson2019} and \pkg{glmmTMB} \citep{brooks2017} \proglang{R} packages (as described in the \texttt{DESCRIPTION} and \texttt{inst/COPYRIGHTS} files).
+Smoother support was possible thanks to \pkg{mgcv} \citep{wood2017a}. 
+We thank the authors of all these packages. 
+We thank S.
+Kotwicki, M.
+Lindmark, M.
+Martin, C.C.
+Monnahan, P.M.
+Regular, J.T.
+Thorson, J.
+Watson, two anonymous reviewers, and the editor for helpful comments that substantially improved the manuscript.
+Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.
 

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Edit

@@
 An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
-The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space.
+The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
 In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
@@
 The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
-The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable far beyond ecology.
+The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.
 
@@
 \subsection{Inspecting the model}
+
+\texttt{summary()} prints standard model information:
 

### 2023-09-25 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
-More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}ab).
-
+More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
 The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some function that defines the rate at which spatial covariance decays with distance,
@@
 \]
-
-\subsection{The Mat\'ern covariance function}
-
-Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
-If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
-\[
-\Phi\left( \bm{s}_j, \bm{s}_k \right) =
-  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
-  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
-  \left( \kappa \lVert h \rVert \right).
-\]
-The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggliness''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
-The parameter $\nu$ controls the smoothness of the covariance function.
-In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
-A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
-A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\sqrt{8} / \kappa$), which has the empirically derived property that correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
-
-% Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .
 
@@
 @
+
+\subsection{The Mat\'ern covariance function}
+
+Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
+If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
+\[
+\Phi\left( \bm{s}_j, \bm{s}_k \right) =
+  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
+  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
+  \left( \kappa \lVert h \rVert \right).
+\]
+The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggliness''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
+The parameter $\nu$ controls the smoothness of the covariance function.
+In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
+A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
+A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\sqrt{8} / \kappa$), which has the empirically derived property that correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
+
+% Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .
 

### 2023-09-26 doc/paper-jss/sdmTMB-paper.Rnw
Add todo
Wrap by sentence; some edits

@@
 
-Data are often collected in space or in space repeatedly over time.
-While such data are a rich source of information \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial}, they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
-While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
+Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial},
+% TODO: other non-ecology refs
+they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
+While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation.
 Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.
@@
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
 % TODO: decide on using GRF throughout
 
-Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data \citep[e.g.,][]{rue2005gmrf, latimer2009}.
+Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data. 
 Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
@@
 The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
+
 % Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.
@@
 \caption[]{
-Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
-notation, and example code. An \pkg{sdmTMB} model can be built from any
-combination of the process components (first six rows) plus an observation
-component (last row). The examples are from an SDM context, but the model can be
-fit to any spatially referenced point data. Notation: We refer to design
-matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
-coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
-symbols represent standard deviations and covariance matrices, respectively. All
-other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
-$\bm{\omega}$ refer to a vector of main effects and spatial random field
-deviations, respectively). See the \pkg{sdmTMB}
-\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
-for a full description of the model. Note that \code{s()} denotes a smoother as
-in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
-`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
-a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
-an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
+Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code. 
+An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row). 
+The examples are from an SDM context, but the model can be fit to any spatially referenced point data. 
+Notation: We refer to design matrices as $\bm{X}$. 
+The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively. 
+The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively. 
+All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively). 
+See the \pkg{sdmTMB} \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model. 
+Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
 \end{figure}
@@
 After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
+% TODO: visreg + ggeffects effects, etc.?
 
@@
 
-We begin with a simple species distribution model of encounter probability of
-Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen
-Charlotte Sound, British Columbia, Canada. The purpose of our example is to
-illustrate the need for spatial random fields. This survey is conducted by
-Fisheries and Oceans Canada and follows a depth-stratified random sampling
-design, resulting in a georeferenced dataset. The data frame \texttt{pcod} is
-available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
-longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and
-encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of
-Pacific Cod for a given survey sample.
+We begin with a simple species distribution model of encounter probability of Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen Charlotte Sound, British Columbia, Canada.
+The purpose of our example is to illustrate the need for spatial random fields. 
+This survey is conducted by Fisheries and Oceans Canada and follows a depth-stratified random sampling design, resulting in a georeferenced dataset. 
+The data frame \texttt{pcod} is available as package data in \pkg{sdmTMB}. 
+Relevant columns include latitude, longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of Pacific Cod for a given survey sample.
 
@@
 
-An \pkg{sdmTMB} model requires a data frame that contains a response column,
-columns for any predictors, and columns for spatial coordinates. Usually it
-makes sense to convert the spatial coordinates to an equidistant projection such
-as UTMs to ensure that distance remains constant throughout the study region
-\citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}. Here we use
-the helper function \texttt{add\_utm\_columns()} to add UTM coordinates with km
-units (so our estimated spatial range parameter is not too big or small). By
-default, the function will guess the UTM zone and create new columns \texttt{X}
-and \texttt{Y}. Since our example data already has these UTM columns, we can
-skip running this code.
+An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. 
+Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}.
+Here we use the helper function \texttt{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small). 
+By default, the function will guess the UTM zone and create new columns \texttt{X} and \texttt{Y}. 
+Since our example data already has these UTM columns, we can skip running this code.
 
@@
 
-\sloppy We then create a mesh object that contains triangulation and projection
-matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
+\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
 % \texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package.
-The argument \texttt{cutoff} defines the minimum allowed distance
-between mesh vertices in the units of \texttt{X} and \texttt{Y} (km).
+The argument \texttt{cutoff} defines the minimum allowed distance between mesh vertices in the units of \texttt{X} and \texttt{Y} (km).
 We could create a basic mesh just specifying this:
@@
 A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
-mesh_pcod2 <- make_mesh(
-  pcod, c("X", "Y"),
-  fmesher_func = fmesher::fm_mesh_2d_inla,
-  cutoff = 8,
-  max.edge = c(10, 40),
-  offset = c(10, 40)
-)
+# mesh_pcod2 <- make_mesh(
+#   pcod, c("X", "Y"),
+#   fmesher_func = fmesher::fm_mesh_2d_inla,
+#   cutoff = 8,
+#   max.edge = c(10, 40),
+#   offset = c(10, 40)
+# )
+# plot(mesh_pcod2)
+inla_mesh <- fmesher::fm_mesh_2d_inla(
+  loc = cbind(pcod$X, pcod$Y), # coordinates
+  max.edge = c(10, 40), # max triangle edge length; inner and outer meshes
+  offset = c(10, 40),  # inner and outer border widths
+  cutoff = 8 # minimum triangle edge length
+)
+mesh_pcod2 <- make_mesh(pcod, c("X", "Y"), mesh = inla_mesh)
 plot(mesh_pcod2)
@@
 
-We will fit a logistic regression of encounter probability with and without
-spatial random fields to illustrate the importance of accounting for spatial
-correlation. In addition to the spatial random field, we include an intercept
-and a quadratic effect of depth on the probability of encounter. Our random
-field model can be written as
+We will fit a logistic regression of encounter probability with and without spatial random fields to illustrate the importance of accounting for spatial correlation. 
+In addition to the spatial random field, we include an intercept and a quadratic effect of depth on the probability of encounter. 
+Our random field model can be written as
 \[
@@
 \]
-where \(\bm{X}^{\mathrm{main}}_{\bm{s}}\) represents a matrix of main effect
-covariates (intercept, quadratic effects of depth), \(\bm{\beta}\) represents
-a vector of estimated main effect coefficients, and \(\omega_{\bm{s}}\)
-represents the estimated spatial random field.
+where \(\bm{X}^{\mathrm{main}}_{\bm{s}}\) represents a matrix of main effect covariates (intercept, quadratic effects of depth), \(\bm{\beta}\) represents a vector of estimated main effect coefficients, and \(\omega_{\bm{s}}\) represents the estimated spatial random field.
 
@@
 
-We can also fit a version that omits the spatial random field by setting
-\texttt{spatial = "off"}. We will use the \texttt{update()} S3 method to refit
-the model while updating any specified arguments:
+We can also fit a version that omits the spatial random field by setting \texttt{spatial = "off"}. 
+We will use the \texttt{update()} S3 method to refit the model while updating any specified arguments:
 
@@
 
-This does not flag any issues. \texttt{sanity()} is checking that the
-\texttt{nlminb()} optimizer reported successful convergence, that the Hessian
-matrix is positive definite, that no extreme or small eigenvalues are detected,
-that no absolute log likelihood gradients with respect to fixed defects are
-$\le$ 0.001, that all fixed effects have reported standard errors that do not
-look unreasonably large ($<$ 100 by default), that random field marginal
-standard deviations are not unexpectedly small or large ($>$
-0.01 and $<$ 100), and that the random field Matérn range parameter does not
-  look unreasonably large ($<$ 1.5 times the largest edge of a bounding box
-  around the observations).
+This does not flag any issues. \texttt{sanity()} is checking that the \texttt{nlminb()} optimizer reported successful convergence,
+that the Hessian matrix is positive definite,
+that no extreme or small eigenvalues are detected,
+that no absolute log likelihood gradients with respect to fixed defects are $\le$ 0.001,
+that all fixed effects have reported standard errors that do not look unreasonably large ($<$ 100 by default),
+that random field marginal standard deviations are not unexpectedly small or large ($>$ 0.01 and $<$ 100),
+and that the random field Matérn range parameter does not look unreasonably large ($<$ 1.5 times the largest edge of a bounding box around the observations).
 
@@
 
-The output indicates our model was fit by maximum (marginal) likelihood
-(\texttt{ML}). We also see the formula, mesh, fitted data, and family. Next we
-see any estimated main effects, the Matérn range distance, the spatial random
-field standard deviation, and the negative log likelihood at convergence.
+The output indicates our model was fit by maximum (marginal) likelihood (\texttt{ML}).
+We also see the formula, mesh, fitted data, and family.
+Next we see any estimated main effects, the Matérn range distance, the spatial random field standard deviation, and the negative log likelihood at convergence.
 
@@
 
-We can test for spatial autocorrelation with a Moran's I test or with a visual
-inspection of the residuals calculated from the \texttt{residuals()} S3 method.
-We omit example code for Moran's I calculation for brevity but include it in the
-code supplement. The p-value for spatial autocorrelation is low for the model
-without a random field (p $<$ 0.01) but large for the model with
-a random field (p = 0.84) suggesting significant residual correlation in the
-simpler model that is alleviated by including the random field.
+We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \texttt{residuals()} S3 method.
+We omit example code for Moran's I calculation for brevity but include it in the code supplement.
+The p-value for spatial autocorrelation is low for the model without a random field (p $<$ 0.01) but large for the model with a random field (p = 0.84) suggesting significant residual correlation in the simpler model that is alleviated by including the random field.
 
@@
 
-We can do additional model comparison with cross validation via
-\texttt{sdmTMB\_cv()}. We will do 10-fold cross validation with the folds
-constructed randomly. Using the \texttt{fold\_ids} argument, we could supply our
-own folds and conduct spatially blocked cross validation. If we set a parallel
-plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will
-be fit in parallel.
+We can do additional model comparison with cross validation via \texttt{sdmTMB\_cv()}.
+We will do 10-fold cross validation with the folds constructed randomly.
+Using the \texttt{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation.
+If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.
 
@@
 
-Indeed, the log likelihood of left-out data is considerably larger for the model
-that includes random fields:
+Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields:
 
@@
 
-\sloppy To visualize our model, we can make predictions with the
-\texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the
-\texttt{newdata} argument to predict on a new data frame with any locations and
-values for the predictor columns. Here, we will predict on a 2 \(\times\) 2 km
-grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can
-visualize the predictions spatially. The grid contains spatial covariate columns
-and all predictors used in the model set at values for which we want to predict.
-The output of \texttt{predict()} is a data frame containing overall estimates in
-link space (\texttt{est}), estimates from the non-random-field components
-(\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the
-individual random field components (here, \texttt{omega\_s}---the spatial
-field). We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2}
-package (Figure~\ref{fig:pcod-predict-maps}).
+\sloppy To visualize our model, we can make predictions with the \texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the \texttt{newdata} argument to predict on a new data frame with any locations and values for the predictor columns.
+Here, we will predict on a 2 \(\times\) 2 km grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
+The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
+The output of \texttt{predict()} is a data frame containing overall estimates in link space (\texttt{est}), estimates from the non-random-field components (\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \texttt{omega\_s}---the spatial field).
+We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
 
@@
 
-As a second example, we will construct a spatiotemporal model of catch rates of
-Pacific Spiny Dogfish (\emph{Squalus suckleyi}) from a trawl survey off the west
-coast of Vancouver Island, Canada. This example extends the spatial model
-described above by also including (1) spatiotemporal fields allowing the
-distribution in each year to be unique, (2) a time-varying intercept as an AR(1)
-process to allow year effects to vary but remain autocorrelated, (3) a smooth
-effect of depth, allowing catch rates to vary non-linearly with depth, and (4)
-spatial anisotropy allowing spatial correlation to be directionally dependent.
-Since catch rates are positive, continuous, and contain zeros, we begin by
-specifying the response family as a Tweedie distribution \citep{tweedie1984}
-with a log link. We then compare alternative families, spatiotemporal random
-field structures, and the exclusion of anisotropy to illustrate the flexibility
-of \pkg{sdmTMB}.
+As a second example, we will construct a spatiotemporal model of catch rates of Pacific Spiny Dogfish (\emph{Squalus suckleyi}) from a trawl survey off the west coast of Vancouver Island, Canada.
+This example extends the spatial model described above by also including (1) spatiotemporal fields allowing the distribution in each year to be unique, (2) a time-varying intercept as an AR(1) process to allow year effects to vary but remain autocorrelated, (3) a smooth effect of depth, allowing catch rates to vary non-linearly with depth, and (4) spatial anisotropy allowing spatial correlation to be directionally dependent.
+Since catch rates are positive, continuous, and contain zeros, we begin by specifying the response family as a Tweedie distribution \citep{tweedie1984} with a log link.
+We then compare alternative families, spatiotemporal random field structures, and the exclusion of anisotropy to illustrate the flexibility of \pkg{sdmTMB}.
 
@@
 
-We next test two additional model formulations: making the spatial correlation
-isotropic (the default) instead of anisotropic, and structuring the positive
-model component spatiotemporal random fields as AR(1) to allow spatiotemporal
-hotspots to persist from year to year.
+We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal hotspots to persist from year to year.
 
@@
 
-We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
-is important here given the elongated shape of the continental shelf with
-a rapid transition to deeper water. We can use \texttt{plot\_anisotropy()} to
-visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
+We find that the anisotropic AR(1) is favoured.
+It makes sense that anisotropy is important here given the elongated shape of the continental shelf with a rapid transition to deeper water.
+We can use \texttt{plot\_anisotropy()} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
 
@@
 
-On inspecting the model output with
-\texttt{summary()}, we notice that the spatial random field SD in the positive
-model component has become very small once AR(1) spatiotemporal fields were
-added. We therefore consider a model where these are turned off in the second
-component of the delta model and choose this model.
+On inspecting the model output with \texttt{summary()}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
+We therefore consider a model where these are turned off in the second component of the delta model and choose this model.
 
@@
 
-We save our chosen model to the object \texttt{fit} to simplify subsequent code,
-run the \texttt{sanity()} check (suppressed for brevity), and inspect
-\texttt{summary()}:
+We save our chosen model to the object \texttt{fit} to simplify subsequent code, run the \texttt{sanity()} check (suppressed for brevity), and inspect \texttt{summary()}:
 
@@
 
-Similarly to the first example, we can visualize model predictions on a grid
-covering the area of interest. Because this is a spatiotemporal model, we first
-need to replicate our grid for each year we will predict on. Since this is
-a common operation, we include the function \texttt{replicate\_df()} to
-replicate a data frame. We then ensure our data frame contains all the
-predictors used in the model (here \texttt{log\_depth}).
+Similarly to the first example, we can visualize model predictions on a grid covering the area of interest.
+Because this is a spatiotemporal model, we first need to replicate our grid for each year we will predict on.
+Since this is a common operation, we include the function \texttt{replicate\_df()} to replicate a data frame.
+We then ensure our data frame contains all the predictors used in the model (here \texttt{log\_depth}).
 
@@
 
-Our prediction data frame is similar to the binomial spatial model, but includes
-columns for the two delta model components (labelled with suffixes \texttt{1}
-and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
-random effects. We can easily generate plots from this data frame using
-\pkg{ggplot2} code with \texttt{geom\_raster()} similar to our spatial example
-with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}). We suppress that code for
-brevity.
+Our prediction data frame is similar to the binomial spatial model, but includes columns for the two delta model components (labelled with suffixes \texttt{1} and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal random effects.
+We can easily generate plots from this data frame using \pkg{ggplot2} code with \texttt{geom\_raster()} similar to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
+We suppress that code for brevity.
 
@@
 
-We can visualize the conditional effect of the bottom depth smoother by
-predicting across a sequence of depths and holding other variables at reference
-values (Figure~\ref{fig:dog-depth-plot}). Here, we pick the last year, specify
-to include both delta model components (\texttt{model = NA}), omit the random
-fields (\texttt{re\_form = NA}), and return standard errors (\texttt{se\_fit\
-= TRUE}).
+We can visualize the conditional effect of the bottom depth smoother by predicting across a sequence of depths and holding other variables at reference values (Figure~\ref{fig:dog-depth-plot}).
+Here, we pick the last year, specify to include both delta model components (\texttt{model = NA}), omit the random fields (\texttt{re\_form = NA}), and return standard errors (\texttt{se\_fit\ = TRUE}).
 
@@
 
-We can generate an area-weighted population index (e.g., a relative or absolute
-index of abundance or biomass) that is independent of sampling locations by
-predicting from the model on a grid covering the area of interest and summing
-the predicted biomass with the \texttt{get\_index()} function (Figure
-\ref{fig:dog-index-plot}). We supply the grid cell area (4
-km\textsuperscript{2}) to the \texttt{area} argument and specify
-\texttt{bias\_correct = TRUE} to enable a bias correction needed due to the
-non-linear transformation of the random effects \citep{thorson2016bias}.
-
+We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \texttt{get\_index()} function (Figure \ref{fig:dog-index-plot}).
+We supply the grid cell area (4 km\textsuperscript{2}) to the \texttt{area} argument and specify \texttt{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
 % TODO and the Laplace?
@@
 
-In our final example, we demonstrate a model with spatially varying coefficient
-(SVC) effects and illustrate combining uncertainty from parameters by working
-with draws from the joint parameter precision matrix. Spatially varying
-coefficient models \citep{hastie1993} are a class of models in which
-coefficients are allowed to vary spatially constrained by some smooth function.
-In addition, we illustrate using restricted maximum likelihood (REML), which
-integrates over the fixed effects as well as the random effects.
-
-Snowy Owls (\emph{Bubo scandiacus}) breed on the arctic tundra and are irruptive
-migrants, meaning that they appear across the mid-latitudes of North America in
-much greater numbers in some winters than others. The reasons for this
-interannual variation in the number of individuals migrating south are not well
-understood but seem to be related to high abundances of food during the breeding
-season and therefore sharp increases in breeding ground population densities
-\citep{robillard2016}. The North Atlantic Oscillation Index (NAO) has been
-linked to productivity of both owls and their prey in Europe \citep{millon2014}.
-Because both productivity and the choice of wintering location could be
-influenced by climate, we modelled a spatially varying effect of annual mean NAO
-index on winter abundance across the southern boundary of their winter
-distribution.
+In our final example, we demonstrate a model with spatially varying coefficient (SVC) effects and illustrate combining uncertainty from parameters by working with draws from the joint parameter precision matrix.
+Spatially varying coefficient models \citep{hastie1993} are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function.
+In addition, we illustrate using restricted maximum likelihood (REML), which integrates over the fixed effects as well as the random effects.
+
+Snowy Owls (\emph{Bubo scandiacus}) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
+The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities \citep{robillard2016}.
+The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe \citep{millon2014}.
+Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
 
@@
 
-We fit counts observed in North America during annual Christmas
-Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011}
-distribution, random intercepts for year, spatial and spatiotemporal random
-fields, and a spatially varying coefficient associated with the NAO.
+We fit counts observed in North America during annual Christmas Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
 
@@
 
-In addition to the output seen for other models, we now have a section for
-random intercepts and a standard deviation for the spatially varying coefficient
-random field. Given our model specification, all random fields are sharing
-a single Matérn range.
+In addition to the output seen for other models, we now have a section for random intercepts and a standard deviation for the spatially varying coefficient random field.
+Given our model specification, all random fields are sharing a single Matérn range.
 
@@
 
-The spatially varying coefficient effect at any point in space is the
-combination of the main effect and spatially varying random effect for
-\texttt{nao}. Mean estimates of the spatially varying random effect are
-available in the output of \texttt{predict.sdmTMB()} in a column starting with
-\texttt{zeta\_s} (here, \texttt{zeta\_s\_nao}). However, we wish to combine the
-fixed and random effects with uncertainty on the combined predictions. We
-illustrate a way of accomplishing this by simulating from the parameter
-precision matrix assuming the joint parameter covariance matrix is multivariate
-normal. We do this by specifying a non-null number of simulation draws to
-\texttt{nsim} in \texttt{predict.sdmTMB()}. By default, \texttt{nsim > 0} will
-return a matrix of draws from the overall prediction. Here, we instead specify
-that we want to return draws from the \texttt{zeta\_s} ($\zeta_{\bm{s}}$) random
-field, which is the spatially varying random field. This returns a matrix where
-each row matches a row of \texttt{newdata} and each column is a simulation draw.
-We then use the function \texttt{spread\_sims()} to draw 200 simulations from
-the parameters themselves, combine the main effect draws (\texttt{sims\$nao})
-with the random field values \texttt{zeta\_s}, and calculate the median for each
-column of data so we can plot them.
+The spatially varying coefficient effect at any point in space is the combination of the main effect and spatially varying random effect for \texttt{nao}.
+Mean estimates of the spatially varying random effect are available in the output of \texttt{predict.sdmTMB()} in a column starting with \texttt{zeta\_s} (here, \texttt{zeta\_s\_nao}).
+However, we wish to combine the fixed and random effects with uncertainty on the combined predictions.
+We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
+We do this by specifying a non-null number of simulation draws to \texttt{nsim} in \texttt{predict.sdmTMB()}.
+By default, \texttt{nsim > 0} will return a matrix of draws from the overall prediction.
+Here, we instead specify that we want to return draws from the \texttt{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the spatially varying random field.
+This returns a matrix where each row matches a row of \texttt{newdata} and each column is a simulation draw.
+We then use the function \texttt{spread\_sims()} to draw 200 simulations from the parameters themselves, combine the main effect draws (\texttt{sims\$nao}) with the random field values \texttt{zeta\_s}, and calculate the median for each column of data so we can plot them.
 
@@
 
-We can make a basic plot using the following code. A more elaborate
-version is shown in Figure~\ref{fig:owl-plot-fancy}.
+We can make a basic plot using the following code. A more elaborate version is shown in Figure~\ref{fig:owl-plot-fancy}.
 
@@
 
-Overall, we find a weak average positive effect of annual mean NAO on overall
-counts, but a southeast to northwest gradient in the intensity of the effect
-(Figure~\ref{fig:owl-plot-fancy}). This result is consistent with owls closest
-to the Atlantic coast and those migrating the furthest south being the most
-affected by the NAO.
+Overall, we find a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure~\ref{fig:owl-plot-fancy}).
+This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
 
@@
 
-We ran a simple speed comparison between \pkg{sdmTMB},
-\proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE
-spatial random field model to 1000 data points with Gaussian error across
-a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
-In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however,
-\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh
-resolution than \pkg{sdmTMB}. \pkg{mgcv} was most affected by mesh resolution.
-Our test was restricted to one core and default \proglang{R} algebra libraries;
-all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries on one core (\pkg{openBLAS}:
-\citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in
-a \textasciitilde10\% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and
-a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
+We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
+In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
+\pkg{mgcv} was most affected by mesh resolution.
+Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
+Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
 
@@
 \includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
-\caption[]{Comparison of time to fit an SPDE spatial random field model with
-  1000 observations, an intercept and one predictor, Gaussian error, and
-  a sequence of SPDE resolutions. Lines represent means and ribbons 95\%
-  quantiles across 50 random iterations. Note the log-distributed x axis.
+\caption[]{
+Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
+Lines represent means and ribbons 95\% quantiles across 50 random iterations.
+Note the log-distributed x axis. 
 \pkg{VAST} should be similar to \pkg{sdmTMB} and so is not shown.
-  \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian
-  approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()}
-  formulation. \pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and
-  discretized covariates \citep{miller2019}. Note that \pkg{spaMM} only fits
-  spatial, not spatiotemporal, models. All platforms were restricted to one core
-and could be faster with parallel computation or optimized algebra
-libraries.}\label{fig:timing}
+\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
+\pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates \citep{miller2019}.
+Note that \pkg{spaMM} only fits spatial, not spatiotemporal, models.
+All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}
+\label{fig:timing}
 \end{figure}
@@
 
-Here, we describe the methods underlying the speed testing in
-Figure~\ref{fig:timing}. We generated a mesh that was consistent across
-simulated data sets for a given mesh resolution
-(Figure~\ref{fig:mesh-vis-timing}). We did this by setting the \texttt{max.edge}
-argument, which controls the largest allowed triangle edge length. We tested
-values of \texttt{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2. In
-Figure~\ref{fig:timing}, we report on the x-axis the number of mesh nodes
-(``vertices'') that result from each of these meshes.
-
-We simulated 1000 spatial observations with both x and y coordinates from
-uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}). Each
-iteration generated unique coordinates, predictor data, Gaussian random field
-values, and observation error. The Gaussian random field was parameterized with
-a range of 0.5 and marginal standard deviation of 0.2. The model included an
-intercept with a value of 0.2 and a normal(0, 1) predictor with an associated
-coefficient of -0.4. The observation error was Gaussian with a standard
-deviation of 0.3.
-
-We conducted 50 simulation iterations per model and mesh resolution and show
-results for the mean, lower, and upper values (Figure~\ref{fig:timing}). The
-models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core
-CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library.
+Here, we describe the methods underlying the speed testing in Figure~\ref{fig:timing}.
+We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Figure~\ref{fig:mesh-vis-timing}).
+We did this by setting the \texttt{max.edge} argument, which controls the largest allowed triangle edge length.
+We tested values of \texttt{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2.
+In Figure~\ref{fig:timing}, we report on the x-axis the number of mesh nodes (``vertices'') that result from each of these meshes.
+
+We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}).
+Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
+The Gaussian random field was parameterized with a range of 0.5 and marginal standard deviation of 0.2.
+The model included an intercept with a value of 0.2 and a normal(0, 1) predictor with an associated coefficient of -0.4.
+The observation error was Gaussian with a standard deviation of 0.3.
+
+We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Figure~\ref{fig:timing}).
+The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library.
 
@@
 
-First define \texttt{smooth.construct.spde.smooth.spec()} and
-\texttt{Predict.matrix.spde.smooth()} from the supplement of \citet{miller2019},
-then:
+First define \texttt{smooth.construct.spde.smooth.spec()} and \texttt{Predict.matrix.spde.smooth()} from the supplement of \citet{miller2019}, then:
 

### 2023-10-06 doc/paper-jss/sdmTMB-paper.Rnw
+ paragraph on applications to other fields + more refs in bib + minor word changes (mostly intro, clarity)
* Noticed there's a mix of \sdmTMB\ and \pkg{sdmTMB} throughout. Change throughout for consistency?

@@
 \Abstract{
-Geostatistical data are common across scientific fields and available in increasingly large datasets. 
-However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users.
-Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
+Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
 SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
@@
 
-Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects are a class of models appropriate for these data \citep{rue2005gmrf, diggle2007, cressie2011}. 
-Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
+Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
+Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
 % TODO: decide on using GRF throughout
 
-Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data. 
+Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
 Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
-These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
+These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
 This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
-The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
+The SPDE approach has been widely adopted, especially via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
 
@@
 
-Software packages for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
+Software packages designed for specifying statistical models which incorporate the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but can be challenging for many applied researchers.
 For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
-Packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE or spatiotemporal functionality.
+While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
 Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
@@
 
-Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices. 
+Here, we introduce the \R\ package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
-Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package. 
+Many features of \pkg{sdmTMB} may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
 One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
 
-This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of TODO  (Section~\ref{discussion}).
+This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).
 
@@
 
-The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
+The design goal of \sdmTMB\ was to develop a flexible, modular, and intuitive inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
 The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.
@@
 
-\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
+\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
@@
 
-TODO add paragraph here on applications to other fields, beyond ecology and fisheries? Specifically thinkgin about:
-- spatial econometrics (e.g. Elhorst, 2012; Lee \& Yu, 2010)
-- disease and medicine (e.g. Wakefield work)
-- agriculture (https://link.springer.com/article/10.1007/s00122-019-03424-y)
+While our focus is primarily in applications to biological data, the SPDE approach and functionality of \sdmTMB\ has applications in many other fields.
+Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
+These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.
 

### 2023-10-18 doc/paper-jss/sdmTMB-paper.Rnw
Restructure user interface section

@@
 \documentclass[article]{jss}
+
+% setwd("doc/paper-jss/")
+% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex")
 
@@
 
-Here, we introduce the \R\ package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
+Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
-Many features of \pkg{sdmTMB} may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
+Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
 One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
@@
 
-\section{Software}\label{sec:software}
-
-The design goal of \sdmTMB\ was to develop a flexible, modular, and intuitive inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
+\section{User interface}\label{sec:software}
+
+The design goal of \sdmTMB\ was to develop a flexible, modular, and intuitive interface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
 The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.
-
-\sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
-\sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
-The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
-Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
-The statistical model itself is written as a \proglang{C++} template for \TMB.
-\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
-After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
-% TODO: visreg + ggeffects effects, etc.?
+\sdmTMB\ relies on several well-established \R\ packages to construct and fit models (Figure~\ref{fig:flowchart}).
 
@@
 The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
-The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()}.
-\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} and \INLA's triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
+The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()} (Figure~\ref{fig:flowchart}).
+\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
+
 Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
-Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).
-The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
-A fitted model summary can be viewed with \texttt{print.sdmTMB()} or \texttt{summary.sdmTMB()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
-\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
-Prediction on fitted or new data is accomplished with \texttt{predict.sdmTMB()}.
-\texttt{residuals.sdmTMB()} calculates various types of residuals.
-New observations can be simulated from a fitted model with \texttt{simulate.sdmTMB()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
-\texttt{sdmTMB\_cv()} facilitates cross validation.
-Derived quantities can be calculated on the fitted model.
-% TODO: more COG refs
-Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
-% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-% Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively).
-% Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
-% Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
-% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-The \sdmTMB\ formula syntax (\texttt{formula}) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
-In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and smoothers.
-Penalized smoothers (i.e., for generalized additive models, GAMs) use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
+Observation distributions and links are specified with standard \R\ family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).
+
+The \sdmTMB\ formula syntax (\texttt{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
+In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and penalized smoothers \texttt{s()} for generalized additive models, GAMS \citep{wood2017a}.
+Penalized smoothers use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
 Supported functionality includes bivarate smoothers \texttt{s(x, y)},
@@
 and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
-In addition to the main formula, \texttt{sdmTMB()} accepts \emph{one-sided} formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).
-
+Beyond the main formula, \texttt{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).
+
+Once the user makes a call to \texttt{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
+If needed, data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+\texttt{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
+
+A fitted model summary can be viewed with \texttt{print()} or \texttt{summary()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
+\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
+Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
+
+Prediction on fitted or new data is accomplished with \texttt{predict()} (\texttt{?predict.sdmTMB}).
 The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
@@
 
-\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
+A variety of model evaluation tools are available.
+A \texttt{residuals()} method calculates various types of residuals.
+New observations can be simulated from a fitted model with \texttt{simulate()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
+\texttt{sdmTMB\_cv()} facilitates cross validation.
+
+% Derived quantities can be calculated on the fitted model.
+% TODO: more COG refs
+% Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
+
+% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+% Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively).
+% Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+% Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
+% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
@@
 These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
-If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
+If a model will be passed to \pkg{tmbstan} with priors, a \texttt{bayesian} logical argument should be set to \texttt{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
 
@@
 
-\subsection{Installation}\label{installation}
+% \subsection{Software dependencies}\label{dependencies}
+
+% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
+% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
+% The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
+% Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+% The statistical model itself is written as a \proglang{C++} template for \TMB.
+% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
+% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
+% % TODO: visreg + ggeffects effects, etc.?
+
+\section{Installation}\label{installation}
 
@@
 <<sdmTMB-install, eval=FALSE, echo=TRUE>>=
-install.packages("sdmTMB", dependencies = TRUE)
-@
-
-Importantly, \texttt{dependencies = TRUE} includes the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is required to generate matrices for the SPDE.
-%Soon, we intend to switch to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.
-
-Alternatively, the development version can be installed with:
+install.packages("sdmTMB")
+@
+
+As an alternative to the CRAN version, the development version can be installed with:
 
@@
 # install.packages("remotes")
-remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
+remotes::install_github("pbs-assess/sdmTMB")
 @
@@
 We omit example code for Moran's I calculation for brevity but include it in the code supplement.
-The p-value for spatial autocorrelation is low for the model without a random field (p $<$ 0.01) but large for the model with a random field (p = 0.84) suggesting significant residual correlation in the simpler model that is alleviated by including the random field.
+% TODO: reword this?
+We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.
 
@@
 
-\sloppy To visualize our model, we can make predictions with the \texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the \texttt{newdata} argument to predict on a new data frame with any locations and values for the predictor columns.
+\sloppy To visualize our model, we can make predictions with the \texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the \texttt{newdata} argument to predict on a new data frame containing locations and all predictor columns.
 Here, we will predict on a 2 \(\times\) 2 km grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
@@
 
+% TODO component confusing:
+
 <<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density.
 Throughout, two example years are shown.
-\\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects.
-The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g.
-habitat).
+(a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, (b) \\code{omega\\_s1} refers to the spatial random field from the first delta model, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and (d) \\code{est} refers to the overall prediction estimate combining all effects.
+The spatial random field is constant through time (i.e., the two panels in b are identical) and represents static biotic or abiotic features not included as covariates (e.g., habitat).
 The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process.
@@
 There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
-\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GRF models.
+\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
 In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
@@
 % TODO: figure out where to put this paragraph
+% TODO: rework this paragraph to be advantages of random fields GAMs and CAR
 The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
 In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
-The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
-The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
-The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\texttt{share\_range} argument).
-
+% The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
+% The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
+% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\texttt{share\_range} argument).
 % TODO: figure out where to put this paragraph
+Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids...
 Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
@@
 There are several planned future additions to the \pkg{sdmTMB} model structure.
-A subset of features to be added in the near future includes (1) multivariate models to allow for simultaneous modelling of data from multiple species, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \proglang{R} rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB}, (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
-The included \pkg{TMB} \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
+A subset of features to be added in the future include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
+(4) The ability to specify observation-level likelihoods to integrate different data types (e.g. REF).
+(5) Conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature.
+(6) Improved visualization methods and integration
+
+% (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
+% The included \pkg{TMB} \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
 % TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations

### 2023-10-18 doc/paper-jss/sdmTMB-paper.Rnw
Wrap paragraph
Commands

@@
 However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
-Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
-SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
-We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
+Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
+We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
 Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
-Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
-We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
+Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
+We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
 }
@@
 This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
-The SPDE approach has been widely adopted, especially via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
+The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
 
@@
 
-Software packages designed for specifying statistical models which incorporate the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but can be challenging for many applied researchers.
-For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
+Software packages designed for specifying statistical models that incorporate the SPDE, such as \INLA\ and \TMB, are flexible and powerful but can be challenging for many applied researchers.
+For example, \TMB\ requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
 While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
@@
 
-Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
+Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \INLA\ to set up SPDE matrices.
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
 Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
-One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
-
-This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
+
+This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).
 
@@
 \caption[]{
-Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
-An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
+Components of an \sdmTMB\ model with illustrations, descriptions, examples, notation, and example code.
+An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
 The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See the \pkg{sdmTMB} \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
-Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
+See the \sdmTMB\ \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
+Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit an intercept.}\label{fig:diagram}
 \end{figure}
@@
 
-% \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
+% \sdmTMB\ fits GLMMs to spatial or spatiotemporal geostatistical data.
 % Geostatistical data refers to data observed at specific spatial coordinates
@@
 % data can be collected across discrete points in time. The process component of
-% an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
+% an \sdmTMB\ model can be formed by any combination of main (``fixed'')
 % effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
@@
 \caption[]{
-  Description of the model fitting procedure in \pkg{sdmTMB}.
+  Description of the model fitting procedure in \sdmTMB.
 }\label{fig:flowchart}
@@
 The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
 After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
@@
 
-\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
+\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
+Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
 These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
@@
 
-% The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
-% a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
+% The \sdmTMB\ model is fit by maximum marginal likelihood. Internally,
+% a \TMB\ \citep{kristensen2016} model template is used to calculate the
 % marginal log likelihood and its gradient, and the negative log likelihood is
 % minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
-% \proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
+% \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
 % maximize the log likelihood conditional on the estimated fixed effects and are
@@
 % rapid model exploration with maximum likelihood, one can optionally pass an
-% \pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
+% \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017,
 % rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
@@
 % goodness of fit according to user-calculated criteria (e.g., mean squared error,
-% area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
+% area under the curve). (3) An \sdmTMB\ model can be passed to the
 % \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
@@
 % \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
-% The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
+% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\texttt{R\_inla} namespace).
 % Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
@@
 % \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
 % After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
@@
 
-\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R}
+\sdmTMB\ can be installed from the Comprehensive \R
 Archive Network (CRAN) at
@@
 This survey is conducted by Fisheries and Oceans Canada and follows a depth-stratified random sampling design, resulting in a georeferenced dataset.
-The data frame \texttt{pcod} is available as package data in \pkg{sdmTMB}.
+The data frame \texttt{pcod} is available as package data in \sdmTMB.
 Relevant columns include latitude, longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of Pacific Cod for a given survey sample.
@@
 
-An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
+An \sdmTMB\ model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
 Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}.
@@
 We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
-% Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
+% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
 For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
@@
 Since catch rates are positive, continuous, and contain zeros, we begin by specifying the response family as a Tweedie distribution \citep{tweedie1984} with a log link.
-We then compare alternative families, spatiotemporal random field structures, and the exclusion of anisotropy to illustrate the flexibility of \pkg{sdmTMB}.
+We then compare alternative families, spatiotemporal random field structures, and the exclusion of anisotropy to illustrate the flexibility of \sdmTMB.
 
@@
 
-There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
-\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
+There are many \R\ packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
+\sdmTMB, \pkg{VAST}, \INLA/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
 In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
-\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
-\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
+\sdmTMB, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \INLA/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
+\sdmTMB\ and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
 To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
-Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
+Of these packages, only \sdmTMB\ and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
 \pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
-% We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.
+% We provide comparisons of the syntax and the reproducibility of results from models fit using \sdmTMB, \INLA, and \pkg{inlabru} in Appendix~\ref{app:inla}.
 
@@
 \toprule
- & \pkg{sdmTMB} & \pkg{VAST} & \proglang{R}-\pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
+ & \sdmTMB\ & \pkg{VAST} & \INLA/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
 \midrule
@@
 
-\caption{Comparison of functionality between several \proglang{R} packages that
+\caption{Comparison of functionality between several \R\ packages that
   can fit geostatistical GLMMs.
@@
 $^2$Penalized smoother GAMs that determine `wiggliness'.
-$^3$\pkg{inlabru} but not \proglang{R}-\pkg{INLA}.
+$^3$\pkg{inlabru} but not \INLA.
 $^4$Spatiotemporal fields as random walk only.
@@
 
-We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
-In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
+We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
+In this test, \sdmTMB\ was the fastest across all mesh resolutions; however, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB.
 \pkg{mgcv} was most affected by mesh resolution.
-Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
+Our test was restricted to one core and default \R\ algebra libraries; all packages could run faster with optimized libraries and parallel processing.
+Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
 
@@
 Note the log-distributed x axis.
-\pkg{VAST} should be similar to \pkg{sdmTMB} and so is not shown.
+\pkg{VAST} should be similar to \sdmTMB\ and so is not shown.
 \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
@@
 Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
-We think users familiar with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
-Users familiar with \proglang{R}-\pkg{INLA} will find \pkg{inlabru} approachable.
-Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
-\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward.
-Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
+We think users familiar with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
+Users familiar with \INLA\ will find \pkg{inlabru} approachable.
+Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
+\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \sdmTMB\ more straightforward.
+Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
 
@@
 % TODO: rework this paragraph to be advantages of random fields GAMs and CAR
-The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+The GLMMs underpinning \sdmTMB\ models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
 In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
-% The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
+% The random fields in \sdmTMB\ are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
 % The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
@@
 Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids...
-Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
+Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
 A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004} (e.g., spatial variance and range---the distance at which points are effectively independent).
 
-With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \proglang{R}-\pkg{INLA}/\pkg{inlabru}.
+With respect to computational speed, \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
-
-Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and simulation from the parameter joint precision matrix.
-
-There are several planned future additions to the \pkg{sdmTMB} model structure.
+However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
+
+Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and simulation from the parameter joint precision matrix.
+
+There are several planned future additions to the \sdmTMB\ model structure.
 A subset of features to be added in the future include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
@@
 % (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
-% The included \pkg{TMB} \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
+% The included \TMB\ \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
 % TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
@@
 
-\pkg{sdmTMB} would not be possible without the \pkg{TMB} \citep{kristensen2016} and \proglang{R}-\pkg{INLA} \citep{rue2009, lindgren2011, lindgren2015} \proglang{R} packages.
-\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} \citep{thorson2019} and \pkg{glmmTMB} \citep{brooks2017} \proglang{R} packages (as described in the \texttt{DESCRIPTION} and \texttt{inst/COPYRIGHTS} files).
+\sdmTMB\ would not be possible without the \TMB\ \citep{kristensen2016} and \INLA\ \citep{rue2009, lindgren2011, lindgren2015} \R\ packages.
+\sdmTMB\ is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} \citep{thorson2019} and \pkg{glmmTMB} \citep{brooks2017} \R\ packages (as described in the \texttt{DESCRIPTION} and \texttt{inst/COPYRIGHTS} files).
 Smoother support was possible thanks to \pkg{mgcv} \citep{wood2017a}.
@@
 
-\subsection[Example sdmTMB model fit]{Example \pkg{sdmTMB} model fit}
+\subsection[Example sdmTMB model fit]{Example \sdmTMB\ model fit}
 

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits + fct + code

@@
 
-Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial},
-% TODO: other non-ecology refs
+Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields,
 they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
@@
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
 % TODO: decide on using GRF throughout
@@
 More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
-The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some function that defines the rate at which spatial covariance decays with distance,
+The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
 \[
 \boldsymbol{\omega} \sim \operatorname{MVNormal}
-  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
+  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),
 \]
+constrained by some function that defines the rate at which spatial covariance decays with distance.
 
@@
 \]
-The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggliness''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
+The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
 The parameter $\nu$ controls the smoothness of the covariance function.
@@
 A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
-A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\sqrt{8} / \kappa$), which has the empirically derived property that correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
+A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
 
@@
 
-In practice, working directly with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
+In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
 and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
@@
 The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
-Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an additional parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
+Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
 \[
@@
 
-We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
+We can extend our spatial model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
 \[
@@
 \]
-In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in Section~\ref{sec:dogar1}).
+In this case, we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in Section~\ref{sec:dogar1}).
 The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
-We demonstrate these model components in the examples below.
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
+We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
 % TODO: add MVN(0, sigma)s
@@
 % expanded description of the statistical model
-% (\texttt{browseVignettes("sdmTMB")} or see the
+% (\code{browseVignettes("sdmTMB")} or see the
 % \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
@@
 
-\section{User interface}\label{sec:software}
+\section{Software design and user interface}\label{sec:software}
 
@@
 The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
-The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()} (Figure~\ref{fig:flowchart}).
-\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
-
-Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
-Observation distributions and links are specified with standard \R\ family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).
-
-The \sdmTMB\ formula syntax (\texttt{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
-In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and penalized smoothers \texttt{s()} for generalized additive models, GAMS \citep{wood2017a}.
-Penalized smoothers use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
-Supported functionality includes bivarate smoothers \texttt{s(x, y)},
-smoothers varying by continuous or categorical variables \texttt{s(x1, by = x2)},
-cyclical smoothers \texttt{s(x, bs = "cc")},
-and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
-Beyond the main formula, \texttt{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).
-
-Once the user makes a call to \texttt{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
-If needed, data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
-\texttt{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
+The user starts by constructing a triangulation mesh for the SPDE approach with \fct{make\_mesh()} (Figure~\ref{fig:flowchart}).
+\fct{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \fct{make\_mesh()}.
+
+Fitting is accomplished with \fct{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \fct{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
+Observation distributions and links are specified with standard \R\ family functions (e.g., \fct{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \fct{nbinom2()}, \fct{delta\_lognormal()}, see \fct{?sdmTMB::Families}).
+
+The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
+In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt()} \citep{barrowman2000}, logistic functions \fct{logistic()}, and penalized smoothers \fct{s()} for generalized additive models, GAMS \citep{wood2017a}.
+Penalized smoothers use the same \fct{s()} and \fct{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
+Supported functionality includes bivarate smoothers \fct{s(x, y)},
+smoothers varying by continuous or categorical variables \fct{s(x1, by = x2)},
+cyclical smoothers \fct{s(x, bs = "cc")},
+and smoothers with specified basis dimensions \fct{s(x, k = 4)} \citep{wood2017a}.
+Beyond the main formula, \fct{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).
+
+Once the user makes a call to \fct{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
+If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+\fct{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
 The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
 After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
 
-A fitted model summary can be viewed with \texttt{print()} or \texttt{summary()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
-\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
-
-Prediction on fitted or new data is accomplished with \texttt{predict()} (\texttt{?predict.sdmTMB}).
-The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
+A fitted model summary can be viewed with \fct{print()} or \fct{summary()} and a set of basic ``sanity'' checks can be run with \fct{sanity()}.
+\fct{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
+Others standard S3 methods are also available such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()}.
+
+Prediction on fitted or new data is accomplished with \fct{predict()} (\fct{?predict.sdmTMB}).
+The \fct{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\fct{newdata}),
 whether to return predictions on the link or response scale (type),
-whether to return standard errors (\texttt{se\_fit},
+whether to return standard errors (\code{se\_fit},
 can be slow if conditioned on random fields),
-whether to condition on the random fields (\texttt{re\_form}),
-whether to condition on the random intercepts (\texttt{re\_form\_iid}),
-which delta model linear predictor to use (\texttt{model}),
-whether to take draws from the joint parameter precision matrix (\texttt{nsim}),
-and whether to use MCMC samples from a \pkg{tmbstan} model fit (\texttt{mcmc\_samples}).
+whether to condition on the random fields (\code{re\_form}),
+whether to condition on the random intercepts (\code{re\_form\_iid}),
+which delta model linear predictor to use (\code{model}),
+whether to take draws from the joint parameter precision matrix (\code{nsim}),
+and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).
 
 A variety of model evaluation tools are available.
-A \texttt{residuals()} method calculates various types of residuals.
-New observations can be simulated from a fitted model with \texttt{simulate()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
-\texttt{sdmTMB\_cv()} facilitates cross validation.
+A \fct{residuals()} method calculates various types of residuals.
+New observations can be simulated from a fitted model with \fct{simulate()} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate()}.
+\fct{sdmTMB\_cv()} facilitates cross validation.
 
@@
 % TODO: more COG refs
-% Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
-
-% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-% Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively).
-% Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
-% Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
-% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
+% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).
+
+% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB()}, \fct{sdmTMB\_simulate()} respectively).
+% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB()}) or calculate residuals (\fct{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).
+% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB()} \code{prior} argument (\code{?sdmTMBpriors}).
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
+Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
 These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
-If a model will be passed to \pkg{tmbstan} with priors, a \texttt{bayesian} logical argument should be set to \texttt{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
+If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
 
@@
 % marginal log likelihood and its gradient, and the negative log likelihood is
-% minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
+% minimized via the non-linear optimization routine \code{stats::nlminb()} in
 % \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
@@
 % approaches to assist this process: (1) The Akaike Information Criterion
-% \citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
+% \citep[AIC,][]{akaike1974} can be calculated with \fct{AIC()}, although AIC
 % has well-documented biases with mixed-effects models \citep{liang2008}. (2)
-% Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
+% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv()} can be used
 % with user-specified or randomly chosen folds for model selection or to evaluate
@@
 % approximation, or perform posterior predictive checks (see
-% \texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
+% \code{?extract\_mcmc}). (4) The \fct{residuals()} method by default returns
 % randomized quantile \citep{dunn1996} or probability integral transform (PIT)
@@
 % of the random effects to avoid this issue \citep{rufener2021} is recommended but
-% slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
-% are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
-% fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
+% slower (\code{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
+% are also possible. (5) The \fct{simulate.sdmTMB()} method can simulate from
+% fitted models and the \fct{sdmTMB\_simulate()} function facilitates
 % simulating data without starting from a fitted model. Models can be fit to these
@@
 % \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
-% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\texttt{R\_inla} namespace).
-% Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\code{R\_inla} namespace).
+% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
 % The statistical model itself is written as a \proglang{C++} template for \TMB.
 % \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
+% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
 % After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
@@
 
-\section{Installation}\label{installation}
+\subsection{Installation}\label{installation}
 
@@
 This survey is conducted by Fisheries and Oceans Canada and follows a depth-stratified random sampling design, resulting in a georeferenced dataset.
-The data frame \texttt{pcod} is available as package data in \sdmTMB.
-Relevant columns include latitude, longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of Pacific Cod for a given survey sample.
+The data frame \code{pcod} is available as package data in \sdmTMB.
+Relevant columns include latitude, longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and encounter (\code{present = 1}) vs.~non-encounter (\code{present = 0}) of Pacific Cod for a given survey sample.
 
@@
 An \sdmTMB\ model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}.
-Here we use the helper function \texttt{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
-By default, the function will guess the UTM zone and create new columns \texttt{X} and \texttt{Y}.
+Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \fct{sf::st\_transform()},][]{pebesma2018}.
+Here we use the helper function \fct{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
+By default, the function will guess the UTM zone and create new columns \code{X} and \code{Y}.
 Since our example data already has these UTM columns, we can skip running this code.
@@
 
-\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
-% \texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package.
-The argument \texttt{cutoff} defines the minimum allowed distance between mesh vertices in the units of \texttt{X} and \texttt{Y} (km).
+\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \fct{make\_mesh()}.
+% \fct{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package.
+The argument \code{cutoff} defines the minimum allowed distance between mesh vertices in the units of \code{X} and \code{Y} (km).
 We could create a basic mesh just specifying this:
@@
 
-We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
-% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
-For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
+We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
+% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh()}.
+For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
 The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).
@@
 We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
-Our mesh has 563 (\texttt{mesh\_pcod2\$mesh\$n}) vertices.
+Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
 Mesh complexity has the single largest influence on the speed of fitting these models.
@@
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
-# mesh_pcod2 <- make_mesh(
-#   pcod, c("X", "Y"),
-#   fmesher_func = fmesher::fm_mesh_2d_inla,
-#   cutoff = 8,
-#   max.edge = c(10, 40),
-#   offset = c(10, 40)
-# )
-# plot(mesh_pcod2)
-inla_mesh <- fmesher::fm_mesh_2d_inla(
-  loc = cbind(pcod$X, pcod$Y), # coordinates
+A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the resulting object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
+mesh_pcod2 <- make_mesh(
+  pcod, c("X", "Y"),
+  fmesher_func = fmesher::fm_mesh_2d_inla,
+  cutoff = 8, # minimum triangle edge length
   max.edge = c(10, 40), # max triangle edge length; inner and outer meshes
-  offset = c(10, 40),  # inner and outer border widths
-  cutoff = 8 # minimum triangle edge length
-)
-mesh_pcod2 <- make_mesh(pcod, c("X", "Y"), mesh = inla_mesh)
+  offset = c(10, 40) # inner and outer border widths
+)
 plot(mesh_pcod2)
@@
 
-We can implement this model with \texttt{sdmTMB()}:
+We can implement this model with \fct{sdmTMB()}:
 
@@
 
-We can also fit a version that omits the spatial random field by setting \texttt{spatial = "off"}.
-We will use the \texttt{update()} S3 method to refit the model while updating any specified arguments:
+We can also fit a version that omits the spatial random field by setting \code{spatial = "off"}.
+We will use the \fct{update()} S3 method to refit the model while updating any specified arguments:
 
@@
 
-We can run some basic checks on our model with the \texttt{sanity()} function:
+We can run some basic checks on our model with the \fct{sanity()} function:
 
@@
 
-This does not flag any issues. \texttt{sanity()} is checking that the \texttt{nlminb()} optimizer reported successful convergence,
+This does not flag any issues. \fct{sanity()} is checking that the \fct{nlminb()} optimizer reported successful convergence,
 that the Hessian matrix is positive definite,
@@
 
-The output indicates our model was fit by maximum (marginal) likelihood (\texttt{ML}).
+The output indicates our model was fit by maximum (marginal) likelihood (\code{ML}).
 We also see the formula, mesh, fitted data, and family.
@@
 
-We can use the \texttt{tidy()} function to obtain a data frame with parameter estimates (standard methods such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()} are also available).
+We can use the \fct{tidy()} function to obtain a data frame with parameter estimates (standard methods such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()} are also available).
 The standard errors on our fixed effects have increased with the spatial random field:
@@
 
-By setting \texttt{effects = "ran\_pars"}, \texttt{tidy()} will return random
-field parameters, where \texttt{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').
+By setting \code{effects = "ran\_pars"}, \fct{tidy()} will return random
+field parameters, where \code{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').
 
@@
 
-We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \texttt{residuals()} S3 method.
+We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals()} S3 method.
 We omit example code for Moran's I calculation for brevity but include it in the code supplement.
@@
 
-We can do additional model comparison with cross validation via \texttt{sdmTMB\_cv()}.
+We can do additional model comparison with cross validation via \fct{sdmTMB\_cv()}.
 We will do 10-fold cross validation with the folds constructed randomly.
-Using the \texttt{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation.
+Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation.
 If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.
@@
 
-\sloppy To visualize our model, we can make predictions with the \texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the \texttt{newdata} argument to predict on a new data frame containing locations and all predictor columns.
-Here, we will predict on a 2 \(\times\) 2 km grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
+\sloppy To visualize our model, we can make predictions with the \fct{predict()} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
+Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
 The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
-The output of \texttt{predict()} is a data frame containing overall estimates in link space (\texttt{est}), estimates from the non-random-field components (\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \texttt{omega\_s}---the spatial field).
-We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
+The output of \fct{predict()} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
+We can plot these with \fct{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
 
@@
 % The Poisson-link models depart from conventional delta models in treating the probability of occurrence at a given location ($p_i$) as a Poisson process rather than a Bernoulli one, $p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals available to be encountered, and $a_i$ is a measure of sampling effort \citep{thorson2018poisson}.
-Note that we choose to omit the spatiotemporal random fields from the binomial model component (\texttt{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
+Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
 
@@
 It makes sense that anisotropy is important here given the elongated shape of the continental shelf with a rapid transition to deeper water.
-We can use \texttt{plot\_anisotropy()} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
+We can use \fct{plot\_anisotropy()} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
 
@@
 
-On inspecting the model output with \texttt{summary()}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
+On inspecting the model output with \fct{summary()}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
 We therefore consider a model where these are turned off in the second component of the delta model and choose this model.
@@
 
-We save our chosen model to the object \texttt{fit} to simplify subsequent code, run the \texttt{sanity()} check (suppressed for brevity), and inspect \texttt{summary()}:
+We save our chosen model to the object \code{fit} to simplify subsequent code, run the \fct{sanity()} check (suppressed for brevity), and inspect \fct{summary()}:
 
@@
 We now have two model components, which are shown one after the other.
-Starting with the binomial component, we have output from the smoother, which includes a linear component (\texttt{slog\_depth}) and the standard deviation on the smoother weights (\texttt{sds(log\_depth)}).
+Starting with the binomial component, we have output from the smoother, which includes a linear component (\code{slog\_depth}) and the standard deviation on the smoother weights (\code{sds(log\_depth)}).
 The smoother summary follows the format used in the \pkg{brms} package \citep{burkner2017}.
@@
 Because this is a spatiotemporal model, we first need to replicate our grid for each year we will predict on.
-Since this is a common operation, we include the function \texttt{replicate\_df()} to replicate a data frame.
-We then ensure our data frame contains all the predictors used in the model (here \texttt{log\_depth}).
+Since this is a common operation, we include the function \fct{replicate\_df()} to replicate a data frame.
+We then ensure our data frame contains all the predictors used in the model (here \code{log\_depth}).
 
@@
 
-Our prediction data frame is similar to the binomial spatial model, but includes columns for the two delta model components (labelled with suffixes \texttt{1} and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal random effects.
-We can easily generate plots from this data frame using \pkg{ggplot2} code with \texttt{geom\_raster()} similar to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
+Our prediction data frame is similar to the binomial spatial model, but includes columns for the two delta model components (labelled with suffixes \code{1} and \code{2}) and adds an \code{epsilon\_st} column for spatiotemporal random effects.
+We can easily generate plots from this data frame using \pkg{ggplot2} code with \fct{geom\_raster()} similar to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
 We suppress that code for brevity.
@@
 We can visualize the conditional effect of the bottom depth smoother by predicting across a sequence of depths and holding other variables at reference values (Figure~\ref{fig:dog-depth-plot}).
-Here, we pick the last year, specify to include both delta model components (\texttt{model = NA}), omit the random fields (\texttt{re\_form = NA}), and return standard errors (\texttt{se\_fit\ = TRUE}).
+Here, we pick the last year, specify to include both delta model components (\code{model = NA}), omit the random fields (\code{re\_form = NA}), and return standard errors (\code{se\_fit\ = TRUE}).
 
@@
 
-We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \texttt{get\_index()} function (Figure \ref{fig:dog-index-plot}).
-We supply the grid cell area (4 km\textsuperscript{2}) to the \texttt{area} argument and specify \texttt{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
+We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index()} function (Figure \ref{fig:dog-index-plot}).
+We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
 % TODO and the Laplace?
@@
 
-\texttt{summary()} prints standard model information:
+\fct{summary()} prints standard model information:
 
@@
 
-The spatially varying coefficient effect at any point in space is the combination of the main effect and spatially varying random effect for \texttt{nao}.
-Mean estimates of the spatially varying random effect are available in the output of \texttt{predict.sdmTMB()} in a column starting with \texttt{zeta\_s} (here, \texttt{zeta\_s\_nao}).
+The spatially varying coefficient effect at any point in space is the combination of the main effect and spatially varying random effect for \code{nao}.
+Mean estimates of the spatially varying random effect are available in the output of \fct{predict.sdmTMB()} in a column starting with \code{zeta\_s} (here, \code{zeta\_s\_nao}).
 However, we wish to combine the fixed and random effects with uncertainty on the combined predictions.
 We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
-We do this by specifying a non-null number of simulation draws to \texttt{nsim} in \texttt{predict.sdmTMB()}.
-By default, \texttt{nsim > 0} will return a matrix of draws from the overall prediction.
-Here, we instead specify that we want to return draws from the \texttt{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the spatially varying random field.
-This returns a matrix where each row matches a row of \texttt{newdata} and each column is a simulation draw.
-We then use the function \texttt{spread\_sims()} to draw 200 simulations from the parameters themselves, combine the main effect draws (\texttt{sims\$nao}) with the random field values \texttt{zeta\_s}, and calculate the median for each column of data so we can plot them.
+We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB()}.
+By default, \code{nsim > 0} will return a matrix of draws from the overall prediction.
+Here, we instead specify that we want to return draws from the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the spatially varying random field.
+This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
+We then use the function \fct{spread\_sims()} to draw 200 simulations from the parameters themselves, combine the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}, and calculate the median for each column of data so we can plot them.
 
@@
 $^7$Zero-inflated NB2 only.
-$^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}.
+$^8$Tweedie power parameter fixed for \fct{mgcv::gamm()}.
 $^9$Possible as log-linked Poisson GLMM with aggregated data.
@@
 \pkg{VAST} should be similar to \sdmTMB\ and so is not shown.
-\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
-\pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates \citep{miller2019}.
+\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like()} formulation.
+\pkg{mgcv} used \fct{bam()}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
 Note that \pkg{spaMM} only fits spatial, not spatiotemporal, models.
@@
 Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
-We think users familiar with \texttt{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
+We think users familiar with \fct{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
 Users familiar with \INLA\ will find \pkg{inlabru} approachable.
@@
 % The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
-% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\texttt{share\_range} argument).
+% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\code{share\_range} argument).
 % TODO: figure out where to put this paragraph
@@
 % (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
-% The included \TMB\ \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
+% The included \TMB\ \code{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
 % TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
@@
 \sdmTMB\ would not be possible without the \TMB\ \citep{kristensen2016} and \INLA\ \citep{rue2009, lindgren2011, lindgren2015} \R\ packages.
-\sdmTMB\ is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} \citep{thorson2019} and \pkg{glmmTMB} \citep{brooks2017} \R\ packages (as described in the \texttt{DESCRIPTION} and \texttt{inst/COPYRIGHTS} files).
+\sdmTMB\ is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} \citep{thorson2019} and \pkg{glmmTMB} \citep{brooks2017} \R\ packages (as described in the \code{DESCRIPTION} and \code{inst/COPYRIGHTS} files).
 Smoother support was possible thanks to \pkg{mgcv} \citep{wood2017a}.
@@
 We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Figure~\ref{fig:mesh-vis-timing}).
-We did this by setting the \texttt{max.edge} argument, which controls the largest allowed triangle edge length.
-We tested values of \texttt{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2.
+We did this by setting the \code{max.edge} argument, which controls the largest allowed triangle edge length.
+We tested values of \code{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2.
 In Figure~\ref{fig:timing}, we report on the x-axis the number of mesh nodes (``vertices'') that result from each of these meshes.
@@
 
-We will illustrate with a \texttt{max.edge} of 0.06:
+We will illustrate with a \code{max.edge} of 0.06:
 
@@
 
-This mesh has \Sexpr{vertices} (\texttt{mesh\$n}) vertices.
+This mesh has \Sexpr{vertices} (\code{mesh\$n}) vertices.
 
@@
 
-First define \texttt{smooth.construct.spde.smooth.spec()} and \texttt{Predict.matrix.spde.smooth()} from the supplement of \citet{miller2019}, then:
+First define \fct{smooth.construct.spde.smooth.spec()} and \fct{Predict.matrix.spde.smooth()} from the supplement of \citet{miller2019}, then:
 

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
fct fix

@@
 An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
-The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
+The examples are from an ecology context, but the model can be fit to any spatially referenced point data.
 Notation: We refer to design matrices as $\bm{X}$.
@@
 All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
-See the \sdmTMB\ \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
-Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit an intercept.}\label{fig:diagram}
+Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint ``hockey-stick'' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit the intercept.}\label{fig:diagram}
 \end{figure}
@@
 
-The expectation for a new data point is then the probability of a non-zero multiplied by the positive rate: $p \cdot r$.
+The expectation for a new data point is then the probability of a non-zero event multiplied by the positive rate: $p \cdot r$.
 
@@
 The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
-The user starts by constructing a triangulation mesh for the SPDE approach with \fct{make\_mesh()} (Figure~\ref{fig:flowchart}).
-\fct{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \fct{make\_mesh()}.
-
-Fitting is accomplished with \fct{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \fct{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
-Observation distributions and links are specified with standard \R\ family functions (e.g., \fct{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \fct{nbinom2()}, \fct{delta\_lognormal()}, see \fct{?sdmTMB::Families}).
+The user starts by constructing a triangulation mesh for the SPDE approach with \fct{make\_mesh} (Figure~\ref{fig:flowchart}).
+\fct{make\_mesh} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \fct{make\_mesh}.
+
+Fitting is accomplished with \fct{sdmTMB}, which has arguments similar to \pkg{glmmTMB}'s \fct{glmmTMB} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
+Observation distributions and links are specified with standard \R\ family functions (e.g., \fct{binomial}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \fct{nbinom2}, \fct{delta\_lognormal}, see \code{?sdmTMB::Families}).
 
 The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
-In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt()} \citep{barrowman2000}, logistic functions \fct{logistic()}, and penalized smoothers \fct{s()} for generalized additive models, GAMS \citep{wood2017a}.
-Penalized smoothers use the same \fct{s()} and \fct{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
-Supported functionality includes bivarate smoothers \fct{s(x, y)},
-smoothers varying by continuous or categorical variables \fct{s(x1, by = x2)},
-cyclical smoothers \fct{s(x, bs = "cc")},
-and smoothers with specified basis dimensions \fct{s(x, k = 4)} \citep{wood2017a}.
-Beyond the main formula, \fct{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).
-
-Once the user makes a call to \fct{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
-If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
-\fct{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
+In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt} \citep{barrowman2000}, logistic functions \fct{logistic}, and penalized smoothers \fct{s} for generalized additive models, GAMS \citep{wood2017a}.
+Penalized smoothers use the same \fct{s} and \fct{t2} syntax as in \pkg{mgcv} \citep{wood2017a}.
+Supported functionality includes bivarate smoothers \code{s(x, y)},
+smoothers varying by continuous or categorical variables \code{s(x1, by = x2)},
+cyclical smoothers \code{s(x, bs = "cc")},
+and smoothers with specified basis dimensions \code{s(x, k = 4)} \citep{wood2017a}.
+Beyond the main formula, \fct{sdmTMB} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).
+
+Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
+If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
 The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
 After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
 
-A fitted model summary can be viewed with \fct{print()} or \fct{summary()} and a set of basic ``sanity'' checks can be run with \fct{sanity()}.
-\fct{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-Others standard S3 methods are also available such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()}.
-
-Prediction on fitted or new data is accomplished with \fct{predict()} (\fct{?predict.sdmTMB}).
-The \fct{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\fct{newdata}),
+A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
+\fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
+Others standard S3 methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
+
+Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
+The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
 whether to return predictions on the link or response scale (type),
@@
 A variety of model evaluation tools are available.
-A \fct{residuals()} method calculates various types of residuals.
-New observations can be simulated from a fitted model with \fct{simulate()} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate()}.
-\fct{sdmTMB\_cv()} facilitates cross validation.
+A \fct{residuals} method calculates various types of residuals.
+New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
+\fct{sdmTMB\_cv} facilitates cross validation.
 
@@
 % TODO: more COG refs
-% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).
-
-% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB()}, \fct{sdmTMB\_simulate()} respectively).
-% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB()}) or calculate residuals (\fct{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
-% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).
-% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB()} \code{prior} argument (\code{?sdmTMBpriors}).
+% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).
+
+% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB}, \fct{sdmTMB\_simulate} respectively).
+% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB}) or calculate residuals (\fct{residuals.sdmTMB}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
+% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).
+% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
+
+\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
@@
 % approaches to assist this process: (1) The Akaike Information Criterion
-% \citep[AIC,][]{akaike1974} can be calculated with \fct{AIC()}, although AIC
+% \citep[AIC,][]{akaike1974} can be calculated with \fct{AIC}, although AIC
 % has well-documented biases with mixed-effects models \citep{liang2008}. (2)
-% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv()} can be used
+% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv} can be used
 % with user-specified or randomly chosen folds for model selection or to evaluate
@@
 % approximation, or perform posterior predictive checks (see
-% \code{?extract\_mcmc}). (4) The \fct{residuals()} method by default returns
+% \code{?extract\_mcmc}). (4) The \fct{residuals} method by default returns
 % randomized quantile \citep{dunn1996} or probability integral transform (PIT)
@@
 % slower (\code{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
-% are also possible. (5) The \fct{simulate.sdmTMB()} method can simulate from
-% fitted models and the \fct{sdmTMB\_simulate()} function facilitates
+% are also possible. (5) The \fct{simulate.sdmTMB} method can simulate from
+% fitted models and the \fct{sdmTMB\_simulate} function facilitates
 % simulating data without starting from a fitted model. Models can be fit to these
@@
 % The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\code{R\_inla} namespace).
-% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
+% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
 % The statistical model itself is written as a \proglang{C++} template for \TMB.
 % \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
+% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
 % After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
@@
 
-\sdmTMB\ can be installed from the Comprehensive \R
-Archive Network (CRAN) at
+\sdmTMB\ can be installed from the Comprehensive \R\ Archive Network (CRAN) at
 \url{https://CRAN.R-project.org/package=sdmTMB}
@@
 
-Additional utilities are maintained in the \pkg{sdmTMBextra} package at\\
+Additional utilities that require heavier package dependencies (such as \INLA\ and \pkg{rstan}) are maintained in the \pkg{sdmTMBextra} package at\\
 \url{https://github.com/pbs-assess/sdmTMBextra}
@@
 An \sdmTMB\ model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
-Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \fct{sf::st\_transform()},][]{pebesma2018}.
-Here we use the helper function \fct{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
+Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \fct{sf::st\_transform},][]{pebesma2018}.
+Here we use the helper function \fct{add\_utm\_columns} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
 By default, the function will guess the UTM zone and create new columns \code{X} and \code{Y}.
@@
 
-\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \fct{make\_mesh()}.
-% \fct{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package.
+\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \fct{make\_mesh}.
+% \fct{make\_mesh} is a wrapper function for functions in the \pkg{fmesher} package.
 The argument \code{cutoff} defines the minimum allowed distance between mesh vertices in the units of \code{X} and \code{Y} (km).
@@
 
-We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
-% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh()}.
+We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
+% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh}.
 For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
@@
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
-A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the resulting object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
+A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
 mesh_pcod2 <- make_mesh(
@@
 
-We can implement this model with \fct{sdmTMB()}:
+We can implement this model with \fct{sdmTMB}:
 
@@
 We can also fit a version that omits the spatial random field by setting \code{spatial = "off"}.
-We will use the \fct{update()} S3 method to refit the model while updating any specified arguments:
+We will use the \fct{update} S3 method to refit the model while updating any specified arguments:
 
@@
 
-We can run some basic checks on our model with the \fct{sanity()} function:
+We can run some basic checks on our model with the \fct{sanity} function:
 
@@
 
-This does not flag any issues. \fct{sanity()} is checking that the \fct{nlminb()} optimizer reported successful convergence,
+This does not flag any issues. \fct{sanity} is checking that the \fct{nlminb} optimizer reported successful convergence,
 that the Hessian matrix is positive definite,
@@
 
-We can use the \fct{tidy()} function to obtain a data frame with parameter estimates (standard methods such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()} are also available).
+We can use the \fct{tidy} function to obtain a data frame with parameter estimates (standard methods such as \fct{fixef}, \fct{confint}, and \fct{vcov} are also available).
 The standard errors on our fixed effects have increased with the spatial random field:
@@
 
-By setting \code{effects = "ran\_pars"}, \fct{tidy()} will return random
+By setting \code{effects = "ran\_pars"}, \fct{tidy} will return random
 field parameters, where \code{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').
@@
 
-We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals()} S3 method.
+We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} S3 method.
 We omit example code for Moran's I calculation for brevity but include it in the code supplement.
@@
 
-We can do additional model comparison with cross validation via \fct{sdmTMB\_cv()}.
+We can do additional model comparison with cross validation via \fct{sdmTMB\_cv}.
 We will do 10-fold cross validation with the folds constructed randomly.
@@
 
-\sloppy To visualize our model, we can make predictions with the \fct{predict()} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
+\sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
 Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
 The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
-The output of \fct{predict()} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
-We can plot these with \fct{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
+The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
+We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
 
@@
 It makes sense that anisotropy is important here given the elongated shape of the continental shelf with a rapid transition to deeper water.
-We can use \fct{plot\_anisotropy()} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
+We can use \fct{plot\_anisotropy} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
 
@@
 
-On inspecting the model output with \fct{summary()}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
+On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
 We therefore consider a model where these are turned off in the second component of the delta model and choose this model.
@@
 
-We save our chosen model to the object \code{fit} to simplify subsequent code, run the \fct{sanity()} check (suppressed for brevity), and inspect \fct{summary()}:
+We save our chosen model to the object \code{fit} to simplify subsequent code, run the \fct{sanity} check (suppressed for brevity), and inspect \fct{summary}:
 
@@
 Because this is a spatiotemporal model, we first need to replicate our grid for each year we will predict on.
-Since this is a common operation, we include the function \fct{replicate\_df()} to replicate a data frame.
+Since this is a common operation, we include the function \fct{replicate\_df} to replicate a data frame.
 We then ensure our data frame contains all the predictors used in the model (here \code{log\_depth}).
@@
 Our prediction data frame is similar to the binomial spatial model, but includes columns for the two delta model components (labelled with suffixes \code{1} and \code{2}) and adds an \code{epsilon\_st} column for spatiotemporal random effects.
-We can easily generate plots from this data frame using \pkg{ggplot2} code with \fct{geom\_raster()} similar to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
+We can easily generate plots from this data frame using \pkg{ggplot2} code with \fct{geom\_raster} similar to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
 We suppress that code for brevity.
@@
 
-We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index()} function (Figure \ref{fig:dog-index-plot}).
+We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
 We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
@@
 
-\fct{summary()} prints standard model information:
+\fct{summary} prints standard model information:
 
@@
 The spatially varying coefficient effect at any point in space is the combination of the main effect and spatially varying random effect for \code{nao}.
-Mean estimates of the spatially varying random effect are available in the output of \fct{predict.sdmTMB()} in a column starting with \code{zeta\_s} (here, \code{zeta\_s\_nao}).
+Mean estimates of the spatially varying random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (here, \code{zeta\_s\_nao}).
 However, we wish to combine the fixed and random effects with uncertainty on the combined predictions.
 We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
-We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB()}.
+We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
 By default, \code{nsim > 0} will return a matrix of draws from the overall prediction.
@@
 This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
-We then use the function \fct{spread\_sims()} to draw 200 simulations from the parameters themselves, combine the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}, and calculate the median for each column of data so we can plot them.
+We then use the function \fct{spread\_sims} to draw 200 simulations from the parameters themselves, combine the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}, and calculate the median for each column of data so we can plot them.
 
@@
 $^7$Zero-inflated NB2 only.
-$^8$Tweedie power parameter fixed for \fct{mgcv::gamm()}.
+$^8$Tweedie power parameter fixed for \fct{mgcv::gamm}.
 $^9$Possible as log-linked Poisson GLMM with aggregated data.
@@
 \pkg{VAST} should be similar to \sdmTMB\ and so is not shown.
-\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like()} formulation.
-\pkg{mgcv} used \fct{bam()}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
+\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
+\pkg{mgcv} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
 Note that \pkg{spaMM} only fits spatial, not spatiotemporal, models.
@@
 Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
-We think users familiar with \fct{stats::glm()}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
+We think users familiar with \fct{stats::glm}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
 Users familiar with \INLA\ will find \pkg{inlabru} approachable.
@@
 
-First define \fct{smooth.construct.spde.smooth.spec()} and \fct{Predict.matrix.spde.smooth()} from the supplement of \citet{miller2019}, then:
+First define \fct{smooth.construct.spde.smooth.spec} and \fct{Predict.matrix.spde.smooth} from the supplement of \citet{miller2019}, then:
 

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Editing

@@
 
-We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
+We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge (\code{max.edge}) length of 10 km and 40 km for the inner and outer mesh, and an \code{offset} width of 10 km and 40 km for the inner and outer meshes borders.
 % Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh}.
@@
 \]
-where \(\bm{X}^{\mathrm{main}}_{\bm{s}}\) represents a matrix of main effect covariates (intercept, quadratic effects of depth), \(\bm{\beta}\) represents a vector of estimated main effect coefficients, and \(\omega_{\bm{s}}\) represents the estimated spatial random field.
+where \(\bm{X}^{\mathrm{main}}_{\bm{s}}\) represents a matrix of main effect covariates (intercept and quadratic effects of depth), \(\bm{\beta}\) represents a vector of estimated main effect coefficients, and \(\omega_{\bm{s}}\) represents the estimated spatial random field at location $\s$.
 
@@
 that random field marginal standard deviations are not unexpectedly small or large ($>$ 0.01 and $<$ 100),
-and that the random field Matérn range parameter does not look unreasonably large ($<$ 1.5 times the largest edge of a bounding box around the observations).
+and that the random field Matérn range parameter does not look unreasonably large ($<$ 1.5 times the largest distance from a bounding box around the observations).
 
@@
 By setting \code{effects = "ran\_pars"}, \fct{tidy} will return random
-field parameters, where \code{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').
+field parameters, where \code{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (``O'' for ``Omega'').
 
@@
 
-We can also see that the AIC of the model with spatial random fields is
+We can also see that the conditional AIC of the model with spatial random fields is
 lower:
@@
 
-Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields:
+Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields indicating better out-of-sample predictions:
 
@@
 @
+
+In practice, we would want to repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 As a second example, we will construct a spatiotemporal model of catch rates of Pacific Spiny Dogfish (\emph{Squalus suckleyi}) from a trawl survey off the west coast of Vancouver Island, Canada.
-This example extends the spatial model described above by also including (1) spatiotemporal fields allowing the distribution in each year to be unique, (2) a time-varying intercept as an AR(1) process to allow year effects to vary but remain autocorrelated, (3) a smooth effect of depth, allowing catch rates to vary non-linearly with depth, and (4) spatial anisotropy allowing spatial correlation to be directionally dependent.
+This example extends the spatial model described above by including (1) spatiotemporal fields, allowing unique spatially correlated latent effects each year; (2) a time-varying intercept as an AR(1) process, allowing year effects to vary but remain autocorrelated; (3) a smooth effect of depth, allowing catch rates to vary non-linearly with depth; and (4) spatial anisotropy, allowing spatial correlation to be directionally dependent.
 Since catch rates are positive, continuous, and contain zeros, we begin by specifying the response family as a Tweedie distribution \citep{tweedie1984} with a log link.
@@
 \]
-where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents spatiotemporal fields.
+where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) (representing latent spatial effects that vary by year).
 The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
@@
 We now have two model components, which are shown one after the other.
-Starting with the binomial component, we have output from the smoother, which includes a linear component (\code{slog\_depth}) and the standard deviation on the smoother weights (\code{sds(log\_depth)}).
+Starting with the binomial component, we have output from the smoother, which includes a linear parameter (\code{slog\_depth}) and the standard deviation on the smoother weights (\code{sds(log\_depth)}).
 The smoother summary follows the format used in the \pkg{brms} package \citep{burkner2017}.
@@
 
-<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density.
+<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction elements from the spatiotemporal model of Pacific Dogfish biomass density.
 Throughout, two example years are shown.
-(a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, (b) \\code{omega\\_s1} refers to the spatial random field from the first delta model, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and (d) \\code{est} refers to the overall prediction estimate combining all effects.
+(a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field elements (here, a smoother for bottom depth and the time-varying year effect) from the first delta model component, (b) \\code{omega\\_s1} refers to the spatial random field from the first delta model component, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model component, and (d) \\code{est} refers to the overall prediction estimate combining all effects.
 The spatial random field is constant through time (i.e., the two panels in b are identical) and represents static biotic or abiotic features not included as covariates (e.g., habitat).
@@
 The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively.
-Other fixed effects are held at constant values and the random fields are omitted from the prediction.", out.width="4in">>=
+Other fixed effects are held at constant values and the random fields are set to their expected value (zero).", out.width="4in">>=
 ggplot(pred_depth, aes(
@@
 
-Our data are read in from supplementary data and contain columns for UTM spatial coordinates, year, year as a factor, and the annual NAO value:
-
+Our data are read in from supplementary data and contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:
 <<owl-data>>=
@@
 
-We fit counts observed in North America during annual Christmas Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+We will fit counts observed in North America during annual Christmas Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
 

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 \pkg{mgcv} was most affected by mesh resolution.
-Our test was restricted to one core and default \R\ algebra libraries; all packages could run faster with optimized libraries and parallel processing.
-Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
+Our test was restricted to one core and default \R\ algebra libraries.
+% all packages could run faster with optimized libraries and parallel processing.
+% Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
 
@@
 
-While our focus is primarily in applications to biological data, the SPDE approach and functionality of \sdmTMB\ has applications in many other fields.
+While our examples focused on applications to ecological data, the SPDE approach and the functionality of \sdmTMB\ has applications in many other fields.
 Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
@@
 
-TODO add paragraph here on uniting functionality not available in any one package?
+\textbf{TODO add paragraph here on uniting functionality not available in any one package?}
 
@@
 % TODO: figure out where to put this paragraph
-Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids...
+Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids.
 Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
-A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004} (e.g., spatial variance and range---the distance at which points are effectively independent).
+A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004}.
 
@@
 
-Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and simulation from the parameter joint precision matrix.
+Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, and time-varying spatiotemporal covariance parameters \citep{ward2022}.
 
 There are several planned future additions to the \sdmTMB\ model structure.
-A subset of features to be added in the future include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
-(4) The ability to specify observation-level likelihoods to integrate different data types (e.g. REF).
-(5) Conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature.
-(6) Improved visualization methods and integration
+A subset of features to be added include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
+(4) the ability to specify observation-level likelihoods to integrate different data types (e.g. REF),
+(5) conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature, and
+(6) improved integration with existing visualization methods such as \pkg{ggeffects} \citep{ggeffects}.
 

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Abstract

@@
 However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
-Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
+Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent Gaussian Markov random fields using an SPDE-(stochastic partial differential equation) based approach.
 SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 \fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-Others standard S3 methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
+Others standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
 
@@
 
-As an alternative to the CRAN version, the development version can be installed with:
+Users who wish to automatically install suggested packages as well may wish to use
+
+<<sdmTMB-install-suggest, eval=FALSE, echo=TRUE>>=
+install.packages("sdmTMB", dependencies = TRUE)
+@
+
+As an alternative to the CRAN version, the development version can be installed with
 
 <<remotes, eval=FALSE, echo=TRUE>>=
-# install.packages("remotes")
-remotes::install_github("pbs-assess/sdmTMB")
+install.packages("remotes")
+remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
 @
@@
 mesh_pcod2 <- make_mesh(
-  pcod, c("X", "Y"),
+  pcod,
+  xy_cols = c("X", "Y"),
   fmesher_func = fmesher::fm_mesh_2d_inla,
-  cutoff = 8, # minimum triangle edge length
-  max.edge = c(10, 40), # max triangle edge length; inner and outer meshes
-  offset = c(10, 40) # inner and outer border widths
+  cutoff = 8,
+  max.edge = c(10, 40),
+  offset = c(10, 40)
 )
@@
 We can also fit a version that omits the spatial random field by setting \code{spatial = "off"}.
-We will use the \fct{update} S3 method to refit the model while updating any specified arguments:
+We will use the \fct{update} method to refit the model while updating any specified arguments:
 
@@
 
-We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} S3 method.
+We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} method.
 We omit example code for Moran's I calculation for brevity but include it in the code supplement.

### 2023-10-20 doc/paper-jss/sdmTMB-paper.Rnw
Fix #15
delta method
Cleanup

@@
 
-% Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.
-
 Software packages designed for specifying statistical models that incorporate the SPDE, such as \INLA\ and \TMB, are flexible and powerful but can be challenging for many applied researchers.
@@
 
-% We provide a more detailed comparison of related software packages in Table~\ref{tab:functionality} and the Discussion (Section~\ref{discussion}).
-
 Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \INLA\ to set up SPDE matrices.
@@
 This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).
-
-% TODO: there is now some duplication between these two paragraphs!?
 
@@
 
-% Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .
-
 \subsection{Geometric anisotropy}
@@
 \begin{aligned}
-% \operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
-% \log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
 \operatorname{logit} (p) &= \eta_1,\\
@@
 \begin{aligned}
-% \log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
-% \log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
 \log (n) &= \eta_1,\\
@@
 
-% \sdmTMB\ fits GLMMs to spatial or spatiotemporal geostatistical data.
-% Geostatistical data refers to data observed at specific spatial coordinates
-% reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
-% data can be collected across discrete points in time. The process component of
-% an \sdmTMB\ model can be formed by any combination of main (``fixed'')
-% effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
-% identically distributed) random intercepts \(\alpha_g\), spatial intercept
-% random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
-% \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
-% varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
-% are discussed in more detail via our examples below.
-
-% This process component is combined with an observation error family (e.g.,
-% Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
-% log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
-% families can be combined into a two-part ``delta'' or ``hurdle'' model
-% \citep{aitchison1955} to model the zero vs.~non-zero observations separately
-% from the positive observations. See the model description vignette for an
-% expanded description of the statistical model
-% (\code{browseVignettes("sdmTMB")} or see the
-% \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).
-
-% By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}).
-% Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
-
 \section{Software design and user interface}\label{sec:software}
@@
 \end{figure}
-
-% TODO: () on end of functions!? see flowchart vs. rest
-% TODO: blah.sdmTMB() for S3 methods?
 
@@
 After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
+
+\textbf{TODO: generalized delta method! and \citep{zheng2021}}
 
@@
 
-% Describe modular design philosopy
-
-% The design goal of...
-
-% flexible, modular, familiar, make SPDE accessible and fast
-
-% combine functionality commonly needed together
-
-% The \sdmTMB\ model is fit by maximum marginal likelihood. Internally,
-% a \TMB\ \citep{kristensen2016} model template is used to calculate the
-% marginal log likelihood and its gradient, and the negative log likelihood is
-% minimized via the non-linear optimization routine \code{stats::nlminb()} in
-% \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
-% maximize the log likelihood conditional on the estimated fixed effects and are
-% integrated over via the Laplace approximation \citep{kristensen2016}. After
-% rapid model exploration with maximum likelihood, one can optionally pass an
-% \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017,
-% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
-% distribution for Bayesian inference.
-
 % TODO: describe this in more detail and how it could apply beyond ecology;
@@
 % `ecology` from the description
-
-% TODO rename this? move this? add additional sections above!?
 
@@
 
-% \subsection{Software dependencies}\label{dependencies}
-
-% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
-% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
-% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\code{R\_inla} namespace).
-% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
-% The statistical model itself is written as a \proglang{C++} template for \TMB.
-% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
-% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
-% % TODO: visreg + ggeffects effects, etc.?
-
 \subsection{Installation}\label{installation}
@@
 We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge (\code{max.edge}) length of 10 km and 40 km for the inner and outer mesh, and an \code{offset} width of 10 km and 40 km for the inner and outer meshes borders.
-% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh}.
 For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
@@
 We omit example code for Moran's I calculation for brevity but include it in the code supplement.
-% TODO: reword this?
 We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.
@@
 Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
-% The Poisson-link models depart from conventional delta models in treating the probability of occurrence at a given location ($p_i$) as a Poisson process rather than a Bernoulli one, $p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals available to be encountered, and $a_i$ is a measure of sampling effort \citep{thorson2018poisson}.
 Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
-
-% TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.
 
@@
 
-% TODO: Need description of anisotropy and why it's important here or somewhere else
-
 We find that the anisotropic AR(1) is favoured.
@@
 We then ensure our data frame contains all the predictors used in the model (here \code{log\_depth}).
-
-% TODO Propose moving this to a short paragraph `Prediction` to highlight modularity
 
@@
 @
-
-% TODO component confusing:
 
@@
 We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
-% TODO and the Laplace?
 
@@
 Our data are read in from supplementary data and contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:
+
 <<owl-data>>=
@@
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
-% We provide comparisons of the syntax and the reproducibility of results from models fit using \sdmTMB, \INLA, and \pkg{inlabru} in Appendix~\ref{app:inla}.
 
@@
 \end{threeparttable}
-\label{tab:functionality}
 
@@
 }
+\label{tab:functionality}
 \end{table}
@@
 Our test was restricted to one core and default \R\ algebra libraries.
-% all packages could run faster with optimized libraries and parallel processing.
-% Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
 
@@
 
-% (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
-% The included \TMB\ \code{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
-% TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
-% removed random slopes -- another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750
+% another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750
 

### 2023-11-24 doc/paper-jss/sdmTMB-paper.Rnw
add uncertainty to owl example

@@
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe \citep{millon2014}.
-Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
+Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on the end of year's abundance across the southern boundary of their winter distribution.
+Annual mean NAO captures the preceding winter's conditions, combined with breeding season and early winter climate.
 
@@
 This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
-We then use the function \fct{spread\_sims} to draw 200 simulations from the parameters themselves, combine the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}, and calculate the median for each column of data so we can plot them.
+We then use the function \fct{spread\_sims} to draw 200 simulations from the parameters themselves, combine the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}, and calculate the median, and upper and lower quantiles for each column of data so we can plot them.
+For this example we use thresholds of 0.1 and 0.9 representing confidence invervals (CI) of 80\%.
 
@@
 combined <- sims$nao + t(zeta_s)
-snow$nao_effect <- apply(exp(combined), 2, median)
+snow$nao_effect <- exp(apply(combined, 2, median))
+snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.1))
+snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.9))
 @
@@
 
-Overall, we find a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure~\ref{fig:owl-plot-fancy}).
+Overall, we find a weak average positive effect of annual mean NAO on overall counts with a southeast to northwest gradient in the intensity of the effect (Figure~\ref{fig:owl-plot-fancy}).
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
-
-% TODO keep this? We provide an extended version of this example in Appendix \ref{app:owls}.
+However, if we recalculated the CI at 95\% level the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
 
@@
 <<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
-Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
 The effect is multiplicative on owl count per NAO unit.
-In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, fig.pos="htbp", out.width="4in">>=
-nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
+In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, out.width="5.25in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
+nao_effect_df <- select(snow, X, Y, nao_effect, nao_effect_lwr, nao_effect_upr) |> distinct() |>
   mutate(X = X * 100000, Y = Y * 100000) # to match CRS
-ggplot(data = nao_effect_df) +
+
+snow_g1 <- ggplot(data = nao_effect_df) +
   geom_point(aes(X, Y, colour = nao_effect)) +
@@
   scale_colour_viridis_c(
+    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
     guide = guide_colourbar(direction = "horizontal", title.position = "left")
   ) +
+  ggtitle("(a) Median") +
   labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
-    theme(legend.position = "bottom", axis.title = element_blank())
+  theme(legend.position = "top",
+          plot.title = element_text(size = 10, hjust = -0.2),
+          axis.title = element_blank())
+
+snow_g2 <- ggplot(data = nao_effect_df) +
+  geom_point(aes(X, Y, colour = nao_effect_lwr)) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  coord_sf(
+    xlim = c(min(snow2$X), max(snow2$X)),
+    ylim = c(min(snow2$Y), max(snow2$Y))
+  ) +
+  scale_colour_viridis_c(
+    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
+  ) +
+  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr),2), " to ", round(max(snow$nao_effect_lwr),2))) +
+  theme(legend.position = "none",
+          plot.title = element_text(size = 10), axis.ticks = element_blank(),
+          axis.title = element_blank(), axis.text = element_blank())
+
+snow_g3 <- ggplot(data = nao_effect_df) +
+  geom_point(aes(X, Y, colour = nao_effect_upr)) +
+  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  coord_sf(
+    xlim = c(min(snow2$X), max(snow2$X)),
+    ylim = c(min(snow2$Y), max(snow2$Y))
+  ) +
+  scale_colour_viridis_c(
+    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
+  ) +
+  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr),2), " to ", round(max(snow$nao_effect_upr),2))) +
+  theme(legend.position = "none",
+          plot.title = element_text(size = 10), axis.ticks = element_blank(),
+          axis.title = element_blank(), axis.text = element_blank())
+
+bottom_row <- cowplot::plot_grid(snow_g2, snow_g3, label_size = 12)
+cowplot::plot_grid(snow_g1, bottom_row, ncol = 1, rel_heights = c(2.18, 1))
 @

### 2023-11-24 doc/paper-jss/sdmTMB-paper.Rnw
slight adjustment to owl fig titles

@@
   ) +
-  ggtitle("(a) Median") +
-  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
+  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect),2), " to ", round(max(snow$nao_effect),2))) +
+  labs(colour = "Estimated effect of NAO\non Snowy Owl count") +
   theme(legend.position = "top",

### 2023-11-24 doc/paper-jss/sdmTMB-paper.Rnw
fancier owl figure

--- 2023-11-24 doc/paper-jss/sdmTMB-paper.Rnw
@@
 
-We can make a basic plot using the following code. A more elaborate version is shown in Figure~\ref{fig:owl-plot-fancy}.
+We can make a basic plot using the following code. A more elaborate version including the CIs is shown in Figure~\ref{fig:owl-plot-fancy}a.
 
@@
 lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
+land <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
+land <- land %>% sf::st_transform(crs = Albers)
 @
@@
 
-<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
+<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficients for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
 The effect is multiplicative on owl count per NAO unit.
-In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, out.width="5.25in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
-nao_effect_df <- select(snow, X, Y, nao_effect, nao_effect_lwr, nao_effect_upr) |> distinct() |>
+In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.
+Point size is scaled to the mean counts in each location.", fig.width=5, out.width="5.25in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
+nao_effect_df <- select(snow, X, Y, count, nao_effect, nao_effect_lwr, nao_effect_upr) |>
+  group_by(X,Y) |> mutate(mean_count = mean(count, na.rm = TRUE)) |> select(-count) |>
+  distinct() |>
   mutate(X = X * 100000, Y = Y * 100000) # to match CRS
@@
 snow_g1 <- ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect)) +
-  geom_sf(data = coast, colour = "gray50") +
+  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
+  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_point(aes(X, Y, colour = nao_effect, size = mean_count), alpha = 0.5) +
+  geom_sf(data = coast, colour = "gray50", fill = NA) +
   geom_sf(data = lakes, colour = "gray50", fill = NA) +
@@
     limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
-    guide = guide_colourbar(direction = "horizontal", title.position = "left")
+    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
+    # guide = guide_colourbar(direction = "horizontal", title.position = "left")
   ) +
-  ggtitle("(a) Median") +
-  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
-  theme(legend.position = "top",
-          plot.title = element_text(size = 10, hjust = -0.2),
+  guides(size = "none") +
+  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect),2), " to ", round(max(snow$nao_effect),2))) +
+  labs(colour = "NAO effect\non Snowy Owl counts") +
+  theme_bw() + theme(
+    legend.position = c(0.25,0.17),
+    legend.title = element_text(size= 9, hjust = 0),
+    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
+    # legend.position = "top",
+        panel.background = element_rect(fill = "grey90", colour = NA),
+        panel.grid.major = element_line(colour = "white", size = 0.5),
+        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
+        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
+        plot.title = element_text(size = 10, hjust = -0.15
+                                  ),
         axis.title = element_blank())
@@
 snow_g2 <- ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect_lwr)) +
+  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
+  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_point(aes(X, Y, colour = nao_effect_lwr, size = mean_count), alpha = 0.5) +
   geom_sf(data = coast, colour = "gray50") +
@@
   ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr),2), " to ", round(max(snow$nao_effect_lwr),2))) +
-  theme(legend.position = "none",
+  theme_bw() + theme(legend.position = "none",
+    # plot.margin = unit(c(0, 0, 0, 0.5), "cm"),
+        panel.background = element_rect(fill = "grey90", colour = NA),
+        panel.grid.major = element_line(colour = "white", size = 0.5),
+        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
+        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
           plot.title = element_text(size = 10), axis.ticks = element_blank(),
@@
 snow_g3 <- ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect_upr)) +
+  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
+  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_point(aes(X, Y, colour = nao_effect_upr, size = mean_count), alpha = 0.5) +
   geom_sf(data = coast, colour = "gray50") +
@@
   ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr),2), " to ", round(max(snow$nao_effect_upr),2))) +
-  theme(legend.position = "none",
+  theme_bw() + theme(legend.position = "none",
+    # plot.margin = unit(c(0, 0, 0, 0.5), "cm"),
+        panel.background = element_rect(fill = "grey90", colour = NA),
+        panel.grid.major = element_line(colour = "white", size = 0.5),
+        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
+        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
           plot.title = element_text(size = 10), axis.ticks = element_blank(),

### 2023-11-25 doc/paper-jss/sdmTMB-paper.Rnw
add count to owl data head
edit owl text and shrink fig

--- 2023-11-24 doc/paper-jss/sdmTMB-paper.Rnw
@@
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe \citep{millon2014}.
-Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on the end of year's abundance across the southern boundary of their winter distribution.
+Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on early winter abundance across the southern boundary of their winter distribution.
 Annual mean NAO captures the preceding winter's conditions, combined with breeding season and early winter climate.
 
-Our data are read in from supplementary data and contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:
+To do this we use counts of Snowy Owls observed by the annual Christmas Bird Counts \citep{cbc} from all locations where they have been recorded and for which there was at least 3 counts conducted between the years 1979 and 2020.
+Our data are read in from supplementary data and also contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:
 
@@
 <<owl-data-head, echo=TRUE>>=
-select(snow, X, Y, year, year_f, nao)
+select(snow, X, Y, year, year_f, nao, count) |> head(n = 5)
 @
@@
 
-We will fit counts observed in North America during annual Christmas Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+We will fit counts using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
 
@@
 
-\subsection{Extracting the spatially varying coefficient effects}
+\subsection{Extracting the spatially varying coefficient effects with uncertainty}
 
 The spatially varying coefficient effect at any point in space is the combination of the main effect and spatially varying random effect for \code{nao}.
-Mean estimates of the spatially varying random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (here, \code{zeta\_s\_nao}).
-However, we wish to combine the fixed and random effects with uncertainty on the combined predictions.
+Mean estimates of the spatially varying random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case would appear as \code{zeta\_s\_nao}).
+However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
 We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
 We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
-By default, \code{nsim > 0} will return a matrix of draws from the overall prediction.
-Here, we instead specify that we want to return draws from the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the spatially varying random field.
+By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
+Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the spatially varying random field.
 This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
-We then use the function \fct{spread\_sims} to draw 200 simulations from the parameters themselves, combine the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}, and calculate the median, and upper and lower quantiles for each column of data so we can plot them.
-For this example we use thresholds of 0.1 and 0.9 representing confidence invervals (CI) of 80\%.
-
-<<p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
+We then use the function \fct{spread\_sims} to draw 200 simulations for the parameters themselves.
+Because the simulations are stored in different dimensions for these two functions, the random field draws must be transposed \code{t()} before combining the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
+Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
+For this example we use thresholds of 0.1 and 0.9 representing confidence intervals (CI) of 80\%.
+
+<<owl-p, message=FALSE, echo=FALSE, eval=FALSE, cache=TRUE>>=
+# not currently including, but might be helpful
+snow <- predict(fit_owl, newdata = snow)
+snow |> select(X, Y, year, nao, count, est, omega_s, epsilon_st, zeta_s_nao) |>
+  head(n = 2)
+@
+
+<<zeta-effect, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
 zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
@@
 sims <- spread_sims(fit_owl, nsim = 200)
+dim(sims)
 combined <- sims$nao + t(zeta_s)
@@
 
-We can make a basic plot using the following code. A more elaborate version is shown in Figure~\ref{fig:owl-plot-fancy}.
+We can make a basic plot using the following code. A more elaborate version including separate panels for each of the CIs is shown in Figure~\ref{fig:owl-plot-fancy}a.
 
@@
 
-<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
+<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficients for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
 The effect is multiplicative on owl count per NAO unit.
-In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, out.width="5.25in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
-nao_effect_df <- select(snow, X, Y, nao_effect, nao_effect_lwr, nao_effect_upr) |> distinct() |>
-  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
+In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.
+Point size is scaled to the mean counts in each location.", fig.width=5, out.width="4in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
+nao_effect_df <- select(snow2, X, Y, count, nao_effect, nao_effect_lwr, nao_effect_upr) |>
+  group_by(X,Y) |> summarise_all(mean)
 
 snow_g1 <- ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect)) +
-  geom_sf(data = coast, colour = "gray50") +
-  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
+  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
+  geom_sf(data = coast, colour = "gray50", fill = NA, lwd = 0.35) +
+  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
   coord_sf(
@@
 snow_g2 <- ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect_lwr)) +
-  geom_sf(data = coast, colour = "gray50") +
-  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
+  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_point(aes(X, Y, colour = nao_effect_lwr, size = count), alpha = 0.5) +
+  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
+  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
   coord_sf(
@@
 snow_g3 <- ggplot(data = nao_effect_df) +
-  geom_point(aes(X, Y, colour = nao_effect_upr)) +
-  geom_sf(data = coast, colour = "gray50") +
-  geom_sf(data = lakes, colour = "gray50", fill = NA) +
+  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
+  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_point(aes(X, Y, colour = nao_effect_upr, size = count), alpha = 0.5) +
+  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
+  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
   coord_sf(

### 2023-11-28 doc/paper-jss/sdmTMB-paper.Rnw
add check on CI for main effect of NAO

@@
 Given our model specification, all random fields are sharing a single Matérn range.
+We can also check the confidence intervals (CIs) on the main effect of NAO and see that they overlap zero.
+
+<<owl-tidy, echo=TRUE>>=
+tidy(fit_owl, conf.int = TRUE)
+@
+
+But given that this is only part of the effect of NAO that we include in our model, how do we assess if this variable has a meaningful effect in some locations, even if not overall?
 
@@
 Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
-For this example we use thresholds of 0.1 and 0.9 representing confidence intervals (CI) of 80\%.
+For this example we use thresholds of 0.1 and 0.9 representing CIs of 80\%.
 

### 2023-11-28 doc/paper-jss/sdmTMB-paper.Rnw
add note on centring SVCs
use SVC shorthand in more places

@@
 
-We will fit counts using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.
+We will fit counts using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a SVC associated with the NAO.
+Centring and scaling variables is especially helpful when estimating SVCs.
+NAO is an index so it can be left as is, but including it as a main effect as well helps ensure that the SVCs will have a mean near 0.
 
@@
 
-In addition to the output seen for other models, we now have a section for random intercepts and a standard deviation for the spatially varying coefficient random field.
+In addition to the output seen for other models, we now have a section for random intercepts and a standard deviation for the SVC random field.
 Given our model specification, all random fields are sharing a single Matérn range.
@@
 
-\subsection{Extracting the spatially varying coefficient effects with uncertainty}
-
-The spatially varying coefficient effect at any point in space is the combination of the main effect and spatially varying random effect for \code{nao}.
-Mean estimates of the spatially varying random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case would appear as \code{zeta\_s\_nao}).
+\subsection{Extracting the spatially varying effects with uncertainty}
+
+The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
+Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case would appear as \code{zeta\_s\_nao}).
 However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
@@
 By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
-Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the spatially varying random field.
+Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field.
 This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.

### 2023-11-28 doc/paper-jss/sdmTMB-paper.Rnw
more shortening of SVC language

@@
 In our final example, we demonstrate a model with spatially varying coefficient (SVC) effects and illustrate combining uncertainty from parameters by working with draws from the joint parameter precision matrix.
-Spatially varying coefficient models \citep{hastie1993} are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function.
+SVC models \citep{hastie1993} are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function.
 In addition, we illustrate using restricted maximum likelihood (REML), which integrates over the fixed effects as well as the random effects.
@@
 <<owl-data-head, echo=TRUE>>=
-select(snow, X, Y, year, year_f, nao, count) |> head(n = 5)
+select(snow, X, Y, year, year_f, nao, count) |> head(n = 6)
 @
@@
 
-<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficients for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
+<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
 The effect is multiplicative on owl count per NAO unit.

### 2023-12-12 doc/paper-jss/sdmTMB-paper.Rnw
Editing owls section

@@
 In our final example, we demonstrate a model with spatially varying coefficient (SVC) effects and illustrate combining uncertainty from parameters by working with draws from the joint parameter precision matrix.
-SVC models \citep{hastie1993} are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function.
+SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993}.
 In addition, we illustrate using restricted maximum likelihood (REML), which integrates over the fixed effects as well as the random effects.
@@
 The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe \citep{millon2014}.
-Because both productivity and the choice of wintering location could be influenced by climate, we modelled a spatially varying effect of annual mean NAO index on early winter abundance across the southern boundary of their winter distribution.
+Because both productivity and the choice of wintering location could be influenced by climate, we model an SVC effect of annual mean NAO index on early winter abundance across the southern boundary of their winter distribution.
 Annual mean NAO captures the preceding winter's conditions, combined with breeding season and early winter climate.
 
-To do this we use counts of Snowy Owls observed by the annual Christmas Bird Counts \citep{cbc} from all locations where they have been recorded and for which there was at least 3 counts conducted between the years 1979 and 2020.
-Our data are read in from supplementary data and also contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:
+To do this, we use counts of Snowy Owls observed by the annual Christmas Bird Counts \citep{cbc} from all locations where they have been recorded and for which there were at least three counts conducted between 1979 and 2020.
+Our data are read from supplementary data and also contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, the annual NAO value, and own count:
 
@@
 <<owl-data-head, echo=TRUE>>=
-select(snow, X, Y, year, year_f, nao, count) |> head(n = 6)
+select(snow, X, Y, year, year_f, nao, count) |> head()
 @
@@
 
-We will fit counts using a negative binomial \citep[NB2,][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and a SVC associated with the NAO.
-Centring and scaling variables is especially helpful when estimating SVCs.
-NAO is an index so it can be left as is, but including it as a main effect as well helps ensure that the SVCs will have a mean near 0.
+We will fit counts using a negative binomial \citep[NB2;][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and an SVC associated with the NAO.
+Centering and scaling variables (say by there mean and standard deviation) is especially helpful when estimating SVCs.
+Here, NAO is an index with a mean near zero and an SD not too far from 1, so we will leave it as is.
+However, we also include NAO as a main effect since the SVC random field is assumed to have a mean of zero.
 
@@
 
-But given that this is only part of the effect of NAO that we include in our model, how do we assess if this variable has a meaningful effect in some locations, even if not overall?
+However, given that this is only part of the effect of NAO that we include in our model, we next assess if this variable has a meaningful effect in some locations, even if not overall.
 
@@
 The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
-Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case would appear as \code{zeta\_s\_nao}).
+Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
 However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
@@
 We then use the function \fct{spread\_sims} to draw 200 simulations for the parameters themselves.
-Because the simulations are stored in different dimensions for these two functions, the random field draws must be transposed \code{t()} before combining the main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
+Because the simulations are stored in different dimensions, the random field draws must be transposed \code{t()} before combining the vector of main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
 Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
-For this example we use thresholds of 0.1 and 0.9 representing CIs of 80\%.
+For this example, we use thresholds of 0.1 and 0.9 representing CIs of 80\%.
 
@@
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
-However, if we recalculated the CI at 95\% level the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
+However, if we recalculated the CI at 95\% level the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.2, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
 

### 2023-12-12 doc/paper-jss/sdmTMB-paper.Rnw
Discussion edits

@@
 We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
-In this test, \sdmTMB\ was the fastest across all mesh resolutions; however, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB.
-\pkg{mgcv} was most affected by mesh resolution.
 Our test was restricted to one core and default \R\ algebra libraries.
+\sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
+These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
+However, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB and \pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing}).
+For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
 
@@
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
+Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
+% TODO: mention this in response doc
+
+% \textbf{TODO add paragraph here on uniting functionality not available in any one package?}
 
@@
 Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
+The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
 These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.
-
-\textbf{TODO add paragraph here on uniting functionality not available in any one package?}
 
@@
 
-With respect to computational speed, \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
-These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
-
-Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, and time-varying spatiotemporal covariance parameters \citep{ward2022}.
-
+Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and integration with visualization tools such as \pkg{ggeffects} \citep{ggeffects}.
 There are several planned future additions to the \sdmTMB\ model structure.
 A subset of features to be added include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
-(4) the ability to specify observation-level likelihoods to integrate different data types (e.g. REF),
-(5) conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature, and
-(6) improved integration with existing visualization methods such as \pkg{ggeffects} \citep{ggeffects}.
+(4) the ability to specify observation-level likelihoods to integrate different data types, and
+(5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
 

### 2023-12-12 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
+% A19
+Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
+bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
+Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristkristensen2016, zheng2021}.
 After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
 
-\textbf{TODO: generalized delta method! and \citep{zheng2021}}
+% \textbf{TODO: generalized delta method! and \citep{zheng2021}}
 
@@
 The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
+The overall predictions combining fixed and random effects are plug-in predictors combining
+
 We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
@@
 
-\subsection{Calculating an area-weighted index}
+\subsection{Calculating an area-weighted index} \label{index-bias-correct}
 

### 2023-12-13 doc/paper-jss/sdmTMB-paper.Rnw
typo

@@
 To do this, we use counts of Snowy Owls observed by the annual Christmas Bird Counts \citep{cbc} from all locations where they have been recorded and for which there were at least three counts conducted between 1979 and 2020.
-Our data are read from supplementary data and also contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, the annual NAO value, and own count:
+Our data are read from supplementary data and also contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, the annual NAO value, and owl count:
 

### 2023-12-14 doc/paper-jss/sdmTMB-paper.Rnw
Copy editing; add JT

@@
 % setwd("doc/paper-jss/")
-% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex")
+% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)
 
@@
 Sean C. Anderson~\orcidlink{0000-0001-9563-1937}\\Pacific Biological
-Station \And Eric J. Ward~\orcidlink{0000-0002-4359-0296}\\Northwest
-Fisheries Science Center \AND Philina A.
-English~\orcidlink{0000-0003-2992-6782}\\Pacific Biological
-Station \And Lewis A. K. Barnett~\orcidlink{0000-0002-9381-8375}\\Alaska
-Fisheries Science Center
+Station \And
+Eric J. Ward~\orcidlink{0000-0002-4359-0296}\\Northwest
+Fisheries\\Science Center \And
+Philina A. English~\orcidlink{0000-0003-2992-6782}\\Pacific Biological
+Station \AND
+Lewis A. K. Barnett~\orcidlink{0000-0002-9381-8375}\\Alaska
+Fisheries Science Center \And
+James T. Thorson~\orcidlink{0000-0001-7415-1010}\\Alaska
+Fisheries Science Center\\
 }
@@
 \Plainauthor{Sean C. Anderson, Eric J. Ward, Philina A. English, Lewis
-A. K. Barnett}
+A. K. Barnett, James T. Thorson}
 \Plaintitle{sdmTMB: An R Package for Fast, Flexible, and Accessible
@@
 \Abstract{
-Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
-However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
-Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent Gaussian Markov random fields using an SPDE-(stochastic partial differential equation) based approach.
-SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
-We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
-Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
+Geostatistical spatial or spatiotemporal data are common across scientific fields.
+% and the dimensions of these data continue to grow in space and time.
+However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
+Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
+SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
+We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
+% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
 Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
@@
 
-\Keywords{Gaussian Markov random fields (GMRF), generalized linear mixed effects
-models (GLMM), INLA, SPDE, species distribution modelling, R package,
-spatio-temporal, spatial-temporal, Template Model Builder}
-
-\Plainkeywords{Gaussian Markov random fields (GMRF), generalized linear mixed
-effects models (GLMM), INLA, SPDE, species distribution modelling, R package,
-spatio-temporal, spatial-temporal, Template Model Builder}
+\Keywords{Gaussian Markov random field (GMRF), GLMM, SPDE, species distribution modelling, R package, spatial-temporal, Template Model Builder (TMB)}
+
+\Plainkeywords{Gaussian Markov random field (GMRF), GLMM, SPDE, species distribution modelling, R package, spatial-temporal, Template Model Builder (TMB)}
 
@@
     Alaska Fisheries Science Center\\
-    National Marine Fisheries Service, National Oceanic and Atmospheric
-Administration, Seattle, WA, USA\\
-
+    National Marine Fisheries Service, National Oceanic and
+    Atmospheric Administration, Seattle, WA, USA\\
+
+    James T. Thorson\\
+    Alaska Fisheries Science Center\\
+    National Marine Fisheries Service, National Oceanic and
+    Atmospheric Administration, Seattle, WA, USA
   }
@@
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
-A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
+A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
 % TODO: decide on using GRF throughout
@@
 This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
-The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
+The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2015ices, thorson2019, osgood-zimmerman2023}.
 
@@
 While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
-Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
-
-Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \INLA\ to set up SPDE matrices.
+Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, lack some functionality, or may be challenging to learn for some users.
+
+Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
@@
 
-A basic GLMM with spatial Gaussian random fields can be written as
+A GLMM with spatial Gaussian random fields can be written as
 \[
@@
 \]
-where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
-That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
+where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_{\bm{s}}$.
+That mean $\mu_{\bm{s}}$ is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta_{\bm{s}}$.
 In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
@@
 In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
-A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
+A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
 A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
@@
 \]
-Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices, thorson2018comp}.
+Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices}.
 
@@
 Instead, we refer to the following literature:
-\citet{lindgren2011} develops the approach.
+\citet{lindgren2011} develop the approach.
 \citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
 The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
-\citet{miller2019} contains an introduction to the SPDE and illustrates its equivalence to penalized smoothing approaches.
-\citet{lindgren2022ten} provides a recent review of the approach and its applications over the last decade.
+\citet{miller2019} summarizes the approach and illustrates its equivalence to penalized smoothing approaches.
+\citet{lindgren2022ten} provide a recent review of the approach and its applications over the last decade.
 
@@
 First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
-This means that GRF models can be computationally fit as GMRFs.
+This means that GRF models can be computationally approximated as GMRFs.
 By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
+where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
 We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
@@
 bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
-Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristkristensen2016, zheng2021}.
-After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
+Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
+After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.
 
@@
 Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
+In this paragraph, we include relevant \fct{predict} arguments in parentheses.
 The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
@@
 whether to return standard errors (\code{se\_fit},
-can be slow if conditioned on random fields),
+which can be slow if conditioned on random fields),
 whether to condition on the random fields (\code{re\_form}),
@@
 These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
-Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
+Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
 These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
@@
 
-\sdmTMB\ can be installed from the Comprehensive \R\ Archive Network (CRAN) at
+\sdmTMB\ can be installed from the Comprehensive \R\ Archive Network (CRAN) at\\
 \url{https://CRAN.R-project.org/package=sdmTMB}
@@
 
-Additional utilities that require heavier package dependencies (such as \INLA\ and \pkg{rstan}) are maintained in the \pkg{sdmTMBextra} package at\\
+Additional utilities, which require heavier package dependencies (such as \INLA\ and \pkg{rstan}) and are used by only a subset of users, are maintained in the \pkg{sdmTMBextra} package at\\
 \url{https://github.com/pbs-assess/sdmTMBextra}
@@
 The argument \code{cutoff} defines the minimum allowed distance between mesh vertices in the units of \code{X} and \code{Y} (km).
-We could create a basic mesh just specifying this:
+We could create a basic mesh specifying this:
 
@@
 For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
-The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).
+The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range to avoid edge effects.
+See \citet{krainski2018} chapters 2.6 and 2.7 for guidance on mesh construction.
 
@@
 Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
-Mesh complexity has the single largest influence on the speed of fitting these models.
+Usually, mesh complexity has the single largest influence on the speed of fitting these models.
 
@@
 <<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
-The locations where lines intersect are referred to as ``vertices''.
-Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
+The locations where lines intersect are referred to as ``vertices'' or sometimes ``knots''.
+Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation to a point.
 A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
@@
 In addition to the spatial random field, we include an intercept and a quadratic effect of depth on the probability of encounter.
-Our random field model can be written as
+Our model can be written as
 \[
@@
 that no extreme or small eigenvalues are detected,
-that no absolute log likelihood gradients with respect to fixed defects are $\le$ 0.001,
+that no absolute log likelihood gradients with respect to fixed effects are $\ge$ 0.001,
 that all fixed effects have reported standard errors that do not look unreasonably large ($<$ 100 by default),
-that random field marginal standard deviations are not unexpectedly small or large ($>$ 0.01 and $<$ 100),
-and that the random field Matérn range parameter does not look unreasonably large ($<$ 1.5 times the largest distance from a bounding box around the observations).
+that random field marginal standard deviations are not unexpectedly small ($<$ 0.01) or large ($>$ 100),
+and that the random field Matérn range parameter does not look unreasonably large ($\ge$ 1.5 times the largest distance from a bounding box around the observations).
 
@@
 We can do additional model comparison with cross validation via \fct{sdmTMB\_cv}.
-We will do 10-fold cross validation with the folds constructed randomly.
+We will do 10-fold cross validation with the folds constructed randomly. We will set the seed each time to ensure the folds are consistent.
 Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation.
@@
 <<pcod-cv, eval=TRUE, echo=TRUE>>=
+set.seed(12928)
 cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
@@
 )
+set.seed(12928)
 cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
@@
 
+% TODO: data citation!?
 \sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
@@
 The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
-The overall predictions combining fixed and random effects are plug-in predictors combining
-
+% The overall predictions combining fixed and random effects are plug-in predictions.
 We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
@@
 \]
-where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) (representing latent spatial effects that vary by year).
+where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
 The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
@@
 Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
-Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
+Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
 
@@
 
-We find that the Poisson-link delta-lognormal model is favoured by AIC. In an applied situation we would inspect the distribution of the residuals and consider comparing models with cross validation.
+We find that the Poisson-link delta-lognormal model \citep{thorson2018poisson} is favoured by conditional AIC.
+In an applied situation, we would inspect the distribution of the residuals and consider comparing models with cross validation.
 
@@
 <<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
-Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled.
+Ellipses are centered at coordinates of zero in the units that the X-Y coordinates are modeled.
 The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
@@
 
-Our prediction data frame is similar to the binomial spatial model, but includes columns for the two delta model components (labelled with suffixes \code{1} and \code{2}) and adds an \code{epsilon\_st} column for spatiotemporal random effects.
-We can easily generate plots from this data frame using \pkg{ggplot2} code with \fct{geom\_raster} similar to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
+Our prediction data frame is similar to the binomial spatial model, but includes columns for the two delta model linear predictors (labelled with suffixes \code{1} and \code{2}) and adds an \code{epsilon\_st} column for spatiotemporal random effects.
+We can easily generate plots from this data frame using \pkg{ggplot2} code with \fct{geom\_raster} similarly to our spatial example with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}).
 We suppress that code for brevity.
@@
 The spatial random field is constant through time (i.e., the two panels in b are identical) and represents static biotic or abiotic features not included as covariates (e.g., habitat).
-The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process.
+The spatiotemporal random fields are different each time step and here are constrained to follow an AR(1) process.
 They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).", fig.width=9, out.width="6in">>=
@@
 Here, we pick the last year, specify to include both delta model components (\code{model = NA}), omit the random fields (\code{re\_form = NA}), and return standard errors (\code{se\_fit\ = TRUE}).
+Alternatively, we could produce a similar plot using \pkg{ggeffects} \citep{ggeffects} with \fct{ggeffects::ggpredict}.
+
+% TODO check if on CRAN yet!
 
@@
 In our final example, we demonstrate a model with spatially varying coefficient (SVC) effects and illustrate combining uncertainty from parameters by working with draws from the joint parameter precision matrix.
-SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993}.
-In addition, we illustrate using restricted maximum likelihood (REML), which integrates over the fixed effects as well as the random effects.
+SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993, thorson2023SVC}.
+In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.
 
@@
 We will fit counts using a negative binomial \citep[NB2;][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and an SVC associated with the NAO.
-Centering and scaling variables (say by there mean and standard deviation) is especially helpful when estimating SVCs.
+Centering and scaling variables (by their mean and standard deviation here) can be especially helpful when estimating SVCs.
 Here, NAO is an index with a mean near zero and an SD not too far from 1, so we will leave it as is.
-However, we also include NAO as a main effect since the SVC random field is assumed to have a mean of zero.
+However, we also include NAO as a main effect since the SVC random field is drawn from a multivariate normal distribution with mean zero.
 
@@
 Overall, we find a weak average positive effect of annual mean NAO on overall counts with a southeast to northwest gradient in the intensity of the effect (Figure~\ref{fig:owl-plot-fancy}).
+At some locations, the lower CI on the exponentiated effect is above 1.
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
-However, if we recalculated the CI at 95\% level the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.2, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
+However, if we recalculated the CI at a 95\% level (not shown) the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
 
@@
 nao_effect_df <- select(snow2, X, Y, count, nao_effect, nao_effect_lwr, nao_effect_upr) |>
-  group_by(X,Y) |> summarise_all(mean)
+  group_by(X, Y) |>
+  summarise_all(mean)
 
 snow_g1 <- ggplot(data = nao_effect_df) +
-  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
-  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
+  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
   geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
@@
     guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
-    # guide = guide_colourbar(direction = "horizontal", title.position = "left")
   ) +
   guides(size = "none") +
-  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect),2), " to ", round(max(snow$nao_effect),2))) +
+  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect), 2), " to ", round(max(snow$nao_effect), 2))) +
   labs(colour = "NAO effect\non Snowy Owl counts") +
-  theme_bw() + theme(
-    legend.position = c(0.25,0.17),
-    legend.title = element_text(size= 9, hjust = 0),
+  theme_bw() +
+  theme(
+    legend.position = c(0.25, 0.17),
+    legend.title = element_text(size = 9, hjust = 0),
     plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
-    # legend.position = "top",
     panel.background = element_rect(fill = "grey90", colour = NA),
     panel.grid.major = element_line(colour = "white", size = 0.5),
-        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
-        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
-        plot.title = element_text(size = 10, hjust = -0.15
-                                  ),
-        axis.title = element_blank())
+    legend.background = element_rect(fill = "transparent", colour = NA),
+    legend.box.background = element_rect(fill = "transparent", colour = NA),
+    plot.title = element_text(size = 10, hjust = -0.15),
+    axis.title = element_blank()
+  )
 
 snow_g2 <- ggplot(data = nao_effect_df) +
-  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
-  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
+  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
   geom_point(aes(X, Y, colour = nao_effect_lwr, size = count), alpha = 0.5) +
@@
   ) +
-  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr),2), " to ", round(max(snow$nao_effect_lwr),2))) +
-  theme_bw() + theme(legend.position = "none",
-    # plot.margin = unit(c(0, 0, 0, 0.5), "cm"),
+  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
+  theme_bw() +
+  theme(
+    legend.position = "none",
     panel.background = element_rect(fill = "grey90", colour = NA),
     panel.grid.major = element_line(colour = "white", size = 0.5),
-        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
-        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
+    legend.background = element_rect(fill = "transparent", colour = NA),
+    legend.box.background = element_rect(fill = "transparent", colour = NA),
     plot.title = element_text(size = 10), axis.ticks = element_blank(),
-          axis.title = element_blank(), axis.text = element_blank())
+    axis.title = element_blank(), axis.text = element_blank()
+  )
 
 snow_g3 <- ggplot(data = nao_effect_df) +
-  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
-  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
+  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
+  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
   geom_point(aes(X, Y, colour = nao_effect_upr, size = count), alpha = 0.5) +
@@
   ) +
-  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr),2), " to ", round(max(snow$nao_effect_upr),2))) +
-  theme_bw() + theme(legend.position = "none",
-    # plot.margin = unit(c(0, 0, 0, 0.5), "cm"),
+  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
+  theme_bw() +
+  theme(
+    legend.position = "none",
     panel.background = element_rect(fill = "grey90", colour = NA),
     panel.grid.major = element_line(colour = "white", size = 0.5),
-        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
-        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
+    legend.background = element_rect(fill = "transparent", colour = NA),
+    legend.box.background = element_rect(fill = "transparent", colour = NA),
     plot.title = element_text(size = 10), axis.ticks = element_blank(),
-          axis.title = element_blank(), axis.text = element_blank())
+    axis.title = element_blank(), axis.text = element_blank()
+  )
 
@@
 We thank the authors of all these packages.
-We thank S.
-Kotwicki, M.
-Lindmark, M.
-Martin, C.C.
-Monnahan, P.M.
-Regular, J.T.
-Thorson, J.
-Watson, two anonymous reviewers, and the editor for helpful comments that substantially improved the manuscript.
+We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J. Watson, two anonymous reviewers, and the editor for helpful comments that substantially improved the manuscript.
 Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.

### 2023-12-14 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 \sdmTMB, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \INLA/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
-\sdmTMB\ and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
+\pkg{mgcv} and \sdmTMB\ (currently) focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
 To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
@@
 These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-However, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB and \pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing}).
+However, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB\ and \pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing}).
 For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
@@
 The GLMMs underpinning \sdmTMB\ models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
+In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions [e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}].
 % The random fields in \sdmTMB\ are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
@@
 Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids.
-Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
+Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including CAR models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
 A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004}.
 
-Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and integration with visualization tools such as \pkg{ggeffects} \citep{ggeffects}.
+Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, and time-varying spatiotemporal covariance parameters \citep{ward2022}.
 There are several planned future additions to the \sdmTMB\ model structure.
 A subset of features to be added include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
-(4) the ability to specify observation-level likelihoods to integrate different data types, and
+(4) the ability to specify observation-specific likelihoods to integrate different data types \citep{gruss2019multiple}, and
 (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
@@
 We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J. Watson, two anonymous reviewers, and the editor for helpful comments that substantially improved the manuscript.
-Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.
+We thank the many individuals who have contributed to collecting the trawl survey data with Fisheries and Oceans Canada that was used in the Pacific Cod and Pacific Spiny Dogfish examples.
+Christmas Bird Count data for the Snowy Owl example were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.
 

### 2023-12-14 doc/paper-jss/sdmTMB-paper.Rnw
Delete empty line

@@
 dat$log_depth <- log(dat$depth)
-
 mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)

### 2023-12-15 doc/paper-jss/sdmTMB-paper.Rnw
Add linenumbers + response input

@@
 \newcommand{\R}{\proglang{R}}
+
+% TODO START FOR REVIEW ONLY
+\usepackage{lineno}
+\definecolor{darkgrey}{HTML}{A9A9A9}
+\renewcommand\linenumberfont{\normalfont\bfseries\small\color{darkgrey}}
+\modulolinenumbers[2]
+\usepackage[dvipsnames]{xcolor}
+\definecolor{niceblue}{HTML}{236899}
+\newcommand{\rev}[1]{{\color{niceblue} #1}}
+\newcommand{\Rev}[1]{\label{#1}\linelabel{#1}}
+\newcommand{\lr}[1]{line~\lineref{#1}}
+% TODO END FOR REVIEW ONLY
 
@@
     Fisheries Service, Seattle, WA, USA\\
+    E-mail: \email{eric.ward@noaa.gov}\\
 
@@
     Fisheries and Oceans Canada, Nanaimo, BC, Canada\\
+    E-mail: \email{philina.english@dfo-mpo.gc.ca}\\
 
@@
     Atmospheric Administration, Seattle, WA, USA\\
+    E-mail: \email{lewis.barnett@noaa.gov}\\
 
@@
     National Marine Fisheries Service, National Oceanic and
-    Atmospheric Administration, Seattle, WA, USA
+    Atmospheric Administration, Seattle, WA, USA\\
+    E-mail: \email{james.thorson@noaa.gov}\\
   }
@@
 \begin{document}
+
+\linenumbers
 
@@
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
-Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
+\Rev{B2}Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
 One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
@@
 
-In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
+\Rev{B4a}In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
 and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
@@
 
-For a user of \sdmTMB, the following are the important elements to understand.
+\Rev{B4b}For a user of \sdmTMB, the following are the important elements to understand.
 First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
@@
 The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
-% A19
-Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
+\Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
 bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
@@
 Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
-These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
+\Rev{A4}These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
 If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
@@
 We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
-We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
+\Rev{A19b}We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
 
@@
 
-The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
+\Rev{B7}The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
 Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
@@
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
-Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
+\Rev{B3}Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
 % TODO: mention this in response doc
@@
 While our examples focused on applications to ecological data, the SPDE approach and the functionality of \sdmTMB\ has applications in many other fields.
-Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
-The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
+\Rev{B9a}Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
+\Rev{B9b}The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
 These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.
@@
 
-<<inlabrufit-timing, echo=TRUE, cache=TRUE>>=
-# convert to sp SpatialPointsDataFrame first:
+% weird eval errors as of 2023-12-14 when in Rnw, hence eval=FALSE
+
+<<inlabrufit-timing, echo=TRUE, cache=TRUE, warning=FALSE, eval=FALSE>>=
 dat_sp <- sp::SpatialPointsDataFrame(
@@
 
-\clearpage
+% \clearpage
+
+% \input{response}
 

### 2024-01-16 doc/paper-jss/sdmTMB-paper.Rnw

@@
 
-Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
+Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation when evaluating the multivariate normal density function.
 Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
@@
 The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2015ices, thorson2019, osgood-zimmerman2023}.
+This SPDE method approximates a Matérn correlation function as arising mechanistically from local diffusion in space and/or time \citep{lindgren2011} and it results in a sparse precision matrix that permits efficient computation using existing sparse-matrix libraries \citep{rue2005gmrf}.
 
@@
 \Rev{B2}Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
-One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.
 
@@
 This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
-Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
+Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise arising from unmodeled processes.
+% Jim: Similar to the recent Ives MEE perspectives, I prefer to discuss unmodeled stochastic processes, rather than the term "error" or "noise", so suggesting a compromise wording here.
 More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
@@
 \Rev{B4a}In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
-and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
-One such approach is the SPDE approach.
+and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{rue2005gmrf,simpson2012}.
+One such approach is the SPDE approach, which approximates a mechanistic process of local diffusion using methods drawn from finite-element analysis.
 A full description of the SPDE approach is beyond the scope of this paper.
@@
 \fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
-The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in C++.
+% Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
+_Matrix: Sparse and Dense Matrix Classes and Methods_.
+R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack.
+If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al.
+(2010).
+“Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
 The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
 \Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
-bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
+it is also possible to apply the epsilon estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
 Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
@@
 % using the Laplace approximation \citep{thygesen2017a}. We provide several
-% approaches to assist this process: (1) The Akaike Information Criterion
-% \citep[AIC,][]{akaike1974} can be calculated with \fct{AIC}, although AIC
+% approaches to assist this process: (1) The (marginal) Akaike Information Criterion
+% \citep[AIC,][]{akaike1974} can be calculated from the marginal likelihood with \fct{AIC}, although AIC
 % has well-documented biases with mixed-effects models \citep{liang2008}. (2)
@@
 
+% Jim:  If this section is re-added, you could also list the \fct{TMB::checkConsistency}
+% which evaluates the accuracy (and provides a bias-correction estimator) for the 
+% Laplace approximation.
+
 \subsection{Installation}\label{installation}
@@
 
-We can also see that the conditional AIC of the model with spatial random fields is
+We can also see that the marginal AIC of the model with spatial random fields is
 lower:
+% JIM: Please feel free to email me, but if you're using the marginal likelihood to compute AIC then it's using marginal AIC.
+I have a paper in revision @ Ecology discussing TMB calculation for conditional AIC, and happy to share or discuss further.
 
@@
 The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
+Some covariates might be fixed at a specified value for all predictions, such that we are predicting the expected value for samples conditional on those specified values.
+We call these covariates \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates that was estimated in fitted data \cite{thorson2019}.
 The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
 % The overall predictions combining fixed and random effects are plug-in predictions.
+% Jim:  I'm adding a sentence here about catchability/detectability covariates.
+I'd be happy to discuss if y'all disagree with including this.
+Obviously I'm concerned about people moving back to the fisheries-standard of not carefully considering what is a catchability vs.
+density covariate, and would love to preserve this as distinction that is highlighted (even if it's not explicit in the model structure itself).
 We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
@@
 
-We find that the Poisson-link delta-lognormal model \citep{thorson2018poisson} is favoured by conditional AIC.
+We find that the Poisson-link delta-lognormal model \citep{thorson2018poisson} is favoured by marginal AIC.
 In an applied situation, we would inspect the distribution of the residuals and consider comparing models with cross validation.
@@
 where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields each year.
+This is equivalent to a separable model GMRF with precision arising from the kronecker product of the spatial precision and an AR1 temporal precision.
+%% Jim:  Feel free to drop the last sentence, but I think statisticians would wanna emphasize that this is a separable space-time formulation.
 
@@
 % TODO check if on CRAN yet!
+% Jim:  It might be worth emphasizing that this calculation of marginal effects will differ from a partial dependence plot (PDP) when depth is correlated with other variables, and that package pdp could be used to additionally compute the PDP.
+Please tell me if you want me to suggest text or a code-demo, etc.
 
@@
 We will fit counts using a negative binomial \citep[NB2;][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and an SVC associated with the NAO.
-Centering and scaling variables (by their mean and standard deviation here) can be especially helpful when estimating SVCs.
+Centering and scaling variables (by their mean and standard deviation here) can be especially helpful to reduce the correlation between the SVC and other random fields that are included in the model when estimating SVCs.
 Here, NAO is an index with a mean near zero and an SD not too far from 1, so we will leave it as is.
@@
 However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
-We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
+We illustrate a way of accomplishing this by simulating from uncertainty about fixed and random effects while assuming that parametric uncertainty is well approximated using a multivariate normal distribution and the joint precision matrix.
+% Jim:  I found the previous wording a bit confusing, so suggested an edit.
 We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
@@
 Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
-\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \sdmTMB\ more straightforward.
-Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
+\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts will likely find \sdmTMB\ more straightforward.
+Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
 \Rev{B3}Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
@@
 % TODO: rework this paragraph to be advantages of random fields GAMs and CAR
-The GLMMs underpinning \sdmTMB\ models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions [e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}].
+The GLMMs underpinning \sdmTMB\ models are spatially explicit and derived from a mechanistic (diffusion) process---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+In contrast, non-parametric approaches do not estimate spatial covariance functions [e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}].
+%  I think semi-parametric is defined differently in different contexts, but don't agree with including it in that claim so removing it rather than adding clarifying text.
 % The random fields in \sdmTMB\ are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
@@
 % TODO: figure out where to put this paragraph
-Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids.
+Approaches such as conditional autoregressive models (CAR) are applicable to areal data, where the spatial domain is discretized into a set of vertices or polygons.
 Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including CAR models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
@@
 (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
+% Are y'all still planning on multivariate models?  Consider removing, or I'd be happy to discuss further?
 

### 2024-01-16 doc/paper-jss/sdmTMB-paper.Rnw
Edits incorporating Jim's comments

@@
 Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise arising from unmodeled processes.
-% Jim: Similar to the recent Ives MEE perspectives, I prefer to discuss unmodeled stochastic processes, rather than the term "error" or "noise", so suggesting a compromise wording here.
 More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
@@
 The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in C++.
-% Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
+% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
 _Matrix: Sparse and Dense Matrix Classes and Methods_.
@@
 lower:
-% JIM: Please feel free to email me, but if you're using the marginal likelihood to compute AIC then it's using marginal AIC.
-I have a paper in revision @ Ecology discussing TMB calculation for conditional AIC, and happy to share or discuss further.
 
@@
 % The overall predictions combining fixed and random effects are plug-in predictions.
-% Jim:  I'm adding a sentence here about catchability/detectability covariates.
-I'd be happy to discuss if y'all disagree with including this.
-Obviously I'm concerned about people moving back to the fisheries-standard of not carefully considering what is a catchability vs.
-density covariate, and would love to preserve this as distinction that is highlighted (even if it's not explicit in the model structure itself).
 We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
@@
 This is equivalent to a separable model GMRF with precision arising from the kronecker product of the spatial precision and an AR1 temporal precision.
-%% Jim:  Feel free to drop the last sentence, but I think statisticians would wanna emphasize that this is a separable space-time formulation.
 
@@
 % TODO check if on CRAN yet!
-% Jim:  It might be worth emphasizing that this calculation of marginal effects will differ from a partial dependence plot (PDP) when depth is correlated with other variables, and that package pdp could be used to additionally compute the PDP.
+% TODO Jim:  It might be worth emphasizing that this calculation of marginal effects will differ from a partial dependence plot (PDP) when depth is correlated with other variables, and that package pdp could be used to additionally compute the PDP.
 Please tell me if you want me to suggest text or a code-demo, etc.
@@
 We illustrate a way of accomplishing this by simulating from uncertainty about fixed and random effects while assuming that parametric uncertainty is well approximated using a multivariate normal distribution and the joint precision matrix.
-% Jim:  I found the previous wording a bit confusing, so suggested an edit.
 We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
@@
 
+% TODO: add multivariate responses to mgcv?
+
 \caption{Comparison of functionality between several \R\ packages that
@@
 There are several planned future additions to the \sdmTMB\ model structure.
-A subset of features to be added include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
-(4) the ability to specify observation-specific likelihoods to integrate different data types \citep{gruss2019multiple}, and
-(5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
-% Are y'all still planning on multivariate models?  Consider removing, or I'd be happy to discuss further?
-
+A subset of features to be added include
+(1) the ability to specify observation-specific likelihoods to integrate different data types \citep{gruss2019multiple},
+(2) inclusion of a dispersion formula similar to \pkg{glmmTMB},
+(3) basic multivariate models to allow for simultaneous modelling of data from multiple groups,
+and
+(4) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++},
+% (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
+% TODO: Jim: Are y'all still planning on multivariate models?  Consider removing, or I'd be happy to discuss further?
+% SA: added 'basic', which should be true (already in a branch... but unlikely to go too far with this
 % another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750

### 2024-02-05 doc/paper-jss/sdmTMB-paper.Rnw
Edits
Eric/Lewis edits

@@
 
-Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields,
+Data are often collected at particular locations in space or in space repeatedly over time. While such data are a rich source of information across many fields,
 they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
 While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation.
-Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.
+Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can provide useful spatial summary statistics \citep[e.g.,][]{thorson2019d, barnett2021}.
 
@@
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
-\Rev{B2}Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
+\Rev{B2}Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
 One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.
 
-This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).
+This paper describes the statistical models underlying \sdmTMB\ (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).
 
@@
 The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent).
-Panel a shows a shorter range than panel b, which results in a ``wigglier'' surface.
-Panel c shows the Mat\\'ern function for these two range values.
+Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface.
+Panel (c) shows the Mat\\'ern function for these two range values.
 The dashed horizontal line shows a correlation of 0.13.">>=
@@
 \Phi\left( \bm{s}_j, \bm{s}_k \right) =
-  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
+  \frac{\sigma_m^2}{2^{\nu - 1}\Gamma(\nu)}
   \left(\kappa \lVert h \rVert \right)^\nu K_\nu
@@
 \]
-The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
+The parameter $\sigma_m^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
 The parameter $\nu$ controls the smoothness of the covariance function.
@@
 \Phi\left( s_j, s_k \right) =
-  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
+  \frac{\sigma_m^2}{2^{\nu - 1}\Gamma(\nu)}
   \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
@@
 where $\kappa$ is the Mat\'ern decorrelation rate as before.
-We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.
+We can calculate the marginal variance of the Mat\'ern random field as $\sigma_m^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.
 
@@
 
-An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
+An additional delta model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
 The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
@@
 <<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
-The locations where lines intersect are referred to as ``vertices'' or sometimes ``knots''.
-Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation to a point.
+The locations where lines intersect are referred to as ``vertices'' or ``knots''.
+Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation, to a point.
 A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
@@
 
-\subsection{Exploring delta-model alternative families}
+\subsection{Exploring delta model alternative families}
 
@@
 On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
-We therefore consider a model where these are turned off in the second component of the delta model and choose this model.
+We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.
 
@@
 
-<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish density.
+<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish population density.
 The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively.
@@
 
-% \input{response}
+\input{response}
 

### 2024-02-10 doc/paper-jss/sdmTMB-paper.Rnw
Add tinyVAST + other edits

@@
 While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
-Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, lack some functionality, or may be challenging to learn for some users.
+Packages such as \pkg{VAST} \citep{thorson2019}, \pkg{tinyVAST} \citep{thorson2024tinyvast}, and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, lack some functionality, or may be challenging to learn for some users.
 
@@
 Starting with the binomial component, we have output from the smoother, which includes a linear parameter (\code{slog\_depth}) and the standard deviation on the smoother weights (\code{sds(log\_depth)}).
-The smoother summary follows the format used in the \pkg{brms} package \citep{burkner2017}.
+The smoother summary follows the format used in the \pkg{brms} package \citep{brms}.
 Next, we have the time-varying intercepts and information on our anisotropic spatial correlation.
@@
 
-We can make a basic plot using the following code. A more elaborate version including separate panels for each of the CIs is shown in Figure~\ref{fig:owl-plot-fancy}a.
+We can make a basic plot using the following code. A more elaborate version including separate panels for each of the CIs is shown in Figure~\ref{fig:owl-plot-fancy}.
 
@@
 There are many \R\ packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
-\sdmTMB, \pkg{VAST}, \INLA/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
+\sdmTMB, \pkg{VAST}, \pkg{tinyVAST}, \INLA/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
 In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
 \sdmTMB, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \INLA/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
-\pkg{mgcv} and \sdmTMB\ (currently) focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
-To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
+% \pkg{mgcv} and \sdmTMB\ (currently) focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
+To our knowledge, \pkg{VAST} and \pkg{tinyVAST} are the only packages to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
 Of these packages, only \sdmTMB\ and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
@@
 log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
-Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
+Multivariate responses & -- & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
 Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
@@
 
-% TODO: add multivariate responses to mgcv?
-
 \caption{Comparison of functionality between several \R\ packages that
   can fit geostatistical GLMMs.
+  At the time of writing, the feature set of \pkg{tinyVAST} is rapidly evolving and so is not shown here.
 Notes:
@@
 Note the log-distributed x axis.
-\pkg{VAST} should be similar to \sdmTMB\ and so is not shown.
+\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
 \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
@@
 Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
-\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts will likely find \sdmTMB\ more straightforward.
+\pkg{VAST} and \pkg{tinyVAST} are the sole options for fitting some multivariate models; alternatively, because these packages focus on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs will likely find \sdmTMB\ more straightforward.
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
@@
 There are several planned future additions to the \sdmTMB\ model structure.
+% Although we may introduce basic multivariate response modelling, we intend \sdmTMB\ to focus on univariate responses given \pkg{tinyVAST}'s focus on related models in a multivariate context.
 A subset of features to be added include
-(1) the ability to specify observation-specific likelihoods to integrate different data types \citep{gruss2019multiple},
+(1) the ability to specify observation-specific likelihoods to integrate different data types \citep[e.g.,][]{gruss2019multiple},
 (2) inclusion of a dispersion formula similar to \pkg{glmmTMB},
-(3) basic multivariate models to allow for simultaneous modelling of data from multiple groups,
+(3) inclusion of random slopes to complement the existing random intercepts,
 and
@@
 % (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
-% TODO: Jim: Are y'all still planning on multivariate models?  Consider removing, or I'd be happy to discuss further?
-% SA: added 'basic', which should be true (already in a branch... but unlikely to go too far with this
 % another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750

### 2024-02-10 doc/paper-jss/make-reprex-jss.R
Update make-reprex-jss.R

@@ -4,14 +4,13 @@ setwd(here::here())
 setwd("doc/paper-jss/")
 dir.create("reprex", showWarnings = FALSE)
 
-knitr::purl("paper-jss.Rmd", documentation = 1L)
+knitr::purl("sdmTMB-paper.Rnw", documentation = 1L)
 
-system("mv paper-jss.R reprex/paper-jss.R")
+system("mv sdmTMB-paper.R reprex/sdmTMB-paper.R")
 
-d <- readLines("reprex/paper-jss.R")
-d[grepl("## ----child=", d)]
-d <- d[!grepl("## ----child=", d)]
-
+d <- readLines("reprex/sdmTMB-paper.R")
+# d[grepl("## ----child=", d)]
+# d <- d[!grepl("## ----child=", d)]
 
 do_replace <- function(find, replacement) {
   cat("found:\n")
@@ -23,8 +22,11 @@ do_replace <- function(find, replacement) {
   d
 }
 
-d[grep("dpi = 140", d)]
-d <- do_replace("dpi = 140", "  dpi = 140")
+d[grep("knitr::render_sweave()", d)]
+d <- do_replace("knitr::render_sweave()", "# knitr::render_sweave()")
+
+d[grep("dpi = 140,", d)]
+d <- do_replace("dpi = 140,", "  dpi = 140,")
 
 d[grep(" cache = TRUE,", d)]
 d <- do_replace(" cache = TRUE,", " # cache = TRUE,")
@@ -36,23 +38,23 @@ d <- do_replace("-strip all", "  # optipng = \"-strip all\"")
 d[grep("optipng = knitr", d)]
 d <- do_replace("optipng = knitr", "# knitr::knit_hooks$set(optipng = knitr::hook_optipng)")
 
-i <- grep("## ----compare-table", d)
-j <- grep("## ----setup-pcod", d)
-d <- d[-seq(i, j-2)]
+# i <- grep("## ----compare-table", d)
+# j <- grep("## ----setup-pcod", d)
+# d <- d[-seq(i, j-2)]
 
-i <- grep("owl-knitr-setup", d)
-j <- grep("setup-owls", d)
-d <- d[-seq(i, j-1)]
+# i <- grep("owl-knitr-setup", d)
+# j <- grep("setup-owls", d)
+# d <- d[-seq(i, j-1)]
 
 
 d[grep("snow <- read", d)]
 d <- do_replace("snow <- read", "snow <- readRDS(\"snow-data.rds\")")
 
-d[grep("owl <- here::", d)]
[…]

### 2024-02-10 doc/paper-jss/sdmTMB-paper.Rnw
Account for delta versions

@@
 
-<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
+<<check-version, results='hide', echo=FALSE, include=FALSE>>=
+# delta_poisson_link_gamma() deprecated in favor of delta_gamma(type = "poisson-link")
+new_deltas <- packageVersion("sdmTMB") > '0.4.2.9000'
+@
+
+<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=new_deltas, eval=new_deltas, include=new_deltas>>=
+fit_dg <- update(fit_tw, family = delta_gamma(),
+  spatiotemporal = list("off", "iid"))
+fit_dl <- update(fit_dg, family = delta_lognormal())
+fit_dpg <- update(fit_dg, family = delta_gamma(type = "poisson-link"))
+fit_dpl <- update(fit_dg, family = delta_lognormal(type = "poisson-link"))
+@
+
+<<dog-update-old, results='hide', message=FALSE, warning=FALSE, echo=!new_deltas, eval=!new_deltas, include=!new_deltas>>=
 fit_dg <- update(fit_tw, family = delta_gamma(),

### 2024-02-10 doc/paper-jss/reprex/replication-code-main.R
Update reprex

@@ -1,116 +1,301 @@
-## ----main-setup, include=FALSE--------------------------
+## ----preliminaries, echo=FALSE, results='hide', include=FALSE, cache=FALSE----
+library(knitr)
+opts_chunk$set(engine = 'R', tidy = FALSE)
+options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+
+
+## ----main-setup, include=FALSE, cache=FALSE---------------------------
 knitr::opts_chunk$set(
   echo = FALSE,
   message = FALSE,
   warning = FALSE,
   collapse = TRUE,
   comment = "#>",
-  # prompt = FALSE,
  # cache = TRUE,
  # autodep = TRUE,
   fig.width = 7,
   fig.asp = 0.618,
   fig.pos = "ht",
+  fig.align = "center",
   cache.comments = TRUE,
   dev = "png",
-  dpi = 140
+  dpi = 140,
   # optipng = "-strip all"
-  # R.options = list(prompt = "R> ", continue = "+ ")
+  highlight = FALSE
 )
 # knitr::knit_hooks$set(optipng = knitr::hook_optipng)
 options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
+opts_chunk$set(prompt = TRUE)
+options(replace.assign = TRUE, width = 72, prompt = "R> ")
+# knitr::render_sweave()
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------
+## ----libraries, echo=FALSE, warning=FALSE, message=FALSE--------------
+library(ggplot2)
+library(dplyr)
+library(sdmTMB)
+
+
+## ----matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface. Panel (c) shows the Mat\\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13."----
[…]

### 2024-02-10 doc/paper-jss/reprex/replication-code-pcod.R

@@ -1,223 +1,0 @@
-library("sdmTMB")
-library("dplyr")
-library("ggplot2")
-library("sf")
-library("patchwork")
-dir.create("figs", showWarnings = FALSE)
-
-if (!require("rnaturalearthhires", quietly = TRUE)) {
-  message(
-    "Please install 'rnaturalearthhires' to use a high-resolution map.\n",
-    "`remotes::install_github('ropensci/rnaturalearthhires')`"
-  )
-  rnaturalearthhires_installed <- FALSE
-} else {
-  rnaturalearthhires_installed <- TRUE
-}
-
-# from remotes::install_github('seananderson/ggsidekick')
-theme_sleek <- function(base_size = 11, base_family = "") {
-  half_line <- base_size / 2
-  theme_light(base_size = base_size, base_family = base_family) +
-    theme(
-      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-      axis.ticks.length = grid::unit(half_line / 2.2, "pt"), strip.background = element_rect(
-        fill = NA,
-        colour = NA
-      ), strip.text.x = element_text(colour = "grey30"),
-      strip.text.y = element_text(colour = "grey30"), axis.text = element_text(colour = "grey30"),
-      axis.title = element_text(colour = "grey30"), legend.title = element_text(
-        colour = "grey30",
-        size = rel(0.9)
-      ), panel.border = element_rect(
-        fill = NA,
-        colour = "grey70", linewidth = 1
-      ), legend.key.size = grid::unit(
-        0.9,
-        "lines"
-      ), legend.text = element_text(
-        size = rel(0.7),
-        colour = "grey30"
-      ), legend.key = element_rect(
-        colour = NA,
-        fill = NA
-      ), legend.background = element_rect(
-        colour = NA,
-        fill = NA
-      ), plot.title = element_text(
-        colour = "grey30",
-        size = rel(1)
-      ), plot.subtitle = element_text(
-        colour = "grey30",
-        size = rel(0.85)
-      )
-    )
-}
-theme_set(theme_sleek())
-
[…]

### 2024-02-10 doc/paper-jss/make-reprex-jss.R
Update make-reprex-jss.R

@@ -82,6 +82,26 @@ d[grepl("here::", d)]
 # i <- grep("## ----inla-knitr-setup", d)
 # d <- c(d[seq(1, i-1)], "\n# INLA comparison appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])
 
+i <- grep("dog-update-old, ", d) + 1
+j <- grep("## fit_dpl <- update", d)
+for (s in i:j) {
+  d[s] <- gsub("^## ", "", d[s])
+}
+
+i <- grep("dog-update,", d) - 1
+j <- grep("dog-update-old,", d) - 2
+d <- c(d[1:i], "if (new_deltas) {", d[(i+1):(j-1)], "}", d[(j):length(d)])
+
+i <- grep("dog-update-old,", d) - 1
+j <- grep("----dog-aic,", d) - 2
+d <- c(d[1:i], "if (!new_deltas) {", d[(i+1):(j-1)], "}", d[(j):length(d)])
+
+i <- grep("inlabrufit-timing", d) + 1
+j <- grep("bru_max_iter = 1, num.threads = \"1:1\"", d) + 2
+for (s in i:j) {
+  d[s] <- gsub("^## ", "", d[s])
+}
+
 # these make figs not show up in html:
 d <- gsub('dev = \\"pdf\\"', 'dev = \\"png\\"', d)
 d <- gsub('out.width=\\"[0-9]+in\\", ', "", d)
@@ -100,7 +120,7 @@ if (RUN_SPIN) system("R -e 'knitr::spin(\"replication-code-main.R\")'")
 
 d <- readLines(here::here("analysis/timing.R"))
 writeLines(d, "replication-code-timing.R")
-# if (RUN_SPIN) system("R -e 'knitr::spin(\"replication-code-timing.R\")'")
+if (RUN_SPIN) system("R -e 'knitr::spin(\"replication-code-timing.R\")'")
 #
 # d <- readLines(here::here("analysis/pcod-fig.R"))
 # # d <- gsub('silent = TRUE', 'silent = FALSE', d)

### 2024-02-10 doc/paper-jss/sdmTMB-paper.Rnw
Make mgcv SPDE code work with new INLA

@@
   } else {
-    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
+    if (!inherits(object$xt$mesh, "inla.mesh")) stop("xt must be NULL or an inla.mesh object")
     mesh <- object$xt$mesh

### 2024-02-10 analysis/timing.R

@@ -80,7 +80,7 @@ smooth.construct.spde.smooth.spec <- function(object, data, knots) {
       stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
     }
   } else {
-    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
+    if (!inherits(object$xt$mesh, "inla.mesh")) stop("xt must be NULL or an inla.mesh object")
     mesh <- object$xt$mesh
   }
   object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))

### 2024-02-10 doc/paper-jss/reprex/replication-code-main.R
Update replication-code-main.R

@@ -4,7 +4,7 @@ opts_chunk$set(engine = 'R', tidy = FALSE)
 options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
 
 
-## ----main-setup, include=FALSE, cache=FALSE---------------------------
+## ----main-setup, include=FALSE, cache=FALSE-------------------
 knitr::opts_chunk$set(
   echo = FALSE,
   message = FALSE,
@@ -30,7 +30,7 @@ options(replace.assign = TRUE, width = 72, prompt = "R> ")
 # knitr::render_sweave()
 
 
-## ----libraries, echo=FALSE, warning=FALSE, message=FALSE--------------
+## ----libraries, echo=FALSE, warning=FALSE, message=FALSE------
 library(ggplot2)
 library(dplyr)
 library(sdmTMB)
@@ -114,26 +114,26 @@ sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range
 cowplot::plot_grid(sim_g1, sim_g2, sim_g3, align = "h", ncol = 3L)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE----------------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------------
 ## install.packages("sdmTMB")
 
 
-## ----sdmTMB-install-suggest, eval=FALSE, echo=TRUE--------------------
+## ----sdmTMB-install-suggest, eval=FALSE, echo=TRUE------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----remotes, eval=FALSE, echo=TRUE-----------------------------------
+## ----remotes, eval=FALSE, echo=TRUE---------------------------
 ## install.packages("remotes")
 ## remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
 
 
-## ----libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE-------
+## ----libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE----
 library(sdmTMB)
 library(dplyr)
 library(ggplot2)
 
 
-## ----setoptions, echo=FALSE-------------------------------------------
+## ----setoptions, echo=FALSE-----------------------------------
 options(
   pillar.print_max = 3,
   pillar.print_min = 3,
@@ -143,16 +143,16 @@ options(
 options(width = 80)
 
 
[…]

### 2024-05-25 doc/paper-jss/sdmTMB-paper.Rnw
Revisions - residuals - BLAS - convergence changes - Moran's I - statsv typo

@@
 Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
-We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
+We hope \sdmTMB will help open this useful class of models to a wider field of geostatistical analysts.
 }
@@
 
-\subsubsection{Traditional delta models}
+\subsubsection{Standard delta models}
 
@@
 \fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
-The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in C++.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in \proglang{C++}.
 % TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
@@
 “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
+The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
 \Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
-it is also possible to apply the epsilon estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
+it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
 Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
@@
 A \fct{residuals} method calculates various types of residuals.
+The default is a form of randomized quantile residuals \citep{dunn1996}.
 New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
@@
 
+Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
+% and \pkg{LAPACK} \citep{lapack}
+library
+with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
+can expect about an order of magnitude increase in model fitting speed.
+Suggestions are included in the package README.
+
 \section{Example: spatial species distribution modelling}\label{pcod}
@@
 
-We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} method.
-We omit example code for Moran's I calculation for brevity but include it in the code supplement.
-We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.
-
-<<morans, echo=FALSE, results='hide'>>=
-test_autocor <- function(obj) {
+We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
+Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
+We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
+
+<<morans, echo=TRUE>>=
+inv_dist_matrix <- 1 / as.matrix(dist(pcod[,c("X", "Y"), ]))
+diag(inv_dist_matrix) <- 0
 set.seed(1)
-  s <- simulate(obj, nsim = 500)
-  pr <- predict(obj, type = "response")$est
-  r <- DHARMa::createDHARMa(
-    simulatedResponse = s,
-    observedResponse = pcod$present,
-    fittedPredictedResponse = pr
-  )
-  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
-}
-(t_no_rf <- test_autocor(fit_bin))
-(t_rf <- test_autocor(fit_bin_rf))
-p_rf <- round(t_rf$p.value, 2)
-@
+r_bin <- residuals(fit_bin, type = "mle-mvn")
+set.seed(1)
+r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")
+
+ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
+ape::Moran.I(r_bin_rf, weight = inv_dist_matrix)$p.value
+@
+
+% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
+We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for the model with a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
+We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals \citep{dunn1996}.
 
@@
 
-Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields indicating better out-of-sample predictions:
+Indeed, the log likelihood of left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:
 
@@
 
-In practice, we would want to repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
+In practice, we would repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
-Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
-
-<<check-version, results='hide', echo=FALSE, include=FALSE>>=
-# delta_poisson_link_gamma() deprecated in favor of delta_gamma(type = "poisson-link")
-new_deltas <- packageVersion("sdmTMB") > '0.4.2.9000'
-@
-
-<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=new_deltas, eval=new_deltas, include=new_deltas>>=
-fit_dg <- update(fit_tw, family = delta_gamma(),
-  spatiotemporal = list("off", "iid"))
-fit_dl <- update(fit_dg, family = delta_lognormal())
-fit_dpg <- update(fit_dg, family = delta_gamma(type = "poisson-link"))
-fit_dpl <- update(fit_dg, family = delta_lognormal(type = "poisson-link"))
-@
-
-<<dog-update-old, results='hide', message=FALSE, warning=FALSE, echo=!new_deltas, eval=!new_deltas, include=!new_deltas>>=
-fit_dg <- update(fit_tw, family = delta_gamma(),
-  spatiotemporal = list("off", "iid"))
-fit_dl <- update(fit_dg, family = delta_lognormal())
-fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
-fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
+Note that we need to add an extra Newton optimization step to the delta-lognormal model to achieve suitable convergence (discovered via a call to \code{sanity(fit\_dl)}).
+% Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
+
+<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE>>=
+fit_dg <- update(fit_tw, family = delta_gamma())
+fit_dl <- update(fit_tw, family = delta_lognormal(),
+  control = sdmTMBcontrol(newton_loops = 2))
+fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))
+fit_dpl <- update(fit_tw, family = delta_lognormal(type = "poisson-link"))
 @
@@
 
-We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal hotspots to persist from year to year.
+We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.
 
@@
 fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
-fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
+fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = "ar1")
 @
@@
 The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
-plot_anisotropy(fit_dpl)
-@
-
-On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
-We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.
-
-<<dog-ar1-2, echo=TRUE, results="hide">>=
-fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
-@
-
-<<dog-aic3, echo=TRUE>>=
-AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
+plot_anisotropy(fit_dpl_ar1)
 @
@@
 <<dog-print, echo=TRUE>>=
-fit <- fit_dpl_ar1_only
+fit <- fit_dpl_ar1
 sanity(fit)
@@
 We then have the second model component (lognormal) with a similar summary structure but with the addition of a dispersion parameter, the AR(1) correlation of the spatiotemporal random fields, and a spatiotemporal random field marginal standard deviation.
+
+We can check simulation-based randomized quantile residuals from our chosen model via the \pkg{DHARMa} package \citep{dharma}.
+To do that, we simulate from our model with the \fct{simulate.sdmTMB} method and pass those simulations to a helper function \fct{dharma\_residuals}.
+
+<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package.
+The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE>>=
+set.seed(1)
+s <- simulate(fit, nsim = 500, type = "mle-mvn")
+dharma_residuals(s, fit)
+@
+
+The quantile-quantile plot suggests that under the model assumptions, the (transformed) residuals are consistent with an independent uniform(0, 1) distribution.
 
@@
 
-\input{response}
+% \input{../../../sdmTMB-paper-response/response}
 

### 2024-05-25 doc/paper-jss/make-reprex-jss.R
Uncommitted old changes

@@ -4,11 +4,14 @@ setwd(here::here())
 setwd("doc/paper-jss/")
 dir.create("reprex", showWarnings = FALSE)
 
+# Needed! otherwise can't find chunk option objects
+new_deltas <- TRUE
 knitr::purl("sdmTMB-paper.Rnw", documentation = 1L)
 
 system("mv sdmTMB-paper.R reprex/sdmTMB-paper.R")
 
 d <- readLines("reprex/sdmTMB-paper.R")
+
 # d[grepl("## ----child=", d)]
 # d <- d[!grepl("## ----child=", d)]
 
@@ -72,15 +75,6 @@ d <- gsub("here::here\\(\"ne_10m_lakes\"\\)", "\"ne_10m_lakes\"", d)
 d <- c(d, "sessionInfo()")
 
 d[grepl("here::", d)]
-
-# i <- grep("## ----setup-owls", d)
-# d <- c(d[seq(1, i-1)], "\n# Snowy Owl appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])
-
-# i <- grep("## ----setup-pcod", d)
-# d <- c(d[seq(1, i-1)], "\n# Pacific Cod appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])
-
-# i <- grep("## ----inla-knitr-setup", d)
-# d <- c(d[seq(1, i-1)], "\n# INLA comparison appendix -----------------------------------------------------------------------\n", d[seq(i, length(d))])
 
 i <- grep("dog-update-old, ", d) + 1
 j <- grep("## fit_dpl <- update", d)

### 2024-05-25 doc/paper-jss/reprex/replication-code-main.R

@@ -4,7 +4,7 @@ opts_chunk$set(engine = 'R', tidy = FALSE)
 options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
 
 
-## ----main-setup, include=FALSE, cache=FALSE-------------------
+## ----main-setup, include=FALSE, cache=FALSE-------------------------------
 knitr::opts_chunk$set(
   echo = FALSE,
   message = FALSE,
@@ -30,7 +30,7 @@ options(replace.assign = TRUE, width = 72, prompt = "R> ")
 # knitr::render_sweave()
 
 
-## ----libraries, echo=FALSE, warning=FALSE, message=FALSE------
+## ----libraries, echo=FALSE, warning=FALSE, message=FALSE------------------
 library(ggplot2)
 library(dplyr)
 library(sdmTMB)
@@ -114,26 +114,26 @@ sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range
 cowplot::plot_grid(sim_g1, sim_g2, sim_g3, align = "h", ncol = 3L)
 
 
-## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------------
+## ----sdmTMB-install, eval=FALSE, echo=TRUE--------------------------------
 ## install.packages("sdmTMB")
 
 
-## ----sdmTMB-install-suggest, eval=FALSE, echo=TRUE------------
+## ----sdmTMB-install-suggest, eval=FALSE, echo=TRUE------------------------
 ## install.packages("sdmTMB", dependencies = TRUE)
 
 
-## ----remotes, eval=FALSE, echo=TRUE---------------------------
+## ----remotes, eval=FALSE, echo=TRUE---------------------------------------
 ## install.packages("remotes")
 ## remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
 
 
-## ----libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE----
+## ----libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE-----------
 library(sdmTMB)
 library(dplyr)
 library(ggplot2)
 
 
-## ----setoptions, echo=FALSE-----------------------------------
+## ----setoptions, echo=FALSE-----------------------------------------------
 options(
   pillar.print_max = 3,
   pillar.print_min = 3,
@@ -143,16 +143,16 @@ options(
 options(width = 80)
 
 
[…]

### 2024-05-27 doc/paper-jss/sdmTMB-paper.Rnw
Revisions

@@
 % knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)
+% R -e 'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)'
 
@@
 
-%% -- Article metainformation (author, title, ...) -----------------------------
+%% -- Article meta information (author, title, ...) -----------------------------
 
@@
 
-Data are often collected at particular locations in space or in space repeatedly over time. While such data are a rich source of information across many fields,
+Data are often collected at particular locations in space or in space repeatedly over time.
+While such data are a rich source of information across many fields,
 they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
@@
 
-Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
+Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011, thorsonkristensen2024}.
 Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
@@
 This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
-The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2015ices, thorson2019, osgood-zimmerman2023}.
+The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2019, osgood-zimmerman2023, thorsonkristensen2024}.
 This SPDE method approximates a Matérn correlation function as arising mechanistically from local diffusion in space and/or time \citep{lindgren2011} and it results in a sparse precision matrix that permits efficient computation using existing sparse-matrix libraries \citep{rue2005gmrf}.
@@
 Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
-\Rev{B2}Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
+Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
 One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.
@@
 
-\Rev{B4a}In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
+In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
 and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{rue2005gmrf,simpson2012}.
@@
 
-\Rev{B4b}For a user of \sdmTMB, the following are the important elements to understand.
+For a user of \sdmTMB, the following are the important elements to understand.
 First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
@@
 “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
-The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
-\Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
+The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
+Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
 it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
@@
 \fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-Others standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
+Other \Rev{A7} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
 
@@
 Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
-\Rev{A4}These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
+These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
 If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.
@@
 Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
-We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
+\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
 
@@
 We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
-\Rev{A19b}We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
+We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
 
@@
   spatiotemporal = "iid",
-  reml = TRUE,
   silent = FALSE
@@
 
-\Rev{B7}The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
+The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
 Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
@@
 <<owl-p, message=FALSE, echo=FALSE, eval=FALSE, cache=TRUE>>=
-# not currently including, but might be helpful
-snow <- predict(fit_owl, newdata = snow)
-snow |> select(X, Y, year, nao, count, est, omega_s, epsilon_st, zeta_s_nao) |>
-  head(n = 2)
+pred_snow <- predict(fit_owl, newdata = snow)
+pred_snow |>
+  select(X, Y, year, nao, count, est, omega_s, epsilon_st, zeta_s_nao) |>
+  head()
 @
@@
 <<zeta-effect, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
+set.seed(1)
 zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
@@
 <<shapes, echo=FALSE>>=
-if (!file.exists(here::here("data/ne_10m_lakes"))) {
-  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
-    "download/10m/physical/ne_10m_lakes.zip")
-  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
-  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
+if (!file.exists(here::here("data/ne_10m_lakes.shp"))) {
+  lakes <- rnaturalearth::ne_download(
+    scale = 10, type = "lakes", category = "physical",
+    destdir = here::here("data")
+  )
 }
@@
 Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
-coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
+coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") |>
   sf::st_transform(crs = Albers)
 lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
+lakes <- lakes[lakes$scalerank == 0, ] |> sf::st_transform(crs = Albers)
 land <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
-land <- land %>% sf::st_transform(crs = Albers)
+land <- land |> sf::st_transform(crs = Albers)
 @
@@
 <<owl-proj2>>=
-# project to get nice axis limits
 snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
@@
 Point size is scaled to the mean counts in each location.", fig.width=5, out.width="4in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
-nao_effect_df <- select(snow2, X, Y, count, nao_effect, nao_effect_lwr, nao_effect_upr) |>
+nao_effect_df <- snow2 |>
   group_by(X, Y) |>
-  summarise_all(mean)
+  summarise(
+    count = mean(count), nao_effect = mean(nao_effect),
+    nao_effect_lwr = mean(nao_effect_lwr), nao_effect_upr = mean(nao_effect_upr),
+    .groups = "drop"
+  )
 
@@
     panel.background = element_rect(fill = "grey90", colour = NA),
-    panel.grid.major = element_line(colour = "white", size = 0.5),
+    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
     legend.background = element_rect(fill = "transparent", colour = NA),
@@
     axis.title = element_blank()
+  )
+
+th <- theme(
+    legend.position = "none",
+    panel.background = element_rect(fill = "grey90", colour = NA),
+    panel.grid.major = element_line(colour = "white", size = 0.5),
+    legend.background = element_rect(fill = "transparent", colour = NA),
+    legend.box.background = element_rect(fill = "transparent", colour = NA),
+    plot.title = element_text(size = 10), axis.ticks = element_blank(),
+    axis.title = element_blank(), axis.text = element_blank()
   )
@@
   theme_bw() +
-  theme(
-    legend.position = "none",
-    panel.background = element_rect(fill = "grey90", colour = NA),
-    panel.grid.major = element_line(colour = "white", size = 0.5),
-    legend.background = element_rect(fill = "transparent", colour = NA),
-    legend.box.background = element_rect(fill = "transparent", colour = NA),
-    plot.title = element_text(size = 10), axis.ticks = element_blank(),
-    axis.title = element_blank(), axis.text = element_blank()
-  )
+  th
 
@@
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
-\Rev{B3}Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
+Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
 % TODO: mention this in response doc
@@
 While our examples focused on applications to ecological data, the SPDE approach and the functionality of \sdmTMB\ has applications in many other fields.
-\Rev{B9a}Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
-\Rev{B9b}The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
+Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
+The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
 These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.
@@
 
-% \input{../../../sdmTMB-paper-response/response}
+\input{../../../sdmTMB-paper-response/response2}
 

### 2024-05-28 doc/paper-jss/sdmTMB-paper.Rnw
Revisions

@@
 Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
-\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
+\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimate (MLE) but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
 
@@
 
-We can also see that the marginal AIC of the model with spatial random fields is
-lower:
+We can also see that the marginal Akaike information criterion (AIC) \citep{akaike1974} of the model with spatial random fields is lower:
 
@@
 
+\Rev{A1}We note that caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
+In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
+Methods to estimate effective degrees of freedom were recently demonstrated in \citep{thorsonXX}, but are not yet included in \sdmTMB.
+
 \subsection{Comparing models with cross validation}
 
-We can do additional model comparison with cross validation via \fct{sdmTMB\_cv}.
-We will do 10-fold cross validation with the folds constructed randomly. We will set the seed each time to ensure the folds are consistent.
-Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation.
+As an alternative to AIC, we can conduct model comparison with cross validation via \fct{sdmTMB\_cv}.
+We will do 10-fold cross validation with the folds constructed randomly.
+We will set the seed each time to ensure the folds are consistent.
+Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation \citep[][]{roberts2017}.
 If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.
@@
 
-Indeed, the log likelihood of left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:
+We can then calculate any performance metric of interest for comparison.
+\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017}
+
+$$
+\mathrm{lpd} = \sum_{i=1}^{n} \ell (\widehat{y_i} | y_i),
+$$
+
+where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLE and random effects at their Empirical Bayes estimate.
+
+Indeed, the log likelihood predictive density for the left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:
 
@@
 
-In practice, we would repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
+In practice, we would repeat this procedure several times to ensure the rank order is insensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 
-% TODO: data citation!?
 \sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
@@
 
-The dataset includes spatial coordinates, year, dogfish catch weight in kg, area
-swept in km, and bottom depth:
+The dataset includes spatial coordinates, year, dogfish catch weight in kg, area swept in km, and bottom depth:
 
@@
 
-We add UTM zone 9 columns, create a log depth column for convenience, and create
-a basic mesh:
+We add UTM zone 9 columns, create a log depth column for convenience, and create a basic mesh:
 
@@
 dat$log_depth <- log(dat$depth)
-mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
+mesh <- make_mesh(dat, xy_cols = c("X", "Y"), n_knots = 200)
 @
@@
 where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$.
+We include the argument \code{extra\_time}, which represents all time steps to include in the latent process, such that any autoregressive processes represent annual time steps as opposed to the irregular years of data collection.
 
@@
   anisotropy = TRUE,
+  extra_time = seq(min(dat$year), max(dat$year)),
   silent = FALSE
@@
 fit_dg <- update(fit_tw, family = delta_gamma())
-fit_dl <- update(fit_tw, family = delta_lognormal(),
-  control = sdmTMBcontrol(newton_loops = 2))
+fit_dl <- update(fit_tw, family = delta_lognormal())
 fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))

### 2024-05-28 analysis/timing.R

@@ -7,8 +7,19 @@ library("INLA")
 library("inlabru")
 library("mgcv")
 library("spaMM")
+# install.packages("RSpectra") # for faster spaMM
+
 dir.create("figs", showWarnings = FALSE)
 dir.create("analysis", showWarnings = FALSE)
+
+is_optimized <- function() {
+  m <- 1e4; n <- 1e3; k <- 3e2
+  X <- matrix(rnorm(m*k), nrow=m)
+  Y <- matrix(rnorm(n*k), ncol=n)
+  s <- system.time(X %*% Y)
+  s[[3]] < 0.5
+}
+(optimized_blas <- is_optimized())
 
 INLA::inla.setOption(num.threads = "1:1")
 
@@ -192,28 +203,13 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
       data = sim_dat,
       mesh = mesh,
       family = family,
-      control = sdmTMBcontrol(newton_loops = 0L),
+      control = sdmTMBcontrol(newton_loops = 0L, multiphase = FALSE),
       priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2))
     )
   })
   times$sdmTMB <- out[["elapsed"]]
 
-  dat_sp <- sp::SpatialPointsDataFrame(cbind(sim_dat$X, sim_dat$Y),
-    proj4string = sp::CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
-+ +y_0=0 +datum=NAD83 +units=km +no_defs"), data = sim_dat
-  )
-  loc.bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
-  segm.bnd <- INLA::inla.mesh.segment(loc.bnd)
-  mesh2 <- INLA::inla.mesh.2d(
-    boundary = segm.bnd,
-    max.edge = c(max.edge, 0.2),
-    offset = c(0.1, 0.05)
-  )
-  spde <- INLA::inla.spde2.pcmatern(
-    mesh = mesh2,
-    prior.range = c(0.05, 0.05),
-    prior.sigma = c(2, 0.05)
-  )
+  cat("inlabru EB\n")
 
   dat <- s$dat
[…]

### 2024-05-28 doc/paper-jss/sdmTMB-paper.Rnw
Revisions

@@
 
+<<check-version, echo=FALSE, cache=FALSE>>=
+if (packageVersion("sdmTMB") < "0.6.0") {
+  stop("Please install a version of sdmTMB >= 0.6.0.", call. = FALSE)
+}
+@
+
 \clearpage
@@
 Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
-One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.
+One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.
 
@@
 \]
-where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_{\bm{s}}$.
+where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is defined as the mean $\mu_{\bm{s}}$.
 That mean $\mu_{\bm{s}}$ is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta_{\bm{s}}$.
@@
 A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
-A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).
+A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}c).
 
@@
 g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
-\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
+\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma}_t +
 \bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
@@
 \]
-where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
+where the $\bm{X}$ represent model matrices, $\bm{\gamma}_t$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
 We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
@@
 
-So far we have described models with one linear predictor and one family (common terminology in \R\ packages for the combination of an observation likelihood and link).
+So far, we have described models with one linear predictor and one family (common terminology in \R\ packages for the combination of an observation likelihood and link).
 Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}.
@@
 \fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
-The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in \proglang{C++}.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
 % TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
@@
 “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
-The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
 Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
@@
 The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
-whether to return predictions on the link or response scale (type),
+whether to return predictions on the link or response scale (\code{type}),
 whether to return standard errors (\code{se\_fit},
@@
 A \fct{residuals} method calculates various types of residuals.
-The default is a form of randomized quantile residuals \citep{dunn1996}.
+The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.
 New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
@@
 <<remotes, eval=FALSE, echo=TRUE>>=
-install.packages("remotes")
-remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
+install.packages("pak")
+pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
 @
@@
 % and \pkg{LAPACK} \citep{lapack}
-library
-with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
-can expect about an order of magnitude increase in model fitting speed.
-Suggestions are included in the package README.
+library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
+can expect up to an order of magnitude increase in model fitting speed.
+Suggestions are included in the package readme file
 
@@
 Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
-Usually, mesh complexity has the single largest influence on the speed of fitting these models.
+Mesh complexity has a large influence on the speed of fitting these models.
 
@@
 r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")
-
 ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
@@
 % We omit example code for Moran's I calculation for brevity but include it in the code supplement.
-We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for the model with a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
+We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
 We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals \citep{dunn1996}.
@@
 In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
-Methods to estimate effective degrees of freedom were recently demonstrated in \citep{thorsonXX}, but are not yet included in \sdmTMB.
+Methods to estimate effective degrees of freedom for similar models were recently demonstrated in \citet{thorson2024cAIC}, but are not yet included in \sdmTMB.
 
@@
 
-As an alternative to AIC, we can conduct model comparison with cross validation via \fct{sdmTMB\_cv}.
+As an alternative to AIC, we can conduct model comparison with cross validation.
+\sdmTMB\ includes the helper function \fct{sdmTMB\_cv} to facilitate this.
 We will do 10-fold cross validation with the folds constructed randomly.
@@
 
-where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLE and random effects at their Empirical Bayes estimate.
+where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
 
@@
 Some covariates might be fixed at a specified value for all predictions, such that we are predicting the expected value for samples conditional on those specified values.
-We call these covariates \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates that was estimated in fitted data \cite{thorson2019}.
+In the context of fish or animal surveys, these are sometimes called \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates \cite{thorson2019}.
 The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
 % The overall predictions combining fixed and random effects are plug-in predictions.
-We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
+We can plot these with \fct{geom\_raster} or \fct{geom\_tile}  from the \pkg{ggplot2} \citep{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).
 
@@
 We can then specify our model.
-We include an offset for the effort variable log area swept such that we are effectively modelling density and our predictions will be for an area swept of 1 km\textsuperscript{2}.
+We include an offset \citep[][p.~206]{mccullagh1989} for the effort variable log area swept such that we are effectively modelling density and our predictions will be for an area swept of 1 km\textsuperscript{2}.
 
@@
 \]
-where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$.
-We include the argument \code{extra\_time}, which represents all time steps to include in the latent process, such that any autoregressive processes represent annual time steps as opposed to the irregular years of data collection.
+where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$ and $\sigma^2_{\gamma}$ represents the marginal variance of this process.
+We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.
 
@@
 
-We can explore four alternative families that may better represent the data.
+We next explore four alternative families that may better represent the data.
 Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
-Note that we need to add an extra Newton optimization step to the delta-lognormal model to achieve suitable convergence (discovered via a call to \code{sanity(fit\_dl)}).
-% Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
 
@@
 
-We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.
-
-The AR(1) fields can be represented as:
+We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.
+
+The AR(1) fields can be represented as
 \[
@@
 where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields each year.
-This is equivalent to a separable model GMRF with precision arising from the kronecker product of the spatial precision and an AR1 temporal precision.
+This is equivalent to a separable model GMRF with precision arising from the Kronecker product of the spatial precision and an AR(1) temporal precision.
 
@@
 <<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package.
-The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE>>=
-set.seed(1)
+The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE, out.width="4in">>=
+set.seed(42)
 s <- simulate(fit, nsim = 500, type = "mle-mvn")
-dharma_residuals(s, fit)
-@
-
-The quantile-quantile plot suggests that under the model assumptions, the (transformed) residuals are consistent with an independent uniform(0, 1) distribution.
+dharma_residuals(s, fit, test_uniformity = TRUE)
+@
+
+The quantile-quantile plot suggests that under the model assumptions, the (transformed) residuals are reasonably consistent with an independent uniform(0, 1) distribution.
 
@@
   scale_fill_viridis_c(trans = "log10") +
-  ggtitle("(a) Non-random-field components; first delta model")
+  ggtitle("(a) Non-random-field components; 1st linear predictor")
 @
@@
   scale_fill_gradient2() +
-  ggtitle("(b) Spatial random field; first delta model")
+  ggtitle("(b) Spatial random field; 1st linear predictor")
 @
@@
   scale_fill_gradient2() +
-  ggtitle("(c) Spatiotemporal random field; second delta model")
+  ggtitle("(c) Spatiotemporal random field; 2nd linear predictor")
 @
@@
 
-<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction elements from the spatiotemporal model of Pacific Dogfish biomass density.
+<<dog-wcvi-pred, fig.asp = 0.8, fig.cap= "Example prediction elements from the spatiotemporal model of Pacific Dogfish biomass density.
 Throughout, two example years are shown.
-(a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field elements (here, a smoother for bottom depth and the time-varying year effect) from the first delta model component, (b) \\code{omega\\_s1} refers to the spatial random field from the first delta model component, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model component, and (d) \\code{est} refers to the overall prediction estimate combining all effects.
+(a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field elements (here, a smoother for bottom depth and the time-varying year effect) from the first linear predictor, (b) \\code{omega\\_s1} refers to the spatial random field from the first linear predictor, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second linear predictor, and (d) \\code{est} refers to the overall prediction estimate combining all effects.
 The spatial random field is constant through time (i.e., the two panels in b are identical) and represents static biotic or abiotic features not included as covariates (e.g., habitat).
@@
 We will fit counts using a negative binomial \citep[NB2;][]{hilbe2011} distribution, random intercepts for year, spatial and spatiotemporal random fields, and an SVC associated with the NAO.
-Centering and scaling variables (by their mean and standard deviation here) can be especially helpful to reduce the correlation between the SVC and other random fields that are included in the model when estimating SVCs.
+Centering and scaling variables (e.g., by their mean and standard deviation) can be helpful to reduce the correlation between the SVC and other random fields that are included in the model when estimating SVCs.
 Here, NAO is an index with a mean near zero and an SD not too far from 1, so we will leave it as is.
@@
   count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ nao,
+  spatial_varying = ~ 0 + nao,
   time = "year",
@@
 However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
-We illustrate a way of accomplishing this by simulating from uncertainty about fixed and random effects while assuming that parametric uncertainty is well approximated using a multivariate normal distribution and the joint precision matrix.
+We illustrate a way of accomplishing this by simulating from the fixed and random effects while assuming that parametric uncertainty is well approximated using a multivariate normal distribution and the joint precision matrix.
 We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
 By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
-Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field.
+Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field (\code{sims\_var = "zeta_s"}).
 This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
@@
 Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
-For this example, we use thresholds of 0.1 and 0.9 representing CIs of 80\%.
+For this example, we use thresholds of 0.025 and 0.975 representing a 95\% CI.
 
@@
 <<zeta-effect, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
-set.seed(1)
-zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
+set.seed(42)
+zeta_s <- predict(fit_owl, newdata = snow, nsim = 300, sims_var = "zeta_s")
 dim(zeta_s)
-sims <- spread_sims(fit_owl, nsim = 200)
+sims <- spread_sims(fit_owl, nsim = 300)
 dim(sims)
@@
 snow$nao_effect <- exp(apply(combined, 2, median))
-snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.1))
-snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.9))
+snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.025))
+snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.975))
 @
@@
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
-However, if we recalculated the CI at a 95\% level (not shown) the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
 
@@
 However, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB\ and \pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing}).
-For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
+For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.
 
@@
 
-% TODO: B9: I kept thinking about potential overlaps with spatial / spatiotemporal econometrics.
-Maybe a short note on how those methods relate to spatial econometrics (e.g.
-Elhorst, 2012; Lee & Yu, 2010) may broaden the interest in the paper.
-
-% TODO: figure out where to put this paragraph
-% TODO: rework this paragraph to be advantages of random fields GAMs and CAR
-The GLMMs underpinning \sdmTMB\ models are spatially explicit and derived from a mechanistic (diffusion) process---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
-In contrast, non-parametric approaches do not estimate spatial covariance functions [e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}].
-%  I think semi-parametric is defined differently in different contexts, but don't agree with including it in that claim so removing it rather than adding clarifying text.
-% The random fields in \sdmTMB\ are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
-% The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
-% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\code{share\_range} argument).
-% TODO: figure out where to put this paragraph
+The GLMMs underpinning \sdmTMB\ models are spatially explicit and derived from a mechanistic diffusion process---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
+In contrast, non-parametric approaches [e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}] and most smooths in \pkg{mgcv} \citep{wood2017a} do not estimate spatial covariance functions.
 Approaches such as conditional autoregressive models (CAR) are applicable to areal data, where the spatial domain is discretized into a set of vertices or polygons.
@@
 and
-(4) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++},
+(4) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base is written in \R\ rather than \proglang{C++}.
 % (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.

### 2024-05-28 analysis/timing.R

@@ -323,7 +323,7 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
     max.edge = c(max.edge, 0.2),
     offset = c(0.1, 0.05)
   )
-  # if (max.edge > 0.07) {  # finest mesh too slow and not plotted
+  if (!(max.edge < 0.07 && !optimized_blas)) {  # finest mesh too slow and not plotted
     out <- system.time({
       fit <- tryCatch(
         {
@@ -341,10 +341,10 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
         error = function(e) NA
       )
     })
-  # } else {
-    # print("Too slow, skipping mgcv here")
-    # fit <- NA
-  # }
+  } else {
+    print("Too slow, skipping mgcv here")
+    fit <- NA
+  }
   times$mgcv_disc <- if (all(is.na(fit))) NA else out[["elapsed"]]
 
   out <- as_tibble(times)
@@ -369,7 +369,7 @@ to_run <- expand.grid(
 # full (slow) run to match paper:
 if (FALSE) {
   # iter <- if (optimized_blas) c(50, 25, 10) else c(25, 10, 5)
-  iter <- if (optimized_blas) c(2, 2, 2) else c(2, 2, 2)
+  iter <- if (optimized_blas) c(25, 10, 10) else c(25, 10, 10)
 
   to_run <- expand.grid(
     n_obs = c(1e3),
@@ -404,8 +404,11 @@ system.time({
   )
 })
 
+saveRDS(out, file = "analysis/timing-cache-2024-05-27-standard-large.rds")
+saveRDS(out, file = "analysis/timing-cache-2024-05-27-optimized-large.rds")
 # saveRDS(out, file = "analysis/timing-cache-parallel-2024-05-26-standard-large.rds")
 # saveRDS(out, file = "analysis/timing-cache-parallel-2024-05-26-optimized-large.rds")
+out2 <- readRDS("analysis/timing-cache-parallel-2024-05-26-optimized-large.rds")
 out1 <- readRDS("analysis/timing-cache-parallel-2024-05-26-standard-large.rds")
 out2 <- readRDS("analysis/timing-cache-parallel-2024-05-26-optimized-large.rds")
 out1$blas_optimized <- FALSE
@@ -484,7 +487,7 @@ g <- out_long_sum |>
[…]

### 2024-06-01 doc/paper-jss/sdmTMB-paper.Rnw
Updates to paper

@@
 \usepackage[utf8]{inputenc}
+
+% temp: track changes
+\usepackage{changes}
 
@@
 \Abstract{
-Geostatistical spatial or spatiotemporal data are common across scientific fields.
+Geostatistical \added{spatial} or spatiotemporal data are common across scientific fields.
 % and the dimensions of these data continue to grow in space and time.
@@
 Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
-We hope \sdmTMB will help open this useful class of models to a wider field of geostatistical analysts.
+We hope \sdmTMB \deleted{'s accessible interface} will help open this useful class of models to a wider field of geostatistical analysts.
 }
@@
 
-<<check-version, echo=FALSE, cache=FALSE>>=
+<<check-version, echo=FALSE, cache=FALSE, eval=TRUE>>=
 if (packageVersion("sdmTMB") < "0.6.0") {
@@
 Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise arising from unmodeled processes.
-More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
+More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}c).
 The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
@@
 \fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
-The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \added{\citep{Matrix}} as \added{an} interface to the \pkg{Eigen} library in \proglang{C++}.
 % TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
@@
 “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2024}.
 Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
-it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
+it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
 Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
@@
 \fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-Other \Rev{A7} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
+\Rev{A7}\added{Other} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
 
@@
 A \fct{residuals} method calculates various types of residuals.
-The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.
+\added{The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.}
 New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
@@
 
+\added{
 Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
@@
 can expect up to an order of magnitude increase in model fitting speed.
-Suggestions are included in the package readme file
+Suggestions are included in the package README file.
+}
 
@@
 Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
-Mesh complexity has a large influence on the speed of fitting these models.
+\added{Mesh complexity has a large influence} on the speed of fitting these models.
 
@@
 
+\added{
 We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
@@
 \Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimate (MLE) but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
+}
 
@@
 % We omit example code for Moran's I calculation for brevity but include it in the code supplement.
+\added{
 We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
-We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals \citep{dunn1996}.
+We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
+}
 
@@
 
+\added{
 \Rev{A1}We note that caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
@@
 Methods to estimate effective degrees of freedom for similar models were recently demonstrated in \citet{thorson2024cAIC}, but are not yet included in \sdmTMB.
+}
 
@@
 
+\added{
 As an alternative to AIC, we can conduct model comparison with cross validation.
 \sdmTMB\ includes the helper function \fct{sdmTMB\_cv} to facilitate this.
+}
 We will do 10-fold cross validation with the folds constructed randomly.
@@
 
+\added{
 We can then calculate any performance metric of interest for comparison.
 \Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017}
-
+}
+
+\added{
 $$
@@
 $$
-
+}
+
+\added{
 where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
-
-Indeed, the log likelihood predictive density for the left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:
+}
+
+Indeed, the log likelihood predictive density \added{for the left-out data} is considerably \added{higher} for the model that includes random fields indicating better out-of-sample predictions:
 
@@
 
-In practice, we would repeat this procedure several times to ensure the rank order is insensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
+In practice, we would \deleted{want to} repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 
-We add UTM zone 9 columns, create a log depth column for convenience, and create a basic mesh:
+\Rev{A8}We add UTM zone 9 columns, create a log depth column for convenience, and create a basic mesh:
 
@@
 
-We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.
+We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the spatiotemporal random fields as AR(1) to allow spatiotemporal \added{patterns} to \added{partially} persist from year to year.
 
@@
 @
+
+\Rev{A4}
+\deleted{
+On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
+We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.
+}
 
@@
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
+\deleted{
+However, if we recalculated the CI at a 95\% level (not shown) the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
+}
 
@@
   sf::st_transform(crs = Albers)
-lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
+lakes <- sf::st_read(here::here("data/ne_10m_lakes.shp"), quiet = TRUE)
 lakes <- lakes[lakes$scalerank == 0, ] |> sf::st_transform(crs = Albers)
@@
 Of these packages, only \sdmTMB\ and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
-\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
+To our understanding, \pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
 There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
@@
 
-We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
+We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
 Our test was restricted to one core and default \R\ algebra libraries.
-\sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
-These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-However, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB\ and \pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing}).
+% \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
+% These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
+\added{
+With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
+At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
+\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\sim$ 200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
+\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial and not spatiotemporal models.
+At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
+}
+Speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
 For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.
@@
 \centering
-\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
+% \includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
+\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel.pdf}
 \caption[]{
-Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
-Lines represent means and ribbons 95\% quantiles across 50 random iterations.
-Note the log-distributed x axis.
+Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
+Lines represent means and ribbons 90\% quantiles across 50 random iterations.
 \pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
-\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
-\pkg{mgcv} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
-Note that \pkg{spaMM} only fits spatial, not spatiotemporal, models.
-All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}
+\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
+\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
+Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
+All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
+See Figure~\ref{fig:timing2} for a version with log-log axes.
+}
 \label{fig:timing}
@@
 
-We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}).
+We simulated 1,000, \added{10,000, or 100,000} spatial observations with both x and y coordinates from uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}).
 Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
@@
 
-We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Figure~\ref{fig:timing}).
-The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library.
+We conducted 50 simulation iterations per model and mesh resolution and show timing results for the mean, 10\%, and 90\% quantile values.
+The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default \pkg{BLAS} \citep{blas} library packaged with \R \citep{r2024}.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel-log.pdf}
+\caption[]{
+  \added{
+  Same as Figure~\ref{fig:timing} but with both axes log distributed.
+}
+}
+\label{fig:timing2}
+\end{figure}
 
@@
 @
+
+\clearpage
 

### 2024-06-01 analysis/timing.R

@@ -323,7 +323,7 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
     max.edge = c(max.edge, 0.2),
     offset = c(0.1, 0.05)
   )
-  if (!(max.edge < 0.07 && !optimized_blas)) {  # finest mesh too slow and not plotted
+  # if (!(max.edge < 0.07 && !optimized_blas)) {  # finest mesh too slow and not plotted
     out <- system.time({
       fit <- tryCatch(
         {
@@ -341,10 +341,10 @@ sim_fit_time <- function(n_obs = 1000, cutoff = 0.1, iter = 1, phi = 0.3, tweedi
         error = function(e) NA
       )
     })
-  } else {
-    print("Too slow, skipping mgcv here")
-    fit <- NA
-  }
+  # } else {
+  #   print("Too slow, skipping mgcv here")
+  #   fit <- NA
+  # }
   times$mgcv_disc <- if (all(is.na(fit))) NA else out[["elapsed"]]
 
   out <- as_tibble(times)
@@ -395,6 +395,9 @@ if (FALSE) {
 to_run$seed <- to_run$iter * 29212
 nrow(to_run)
 
+to_run <- filter(to_run, n_obs == 1e3)
+nrow(to_run)
+
 # don't run in parallel to ensure all timing is comparable:
 system.time({
   out <- purrr::pmap_dfr(
@@ -404,23 +407,27 @@ system.time({
   )
 })
 
-saveRDS(out, file = "analysis/timing-cache-2024-05-27-standard-large.rds")
+saveRDS(out, file = "analysis/timing-cache-2024-05-31-standard-large.rds")
 saveRDS(out, file = "analysis/timing-cache-2024-05-27-optimized-large.rds")
+
 # saveRDS(out, file = "analysis/timing-cache-parallel-2024-05-26-standard-large.rds")
 # saveRDS(out, file = "analysis/timing-cache-parallel-2024-05-26-optimized-large.rds")
-out2 <- readRDS("analysis/timing-cache-parallel-2024-05-26-optimized-large.rds")
-out1 <- readRDS("analysis/timing-cache-parallel-2024-05-26-standard-large.rds")
[…]

### 2024-06-01 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
 \added{
-Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
+Users willing to replace the default \R\ \pkg{BLAS} (Basic Linear Algebra Subprograms) \citep{blas}
 % and \pkg{LAPACK} \citep{lapack}
 library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
-can expect up to an order of magnitude increase in model fitting speed.
+can expect up to an order of magnitude increase in model fitting speed for complex models.
 Suggestions are included in the package README file.
@@
 We can then calculate any performance metric of interest for comparison.
-\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017}
+\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017},
 }
@@
 $$
-\mathrm{lpd} = \sum_{i=1}^{n} \ell (\widehat{y_i} | y_i),
+\mathrm{lpd} = \sum_{i=1}^{n} \log \mathcal{L} (\widehat{y_i} | y_i),
 $$
@@
 \added{
-where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
+where $n$ represents the number of left-out data points, $\log \mathcal{L}$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
 }
@@
 where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$ and $\sigma^2_{\gamma}$ represents the marginal variance of this process.
-We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.
+\added{We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.}
 
@@
 We can use \fct{plot\_anisotropy} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).
+
+\added{
+As an example of estimation time for these complex spatiotemporal models, the Tweedie model (\code{fit\_tw}), Poisson-link delta-lognormal model (\code{fit\_dpl}), and Poisson-link delta-lognormal model with autogressive random fields (\code{fit\_dpl\_ar1}) took approximately 10, 30, and 90 seconds to fit, respectfully, on an Apple MacBook Pro with an M2 Pro processor and with Apple's \pkg{vecLib} implementation of \pkg{BLAS} in \R\ 4.4.0.
+}
 
@@
 \added{
-With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
+With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three- to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
 At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
 \pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\sim$ 200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
-\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial and not spatiotemporal models.
+\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial GMRFs.
 At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
@@
 \caption[]{
-Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
+Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to \added{three dataset sizes}.
 Lines represent means and ribbons 90\% quantiles across 50 random iterations.
 \pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
-\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
-\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
-Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
+\pkg{inlabru} \added{(version 2.10.1)} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
+\pkg{mgcv} \added{(version 1.9.1)} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
+Note that \pkg{spaMM} \added{(version 4.4.16)} only fits spatial, not spatiotemporal, models.
 All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
-See Figure~\ref{fig:timing2} for a version with log-log axes.
+\added{See Figure~\ref{fig:timing2} for a version with log-log axes.}
 }
@@
 We conducted 50 simulation iterations per model and mesh resolution and show timing results for the mean, 10\%, and 90\% quantile values.
-The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default \pkg{BLAS} \citep{blas} library packaged with \R \citep{r2024}.
+The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.4.0 and the default \pkg{BLAS} library packaged with \R \citep{r2024}.
 

### 2024-06-01 doc/paper-jss/sdmTMB-paper.Rnw
Drop first 'we note'
Drop 2nd we note

@@
 We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
-We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
+The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
 }
@@
 \added{
-\Rev{A1}We note that caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
+\Rev{A1}Caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
 In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
@@
 
-\input{../../../sdmTMB-paper-response/response2}
+% \input{../../../sdmTMB-paper-response/response2}
 

### 2024-06-07 doc/paper-jss/sdmTMB-paper.Rnw
Note R and Matrix versions in timing
Edits to timing section

@@
 
-We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
-Our test was restricted to one core and default \R\ algebra libraries.
-% \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
-% These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-\added{
+We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
+Our test was restricted to one core and default \R\ algebra libraries \added{using \R\ 4.4.0 and \pkg{Matrix} version 1.7.0.
 With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three- to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
 At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
-\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\sim$ 200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
-\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial GMRFs.
+\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\approx$~200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
+The \pkg{mgcv} model was fit with the \fct{bam} function instead of the standard \fct{gam}.
+The \fct{bam} function here uses numerical methods optimized for large datasets, including a discretization of covariate values \citep{wood2015, wood2017b, li2020a}.
+The standard \fct{gam} function was approximately 60 times slower than \fct{bam} for the largest dataset with the most complex mesh (not shown).
+\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but to our knowledge can only fit spatial GMRFs.
 At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
 }
-Speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
-For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.
+Speed increases can allow for more rapid and thorough model exploration and experimentation with this class of computationally intensive models.
+For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.
 

### 2024-06-10 doc/paper-jss/diff.R
Create diff.R

@@ -1,0 +1,48 @@
+old_folder <- "~/Downloads/sdmTMB-paper-53264baf950427e13209950484b881712f1d74c4/doc/paper-jss/"
+new_folder <- "~/src/sdmTMB-paper/doc/paper-jss/"
+
+wd <- getwd()
+setwd(new_folder)
+
+f2 <- readLines(paste0(new_folder, "sdmTMB-paper.tex"))
+for (i in seq_along(f2)) {
+  x <- f2[i]
+  x <- gsub("\\\\Rev\\{[a-zA-Z0-9]+\\}", "", x) # reviewer reference tags
+  f2[i] <- x
+}
+
+# remove response letter at end:
+
+resp <- grep("response2", f2)
+f2 <- c(f2[seq(1, resp-1)], "\\end{document}")
+
+writeLines(f2, paste0(new_folder, "/main-clean.tex"))
+system(paste0("latexdiff ", old_folder, "/sdmTMB-paper.tex main-clean.tex > diff.tex"))
+
+system("pdflatex diff.tex")
+system("bibtex diff")
+system("pdflatex diff.tex")
+system("pdflatex diff.tex")
+system("pdflatex diff.tex")
+system("pdflatex diff.tex")
+
+system("pdflatex main-clean.tex")
+system("bibtex main-clean")
+system("pdflatex main-clean.tex")
+system("pdflatex main-clean.tex")
+system("pdflatex main-clean.tex")
+system("pdflatex main-clean.tex")
+
+system("cp main-clean.pdf anderson-etal-main.pdf")
+system("cp diff.pdf anderson-etal-diff.pdf")
+
+system("pdflatex main.tex")
+system("bibtex main")
+system("pdflatex main.tex")
+system("pdflatex main.tex")
+system("pdflatex main.tex")
+system("pdflatex main.tex")
+
+system("open .")
+
+setwd(wd)

### 2024-06-10 doc/paper-jss/sdmTMB-paper.Rnw
Edits

@@
   scale_x_continuous(breaks = r) +
-  theme(legend.position = c(0.7, 0.7)) +
+  theme(legend.position = "inside", legend.position.inside = c(0.7, 0.7)) +
   ggtitle("(c) Matérn correlation function")
@@
 \added{
-We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
+We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
 The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
@@
 \added{
-where $n$ represents the number of left-out data points, $\log \mathcal{L}$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
+where $n$ represents the number of left-out data points, $\log \mathcal{L}$ is a symbol for log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
 }
@@
 
-In practice, we would \deleted{want to} repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
+In practice, we would repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 <<dog-index, echo=TRUE>>=
+grid$area <- 4 # 2 km x 2 km
 pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
-ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
+ind <- get_index(pred2, bias_correct = TRUE, area = grid$area)
 @
@@
 SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993, thorson2023SVC}.
-In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.
+\deleted{In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.}
 
@@
   theme(
-    legend.position = c(0.25, 0.17),
+    legend.position = "inside",
+    legend.position.inside = c(0.25, 0.17),
     legend.title = element_text(size = 9, hjust = 0),
@@
   ) +
-  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
+  ggtitle(paste0("(b) Lower 95% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
   theme_bw() +
@@
   ) +
-  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
+  ggtitle(paste0("(c) Upper 95% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
   theme_bw() +
@@
 \centering
-% \includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
-\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel.pdf}
+\includegraphics[width=\textwidth]{../../figs/timing-spatial-multipanel.pdf}
 \caption[]{
@@
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
-Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
-% TODO: mention this in response doc
+Overall, \deleted{we think there is value in that} \sdmTMB\ unites functionality useful in many applied settings into a single package.
 
@@
 \centering
-\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel-log.pdf}
+\includegraphics[width=\textwidth]{../../figs/timing-spatial-multipanel-log.pdf}
 \caption[]{
@@
 % \clearpage
-
 % \input{../../../sdmTMB-paper-response/response2}

### 2024-06-11 doc/paper-jss/prep-submission.R
prep submission script

@@ -1,0 +1,38 @@
+setwd(here::here())
+setwd("doc/paper-jss/")
+
+system('R -e \'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)\'')
+
+system("gs -dBATCH -dNOPAUSE -q -sDEVICE=pdfwrite -dFirstPage=51 -dLastPage=54 -sOUTPUTFILE=reviewer-response2.pdf sdmTMB-paper.pdf")
+
+d <- readLines("sdmTMB-paper.Rnw")
+i <- grep("response2}", d)
+d[seq(i-1, i+1)] <- paste0("% ", d[seq(i-1, i+1)])
+writeLines(d, "sdmTMB-paper.Rnw")
+
+system('R -e \'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)\'')
+
+system("mv sdmTMB-paper.pdf sdmTMB-paper-V3-changes.pdf")
+
+d <- readLines("sdmTMB-paper.Rnw")
+i <- grep("\\[final\\]", d)
+d[i] <- "\\usepackage[final]{changes}"
+d[i-1] <- "% \\usepackage{changes}"
+writeLines(d, "sdmTMB-paper.Rnw")
+
+system('R -e \'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)\'')
+
+system("mv sdmTMB-paper.pdf sdmTMB-paper-V3.pdf")
+
+## revert state:
+d <- readLines("sdmTMB-paper.Rnw")
+i <- grep("response2}", d)
+d[seq(i-1, i+1)] <- gsub("\\% ", "", d[seq(i-1, i+1)])
+i <- grep("\\[final\\]", d)
+d[i] <- "% \\usepackage[final]{changes}"
+d[i-1] <- "\\usepackage{changes}"
+writeLines(d, "sdmTMB-paper.Rnw")
+
+# system('R -e \'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)\'')
+
+setwd(here::here())

### 2025-06-13 doc/jss-accepted-version/tex/jss4956.tex
Add files sent to JSS

--- 2024-06-10 doc/paper-jss/sdmTMB-paper.Rnw
@@
-\documentclass[article]{jss}
-
-% setwd("doc/paper-jss/")
-% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)
-% R -e 'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)'
+\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[dvipsnames]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\usepackage{Sweave}
 
@@
 
-% temp: track changes
+% \usepackage{changes}
+\usepackage[final]{changes}
 \usepackage{changes}
@@
 \newcommand{\R}{\proglang{R}}
-
-% TODO START FOR REVIEW ONLY
-\usepackage{lineno}
-\definecolor{darkgrey}{HTML}{A9A9A9}
-\renewcommand\linenumberfont{\normalfont\bfseries\small\color{darkgrey}}
-\modulolinenumbers[2]
-\usepackage[dvipsnames]{xcolor}
-\definecolor{niceblue}{HTML}{236899}
-\newcommand{\rev}[1]{{\color{niceblue} #1}}
-\newcommand{\Rev}[1]{\label{#1}\linelabel{#1}}
-\newcommand{\lr}[1]{line~\lineref{#1}}
-% TODO END FOR REVIEW ONLY
 
@@
 \Abstract{
-Geostatistical \added{spatial} or spatiotemporal data are common across scientific fields.
+Geostatistical spatial or spatiotemporal data are common across scientific fields.
 % and the dimensions of these data continue to grow in space and time.
@@
 
-% \widowpenalty1000
-% \clubpenalty1000
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
-
-\linenumbers
-
-<<preliminaries, echo=FALSE, results='hide', include=FALSE, cache=FALSE>>=
-library(knitr)
-opts_chunk$set(engine = 'R', tidy = FALSE)
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-@
-
-<<main-setup, include=FALSE, cache=FALSE>>=
-knitr::opts_chunk$set(
-  echo = FALSE,
-  message = FALSE,
-  warning = FALSE,
-  collapse = TRUE,
-  comment = "#>",
-  cache = TRUE,
-  autodep = TRUE,
-  fig.width = 7,
-  fig.asp = 0.618,
-  fig.pos = "ht",
-  fig.align = "center",
-  cache.comments = TRUE,
-  dev = "pdf",
-  dpi = 140,
-  optipng = "-strip all",
-  highlight = FALSE
-)
-knitr::knit_hooks$set(optipng = knitr::hook_optipng)
-options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
-opts_chunk$set(prompt = TRUE)
-options(replace.assign = TRUE, width = 72, prompt = "R> ")
-knitr::render_sweave()
-@
-
-<<libraries, echo=FALSE, warning=FALSE, message=FALSE>>=
-library(ggplot2)
-library(dplyr)
-library(sdmTMB)
-@
-
-<<check-version, echo=FALSE, cache=FALSE, eval=TRUE>>=
-if (packageVersion("sdmTMB") < "0.6.0") {
-  stop("Please install a version of sdmTMB >= 0.6.0.", call. = FALSE)
-}
-@
 
@@
 A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
-% TODO: decide on using GRF throughout
 
@@
 
-<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values.
-The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent).
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=\textwidth]{Figures/matern-range-1}
+
+}
+
+\caption[Example Gaussian random fields for two range values]{Example Gaussian random fields for two range values.
+The range describes the distance at which spatial correlation decays to $\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent).
 Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface.
-Panel (c) shows the Mat\\'ern function for these two range values.
-The dashed horizontal line shows a correlation of 0.13.">>=
-predictor_dat <- expand.grid(
-  x = seq(0, 1, length.out = 100),
-  y = seq(0, 1, length.out = 100),
-  year = seq_len(6)
-)
-sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
-s1 <- sdmTMB_simulate(
-  formula = ~1,
-  data = predictor_dat,
-  mesh = sim_mesh,
-  range = 0.2,
-  phi = 0.1,
-  sigma_O = 0.2,
-  seed = 1,
-  B = 0
-)
-sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
-  geom_raster(show.legend = FALSE) +
-  scale_fill_viridis_c(option = "C") +
-  coord_fixed(expand = FALSE) +
-  theme_light() +
-  # theme(axis.title = element_blank()) +
-  ggtitle("(a) Range = 0.2") +
-  labs(x = "X", y = "Y")
-
-s2 <- sdmTMB_simulate(
-  formula = ~1,
-  data = predictor_dat,
-  mesh = sim_mesh,
-  range = 0.6,
-  phi = 0.1,
-  sigma_O = 0.2,
-  seed = 1,
-  B = 0
-)
-sim_g2 <- ggplot(s2, aes(x, y, fill = mu)) +
-  geom_raster(show.legend = FALSE) +
-  scale_fill_viridis_c(option = "C") +
-  coord_fixed(expand = FALSE) +
-  theme_light() +
-  # theme(axis.title = element_blank()) +
-  ggtitle("(b) Range = 0.6") +
-  labs(x = "X", y = "Y")
-
-x <- seq(0, 1, length.out = 200)
-r <- seq(0.2, 1, 0.2)
-r <- c(0.2, 0.6)
-df <- data.frame(
-  x = rep(x, length(r)),
-  range = rep(r, each = length(x))
-)
-matern <- function(h, sigma = 1, kappa, nu = 1) {
-  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
-    ((kappa * abs(h))^nu) *
-    besselK(kappa * abs(h), nu)
-  ret
-  ret[x == 0] <- sigma^2
-  ret
-}
-blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
-df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
-sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
-  geom_line() +
-  theme_light() +
-  xlab("Distance") +
-  ylab("Correlation") +
-  labs(colour = "Range") +
-  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
-  scale_colour_manual(values = blues) +
-  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
-  scale_x_continuous(breaks = r) +
-  theme(legend.position = "inside", legend.position.inside = c(0.7, 0.7)) +
-  ggtitle("(c) Matérn correlation function")
-
-cowplot::plot_grid(sim_g1, sim_g2, sim_g3, align = "h", ncol = 3L)
-@
+Panel (c) shows the Mat\'ern function for these two range values.
+The dashed horizontal line shows a correlation of 0.13.}\label{fig:matern-range}
+\end{figure}
+\end{Schunk}
 
@@
 We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
-% TODO: add MVN(0, sigma)s
 
@@
 \centering
-\includegraphics[width=\textwidth]{../../figs/diagram}
+\includegraphics[width=\textwidth]{Figures/diagram}
 \caption[]{
@@
 \centering
-\includegraphics[width=3.4in]{flowchart}
+\includegraphics[width=3.4in]{Figures/flowchart}
 \caption[]{
@@
 \fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
-The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \added{\citep{Matrix}} as \added{an} interface to the \pkg{Eigen} library in \proglang{C++}.
-% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
-_Matrix: Sparse and Dense Matrix Classes and Methods_.
-R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack.
-If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al.
-(2010).
-“Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2024}.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using A6 \fct{stats::optimHess} \citep{gay1990, r2024}.
 Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
@@
 
-% \textbf{TODO: generalized delta method! and \citep{zheng2021}}
-
 A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
 \fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-\Rev{A7}\added{Other} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
+Other standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
 
@@
 A \fct{residuals} method calculates various types of residuals.
-\added{The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.}
+The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.
 New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
 \fct{sdmTMB\_cv} facilitates cross validation.
-
-% Derived quantities can be calculated on the fitted model.
-% TODO: more COG refs
-% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).
-
-% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
-
-% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB}, \fct{sdmTMB\_simulate} respectively).
-% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB}) or calculate residuals (\fct{residuals.sdmTMB}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
-% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).
-% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.
 
@@
 
-% TODO: describe this in more detail and how it could apply beyond ecology;
-% maybe a derived variables subsection? Expanded this sentence a bit -- COGs are
-% pretty common in other fields, but I don't know about indices -- removed
-% `ecology` from the description
-
-% \section{Model validation and selection}\label{model-validation-and-selection}
-
-% Validation and selection of state-space models is challenging, particularly when
-% using the Laplace approximation \citep{thygesen2017a}. We provide several
-% approaches to assist this process: (1) The (marginal) Akaike Information Criterion
-% \citep[AIC,][]{akaike1974} can be calculated from the marginal likelihood with \fct{AIC}, although AIC
-% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
-% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv} can be used
-% with user-specified or randomly chosen folds for model selection or to evaluate
-% goodness of fit according to user-calculated criteria (e.g., mean squared error,
-% area under the curve). (3) An \sdmTMB\ model can be passed to the
-% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
-% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
-% approximation, or perform posterior predictive checks (see
-% \code{?extract\_mcmc}). (4) The \fct{residuals} method by default returns
-% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
-% \citep{smith1985} residuals. For state-space models, these residuals have known
-% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
-% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
-% of the random effects to avoid this issue \citep{rufener2021} is recommended but
-% slower (\code{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
-% are also possible. (5) The \fct{simulate.sdmTMB} method can simulate from
-% fitted models and the \fct{sdmTMB\_simulate} function facilitates
-% simulating data without starting from a fitted model. Models can be fit to these
-% simulated data to ensure identifiability, evaluate bias and precision in
-% parameter estimation, or evaluate the consequences of model misspecification.
-
-% Jim:  If this section is re-added, you could also list the \fct{TMB::checkConsistency}
-% which evaluates the accuracy (and provides a bias-correction estimator) for the
-% Laplace approximation.
-
 \subsection{Installation}\label{installation}
@@
 
-<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
-install.packages("sdmTMB")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> install.packages("sdmTMB")
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<sdmTMB-install-suggest, eval=FALSE, echo=TRUE>>=
-install.packages("sdmTMB", dependencies = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> install.packages("sdmTMB", dependencies = TRUE)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<remotes, eval=FALSE, echo=TRUE>>=
-install.packages("pak")
-pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> install.packages("pak")
+R> pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\added{
 Users willing to replace the default \R\ \pkg{BLAS} (Basic Linear Algebra Subprograms) \citep{blas}
@@
 Suggestions are included in the package README file.
-}
 
@@
 
-<<libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE>>=
-library(sdmTMB)
-library(dplyr)
-library(ggplot2)
-@
-
-<<setoptions, echo=FALSE>>=
-options(
-  pillar.print_max = 3,
-  pillar.print_min = 3,
-  pillar.advice = FALSE,
-  pillar.width = 80
-)
-options(width = 80)
-@
-
-<<libs-extras, cache=FALSE, echo=FALSE>>=
-theme_set(theme_light())
-options(ggplot2.continuous.fill = "viridis")
-@
-
-<<pcod-head, echo=TRUE>>=
-select(pcod, lat, lon, X, Y, depth, present)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> library(sdmTMB)
+R> library(dplyr)
+R> library(ggplot2)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+R> select(pcod, lat, lon, X, Y, depth, present)
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 2,143 x 6
+    lat   lon     X     Y depth present
+  <dbl> <dbl> <dbl> <dbl> <dbl>   <dbl>
+1  52.3 -130.  446. 5793.   201       1
+2  52.3 -130.  446. 5800.   212       1
+3  52.4 -130.  449. 5802.   220       0
+# i 2,140 more rows
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<pcod-utms-eval, echo=TRUE, eval=FALSE>>=
-pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
-mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
+\end{Sinput}
+\end{Schunk}
 
@@
 Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
-\added{Mesh complexity has a large influence} on the speed of fitting these models.
-
-% https://haakonbakkagit.github.io/btopic104.html
-% https://becarioprecario.bitbucket.io/spde-gitbook/index.html
-
-<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points).
+Mesh complexity has a large influence on the speed of fitting these models.
+
+\begin{Schunk}
+\begin{Sinput}
+R> mesh_pcod2 <- make_mesh(
++    pcod,
++    xy_cols = c("X", "Y"),
++    fmesher_func = fmesher::fm_mesh_2d_inla,
++    cutoff = 8,
++    max.edge = c(10, 40),
++    offset = c(10, 40)
++  )
+R> plot(mesh_pcod2)
+\end{Sinput}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=3in]{Figures/dog-binomial-mesh2-1}
+
+}
+
+\caption[SPDE mesh (lines) combined with the trawl survey observations (points)]{SPDE mesh (lines) combined with the trawl survey observations (points).
 The locations where lines intersect are referred to as ``vertices'' or ``knots''.
 Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation, to a point.
-A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
-mesh_pcod2 <- make_mesh(
-  pcod,
-  xy_cols = c("X", "Y"),
-  fmesher_func = fmesher::fm_mesh_2d_inla,
-  cutoff = 8,
-  max.edge = c(10, 40),
-  offset = c(10, 40)
-)
-plot(mesh_pcod2)
-@
+A greater degree of control over the mesh construction can be achieved by using \pkg{fmesher} or \proglang{R}-\pkg{INLA} directly and supplying the object to \code{make\_mesh()}.}\label{fig:dog-binomial-mesh2}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<pcod-fit, echo=TRUE>>=
-fit_bin_rf <- sdmTMB(
-  present ~ poly(log(depth), 2),
-  data = pcod,
-  mesh = mesh_pcod2,
-  spatial = "on",
-  family = binomial(link = "logit")
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit_bin_rf <- sdmTMB(
++    present ~ poly(log(depth), 2),
++    data = pcod,
++    mesh = mesh_pcod2,
++    spatial = "on",
++    family = binomial(link = "logit")
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<pcod-fit-off, echo=TRUE>>=
-fit_bin <- update(fit_bin_rf, spatial = "off")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit_bin <- update(fit_bin_rf, spatial = "off")
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE>>=
-sanity(fit_bin_rf)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> sanity(fit_bin_rf)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<pcod-bin-summary, echo=TRUE>>=
-summary(fit_bin_rf)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> summary(fit_bin_rf)
+\end{Sinput}
+\begin{Soutput}
+Spatial model fit by ML ['sdmTMB']
+Formula: present ~ poly(log(depth), 2)
+Mesh: mesh_pcod2 (isotropic covariance)
+Data: pcod
+Family: binomial(link = 'logit')
+
+                     coef.est coef.se
+(Intercept)             -0.44    0.42
+poly(log(depth), 2)1   -75.72    9.30
+poly(log(depth), 2)2  -109.79    9.27
+
+Matérn range: 41.43
+Spatial SD: 1.69
+ML criterion at convergence: 1034.744
+
+See ?tidy.sdmTMB to extract these values as a data frame.
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<pcod-tidy, eval=TRUE, echo=TRUE, results='markup'>>=
-tidy(fit_bin_rf, conf.int = TRUE)
-tidy(fit_bin, conf.int = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> tidy(fit_bin_rf, conf.int = TRUE)
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 3 x 5
+  term                 estimate std.error conf.low conf.high
+  <chr>                   <dbl>     <dbl>    <dbl>     <dbl>
+1 (Intercept)            -0.439     0.424    -1.27     0.393
+2 poly(log(depth), 2)1  -75.7       9.30    -94.0    -57.5  
+3 poly(log(depth), 2)2 -110.        9.27   -128.     -91.6  
+\end{Soutput}
+\begin{Sinput}
+R> tidy(fit_bin, conf.int = TRUE)
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 3 x 5
+  term                 estimate std.error conf.low conf.high
+  <chr>                   <dbl>     <dbl>    <dbl>     <dbl>
+1 (Intercept)            -0.426    0.0573   -0.538    -0.314
+2 poly(log(depth), 2)1  -31.7      3.03    -37.6     -25.8  
+3 poly(log(depth), 2)2  -66.9      4.09    -74.9     -58.9  
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<pcod-eg1-tidy-re, eval=TRUE, echo=TRUE>>=
-tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 2 x 5
+  term    estimate std.error conf.low conf.high
+  <chr>      <dbl>     <dbl>    <dbl>     <dbl>
+1 range      41.4      9.38     26.6      64.6 
+2 sigma_O     1.69     0.233     1.29      2.22
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\added{
 We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
 Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
-\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimate (MLE) but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
-}
-
-<<morans, echo=TRUE>>=
-inv_dist_matrix <- 1 / as.matrix(dist(pcod[,c("X", "Y"), ]))
-diag(inv_dist_matrix) <- 0
-set.seed(1)
-r_bin <- residuals(fit_bin, type = "mle-mvn")
-set.seed(1)
-r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")
-ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
-ape::Moran.I(r_bin_rf, weight = inv_dist_matrix)$p.value
-@
-
-% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
-\added{
+We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimate (MLE) but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
+
+\begin{Schunk}
+\begin{Sinput}
+R> inv_dist_matrix <- 1 / as.matrix(dist(pcod[,c("X", "Y"), ]))
+R> diag(inv_dist_matrix) <- 0
+R> set.seed(1)
+R> r_bin <- residuals(fit_bin, type = "mle-mvn")
+R> set.seed(1)
+R> r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")
+R> ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
+\end{Sinput}
+\begin{Soutput}
+[1] 0
+\end{Soutput}
+\begin{Sinput}
+R> ape::Moran.I(r_bin_rf, weight = inv_dist_matrix)$p.value
+\end{Sinput}
+\begin{Soutput}
+[1] 0.8817066
+\end{Soutput}
+\end{Schunk}
+
 We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
 The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
-}
 
@@
 
-<<pcod-aic, eval=TRUE, echo=TRUE, results='markup'>>=
-AIC(fit_bin_rf, fit_bin)
-@
-
-\added{
-\Rev{A1}Caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
+\begin{Schunk}
+\begin{Sinput}
+R> AIC(fit_bin_rf, fit_bin)
+\end{Sinput}
+\begin{Soutput}
+           df      AIC
+fit_bin_rf  5 2079.487
+fit_bin     3 2392.070
+\end{Soutput}
+\end{Schunk}
+
+Caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
 In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
 Methods to estimate effective degrees of freedom for similar models were recently demonstrated in \citet{thorson2024cAIC}, but are not yet included in \sdmTMB.
-}
 
@@
 
-\added{
 As an alternative to AIC, we can conduct model comparison with cross validation.
 \sdmTMB\ includes the helper function \fct{sdmTMB\_cv} to facilitate this.
-}
 We will do 10-fold cross validation with the folds constructed randomly.
@@
 
-<<pcod-cv-future, eval=FALSE, echo=TRUE>>=
-library(future)
-plan(multisession)
-@
-
-<<pcod-cv, eval=TRUE, echo=TRUE>>=
-set.seed(12928)
-cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
-  data = pcod, mesh = mesh_pcod, spatial = "on",
-  family = binomial(), k_folds = 10
-)
-set.seed(12928)
-cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
-  data = pcod, mesh = mesh_pcod, spatial = "off",
-  family = binomial(), k_folds = 10
-)
-@
-
-\added{
+\begin{Schunk}
+\begin{Sinput}
+R> library(future)
+R> plan(multisession)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(12928)
+R> cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
++    data = pcod, mesh = mesh_pcod, spatial = "on",
++    family = binomial(), k_folds = 10
++  )
+R> set.seed(12928)
+R> cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
++    data = pcod, mesh = mesh_pcod, spatial = "off",
++    family = binomial(), k_folds = 10
++  )
+\end{Sinput}
+\end{Schunk}
+
 We can then calculate any performance metric of interest for comparison.
-\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017},
-}
-
-\added{
+A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017},
+
 $$
@@
 $$
-}
-
-\added{
+
 where $n$ represents the number of left-out data points, $\log \mathcal{L}$ is a symbol for log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
-}
-
-Indeed, the log likelihood predictive density \added{for the left-out data} is considerably \added{higher} for the model that includes random fields indicating better out-of-sample predictions:
-
-<<pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'>>=
-cv_bin_rf$sum_loglik
-cv_bin$sum_loglik
-@
-
-In practice, we would repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
+
+Indeed, the log likelihood predictive density for the left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:
+
+\begin{Schunk}
+\begin{Sinput}
+R> cv_bin_rf$sum_loglik
+\end{Sinput}
+\begin{Soutput}
+[1] -1006.435
+\end{Soutput}
+\begin{Sinput}
+R> cv_bin$sum_loglik
+\end{Sinput}
+\begin{Soutput}
+[1] -1195.931
+\end{Soutput}
+\end{Schunk}
+
+In practice, we would repeat this procedure several times to ensure the rank order is insensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 
-<<pcod-predict, echo=TRUE, results='markup'>>=
-p <- predict(fit_bin_rf, newdata = qcs_grid)
-select(p, X, Y, depth, est, est_non_rf, omega_s) |>
-  as_tibble() |> head(n = 2)
-@
-
-<<pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod.
+\begin{Schunk}
+\begin{Sinput}
+R> p <- predict(fit_bin_rf, newdata = qcs_grid)
+R> select(p, X, Y, depth, est, est_non_rf, omega_s) |>
++    as_tibble() |> head(n = 2)
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 2 x 6
+      X     Y depth    est est_non_rf omega_s
+  <dbl> <dbl> <dbl>  <dbl>      <dbl>   <dbl>
+1   456  5636  347. -6.52      -6.23   -0.291
+2   458  5636  223. -0.861     -0.600  -0.261
+\end{Soutput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=6.1in]{Figures/pcod-predict-maps-1}
+
+}
+
+\caption[Prediction components from the binomial species distribution model of Pacific Cod]{Prediction components from the binomial species distribution model of Pacific Cod.
 Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link (logit) space, and (c) the overall prediction, which here is the combination of panels a and b.
 The spatial random field represents spatially correlated latent effects not accounted for by the fixed effects.
-Note the difference between predictions from depth alone (a) and predictions including a spatial random field (c).">>=
-plot_spatial_map <- function(dat, column, title) {
-  ggplot(dat, aes(X, Y, fill = {{ column }})) +
-    geom_raster() +
-    coord_fixed() +
-    theme(legend.position= "bottom") +
-    ggtitle(title) +
-    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
-}
-g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
-g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
-g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
-cowplot::plot_grid(g1, g2, g3, ncol = 3)
-@
+Note the difference between predictions from depth alone (a) and predictions including a spatial random field (c).}\label{fig:pcod-predict-maps}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<dog-head-dat, echo=TRUE>>=
-dat <- select(dogfish, lon = longitude, lat = latitude, year,
-  catch_weight, area_swept, depth)
-dat
-@
-
-\Rev{A8}We add UTM zone 9 columns, create a log depth column for convenience, and create a basic mesh:
-
-<<dog-utms, echo=TRUE>>=
-dat <- add_utm_columns(dat, c("lon", "lat"),
-  units = "km", utm_crs = 32609)
-dat$log_depth <- log(dat$depth)
-mesh <- make_mesh(dat, xy_cols = c("X", "Y"), n_knots = 200)
-@
-
-<<dog-mesh2, eval=FALSE>>=
-plot(mesh)
-mesh$mesh$n
-@
+\begin{Schunk}
+\begin{Sinput}
+R> dat <- select(dogfish, lon = longitude, lat = latitude, year,
++    catch_weight, area_swept, depth)
+R> dat
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 1,458 x 6
+    lon   lat  year catch_weight area_swept depth
+  <dbl> <dbl> <int>        <dbl>      <dbl> <dbl>
+1 -125.  48.7  2004         22.7      0.103    73
+2 -126.  48.2  2004          0        0.103   455
+3 -126.  48.3  2004          0        0.116   171
+# i 1,455 more rows
+\end{Soutput}
+\end{Schunk}
+
+We add UTM zone 9 columns, create a log depth column for convenience, and create a basic mesh:
+
+\begin{Schunk}
+\begin{Sinput}
+R> dat <- add_utm_columns(dat, c("lon", "lat"),
++    units = "km", utm_crs = 32609)
+R> dat$log_depth <- log(dat$depth)
+R> mesh <- make_mesh(dat, xy_cols = c("X", "Y"), n_knots = 200)
+\end{Sinput}
+\end{Schunk}
 
@@
 where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$ and $\sigma^2_{\gamma}$ represents the marginal variance of this process.
-\added{We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.}
+We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.
 
@@
 
-<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
-fit_tw <- sdmTMB(
-  catch_weight ~ s(log_depth),
-  data = dat,
-  mesh = mesh,
-  family = tweedie(),
-  offset = log(dat$area_swept),
-  time = "year",
-  time_varying = ~ 1,
-  time_varying_type = "ar1",
-  spatial = "on",
-  spatiotemporal = "iid",
-  anisotropy = TRUE,
-  extra_time = seq(min(dat$year), max(dat$year)),
-  silent = FALSE
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit_tw <- sdmTMB(
++    catch_weight ~ s(log_depth),
++    data = dat,
++    mesh = mesh,
++    family = tweedie(),
++    offset = log(dat$area_swept),
++    time = "year",
++    time_varying = ~ 1,
++    time_varying_type = "ar1",
++    spatial = "on",
++    spatiotemporal = "iid",
++    anisotropy = TRUE,
++    extra_time = seq(min(dat$year), max(dat$year)),
++    silent = FALSE
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE>>=
-fit_dg <- update(fit_tw, family = delta_gamma())
-fit_dl <- update(fit_tw, family = delta_lognormal())
-fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))
-fit_dpl <- update(fit_tw, family = delta_lognormal(type = "poisson-link"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit_dg <- update(fit_tw, family = delta_gamma())
+R> fit_dl <- update(fit_tw, family = delta_lognormal())
+R> fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))
+R> fit_dpl <- update(fit_tw, family = delta_lognormal(type = "poisson-link"))
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<dog-aic, echo=TRUE>>=
-AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |>
-  mutate(delta_AIC = AIC - min(AIC)) |>
-  arrange(delta_AIC)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |>
++    mutate(delta_AIC = AIC - min(AIC)) |>
++    arrange(delta_AIC)
+\end{Sinput}
+\begin{Soutput}
+        df      AIC delta_AIC
+fit_dpl 19 11043.63   0.00000
+fit_dpg 19 11112.30  68.66264
+fit_dl  19 11133.52  89.89018
+fit_dg  19 11208.90 165.26222
+fit_tw  12 11356.57 312.93803
+\end{Soutput}
+\end{Schunk}
 
@@
 
-We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the spatiotemporal random fields as AR(1) to allow spatiotemporal \added{patterns} to \added{partially} persist from year to year.
+We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.
 
@@
 
-<<dog-ar1, results='hide', echo=TRUE>>=
-fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
-fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = "ar1")
-@
-
-<<dog-aci2, echo=TRUE>>=
-AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
+R> fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = "ar1")
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+R> AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
+\end{Sinput}
+\begin{Soutput}
+            df      AIC
+fit_dpl_ar1 21 11038.88
+fit_dpl     19 11043.63
+fit_dpl_iso 17 11069.41
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\added{
 As an example of estimation time for these complex spatiotemporal models, the Tweedie model (\code{fit\_tw}), Poisson-link delta-lognormal model (\code{fit\_dpl}), and Poisson-link delta-lognormal model with autogressive random fields (\code{fit\_dpl\_ar1}) took approximately 10, 30, and 90 seconds to fit, respectfully, on an Apple MacBook Pro with an M2 Pro processor and with Apple's \pkg{vecLib} implementation of \pkg{BLAS} in \R\ 4.4.0.
-}
-
-<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}.
+
+\begin{Schunk}
+\begin{Sinput}
+R> plot_anisotropy(fit_dpl_ar1)
+\end{Sinput}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=4in]{Figures/dog-aniso-1}
+
+}
+
+\caption[A visualization of anisotropy from the function \code{plot\_anisotropy()}]{A visualization of anisotropy from the function \code{plot\_anisotropy()}.
 Ellipses are centered at coordinates of zero in the units that the X-Y coordinates are modeled.
-The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
-plot_anisotropy(fit_dpl_ar1)
-@
-
-\Rev{A4}
+The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).}\label{fig:dog-aniso}
+\end{figure}
+\end{Schunk}
+
 \deleted{
@@
 
-<<dog-print, echo=TRUE>>=
-fit <- fit_dpl_ar1
-sanity(fit)
-summary(fit)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit <- fit_dpl_ar1
+R> sanity(fit)
+R> summary(fit)
+\end{Sinput}
+\begin{Soutput}
+Spatiotemporal model fit by ML ['sdmTMB']
+Formula: catch_weight ~ s(log_depth)
+Mesh: mesh (anisotropic covariance)
+Time column: year
+Data: dat
+Family: delta_lognormal(link1 = 'log', link2 = 'log', type = 'poisson-link')
+
+Delta/hurdle model 1: -----------------------------------
+Family: binomial(link = 'log') 
+            coef.est coef.se
+(Intercept)     2.35    0.53
+slog_depth      0.20    0.67
+
+Smooth terms:
+               Std. Dev.
+sds(log_depth)      3.23
+
+Time-varying parameters:
+                 coef.est coef.se
+(Intercept)-2004     0.34    0.61
+(Intercept)-2005     0.39    0.65
+(Intercept)-2006     0.45    0.68
+(Intercept)-2007     0.35    0.61
+(Intercept)-2008     0.26    0.55
+(Intercept)-2009     0.25    0.54
+(Intercept)-2010     0.25    0.52
+(Intercept)-2011     0.08    0.47
+(Intercept)-2012    -0.09    0.49
+(Intercept)-2013    -0.10    0.47
+(Intercept)-2014    -0.10    0.45
+(Intercept)-2015    -0.05    0.43
+(Intercept)-2016     0.01    0.43
+(Intercept)-2017    -0.01    0.45
+(Intercept)-2018    -0.04    0.44
+(Intercept)-2019    -0.15    0.47
+(Intercept)-2020    -0.26    0.53
+(Intercept)-2021    -0.38    0.60
+(Intercept)-2022    -0.30    0.53
+
+Spatiotemporal AR1 correlation (rho): 0.74
+Matérn anisotropic range (spatial): 17.5 to 183.6 at 130 deg.
+Spatial SD: 0.82
+Spatiotemporal marginal AR1 SD: 1.34
+
+Delta/hurdle model 2: -----------------------------------
+Family: lognormal(link = 'log') 
+            coef.est coef.se
+(Intercept)     2.20    0.20
+slog_depth      0.11    0.24
+
+Smooth terms:
+               Std. Dev.
+sds(log_depth)      0.59
+
+Time-varying parameters:
+                 coef.est coef.se
+(Intercept)-2004     0.43    0.27
+(Intercept)-2005     0.25    0.31
+(Intercept)-2006     0.13    0.24
+(Intercept)-2007     0.03    0.30
+(Intercept)-2008    -0.06    0.26
+(Intercept)-2009     0.20    0.30
+(Intercept)-2010     0.50    0.26
+(Intercept)-2011     0.30    0.30
+(Intercept)-2012     0.18    0.23
+(Intercept)-2013     0.10    0.29
+(Intercept)-2014     0.04    0.23
+(Intercept)-2015    -0.10    0.29
+(Intercept)-2016    -0.26    0.23
+(Intercept)-2017    -0.24    0.30
+(Intercept)-2018    -0.27    0.23
+(Intercept)-2019    -0.26    0.35
+(Intercept)-2020    -0.30    0.35
+(Intercept)-2021    -0.41    0.24
+(Intercept)-2022    -0.20    0.25
+
+Dispersion parameter: 1.10
+Spatiotemporal AR1 correlation (rho): 0.22
+Matérn anisotropic range (spatial): 4.1 to 43.3 at 130 deg.
+Spatial SD: 0.41
+Spatiotemporal marginal AR1 SD: 0.82
+
+ML criterion at convergence: 5498.439
+
+See ?tidy.sdmTMB to extract these values as a data frame.
+See ?plot_anisotropy to plot the anisotropic range.
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package.
-The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE, out.width="4in">>=
-set.seed(42)
-s <- simulate(fit, nsim = 500, type = "mle-mvn")
-dharma_residuals(s, fit, test_uniformity = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(42)
+R> s <- simulate(fit, nsim = 500, type = "mle-mvn")
+R> dharma_residuals(s, fit, test_uniformity = TRUE)
+\end{Sinput}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=4in]{Figures/dog-residuals-1}
+
+}
+
+\caption[Simulation-based randomized quantile residuals from the \pkg{DHARMa} package]{Simulation-based randomized quantile residuals from the \pkg{DHARMa} package.
+The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.}\label{fig:dog-residuals}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<dog-grid, echo=TRUE>>=
-grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
-grid$log_depth <- log(grid$depth)
-head(grid, n = 2)
-@
-
-<<dog-pred1, echo=TRUE>>=
-pred <- predict(fit, newdata = grid, type = "response")
-@
-
-<<dog-pred2, echo=TRUE>>=
-names(pred)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
+R> grid$log_depth <- log(grid$depth)
+R> head(grid, n = 2)
+\end{Sinput}
+\begin{Soutput}
+         X        Y depth year log_depth
+1 733.9852 5346.334   447 2004  6.102559
+2 735.9852 5346.334   442 2004  6.091310
+\end{Soutput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+R> pred <- predict(fit, newdata = grid, type = "response")
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+R> names(pred)
+\end{Sinput}
+\begin{Soutput}
+ [1] "X"           "Y"           "depth"       "year"        "log_depth"  
+ [6] "est1"        "est2"        "est_non_rf1" "est_non_rf2" "est_rf1"    
+[11] "est_rf2"     "omega_s1"    "omega_s2"    "epsilon_st1" "epsilon_st2"
+[16] "est"        
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<plot-map, echo=FALSE>>=
-plot_map <- function(dat, column) {
-  ggplot(dat, aes(X, Y, fill = {{ column }})) +
-    geom_raster() +
-    facet_wrap(vars(year)) +
-    coord_fixed() +
-    theme(legend.position= "bottom") +
-    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
-}
-@
-
-<<dog-plot1, echo=FALSE>>=
-g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |>
-  plot_map(est_non_rf1) +
-  scale_fill_viridis_c(trans = "log10") +
-  ggtitle("(a) Non-random-field components; 1st linear predictor")
-@
-
-<<dog-plot2, echo=FALSE>>=
-g_omega <- pred |> filter(year %in% c(2004, 2022)) |>
-  plot_map(omega_s1) +
-  scale_fill_gradient2() +
-  ggtitle("(b) Spatial random field; 1st linear predictor")
-@
-
-<<dog-plot3, echo=FALSE>>=
-g_eps <- pred |> filter(year %in% c(2004, 2022)) |>
-  plot_map(epsilon_st2) +
-  scale_fill_gradient2() +
-  ggtitle("(c) Spatiotemporal random field; 2nd linear predictor")
-@
-
-<<dog-plot4, echo=FALSE>>=
-g_est <- pred |> filter(year %in% c(2004, 2022)) |>
-  plot_map(est) +
-  # labs(fill = "Density") +
-  scale_fill_viridis_c(trans = "log10") +
-  ggtitle("(d) Overall prediction")
-@
-
-<<dog-wcvi-pred, fig.asp = 0.8, fig.cap= "Example prediction elements from the spatiotemporal model of Pacific Dogfish biomass density.
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=6in]{Figures/dog-wcvi-pred-1}
+
+}
+
+\caption[Example prediction elements from the spatiotemporal model of Pacific Dogfish biomass density]{Example prediction elements from the spatiotemporal model of Pacific Dogfish biomass density.
 Throughout, two example years are shown.
-(a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field elements (here, a smoother for bottom depth and the time-varying year effect) from the first linear predictor, (b) \\code{omega\\_s1} refers to the spatial random field from the first linear predictor, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second linear predictor, and (d) \\code{est} refers to the overall prediction estimate combining all effects.
+(a) \code{est\_non\_rf1} refers to the prediction from all non-random-field elements (here, a smoother for bottom depth and the time-varying year effect) from the first linear predictor, (b) \code{omega\_s1} refers to the spatial random field from the first linear predictor, (c) \code{epsilon\_st2} refers to spatiotemporal random fields from the second linear predictor, and (d) \code{est} refers to the overall prediction estimate combining all effects.
 The spatial random field is constant through time (i.e., the two panels in b are identical) and represents static biotic or abiotic features not included as covariates (e.g., habitat).
 The spatiotemporal random fields are different each time step and here are constrained to follow an AR(1) process.
-They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).", fig.width=9, out.width="6in">>=
-cowplot::plot_grid(
-  g_nonrf,
-  g_omega,
-  g_eps,
-  g_est,
-  ncol = 2L
-)
-@
+They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).}\label{fig:dog-wcvi-pred}
+\end{figure}
+\end{Schunk}
 
@@
 
-% TODO check if on CRAN yet!
-% TODO Jim:  It might be worth emphasizing that this calculation of marginal effects will differ from a partial dependence plot (PDP) when depth is correlated with other variables, and that package pdp could be used to additionally compute the PDP.
-Please tell me if you want me to suggest text or a code-demo, etc.
-
-<<dog-depth, echo=TRUE, out.width="5in">>=
-nd <- data.frame(
-  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100),
-  year = max(dat$year)
-)
-pred_depth <- predict(
-  fit, newdata = nd,
-  model = NA, re_form = NA, se_fit = TRUE
-)
-@
-
-<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish population density.
-The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively.
-Other fixed effects are held at constant values and the random fields are set to their expected value (zero).", out.width="4in">>=
-ggplot(pred_depth, aes(
-  exp(log_depth), exp(est),
-  ymin = exp(est - 2 * est_se),
-  ymax = exp(est + 2 * est_se))) +
-  geom_ribbon(fill = "grey90") +
-  geom_line() +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
-  labs(x = "Depth (m)", y = "Density")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> nd <- data.frame(
++    log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100),
++    year = max(dat$year)
++  )
+R> pred_depth <- predict(
++    fit, newdata = nd,
++    model = NA, re_form = NA, se_fit = TRUE
++  )
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=4in]{Figures/dog-depth-plot-1}
+
+}
+
+\caption[The conditional effect of ocean bottom depth on Pacific Spiny Dogfish population density]{The conditional effect of ocean bottom depth on Pacific Spiny Dogfish population density.
+The line and shaded ribbon represent the mean and 95\% confidence interval, respectively.
+Other fixed effects are held at constant values and the random fields are set to their expected value (zero).}\label{fig:dog-depth-plot}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<dog-index, echo=TRUE>>=
-grid$area <- 4 # 2 km x 2 km
-pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
-ind <- get_index(pred2, bias_correct = TRUE, area = grid$area)
-@
-
-<<dog-index-plot, echo=FALSE, fig.cap="Area-weighted index of relative biomass over time for Pacific Spiny Dogfish.
-Dots and line segments represent means and 95\\% confidence intervals.", out.width="4in">>=
-ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
-  geom_pointrange() +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
-  labs(y = "Biomass", x = "Year")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> grid$area <- 4 # 2 km x 2 km
+R> pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
+R> ind <- get_index(pred2, bias_correct = TRUE, area = grid$area)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=4in]{Figures/dog-index-plot-1}
+
+}
+
+\caption[Area-weighted index of relative biomass over time for Pacific Spiny Dogfish]{Area-weighted index of relative biomass over time for Pacific Spiny Dogfish.
+Dots and line segments represent means and 95\% confidence intervals.}\label{fig:dog-index-plot}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<owl-data>>=
-snow <- readRDS(here::here("data/SNOW_data.rds"))
-@
-
-<<owl-data-head, echo=TRUE>>=
-select(snow, X, Y, year, year_f, nao, count) |> head()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> select(snow, X, Y, year, year_f, nao, count) |> head()
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 6 x 6
+      X     Y  year year_f   nao count
+  <dbl> <dbl> <int> <fct>  <dbl> <dbl>
+1 -12.0  14.2  1979 1979    0.19     1
+2 -11.1  16.9  1979 1979    0.19     3
+3 -11.7  17.1  1979 1979    0.19     0
+# i 3 more rows
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<owl-fit, echo=TRUE, results = "hide">>=
-mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
-fit_owl <- sdmTMB(
-  count ~ 1 + nao + (1 | year_f),
-  spatial_varying = ~ 0 + nao,
-  time = "year",
-  data = snow,
-  mesh = mesh_snow,
-  family = nbinom2(link = "log"),
-  spatial = "on",
-  spatiotemporal = "iid",
-  silent = FALSE
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
+R> fit_owl <- sdmTMB(
++    count ~ 1 + nao + (1 | year_f),
++    spatial_varying = ~ 0 + nao,
++    time = "year",
++    data = snow,
++    mesh = mesh_snow,
++    family = nbinom2(link = "log"),
++    spatial = "on",
++    spatiotemporal = "iid",
++    silent = FALSE
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<owl-sanity, eval=FALSE>>=
-sanity(fit_owl)
-@
-
-<<owl-print, echo=TRUE>>=
-summary(fit_owl)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> summary(fit_owl)
+\end{Sinput}
+\begin{Soutput}
+Spatiotemporal model fit by ML ['sdmTMB']
+Formula: count ~ 1 + nao + (1 | year_f)
+Mesh: mesh_snow (isotropic covariance)
+Time column: year
+Data: snow
+Family: nbinom2(link = 'log')
+
+            coef.est coef.se
+(Intercept)    -3.23    5.44
+nao             0.36    0.22
+
+Random intercepts:
+       Std. Dev.
+year_f       0.3
+
+Dispersion parameter: 0.46
+Matérn range: 25.21
+Spatial SD: 6.97
+Spatially varying coefficient SD (nao): 0.17
+Spatiotemporal IID SD: 0.79
+ML criterion at convergence: 17512.366
+
+See ?tidy.sdmTMB to extract these values as a data frame.
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<owl-tidy, echo=TRUE>>=
-tidy(fit_owl, conf.int = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> tidy(fit_owl, conf.int = TRUE)
+\end{Sinput}
+\begin{Soutput}
+# A tibble: 2 x 5
+  term        estimate std.error conf.low conf.high
+  <chr>          <dbl>     <dbl>    <dbl>     <dbl>
+1 (Intercept)   -3.23      5.44  -13.9        7.42 
+2 nao            0.356     0.224  -0.0828     0.795
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<owl-p, message=FALSE, echo=FALSE, eval=FALSE, cache=TRUE>>=
-pred_snow <- predict(fit_owl, newdata = snow)
-pred_snow |>
-  select(X, Y, year, nao, count, est, omega_s, epsilon_st, zeta_s_nao) |>
-  head()
-@
-
-<<zeta-effect, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
-set.seed(42)
-zeta_s <- predict(fit_owl, newdata = snow, nsim = 300, sims_var = "zeta_s")
-dim(zeta_s)
-sims <- spread_sims(fit_owl, nsim = 300)
-dim(sims)
-combined <- sims$nao + t(zeta_s)
-snow$nao_effect <- exp(apply(combined, 2, median))
-snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.025))
-snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.975))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(42)
+R> zeta_s <- predict(fit_owl, newdata = snow, nsim = 300, sims_var = "zeta_s")
+R> dim(zeta_s)
+\end{Sinput}
+\begin{Soutput}
+[1] 30392   300
+\end{Soutput}
+\begin{Sinput}
+R> sims <- spread_sims(fit_owl, nsim = 300)
+R> dim(sims)
+\end{Sinput}
+\begin{Soutput}
+[1] 300   9
+\end{Soutput}
+\begin{Sinput}
+R> combined <- sims$nao + t(zeta_s)
+R> snow$nao_effect <- exp(apply(combined, 2, median))
+R> snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.025))
+R> snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.975))
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<owl-plot-basic, echo=TRUE, eval=FALSE>>=
-ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<shapes, echo=FALSE>>=
-if (!file.exists(here::here("data/ne_10m_lakes.shp"))) {
-  lakes <- rnaturalearth::ne_download(
-    scale = 10, type = "lakes", category = "physical",
-    destdir = here::here("data")
-  )
-}
-@
-
-<<shapes-read, echo=FALSE>>=
-Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
-coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") |>
-  sf::st_transform(crs = Albers)
-lakes <- sf::st_read(here::here("data/ne_10m_lakes.shp"), quiet = TRUE)
-lakes <- lakes[lakes$scalerank == 0, ] |> sf::st_transform(crs = Albers)
-land <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
-land <- land |> sf::st_transform(crs = Albers)
-@
-
-<<owl-proj2>>=
-snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
-snow2 <- snow2 |> mutate(x = X, y = Y) |>
-  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
-@
-
-<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
+\begin{Schunk}
+\begin{figure}[htbp]
+
+{\centering \includegraphics[width=4in]{Figures/owl-plot-fancy-1}
+
+}
+
+\caption[Spatially varying effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US]{Spatially varying effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US.
 The effect is multiplicative on owl count per NAO unit.
 In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.
-Point size is scaled to the mean counts in each location.", fig.width=5, out.width="4in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
-nao_effect_df <- snow2 |>
-  group_by(X, Y) |>
-  summarise(
-    count = mean(count), nao_effect = mean(nao_effect),
-    nao_effect_lwr = mean(nao_effect_lwr), nao_effect_upr = mean(nao_effect_upr),
-    .groups = "drop"
-  )
-
-snow_g1 <- ggplot(data = nao_effect_df) +
-  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
-  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
-  geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
-  geom_sf(data = coast, colour = "gray50", fill = NA, lwd = 0.35) +
-  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
-  coord_sf(
-    xlim = c(min(snow2$X), max(snow2$X)),
-    ylim = c(min(snow2$Y), max(snow2$Y))
-  ) +
-  scale_colour_viridis_c(
-    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
-    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
-  ) +
-  guides(size = "none") +
-  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect), 2), " to ", round(max(snow$nao_effect), 2))) +
-  labs(colour = "NAO effect\non Snowy Owl counts") +
-  theme_bw() +
-  theme(
-    legend.position = "inside",
-    legend.position.inside = c(0.25, 0.17),
-    legend.title = element_text(size = 9, hjust = 0),
-    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
-    panel.background = element_rect(fill = "grey90", colour = NA),
-    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
-    legend.background = element_rect(fill = "transparent", colour = NA),
-    legend.box.background = element_rect(fill = "transparent", colour = NA),
-    plot.title = element_text(size = 10, hjust = -0.15),
-    axis.title = element_blank()
-  )
-
-th <- theme(
-    legend.position = "none",
-    panel.background = element_rect(fill = "grey90", colour = NA),
-    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
-    legend.background = element_rect(fill = "transparent", colour = NA),
-    legend.box.background = element_rect(fill = "transparent", colour = NA),
-    plot.title = element_text(size = 10), axis.ticks = element_blank(),
-    axis.title = element_blank(), axis.text = element_blank()
-  )
-
-snow_g2 <- ggplot(data = nao_effect_df) +
-  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
-  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
-  geom_point(aes(X, Y, colour = nao_effect_lwr, size = count), alpha = 0.5) +
-  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
-  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
-  coord_sf(
-    xlim = c(min(snow2$X), max(snow2$X)),
-    ylim = c(min(snow2$Y), max(snow2$Y))
-  ) +
-  scale_colour_viridis_c(
-    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
-  ) +
-  ggtitle(paste0("(b) Lower 95% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
-  theme_bw() +
-  th
-
-snow_g3 <- ggplot(data = nao_effect_df) +
-  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
-  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
-  geom_point(aes(X, Y, colour = nao_effect_upr, size = count), alpha = 0.5) +
-  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
-  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
-  coord_sf(
-    xlim = c(min(snow2$X), max(snow2$X)),
-    ylim = c(min(snow2$Y), max(snow2$Y))
-  ) +
-  scale_colour_viridis_c(
-    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
-  ) +
-  ggtitle(paste0("(c) Upper 95% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
-  theme_bw() +
-  th
-
-bottom_row <- cowplot::plot_grid(snow_g2, snow_g3, label_size = 12)
-cowplot::plot_grid(snow_g1, bottom_row, ncol = 1, rel_heights = c(2.18, 1))
-@
+Point size is scaled to the mean counts in each location.}\label{fig:owl-plot-fancy}
+\end{figure}
+\end{Schunk}
 
@@
 
-We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
-Our test was restricted to one core and default \R\ algebra libraries \added{using \R\ 4.4.0 and \pkg{Matrix} version 1.7.0.
+We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, 10,000, or 100,000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
+Our test was restricted to one core and default \R\ algebra libraries using \R\ 4.4.0 and \pkg{Matrix} version 1.7.0.
 With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three- to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
@@
 At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
-}
 Speed increases can allow for more rapid and thorough model exploration and experimentation with this class of computationally intensive models.
@@
 \centering
-\includegraphics[width=\textwidth]{../../figs/timing-spatial-multipanel.pdf}
+\includegraphics[width=\textwidth]{Figures/timing-spatial-multipanel}
 \caption[]{
-Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to \added{three dataset sizes}.
+Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
 Lines represent means and ribbons 90\% quantiles across 50 random iterations.
 \pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
-\pkg{inlabru} \added{(version 2.10.1)} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
-\pkg{mgcv} \added{(version 1.9.1)} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
-Note that \pkg{spaMM} \added{(version 4.4.16)} only fits spatial, not spatiotemporal, models.
+\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
+\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
+Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
 All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
-\added{See Figure~\ref{fig:timing2} for a version with log-log axes.}
+See Figure~\ref{fig:timing2} for a version with log-log axes.
 }
@@
 Overall, \deleted{we think there is value in that} \sdmTMB\ unites functionality useful in many applied settings into a single package.
-
-% \textbf{TODO add paragraph here on uniting functionality not available in any one package?}
 
@@
 
-\bibliography{refs.bib}
+\bibliography{jss4956.bib}
 
@@
 \section{Speed testing related packages}\label{app:speed}
-
 Here, we describe the methods underlying the speed testing in Figure~\ref{fig:timing}.
@@
 
-We simulated 1,000, \added{10,000, or 100,000} spatial observations with both x and y coordinates from uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}).
+We simulated 1,000, 10,000, or 100,000 spatial observations with both x and y coordinates from uniform(0, 1) distributions (Figure~\ref{fig:sim-dat-plot-timing}).
 Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
@@
 \centering
-\includegraphics[width=\textwidth]{../../figs/timing-spatial-multipanel-log.pdf}
+\includegraphics[width=\textwidth]{Figures/timing-spatial-multipanel-log}
 \caption[]{
-  \added{
   Same as Figure~\ref{fig:timing} but with both axes log distributed.
-}
 }
@@
 
-% The following package versions were used:
-
-<<pkg-versions, echo=FALSE, eval=FALSE>>=
-packageVersion("INLA")
-packageVersion("inlabru")
-packageVersion("mgcv")
-packageVersion("spaMM")
-packageVersion("sdmTMB")
-@
-
 \clearpage
@@
 
-<<pkgs, warning=FALSE, message=FALSE, cache=FALSE, echo=FALSE>>=
-library("INLA")
-library("inlabru")
-library("ggplot2")
-library("sdmTMB")
-library("mgcv")
-library("spaMM")
-@
-
 We will illustrate with a \code{max.edge} of 0.06:
 
-<<mesh-timing, echo=TRUE>>=
-max_edge <- 0.06
-loc_bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
-segm_bnd <- INLA::inla.mesh.segment(loc_bnd)
-mesh <- INLA::inla.mesh.2d(
-  boundary = segm_bnd,
-  max.edge = c(max_edge, 0.2),
-  offset = c(0.1, 0.05)
-)
-@
-
-<<get-vertices, echo=FALSE>>=
-vertices <- mesh$n
-@
-
-This mesh has \Sexpr{vertices} (\code{mesh\$n}) vertices.
-
-<<mesh-vis-timing, fig.width=7, fig.asp=1.1, out.width="4.3in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
-max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
-g <- list()
-for (i in seq_along(max_edge_vec)) {
-  .mesh <- INLA::inla.mesh.2d(
-    boundary = segm_bnd,
-    max.edge = c(max_edge_vec[i], 0.2),
-    offset = c(0.1, 0.05)
-  )
-  g[[i]] <- ggplot() +
-    gg(.mesh) +
-    theme_light() +
-    coord_equal(expand = FALSE) +
-    ggtitle(paste("Mesh vertices =", .mesh$n))
-}
-cowplot::plot_grid(plotlist = g, ncol = 2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> max_edge <- 0.06
+R> loc_bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
+R> segm_bnd <- INLA::inla.mesh.segment(loc_bnd)
+R> mesh <- INLA::inla.mesh.2d(
++    boundary = segm_bnd,
++    max.edge = c(max_edge, 0.2),
++    offset = c(0.1, 0.05)
++  )
+\end{Sinput}
+\end{Schunk}
+
+This mesh has 1165 (\code{mesh\$n}) vertices.
+
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=4.3in]{Figures/mesh-vis-timing-1}
+
+}
+
+\caption[The meshes used in simulations from least to most vertices]{The meshes used in simulations from least to most vertices.}\label{fig:mesh-vis-timing}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<sim-timing, echo=TRUE>>=
-set.seed(123)
-n_obs <- 1000
-predictor_dat <- data.frame(
-  X = runif(n_obs), Y = runif(n_obs),
-  a1 = rnorm(n_obs)
-)
-mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)
-
-sim_dat <- sdmTMB_simulate(
-  formula = ~ 1 + a1,
-  data = predictor_dat,
-  mesh = mesh_sdmTMB,
-  family = gaussian(),
-  range = 0.5,
-  phi = 0.3,
-  sigma_O = 0.2,
-  B = c(0.2, -0.4) # B0 = intercept, B1 = a1 slope
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(123)
+R> n_obs <- 1000
+R> predictor_dat <- data.frame(
++    X = runif(n_obs), Y = runif(n_obs),
++    a1 = rnorm(n_obs)
++  )
+R> mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)
+R> 
+R> sim_dat <- sdmTMB_simulate(
++    formula = ~ 1 + a1,
++    data = predictor_dat,
++    mesh = mesh_sdmTMB,
++    family = gaussian(),
++    range = 0.5,
++    phi = 0.3,
++    sigma_O = 0.2,
++    B = c(0.2, -0.4) # B0 = intercept, B1 = a1 slope
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<sim-dat-plot-timing, out.width="4in", echo=FALSE, fig.cap="Example simulated dataset with a spatial random field.", fig.pos='ht', fig.align='center'>>=
-ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
-  geom_point() +
-  scale_color_gradient2() +
-  coord_fixed() +
-  theme_light()
-@
+\begin{Schunk}
+\begin{figure}[ht]
+
+{\centering \includegraphics[width=4in]{Figures/sim-dat-plot-timing-1}
+
+}
+
+\caption[Example simulated dataset with a spatial random field]{Example simulated dataset with a spatial random field.}\label{fig:sim-dat-plot-timing}
+\end{figure}
+\end{Schunk}
 
@@
 
-<<sdmTMBfit-timing, echo=TRUE>>=
-fit_sdmTMB <- sdmTMB(
-  observed ~ a1,
-  data = sim_dat,
-  mesh = mesh_sdmTMB,
-  family = gaussian(),
-  priors = sdmTMBpriors(
-    matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)
-  )
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fit_sdmTMB <- sdmTMB(
++    observed ~ a1,
++    data = sim_dat,
++    mesh = mesh_sdmTMB,
++    family = gaussian(),
++    priors = sdmTMBpriors(
++      matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)
++    )
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<spaMMfit-timing, echo=TRUE, cache=TRUE>>=
-spde <- INLA::inla.spde2.pcmatern(
-  mesh = mesh,
-  prior.range = c(0.05, 0.05),
-  prior.sigma = c(2, 0.05)
-)
-
-fit_spaMM <- fitme(
-  observed ~ a1 + IMRF(1 | X + Y, model = spde),
-  family = gaussian(),
-  data = sim_dat
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> spde <- INLA::inla.spde2.pcmatern(
++    mesh = mesh,
++    prior.range = c(0.05, 0.05),
++    prior.sigma = c(2, 0.05)
++  )
+R> 
+R> fit_spaMM <- fitme(
++    observed ~ a1 + IMRF(1 | X + Y, model = spde),
++    family = gaussian(),
++    data = sim_dat
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-% weird eval errors as of 2023-12-14 when in Rnw, hence eval=FALSE
-
-<<inlabrufit-timing, echo=TRUE, cache=TRUE, warning=FALSE, eval=FALSE>>=
-dat_sp <- sp::SpatialPointsDataFrame(
-  cbind(sim_dat$X, sim_dat$Y),
-  proj4string = sp::CRS(
-    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
-+ +y_0=0 +datum=NAD83 +units=km +no_defs"
-  ), data = sim_dat
-)
-components <- observed ~ -1 + Intercept(1) + a1 +
-  spatrf(main = coordinates, model = spde)
-like <- like(observed ~ Intercept + a1 + spatrf,
-  family = "gaussian", data = dat_sp
-)
-fit_bru <- bru(
-  like,
-  components = components,
-  options = bru_options(
-    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
-    bru_max_iter = 1, num.threads = "1:1"
-  )
-)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> dat_sp <- sp::SpatialPointsDataFrame(
++    cbind(sim_dat$X, sim_dat$Y),
++    proj4string = sp::CRS(
++      "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
++  + +y_0=0 +datum=NAD83 +units=km +no_defs"
++    ), data = sim_dat
++  )
+R> components <- observed ~ -1 + Intercept(1) + a1 +
++    spatrf(main = coordinates, model = spde)
+R> like <- like(observed ~ Intercept + a1 + spatrf,
++    family = "gaussian", data = dat_sp
++  )
+R> fit_bru <- bru(
++    like,
++    components = components,
++    options = bru_options(
++      control.inla = list(int.strategy = "eb", strategy = "gaussian"),
++      bru_max_iter = 1, num.threads = "1:1"
++    )
++  )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE>>=
-# -------------------------------------------------------------------------
-# Code in this chunk is from:
-#
-# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
-# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
-# https://doi.org/10.1007/s13253-019-00377-z
-#
-# Re-used here under a Creative Commons Attribution 4.0 International License
-# http://creativecommons.org/licenses/by/4.0/
-smooth.construct.spde.smooth.spec <- function(object, data, knots) {
-  dim <- length(object$term)
-  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
-  if (dim == 1) {
-    x <- data[[object$term]]
-  } else {
-    x <- matrix(0, nr = length(data[[1]]), nc = 2)
-    x[, 1] <- data[[object$term[1]]]
-    x[, 2] <- data[[object$term[2]]]
-  }
-  if (is.null(object$xt)) {
-    if (dim == 1) {
-      t <- seq(min(x), max(x), len = object$bs.dim)
-      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
-    } else {
-      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
-    }
-  } else {
-    if (!inherits(object$xt$mesh, "inla.mesh")) stop("xt must be NULL or an inla.mesh object")
-    mesh <- object$xt$mesh
-  }
-  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
-  inlamats <- INLA::inla.mesh.fem(mesh)
-  object$S <- list()
-  object$S[[1]] <- as.matrix(inlamats$c1)
-  object$S[[2]] <- 2 * as.matrix(inlamats$g1)
-  object$S[[3]] <- as.matrix(inlamats$g2)
-  object$L <- matrix(c(2, 2, 2, 4, 2, 0), ncol = 2)
-  object$rank <- rep(object$bs.dim, 3)
-  object$null.space.dim <- 0
-  object$mesh <- mesh
-  object$df <- ncol(object$X)
-  class(object) <- "spde.smooth"
-  return(object)
-}
-Predict.matrix.spde.smooth <- function(object, data) {
-  dim <- length(object$term)
-  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
-  if (dim == 1) {
-    x <- data[[object$term]]
-  } else {
-    x <- matrix(0, nr = length(data[[1]]), nc = 2)
-    x[, 1] <- data[[object$term[1]]]
-    x[, 2] <- data[[object$term[2]]]
-  }
-  Xp <- INLA::inla.spde.make.A(object$mesh, x)
-  return(as.matrix(Xp))
-}
-# End of code from Miller et al. 2020.
-# -------------------------------------------------------------------------
-@
-
 First define \fct{smooth.construct.spde.smooth.spec} and \fct{Predict.matrix.spde.smooth} from the supplement of \citet{miller2019}, then:
 
-<<mgcvfit-timing, echo=TRUE, eval=TRUE>>=
-class(mesh) <- "inla.mesh"
-fit_bam <- bam(
-  observed ~ a1 + s(X, Y,
-    bs = "spde", k = mesh$n,
-    xt = list(mesh = mesh)
-  ),
-  data = sim_dat,
-  family = gaussian(),
-  method = "fREML",
-  control = gam.control(scalePenalty = FALSE),
-  discrete = TRUE
-)
-@
-
-% \clearpage
-% \input{../../../sdmTMB-paper-response/response2}
+\begin{Schunk}
+\begin{Sinput}
+R> class(mesh) <- "inla.mesh"
+R> fit_bam <- bam(
++    observed ~ a1 + s(X, Y,
++      bs = "spde", k = mesh$n,
++      xt = list(mesh = mesh)
++    ),
++    data = sim_dat,
++    family = gaussian(),
++    method = "fREML",
++    control = gam.control(scalePenalty = FALSE),
++    discrete = TRUE
++  )
+\end{Sinput}
+\end{Schunk}
 

### 2025-06-13 doc/jss-accepted-version/tex/Code/replication-code-main.R

@@ -1,0 +1,811 @@
+## ----libraries, echo=FALSE, warning=FALSE, message=FALSE------------------
+library(ggplot2)
+library(dplyr)
+library(sdmTMB)
+
+# using a non-optimized BLAS library will be considerably slower
+is_optimized <- function() {
+  m <- 1e4
+  n <- 1e3
+  k <- 3e2
+  X <- matrix(rnorm(m * k), nrow = m)
+  Y <- matrix(rnorm(n * k), ncol = n)
+  s <- system.time(X %*% Y)
+  s[[3]] < 0.5
+}
+(optimized_blas <- is_optimized())
+if (!optimized_blas) {
+  warning("Detected a non-optimized BLAS library. We recommend installing an optimized BLAS library for considerably faster fitting of TMB/sdmTMB models. There are additional details in the 'Installation' section of the sdmTMB README.md file. https://github.com/pbs-assess/sdmtmb?tab=readme-ov-file#installation")
+}
+
+## ----check-version, echo=FALSE, cache=FALSE, eval=TRUE--------------------
+if (packageVersion("sdmTMB") < "0.6.0") {
+  stop("Please install a version of sdmTMB >= 0.6.0.", call. = FALSE)
+}
+
+
+## ----matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface. Panel (c) shows the Mat\\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13."----
+predictor_dat <- expand.grid(
+  x = seq(0, 1, length.out = 100),
+  y = seq(0, 1, length.out = 100),
+  year = seq_len(6)
+)
+sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
+s1 <- sdmTMB_simulate(
+  formula = ~1,
+  data = predictor_dat,
+  mesh = sim_mesh,
+  range = 0.2,
+  phi = 0.1,
+  sigma_O = 0.2,
[…]

### 2025-06-13 doc/jss-accepted-version/tex/Code/replication-code-timing.R

@@ -1,0 +1,445 @@
+# Timing figure -----------------------------------------------------------
+
+library("sdmTMB")
+library("dplyr")
+library("ggplot2")
+library("INLA")
+library("inlabru")
+library("mgcv")
+library("spaMM")
+# install.packages("RSpectra") # for faster spaMM
+
+dir.create("figs", showWarnings = FALSE)
+dir.create("analysis", showWarnings = FALSE)
+
+# using a non-optimized BLAS library will be considerably slower
+is_optimized <- function() {
+  m <- 1e4
+  n <- 1e3
+  k <- 3e2
+  X <- matrix(rnorm(m * k), nrow = m)
+  Y <- matrix(rnorm(n * k), ncol = n)
+  s <- system.time(X %*% Y)
+  s[[3]] < 0.5
+}
+(optimized_blas <- is_optimized())
+if (!optimized_blas) {
+  warning("Detected a non-optimized BLAS library. We recommend installing an optimized BLAS library for considerably faster fitting of TMB/sdmTMB models. There are additional details in the 'Installation' section of the sdmTMB README.md file. https://github.com/pbs-assess/sdmtmb?tab=readme-ov-file#installation")
+}
+
+# Set INLA to use 1 thread to match other package defaults:
+INLA::inla.setOption(num.threads = "1:1")
+
+# from remotes::install_github('seananderson/ggsidekick')
+theme_sleek <- function(base_size = 11, base_family = "") {
+  half_line <- base_size / 2
+  theme_light(base_size = base_size, base_family = base_family) +
+    theme(
+      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+      axis.ticks.length = grid::unit(half_line / 2.2, "pt"), strip.background = element_rect(
+        fill = NA,
+        colour = NA
+      ), strip.text.x = element_text(colour = "grey30"),
+      strip.text.y = element_text(colour = "grey30"), axis.text = element_text(colour = "grey30"),
+      axis.title = element_text(colour = "grey30"), legend.title = element_text(
+        colour = "grey30",
+        size = rel(0.9)
[…]

### 2025-06-25 doc/paper-jss/sdmTMB-paper.Rnw
prep submission script
Updates to work with latest INLA

@@
 
-% temp: track changes
+\usepackage{changes}
+% \usepackage[final]{changes}
 \usepackage{changes}
@@
   }
-  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
+
+  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, loc = x))
   inlamats <- INLA::inla.mesh.fem(mesh)
@@
 <<mgcvfit-timing, echo=TRUE, eval=TRUE>>=
-class(mesh) <- "inla.mesh"
 fit_bam <- bam(
@@
 
-% \clearpage
-% \input{../../../sdmTMB-paper-response/response2}
+\clearpage
+\input{../../../sdmTMB-paper-response/response2}
+\clearpage
 

### 2025-06-25 doc/jss-accepted-version/tex/Code/replication-code-timing.R

@@ -100,7 +100,7 @@ smooth.construct.spde.smooth.spec <- function(object, data, knots) {
     if (!inherits(object$xt$mesh, "inla.mesh")) stop("xt must be NULL or an inla.mesh object")
     mesh <- object$xt$mesh
   }
-  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
+  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, loc = x))
   inlamats <- INLA::inla.mesh.fem(mesh)
   object$S <- list()
   object$S[[1]] <- as.matrix(inlamats$c1)

### 2025-06-25 doc/paper-jss/sdmTMB-paper.Rnw
Don't include p-values for residuals

@@
 
-<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package.
-The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE, out.width="4in">>=
+<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package.", fig.asp=0.68, echo=TRUE, out.width="4in">>=
 set.seed(42)
 s <- simulate(fit, nsim = 500, type = "mle-mvn")
-dharma_residuals(s, fit, test_uniformity = TRUE)
+dharma_residuals(s, fit)
 @

### 2025-06-25 doc/jss-accepted-version/tex/Code/replication-code-main.R

@@ -324,7 +324,7 @@ summary(fit)
 ## ----dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package. The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE, out.width="4in"----
 set.seed(42)
 s <- simulate(fit, nsim = 500, type = "mle-mvn")
-dharma_residuals(s, fit, test_uniformity = TRUE)
+dharma_residuals(s, fit)
 
 
 ## ----dog-grid, echo=TRUE--------------------------------------------------

### 2025-06-25 doc/paper-jss/sdmTMB-paper.Rnw
Drop \added{} and \deleted{} and \Rev{}

@@
 Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
-We hope \sdmTMB \deleted{'s accessible interface} will help open this useful class of models to a wider field of geostatistical analysts.
+We hope \sdmTMB 's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
 }
@@
 \fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
-The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \added{\citep{Matrix}} as \added{an} interface to the \pkg{Eigen} library in \proglang{C++}.
+The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
 % TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023).
@@
 \fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
-\Rev{A7}\added{Other} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
+\Rev{A7}Other standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.
 
@@
 A \fct{residuals} method calculates various types of residuals.
-\added{The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.}
+The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.
 New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
@@
 install.packages("pak")
-pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
+pak::pak("pbs-assess/sdmTMB", dependencies = TRUE)
 @
@@
 Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
-\added{Mesh complexity has a large influence} on the speed of fitting these models.
+Mesh complexity has a large influence on the speed of fitting these models.
 
@@
 
-\added{
 We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
@@
 \Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimate (MLE) but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.
-}
 
@@
 % We omit example code for Moran's I calculation for brevity but include it in the code supplement.
-\added{
 We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
 The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
-}
 
@@
 
-\added{
 \Rev{A1}Caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
@@
 Methods to estimate effective degrees of freedom for similar models were recently demonstrated in \citet{thorson2024cAIC}, but are not yet included in \sdmTMB.
-}
 
@@
 
-\added{
 We can then calculate any performance metric of interest for comparison.
 \Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017},
-}
-
-\added{
+
 $$
@@
 $$
-}
-
-\added{
+
 where $n$ represents the number of left-out data points, $\log \mathcal{L}$ is a symbol for log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
-}
-
-Indeed, the log likelihood predictive density \added{for the left-out data} is considerably \added{higher} for the model that includes random fields indicating better out-of-sample predictions:
+
+Indeed, the log likelihood predictive density for the left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:
 
@@
 
-In practice, we would repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
+In practice, we would repeat this procedure several times to ensure the rank order is insensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.
 
@@
 
-\added{
 As an example of estimation time for these complex spatiotemporal models, the Tweedie model (\code{fit\_tw}), Poisson-link delta-lognormal model (\code{fit\_dpl}), and Poisson-link delta-lognormal model with autogressive random fields (\code{fit\_dpl\_ar1}) took approximately 10, 30, and 90 seconds to fit, respectfully, on an Apple MacBook Pro with an M2 Pro processor and with Apple's \pkg{vecLib} implementation of \pkg{BLAS} in \R\ 4.4.0.
-}
 
@@
 \Rev{A4}
-\deleted{
 On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
 We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.
-}
 
@@
 SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993, thorson2023SVC}.
-\deleted{In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.}
+In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.
 
@@
 This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.
-\deleted{
 However, if we recalculated the CI at a 95\% level (not shown) the full range of estimates for the lower CI would be below 1 meaning that at an alpha level of 0.05, we cannot conclude that the effect of NAO at an annual scale differs significantly from 1.
-}
 
@@
 \caption[]{
-Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to \added{three dataset sizes}.
+Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
 Lines represent means and ribbons 90\% quantiles across 50 random iterations.
 \pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
-\pkg{inlabru} \added{(version 2.10.1)} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
-\pkg{mgcv} \added{(version 1.9.1)} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
-Note that \pkg{spaMM} \added{(version 4.4.16)} only fits spatial, not spatiotemporal, models.
+\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
+\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
+Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
 All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
-\added{See Figure~\ref{fig:timing2} for a version with log-log axes.}
+See Figure~\ref{fig:timing2} for a version with log-log axes.
 }
@@
 Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
-Overall, \deleted{we think there is value in that} \sdmTMB\ unites functionality useful in many applied settings into a single package.
+Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
 
@@
 \caption[]{
-  \added{
   Same as Figure~\ref{fig:timing} but with both axes log distributed.
-}
 }

### 2025-06-25 doc/jss-accepted-version/tex/jss4956.tex
Switch to pak::pak
Update tex output with updated packages/text

@@
 
-In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
+In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts or slopes by grouping factors (Figure~\ref{fig:diagram}).
 All these components are additive in link space.
@@
 The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
-In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt} \citep{barrowman2000}, logistic functions \fct{logistic}, and penalized smoothers \fct{s} for generalized additive models, GAMS \citep{wood2017a}.
+In addition to standard main effects, the user can include random intercepts and slopes (e.g., \code{+ (1 + x | group)}), threshold shaped hockey-stick models \fct{breakpt} \citep{barrowman2000}, logistic functions \fct{logistic}, and penalized smoothers \fct{s} for generalized additive models, GAMS \citep{wood2017a}.
 Penalized smoothers use the same \fct{s} and \fct{t2} syntax as in \pkg{mgcv} \citep{wood2017a}.
@@
 The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
-The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using A6 \fct{stats::optimHess} \citep{gay1990, r2024}.
+The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2024}.
 Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
@@
 whether to condition on the random fields (\code{re\_form}),
-whether to condition on the random intercepts (\code{re\_form\_iid}),
+whether to condition on the random intercepts and slopes (\code{re\_form\_iid}),
 which delta model linear predictor to use (\code{model}),
@@
 R> install.packages("pak")
-R> pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
+R> pak::pak("pbs-assess/sdmTMB", dependencies = TRUE)
 \end{Sinput}
@@
 
+Conditional model:
                      coef.est coef.se
@@
 Family: binomial(link = 'log')
+Conditional model:
             coef.est coef.se
@@
 (Intercept)-2022    -0.30    0.53
+rho-(Intercept)      0.85    0.28
 
@@
 Family: lognormal(link = 'log')
+Conditional model:
             coef.est coef.se
@@
 (Intercept)-2022    -0.20    0.25
+rho-(Intercept)      0.62    0.40
 
@@
 R> s <- simulate(fit, nsim = 500, type = "mle-mvn")
-R> dharma_residuals(s, fit, test_uniformity = TRUE)
+R> dharma_residuals(s, fit)
 \end{Sinput}
@@
 
-\caption[Simulation-based randomized quantile residuals from the \pkg{DHARMa} package]{Simulation-based randomized quantile residuals from the \pkg{DHARMa} package.
-The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.}\label{fig:dog-residuals}
+\caption[Simulation-based randomized quantile residuals from the \pkg{DHARMa} package]{Simulation-based randomized quantile residuals from the \pkg{DHARMa} package.}\label{fig:dog-residuals}
 \end{figure}
@@
 
+Random intercepts and/or slopes:
+
+Conditional model:
+     Groups        Name    Variance    Std.Dev.
+     year_f (Intercept)        0.09        0.30
+
+Conditional model:
             coef.est coef.se
@@
 nao             0.36    0.22
-
-Random intercepts:
-       Std. Dev.
-year_f       0.3
 
@@
 (1) the ability to specify observation-specific likelihoods to integrate different data types \citep[e.g.,][]{gruss2019multiple},
-(2) inclusion of a dispersion formula similar to \pkg{glmmTMB},
-(3) inclusion of random slopes to complement the existing random intercepts,
-and
-(4) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base is written in \R\ rather than \proglang{C++}.
+(2) inclusion of a dispersion formula similar to \pkg{glmmTMB}, and
+(3) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base is written in \R\ rather than \proglang{C++}.
 % (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.

### 2025-06-25 doc/jss-accepted-version/tex/jss4956.tex
Trivial whitespace changes etc. in .tex output

@@
 
-% \usepackage{changes}
-\usepackage[final]{changes}
 \usepackage{changes}
@@
 
+% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
 We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
@@
 
-\deleted{
 On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
 We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.
-}
 
@@
 \centering
-\includegraphics[width=\textwidth]{Figures/timing-spatial-multipanel}
+\includegraphics[width=\textwidth]{Figures/timing-spatial-multipanel.pdf}
 \caption[]{
@@
 \section{Speed testing related packages}\label{app:speed}
+
 Here, we describe the methods underlying the speed testing in Figure~\ref{fig:timing}.
